### 2016-05-24 part1/lafaySimulatedScenePerception1.tex
initial commit

% Instructions for Authors fuer Acta Acustica

\documentclass[twoside,twocolumn]{article}
\usepackage{acta}
\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage{setspace}

\begin{document}
\Eingang{00}{00}{0000} \Annahme{00}{00}{0000} \Sachgebiet{1}
\PACS{43.50.Lj, 43.50.Rq, 43.50.Qp, 43.66.Lj, 43.66.Ba}
\Band{0} \Jahr{0000} \Heft{} \Ersteseite{1} \Letzteseite{}

\AuthorsI{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J.F. Petiot$^{1)}$}
\AddressI{$^{1)}$ Institut de Recherche en Communications et  Cybern\'etique de Nantes (IRCCYN), Nantes, France.\\
\hspace*{8pt}gregoire.lafay@irccyn.ec-nantes.fr\\
$^{2)}$ Institut de Recherche et Coordination Acoustique/Musique (IRCAM), Paris, France}

% \footnotemark[1] \Authornotes{\footnotetext[1]{} }

\newcommand{\gl}[1]{\textcolor{blue}{GL : #1}}
\newcommand{\todo}[1]{\textcolor{red}{todo : #1}}
\newcommand{\ml}[1]{\textcolor{magenta}{ML : #1}}
\newcommand{\mr}[1]{\textcolor{green}{MR : #1}}

\Englishtitle{Investigating soundscapes perception through acoustic scenes simulation. Part I: simulation protocol presentation and case study} \Germanorfrenchtitle{}

\Kolumnentitel{Lafay et al. : Approaching mental representations}

\Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process. Subjects are asked to create a full soundscape by means of a  software dedicated to sound arrangement, coupled with a structured sound data set. This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes. A rating experiment of the soundscape pleasantness using a 7-points bipolar scale is conducted to further refine the analysis of the simulated urban acoustic scenes. Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.}
\Germanorfrenchabstract{}

\ScientificPaper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%   INTRODUCTION      %%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\onecolumn
\setlength{\parindent}{5ex}
%\setstretch{2}
\section{Introduction}

%% Original intro

%% The notion of soundscape has been introduced by Schafer \cite{schafer_new_1969, schafer1977tuning} in the 1970s as the auditory equivalent to landscape. Following this paradigm, a sonic environment is described by focusing on the listener's evaluation, rather than only taking into account the acoustic parameters of the sound.

%% Only during the 1980s did policy-makers started taking into account the link between \textit{noise} and pollution, considering noise as a significant degradation of the quality of life. To fight this phenomenon, the first approach consisted in identifying unwanted sounds, and lowering their intensities. Thus, in the wake of this realization, several regulations have emerged, often limited to enforcing sound level thresholds.

%% However, it is now widely accepted that acoustic comfort is a complex concept that cannot be described solely with the help of objective acoustical measures such as $LA_{(eq,V)}$, but needs to take into account other perceptive factors \cite{schulte-fortkamp_soundscape_2006, aletta2016soundscape, Yang2005211, kang_semantic_2010}. ``\,Noise\,''  is a cognitive object which depends upon a listener's appreciation and the context in which it is heard. Many urban ``\,noises\,'', such as that of a siren, can annoy as well as warn of a danger, and many town districts are appreciated because of a lively and animated atmosphere which often results in higher sound levels. Conventional noise regulations concentrate on high sound levels from transportation and industry, and their scope does not extend beyond unwanted sound sources. In other words, they do not try to improve the sonic environment by identifying and reinforcing pleasant or important sounds. This latter approach was pioneered by Schafer, who named it the ``\,positive\,'' approach.

%% For two decades, the soundscape paradigm has been considered to study the perception of acoustic scenes in order to provide urban planners with meaningful sound quality indicators (see \cite{dubois2006cognitive} and  \cite{aletta2016soundscape} for extensive reviews). Most studies follow the \emph{modus operandi} of cognitive psychology \cite{maffiolo_caracterisation_1999}, by first describing the soundscape using both acoustic measurements and perceptive factors, and then investigating the interactions between those descriptors to understand the cognitive processes involved in the perception of the acoustic scenes. Soundscape studies are thus interdisciplinary \cite{davies_perception_2013, aletta2016soundscape}, borrowing tools from various fields such as acoustics, psycho-physics or cognitive psychology, but also psycho-linguistics, sociology and data mining.

%% The soundscape approach appears to be a powerful tool to develop perceptively motivated indicators \cite{schulte-fortkamp_soundscape:_2007}. That said, the use of various experimental protocols to assess and measure soundscapes makes the integration of results difficult \cite{davies_perception_2013}. Also, there still exist no definite consensus in the soundscape community about the acoustic descriptors to be used \cite{aletta2016soundscape}. This issue prevents experimenters to provide  urban planners with generic and clear indicators or models able to predict sound environment quality \cite{schulte-fortkamp_soundscape:_2007, aletta2016soundscape}. Recently, several projects such as the European Cooperation in Science and Technology Action (TD0804: soundscape of European Cities and Landscapes) or the Positive Soundscape project \cite{davies_perception_2013} have been undertaken to standardize soundscape assessment and descriptors \cite{schulte2013soundscape}, but those matters are still considered as open problems.

%% We believe that soundscapes studies greatly suffer from one important issue: to the best of our knowledge, there is no morphological definition of a soundscape, and therefore no standardized way to decompose it in distinct elements. It is thus difficult to study precisely which component of the environment is the cause of potential variations in the perceptive measurements.

%% We propose to tackle this issue by introducing a new experimental protocol, in which  subjects are asked to create a complex sonic environment thanks to a dedicated software tool linked to a dataset of isolated environmental sound samples. The simulation provides us with simulated scenes fully annotated in terms of source presence and sound levels. Those annotations enable us to ground the analyses of further perceptive descriptors obtained via standard perceptive evaluations of the simulated scenes, as well as to study the contribution of each sound source.

%% As a case study, we focus in this paper on urban soundscape pleasantness. We first give an overview of the tools used in soundscape perception studies, and justify the benefits of conducting this kind of simulation experiment as a data generation process. We then describe the simulation process, and present the results of the case study. The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale.

%% Results show that 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding,  2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterise the appraisal than global descriptors.

%% Version MR 19/05/2016

The study of our sonic environments, both natural and man-made, has gathered growing interest in the past decades, finding applications in fields as varied as context inference and surveillance \cite{heittola13,1540194,park14}, evaluation of urban noise pollution \cite{raimbault_urban_2005} or the emerging area of eco-acoustics \cite{ECOACOUSTICS2014, krause}. The notion of \emph{soundscape}, introduced by Schafer \cite{schafer_new_1969, schafer1977tuning} in the 1970s as the auditory equivalent to landscape, is a theoretical framework of choice for that research. Following this paradigm, a sonic environment is described by focusing on the listener's evaluation, rather than only taking into account the acoustic parameters of the sound scene.

Most soundscape studies follow the \emph{modus operandi} of cognitive psychology \cite{maffiolo_caracterisation_1999}, by first describing the soundscape using both acoustic measurements and perceptive factors, and then investigating the interactions between those descriptors to understand the cognitive processes involved in the perception of the acoustic scenes. Soundscape studies are thus interdisciplinary \cite{davies_perception_2013, aletta2016soundscape}, borrowing tools from various fields such as acoustics, psycho-physics or cognitive psychology, but also psycho-linguistics, sociology and data mining.

While the soundscape approach appears to be a powerful tool to develop perceptively motivated indicators \cite{schulte-fortkamp_soundscape:_2007}, the use of various experimental protocols to assess and measure soundscapes makes the integration of results difficult \cite{davies_perception_2013}. Moreover, there still exist no definite consensus in the soundscape community about the acoustic descriptors to be used \cite{aletta2016soundscape}. We believe that soundscapes studies greatly suffer from one important issue: to the best of our knowledge, there is no morphological definition of a soundscape, and therefore no standardized way to decompose it into distinct elements. It is thus difficult to study precisely which component of the environment is the cause of potential variations in the perceptive measurements.

We propose to tackle this issue by introducing a new experimental protocol, in which subjects are asked to create a complex sonic environment thanks to a dedicated software tool linked to a dataset of isolated environmental sound samples. The simulation provides us with simulated scenes fully annotated in terms of source presence and sound levels. Those annotations enable us to ground the analyses of further perceptive descriptors obtained via standard perceptive evaluations of the simulated scenes, as well as to study the contribution of each sound source.

As a case study to demonstrate the interest of that new approach, we focus in this paper on the issue of urban soundscape pleasantness as tackled in \cite{guastavino_ideal_2006}.

Only during the 1980s did policy-makers start taking into account the link between \textit{noise} and pollution, considering noise as a significant degradation of the quality of life. To fight this phenomenon, the first approach consisted in identifying unwanted sounds, and lowering their intensities. Thus, in the wake of this realization, several regulations have emerged, often limited to enforcing sound level thresholds.

However, it is now widely accepted that acoustic comfort is a complex concept that cannot be described solely with the help of objective acoustical measures such as $LA_{(eq,V)}$, but needs to take into account other perceptive factors \cite{schulte-fortkamp_soundscape_2006, aletta2016soundscape, Yang2005211, kang_semantic_2010}. ``\,Noise\,''  is a cognitive object which depends upon a listener's appreciation and the context in which it is heard. Many urban ``\,noises\,'', such as that of a siren, can annoy as well as warn of a danger, and many town districts are appreciated because of a lively and animated atmosphere which often results in higher sound levels. Conventional noise regulations concentrate on high sound levels from transportation and industry, and their scope does not extend beyond unwanted sound sources. In other words, they do not try to improve the sonic environment by identifying and reinforcing pleasant or important sounds. This latter approach was pioneered by Schafer, who named it the ``\,positive\,'' approach.

The soundscape paradigm has been employed for two decades to study the perception of urban acoustic scenes (see \cite{dubois2006cognitive} and  \cite{aletta2016soundscape} for extensive reviews), however, due to the limitations we have pointed out, experiments have failed to provide  urban planners with generic and clear indicators or models able to predict sound environment quality \cite{schulte-fortkamp_soundscape:_2007, aletta2016soundscape}. Recently, several projects such as the European Cooperation in Science and Technology Action (TD0804: soundscape of European Cities and Landscapes) or the Positive Soundscape project \cite{davies_perception_2013} have been undertaken to standardize soundscape assessment and descriptors \cite{schulte2013soundscape}, but those matters are still considered as open problems. We explore in this paper what additional insights our proposed ``\,generative\,'' approach to soundscape perception analysis can bring to light.

We first give an overview of the tools used in soundscape perception studies, and justify the benefits of conducting this kind of simulation experiment as a data generation process. We then describe the simulation process, and present the results of the case study. The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale.

Results show that 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding, 2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterise the appraisal than global descriptors.

\section{Background and motivations}
\label{statoftheart}

Two approaches, named in this paper \textit{categorical} and \textit{dimensional}, are commonly used to gain knowledge on soundscape perception.

The \textit{categorical} approach investigates both the existence of 1) soundscape categories and 2) soundscape components (categories of sound sources) using both sorting \cite{maffiolo_caracterisation_1999, guastavino_categorization_2007} and descriptive \cite{axelsson2005soundscape, raimbault_urban_2005, raimbault2006qualitative, guastavino_ideal_2006} tasks (questionnaires or interviews). To identify categories, those methods usually require psycho-linguistic and lexical analyses. Their results tend to show that soundscape perception relies upon the identification and assessment of the sound sources which compose the soundscape \cite{dubois2006cognitive}. Maffiolo \cite{maffiolo_caracterisation_1999} shows the existence of two generic soundscape categories, namely the ``\,event sequences\,'' and the ``\,amorphous sequences\,''. Similar findings were made by \cite{guastavino_ideal_2006}. Guastavino \cite{guastavino_categorization_2007} as well as Niessen \emph{et al.} \cite{niessen_categories_2010}  show the prominence of specific sound source categories in soundscape perception: ``\,natural\,'', ``\,human\,'', ``\,traffic\,'' and ``\,mechanical\,''. Other studies not using the \textit{categorical} approach reveal that sound source identification underlines the perception of others factors, for example Kuwano \emph{et al.} \cite{kuwano_memory_2003-1} show that overall judgment of loudness of a sound environment sequence is not statistically different from average instantaneous judgment of the recalled sound events of the sequence.  As reported by \cite{davies_perception_2013}, while the categorical approach does help to understand how a soundscape is composed, it does not provide information on how soundscape components are perceived and influence the qualitative evaluation.

Instead of asking subjects to break soundscapes down into categories, the \textit{dimensional} approach makes use of bipolar semantic scales to describe the soundscape either in a holistic way \cite{jeon2013soundwalk, kang_semantic_2010, cain_development_2013, torija2013application} or by considering the influence of specific sound sources \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}. From those descriptors, it is possible to use dimensional analysis techniques such as Principal Component Analysis to find relevant perceptual axes \cite{cain_development_2013}, or to connect those perceptive outcomes with objective acoustics measurements using various statistical analysis or modelling techniques such as multiple linear regression \cite{lavandier2006contribution,ricciardi2015sound} or artificial neural network \cite{yu2009modeling}. Several semantic scales show promising results when it commes to modeling perceived affective quality, namely ``\,Calmness\,'' and ``\,Vibrancy\,'' \cite{cain_development_2013, axelsson2010principal}, or ``\,Pleasantness\,'' and ``\,Eventfulness\,'' \cite{axelsson2010principal}; but it has been shown that others context dependent (``\,Appropriateness\,'',``\,Familiarity\,'') or music related (``\,Music-likeness\,'' ) scales provide useful complementary information on soundscape perception \cite{aletta2016soundscape}. In addition, \cite{yu2009modeling} shows that multiple factors such as physical, behavioral, social, demographical, and psychological are involved when evaluating acoustic comfort.  Similar findings are made in \cite{yu2010factors} showing that age and education level significantly influence the sound preference.

Studies investigating the influence of sound sources in soundscape perception show that global sound level (perceived or measured) cannot entirely explain the quality of a sonic environment. \cite{guyot2005urban} shows that a rise in soundscape quality between two street soundscapes having a similar time-ratio of traffic sounds and a similar global sound level may result from the addition of a positively connotated sound source, e.g. belonging to the ``\,human\,'' category. Using multiple linear regression to predict unpleasantness from descriptors characterizing global sound level as well as time presence and sound level of each source, \cite{lavandier2006contribution} shows that the percentage of explained variance increases when the source descriptors are added to the global sound level. Similar findings are made in \cite{ricciardi2015sound}, showing that a combined use of global and source specific descriptors allows to build more relevant model of sound quality. Lastly, \cite{lavandier2006contribution} shows that global sound level is not always appropriate to characterize the appraisal, especially in the case of positively rated soundscapes such as \textit{park} or \textit{market}.

In order to be able to study the relative perceptual importance of each sound source, it is beneficial to first gain knowledge about the presence and level of those sound sources within the scene. Acquiring such data may be a serious challenge. From most of the studies cited above \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}, subjects are asked to evaluate the time presence and/or the sound level of specific sound source categories using bipolar scales. To the best of our knowledge, only \cite{lavandier2006contribution} relies on objective measurement of those indicators by manually annotating the recordings used as stimuli. This annotation step is hardly manageable for \emph{in-situ} studies such as \cite{ricciardi2015sound} involving a large number of participants. Furthermore, only a pre-imposed restricted set of sound categories are evaluated, preventing experimenters to investigate the influence of more precise or alternative organisations of the sound sources.

There is thus a need to 1) ground the study into a precise taxonomy of sound sources, and 2) generate stimuli that are precisely annotated in terms of sound source presence and level. We propose to tackle this issue through the use of a simulation protocol in which subjects are asked to simulate specific complex sound environments thanks to a soundscape simulator linked to a dataset of isolated environmental sound samples. This approach leads us to consider the soundscape as a sequence of sound sources \cite{nelken_ear_2013}, following a morphological approach of great potential coined as a ``\,skeleton of events on a bed of texture\,''. Experimental subjects are able to create a sound environment by controlling separately each of the sources that make it up. Consequently, detailed post-analysis measurement of the impact of each source on the global perception of the simulated sound environment can be made.

Previous works conducted by Bruce \textit{et al.} also use a soundscape simulator to study soundscape perception \cite{bruce2009development,bruce2014effects}. However, in their approach, subjects may only add or remove a reduced set of long duration recordings of foreground sounds, and adjust their levels. In our approach, subjects are presented with a full dataset of classes of isolated sound samples, that they may finely manipulate in terms of sound intensity, repetition and time positioning. Important questions about the representativity of the  dataset are studied in Sections~\ref{sec:SoundDataset} and  \ref{sec:datasetAnalyses}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%   PROTOCOL      %%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Simulation protocol}

\subsection{Sound dataset}
\label{sec:SoundDataset}

The proposed protocol can be seen as the inverse of that of a description task (see Figure~\ref{psycholing}). In contrast with interviews and questionnaires, which both require subjects to describe a soundscape, \textit{i.e.} to decompose it into elements, subjects are here asked to recompose the soundscape from a sound dataset of urban environmental sounds.

To simulate a verisimilar sound environment, we need to provide the subject with a reasonable and representative number of isolated sounds. This dataset represents the ``\,sonic world\,'', or more precisely a semantic discretization of it in terms of  ``\,sound sources categories\,''. The design of the dataset is largely inspired by early work by Rosch \cite{roach1978cognition} on conceptual categorization which states that isolated sound classification operates at several levels of abstraction \cite{houix_lexical_2012}. Following this consideration, the dataset is structured along two dimensions:

\begin{itemize}
\item object dimension: where we make the difference between distinct sounds sources such as \textit{voice}, or \textit{car}. 
\item abstraction dimension: where we make the difference between distinct level of abstraction use to describe a sound (``\,human\,''>``\,voice\,''>``\,laugh\,''). 
\end{itemize}

Figure~\ref{datasetexample} shows a subset of the described hierarchical organization. The top classes our taxonomy represent concepts such as ``\,urban transport\,'', encompassing classes of sound sources such as ``\,boat\,'' or  ``\,cars\,'', which are themselves split into sub-classes. At the leaf level, classes are described using sound source names and actions (\textit{car passing, male-voice yelling}). The deeper the class level, the lower the variability between the exemplars lying under the class. In this paper the class levels will be referred to as \emph{semantic levels}, the top classes being at semantic level 0, the first subclasses at semantic level 1 and so on. Leaf classes are collections of recorded isolated sounds.

The dataset taxonomy being the same for each subject, it acts as a reference for the analysis, allowing us to get a non ambiguous identification of the sounds populating a scene. Most soundscape studies use recordings or \textit{in situ} sound environment as stimuli. In both cases, these stimuli remain specific exemplars of many potential soundscapes which could occur at the same location and the same time. The re-composition process frees experimenters from this bias as the simulated soundscape is directly related to the subject's mental representation. While expressivity in this approach is still limited by the diversity of the available sound dataset, we nonetheless believe that it allows the subjects' responses to be less constrained than by an imposed soundscape exemplar, provided that the sound dataset is carefully designed.

A soundscape  is here understood as being  ``\,a skeleton of events on a bed of texture\,'' \cite{nelken_ear_2013}, where sound events are short isolated sounds such as \textit{car passing} or \textit{baby crying}, whereas sound textures are long stationary  sounds such as \textit{traffic} or \textit{crowd sound}, in which the physical characteristics remain stable over long periods of time~\cite{saint-arnaud_classification_1995}. Several studies point out the fact that textures and events drive two distinct cognitive processes \cite{maffiolo_caracterisation_1999, mcdermott2013summary}. We thus clearly distinguish between ``\,sound events\,'' and ``\,sound textures\,'' and use two distinct datasets.

For ease of reading, we introduce some terms that will be used throughout the paper:

\begin{description}
\item[Sample]: a recording of an isolated sound event or texture.
\item[Sound class]: either a collection of samples or a collection of sub-classes, depending of the semantic level of the class.
\item[Track]: a temporal sequence of samples belonging to the same sound class. As samples of a texture track are concatenated seamlessly (see section \ref{sec:simulationProcess}), a texture track can be seen as being made of one sample.
\end{description}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.4\paperwidth]{../gfx/1.eps}
  \caption{\label{psycholing} Psycho-linguistic analysis paradigm (top) and proposed approach (bottom).}
  \end{center}
\end{figure}

\begin{figure}[t!]
\begin{center}
\includegraphics[width=.4\paperwidth]{../gfx/3.eps}
\caption{\label{datasetexample} Hierarchical structure of the event and texture data sets. Depending on the considered top class, there could be more or less than 3 semantic levels.}
\end{center}
\end{figure}

\subsection{Sample Selection process}

The sample selection process is designed to rely only on the listening of sounds themselves, in order not to influence subjects with \textit{a priori} associated semantic values. To explore the dataset without any written textual help, we have designed a selection interface where each leaf class is depicted by a point in 2D space \cite{lafay:hal-01300399}. Subjects may only interact with the leaf classes of the datasets but the location of the points in space is constrained by the taxonomy, \textit{i.e.} subclasses belonging to a same class are close to one another. This organization is chosen to facilitate the sound class selection process and allow the subject to quickly apprehend the full range of the dataset. This may also reduce several potential biases of studies relying on verbal data to describe a soundscape: 1) the variable fluency of the subject in the experiment language, 2) the lack of definite terms to fully describe sounds or complex sound environments \cite{guastavino_ideal_2006}, and 3) the high inter-subject variability in the verbal description of a same sound.

\subsection{Simulation process and parameters}
\label{sec:simulationProcess}

Following the above described terminology, a soundscape is a sum of tracks. Each track is modeled by a temporal sequence of sound samples belonging to the same sound class. Thus, each track is related to one specific leaf sound class (see \ref{subsection:Sound dataset}). To generate a soundscape, subjects may interact with the track parameters, but not with specific samples; for example, if a subject wishes to put a sequence of car sounds, he may specify how often (on average) cars pass by, and how loud they should be on average, but he cannot control each event individually. This allows to streamline the creation process, and keep the user focused on a global perception of the soundscape.

The simulation process involves two steps:

\begin{enumerate}
\item \textit{Sound Class selection:}  the subject selects a leaf class of sound (``\,car passing\,'', ``\,heavy rain\,'', \ldots{}). Once a class is selected, a track made of randomly selected samples belonging to the considered sound class is created. For a texture track, samples are sequenced without any gap, using crossfades to guarantee seamless transitions. The soundscape simulator guarantees that a texture sample will never transition to itself in order to avoid any unnatural effect due to the repetition of identical sounds, that a human would easily detect \cite{agus_rapid_2010}.
\item \textit{Sound Class parametrization:} the subject may tune the parameters (time positioning and sound levels) of the track.
\end{enumerate}

Five audio parameters allow the subject to parametrize a track. The parameters always affect all the samples of a track and never a particular isolated sound. Two parameters, namely \textit{Inter onset spacing time} and \textit{Sample fade In/Out}, are only used with event tracks.

\begin{enumerate}
\item \textit{Sound level} (dB): for each track, the sound levels of individual samples are drawn from a normal distribution. Subjects may control the sound level of a track by setting the mean and variance of the distribution. 
\item \textit{Global fade In/Out} (sec):  allows the subject to set a fade effect on the whole sequence of samples. 
\item \textit{Start and Stop position} (sec): allows the subject to set the beginning and the end times of a temporal sequence.
\item \textit{Inter onset spacing time} (sec): similarly to the sound level, inter onset spacing times between events are drawn from a normal distribution that may be parametrized by the subject for each event track.
\item \textit{Sample fade In/Out} (sec): allows the subject to set a same fade effect individually on each sample of an event track.
\end{enumerate}

Considering a soundscape as a sum of isolated samples is in line with the Auditory Scene Analysis (ASA) theory \cite{bregman1994auditory}, which states that auditory stimuli are decomposed into ``\,auditory objects\,'' called ``\,streams\,'', which may be regarded as sequences of auditory events emitted by putative sound sources \cite{ciocca2007auditory,winkler2009modeling}, and also with recent neuroscience studies suggesting that the primary auditory cortex (A1) produces representations of such ``\,auditory objects\,'' that act as bases for high level cognitive processing \cite{nelken2008neurons}. Thus, the track can be seen as the generative counterpart of the stream, which is rather a cognitive object.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%   EXPERIMENT Plan %%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Case Study: urban soundscape pleasantness}
\label{sec:CaseStudyUrbanSoundscape}
\subsection{Experimental planning}

We apply the proposed protocol to the field of urban soundscapes appraisal, by asking a panel of subjects to simulate two antagonistic urban soundscapes:  one ``\,ideal\,'', or pleasant (i) and the other ``\,non-ideal\,'' or unpleasant (ni). The two labels have been chosen to compare the results with those of a previous study addressing ideal urban sound environment representations \cite{guastavino_ideal_2006}. The simulation protocol used in this experiment was previously tested and validated thanks to a pilot study~\cite{soundscape3} run with 10 subjects.

To further characterize the simulated scenes, we run a complementary experiment where 10 subjects who did not take part in the first experiment are asked to evaluate the pleasantness of the simulated scenes using a bipolar semantic scale. This second experiment has two goals:

\begin{enumerate}
\item gain finer knowledge about the pleasantness of each simulated scene,
\item detect possible scene outliers. As the distinction i/ni acts as our reference throughout the analysis, one has to ensure that all the scenes are indeed perceived as ideal or non ideal scenes without ambiguity, \textit{i.e.} the highest average perceptual rating of the ni-scenes has to remain inferior to the lowest average rating of the i-scenes.
\end{enumerate}

The simulation experiment is referred as experiment-1 and the rating experiment as experiment-2. The complete experimental plan is described in Figure~\ref{fig:expLan}.

The simulated scenes as well as their annotations are available on an Archive Data Repository \footnote{\url{https://archive.org/details/soundSimulatedUrbanScene}}. Supplementary information including the experiment webpage and raw data for each experiment can be found on the companion website \footnote{\url{http://www.irccyn.ec-nantes.fr/~lagrange/demonstrations/simScene.html}}.

\begin{figure}[t]
\begin{center}
\includegraphics[width=.4\paperwidth]{../gfx/5.eps}
\caption{\label{fig:expLan} Experimental planning.}
\end{center}
\end{figure}

\subsection{Dataset of isolated samples}

The taxonomy, whose first two levels are depicted on Figure~\ref{fig:taxonomie}, is built following the category names found in previous studies addressing urban sound source categories \cite{maffiolo_caracterisation_1999,dubois2006cognitive,guastavino_categorization_2007, guastavino_ideal_2006, niessen_categories_2010, brown_towards_2011}, and is very similar to the one obtained in an other recent study on urban environmental sounds \cite{Salamon14}.

483 monophonic urban environmental sounds are collected, including 381 sound events and 102 textures are chosen to build the dataset of isolated sound samples. Among them, 260 events and 72 textures are recorded by the authors using a \textit{Audio Technica AT8035} shotgun microphone (to isolate sounds from their background) connected to a \textit{ZOOM H4n} recorder. The remaining come from existing sound banks. All sounds are normalized to the same root mean square (RMS) level.

\begin{figure}[t]
\begin{center}
\includegraphics[width=.4\paperwidth]{../gfxHierarchy/taxonomy.eps}
\caption{\label{fig:taxonomie} Taxonomy of (a) sound events, and (b) sound textures for the semantic levels 0 and 1, as well as an intermediate grouping level denoted $0^{+}$.}
\end{center}
\end{figure}

\subsection{Data analysis}

As experiment-1 generates a large amount of data suggesting numerous avenues of investigation, we here choose to restrict ourselves to the analysis of 1) selected sound classes, 2) sound event density, and 3) scene sound levels, other aspects being intentionally left for future research.

For ease of reading, we introduce the following terms and concepts that will be used throughout the analysis: 
\begin{description}
\item[$Diversity$:] number of distinct sound classes used in a scene. This number depends on the considered semantic level. For example, considering the two subclasses of the semantic level 2 \textit{car-passing} and \textit{car-starting}, which both belong to the same class \textit{car}, we count 2 distinct sound classes for semantic levels 2 and 3, but only 1 sound class for semantic levels 0 and 1. 
\item[$Density$:] logarithm of the number of event samples per chunk of 1 second, averaged over the scene duration. Density may be computed considering all sound classes, or only a subset of them. If nothing is specified, then the density accounts for all classes.
\item[$LA_{(eq,V)}$:]  Decibel of the RMS voltage relative to 1 volt, computed every second and averaged over the scene duration. An A-weighting filtering is performed before the RMS computation. Computing other statistical descriptors, namely $LA_{(min,V)}$, $LA_{(max,V)}$ and $LA_{(10-90,V)}$, shows that they present a high correlation with $LA_{(eq,V)}$ ($r\geq0.85$, $p<0.01$), therefore only the latter is considered.
\end{description}

We measure significance between the i- and ni-scenes using paired Student's t-test. All correlations are computed using the Pearson product-moment correlation coefficient. All statistical tests are performed at an $\alpha=5\%$ significance level.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%  EXPERIMENT Set Up  %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Experiment-1 setup}
\subsubsection*{Task}

Subjects are asked to successively create two urban soundscapes, lasting one minute each. The first must be ideal (a soundscape in which the subject would like to live), the second non-ideal (one in which they would try to avoid living). They are asked to mimic a static listener. Subjects are not restricted in their design choices, but are forbidden to create physically impossible situations such as ``\,a dog barking every 10 milliseconds\,''. At the beginning of the experiment, a 20 minutes tutorial is planned in order to familiarize the subjects with the software environment. The experiment is scheduled to last from two to three hours.

Due to time constraints for the experiment-2, we trim 15 seconds from the start and end of each simulated scene, giving us 30 second scenes.

\subsubsection*{Participants and Apparatus}

44 post-graduate students (14 females) of the {\'E}cole Centrale de Nantes (French engineering school) take part in the experiment. They are about the same age (M: 21.6, SD: 2). All of them have been living in the same large French city (Nantes) for at least two years before the experiment. The experiment is run simultaneously with the 44 subjects spread into three identical rooms with a calm environment. Subjects are forbidden to talk to one another and asked to adjust volume to a comfortable level once, at the beginning of the experiment. Audio is presented diotically to each participant via headphones. Three experimenters are always present (one in each room) to give instructions and answer queries if needed.

Due to a misunderstanding of the instructions or a failure to proceed within the time limits, four subjects are excluded from the analysis. 40 subjects completed the experiment successfully, providing 40 ideal urban scenes (i-scenes) and 40 non-ideal urban scenes (ni-scenes).

\subsection{Set-up experiment-2}
\label{sec:Experiment-2}
\subsubsection*{Task}

Subjects are asked to evaluate the pleasantness of the 80 simulated scenes using a bipolar scale of 7 points going from -3 (non ideal, very unpleasant) to +3 (ideal, very pleasant). The first ten scenes are used to give the subject an idea of the pleasantness range of the scenes. Those ten scenes are repeated at the end of the experiment and the ratings of their first occurrences are not taken into account in the analysis. Subjects listen to at least the 20 first seconds of the scenes and are free to skip the last 10 seconds at their convenience. Scenes are presented in a randomized order, but the ten calibration scenes are five i-scenes and five ni-scenes.

\subsubsection*{Participants and Apparatus}

10 post-graduate students (2 females) of the {\'E}cole Centrale de Nantes took part in the experiment. None of them did the first experiment. They are about the same age (M: 23.1, SD: 1.8). All of them have been living in the same large French city (Nantes) for at least two years. The experiment is run simultaneously with the 10 subjects in a calm environment. Audio is presented diotically to each participant via \textit{BeyerDynamic DT 990 Pro} semi open headphones at -12 dB (FS). One experimenter is always present to give instructions and answer queries if needed. All the  subjects successfully completed the experiment.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%  Results       %%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Results}

\subsection{Validity of the  dataset and the selection interface}
\label{sec:datasetAnalyses}

The dataset and selection interface are now assessed by considering the subject comments.

Looking at the database criticisms, 28 subjects stated that they could not find one or several desired sounds, with a maximum of 4 sounds for one subject. From all the missing sounds we identified 26 sound classes at different semantic levels. Among those classes, 16 were effectively present in the database, 1 referred to musical sounds which we had chosen to exclude from the database, and only 9 were effectively absent. Except for one sound class (``\,stroller pushchair\,''), each of the 16 classes which were reported as missing but were present in the database were used by at least one other subject. Among the 9 truly missing ones, we find very specific sounds as ``\,sport car\,'' or ``\,teenager voice\,''. We believe that those results show that the database diversity was sufficient for the purpose of the study.

If we look at the interface criticisms, 32.5\% of all the subject clearly stated that the selection interface was a ``\,straightforward yet effective way\,'' to find a sound whereas only 10\% indicated that they encountered difficulties. The remaining 57.5\% did not report difficulties. Those results tend to confirm that both the data set and the selection interface are well suited for the experiment.

\subsection{Consistency of the data}

To ensure consistency of the simulated scenes, no ni-scene should be rated in average as more pleasant as any i-scene, and reciprocally. Figure~\ref{xp2_1}.a shows the distributions of the averaged ratings considering scenes as observations. Four scenes (2 i-scenes an 2 ni-scenes) appear to be ambiguous, and are henceforth removed from the analysis, as well as their antagonist scenes. This operation leaves us with 36 i-scenes and 36 ni-scenes. Corrected distributions are displayed on Figure~\ref{xp2_1}.b. Figure~\ref{xp2_1}.c shows the distributions of the averaged ratings, considering subjects as observations, without the scene outliers.

It is clear that the pleasantness levels of the scenes, as intended by experiment-1 subjects, are well perceived by experiment-2 subjects ($p<0.01$).

\begin{figure}[t]
\begin{center}
\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp2_1.eps}
\caption{\label{xp2_1} Boxplot of the averaged ratings given by the subject considering (a) scenes as observations, (b) scenes as observations without scene outliers and (c) subjects as observations without scene outliers.} 
\end{center}
\end{figure}

\subsection{Sound level, density and diversity}
\label{sec:levelDensityDiversity}

\begin{figure}[t!]
\centering
\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_deSoLv_1.eps} 
\caption{\label{fig:xp1_deSoLv_1}  Boxplot of  the scene (a) $Density$, (b) $LA_{(eq,V)}$, (c) $Density$ \emph{vs} rating and (d) $LA_{(eq,V)}$ \emph{vs} rating}
\end{figure}

\begin{table}[t]
\centering
\begin{tabular}{l r r} 
                & $LA_{(eq,V)}$    & $Density$    \\
\hline
all             & \textbf{-0.76} ($p<0.01$)    & \textbf{-0.35} ($p<0.01$) \\
i-scenes        & -0.32 ($p=0.06$)    & -0.28 ($p=0.10$) \\
ni-scenes       & \textbf{-0.66} ($p<0.01$)    & \textbf{-0.47} ($p<0.01$) \\

\hline
\end{tabular}
\vspace{0.5mm}
\caption{\label{tab:numGlobal} Correlations computed between ratings \emph{vs} $LA_{(eq,V)}$ and $Density$}
\end{table}

\begin{figure}[t!]
\centering
\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_div_1.eps} 
\caption{\label{fig:analyseDensitySnrGlobal3} Diversity across semantic levels for event and texture classes.}
\end{figure}

The event density is slightly higher for ni-scenes than for i-scenes (Figure~\ref{fig:xp1_deSoLv_1}.a). An average density of $4.4$ for the i-scenes and $6$ for the ni-scenes is found, with a significant effect of the scene type ($p<0.05$). The density varies widely from subject to subject. Concerning $LA_{(eq,V)}$ (Figure~\ref{fig:xp1_deSoLv_1}.b), the  sound level is significantly higher for ni-scenes than for i-scenes ($p<0.01$).

To study the ability of those descriptors to characterize the appraisal, we compute the correlations between the ratings, the density and the $LA_{(eq,V)}$, considering all the scenes, as well as the i- and ni-scenes separately. Results are shown in Table~\ref{tab:numGlobal}, as well as in Figures~\ref{fig:xp1_deSoLv_1}.c and~\ref{fig:xp1_deSoLv_1}.d;

Not surprisingly, these results show that ni-scenes are on average louder than the i-scenes, and the strong correlation between $LA_{(eq,V)}$ and the ratings indicates that the sound level does influence the pleasantness. However, the correlations computed separately for the i- and ni-scenes suggest $LA_{(eq,V)}$ does not characterize the notion of pleasantness in a equal manner for the two type of scenes:  there is still a good correlation between $LA_{(eq,V)}$ and the ratings for the ni-scenes, but there is no correlation if we take only the i-scenes. This is in line with the trend observed by \cite{lavandier2006contribution} that sound level may be a sufficient indicator of comfort for negatively rated acoustics scenes, but not for positively rated ones.

Considering the event densities, ni-scenes seem to be slightly more populated that i-scenes. However, it appears that a global measure of sound event density fails to characterize the appraisal if all the scenes are considered at once. If we consider the two types of scenes separately, the correlation remains weak for the ni-scenes, and is non-existent for the i-scenes. This may suggest that either the number of events is an irrelevant indicator, or that only the densities of specific event classes are involved. Still, although there is no clear correlation between density and pleasantness for the ni-scenes, the density seems to play an indirect role as it is correlated with $LA_{(eq,V)}$ ($0.62$, $p<0.01$).

Averaged diversities are shown on Figure~\ref{fig:analyseDensitySnrGlobal3}. Except for semantic level 0, the diversity is significantly higher for the event classes of the ni-scenes (levels 1, 2 and 3: $p<0.01$;), revealing that ni-scenes are composed of a larger variety of sound events than i-scenes. Diversity remains constant across semantic levels for texture classes of i- and ni-scenes, with no statistical differences. The correlations computed between the event diversities and the averaged pleasantness ratings remain weak ($r<0.45$, $p<0.01$), irrespective of the considered semantic level. If we look at the two type of scenes separately, no correlation is observed. This suggests that diversity plays no role in pleasantness evaluation.

\subsection{Class selection}
\label{sec:classSelection}

We investigate the sound class selection by counting the number of times a class has been used to simulate the i- or ni-scenes. Figures~\ref{fig:xp1_class_1} and \ref{fig:xp1_class_2} show the class distributions for the event and texture classes, respectively. For ease of reading, we choose an intermediate semantic level, denoted $0^{+}$, laying between semantic levels 0 and 1 (see Figure~\ref{fig:taxonomie}).

There is a considerable difference between the sound class selections of the i- and ni-scenes, showing that semantics play an important role in soundscape qualitative evaluation. The sound classes shown in Figures~\ref{fig:xp1_class_1} and~\ref{fig:xp1_class_2} are very close to those found by Guastavino \cite{guastavino_ideal_2006}, \textit{i.e.} classes that suggest human presence and nature are the most present in i-scenes while classes referring to mechanical and construction work sounds are used for ni-scenes.

However, some differences have to be reported: \cite{guastavino_ideal_2006} reports that ``\,public transports\,'' are typical sounds of an ideal urban environment. This observation is attributed to the fact that urban environment perception is driven by social values. Public transport sounds are thus better accepted than those of private vehicles, as they are usually positively connotated. Results shown in Figure~\ref{fig:xp1_class_1} tend to qualify this claim  by showing that sounds of public transports (\textit{bus} and \textit{train}) were chosen by $27.5\%$ of the subjects for the i-scenes and by $45\%$ of the subjects for the ni-scenes. It is important to note that in \cite{guastavino_ideal_2006} subjects are asked to recall an environment in memory, that is, without any sound support. Despite the fact that the idea of public transport sounds are well accepted due to societal considerations (\textit{bus} sounds correspond indeed to $27.5\%$ of the event classes used in the i-scenes, as much as \textit{bicycle} sounds and more than any other vehicle sounds), the sound itself still remains similar to that of any other vehicle. Thus the class \textit{public\_transport} does appear in ni-scenes, more than \textit{car} or \textit{truck}. Similar findings showing that bus presence has a negative effect on soundscape appraisal are reported in \cite{lavandier2006contribution}.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_class_1.eps}
    \caption{\label{fig:xp1_class_1} Percentage of subjects using event sound classes of the semantic level $0^{+}$ to simulate the i-scenes (left-green) and the ni-scenes (right-red).}
        \end{center}
\end{figure}

\begin{figure}[t!]
\begin{center}
\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_class_2.eps}
    \caption{\label{fig:xp1_class_2} Percentage of subjects using texture sound classes of the semantic level $0^{+}$ to simulate the i-scenes (left-green) and the ni-scenes (right-red).}
    \end{center}
\end{figure}

\subsection{Markers}
\label{sec:markers}

As semantics is found to play an important role in the qualitative perception of the soundscape, we now investigate the existence of sound markers, \textit{i.e.} sound classes of great importance when considering the pleasantness of a soundscape. Here, an event class is considered as a marker if it has been mostly used in one type of soundscape (i or ni). To identify markers, a two-tailed V-test statistical test is performed for each semantic level. The V-test tests the null hypothesis that the $n_k$ event classes used for the scene type $k$ are randomly drawn from all the $n$ event classes available. A rejection of that hypothesis provides evidences that an event class has been significantly used for one type of scene only. Since the test concerns many event classes, Bonferroni correction is applied.

Results are shown on Table~\ref{tab:markers}. Nine markers are found across all semantic levels. As suggested by Figures~\ref{fig:xp1_class_1} and~\ref{fig:xp1_class_2}, sounds evoking human activity (\textit{male footsteps concrete}) and nature (\textit{animal}, \textit{birds} and \textit{birds singing}) are markers of the i-scene. The presence of \textit{church bell} and \textit{bell bike} as strong markers of the i-scenes can be due to the socio-cultural background of the subjects, in great majority European citizens. It confirms Schafer's claim that sounds that are recognized by a community as integral part of its sonic environment are popular \cite{schafer1977tuning}.

Markers of the ni-scenes are sounds due to construction work (\textit{construction work}), and sounds suggesting dense traffic (\textit{horn}, \textit{siren}). Although these markers are rather intuitive, none of the event classes related to motorized vehicles is a marker. It seems that subjects chose to illustrate unpleasant traffic sounds using horns and sirens.

\begin{table}[t]
 \setlength{\tabcolsep}{0.1pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c} 
Semantic   &  \multicolumn{2}{c}{Markers} \\
level & i-scenes & ni-scenes \\
\hline
0  &              &  construction work  \\
\hline
1  & church bell  & horn  \\
   & bell bike    & siren \\
   & animal       &       \\
   \hline
2  & birds         & horn  \\
   & church bell   & siren \\
   & bell bike     &       \\
   \hline
3  & birds singing & horn  \\
   & church bell   & siren \\
   & bell bike     &       \\
   & male footsteps concrete &  \\
  \hline
\end{tabular}
}
\vspace{0.5mm}
\caption{\label{tab:markers} Event classes found to be markers. In each cell, markers are ordered from top to bottom by decreasing V-test values.}
\end{table}

\subsection{Quantifying class selection}
\label{sec:QuantifyingClassSelection}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.4\paperwidth]{../gfxMatlab/pa5_1.eps}
\caption{\label{fig:pa5} Precision at rank 5 ($P@5$) computed from the \textit{Hamming} distances between the scenes at different semantic levels.}
\end{center}
\end{figure}

To quantify how class selection characterizes the pleasantness of a urban environment, we now follow a class-based classification of the scenes evaluated with the standard precision at rank 5 ($P@5$) metric.

Each simulated scene $s_i$ is represented by a boolean vector of dimension $n$: $S_i = (x_1, x_2, \ldots{}, x_{n})$, $i\in[1,72]$. Each dimension corresponds to a sound class (event and texture) of a particular semantic level (for example: $n=44$ classes for semantic level 1). Thus $x_1 = 1$ indicates that the sound class $x_1$ is present in scene $s_i$ (not present if $x_1=0$). A \textit{Hamming} distance is computed between all pairs of $S_i$. Then, for each vector $S_i$ the number of scenes sharing the same label (i or ni) as $S_i$  among the five scenes $s_j$ ($j=1,...,5$) that have the smallest distance to it is computed. The $P@5$ precision is then obtained by averaging the results over all the scenes $s_i$ ($i=1, \ldots{}, 72$).

Results considering both the event and texture classes ($ET$) to build the vectors  $S_i$ are shown on Figure~\ref{fig:pa5}. The $P@5$ is of 81\% for semantic level 0 and remains above 89\% starting at semantic level 1. These result strongly indicates that it is possible to make a clear distinction between the i- and ni-scenes based solely on the class presence/absence. We also note that the deeper the hierarchical level, the higher the $P@5$; in other words an increased precision in identifying scene elements allows a better distinction between i- and ni-scenes.

To refine the analysis, only event classes ($E$), texture classes ($T$), event markers ($E_m$) and event classes without event markers ($E_{w/o m}$) are also considered on Figure~\ref{fig:pa5}. The constant deviation of circa $10\%$ from the semantic level 1 between $E$ and $T$, and the fact that $P@5$ are similar for $ET$ and $E$ indicate that soundscape quality is mostly conveyed by sound events.

Since similar precision are achieved by $ET$, $E$ and $E_m$, we may conclude that all the event classes are not equivalently important to distinguish between the two scene types. Indeed, $E_m$ uses very partial information but still reaches quasi-equivalent results. Moreover, the drop in $P@5$ considering $E_{w/o m}$ from the semantic level 1 confirms the fact that soundscape quality mostly relies on the presence of specific sound markers.

\subsection{Marker-wise pleasantness descriptors}
\label{sec:QuantifyingMarkers}

\begin{table}[t]
\centering
\begin{tabular}{l r r} 
                    &   i-scenes   & ni-scenes \\
\hline
$LA_{(eq,V)\_m}$    & 0.19  ($p=0.28$)           & \textbf{-0.60} ($p<0.01$) \\
$LA_{(eq,V)\_b}$    & \textbf{-0.51} ($p<0.01$)  & -0.16 ($p=0.36$) \\
$LA_{(eq,V)\_m-b}$  & \textbf{0.69} ($p<0.01$)   & -0.31 ($p=0.06$) \\
$Density_{m}$       & -0.20  ($p=0.24$)          & \textbf{-0.58} ($p<0.01$)  \\
$Density_{b}$       & -0.23  ($p=0.18$)          & -0.30 ($p=0.07$)  \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{\label{tab:numMarkers} Correlations computed between rating \emph{vs.} $LA_{(eq,V)\_m}$, $LA_{(eq,V)\_b}$, $LA_{(eq,V)\_m-b}$ and $Density_m$.}
\end{table}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_deSoLvMar_1.eps}
  \caption{\label{fig:numMarkers} Averaged pleasantness vs. (a)  $LA_{(eq,V)\_m-b}$, (b) $LA_{(eq,V)\_m}$, (c) $LA_{(eq,V)\_b}$ and (d) $Density_m$.}
  \end{center}
\end{figure}

In order to check if the physical characteristics of the markers alone have an influence on the pleasantness, we now consider the densities and the $LA_{(eq,V)}$ with on one hand the tracks of the markers ($LA_{(eq,V)\_m}$, $Density_{m}$) and on the other hand the remaining sounds ($LA_{(eq,V)\_b}$, $Density_{b}$). The marker durations is rather short, and the number of markers in a scene is rather low, a window length of 125 ms is used to compute $LA_{(eq,V)\_m}$ and $LA_{(eq,V)\_b}$. For the i-scenes we consider only the markers of the i-scenes (i-markers), and for the ni-scenes only the markers of the ni-scenes (ni-markers).

The difference $LA_{(eq,V)\_m-b}=LA_{(eq,V)\_m}-LA_{(eq,V)\_b}$ is also considered as it illustrates how the markers emerge from the other sounds. Correlations between those descriptors and the average perceptual ratings are depicted in Table~\ref{tab:numMarkers} and Figure~\ref{fig:numMarkers}.

Considering $Density_{m}$ (Figure~\ref{fig:numMarkers}.d), results are almost the same than to those found in section~\ref{sec:levelDensityDiversity}. There is no correlation for the i-scenes, but the correlation increases for the ni-scenes. No correlation is found for $Density_{b}$. Sound levels descriptors are much more informative. Little to no information is lost by considering only the markers to compute $LA_{(eq,V)\_m}$, \emph{i.e.} there is still a good correlation for the ni-scenes (although inferior to the one observed in Section~\ref{sec:levelDensityDiversity}), and no correlation for the i-scenes.

Considering the ni-scenes, the good correlation for $LA_{(eq,V)\_m}$  (Figure~\ref{fig:numMarkers}.b) and the absence of correlation for $LA_{(eq,V)\_b}$  (Figure~\ref{fig:numMarkers}.c) suggests that the sound levels of the ni-markers convey most of the annoyance. Also, no correlation is observed for $LA_{(eq,V)\_m-b}$ (Figure~\ref{fig:numMarkers}.a). This indicates that the predominant influence of the marker sound levels remains true irrespective of the sound level of the other sounds. A loud horn sound is expected to increase the annoyance level of a scene (more so than a soft klaxon sound), whether or not the remaining of the sonic environment is ``\,noisy\,''.

That said, this last statement is not valid for the i-scenes. The strong uphill correlation found between pleasantness and  $LA_{(eq,V)\_m-b}$ (Figure~\ref{fig:numMarkers}.a) shows that the higher the markers sound levels relative to the other sounds, the higher the pleasantness. But the moderate downhill correlation observed for $LA_{(eq,V)\_b}$ (Figure~\ref{fig:numMarkers}.c) shows that the levels of those remaining sounds may be a cause of annoyance if perceived as too loud. It is thus the emergence of the i-markers from the others sounds that drive the appraisal. Considering for example the sonic environment of a park and following those conclusions, the global appraisal will increase with the sound level of the birds, provided other sounds like human voices or traffic hubbub are not too loud.

These results are strong indicators that global sound level is not a good indicator to characterize the appraisal of \emph{a priori} positive sound environments such as a park or a quiet street. On the contrary, the contribution of each sound source needs to be considered separately, as for certain type of sounds, the pleasantness does increase with their sound levels.

Those observations may also serve as guidelines for urban planner, showing that:

\begin{itemize}
\item in the case of negative environment such as street with public work or busy traffic, where it is often difficult to diminish the global sound level, focusing only on the noise regulation of specific sound sources (jack-hammer or horn) may significantly reduce the overall annoyance;
\item in the case of positive environment such as park or quiet street, the appraisal may be enhanced by increasing the presence of specific sound sources such as birds singing.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%    DISCUSSION AND PERSPECTIVES   %%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Discussion and perspectives}

We introduce in this paper a novel experimental protocol to study the perception of soundscapes, where subjects are asked to recreate an auditory environment using a tool specifically developed for that purpose and a structured database of isolated sounds. This protocol allows us to generate soundscapes whose composition, in terms of nature, timing and intensity of sound sources, is known with perfect accuracy. Using the data thus produced in a series of analyses on the issue of ideal \emph{vs} non-ideal urban soundscape classification and characterization, we show that:

\begin{itemize}
\item The data obtained with the simulation allows us to reach results that are consistent with previous perceptual studies relying on recorded scenes \cite{lavandier2006contribution} or the experimental subjects' memory \cite{guastavino_ideal_2006}.
\item From the global descriptors sound level, class diversity and event density, only sound level seems to play a role in pleasantness evaluation.
\item The ability of sound level to characterize pleasantness relies upon the quality of the scene itself. For positively rated scenes (i-scenes), it is not relevant.
\item A semantic description of the soundscapes in terms of presence~/ absence of sound sources is an effective criterion to distinguish between ideal and non-ideal urban environments.
\item A few event markers, appearing at various semantic levels, suffice to perform efficient discrimination between ideal and non-ideal urban environments, although the type of markers found in this study may depend on the socio-cultural background of the subjects. 
\item The sound level computed on the markers only is a better predictor of perceived agreement than the global sound level.
\end{itemize}

Compared to more traditional survey based methods, the proposed method provides an increased precision and reliability of soundscape description useful for the analysis of subjects' judgments. This article was intentionally focused on the analysis of the scene composition; further research shall be dedicated to 1) a thorough study of the influence of the physical characteristics of the sound markers on pleasantness evaluation, as well as categorisation process, and 2) the elaboration of a predictive method based on that analysis, allowing to anticipate the agreement of an acoustic environment from its composition.

In addition, more understanding shall be gained about the actual impact of sound level on soundscape perception: is there a significant difference in evaluation between two scenes having the exact same composition, but different levels? Is loudness but a consequence of scene composition ?

\noindent \textbf{Acknowledgements}\\
\setlength{\parindent}{0.7cm} 
Research project partly funded by ANR-11-JS03-005-01. The authors would like to thank the students of the Ecole Centrale de Nantes for their willing participation.

%----------------------------------------------------------------------------------------
%	REFERENCE LIST
%----------------------------------------------------------------------------------------

\References{}{../biblio}{actalit}

\end{document}

### 2016-05-24 part2/lafaySimulatedScenePerception2.tex
initial commit

% Instructions for Authors fuer Acta Acustica

\documentclass[twoside,twocolumn]{article}
\usepackage{acta}
\usepackage{url}
 \usepackage[utf8]{inputenc}

\begin{document}
\Eingang{00}{00}{0000} \Annahme{00}{00}{0000} \Sachgebiet{1}
\PACS{43.50.Lj, 43.50.Rq, 43.50.Qp, 43.66.Lj, 43.66.Ba}
\Band{0} \Jahr{0000} \Heft{} \Ersteseite{1} \Letzteseite{}

\AuthorsI{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J.F. Petiot$^{1)}$}
\AddressI{$^{1)}$ Institut de Recherche en Communications et  Cybern\'etique de Nantes (IRCCYN), Nantes, France.\\
\hspace*{8pt}gregoire.lafay@irccyn.ec-nantes.fr\\
$^{2)}$ Institut de Recherche et Coordination Acoustique/Musique (IRCAM), Paris, France}

% \footnotemark[1] \Authornotes{\footnotetext[1]{} }

\Englishtitle{Investigating soundscapes perception through acoustic scenes simulation. Part II} \Germanorfrenchtitle{}

\Kolumnentitel{Lafay et al. : Approaching mental representations}

\Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process. Subjects are asked to create a full soundscape by means of a  software dedicated to sound arrangement, coupled with a structured sound data set. This paradigm is used to characterize urban sound environment representations, by analyzing the sound classes that were used to simulate the auditory scenes. Three others perceptive experiments are conducted to refine the analysis of the simulated urban acoustic scenes. Results show that a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes and potentially sufficient for some description tasks.}
\Germanorfrenchabstract{}

\ScientificPaper

\newcommand{\gl}[1]{\textcolor{blue}{GL : #1}}
\newcommand{\partI}[1]{\textcolor{red}{From part I : #1}}
\newcommand{\ml}[1]{\textcolor{magenta}{ML : #1}}
\newcommand{\mr}[1]{\textcolor{green}{MR : #1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%   INTRODUCTION      %%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{introduction}

\partI{For two decades, the soundscape paradigm has been considered to study the perception of acoustic scenes in order to provide urban planners with meaningful sound quality indicators (see \cite{dubois2006cognitive} and  \cite{aletta2016soundscape} for extensive reviews). Most studies follow the modus operandi of cognitive psychology \cite{maffiolo_caracterisation_1999}, by first describing the soundscape using both acoustic measurements and perceptive factors, and then investigating the interactions between those descriptors to understand the cognitive processes involved in the perception of the acoustic scenes. Soundscape studies are thus interdisciplinary \cite{davies_perception_2013, aletta2016soundscape}, borrowing tools from various fields such as acoustics, psycho-physics, cognitive psychology but also psycho-linguistics, sociology and data mining.}

\partI{The soundscape approach appears to be a powerful tool to develop perceptively motivated indicators \cite{schulte-fortkamp_soundscape:_2007}. That said, the use of various experimental protocols to assess and measure soundscapes makes the integration of results difficult \cite{davies_perception_2013}. Also, there still exist no definite consensus in the soundscape community about the acoustic descriptors to be used \cite{aletta2016soundscape}. This issue prevents experimenters to provide  urban planners with generic and clear indicators or models able to predict sound environment quality \cite{schulte-fortkamp_soundscape:_2007, aletta2016soundscape}. Recently, several projects such as the European Cooperation in Science and Technology Action (TD0804: soundscape of European Cities and Landscapes) or the Positive Soundscape project \cite{davies_perception_2013} have been undertaken to standardize soundscape assessment and descriptors \cite{schulte2013soundscape}, but those matters are still considered as open problems.}

\partI{We believe that soundscapes studies greatly suffer from one important issue: to the best of our knowledge, there is no morphological definition of a soundscape, and therefore no standardized way to decompose it in distinct elements. It is thus difficult to study precisely which component of the environment is the cause of potential variations in the perceptive measurements.} The fact that sound sources does not influence the perception in a equal manner is not a novel idea. In \cite{schafer1977tuning} Schafer had already introduced the notion of \emph{soundmark}, which refers to sounds with subjective qualities that bring them into focus for people used to them.

To tackle this issue, we introduced in the first part of this paper \cite{lafayPartI} a simulation protocol where subjects were asked to simulate a full acoustic environment thanks to a soundscape simulator linked to a dataset of isolated sound samples. The simulation allows us to generate acoustic environments fully characterized in term of both source presences and sound levels, thus enabling us to study contribution of each sound source on soundscape perception. We presented the results of a case study focused on urban soundscapes pleasantness. First a simulation experiment was run giving us 36 ideal urban soundscapes (i-scenes) and 36 non-ideal ones (ni-scenes). Second a pleasantness rating experiment was run with a different pool of subject using a 7-points semantic scale. The combined analysis of the two experiment allowed us to draw the following outcomes:

\begin{itemize}
\item The ability of the sound level to characterize the pleasantness relies upon the quality of the scene itself. For positively rated scenes (i-scenes), it is not relevant.
\item  A semantic description of the soundscapes in terms of presence~/ absence of sound sources is an effective criterion to distinguish between ideal or non-ideal urban environment.
\item  A few event markers (see Table~\ref{tab:markers}), appearing at various semantic levels, suffice to perform efficient discrimination between the ideal or non-ideal urban environment, although the type of markers found in this study may depend on the socio-cultural background of the subjects. 
\item Sound level computed on the markers better characterize the appraisal than global sound level.
\end{itemize}

Following those first results, we conducted two complementary experiments to further our understanding of the sound source contribution on scene perception:\\

\textbf{Experiment-1}: a pleasantness rating experiment similar. In order to check the effect of markers, subjects must rate 144 scenes comprising the 72 initial scenes, and the same 72 scenes but re-simulated without the sound markers of semantic level 1. The goal is to check whether markers are such salient discriminant factors that removing them will blur the distinction between i- and ni-scenes.\\

\textbf{Experiment-2}:  a free sorting task experiment. This experiment has multiples goals:
\begin{enumerate}
\item to confirm that the categorizations of the simulated scenes rely on sound sources identification. This fact has been shown considering recordings of complex sound environment mostly  \cite{maffiolo_caracterisation_1999,dubois2006cognitive}
\item to identify what sound categories are spontaneously used by the subjects to categorize the scenes, and observe at what level of abstraction those categories are expressed.
\item to see whether  the event markers  \gl{, caractéristiques de qualité d'un environnement} correlate with those categories  
\item  \gl{Autour de quelles classes/categories s'opèrent les regroupements par similarité}\\
\end{enumerate}

\gl{HERE RESULTS}

\begin{table}[t]
 \setlength{\tabcolsep}{0.1pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c} 
Semantic   &  \multicolumn{2}{c}{Markers} \\
level & i-scenes & ni-scenes \\
\hline
0  &              &  construction work  \\
\hline
1  & church bell  & horn  \\
   & bell bike    & siren \\
   & animal       &       \\
   \hline
2  & birds         & horn  \\
   & church bell   & siren \\
   & bell bike     &       \\
   \hline
3  & birds singing & horn  \\
   & church bell   & siren \\
   & bell bike     &       \\
   & male footsteps concrete &  \\
  \hline
\end{tabular}
}
\vspace{0.5mm}
\caption{\label{tab:markers} Event classes found to be markers. In each cell, markers are ordered from top to bottom by decreasing V-test values.}
\end{table}

\section{Experiment-1 set-up}
\subsection{Task}

Subjects are asked to evaluate the pleasantness of 144 urban environments, comprising the 72 original simulated scenes, and the same 72 scenes but generated without the sound markers. Evaluation is made using a bipolar semantic scale of 11 points going from -5 (non ideal, very unpleasant) to +5 (ideal, very pleasant). The first ten scenes are used to give the subject an idea of the pleasantness range. Those scenes are repeated at the end of the experiment and their first occurrences are not taken into account for the analysis. Subject are forced to listen to at least the 20 first seconds of the scenes, but are free to skip the 10 last seconds if needed. Scenes are presented in a randomized order, but we make sure that the ten first scenes 1) contain sound markers and 2) are evenly balanced between i- and ni-scenes.

\subsection{Participants and apparatus}

12 subjects take part in experiment-1. None of them did the simulation experiment nor the first pleasantness rating experiment. The experiment is run simultaneously for the firsts 5 subjects.  The remaining of the subjects did the experiment separately. All the subjects did the experiment in a calm environment. Audio is presented diotically to each participant via \textit{BeyerDynamic DT 990 Pro} semi open headphones at -12 dB (FS). One experimenter is always present to give instructions and answer queries if needed. The experiment lasts one hour and a half. All the subjects successfully completed the experiment.

\subsection{Statistical analysis}

We measure significance using paired Student's t-test and two-way repeated-measure ANOVA. All the tests are preformed at a $\alpha=5\%$ significance level. For repeated measures designs, sphericity is assessed using Mauchly's sphericity test. If the sphericity assumption does not hold, we use the Greenhouse-Geisser correction, and in that case the corrected $p$-value is noted $p_{gg}$. All correlations are computed using the Pearson product-moment correlation coefficient.

\section{Experiment-1 results}

Among the 12 subjects, one gives an averaged rate for the ni-scenes superior to 2*IQR of the median averaged rates of the other subjects. That subject is therefore considered as an being an outlier and  removed from the data.

Figure~\ref{fig:xp4_3}.a depicts the distributions of the averaged ratings for the i- and ni-scenes, with and without markers, considering the scenes as observations. We note that the impact of the scene outlier detection procedure run in \cite{lafayPartI} stands in this experiment: none of the i-scenes (resp. ni-scenes) has an averaged rating inferior to the maximum averaged rating of the ni-scenes. Furthermore we have a correlation of $r=0.95$ ($p<0.01$) between the scene averaged ratings of the two experiments, denoting  the consistency of the results.

We show the averaged ratings per subjects for the i- and ni-scenes with and without markers in Figure~\ref{fig:xp4_3}.b. A two way repeated measure ANOVA considering the presence of markers as well as the scene type (i/ni) as independent variables reveals significant effect of both the marker presence ($F[1,10]=14$, $p_{gg}<0.01$), the scene type ($F[1,10]=116$, $p_{gg}<0.05$) and the interaction ($F[1,10]=49$, $p_{gg}<0.01$). \emph{Post hoc} analysis shows significant differences between the mean averaged ratings with and without markers for both the ni-scenes ($p<0.01$) and  the i-scenes ($p<0.05$).

Figures~\ref{fig:xp4_3}.c and~\ref{fig:xp4_3}.d illustrate the effect of scene type and presence/absence of markers on $LAeq$ and density, respectively. Concerning $LAeq$, a two way repeated measure ANOVA shows a significant effect of both the markers presence ($F[1,35]=123$, $p_{gg}<0.01$) and the scene type ($F[1,35]=77$, $p_{gg}<0.01$), with \emph{post hoc analysis} showing significant differences between the mean $LAeq$ for both the ni-scenes ($p<0.01$) and the i-scenes ($p<0.05$) with and without markers. Considering the densities, we also find a significant effect of both the markers presence ($F[1,35]=52$, $p_{gg}<0.01$) and the scene type ($F[1,35]=6$, $p_{gg}<0.05$), with \emph{post hoc} analysis revealing once again significant differences for both the ni-scenes  ($p<0.01$) and the i-scenes ($p<0.05$) with and without markers. Those results show us that, as expected, the removal of the markers does have an effect on scene morphology, since this operation reduces both the densities and the sound levels for both the i- and ni-scenes.

This results validates our hypothesis that the presence/absence of specific sounds classes directly influence the pleasantness of an acoustic scenes. As expected, the removal of the ni-makers increases the pleasantness of ni-scenes, by 1 in average. The removal of the i-markers does decrease the pleasantness of i-scenes, but by a lower amount (0.4), more surprisingly.

While $LAeq$ does have an influence on pleasantness, we can conclude here that sound markers are more important. Indeed, removing positive markers such as \textit{bike bell}, \textit{church bell} and \textit{birds} from a scene decreases the sound level, and should \emph{a priori} have a positive impact on pleasantness, but in fact the scene rating decreases significantly.

\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{gfxMatlab/xp4_3.eps}
\caption{\label{fig:xp4_3} Boxplot of the averaged ratings considering (a) subject as observations and (b) scenes as observations. Boxplot of (c) the scene $LAeq$ and (d) the scene densities; \textit{w/ m} and \textit{w/o m} indicate the presence or absence of sound markers respectively.}
\end{center}
\end{figure}

\subsection{pleasantness and sound level}

\begin{table}[t]
 \centering
\begin{tabular}{c c c} 
            & w/ markers          & w/o markers       \\
\hline
all         & -0.82 ($p<0.01$)    & -0.79 ($p<0.01$)  \\
i-scenes    & -0.49 ($p<0.01$)    & -0.53 ($p<0.01$)  \\
ni-scenes   & -0.85 ($p<0.01$)    & -0.68 ($p<0.01$)  \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{\label{tab:plCorr} Correlations between $LAeq$ and pleasantness.}
\end{table}

To study the impact of the markers removal on how sound level is able to characterize the appraisal, we compute the correlations between $LAeq$ and the averaged pleasantness. Results are depicted in Table~\ref{tab:plCorr}.

Correlations observed for the scenes with markers confirm the trends observed in \cite{lafayPartI}. The strong correlation observed for the ni-scenes with markers and the weak correlation observed for the i-scenes reveal that 1) sound level does not characterize appraisal in a equal manner for the i- and ni-scenes and 2)  while sound level is a good descriptor of pleasantness for ni-scenes, it is not the case for positively rated ones.

Moreover results show that removing the markers tends to slightly increase the correlation for the i-scenes, and decrease the correlation for the ni-scenes. Considering the i-scenes, we sow in \cite{lafayPartI} that the louder the i-markers relative to the remaining of the sounds, the higher the pleasantness. It appears that in absence of the i-markers, the negative influence of the sound level on the appraisal becomes more predominant. Considering the ni-scenes, the inverse effect is observed. Indeed results of \cite{lafayPartI} showed a strong negative correlation between the $Laeq$ of the ni-markers and the pleasantness. It appears here that the removal of those markers drastically diminishes the negative impact of the sound level.

These results are strong indicators that all the sources do not equivalently participated in pleasantness perception. In particular they show that the ability of sound level to predict pleasantness is only due to the presence of some sources, \emph{i.e.} the ni-markers. In the absence of such sources, appraisal seems to rely on other descriptors.

\section{Experiment-2 set-up}
\subsection{Task}
Les mêmes resultats sont observé dans \cite{guyot2005urban, ricciardi2015sound}.

Subjects are asked to group the 72 simulated scenes (36 i-scenes and 36 ni-scenes) based on their similarity. They are allowed to create as many clusters as they wish. The scenes are represented as dots on a plane. Subjects may listen to a scene by clicking on a dot, move the dots, and use color to specify clusters. After the  sorting part, subjects are asked to label each cluster using a word, a list of words, or a short sentence.

\subsection{Participants and Apparatus}

Experiment-2 is run with the same subjects who took part in the rating experiment in \cite{lafayPartI}, but two weeks after. The 10 subjects  take part in the experiment simultaneously, in a calm environment. Audio is presented diotically to each participant via \textit{BeyerDynamic DT 990 Pro} semi open headphones at -12 dB (FS). One experimenter is always present to give instructions and answer queries if needed. All the subjects successfully completed the experiment.

\subsection{Statistical analysis}

We measure significance using one way. All the tests are preformed at a $\alpha=5\%$ significance level. For multiple comparison procedures we use Bonferroni correction, and in that case we designate by $p_{bonf}$  the corrected $p$-value. All correlations are computed using the Pearson product-moment correlation coefficient.

\section{Experiment-2 results}
\subsection{Categorization strategies}

\setlength\tabcolsep{0.7mm}
\begin{table}[t]
 \centering
{\renewcommand{\arraystretch}{0.9}
\begin{tabular}{lccccc}
subject & source & location & quality & volume & frequency  \\
\hline
    $1^*$&        &          &    x    &        &           \\
    2    &   x    &          &         &        &           \\
    $3^*$&        &          &         &    x   &     x     \\
    4    &   x    &          &    x    &        &           \\
    5    &   x    &    x     &    x    &        &           \\        
    $6^*$&        &          &    x    &        &           \\
    7    &   x    &          &         &    x   &           \\
    $8^*$&        &          &    x    &        &           \\
    9    &   x    &    x     &         &        &           \\
    10   &   x    &          &         &        &           \\
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{\label{tab:StratSate} Strategies adopted by the subjects to categorize the scenes. Subjects removed from the data are indicated by $*$.}
\end{table}

\setlength\tabcolsep{0.7mm}
\begin{table}[t]
 \centering
{\renewcommand{\arraystretch}{0.9}
\begin{tabular}{ccc}
labels & categories  & classes\\
       & (merging)   &        \\
\hline
alarm (2)             & alarm            &  \textit{car\_fire\_alarm} (sl-1) \\

horn  (2)             & horn             &  \textit{horn} (sl-1)             \\
                      & (siren/horn)     &                                    \\
siren (3)             & siren            &  \textit{siren} (sl-1)            \\
                      & (siren/horn)     &                                    \\
birds (4)             & birds            &  \textit{birds} (sl-2)             \\
\hline
church bell (4)       & church bell      &  \textit{church bell} (sl-1)      \\
church (2)            &                  &                                   \\
\hline
mechanical            & construction     &  \textit{hand tool} (sl-1)       \\
public work  (4)      &                  &  \textit{power\ tools} (sl-1)   \\
tools                 &                  &  \textit{jackhammer} (sl-1)       \\
                      &                  &  \textit{metal\_work} (sl-1)       \\
			          &                  &  \textit{vehicle\_work} (sl-1)     \\  	
\hline			                                                             
human (4)             & human            &   \textit{voice} (sl-0)           \\      
people, crowd         &                  &   \textit{footstep} (sl-1)       \\      
footstep (2)          &                  &                                   \\  
\hline
traffic (2)           & traffic          &  \textit{scooter} (sl-1)         \\
car (2)               &                  &  \textit{motorcycle} (sl-1)       \\
                      &                  &  \textit{car} (sl-1) \textit{truck} (sl-1) \\ 
\hline
water, rain (2)       & water            &   \textit{water}$^T$ (sl-1)          \\ 
bad weather           &                  &                                  \\ 
\hline
market                & market           &   \textit{market}$^T$ (sl-1)          \\ 
                      & (human)          &                                    \\
park                  & park             &   \textit{park}$^T$ (sl-1)           \\ 
                      & (bird)           &                                    \\
nature (3)            & nature           &   \textit{nature}$^T$ (sl-0)          \\  
background$*$         & ---              &    ---                              \\ 
noise$*$ (2)          & ---              &    ---                              \\               
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{\label{tab:association}  Associations made between labels, categories and classes of the class taxonomy taken at specific semantic levels (sl). Numbers in parentheses indicate the number of subjects who use this label. ($*$) indicates labels not taken into account in this study. ($^T$) indicates texture classes}
\end{table}

From the labels used by the subjects to describe their clusters, 5 categorisation strategies can be identified (Table~\ref{tab:StratSate}). Most subjects ($70\%$) choose to categorize the scenes based on the presence/absence of sound sources. Other strategies rely on location (ex: ``\,park\,'', ``\,market\,''), affective quality (ex: ``\,calm,'', ``\,relaxing\,'', ``\,annoying\,''), volume (ex: ``\,loud\,'') and frequency (ex: ``\,presence of bass frequencies\,'').  This first result confirms previous findings showing that the categorization process relies mostly on sound source identification \cite{dubois2006cognitive,maffiolo_caracterisation_1999}, and is in line with previous studies stating that quality and location play a role in soundscape categorization \cite{aletta2016soundscape}.

\gl{Validation Claim 1}

To ensure the consistency of the analysis, we remove from the study the 4 subjects who do not use the ``\,sound source\,'' strategy, and only keep the source and location labels given by the remaining subjects to describe their clusters.

\subsection{From labels to sound source categories}
\label{sec:fromLab2Cat}

\begin{figure*}[t!]
\centering
\includegraphics[width=.8\paperwidth]{gfxMatlab/xp3_1.eps} 
\caption{\label{fig:xp3_1} (a) Number of time each semantic category has been used to describe a scene; (b) two-dimensional plot of the dimensions 1 and 2 resulting from the MDS analysis with cluster membership; (c) dendrogram resulting of the HCA analysis, x label digits indicates cluster membership while x label colors indicate scene type (green: ideal, red: non-ideal).}
\end{figure*}

We first analysed the labels used by the subjects to describe their clusters. From the descriptions, 23 labels can be isolated (see Table~\ref{tab:association} first column). The labels that designate a same sound source are then grouped into 14 categories (see Table~\ref{tab:association} second column).

Among the 14 categories obtained, 12 of them can be directly linked with sound sources and 2 are referring to ambiguous sounds (``\,noise\,'' and ``\,background\,''). To measure the relative importance of the categories in the categorization process, we count the number of time each category has been used to describe a scene (Figure~\ref{fig:xp3_1}.a). As the categories ``\,noise\,'' and ``\,background\,'' are only used to describe less than 5 scenes, we remove them from the analysis.

It can be seen that some general categories may encompassed others (``\,human\,''>``\,market\,''), which may suggest that two subjects could have designate a same object using different levels of abstraction/specificity. In order to check that each category is effectively used to designate a specific sound source, we count the number of time two categories co-occurred in a same scene. A co-occurrence rate of $100\%$ is obtained for 3 categories associations, namely:  ``\,park\,''$\subset$``\,bird\,'', ``\,market\,''$\subset$``\,human\,'', ``\,horn\,''$\subset$``\,siren\,''. As a results, those categories are merged (see Table~\ref{tab:association} second column). It has to be noted that the co-occurence rates of the associations ``\,water\,''/``\,nature\,'' and ``\,nature\,''/``\,water\,'' are of $36\%$ and $21\%$, showing that those two categories refer to two distinct sound objects.

The remaining 9 categories show that sound source identification operates at diverse semantic levels, from the general (``\,construction\,'', ``\,human\,'') to the specific (``\,horn\,'', ``\,siren\,'', ``\,market\,''). For some categories, as for example  ``\,human\,'',  it seems that subjects mostly choose a higher abstraction level (close to semantic level 1 of the taxonomy) to describe human presence, not needing to distinguish between footsteps or voice noises. The fact that the sound reveal human activity seems to be a sufficient descriptor.

There is a considerable overlap between the event markers of the i- and ni-scenes extracted by statistical analysis (see Table~\ref{tab:markers}), and the categories spontaneously used by the subjects to categorize the scenes. Indeed among the 8 categories that could be linked to our event sound classes (only texture classes could be connected with the category ``\,water\,'' ), 4 are directly linked with \emph{event markers} of semantic level 0 and 1: ``\,horn/siren\,'', ``\,construction\,'', ``\,birds\,'' and ``\,church bell\,''. The marker \textit{male footsteps concrete} is comprised in the category ``\,human\,'', although this category may encompass other human sounds.  Of the event markers identified, only  \textit{bell bike} does not appear in the labels used by subjects.

\gl{Les marquers, trouvés comme étant caractéristique de la qualité d'un environnement, semblent donc aussi intervenir dans un processus plus large de categorization. D'un autre coté, trois classes qui ne semblaient pas jouer de role prédominant dans la distinction des scènes idéales et non idéales font ici leur apparition (``\,traffic\,'', ``\,nature\,'' and ``\,alarm\,''), parmi elles, la categorie traffic semble avoir joué un role important dans la similarité inter-scene, celle-ci charactérisant à elle seule plus de 40 scènes.}

\gl{Validation Claim 2}
\gl{Validation Claim 3}

\subsection{Averaging the categorizations}

In order to gain a better understanding of the way in which class presence\,/ absence affects scene similarity, we generate an ``\,average\,'' of the clusterings performed by the subjects. For each participant, a similarity matrix $\delta_{i,j}$ is defined as: \\

\begin{itemize}
\item $\delta_{i,j}=1$ if scenes $i$ and $j$ are in the same cluster, \\
\item $\delta_{i,j}=0$ otherwise.\\
\end{itemize}

A global similarity matrix is computed by summing the individual similarity matrices, then submitted to a nonmetric  Multi-Dimensional Scaling (MDS) analysis using squared stress1 criterion and 10 replications. A 2-dimensional solution is chosen ($stress=0.07$). To make the emerging clusters explicit, we apply Hierarchical Clustering Analysis (HCA) using average between-groups linkage, with a Euclidean distance metric. HCA analysis is run for 3 to 10 clusters, and a clustering evaluation using the silhouette criterion gives us an optimal number of clusters of 5. Cluster cardinalities are shown on Table~\ref{tab:ClustersHCA}. A two-dimensional plot of the scenes coordinates obtained from the MDS analysis, also showing cluster memberships, is shown on Figure~\ref{fig:xp3_1}.b and the dendrogram resulting of the HCA analysis is shown on Figure~\ref{fig:xp3_1}.c.

\subsection{Using sound categories and sound classes to characterize the markers}

\begin{figure*}[t!]
\centering
\includegraphics[width=.8\paperwidth]{gfxMatlab/xp3_2.eps} 
\caption{\label{fig:xp3_2} For each HCA cluster: (a) Percentage of scenes described by each category; (b) Percentage of scenes simulated using the classes connected to each category; (c) Averaged densities computed considering only the samples belonging to the classes connected to each category; (d) Boxplot of averaged rates resulting of experiments-2 (observations: subjects); (d) Boxplot of averaged $LAeq$ (observations: scenes). For ease of reading, ``\,horn\,'' and ``\,siren\,'' are grouped together.}
\end{figure*}

Since the protocol based on scene simulation lets us know exactly what sound sources are present in the scenes, we can confront the categorization based on sound sources with our analysis of that ground truth.

The categories found in section ~\ref{sec:fromLab2Cat} gives us indications on the sound classes as well as the semantic levels to use to describe the clusters. The chosen sound classes are indicates in Table~\ref{tab:association}.  The descendants of sound classes linked with a category label used by subjects to categorize scenes cover $XX\%$ of the taxonomy at level 3, which shows that a great majority of sound sources are used by listeners in the appreciation of a scene, albeit at various levels of abstraction.

The percentage of scenes labeled with each semantic category (Figure~\ref{fig:xp3_2}.a) is computed for each cluster, as well as the percentage of sound classes (Figure~\ref{fig:xp3_2}.b). Small anomalies, \textit{i.e.} when a label is given and the corresponding class is absent, may be due to 1) the influence of the texture sounds, where some events may be detected, or 2) the fact that subjects may describe clusters using sources that are present in most scenes but not all of them.

\gl{description categories}

Considering the categories, Figure~\ref{fig:xp3_2}.a  shows that two of the clusters resulting from the HCA analysis are formed based on the presence of one category: $100\%$ of the cluster-1 scenes are labelled ``\,birds\,'' and  $100\%$ of the cluster-3 scenes are labelled  ``\,church bell\,''. The remaining clusters seems to be characterized by a combination of multiple categories: clusters-2  ``\,human\,'' ($93\%$) $+$  ``\,church bell\,'' ($80\%$),  clusters-4  ``\,human\,'' ($67\%$) $+$  ``\,construction\,'' ($87\%$) $+$ ``\,traffic\,'' ($80\%$)   and clusters-5  ``\,construction\,'' ($100\%$) $+$ ``\,traffic\,'' ($100\%$) $+$ ``\,alarm\,'' ($61\%$) $+$ ``\,horn/siren\,'' ($57\%$).

\gl{description classes, combien texture, combien event}

\gl{Confrontation}

The ``\,traffic\,'' category exhibits a peculiar behaviour: while classes belonging in that category are present in all clusters, it only appears when the classes \textit{horn} and  \textit{siren} are present in the cluster. This seems to indicate that traffic noises are only considered a nuisance above a certain threshold, and that moderate traffic is accepted as an integral part of an ordinary urban soundscape.

\gl{Validation Claim 5}

\subsection{Categorization based on sound source influence pleasantness and sound level}

As can be seen on Figure~\ref{fig:xp3_1}.c, scene clusters mostly obey the ideal~/ non-ideal distinction: clusters 1-2-3 contain exclusively i-scenes and cluster 5 is made up of $96\,\%$ ni-scenes, while cluster 4 is the only mixed one, with $40\,\%$ i-scenes. A cursory examination shows that this observation carries on to scene descriptions, since the event classes used to characterize scenes in clusters 1-2-3 are mostly positively connotated, whereas for those in cluster 5, all labels are characteristic of ni-scenes, while cluster 4 seems mixed.

To quantify the appraisal of the clusters, the mean average ratings from the rating experiment of \cite{lafayPartI} are computed individually for each cluster (Figure~\ref{fig:xp3_2}.d). A one-way repeated measure ANOVA shows a significant effect of the clusters on the mean averaged rates ($F[4,36]=21, p<0.01$), but \emph{post hoc} analysis shows significant effects only between clusters 1-2-3 vs 4, 1-2-3 vs 5 and 4 vs 5. It therefore seems that, from the point of view of pleasantness, the combinations of semantic categories characterizing clusters 1-2-3 are equivalent. Cluster 4 paints a more complex picture: the occasional presence of i-scene markers (``\,church bell\,'', ``\,birds\,'') and the high density of events belonging to the ``\,human\,'' category operate a clear distinction with the scenes of cluster 5, which are mostly characterized with ni-scene markers. This mixed composition of cluster 4 is reflected in the average pleasantness rating, with a score falling between those of clusters 1-2-3 and that of cluster 5. This observation suggests that adding positively connotated sounds to an \emph{a priori} unpleasant environment can improve its overall pleasantness.

A similar study bearing on the scene acoustic level $LAeq$ (Figure~\ref{fig:xp3_2}.e) yields congruent results ($F[4,67]=234$, $p<0.01$; post hoc analysis: $p_{bonf}<0.01$ for clusters 1-2-3 \emph{vs} 4, 1-2-3 \emph{vs} 5 and 4 \emph{vs} 5). Results indicate that not only does averaged pleasantness of the clusters correlate to their averaged acoustic level, but the latter also seems to be influenced by the type of sources making up the scene.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%    DISCUSSION AND PERSPECTIVES   %%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusions and perspectives}

From the outcome of experiment-2, we conclude that:

\begin{itemize}
\item Sound scene categorization relies greatly on sound source identification.
\item Identification takes places at several semantic levels, and mostly hinges on 8 well-defined categories: ``\,church bell\,'', ``\,birds\,'', ``\,human\,'', ``\,horn\,'', ``\,siren\,'', ``\,alarm\,'', ``\,construction\,'', ``\,traffic\,''.
\item The scene markers coincide greatly with the categories spontaneously used by subjects to describe and characterize scene clusters.
\end{itemize}

\noindent \textbf{Acknowledgements}\\
\setlength{\parindent}{0.7cm} 
Research project partly funded by ANR-11-JS03-005-01. The authors thank the students of the Ecole Centrale de Nantes for their willing participation.

%----------------------------------------------------------------------------------------
%	REFERENCE LIST
%----------------------------------------------------------------------------------------

\References{}{biblio}{actalit}

\end{document}

### 2016-07-28 part1/lafaySimulatedScenePerception1.tex
update part I with NM comments

@@
 \hspace*{8pt}gregoire.lafay@irccyn.ec-nantes.fr\\
-$^{2)}$ Institut de Recherche et Coordination Acoustique/Musique (IRCAM), Paris, France}
+$^{2)}$ STMS Ircam-CNRS-UPMC Institut de Recherche et Coordination Acoustique/Musique, Paris, France}
 
@@
 \Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process.
-Subjects are asked to create a full soundscape by means of a  software dedicated to sound arrangement, coupled with a structured sound data set.
+Subjects are asked to create a full soundscape by means of a  software dedicated to sound \gl{edition}, coupled with a structured sound data set.
 This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes.
@@
 The notion of \emph{soundscape}, introduced by Schafer \cite{schafer_new_1969, schafer1977tuning} in the 1970s as the auditory equivalent to landscape, is a theoretical framework of choice for that research.
-Following this paradigm, a sonic environment is described by focusing on the listener's evaluation, rather than only taking into account the acoustic parameters of the sound scene.
+Following this paradigm, a sonic environment is described by focusing on the listener's evaluation \gl{and perception}, rather than only taking into account the acoustic parameters of the sound scene.
 
-Most soundscape studies follow the \emph{modus operandi} of cognitive psychology \cite{maffiolo_caracterisation_1999}, by first describing the soundscape using both acoustic measurements and perceptive factors, and then investigating the interactions between those descriptors to understand the cognitive processes involved in the perception of the acoustic scenes.
-Soundscape studies are thus interdisciplinary \cite{davies_perception_2013, aletta2016soundscape}, borrowing tools from various fields such as acoustics, psycho-physics or cognitive psychology, but also psycho-linguistics, sociology and data mining.
+Most soundscape studies follow the \emph{modus operandi} of cognitive psychology \cite{maffiolo_caracterisation_1999}, by first describing the soundscape using both acoustic measurements and perceptual factors, and then investigating the interactions between those descriptors to understand the cognitive processes involved in the perception of the acoustic scenes.
+Soundscape studies are thus interdisciplinary \cite{davies_perception_2013, aletta2016soundscape}, borrowing tools from various fields such as acoustics, psychophysics or cognitive psychology, but also psycho-linguistics, sociology and data mining.
 
 While the soundscape approach appears to be a powerful tool to develop perceptively motivated indicators \cite{schulte-fortkamp_soundscape:_2007}, the use of various experimental protocols to assess and measure soundscapes makes the integration of results difficult \cite{davies_perception_2013}.
-Moreover, there still exist no definite consensus in the soundscape community about the acoustic descriptors to be used \cite{aletta2016soundscape}.
+Moreover, there still exist no definite consensus in the soundscape community about the acoustic descriptors to be used \cite{brocolini2012prediction,aletta2016soundscape}.
 We believe that soundscapes studies greatly suffer from one important issue: to the best of our knowledge, there is no morphological definition of a soundscape, and therefore no standardized way to decompose it into distinct elements.
-It is thus difficult to study precisely which component of the environment is the cause of potential variations in the perceptive measurements.
+It is thus difficult to study precisely which component of the environment is the cause of potential variations in the perceptual measurements.
 
@@
 The simulation provides us with simulated scenes fully annotated in terms of source presence and sound levels.
-Those annotations enable us to ground the analyses of further perceptive descriptors obtained via standard perceptive evaluations of the simulated scenes, as well as to study the contribution of each sound source.
+Those annotations enable us to ground the analyses of further perceptual descriptors obtained via standard perceptual evaluations of the simulated scenes, as well as to study the contribution of each sound source.
 
-As a case study to demonstrate the interest of that new approach, we focus in this paper on the issue of urban soundscape pleasantness as tackled in \cite{guastavino_ideal_2006}.
+As a case study to demonstrate the interest of that new approach, we focus in this paper on the issue of urban soundscape pleasantness as tackled in \cite{guastavino_ideal_2006} \gl{NM comments}.
 
-Only during the 1980s did policy-makers start taking into account the link between \textit{noise} and pollution, considering noise as a significant degradation of the quality of life.
+Only during the 1980s did policy-makers start taking into account the link between \gl{``\,community noise\,''} and pollution, considering noise as a significant degradation of the life quality.
 To fight this phenomenon, the first approach consisted in identifying unwanted sounds, and lowering their intensities.
@@
 
-However, it is now widely accepted that acoustic comfort is a complex concept that cannot be described solely with the help of objective acoustical measures such as $LA_{(eq,V)}$, but needs to take into account other perceptive factors \cite{schulte-fortkamp_soundscape_2006, aletta2016soundscape, Yang2005211, kang_semantic_2010}.
-``\,Noise\,''  is a cognitive object which depends upon a listener's appreciation and the context in which it is heard.
+However, it is now widely accepted that acoustic comfort is a complex concept that cannot be described solely with the help of objective acoustical measures such as $L_{Aeq}$, but needs to take into account other perceptual factors \cite{schulte-fortkamp_soundscape_2006, aletta2016soundscape, Yang2005211, kang_semantic_2010}.
+``\,Noise\,''  is a cognitive object which depends upon a listener's \gl{assessment} and the context in which it is heard.
 Many urban ``\,noises\,'', such as that of a siren, can annoy as well as warn of a danger, and many town districts are appreciated because of a lively and animated atmosphere which often results in higher sound levels.
@@
 
-The soundscape paradigm has been employed for two decades to study the perception of urban acoustic scenes (see \cite{dubois2006cognitive} and  \cite{aletta2016soundscape} for extensive reviews), however, due to the limitations we have pointed out, experiments have failed to provide  urban planners with generic and clear indicators or models able to predict sound environment quality \cite{schulte-fortkamp_soundscape:_2007, aletta2016soundscape}.
-Recently, several projects such as the European Cooperation in Science and Technology Action (TD0804: soundscape of European Cities and Landscapes) or the Positive Soundscape project \cite{davies_perception_2013} have been undertaken to standardize soundscape assessment and descriptors \cite{schulte2013soundscape}, but those matters are still considered as open problems.
+The soundscape paradigm has been employed for two decades to study the perception of urban acoustic scenes (see \cite{kang2006urban,dubois2006cognitive,aletta2016soundscape} for extensive reviews), however, due to the limitations we have pointed out, experiments have failed to provide  urban planners with generic and clear indicators or models able to predict sound environment quality \cite{schulte-fortkamp_soundscape:_2007, aletta2016soundscape}.
+Recently, several projects such as the European Cooperation in Science and Technology Action (TD0804: soundscape of European Cities and Landscapes) or the Positive Soundscape project \cite{davies_perception_2013} have been undertaken to standardize soundscape assessment and descriptors \cite{schulte2013soundscape}, but those matters are still considered as open problems \cite{ribeiro2013heart}.
 We explore in this paper what additional insights our proposed ``\,generative\,'' approach to soundscape perception analysis can bring to light.
@@
 We then describe the simulation process, and present the results of the case study.
-The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale.
+The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale \gl{NM comments}.
 
-Results show that 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding, 2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterise the appraisal than global descriptors.
+Results show that \gl{NM comments} 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding, 2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterise the appraisal than global descriptors.
 
@@
 
-Two approaches, named in this paper \textit{categorical} and \textit{dimensional}, are commonly used to gain knowledge on soundscape perception.
+Two approaches, named in this paper \textit{categorical} and \textit{dimensional}, are commonly used to gain knowledge on soundscape perception \gl{NM comments}.
 
@@
 Maffiolo \cite{maffiolo_caracterisation_1999} shows the existence of two generic soundscape categories, namely the ``\,event sequences\,'' and the ``\,amorphous sequences\,''.
-Similar findings were made by \cite{guastavino_ideal_2006}.
+\gl{This distinction is also observed in \cite{guastavino_ideal_2006}}.
 Guastavino \cite{guastavino_categorization_2007} as well as Niessen \emph{et al.} \cite{niessen_categories_2010}  show the prominence of specific sound source categories in soundscape perception: ``\,natural\,'', ``\,human\,'', ``\,traffic\,'' and ``\,mechanical\,''.
-Other studies not using the \textit{categorical} approach reveal that sound source identification underlines the perception of others factors, for example Kuwano \emph{et al.} \cite{kuwano_memory_2003-1} show that overall judgment of loudness of a sound environment sequence is not statistically different from average instantaneous judgment of the recalled sound events of the sequence.
+Other studies not using \gl{NM comments} the \textit{categorical} approach reveal that sound source identification underlines the perception of others factors, for example Kuwano \emph{et al.} \cite{kuwano_memory_2003-1} show that overall judgment of loudness of a sound environment sequence is not statistically different from average instantaneous judgment of the recalled sound events of the sequence \gl{NM comments}.
 As reported by \cite{davies_perception_2013}, while the categorical approach does help to understand how a soundscape is composed, it does not provide information on how soundscape components are perceived and influence the qualitative evaluation.
@@
 From those descriptors, it is possible to use dimensional analysis techniques such as Principal Component Analysis to find relevant perceptual axes \cite{cain_development_2013}, or to connect those perceptive outcomes with objective acoustics measurements using various statistical analysis or modelling techniques such as multiple linear regression \cite{lavandier2006contribution,ricciardi2015sound} or artificial neural network \cite{yu2009modeling}.
-Several semantic scales show promising results when it commes to modeling perceived affective quality, namely ``\,Calmness\,'' and ``\,Vibrancy\,'' \cite{cain_development_2013, axelsson2010principal}, or ``\,Pleasantness\,'' and ``\,Eventfulness\,'' \cite{axelsson2010principal}; but it has been shown that others context dependent (``\,Appropriateness\,'',``\,Familiarity\,'') or music related (``\,Music-likeness\,'' ) scales provide useful complementary information on soundscape perception \cite{aletta2016soundscape}.
-In addition, \cite{yu2009modeling} shows that multiple factors such as physical, behavioral, social, demographical, and psychological are involved when evaluating acoustic comfort.
-Similar findings are made in \cite{yu2010factors} showing that age and education level significantly influence the sound preference.
+Several semantic scales show promising results when it comes to modeling perceived affective quality, namely ``\,Calmness\,'' and ``\,Vibrancy\,'' \cite{cain_development_2013, axelsson2010principal}, or ``\,Pleasantness\,'' and ``\,Eventfulness\,'' \cite{axelsson2010principal}; but it has been shown that others context dependent (``\,Appropriateness\,'',``\,Familiarity\,'') or music related (``\,Music-likeness\,'' ) scales provide useful complementary information on soundscape perception \cite{aletta2016soundscape}.
+In addition, several studies have shown that multiple factors such as physical, behavioral, social, demographical, and psychological are involved when evaluating acoustic comfort \cite{marquis2005noise, yu2009modeling}.
+\gl{In particular} \cite{yu2010factors} shows that age and education level significantly influence the sound preference \gl{NM comments, attention \cite{marquis2005noise} est une review}.
 
-Studies investigating the influence of sound sources in soundscape perception show that global sound level (perceived or measured) cannot entirely explain the quality of a sonic environment.
+\gl{Recent} studies investigating the influence of sound sources in soundscape perception show that global sound level (perceived or measured) cannot entirely explain the quality of a sonic environment \gl{NM comments, plus développer, ajout article ?}.
 \cite{guyot2005urban} shows that a rise in soundscape quality between two street soundscapes having a similar time-ratio of traffic sounds and a similar global sound level may result from the addition of a positively connotated sound source, e.g.
-belonging to the ``\,human\,'' category.
+belonging to the ``\,human\,'' category \gl{NM comments: eg means exempli gratia (par exemple)}.
 Using multiple linear regression to predict unpleasantness from descriptors characterizing global sound level as well as time presence and sound level of each source, \cite{lavandier2006contribution} shows that the percentage of explained variance increases when the source descriptors are added to the global sound level.
@@
 
-In order to be able to study the relative perceptual importance of each sound source, it is beneficial to first gain knowledge about the presence and level of those sound sources within the scene.
+In order to be able to study the relative perceptual importance of each sound source, it is beneficial to first gain knowledge about the presence and level of those sound sources within \gl{inside ?} the scene.
 Acquiring such data may be a serious challenge.
@@
 
-To simulate a verisimilar sound environment, we need to provide the subject with a reasonable and representative number of isolated sounds.
-This dataset represents the ``\,sonic world\,'', or more precisely a semantic discretization of it in terms of  ``\,sound sources categories\,''.
-The design of the dataset is largely inspired by early work by Rosch \cite{roach1978cognition} on conceptual categorization which states that isolated sound classification operates at several levels of abstraction \cite{houix_lexical_2012}.
-Following this consideration, the dataset is structured along two dimensions:
+\gl{To simulate a verisimilar sound environment, we need to provide the subject with a reasonable and representative number of isolated sounds.
+This dataset represents the ``\,sonic world\,'', or more precisely a semantic discretization of it in terms of  ``\,sound sources categories\,'', where the term ``\,category\,'' refers to a group of objects (in our case sound sources) considered as equivalent.}
+
+\gl{The design of the dataset is largely inspired by early work by Rosch \cite{roach1978cognition} on the formation of categories of concrete objects.
+According to Rosch, the category system may be viewed as a two dimensional space, where the ``\,vertical\,'' dimension deals with the inclusiveness of the categories (\emph{e.g.} how are ordered ``\,human\,'', ``\,voice\,'' and ``\,laugh\,'') while the ``\,horizontal\,'' dimension concerns the segmentation of categories lying in the same level of inclusiveness (\emph{e.g.} ``\,voice\,'' and ``\,car\,'').
+The ``\,vertical\,'' dimension is a taxonomy, where categories are related to one another by means of class inclusion.
+Rosch formalized three levels of abstraction to structure the ``\,vertical\,'' dimension: superordinate, basic and subordinate.
+The greater the inclusiveness of a category the higher the abstraction level.
+Several studies focusing on sound classification  have already used such hierarchical system \cite{houix_lexical_2012}.
+Following these considerations, our dataset is also structured along two dimensions:}
 
 \begin{itemize}
-\item object dimension: where we make the difference between distinct sounds sources such as \textit{voice}, or \textit{car}. 
-\item abstraction dimension: where we make the difference between distinct level of abstraction use to describe a sound (``\,human\,''>``\,voice\,''>``\,laugh\,''). 
+\item object dimension: where we make the difference between distinct sounds sources such as ``\,voice\,'', or ``\,car\,'' and which is closely related to the ``\,horizontal\,'' dimension of Rosh.
+\item abstraction dimension: where we make the difference between distinct level of abstraction use to describe a sound (``\,human\,''>``\,voice\,''>``\,laugh\,''), and which is closely related to the ``\,vertical\,'' dimension of Rosh.
 \end{itemize}
@@
 
-A soundscape  is here understood as being  ``\,a skeleton of events on a bed of texture\,'' \cite{nelken_ear_2013}, where sound events are short isolated sounds such as \textit{car passing} or \textit{baby crying}, whereas sound textures are long stationary  sounds such as \textit{traffic} or \textit{crowd sound}, in which the physical characteristics remain stable over long periods of time~\cite{saint-arnaud_classification_1995}.
+\gl{As stated is section \ref{statoftheart}}, a soundscape  is here understood as being  ``\,a skeleton of events on a bed of texture\,'' \cite{nelken_ear_2013}, where sound events are short isolated sounds such as \textit{car passing} or \textit{baby crying}, whereas sound textures are long stationary  sounds such as \textit{traffic} or \textit{crowd sound}, in which the physical characteristics remain stable over long periods of time~\cite{saint-arnaud_classification_1995}.
 Several studies point out the fact that textures and events drive two distinct cognitive processes \cite{maffiolo_caracterisation_1999, mcdermott2013summary}.
@@
 \begin{description}
-\item[Sample]: a recording of an isolated sound event or texture.
+\item[Sample]: the recording of an isolated sound event or texture.
 \item[Sound class]: either a collection of samples or a collection of sub-classes, depending of the semantic level of the class.
@@
 The sample selection process is designed to rely only on the listening of sounds themselves, in order not to influence subjects with \textit{a priori} associated semantic values.
-To explore the dataset without any written textual help, we have designed a selection interface where each leaf class is depicted by a point in 2D space \cite{lafay:hal-01300399}.
+To explore the dataset without any written textual help, we have designed a selection interface where each leaf class is depicted by a point in 2D space, \gl{which was previously tested in an other study} \cite{lafayJAES}.
 Subjects may only interact with the leaf classes of the datasets but the location of the points in space is constrained by the taxonomy, \textit{i.e.} subclasses belonging to a same class are close to one another.
@@
 Each track is modeled by a temporal sequence of sound samples belonging to the same sound class.
-Thus, each track is related to one specific leaf sound class (see \ref{subsection:Sound dataset}).
+Thus, each track is related to one specific leaf sound class (see Section \ref{sec:SoundDataset}).
 To generate a soundscape, subjects may interact with the track parameters, but not with specific samples; for example, if a subject wishes to put a sequence of car sounds, he may specify how often (on average) cars pass by, and how loud they should be on average, but he cannot control each event individually.
@@
 \item \textit{Start and Stop position} (sec): allows the subject to set the beginning and the end times of a temporal sequence.
-\item \textit{Inter onset spacing time} (sec): similarly to the sound level, inter onset spacing times between events are drawn from a normal distribution that may be parametrized by the subject for each event track.
-\item \textit{Sample fade In/Out} (sec): allows the subject to set a same fade effect individually on each sample of an event track.
+\item \textit{Inter onset spacing time} (sec): (event tracks only) similarly to the sound level, inter onset spacing times between events are drawn from a normal distribution that may be parametrized by the subject for each event track.
+\item \textit{Sample fade In/Out} (sec): (event tracks only) allows the subject to set a same fade effect individually on each sample of an event track.
 \end{enumerate}
@@
 Considering a soundscape as a sum of isolated samples is in line with the Auditory Scene Analysis (ASA) theory \cite{bregman1994auditory}, which states that auditory stimuli are decomposed into ``\,auditory objects\,'' called ``\,streams\,'', which may be regarded as sequences of auditory events emitted by putative sound sources \cite{ciocca2007auditory,winkler2009modeling}, and also with recent neuroscience studies suggesting that the primary auditory cortex (A1) produces representations of such ``\,auditory objects\,'' that act as bases for high level cognitive processing \cite{nelken2008neurons}.
-Thus, the track can be seen as the generative counterpart of the stream, which is rather a cognitive object.
+Thus, the track can be seen as the generative counterpart of the stream, which is rather a cognitive object.\gl{NM comments}
 
@@
 The two labels have been chosen to compare the results with those of a previous study addressing ideal urban sound environment representations \cite{guastavino_ideal_2006}.
-The simulation protocol used in this experiment was previously tested and validated thanks to a pilot study~\cite{soundscape3} run with 10 subjects.
+The simulation protocol used in this experiment was previously tested and validated thanks to a pilot study~\cite{soundscape3}.
+
+%run with 10 subjects.
 
@@
 Supplementary information including the experiment webpage and raw data for each experiment can be found on the companion website \footnote{\url{http://www.irccyn.ec-nantes.fr/~lagrange/demonstrations/simScene.html}}.
+\gl{NM: Pas d'accord commentaires section IV}
 
@@
 Among them, 260 events and 72 textures are recorded by the authors using a \textit{Audio Technica AT8035} shotgun microphone (to isolate sounds from their background) connected to a \textit{ZOOM H4n} recorder.
-The remaining come from existing sound banks.
+The remaining comes from existing sound banks \gl{NM préciser lesquels ?}.
 All sounds are normalized to the same root mean square (RMS) level.
@@
 For example, considering the two subclasses of the semantic level 2 \textit{car-passing} and \textit{car-starting}, which both belong to the same class \textit{car}, we count 2 distinct sound classes for semantic levels 2 and 3, but only 1 sound class for semantic levels 0 and 1.
-\item[$Density$:] logarithm of the number of event samples per chunk of 1 second, averaged over the scene duration.
+\item[$Density$ ($D$):] logarithm of the number of event samples per chunk of 1 second, averaged over the scene duration.
 Density may be computed considering all sound classes, or only a subset of them.
 If nothing is specified, then the density accounts for all classes.
-\item[$LA_{(eq,V)}$:]  Decibel of the RMS voltage relative to 1 volt, computed every second and averaged over the scene duration.
+\gl{Empirically, we observed that the original density distributions were skewed.
+The use of the logarithm compression allowed us to reduce this skewness.}
+\item[$Sound level$ ($L$):]  Decibel of the RMS voltage relative to 1 volt, computed every second and averaged over the scene duration.
 An A-weighting filtering is performed before the RMS computation.
-Computing other statistical descriptors, namely $LA_{(min,V)}$, $LA_{(max,V)}$ and $LA_{(10-90,V)}$, shows that they present a high correlation with $LA_{(eq,V)}$ ($r\geq0.85$, $p<0.01$), therefore only the latter is considered.
+Computing other statistical descriptors, namely the minimum, maximum or the 10-90th quantiles, shows that they all present a high correlation with the mean ($r\geq0.85$, $p<0.01$), therefore only the latter is considered.
 \end{description}
@@
 
-\subsection{Experiment-1 setup}
-\subsubsection*{Task}
+\subsection{Set-up Experiment-1}
+\subsubsection*{Procedure}
 
@@
 \label{sec:Experiment-2}
-\subsubsection*{Task}
+\subsubsection*{Procedure}
 
-Subjects are asked to evaluate the pleasantness of the 80 simulated scenes using a bipolar scale of 7 points going from -3 (non ideal, very unpleasant) to +3 (ideal, very pleasant).
+Subjects are asked to evaluate the pleasantness of the 80 simulated scenes using a bipolar scale (also known as Likert scale) of 7 points going from -3 (non ideal, very unpleasant) to +3 (ideal, very pleasant).
 The first ten scenes are used to give the subject an idea of the pleasantness range of the scenes.
@@
 Scenes are presented in a randomized order, but the ten calibration scenes are five i-scenes and five ni-scenes.
+\gl{NM comments: la randomization est déjà précisée.}
 
@@
 The experiment is run simultaneously with the 10 subjects in a calm environment.
-Audio is presented diotically to each participant via \textit{BeyerDynamic DT 990 Pro} semi open headphones at -12 dB (FS).
+Audio is presented diotically to each participant via \textit{BeyerDynamic DT 990 Pro} semi open headphones at -12 dB (full scale).
 One experimenter is always present to give instructions and answer queries if needed.
@@
 The remaining 57.5\% did not report difficulties.
+
 Those results tend to confirm that both the data set and the selection interface are well suited for the experiment.
@@
 Figure~\ref{xp2_1}.a shows the distributions of the averaged ratings considering scenes as observations.
-Four scenes (2 i-scenes an 2 ni-scenes) appear to be ambiguous, and are henceforth removed from the analysis, as well as their antagonist scenes.
+Four scenes (2 i-scenes and 2 ni-scenes) appear to be ambiguous, and are henceforth removed from the analysis, as well as their antagonist scenes.
 This operation leaves us with 36 i-scenes and 36 ni-scenes.
@@
 Figure~\ref{xp2_1}.c shows the distributions of the averaged ratings, considering subjects as observations, without the scene outliers.
-
 It is clear that the pleasantness levels of the scenes, as intended by experiment-1 subjects, are well perceived by experiment-2 subjects ($p<0.01$).
@@
 
-\subsection{Sound level, density and diversity}
+\subsection{Density, sound level,  and diversity}
 \label{sec:levelDensityDiversity}
@@
 \includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_deSoLv_1.eps} 
-\caption{\label{fig:xp1_deSoLv_1}  Boxplot of  the scene (a) $Density$, (b) $LA_{(eq,V)}$, (c) $Density$ \emph{vs} rating and (d) $LA_{(eq,V)}$ \emph{vs} rating}
+\caption{\label{fig:xp1_deSoLv_1}  Boxplot of (a) the scene densities ($D$) and (b) the scene sound levels ($L$). Scatter plot of (c) $D$ \emph{vs} ratings and (d) $L$ \emph{vs} rating}
 \end{figure}
@@
 \centering
-\begin{tabular}{l r r} 
-                & $LA_{(eq,V)}$    & $Density$    \\
+\begin{tabular}{l c c} 
+                & $L$                          & $D$    \\
 \hline
@@
 \vspace{0.5mm}
-\caption{\label{tab:numGlobal} Correlations computed between ratings \emph{vs} $LA_{(eq,V)}$ and $Density$}
+\caption{\label{tab:numGlobal} Correlations computed between ratings \emph{vs} sound level ($L$) and event density ($D$)}
 \end{table}
@@
 
-The event density is slightly higher for ni-scenes than for i-scenes (Figure~\ref{fig:xp1_deSoLv_1}.a).
-An average density of $4.4$ for the i-scenes and $6$ for the ni-scenes is found, with a significant effect of the scene type ($p<0.05$).
-The density varies widely from subject to subject.
-Concerning $LA_{(eq,V)}$ (Figure~\ref{fig:xp1_deSoLv_1}.b), the  sound level is significantly higher for ni-scenes than for i-scenes ($p<0.01$).
+The event density ($D$) is slightly higher for ni-scenes than for i-scenes (Figure~\ref{fig:xp1_deSoLv_1}.a).
+The averaged $D$ are $1.2$ for the i-scenes, and $1.6$ for the ni-scenes, with a significant effect of the scene type ($p<0.05$).
+$D$ varies widely from subject to subject.
 
-To study the ability of those descriptors to characterize the appraisal, we compute the correlations between the ratings, the density and the $LA_{(eq,V)}$, considering all the scenes, as well as the i- and ni-scenes separately.
-Results are shown in Table~\ref{tab:numGlobal}, as well as in Figures~\ref{fig:xp1_deSoLv_1}.c and~\ref{fig:xp1_deSoLv_1}.d;
+Concerning the sound level (Figure~\ref{fig:xp1_deSoLv_1}.b), $L$ is significantly higher for ni-scenes than for i-scenes ($p<0.01$).
 
-Not surprisingly, these results show that ni-scenes are on average louder than the i-scenes, and the strong correlation between $LA_{(eq,V)}$ and the ratings indicates that the sound level does influence the pleasantness.
-However, the correlations computed separately for the i- and ni-scenes suggest $LA_{(eq,V)}$ does not characterize the notion of pleasantness in a equal manner for the two type of scenes:  there is still a good correlation between $LA_{(eq,V)}$ and the ratings for the ni-scenes, but there is no correlation if we take only the i-scenes.
+To study the ability of those descriptors to characterize the appraisal, we compute the correlations between the ratings, $D$ and $L$, considering separately the i- and ni-scenes, as wel as all the scenes altogether.
+Results are shown in Table~\ref{tab:numGlobal} and in Figures~\ref{fig:xp1_deSoLv_1}.c and~\ref{fig:xp1_deSoLv_1}.d; \gl{NM comments: je ne pense pas qu'une MANOVA s'utilise comme cela}
+
+Not surprisingly, these results show that ni-scenes are on average louder than i-scenes.
+The strong correlation of $-76\%$ between $L$ and the ratings indicates that the sound level does influence the pleasantness.
+However, the correlations computed separately for the i- and ni-scenes suggest $L$ does not characterize the notion of pleasantness in a equal manner for the two type of scenes:  there is still a good correlation between $L$ and the ratings for the ni-scenes, but there is no correlation if we take only the i-scenes.
 This is in line with the trend observed by \cite{lavandier2006contribution} that sound level may be a sufficient indicator of comfort for negatively rated acoustics scenes, but not for positively rated ones.
 
-Considering the event densities, ni-scenes seem to be slightly more populated that i-scenes.
-However, it appears that a global measure of sound event density fails to characterize the appraisal if all the scenes are considered at once.
-If we consider the two types of scenes separately, the correlation remains weak for the ni-scenes, and is non-existent for the i-scenes.
+\gl{Considering the event density, ni-scenes seem to be slightly more populated that i-scenes.
+However, the weak correlation between $D$ and the ratings ($-0.35$) shows that a global measure of $D$ fails to characterize the appraisal if all the scenes are considered altogether.
+If we consider the two types of scenes separately, the correlation remains moderate for the ni-scenes, and is non-existent for the i-scenes.
 This may suggest that either the number of events is an irrelevant indicator, or that only the densities of specific event classes are involved.
-Still, although there is no clear correlation between density and pleasantness for the ni-scenes, the density seems to play an indirect role as it is correlated with $LA_{(eq,V)}$ ($0.62$, $p<0.01$).
+Still, although there is no clear correlation between $D$ and pleasantness for the ni-scenes, $D$ seems to play an indirect role as it is correlated with $L$ ($0.62$, $p<0.01$).
+In other words, $D$ does have an impact on the appraisal only when the addition of new events increases the sound level.}
 
@@
 Diversity remains constant across semantic levels for texture classes of i- and ni-scenes, with no statistical differences.
-The correlations computed between the event diversities and the averaged pleasantness ratings remain weak ($r<0.45$, $p<0.01$), irrespective of the considered semantic level.
-If we look at the two type of scenes separately, no correlation is observed.
-This suggests that diversity plays no role in pleasantness evaluation.
+The correlations computed between the event diversities and the averaged pleasantness ratings remain weak or moderate ($r<0.45$ \gl{NM comments: non}, $p<0.01$), irrespective of the considered semantic level.
+\gl{NM comments: non} If we look at the two type of scene separately, no correlation is observed.
+\gl{Although the diversity varies between the i- and ni-scenes, this last result shows that it is not a good indicator to precisely characterize the pleasantness.}
 
@@
 
-However, some differences have to be reported: \cite{guastavino_ideal_2006} reports that ``\,public transports\,'' are typical sounds of an ideal urban environment.
-This observation is attributed to the fact that urban environment perception is driven by social values.
+\gl{However, some differences have to be reported: \cite{guastavino_ideal_2006} reports that ``\,public transports\,'' are typical sounds of an ideal urban environment.
+This observation is attributed to the fact that urban environment perception is driven by social considerations.
 Public transport sounds are thus better accepted than those of private vehicles, as they are usually positively connotated.
-Results shown in Figure~\ref{fig:xp1_class_1} tend to qualify this claim  by showing that sounds of public transports (\textit{bus} and \textit{train}) were chosen by $27.5\%$ of the subjects for the i-scenes and by $45\%$ of the subjects for the ni-scenes.
+Results shown in Figure~\ref{fig:xp1_class_1} tend to qualify \gl{NM comments: non} this claim  by showing that sounds of public transports (\textit{bus} and \textit{train}) were chosen by $28\%$ of the subjects for the i-scenes and by $42\%$ of the subjects for the ni-scenes.
 It is important to note that in \cite{guastavino_ideal_2006} subjects are asked to recall an environment in memory, that is, without any sound support.
-Despite the fact that the idea of public transport sounds are well accepted due to societal considerations (\textit{bus} sounds correspond indeed to $27.5\%$ of the event classes used in the i-scenes, as much as \textit{bicycle} sounds and more than any other vehicle sounds), the sound itself still remains similar to that of any other vehicle.
-Thus the class \textit{public\_transport} does appear in ni-scenes, more than \textit{car} or \textit{truck}.
-Similar findings showing that bus presence has a negative effect on soundscape appraisal are reported in \cite{lavandier2006contribution}.
+Tacking the reality of the sound into account, results show that public transport sounds cannot be considered as peculiar to an ideal urban soundscape.
+The class \textit{public\_transport} does appear in ni-scenes, more than \textit{car} or \textit{truck}.
+Similar findings showing that bus presence may have a negative effect on soundscape appraisal are reported in \cite{lavandier2006contribution}.
+One should note that the data still does not contradict the idea that public transport sounds are well accepted: \textit{bus} sounds were chosen by $25\%$ of the subjects for the i-scenes, circa as much as \textit{bike} sounds and more than any other vehicle sounds.
+}
 
@@
 \hline
-$LA_{(eq,V)\_m}$    & 0.19  ($p=0.28$)           & \textbf{-0.60} ($p<0.01$) \\
-$LA_{(eq,V)\_b}$    & \textbf{-0.51} ($p<0.01$)  & -0.16 ($p=0.36$) \\
-$LA_{(eq,V)\_m-b}$  & \textbf{0.69} ($p<0.01$)   & -0.31 ($p=0.06$) \\
-$Density_{m}$       & -0.20  ($p=0.24$)          & \textbf{-0.58} ($p<0.01$)  \\
-$Density_{b}$       & -0.23  ($p=0.18$)          & -0.30 ($p=0.07$)  \\
+$L_m$         & 0.19  ($p=0.28$)           & \textbf{-0.60} ($p<0.01$) \\
+$L_b$         & \textbf{-0.51} ($p<0.01$)  & -0.16 ($p=0.36$) \\
+$L_{m-b}$     & \textbf{0.69} ($p<0.01$)   & -0.31 ($p=0.06$) \\
+$D_{m}$       & -0.20  ($p=0.24$)          & \textbf{-0.58} ($p<0.01$)  \\
+$D_{b}$       & -0.23  ($p=0.18$)          & -0.30 ($p=0.07$)  \\
 \hline
@@
 \vspace{0.5mm}
-\caption{\label{tab:numMarkers} Correlations computed between rating \emph{vs.} $LA_{(eq,V)\_m}$, $LA_{(eq,V)\_b}$, $LA_{(eq,V)\_m-b}$ and $Density_m$.}
+\caption{\label{tab:numMarkers} Correlations computed between rating \emph{vs.} $L_m$, $L_b$, $L_{m-b}$ and $D_m$.}
 \end{table}

### 2016-08-01 part1/lafaySimulatedScenePerception1.tex
udpate NM comments

@@
 \newcommand{\gl}[1]{\textcolor{blue}{GL : #1}}
-\newcommand{\todo}[1]{\textcolor{red}{todo : #1}}
+\newcommand{\nm}[1]{\textcolor{red}{nm : #1}}
 \newcommand{\ml}[1]{\textcolor{magenta}{ML : #1}}
@@
 
-As a case study to demonstrate the interest of that new approach, we focus in this paper on the issue of urban soundscape pleasantness as tackled in \cite{guastavino_ideal_2006} \gl{NM comments}.
+As a case study to demonstrate the interest of that new approach, we focus in this paper on the issue of urban soundscape pleasantness as tackled in \cite{guastavino_ideal_2006} \nm{dans [14], il est plutot question de ideal / non ideal urban soundscapes .....
+}.
 
@@
 We then describe the simulation process, and present the results of the case study.
-The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale \gl{NM comments}.
+The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale.
+\nm{en revoyant cette description, je me dis qu'il y a pt-etre un petit probleme methodologique dans la mesure ou l'on contruit les scenes selon un attribut semantique (l'idealite) et on les evalue sur une autre echelle (l'agrement) .....
+Je pense qu'il faudra bien argumenter c point plus bas dans le texte}
 
-Results show that \gl{NM comments} 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding, 2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterise the appraisal than global descriptors.
+Results show that \nm{Cette section sur les resultats est etonnemment placee, et a priori, de maniere pas tres orthodoxe.
+Je verrai plutot cette synthese des resultats dans l'abstract ou dans la conclusion ....} 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding, 2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterize the appraisal than global descriptors.
 
@@
 
-Two approaches, named in this paper \textit{categorical} and \textit{dimensional}, are commonly used to gain knowledge on soundscape perception \gl{NM comments}.
+Two approaches, named in this paper \textit{categorical} and \textit{dimensional}, are commonly used to gain knowledge on soundscape perception \nm{this is not really specific to soundscape ...
+but a global approach of perception mechanisms.
+We can find this distinction also, for instance, in the works about musical timbre}.
 
@@
 Guastavino \cite{guastavino_categorization_2007} as well as Niessen \emph{et al.} \cite{niessen_categories_2010}  show the prominence of specific sound source categories in soundscape perception: ``\,natural\,'', ``\,human\,'', ``\,traffic\,'' and ``\,mechanical\,''.
-Other studies not using \gl{NM comments} the \textit{categorical} approach reveal that sound source identification underlines the perception of others factors, for example Kuwano \emph{et al.} \cite{kuwano_memory_2003-1} show that overall judgment of loudness of a sound environment sequence is not statistically different from average instantaneous judgment of the recalled sound events of the sequence \gl{NM comments}.
+Other studies not using the \textit{categorical} approach reveal that sound source identification underlines the perception of others factors, for example Kuwano \emph{et al.} \cite{kuwano_memory_2003-1} show that overall judgment of loudness of a sound environment sequence is not statistically different from average instantaneous judgment of the recalled sound events of the sequence \nm{Cet exemple n'est pas completement clair pour moi :
+certes, la loudness est bien une dimension continue (et donc en cela rentre bien dans ta terminologie de depart) mais je ne vois pas trop en quoi elle a un rapport avec la question de l'identification des sources ...
+a preciser, je pense.}.
 As reported by \cite{davies_perception_2013}, while the categorical approach does help to understand how a soundscape is composed, it does not provide information on how soundscape components are perceived and influence the qualitative evaluation.
@@
 In addition, several studies have shown that multiple factors such as physical, behavioral, social, demographical, and psychological are involved when evaluating acoustic comfort \cite{marquis2005noise, yu2009modeling}.
-\gl{In particular} \cite{yu2010factors} shows that age and education level significantly influence the sound preference \gl{NM comments, attention \cite{marquis2005noise} est une review}.
+\gl{In particular} \cite{yu2010factors} shows that age and education level significantly influence the sound preference \nm{Je pense qu'il faut etre un peu plus critique sur ces questions la (et celles developpees dans la phrase precedente) et sur ce sujet qu'il faut absolument citer le papier de synthese de C.
+Marquis-Favre :} \gl{Attention \cite{marquis2005noise} est une review ...}.
 
-\gl{Recent} studies investigating the influence of sound sources in soundscape perception show that global sound level (perceived or measured) cannot entirely explain the quality of a sonic environment \gl{NM comments, plus développer, ajout article ?}.
-\cite{guyot2005urban} shows that a rise in soundscape quality between two street soundscapes having a similar time-ratio of traffic sounds and a similar global sound level may result from the addition of a positively connotated sound source, e.g.
-belonging to the ``\,human\,'' category \gl{NM comments: eg means exempli gratia (par exemple)}.
+\gl{Recent} studies investigating the influence of sound sources in soundscape perception show that global sound level (perceived or measured) cannot entirely explain the quality of a sonic environment.
+\gl{voir remarque precedente; sur cette question, je privilegierais plutot un artcile de revue (AAA), voire un bouquin (Kang,2006) qu'un proc.
+de conf.
+...} \cite{guyot2005urban} shows that a rise in soundscape quality between two street soundscapes having a similar time-ratio of traffic sounds and a similar global sound level may result from the addition of a positively connotated sound source, e.g.
+belonging to the ``\,human\,'' category \nm{mais pas que de la categorie human; cf.
+le resultat que tu evoques  plus bas de [13] avec les sources type "public transport" egalement connotes plus positif que des "vehicules legers" du fait de leur caractere ecologique et collectif plus important que la voiture individuelle} \gl{eg means exempli gratia (par exemple), de plus nous montrons que bien  que les bus soient acceptés (presence dans les i-scenes), ils ne sont pas caracteristiques d'un environnement ideal (presence dans les ni-scenes)}.
 Using multiple linear regression to predict unpleasantness from descriptors characterizing global sound level as well as time presence and sound level of each source, \cite{lavandier2006contribution} shows that the percentage of explained variance increases when the source descriptors are added to the global sound level.
@@
 
-In order to be able to study the relative perceptual importance of each sound source, it is beneficial to first gain knowledge about the presence and level of those sound sources within \gl{inside ?} the scene.
+In order to be able to study the relative perceptual importance of each sound source, it is beneficial to first gain knowledge about the presence and level of those sound sources within \nm{inside ?} the scene.
 Acquiring such data may be a serious challenge.
@@
 \gl{To simulate a verisimilar sound environment, we need to provide the subject with a reasonable and representative number of isolated sounds.
-This dataset represents the ``\,sonic world\,'', or more precisely a semantic discretization of it in terms of  ``\,sound sources categories\,'', where the term ``\,category\,'' refers to a group of objects (in our case sound sources) considered as equivalent.}
+This dataset represents the ``\,sonic world\,'', or more precisely a semantic discretization of it in terms of  ``\,sound sources categories\,''.
+The term ``\,category\,'' refers to a group of objects (in our case sound sources) considered as equivalent.}
 
@@
 Figure~\ref{datasetexample} shows a subset of the described hierarchical organization.
-The top classes our taxonomy represent concepts such as ``\,urban transport\,'', encompassing classes of sound sources such as ``\,boat\,'' or  ``\,cars\,'', which are themselves split into sub-classes.
+The top classes of our taxonomy represent concepts such as ``\,urban transport\,'', encompassing classes of sound sources such as ``\,boat\,'' or  ``\,cars\,'', which are themselves split into sub-classes.
 At the leaf level, classes are described using sound source names and actions (\textit{car passing, male-voice yelling}).
@@
 The sample selection process is designed to rely only on the listening of sounds themselves, in order not to influence subjects with \textit{a priori} associated semantic values.
-To explore the dataset without any written textual help, we have designed a selection interface where each leaf class is depicted by a point in 2D space, \gl{which was previously tested in an other study} \cite{lafayJAES}.
+To explore the dataset without any written textual help, we have designed a selection interface where each leaf class is depicted by a point in 2D space.
+\gl{The interface performances are extensively tested in the following study} \cite{lafayJAES}.
 Subjects may only interact with the leaf classes of the datasets but the location of the points in space is constrained by the taxonomy, \textit{i.e.} subclasses belonging to a same class are close to one another.
@@
 Considering a soundscape as a sum of isolated samples is in line with the Auditory Scene Analysis (ASA) theory \cite{bregman1994auditory}, which states that auditory stimuli are decomposed into ``\,auditory objects\,'' called ``\,streams\,'', which may be regarded as sequences of auditory events emitted by putative sound sources \cite{ciocca2007auditory,winkler2009modeling}, and also with recent neuroscience studies suggesting that the primary auditory cortex (A1) produces representations of such ``\,auditory objects\,'' that act as bases for high level cognitive processing \cite{nelken2008neurons}.
-Thus, the track can be seen as the generative counterpart of the stream, which is rather a cognitive object.\gl{NM comments}
+Thus, the track can be seen as the generative counterpart of the stream, which is rather a cognitive object.\nm{ce paragraphe est interessant mais est, a mon avis, mal place; il legitime la structure du simulateur, je le mettrais donc plutot dans le §3.1 en le developpant pt-etre un peu plus notamment les aspects neurosciences ...}
 
@@
 
-We apply the proposed protocol to the field of urban soundscapes appraisal, by asking a panel of subjects to simulate two antagonistic urban soundscapes:  one ``\,ideal\,'', or pleasant (i) and the other ``\,non-ideal\,'' or unpleasant (ni).
+We apply the proposed protocol to the field of urban soundscapes appraisal, by asking a panel of subjects to simulate two antagonistic urban soundscapes:  one ``\,ideal\,'', or pleasant (i) and the other ``\,non-ideal\,'' or unpleasant (ni) \nm{voir ma remarque en intro.
+sur la concordance ideal/pleasant ...}.
 The two labels have been chosen to compare the results with those of a previous study addressing ideal urban sound environment representations \cite{guastavino_ideal_2006}.
@@
 \begin{enumerate}
-\item gain finer knowledge about the pleasantness of each simulated scene,
+\item gain finer knowledge about the pleasantness of each simulated scene, \nm{trop vague a mon sens : a developper}
 \item detect possible scene outliers.
-As the distinction i/ni acts as our reference throughout the analysis, one has to ensure that all the scenes are indeed perceived as ideal or non ideal scenes without ambiguity, \textit{i.e.} the highest average perceptual rating of the ni-scenes has to remain inferior to the lowest average rating of the i-scenes.
+As the distinction i/ni \nm{cet encodage n'a  jamais ete defini avant - meme si on devine bien ce qu'il veut dire ....
+;)} \gl{cf.
+plus haut} acts as our reference throughout the analysis, one has to ensure that all the scenes are indeed perceived as ideal or non ideal \nm{il ya confusion entre l'evalution ideal/non ideal et sur l'echelle de pleasantness ....
+est-ce la meme expe.
+?
+a clarifier} scenes without ambiguity, \textit{i.e.} the highest average perceptual rating of the ni-scenes has to remain inferior to the lowest average rating of the i-scenes.
 \end{enumerate}
 
-The simulation experiment is referred as experiment-1 and the rating experiment as experiment-2. The complete experimental plan is described in Figure~\ref{fig:expLan}.
+The simulation experiment is referred as experiment-1 and the rating experiment as experiment-2.
+The complete experimental plan is described in Figure~\ref{fig:expLan}.
+\nm{Du coup, cette info; doit etre mise avant dans le paragraphe histoire d'evacuer toutes les confusions (cf.
+mes remarques plus haut)} \gl{non, je pense qu'à moins de deux paragraphes d'intervales, les lecteurs vont s'y retrouver.}
 
@@
 Supplementary information including the experiment webpage and raw data for each experiment can be found on the companion website \footnote{\url{http://www.irccyn.ec-nantes.fr/~lagrange/demonstrations/simScene.html}}.
-\gl{NM: Pas d'accord commentaires section IV}
 
@@
 Among them, 260 events and 72 textures are recorded by the authors using a \textit{Audio Technica AT8035} shotgun microphone (to isolate sounds from their background) connected to a \textit{ZOOM H4n} recorder.
-The remaining comes from existing sound banks \gl{NM préciser lesquels ?}.
+The remaining comes from existing sound banks \nm{preciser lesquelles ?} \gl{non}.
 All sounds are normalized to the same root mean square (RMS) level.
@@
 
-\subsection{Data analysis}
-
-As experiment-1 generates a large amount of data suggesting numerous avenues of investigation, we here choose to restrict ourselves to the analysis of 1) selected sound classes, 2) sound event density, and 3) scene sound levels, other aspects being intentionally left for future research.
-
-For ease of reading, we introduce the following terms and concepts that will be used throughout the analysis: 
-\begin{description}
-\item[$Diversity$:] number of distinct sound classes used in a scene.
-This number depends on the considered semantic level.
-For example, considering the two subclasses of the semantic level 2 \textit{car-passing} and \textit{car-starting}, which both belong to the same class \textit{car}, we count 2 distinct sound classes for semantic levels 2 and 3, but only 1 sound class for semantic levels 0 and 1.
-\item[$Density$ ($D$):] logarithm of the number of event samples per chunk of 1 second, averaged over the scene duration.
-Density may be computed considering all sound classes, or only a subset of them.
-If nothing is specified, then the density accounts for all classes.
-\gl{Empirically, we observed that the original density distributions were skewed.
-The use of the logarithm compression allowed us to reduce this skewness.}
-\item[$Sound level$ ($L$):]  Decibel of the RMS voltage relative to 1 volt, computed every second and averaged over the scene duration.
-An A-weighting filtering is performed before the RMS computation.
-Computing other statistical descriptors, namely the minimum, maximum or the 10-90th quantiles, shows that they all present a high correlation with the mean ($r\geq0.85$, $p<0.01$), therefore only the latter is considered.
-\end{description}
-
-We measure significance between the i- and ni-scenes using paired Student's t-test.
-All correlations are computed using the Pearson product-moment correlation coefficient.
-All statistical tests are performed at an $\alpha=5\%$ significance level.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Subjects are asked to successively create two urban soundscapes, lasting one minute each.
-The first must be ideal (a soundscape in which the subject would like to live), the second non-ideal (one in which they would try to avoid living).
+The first must be ideal (a soundscape in which the subject would like to live), the second non-ideal (one in which they would try to avoid living) \nm{reprendre la formulation exacte de la consigne} \gl{C'est la formulation, bien que cette dernière était transmise oralement}.
 They are asked to mimic a static listener.
@@
 
-Subjects are asked to evaluate the pleasantness of the 80 simulated scenes using a bipolar scale (also known as Likert scale) of 7 points going from -3 (non ideal, very unpleasant) to +3 (ideal, very pleasant).
+Subjects are asked to evaluate the pleasantness of the 80 simulated scenes using a bipolar scale (also known as Likert scale) of 7 points going from -3 (non ideal, very unpleasant) \nm{a nouveau cet amalgame des 2 termes risque d'etre critiquee et doit etre argumenter et expliciter, je pense.} to +3 (ideal, very pleasant).
 The first ten scenes are used to give the subject an idea of the pleasantness range of the scenes.
@@
 Scenes are presented in a randomized order, but the ten calibration scenes are five i-scenes and five ni-scenes.
-\gl{NM comments: la randomization est déjà précisée.}
+\nm{also randomized in the presentation along subjects ?} \gl{a vérifier}
 
@@
 All the  subjects successfully completed the experiment.
+
+\subsection{Data analysis}
+
+As experiment-1 generates a large amount of data suggesting numerous avenues of investigation, we here choose to restrict ourselves to the analysis of 1) selected sound classes, 2) sound event density, and 3) scene sound levels, other aspects being intentionally left for future research.
+
+For ease of reading, we introduce the following terms and concepts that will be used throughout the analysis: 
+\begin{description}
+\item[$Diversity$:] number of distinct sound classes used in a scene.
+This number depends on the considered semantic level.
+For example, considering the two subclasses of the semantic level 2 \textit{car-passing} and \textit{car-starting}, which both belong to the same class \textit{car}, we count 2 distinct sound classes for semantic levels 2 and 3, but only 1 sound class for semantic levels 0 and 1.
+\item[$Density$ ($D$):] logarithm of the number of event samples per chunk of 1 second, averaged over the scene duration.
+Density may be computed considering all sound classes, or only a subset of them.
+If nothing is specified, then the density accounts for all classes.
+\gl{Empirically, we observed that the original distributions were skewed.
+The use of the logarithm compression allowed us to reduce this skewness.}
+\item[$Sound level$ ($L$):]  Decibel of the RMS voltage relative to 1 volt, computed every second and averaged over the scene duration.
+\nm{est-ce que cette reference correspond a une quelconque calibration du microphone
+de prise de son (p.
+ex., X dB a 1 m.) ?
+Si oui, donner plus de details et expliquer comment tu as fait pour les samples issus de base de donnees - dont on ne connait pas a priori ces elements de calibration .....} \gl{non} An A-weighting filtering is performed before the RMS computation.
+Computing other statistical descriptors, namely the minimum, maximum or the 10-90th quantiles, shows that they all present a high correlation with the mean ($r\geq0.85$, $p<0.01$), therefore only the latter is considered.
+\end{description}
+
+We measure significance between the i- and ni-scenes using paired Student's t-test.
+All correlations are computed using the Pearson product-moment correlation coefficient.
+All statistical tests are performed at an $\alpha=5\%$ significance level.
 
@@
 Among the 9 truly missing ones, we find very specific sounds as ``\,sport car\,'' or ``\,teenager voice\,''.
-We believe that those results show that the database diversity was sufficient for the purpose of the study.
+We believe \nm{n'est-ce pas unpeu faible comme argumentaire ....?
+Pt-etre mettre les resultats en \% pour les rendre plus facilement intrerpretables ?} that those results show that the database diversity was sufficient for the purpose of the study.
 
@@
 \begin{center}
-\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp2_1.eps}
-\caption{\label{xp2_1} Boxplot of the averaged ratings given by the subject considering (a) scenes as observations, (b) scenes as observations without scene outliers and (c) subjects as observations without scene outliers.}
+\includegraphics[width=\columnwidth]{../gfxMatlab/xp2_1.eps}
+\caption{\label{xp2_1} Boxplot of the averaged ratings given by the subject considering (a) scenes as observations, (b) scenes as observations without scene outliers and (c) subjects as observations without scene outliers.
+\nm{les figures sont petites et le chiox des couleurs contribue a leur faible lisibilite} \gl{?}}
 \end{center}
@@
 \centering
-\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_deSoLv_1.eps} 
+\includegraphics[width=\columnwidth]{../gfxMatlab/xp1_deSoLv_1.eps} 
 \caption{\label{fig:xp1_deSoLv_1}  Boxplot of (a) the scene densities ($D$) and (b) the scene sound levels ($L$). Scatter plot of (c) $D$ \emph{vs} ratings and (d) $L$ \emph{vs} rating}
@@
 \includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_div_1.eps} 
-\caption{\label{fig:analyseDensitySnrGlobal3} Diversity across semantic levels for event and texture classes.}
+\caption{\label{fig:analyseDensitySnrGlobal3} Diversity across semantic levels for event and texture classes.
+\nm{il faut mieux differencier les donnees provenant du corpus event et du corpus texture}}
 \end{figure}
@@
 To study the ability of those descriptors to characterize the appraisal, we compute the correlations between the ratings, $D$ and $L$, considering separately the i- and ni-scenes, as wel as all the scenes altogether.
-Results are shown in Table~\ref{tab:numGlobal} and in Figures~\ref{fig:xp1_deSoLv_1}.c and~\ref{fig:xp1_deSoLv_1}.d; \gl{NM comments: je ne pense pas qu'une MANOVA s'utilise comme cela}
+Results are shown in Table~\ref{tab:numGlobal} and in Figures~\ref{fig:xp1_deSoLv_1}.c and~\ref{fig:xp1_deSoLv_1}.d; \nm{pour rendre compte de cet effet, ne peut-on pas egalement calculer effectuer une MANOVA qui renseignera sur l'interaction de 1er ordre entre la variable dependante (ratings) et les variables independantes / facteurs expe.
+(density et Leq) ?} \gl{NM comments: je ne pense pas qu'une MANOVA s'utilise comme cela, comme toute analyse de variance, la variable independante est categorielle.
+De plus dans une MANOVA, le M de multivarié s'applique aux variales dépendantes, la proposition relève de la two way anova (mais le rating étant continu cela ne marche pas).
+Il faudrait faire une regression multiple avec terme d'interaction P=A+B+A*B, et interpréter A*B...}
 
@@
 Diversity remains constant across semantic levels for texture classes of i- and ni-scenes, with no statistical differences.
-The correlations computed between the event diversities and the averaged pleasantness ratings remain weak or moderate ($r<0.45$ \gl{NM comments: non}, $p<0.01$), irrespective of the considered semantic level.
-\gl{NM comments: non} If we look at the two type of scene separately, no correlation is observed.
+\nm{qu'en est-il pour les classes event ?
+Est-ce que les differences observees sur fig.
+7 (a partir de semantic level 1) sont statist.
+siginificatives ?} \gl{cf.
+phrase précédente} The correlations computed between the event diversities and the averaged pleasantness ratings remain weak or moderate ($r<0.45$ \nm{r = 0.45 ?} \gl{NM comments: non, toutes les correlations sont inf à .45}, $p<0.01$), irrespective of the considered semantic level.
+If we look at the two types of scene separately, no correlation is observed.
 \gl{Although the diversity varies between the i- and ni-scenes, this last result shows that it is not a good indicator to precisely characterize the pleasantness.}
@@
 We investigate the sound class selection by counting the number of times a class has been used to simulate the i- or ni-scenes.
-Figures~\ref{fig:xp1_class_1} and \ref{fig:xp1_class_2} show the class distributions for the event and texture classes, respectively.
+Figures~\ref{fig:xp1_class_1}.a and~\ref{fig:xp1_class_1}.b  shows the class distributions for the event and texture classes, respectively.
 For ease of reading, we choose an intermediate semantic level, denoted $0^{+}$, laying between semantic levels 0 and 1 (see Figure~\ref{fig:taxonomie}).
@@
 There is a considerable difference between the sound class selections of the i- and ni-scenes, showing that semantics play an important role in soundscape qualitative evaluation.
-The sound classes shown in Figures~\ref{fig:xp1_class_1} and~\ref{fig:xp1_class_2} are very close to those found by Guastavino \cite{guastavino_ideal_2006}, \textit{i.e.} classes that suggest human presence and nature are the most present in i-scenes while classes referring to mechanical and construction work sounds are used for ni-scenes.
+The sound classes shown in Figures~\ref{fig:xp1_class_1}.a and~\ref{fig:xp1_class_1}.b are very close to those found by Guastavino \cite{guastavino_ideal_2006}, \textit{i.e.} classes that suggest human presence and nature are the most present in i-scenes while classes referring to mechanical and construction work sounds are used for ni-scenes.
 
@@
 Public transport sounds are thus better accepted than those of private vehicles, as they are usually positively connotated.
-Results shown in Figure~\ref{fig:xp1_class_1} tend to qualify \gl{NM comments: non} this claim  by showing that sounds of public transports (\textit{bus} and \textit{train}) were chosen by $28\%$ of the subjects for the i-scenes and by $42\%$ of the subjects for the ni-scenes.
+Results shown in Figure~\ref{fig:xp1_class_1}.c tend to qualify this claim  by showing that sounds of public transports (\textit{bus} and \textit{train}) were chosen by $28\%$ of the subjects for the i-scenes and by $42\%$ of the subjects for the ni-scenes.
 It is important to note that in \cite{guastavino_ideal_2006} subjects are asked to recall an environment in memory, that is, without any sound support.
@@
 
-\begin{figure}[t!]
+\begin{figure*}[t!]
 \begin{center}
-\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_class_1.eps}
-    \caption{\label{fig:xp1_class_1} Percentage of subjects using event sound classes of the semantic level $0^{+}$ to simulate the i-scenes (left-green) and the ni-scenes (right-red).}
+\includegraphics[scale=1]{../gfxMatlab/xp1_class_1.eps}
+    \caption{\label{fig:xp1_class_1} (a) Percentage of subjects using (a) event sound classes of the semantic level $0^{+}$, (b) texture sound classes of the semantic level $0^{+}$ and (c) event subclasses of the $0^{+}$ classes \emph{car} and \emph{public transport}, to simulate the i-scenes (left-green) and the ni-scenes (right-red).}
 \end{center}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_class_2.eps}
-    \caption{\label{fig:xp1_class_2} Percentage of subjects using texture sound classes of the semantic level $0^{+}$ to simulate the i-scenes (left-green) and the ni-scenes (right-red).}
-    \end{center}
-\end{figure}
+%\begin{figure}[t!]
+%\begin{center}
+%\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_class_2.eps}
+%    \caption{\label{fig:xp1_class_2} Percentage of subjects using  to simulate the i-scenes (left-green) and the ni-scenes (right-red).}
+%    \end{center}
+%\end{figure}
 
@@
 
-As semantics is found to play an important role in the qualitative perception of the soundscape, we now investigate the existence of sound markers, \textit{i.e.} sound classes of great importance when considering the pleasantness of a soundscape.
+As semantics is found to play an important role in the qualitative perception of the soundscape, we now investigate the existence of sound markers \nm{est-ce l'accepation de Schafer ?
+d'apres ce qui est ecrit juste apres je ne pense pas; par contre, d'apres ce qui est ecrit un peu plus bas (en citant justement Schafer) ca peut l'etre ....
+Il y a ici une ambiguite sur le terme qu'il serait bon de clarifier.} \gl{Non ce n'est pas  l'interprétation de Schaffer, on expliquait plus en détail avant les termes, mais c'est trop long, et non necessaire pour la compréhension du papier}, \textit{i.e.} sound classes of great importance when considering the pleasantness of a soundscape.
 Here, an event class is considered as a marker if it has been mostly used in one type of soundscape (i or ni).
@@
 \includegraphics[width=.4\paperwidth]{../gfxMatlab/pa5_1.eps}
-\caption{\label{fig:pa5} Precision at rank 5 ($P@5$) computed from the \textit{Hamming} distances between the scenes at different semantic levels.}
+\caption{\label{fig:pa5} Precision at rank 5 ($P@5$) computed from the \textit{Hamming} distances between the scenes at different semantic levels, considering the event and texture classes ($ET$), the event classes ($E$), the texture classes ($T$), the event markers ($E_m$) and event classes without event markers ($E_{w/o m}$) .}
 \end{center}
@@
 
-Results considering both the event and texture classes ($ET$) to build the vectors  $S_i$ are shown on Figure~\ref{fig:pa5}.
+Results considering both the event and texture classes ($ET$) \nm{symbolisme pas clair} \gl{pk ?} to build the vectors  $S_i$ are shown on Figure~\ref{fig:pa5}.
 The $P@5$ is of 81\% for semantic level 0 and remains above 89\% starting at semantic level 1.

### 2016-08-01 part1/lafaySimulatedScenePerception1.tex
end NM comments

@@
 \centering
-\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_div_1.eps} 
+\includegraphics[width=\columnwidth]{../gfxMatlab/xp1_div_1.eps} 
 \caption{\label{fig:analyseDensitySnrGlobal3} Diversity across semantic levels for event and texture classes.
@@
 \begin{center}
-\includegraphics[scale=1]{../gfxMatlab/xp1_class_1.eps}
-    \caption{\label{fig:xp1_class_1} (a) Percentage of subjects using (a) event sound classes of the semantic level $0^{+}$, (b) texture sound classes of the semantic level $0^{+}$ and (c) event subclasses of the $0^{+}$ classes \emph{car} and \emph{public transport}, to simulate the i-scenes (left-green) and the ni-scenes (right-red).}
+\includegraphics[width=2\columnwidth]{../gfxMatlab/xp1_class_1.eps}
+    \caption{\label{fig:xp1_class_1} (a) Percentage of subjects using (a) event sound classes of the semantic level $0^{+}$, (b) texture sound classes of the semantic level $0^{+}$ and (c) event classes of the semantic level 1 belonging to the event classes \emph{traffic} and \emph{public transport} of the semantic level $0^{+}$; ni-scenes (left-red), i-scenes (right-green).}
 \end{center}
@@
 \begin{table}[t]
- \setlength{\tabcolsep}{0.1pt}
+ \setlength{\tabcolsep}{0.2pt}
  \centering
@@
 \hline
-0  &              &  construction work  \\
+0  &                                &  construction work (3.78)  \\
 \hline
-1  & church bell  & horn  \\
-   & bell bike    & siren \\
-   & animal       &       \\
+  & church bell  (4.5)             & horn  (3.9) \\
+1   & bell bike    (4.3)             & siren (3.9)\\
+   & animal       (4.2)             &       \\
    \hline
-2  & birds         & horn  \\
-   & church bell   & siren \\
-   & bell bike     &       \\
+  & birds        (4.8)             & horn  (4.0)\\
+2   & church bell  (4.4)             & siren (4.0)\\
+   & bell bike    (4.2)             &       \\
    \hline
-3  & birds singing & horn  \\
-   & church bell   & siren \\
-   & bell bike     &       \\
-   & male footsteps concrete &  \\
+  & birds singing (4.8)            & horn  (4.1)\\
+   & church bell   (4.3)            & siren (4.0)\\
+3   & bell bike     (4.2)            &       \\
+   & male footsteps                 &  \\
+   &   concrete (3.6)               &  \\
   \hline
@@
 \begin{center}
-\includegraphics[width=.4\paperwidth]{../gfxMatlab/pa5_1.eps}
+\includegraphics[width=\columnwidth]{../gfxMatlab/pa5_1.eps}
 \caption{\label{fig:pa5} Precision at rank 5 ($P@5$) computed from the \textit{Hamming} distances between the scenes at different semantic levels, considering the event and texture classes ($ET$), the event classes ($E$), the texture classes ($T$), the event markers ($E_m$) and event classes without event markers ($E_{w/o m}$) .}
@@
 \includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_deSoLvMar_1.eps}
-  \caption{\label{fig:numMarkers} Averaged pleasantness vs. (a)  $LA_{(eq,V)\_m-b}$, (b) $LA_{(eq,V)\_m}$, (c) $LA_{(eq,V)\_b}$ and (d) $Density_m$.}
+  \caption{\label{fig:numMarkers} Averaged pleasantness vs. (a)  $L_{m-b}$, (b) $L_{m}$, (c) $L_{b}$ and (d) $D_m$.}
   \end{center}
@@
 
-In order to check if the physical characteristics of the markers alone have an influence on the pleasantness, we now consider the densities and the $LA_{(eq,V)}$ with on one hand the tracks of the markers ($LA_{(eq,V)\_m}$, $Density_{m}$) and on the other hand the remaining sounds ($LA_{(eq,V)\_b}$, $Density_{b}$).
-The marker durations is rather short, and the number of markers in a scene is rather low, a window length of 125 ms is used to compute $LA_{(eq,V)\_m}$ and $LA_{(eq,V)\_b}$.
+In order to check if the physical characteristics of the markers alone have an influence on the pleasantness, we now consider the $D$ and the $L$ with on one hand the tracks of the markers ($L_m$, $D_{m}$) and on the other hand the remaining sounds ($L_b$, $D_b$).
+As the marker durations are rather short, and the number of markers in a scene are rather low, a window length of 125 ms is used to compute $L_m$ and $L_b$ \nm{pourquoi cette precision methodologique n'a t'elle pas ete donnee aussi au moment de definir ce parametre ?} \gl{plus simple ici}.
 For the i-scenes we consider only the markers of the i-scenes (i-markers), and for the ni-scenes only the markers of the ni-scenes (ni-markers).
+\nm{cela parait evident, pourquoi y'a-t'il besoin de le preciser ?} \gl{fixe bien les idées}
 
-The difference $LA_{(eq,V)\_m-b}=LA_{(eq,V)\_m}-LA_{(eq,V)\_b}$ is also considered as it illustrates how the markers emerge from the other sounds.
+The difference $L_{m-b}=L_m-L_b$ is also considered as it illustrates how the markers emerge from the other sounds.
 Correlations between those descriptors and the average perceptual ratings are depicted in Table~\ref{tab:numMarkers} and Figure~\ref{fig:numMarkers}.
 
-Considering $Density_{m}$ (Figure~\ref{fig:numMarkers}.d), results are almost the same than to those found in section~\ref{sec:levelDensityDiversity}.
+Considering $D_{m}$ (Figure~\ref{fig:numMarkers}.d), results are almost the same than to those found in section~\ref{sec:levelDensityDiversity}.
 There is no correlation for the i-scenes, but the correlation increases for the ni-scenes.
-No correlation is found for $Density_{b}$.
+No correlation is found for $D_{b}$.
 Sound levels descriptors are much more informative.
-Little to no information is lost by considering only the markers to compute $LA_{(eq,V)\_m}$, \emph{i.e.} there is still a good correlation for the ni-scenes (although inferior to the one observed in Section~\ref{sec:levelDensityDiversity}), and no correlation for the i-scenes.
+Little to no information is lost by considering only the markers to compute $L_m$, \emph{i.e.} there is still a good correlation for the ni-scenes (although inferior to the one observed in Section~\ref{sec:levelDensityDiversity}), and no correlation for the i-scenes.
 
-Considering the ni-scenes, the good correlation for $LA_{(eq,V)\_m}$  (Figure~\ref{fig:numMarkers}.b) and the absence of correlation for $LA_{(eq,V)\_b}$  (Figure~\ref{fig:numMarkers}.c) suggests that the sound levels of the ni-markers convey most of the annoyance.
-Also, no correlation is observed for $LA_{(eq,V)\_m-b}$ (Figure~\ref{fig:numMarkers}.a).
+Considering the ni-scenes, the good correlation for $L_m$  (Figure~\ref{fig:numMarkers}.b) and the absence of correlation for $L_{b}$  (Figure~\ref{fig:numMarkers}.c) suggests that the sound levels of the ni-markers convey most of the unpleasantness.
+Also, no correlation is observed for $L_{m-b}$ (Figure~\ref{fig:numMarkers}.a).
 This indicates that the predominant influence of the marker sound levels remains true irrespective of the sound level of the other sounds.
@@
 That said, this last statement is not valid for the i-scenes.
-The strong uphill correlation found between pleasantness and  $LA_{(eq,V)\_m-b}$ (Figure~\ref{fig:numMarkers}.a) shows that the higher the markers sound levels relative to the other sounds, the higher the pleasantness.
-But the moderate downhill correlation observed for $LA_{(eq,V)\_b}$ (Figure~\ref{fig:numMarkers}.c) shows that the levels of those remaining sounds may be a cause of annoyance if perceived as too loud.
+The strong uphill correlation found between pleasantness and  $L_{m-b}$ (Figure~\ref{fig:numMarkers}.a) shows that the higher the markers sound levels relative to the other sounds, the higher the pleasantness.
+But the moderate downhill correlation observed for $L_{b}$ (Figure~\ref{fig:numMarkers}.c) shows that the levels of those remaining sounds may be a cause of annoyance if perceived as too loud.
 It is thus the emergence of the i-markers from the others sounds that drive the appraisal.
@@
 
+\nm{n'est ce pas au bout du compte, 2 propositions contraposees qui reviennent a dire qu'il faut au 1er ordre s'interesser a certaines sources particulieres (et a leur niveau) plutot que considerer un percept global (notamment en terme de niveau)} \gl{rien compris}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 This protocol allows us to generate soundscapes whose composition, in terms of nature, timing and intensity of sound sources, is known with perfect accuracy.
-Using the data thus produced in a series of analyses on the issue of ideal \emph{vs} non-ideal urban soundscape classification and characterization, we show that:
+Using the data thus produced in a series of analyses on the issue of ideal \emph{vs} non-ideal urban soundscape classification and characterization, we showed that:
 
@@
 Compared to more traditional survey based methods, the proposed method provides an increased precision and reliability of soundscape description useful for the analysis of subjects' judgments.
-This article was intentionally focused on the analysis of the scene composition; further research shall be dedicated to 1) a thorough study of the influence of the physical characteristics of the sound markers on pleasantness evaluation, as well as categorisation process, and 2) the elaboration of a predictive method based on that analysis, allowing to anticipate the agreement of an acoustic environment from its composition.
+This article was intentionally focused on the analysis of the scene composition; further research shall be dedicated to 1) a thorough study of the influence of the physical characteristics of the sound markers on pleasantness evaluation, as well as categorisation process, and 2) the elaboration of a predictive method based on that analysis, allowing to anticipate the agreement \nm{encore un autre terme : synonyme de pleasantness ?} of an acoustic environment from its composition.
 
@@
 Is loudness but a consequence of scene composition ?
+\nm{cette derniere phrase est incomprehensible ....
+bien qu'elle semble importante car elle introduit la notion de loudness (perceived intensity) en oppposition a la notion de sound level (measured intensity) utilisee jusqu'a present : a eclaircir et approfondir davantage, je pense .....}
+
+\nm{je ne suis pas trop du genre a raisonner comme cela mais je me dis que : vu le type de revue (AAA) et la thematique de l'article , il est possible que nous ayons JD.
+Polack comme potentiel reviewer, or nous ne citons aucun de ses travaux en la matiere ......}
 

### 2017-03-03 lafaySimulatedScenePerception.tex
archive old version

--- 2016-08-01 part1/lafaySimulatedScenePerception1.tex
@@
 Nine markers are found across all semantic levels.
-As suggested by Figures~\ref{fig:xp1_class_1} and~\ref{fig:xp1_class_2}, sounds evoking human activity (\textit{male footsteps concrete}) and nature (\textit{animal}, \textit{birds} and \textit{birds singing}) are markers of the i-scene.
+As suggested by Figures~\ref{fig:xp1_class_1}.a, sounds evoking human activity (\textit{male footsteps concrete}) and nature (\textit{animal}, \textit{birds} and \textit{birds singing}) are markers of the i-scene.
 The presence of \textit{church bell} and \textit{bell bike} as strong markers of the i-scenes can be due to the socio-cultural background of the subjects, in great majority European citizens.
@@
 \includegraphics[width=\columnwidth]{../gfxMatlab/pa5_1.eps}
-\caption{\label{fig:pa5} Precision at rank 5 ($P@5$) computed from the \textit{Hamming} distances between the scenes at different semantic levels, considering the event and texture classes ($ET$), the event classes ($E$), the texture classes ($T$), the event markers ($E_m$) and event classes without event markers ($E_{w/o m}$) .}
+\caption{\label{fig:pa5} Precision at rank 5 ($P@5$) computed from the \textit{Hamming} distances between the scenes at different semantic levels, considering the event and texture classes ($ET$), the event classes ($E$), the texture classes ($T$), the event markers ($E_m$) and event classes without event markers ($E_{w/o,m}$) .}
 \end{center}
@@
 
-Each simulated scene $s_i$ is represented by a boolean vector of dimension $n$: $S_i = (x_1, x_2, \ldots{}, x_{n})$, $i\in[1,72]$.
+Each simulated scene $s_i$ is represented by a boolean vector of dimension $n$: $S_i = (x_1, x_2, \ldots{}, x_{n})$, $i=\lbrace 1, \ldots, 72 \rbrace$.
 Each dimension corresponds to a sound class (event and texture) of a particular semantic level (for example: $n=44$ classes for semantic level 1).
@@
 A \textit{Hamming} distance is computed between all pairs of $S_i$.
-Then, for each vector $S_i$ the number of scenes sharing the same label (i or ni) as $S_i$  among the five scenes $s_j$ ($j=1,...,5$) that have the smallest distance to it is computed.
-The $P@5$ precision is then obtained by averaging the results over all the scenes $s_i$ ($i=1, \ldots{}, 72$).
+Then, for each scene $s_i$, the number of scenes sharing the same label (i or ni) among the five scenes $s_j$ that have the smallest distance to $s_i$ is computed.
+The $P@5$ precision is then obtained by averaging the results over the 72 scenes $s_i$.
 
 Results considering both the event and texture classes ($ET$) \nm{symbolisme pas clair} \gl{pk ?} to build the vectors  $S_i$ are shown on Figure~\ref{fig:pa5}.
-The $P@5$ is of 81\% for semantic level 0 and remains above 89\% starting at semantic level 1.
+The $P@5$ is of 76\% for semantic level 0 and remains above 86\% starting at semantic level 1.
 These result strongly indicates that it is possible to make a clear distinction between the i- and ni-scenes based solely on the class presence/absence.
@@
 
-To refine the analysis, only event classes ($E$), texture classes ($T$), event markers ($E_m$) and event classes without event markers ($E_{w/o m}$) are also considered on Figure~\ref{fig:pa5}.
-The constant deviation of circa $10\%$ from the semantic level 1 between $E$ and $T$, and the fact that $P@5$ are similar for $ET$ and $E$ indicate that soundscape quality is mostly conveyed by sound events.
+To refine the analysis, only event classes ($E$), texture classes ($T$), event markers ($E_m$) and event classes without event markers ($E_{w/o,m}$) are also considered on Figure~\ref{fig:pa5}.
+The constant deviation of circa $15\%$ from the semantic level 1 between $E$ and $T$, and the fact that $P@5$ are similar for $ET$ and $E$ indicate that soundscape quality is mostly conveyed by sound events.
 
@@
 Indeed, $E_m$ uses very partial information but still reaches quasi-equivalent results.
-Moreover, the drop in $P@5$ considering $E_{w/o m}$ from the semantic level 1 confirms the fact that soundscape quality mostly relies on the presence of specific sound markers.
+Moreover, the drop in $P@5$ considering $E_{w/o,m}$ from the semantic level 1 confirms the fact that soundscape quality mostly relies on the presence of specific sound markers.
 
@@
 
-In order to check if the physical characteristics of the markers alone have an influence on the pleasantness, we now consider the $D$ and the $L$ with on one hand the tracks of the markers ($L_m$, $D_{m}$) and on the other hand the remaining sounds ($L_b$, $D_b$).
-As the marker durations are rather short, and the number of markers in a scene are rather low, a window length of 125 ms is used to compute $L_m$ and $L_b$ \nm{pourquoi cette precision methodologique n'a t'elle pas ete donnee aussi au moment de definir ce parametre ?} \gl{plus simple ici}.
-For the i-scenes we consider only the markers of the i-scenes (i-markers), and for the ni-scenes only the markers of the ni-scenes (ni-markers).
-\nm{cela parait evident, pourquoi y'a-t'il besoin de le preciser ?} \gl{fixe bien les idées}
-
+In order to check if the physical characteristics of the markers alone have an influence on the pleasantness, we now compute $D$ and $L$ considering on one hand the tracks of the markers ($L_m$, $D_{m}$) only, and on the other hand the remaining of the sounds ($L_b$, $D_b$).
+As the marker durations are rather short, and the number of markers in a scene rather low, a window length of 125 ms is used to compute $L_m$ and $L_b$ \nm{pourquoi cette precision methodologique n'a t'elle pas ete donnee aussi au moment de definir ce parametre ?} \gl{non, elle change ici}.
 The difference $L_{m-b}=L_m-L_b$ is also considered as it illustrates how the markers emerge from the other sounds.
-Correlations between those descriptors and the average perceptual ratings are depicted in Table~\ref{tab:numMarkers} and Figure~\ref{fig:numMarkers}.
+Correlations between those descriptors and the average ratings are depicted in Table~\ref{tab:numMarkers} and Figure~\ref{fig:numMarkers}.
 
@@
 
-Considering the ni-scenes, the good correlation for $L_m$  (Figure~\ref{fig:numMarkers}.b) and the absence of correlation for $L_{b}$  (Figure~\ref{fig:numMarkers}.c) suggests that the sound levels of the ni-markers convey most of the unpleasantness.
-Also, no correlation is observed for $L_{m-b}$ (Figure~\ref{fig:numMarkers}.a).
-This indicates that the predominant influence of the marker sound levels remains true irrespective of the sound level of the other sounds.
+Considering the ni-scenes, the good correlation for $L_m$  (Figure~\ref{fig:numMarkers}.b) and the absence of correlation for $L_{b}$  (Figure~\ref{fig:numMarkers}.c) suggest that sound levels of the ni-markers convey most of the unpleasantness.
+The fact that no correlation is observed for $L_{m-b}$ (Figure~\ref{fig:numMarkers}.a) indicates that the predominant influence of the marker sound levels remains true irrespective of the sound level of the other sounds.
 A loud horn sound is expected to increase the annoyance level of a scene (more so than a soft klaxon sound), whether or not the remaining of the sonic environment is ``\,noisy\,''.
@@
 
-These results are strong indicators that global sound level is not a good indicator to characterize the appraisal of \emph{a priori} positive sound environments such as a park or a quiet street.
+These results are strong indicators that global sound level alone is not sufficient to properly characterize the appraisal of \emph{a priori} positive sound environments such as a park or a quiet street.
 On the contrary, the contribution of each sound source needs to be considered separately, as for certain type of sounds, the pleasantness does increase with their sound levels.

### 2017-03-03 material/LaTeX_source_files/IFA.TEX
archive old version

% Instructions for Authors fuer Acta Acustica

\documentclass[twoside,twocolumn]{article}
\usepackage{acta}

\begin{document}
\Eingang{27}{10}{2012} \Annahme{6}{12}{2012} \Sachgebiet{1}
\PACS{43 \dots}
\Band{98} \Jahr{2012} \Heft{} \Ersteseite{1} \Letzteseite{}

\AuthorsI{D. Botteldooren$^{1)}$, J. Wempen$^{1)}$}
\AddressI{$^{1)}$ Department of Information Technology, Ghent
University, Sint Pietersnieuwstraat 41, 9000 Gent, Belgium.\\
\hspace*{8pt}aa.a@intec.ugent.be\\
$^{2)}$ Daucherstr. 98, 85053 Ingolstadt, Germany. wempen@t-online.de}

% \footnotemark[1] \Authornotes{\footnotetext[1]{} }

\Englishtitle{Instructions for Authors} \Germanorfrenchtitle{}

\Kolumnentitel{Botteldooren, Wempen: Instructions for Authors}

\Englishabstract{Instead of providing a classic template to be filled in
by the authors, we decided that it is rather sensful to present an
example on how to build \LaTeX{} source files to be published in Acta
Acustica united with Acustica. Examples of a .tex and a .bib file,
together with corresponding style files, can be found in a .zip file
downloadable from the Acta Acustica united with Acustica webpage.}
\Germanorfrenchabstract{}

\ScientificPaper

\section{Function and scope}

Acta Acustica united with Acustica, published together with the European
Acoustics Association (EAA), is an international, peer-reviewed journal
on acoustics. It publishes original articles on all subjects in the
field of acoustics, such as

\begin{quote}\em
General Linear Acoustics, Nonlinear Acoustics, Macrosonics,
Aeroacoustics, Atmospheric Sound, Underwater Sound, Ultrasonics,
Physical Acoustics, Structural Acoustics, Noise Control, Active Control,
Environmental Noise, Building Acoustics, Room Acoustics, Acoustic
Materials, Electroacoustics and Signal Processing, Computational and
Numerical Acoustics, Hearing, Audiology and Psychoacoustics, Speech,
Musical Acoustics, Virtual Acoustics, Auditory Quality of Systems,
History of Acoustics.
\end{quote}

The journal reports on original scientific research in acoustics and on
engineering applications. The journal considers scientific papers,
technical and applied papers, book reviews, short communications,
doctoral thesis abstracts, etc. In irregular intervals also special
issues and review articles are published.

\section{Languages}

Accepted languages: English, French and German. Authors seeking to reach
a wide readership may wish to consider publishing in the English
language. Authors of contributions in English, who want linguistic
support, may indicate this upon submission of their manuscripts. If the
contribution is not in English, a title and a summary in English must be
submitted.

\section{Manuscripts}

Manuscript submission guidelines have changed on Oct. 1st 2012. The
review process of all manuscripts submitted prior to that date will be
completed following the instructions for authors valid at the time of
submission.

\subsection{Submission}

New and revised manuscripts shall be submitted online at
http://www.editorialmanager.com/aaa/

The following information will need to be provided during submission:

\subsubsection{Article type}

Acta Acustica united with Acustica publishes review papers, scientific
papers, short communications, technical and applied papers, and letters
to the editor.

\begin{itemize}
\item Scientific Papers: contain original material (ideas, models,
   experiments) not published elsewhere that contributes substantially
   to the advance of science in the field of acoustics; they should
   clearly establish the relation between the work reported on, and the
   state-of-the-art; scientific papers are typically not longer than 12
   pages;
\item Review Papers: discuss the state-of-the-art in an area of
   acoustics that seems of particular interest to the community at this
   point in time; review papers should be particularly complete in
   covering literature and should mention all ideas, models,
   experiments, even if they do not correspond to the authors own
   opinion;
\item Technical and Applied Papers: report on original applications of
   an existing technique, concept, or measurement method to a new area;
   it is essential that a technical and applied paper is of sufficient
   interest to a group of researchers and engineers beyond the specific
   application or geographic area;
\item Short Communications: are short scientific papers typically not
   longer than 4 pages; one can decide for a short communication to
   report on original, new ideas, models, or experimental results, that
   do not require a substantial discussion; they are NOT scientific
   papers of poor quality;
\item Letters to the Editor: are promoted as a forum to simulate
   scientific discussion; typically they are critical comments on or
   additions to published work by the author or his peers; they may also
   introduce new ideas for future research without the proof required
   for a scientific paper; the scientific level needs nevertheless be
   guaranteed.
\end{itemize}

\subsubsection{Title}

The manuscript title should be chosen carefully to reflect the content
of the manuscript.

\subsubsection{Author(s) and their affiliation}

The corresponding author should be fully registered in the online
manuscript handling system. For other authors that are not registered,
entering name and affiliation is sufficient. The submitting author can
select other authors to become the corresponding author.

\subsubsection{Field (Section/Category)}

Your manuscript will be published and indexed under this Field in Acta
Acustica united with Acustica.

\subsubsection{Abstract}

The abstract should be a brief and precise representation of the content
of your manuscript. The choice of wording is of utmost importance since
searching in different indexes is mainly based on the abstract text.

\subsubsection{Classification (PACS number)}

For additional classification and for selecting reviewers, the PACS
(Physics and Astronomy Classification Scheme) is used. Please select one
to five keywords accurately describing the area of your manuscript from
the list.

\subsubsection{Statement concerning the originality of the work}

A manuscript can only be considered for publication if it is not under
review or in print with another journal. Previous submission to another
journal is no objection but the reason for submitting to Acta Acustica
united with Acustica should be clarified.

\subsubsection{Comments to the Editorial Office (optional)}

Optionally, additional comments to the editorial board can be added.
These comments will not be forwarded to the reviewers.

\subsubsection{Suggested reviewers}

The author is requested to suggest reviewers. The editors are free to
follow these suggestions or not. Optionally, the author can oppose
possible reviewers. Please give a good reason to oppose a reviewer. The
editors preserve the right to ignore these objections.

\subsubsection{Manuscript files and figures}

Files containing the manuscript and figures should be uploaded online.
At least one item ``Manuscript'' should be attached to the submission. If
multiple documents are added, they will be included in the automatically
generated PDF file used for review in the order given in the submission.
Manuscripts should be submitted preferably in \LaTeX{} \cite{1,2}.
Please submit only one .tex file. Additional files such as bibliographic
data have to be added as separate documents if not included in the main
file(s). Authors not familiar with \LaTeX{} may also submit their
manuscript as Word, WordPerfect, or RTF.

Figures should be submitted as separate files in a vector graphic
format: Encapsulated postscript, .eps; Windows specific formats like
.wmf or .emf, Portable document format .pdf or any other true vector
format. Bitmap formats like .bmp, .tif, .jpg etc. are only allowed for
photographs and similar graphic material.

In case of any problems while uploading a manuscript, please contact the
Editor-in-Chief's office at the address given below. Once the above
information is entered into the system, a manuscript PDF file will be
generated. The submitting author should approve this automatically build
file. Your manuscript will then be assigned a manuscript number and the
Editor-in-Chief will forward it to the most appropriate Associate
Editor. The Associate Editor will send the manuscript to referees,
supervise the reviewing process and will inform you of his/her decision.
Final disposition is up to the Editor-in-Chief. You can follow the
review process at any time by logging in at
http://www.editorialmanager.com/aaa/

\subsection{Style}

Any submitted manuscript should be in a form that allows the referees
efficient study; it should be easily readable and offer enough space to
add comments onto it. The Editors of Acta Acustica united with Acustica
recommend that manuscripts should be written in \LaTeX. This assures that
the manuscript is easily readable and greatly reduces costs. As a
benefit to the author, the paper can be published considerably faster.
Any \LaTeX-style may be used. If an author is not familiar with \LaTeX, no
special formatting is required. The manuscript should be typewritten
wide-spaced, pages should be consecutively numbered and the name of the
first author should be repeated in the upper right-hand corner of the
pages. The sheets of the manuscript should not be bound or stapled.
Formulae should be clearly written using standard symbols which are
explained at its first appearance. Nomenclatures or list of symbols will
be dropped.

\subsubsection{Figures}

In order to assure a readable appearance in the printed version, the
figures must be carefully designed. In particular, the thickness of
lines and the height of text must be chosen such that the lines are
still clearly visible and the text legible after the figures are reduced
to column size. Care should also be taken to ensure line types (dashed,
dotted, different thickness) and line markers are such that lines can
still be differentiated after the figures are reduced to their final
size, i.e. scaled to about 75\,mm width. In rare cases, double-column
figures are allowed.

Please do not use larger lettering than necessary, avoid bold fonts and
keep the letter size constant either in all parts of each figure or
throughout the entire set of figures. If the figures are on separate
pages, the number of each figure and the name of the first author must
be marked.

Original figures should be printed or drawn in black and white.
Unnecessary details should be avoided as well as extra frames, headings,
grids and extensive use of hatched areas. If an author wishes to include
figures containing greyscales, it is obligatory to deliver data files of
these pictures. Where coloured figures are needed, the online version of
the journal will show the coloured figures, while the figures will
normally be reproduced only in black and white in the print version. The
publication of coloured figures in the print version requires a special
fee to be paid to the publisher. If the figures are produced by using a
computer, the authors are asked to provide (in the finally accepted
version) separate data files of the figures. Any vector format is
preferred, standard encapsulated postscript is recommended. The data
files should be consecutively numbered according to their appearance in
the text.

\subsubsection{Bibliography}

Articles submitted to Acta Acustica united with Acustica should cite all
relevant work, part of which should have appeared in Acoustics journals
such as Acta Acustica united with Acustica. This criterion will be
applied to determine whether the article is within scope of the journal.

The number citations should not exceed 80 except for review papers.
References are cited by number in rectangular brackets, like [13]. The
reference list is placed at the end of the manuscript. The order of
references is according to their appearance in the article. All author
names should be included in the reference (cf. References below). An
example biblio.bib file for use with the BiBTeX program is provided
together with the example \TeX-source file ifa.tex of this document in
a .zip file downloadable from the Acta Acustica united with Acustica
webpage.

\subsubsection{Proofs}

Proofs will be sent to the person indicated as the corresponding author
at submission of the manuscript on the first page of the manuscript. The
corrected proofs should be returned within 8 days; corrections should be
carefully done and clearly indicated. No new material can be inserted in
the text at the time of proof reading.

\section{Copyright, self-archiving policy and offprints}

\subsection{Copyright assignment}

The submitting/corresponding author will need to fill in and sign the
Copyright Transfer Statement for the journal and return it to the
publisher. Please see this form on the journal's website for information
on what rights you retain as an author and on the publisher's author
self-archiving policy.

\subsection{Offprints}

No honorarium is paid. The first author of an article will receive 25
offprints free-of-charge. Further offprints (in multiples of 25) may be
ordered at extra costs. A price list is available from the publisher.

\section{Addresses}

Correspondence concerning manuscript status should be directed to the
Editor-in-Chief:

\begin{quote}
Prof. Dr. Dick Botteldooren\\
Acoustics Group\\
Department of Information Technology\\
Ghent University\\
Sint Pietersnieuwstraat 41\\
B-9000 Gent, Belgium\\
Phone +32 9 2649968\\
Fax +32 9 2649960\\
E-mail: aa.a@intec.ugent.be
\end{quote}

Inquiries concerning manuscript style, file formats, layout of tables
and figures and of the final paper may be sent to the Production
Manager:

\begin{quote}
Dr. Johann Wempen\\
Daucherstr. 98\\
D-85053 Ingolstadt\\
Phone +49 841 97 5852\\
E-mail: wempen@t-online.de
\end{quote}

\noindent
Publisher:

\begin{quote}
S. Hirzel Verlag\\
Birkenwaldstr. 44\\
70191 Stuttgart\\
PO Box 10 10 61\\
70009 Stuttgart\\
Germany\\
Phone +49 (0)711 25 82-0\\
Fax +49 (0)711 25 82-290\\
E-mail: ActaAcustica@Hirzel.de
\end{quote}

\subsection*{Acknowledgement}

If appropriate.

\References{}{Biblio}{actalit}

\end{document}

### 2017-03-03 material/jasaMentalRepresentations.tex
archive old version

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[12pt, titlepage, reqno]{article} % Document font size and equations flushed left
 \usepackage[super,sort&compress]{natbib}
 \usepackage{amssymb, latexsym}
 \usepackage{amsmath} 
 \usepackage{amsthm}
 \usepackage{bm}
 \usepackage[mathscr]{eucal}
 \usepackage{enumerate}
\usepackage{url}
\usepackage{graphicx}% http://ctan.org/pkg/graphicx
\usepackage{color}
\renewcommand{\thefootnote}{\alph{footnote})}    
 \renewcommand{\baselinestretch}{2}
 \renewcommand{\section}[1]{\medskip \addtocounter{section}{1}\raggedright 
     \textbf{\Roman{section}. \ #1}\medskip \setcounter{subsection}{0}
    \setlength{\parindent}{5ex}
 }
 \renewcommand{\subsection}[1]{\medskip \addtocounter{subsection}{1}\raggedright
    \textbf{\Alph{subsection}. \ #1} \medskip \setcounter{subsubsection}{0}\setlength{\parindent}{5ex}
}
 \renewcommand{\bibsection}{\medskip \raggedright  \textbf{REFERENCES}
    \medskip}
 \renewcommand{\bibnumfmt}[1]{\textbf{{#1}.}\ \ }
 \pagestyle{myheadings}
 \markboth{\hfill Lafay et al., JASA, p.\ }{\hfill Lafay et al., JASA, p.\ }
 \setlength{\parindent}{5ex}

 \setlength{\paperwidth}{8.5in}
 \setlength{\textwidth}{6.5in}
 \setlength{\oddsidemargin}{0in}
 \setlength{\evensidemargin}{0in}
\usepackage{setspace}
\usepackage[format=hang,font={scriptsize,onehalfspacing},labelfont=bf]{caption}

\begin{document}
\begin{titlepage}

 \begin{center}

 \textbf{Approaching mental representations of urban soundscape using auditory scenes simulation}\\

 \vspace{10ex}

\textbf{Gr\'egoire Lafay\footnote{e-mail: gregoire.lafay@irccyn.ec-nantes.fr}, Mathieu Lagrange and Jean-Fran\c cois Petiot}\\

Institut de Recherche en Communications et  Cybernétique de Nantes (IRCCYN) \\

Ecole centrale de Nantes, France \\

\textbf{Mathias Rossignol and Nicolas Misdariis }\\

Institut de Recherche et Coordination Acoustique/Musique (IRCAM) \\

Paris, France \\

\textbf{pacs: 43.50.Lj, 43.50.Rq, 43.50.Qp, 43.66.Lj, 43.66.Ba}
 \end{center}

 \end{titlepage}

%\renewcommand{\baselinestretch}{1}

%----------------------------------------------------------------------------------------
%	ARTICLE INFORMATION
%----------------------------------------------------------------------------------------

%

%\title{Approaching mental representations of urban soundscape using auditory scenes simulation}
%\author{Grégoire Lafay}
%\affiliation{Institut de Recherche en Communications et  Cybernétique de Nantes (IRCCYN), Ecole centrale de Nantes, France}
%\author{Mathias Rossignol}
%\affiliation{Institut de Recherche et Coordination Acoustique/Musique (IRCAM), France}
%\author{Nicolas Misdariis}
%\affiliation{Institut de Recherche et Coordination Acoustique/Musique (IRCAM), France}
%\author{Mathieu Lagrange}
%\affiliation{Institut de Recherche en Communications et  Cybernétique de Nantes (IRCCYN), Ecole centrale de Nantes, France}
%\author{Jean François Petiot}
%\affiliation{Institut de Recherche en Communications et  Cybernétique de Nantes (IRCCYN), Ecole centrale de Nantes, France}

\begin{abstract}
This paper introduces a subject-centered experimental protocol to study mental representations of urban soundscapes through a simulation process. Subjects are asked to recreate a full sound environment by means of a structured sound data set and a  software dedicated to sound manipulation. This paradigm is used to characterize urban sound environment representations, by analyzing the sound classes that were used to simulate the auditory scenes. Results show that a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban sound environments.
\end{abstract}

%\maketitle % Print the title and abstract box

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%   INTRODUCTION     %%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

 \addtocounter{page}{2}

 \section{Introduction}

 \setlength{\parindent}{5ex}

The notion of soundscape has been introduced by Schafer \cite{schafer_new_1969, schafer1977tuning} in the 1970s as the auditory equivalent to landscape. Following this paradigm, a sonic environment is described by focusing on the listener's evaluation, rather than only taking into account the acoustic parameters of the sound. Focusing on subjective criteria, Schafer proposed to decompose the soundscape between three main components called \textit{keynote sounds}, \textit{signals} and \textit{soundmarks}. The notion of soundmark refers to sounds with subjective qualities that bring them into focus for a certain category of people, whereas keynote sounds and signals may roughly be considered as respectively background and foreground sounds. Schafer also claims that the improvement of the sonic environment requires a ``\,positive\,'' approach to the soundscape, involving the identification and reinforcement of pleasant, important or meaningful sounds in the environment. The soundscape thus appears as a powerful tool to develop perceptively motivated acoustical indicators \cite{schulte-fortkamp_soundscape:_2007}.

The main goals of studies addressing soundscape perception may be summarized as follow:

\begin{enumerate}
\item \textit{Describing the soundscape}: How does the human brain identify different types of soundscapes? What are the elements composing each type of soundscape? How do they differ from one soundscape type to another? 
\item \textit{Evaluating the soundscape}: How do those elements influence the qualitative evaluations of a soundscape, such as ``\,noise annoyance\,'' or ``\,pleasantness\,'', to name but a few?
\end{enumerate}

The second question finds its origin in the growing need to improve the quality of artificial sonic environments, such as the urban soundscape. It is only during the 1980s that policy-makers started taking into account the link between \textit{noise} and pollution, considering noise as a significant degradation of the quality of life. To fight this pollution, the first approach consisted in identifying unwanted sounds, and lowering their intensities. Thus, in the wake of this realization, several regulations have emerged, often limited to enforcing sound level thresholds. However, several studies showed that urban ``\,noise\,'' perception and evaluation is a complex phenomenon that cannot be described solely with the help of objective acoustical measures such as A-weighted levels and $L_{Aeq}$ \cite{schulte-fortkamp_soundscape_2006, schulte2013soundscape,dubois2006cognitive,dubois2000categories,raimbault_urban_2005,guastavino_etude_2003}. More precisely, Yang and Kang \cite{Yang2005211} as well as Kang and Zhang \cite{kang_semantic_2010} found significant differences between objective acoustic comfort measurements and subjective evaluations of sonic environments, thus confirming that ``\,noise\,''  is a cognitive object which depends on a listener's appreciation and the context in which the noise is heard. Many urban ``\,noises\,'' such as that of a \textit{siren}, can annoy as well as warn of a danger. Many town districts are appreciated because of their lively and animated atmosphere which often results in higher sound levels.

To summarize, we believe that even if acoustical regulations  are to some extent effective, there is a need to describe urban soundscapes not only through objective acoustical measures. Qualitative attributes also need to be considered \cite{torija2013application, cain_development_2013} in order to understand better the relationship between humans and their acoustic environment\cite{schafer1977tuning}.

\section{Background}

The strength of the soundscape approach is, to some extent, also its weakness, as it implies to  fully appreciate the interaction of many factors. Perception of soundscape is thus an interdisciplinary field of investigation, each discipline coming with its well identified experimental protocol. In each one, integration of the results can be difficult \cite{davies_perception_2013, schulte-fortkamp_soundscape:_2007}. Recently, ambitious projects have been undertaken as the European Cooperation in Science and Technology Action TD0804-Soundscape of European Cities and Landscapes to standardize soundscape assessment and indicators.

We believe that studies addressing soundscape perception may be roughly divided into two approaches according to their methodologies \cite{davies_perception_2013}.

The first approach is the dimensional approach, which tends to derive relevant emotional dimensions using semantic differential analysis  \cite{cain_development_2013, jeon2013soundwalk, kang_semantic_2010}, and to link those dimensions with acoustical descriptors \cite{torija2013application}. The idea is to investigate the soundscape using both acoustical measurements and semantic data.  In order to establish the semantic axis \cite{davies_perception_2013} prior to the dimensional analysis, those semantic data usually come from qualitative surveys. Using a Principal Component Analysis (PCA) on descriptors of urban soundscapes, Cain and al.\citep{cain_development_2013,davies_perception_2013} found two independent emotional dimensions named ``\,Calmness\,'' and ``\,Vibrancy\,''.  Using a hierarchical cluster analysis on both semantic differential attributes and acoustical descriptors, Torija and al. \cite{torija2013application}  derive 15 soundscape typologies and show that crest factor and sound level at 125Hz are relevant acoustical variables to recognize types of soundscape. Although the dimensional approach provides important cues to describe soundscape perception, it remains a purely holistic approach as it bypasses the way in which the human brain decomposes soundscapes into perceptual auditory objects \cite{bregman1994auditory, winkler2009modeling} and how each of those objects may influence the qualitative judgment.

The second approach is the categorical approach, which investigates mental representations of soundscape. Mental representations can be seen  as mirrors of the perceived external realities (see Dubois et al. \cite{dubois2006cognitive} p.869 for a definition). They are the memory of the knowledge acquired by a subject, and act as the base of \textit{top-down } cognitive systems. As explained by Dubois et al.  \cite{dubois2006cognitive} ``\,semantic categories can be seen as mediating individual sensory experiences to collective representations by means of a shared language\,''. As they cannot be observed, experimenters must use objectivation methods to reach them. Traditional methods used to objectify human mental representations attempt to derive mental categories and perceptual components, either by relying on verbal descriptions coming from questionnaires or interviews \cite{raimbault_urban_2005, guastavino_ideal_2006, axelsson2005soundscape} (categories of sound sources) or by sorting tasks \cite{guastavino_categorization_2007, maffiolo_caracterisation_1999, houix_lexical_2012} (categories of sound sources and soundscapes).  To identify the mental categories, these methods require psycho-linguistic and lexical analyses. Results from those studies tend to show that the soundscape appreciation depends upon the identification and the assessment of the sound sources which compose the soundscape \cite{raimbault_urban_2005, maffiolo_caracterisation_1999, dubois2006cognitive, guastavino_ideal_2006, guastavino_categorization_2007}. The categorical approach thus agrees with the Auditory Scene Analysis (ASA) theory \cite{bregman1994auditory}, which stands that auditory stimuli are decomposed into ``\,auditory objects\,'' called ``\,streams\,'', which may be regarded as sequences of auditory events emitted by putative sound sources \cite{ciocca2007auditory,winkler2009modeling}. This view is in line with recent cognitive neuroscience studies suggesting that the primary auditory cortex (A1) produces representations of such auditory objects that act as bases for high level cognitive processing \cite{nelken2008neurons}.

As underlined by Davies and al. \cite{davies_perception_2013}, while the categorical approach provides useful information to understand how a soundscape is composed, it does not provide information on how the relations (acoustic or semantic) between the soundscape components (sequences of audio events emitted by sound sources) are perceived and influenced the qualitative evaluation. Furthermore, as category ``\,names\,'' may occur at different semantic levels, categorical approaches based on linguistic analysis are not amenable to facilitating comparison between studies \citep{niessen_categories_2010, brown_towards_2011}. To go further in the categorical approach, there is a need for methods  providing 1) a finer characterization of sound source categories with semantic and numeric data, 2) a description of the inter-relations existing between these sound sources, and 3) subjective data that may be used as basis for inter-studies comparisons.

\section{Method overview}

With those considerations in mind, we propose a new  subject-centered experimental protocol to characterize sound environments with qualitative data related to quantitative data. In this protocol, subjects are asked to simulate complex sound environments with pre-imposed associated qualitative appreciations. To do so, they have access to a soundscape simulator that uses classes of sounds as base elements. The soundscape simulator is thus bound to a sound data set of urban environmental sounds, organized hierarchically around sound categories found in previous studies \cite{maffiolo_caracterisation_1999,dubois2006cognitive, guastavino_categorization_2007, guastavino_ideal_2006, niessen_categories_2010, brown_towards_2011}. The proposed protocol was previously tested and validated thanks to a pilot study\cite{soundscape3} with 10 subjects.

Previous works done by Bruce et al. also use a soundscape simulator to study soundscape perception \cite{bruce2009development,bruce2014effects}. But in their approach, subjects may only add or remove a short set of long recordings of foreground sounds, and adjust their levels. In our approach, subjects are presented with a full data set of classes of urban environmental sounds, that they may manipulate in terms of sound intensity and time positioning.

The proposed protocol is used to characterize two antagonistic soundscapes: subjects are asked to simulate two full urban sonic environments, one ``\,ideal\,'' and the other ``\,non-ideal\,''. The names ``\,ideal\,'' and ``\,non-ideal\,'' have been chosen in order to compare the results with those of a previous study addressing ideal urban sound environment representations \cite{guastavino_ideal_2006}. In this paper, the focus is put  on the acoustical analysis of the simulated scenes to figure out what the characteristics of an ideal (resp. non-ideal) urban environment are, in terms of sound source composition, sound event densities and sound levels. To refine the analysis, sound sources categories and their related quantitative data are investigated according to different semantic levels of categorization (\textit{urban transport}$>$\textit{car}$>$\textit{car-passing}) \cite{niessen_categories_2010, brown_towards_2011}.

The simulated scenes as well as the annotations are available on the archive Data Repository: \url{https://archive.org/details/soundSimulatedUrbanScene}
The experiment web-page is available via the link \url{http://soundthings.org/simScene/}. It should be run on the Chrome browser or the Mozilla Firefox browser.

\section{Experimental paradigm}

The proposed experimental protocol is largely inspired  by hypotheses coming from cognitive psychology\cite{dubois2000categories,dubois2006cognitive} and categorization theory \cite{roach1978cognition}. A sound is considered both for its physical properties and for its semantic value which both depend on the subject and the context (urban environment). The purpose of our  protocol is to objectify human mental representations of the sonic world without the need for a linguistic analysis, by characterizing them with combined semantic and numerical data.

To that end, subjects are asked to recreate a complex sound environment, making use of an environmental sound data set that they may explore without any written textual help thanks to a selection interface designed for this study. The selection process  has been designed to rely only on the listening of sounds themselves, in order not to influence subjects with \textit{a priori} associated semantic values. Once a sound element is selected, subjects must name it. They may then modify some of its physical parameters, thanks to a set of audio controllers.  This framework is depicted on Figure~\ref{Paradigm}. It exposes three types of data to objectify mental representations:

\begin{enumerate}
\item  \textit{Generic semantic data}: the ``\,tags\,'' of the sounds chosen by the subject. A tag is relative to the nomenclature of our typology, and thus pre-defined by experimenters. (ex: \textit{male-voice})
\item \textit{Quantitative data}: the set of ``\,audio parameters\,'' attributed to the sounds by the subject (ex: \textit{sound level} dB).
\item \textit{Non-generic semantic data}: all the subjective verbal data including  the \emph{names} given by the subject to the sounds they selected (ex: \textit{the cries of a man}), a general \emph{title} of the simulated scene, and a \emph{free comment} concerning the creation process.
\end{enumerate}

The proposed paradigm can thus be seen as the inverse of that of a description task (questionnaire, interviews) that uses audio data as input (see Figure~\ref{psycholing}). In contrast with interviews and questionnaires, which both require subjects to describe a soundscape, \textit{i.e.} to decompose the soundscape in elements, subjects are asked to recompose the soundscape from a sound data set of urban environmental sounds. This sound data set  represents the ``\,sonic world\,'': a semantic discretization of it in term of  ``\.sound sources\,''. It ideally  provides the subject with all the sound diversity he may desire.

The proposed approach questions the following: ``\,What sounds have been \emph{used}?\,'', ``\,how have they been \emph{used}?\,'' and  ``\,how have they been \emph{named}?\,'', whereas descriptions  tasks respond to:  ``\,What sounds have been \emph{named}?\,'' and ``\,how have they been \emph{described}?\,''.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=1]{1.eps}
  \caption{\label{psycholing} Sketch illustrating the psycho-linguistic analysis paradigm (top) and the proposed approach (bottom)}
  \end{center}
\end{figure}

 \begin{figure*}[t]
\begin{center}
\includegraphics[scale=1]{2.eps}
  \caption{\label{Paradigm} Sketch illustrating the paradigm of the proposed experimental protocol. \textit{square}: non-generic semantic data; \textit{star}: generic semantic data; \textit{circle}: audio and quantitative data; \textit{Dashed-line}: data not addressed in this paper; \textit{Line}: data addressed in this paper}
  \end{center}
\end{figure*}

\section{Simulating the soundscape}

\subsection{Soundscape model}

In order to allow the creation of a complex sound environment, a model of soundscape has been designed whose key aspects are now presented. A soundscape is assumed to be a sum of tracks. Each track is modeled by a temporal sequence of sound samples that belong to the same sound class. To this sense, each track is related to one specific sound class (leaf sound class more specifically, see \ref{subsection:Sound dataset}). To generate a soundscape, subjects may interact with the tracks, but not with specific sound samples. In other words, if a subject wishes to put a sequence of car sounds, he may choose the sound class \textit{car passing} and manipulate its associated track, but he cannot interact with individual sound samples of \textit{car passing}.

Concerning the sound classes, the commonly accepted distinction made between ``\,sound events\,'' and ``\,sound textures\,''  is taken into account. A  soundscape  is thus regarded as  ``\,a skeleton of events on a bed of texture\,'' \cite{nelken_ear_2013}. Several studies point out the fact that textures and events drive two distinct cognitive processes \cite{maffiolo_caracterisation_1999, saint-arnaud_classification_1995, mcdermott2013summary}. If there is to some extend an analogy between the  notion of ``\,sound event\,'', and the Schaferian notion of ``\,signal\,'', the analogy is not obvious between the notions of ``\,texture\,'' and keynote. Schafer's definition of keynote is perceptively motivated whereas most  definitions of texture  adopt a morphological approach (see \cite{saint-arnaud_classification_1995} for a definition of texture, which is the one adopted in this paper).

Following this distinction (events \textit{vs} textures), two separate treatments are handled by the soundscape simulator:

\begin{itemize}
\item \textit{For the sound event classes} the temporal sequence is made of randomly selected sound samples belonging to the considered sound class of events. the temporal sequence is scaled by the mean/average spacing time between the sound events, the beginning and the end of the sequence.
\item \textit{For the sound texture classes} the temporal sequence is made of randomly selected texture samples belonging to the considered texture class, which are sequenced without time space (crossfading is used to guarantee that the transition is seamless). Only the beginning and the end of the sample sequence may be set by the subject. Furthermore, in order to avoid any unnatural effect due to the repetition of identical sounds \cite{agus_rapid_2010}, the soundscape simulator guaranties that a sample will never transition to itself.
\end{itemize}

Six audio parameters allow the subject to manipulate sound sequences of each selected class, in terms of sound levels and time positioning, see Table~\ref{Audio parameters}. The parameters have an effect on all the samples of a sound sequence. To this sense, sound levels, and time spaces between events (only for events classes) are controlled in terms of average and variance between all the sound samples. Global parameters as global fade in/out and start/stop positions act on all the sequence. A last parameters called ``\,sample fade in/out\,'' allows subject to set a same fade effect separately on each sample of an events sequence (only for events classes).

\subsection{Sound data set}
\label{subsection:Sound dataset}

Segmenting the sonic world to provide the subject with a reasonable and representative number of sound events and textures is a crucial step of our experiment. In order to create our sound data set, we rely on previous studies addressing urban sound source categories\cite{maffiolo_caracterisation_1999, dubois2006cognitive,guastavino_categorization_2007, guastavino_ideal_2006, niessen_categories_2010, brown_towards_2011}. Based on the names of those categories, we build two hierarchical structures of event and texture sound classes,  respectively. Top classes of those structures represent concepts as ``\,urban transport\,'' grouping  sound classes such as ``\,boat\,'' or  ``\,cars\,'', which are themselves grouped in sub-classes. The deeper the class level, the lower the variability between the exemplars lying under the class. In this paper, the class levels will be referred as \textbf{semantic levels}, the top classes being at the semantic level 0, the first subclasses at the semantic level 1 and so forth until the leaf classes are reached. Leaf classes are collections of recorded sound samples. Subjects may only interact with the leaf classes. The Figure \ref{datasetexample} illustrates the hierarchical structure.

\subsection{Simulation process}
The simulation process involves two steps (see Figure~\ref{task}), each of them relying on particular data resources and software interfaces. The two steps are:

\begin{enumerate}
\item \textit{Sound Class selection}: where the subject has to select a leaf class of sound (``\,car passing\,'', ``\,heavy rain\,'', \ldots{}).  The sele	ction is made without any written verbal help thanks to a particular selection interface.
\item \textit{Sound Class modification}: where the subject may tune the parameters (time and intensity) of the temporal sequence formed by sounds belonging to the selected sound class.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%   METHODOLOGY     %%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Experiment}
\subsection{Participants}
44 post-graduate students of the {\'E}cole Centrale de Nantes (French engineering school), were asked to take part in the experiment. They were 30 males and 14 females and were about the same age (M: 21.6, STD: 2). All of them have been living in the same large French city (Nantes) for at least two years.

\subsection{Apparatus}
483 urban environmental sounds were collected, of which 381 sound events and 102 textures. Among them, 260 events and 72 textures were recorded by the authors. The rest, which proved to be particularly difficult to record, came from existing sound banks. All recordings were performed using the shotgun microphone \textit{Audio Technica AT8035} connected to a \textit{ZOOM H4n} recorder. The use of a shotgun microphone allowed us to isolate as much as possible sound recordings from undesired events. All the sound were normalized to the same root mean square (RMS) level. The experiment was run simultaneously with the 44 subjects spread in three identical rooms with calm environment. Subjects were forbidden to talk to one another. Audio was presented diotically to each participant via \textit{BeyerDynamic DT 990 Pro} semi open headphones. Three experimenters were always present (one in each room) to give instructions and answer queries if needed, including explanations, software installation and warm up session.

\begin{figure}[t!]
\begin{center}
\includegraphics[scale=1]{3.eps}
\caption{\label{datasetexample} Sketch illustrating  the hierarchical structure of both the event and texture data sets. Depending on the considered top class, there could be more than 3 semantic levels.}
\end{center}
\end{figure}
\subsection{Task}

Subjects are asked to successively create two urban soundscapes. The first must be ideal (\textit{ie} the favorite urban soundscape of the subject in which they would like to live), the second non-ideal (\textit{ie} the worst urban soundscape of the subject in which they would not like to live). They are asked to mimic a static listener. Subjects are absolutely not restricted in their design choices, but are forbidden to create physically impossible situations such as ``\,a dog barking every 10 milliseconds\,''. Each simulation involves six steps which are exposed in Figure~\ref{task}. At the start of the experiment, a small tutorial of 20 minutes has been planed in order to familiarize the subject with the software environment. The experiment is scheduled to last about two hours.

\subsection{Data collection and analysis}
\label{Data collection}
Four subjects were excluded from the analysis due to a  misunderstanding of the instructions or a failure to comply with the time limits. 40 subjects completed the experiment successfully, giving us 40 ideal urban auditory scenes (i-scenes) and 40 non-ideal urban auditory scenes (ni-scenes).

The proposed experimental protocol generated a large range of data suggesting numerous avenues of investigations. We here choose to restrict ourselves to 1) the  analysis of the holistic properties of the created scenes such as sound levels, event density and diversity, and 2) a class wise analysis, \textit{i.e.} the analysis of the properties attached to each specific sound class, and a study of the distribution of the classes among the ideal and non ideal scenes. Other aspects are intentionally left for future research. Thus, among the collected data, we focus on 2 types of data: 1) the generic semantic data being the ``\,tags\,'' of the chosen sound classes, and 2) the quantitative data being the sound levels attached to the sound classes. For the holistic analysis, the data are averaged over the scenes, without looking at particular sound classes. For the class-wise analysis, the scene content is investigated with respect to the different sound classes used.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%   RESULTS   %%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[t]
\begin{center}
\includegraphics[scale=1]{4.eps}
  \caption{\label{task}Sketch illustrating the simulation process}
  \end{center}
\end{figure}

\section{Holistic analysis }

For the holistic analysis, all the statistical tests are performed using a Wilcoxon signed ranks test at the 5\% significance level. Observations are obtained by averaging the data related to each scene. Results are presented in the tables using the convention \textit{average(standard deviation)}.

\subsection{Sound levels}

Table~\ref{SNR} shows the results for the mean sound levels averaged over all subjects. For each scene, the mean sound level is obtained by averaging the sample levels. Samples are the elements which composed a temporal sequence of a particular sound class: if a subject selects the event sound class \textit{car passing} and creates a sequence  composed of 10 audio events, 10 samples are considered. This measure is thus not equivalent to the global sound level, but  can reasonably be considered a good indicator of the overall sound level of a scene.
Considering the event samples of the i- and ni-scenes, results show that the sound levels are significantly higher ($p=2.08*10^{-5}$) for the ni-scenes, indicating that sound levels are indeed on average higher for unpleasant sound environments. But if there is a blatant difference between the event sound levels of the i- and ni-scenes, the deviation between the texture sound levels is not significant($p=0.14$). This result suggests that textures have less direct influence on the sound level perception. This could be due to the fact that textures are sounds with low semantic weight \cite{saint-arnaud_classification_1995}, in other words, sounds which are less easy to identify. This is in line with early results of Kuwano et al. \cite{kuwano_memory_2003-1} showing that overall judgment of loudness of a sound environment sequence is not statistically different with average instantaneous judgment of the recalled sound events of the sequence.

\subsection{Diversity and coverage}

The notion of diversity is addressed by observing the number of distinct sound classes (events and textures) used in each i- and ni-scenes. This number depends on the semantic level considered. For example, let us consider two ni-scenes, one having a sound class \textit{car-passing} of the semantic level 2, and the other a sound class \textit{car-starting}, also of the semantic level 2. Both classes  belong to the hierarchical structure \textit{transport} (level 0) $>$ \textit{car} (level 1). We will count 2 distinct sound classes for the semantic levels 2 and 3 (as \textit{car-starting} and \textit{car-passing} have no subclass), and only 1 sound class for the semantic levels 0 and 1.

Averaged results  are shown in Table \ref{Diversity}. Except for the semantic level 0, the diversity is significantly higher for the ni-scenes (level 1: $p = 2.5*10^{-4}$;  level 2: $p =0.006$;  level 3: $p = 0.011$;). This reveals that ni-scenes are composed of a larger variety of sounds than i-scenes, suggesting that a non ideal urban environment contains more distinct sound sources than an ideal urban environment.

To qualify this measure of diversity, and verify that it is not biased by the selection interface, we look at the coverage. The coverage is defined by the percentage of the sound classes (events and textures) of the data set that have been used by all the subjects. For example, if all the classes of the semantic level 1 have been chosen by at least one subject, the coverage for the semantic level 1 will be of 100\%. A low coverage would suggest that some parts of the data set have not been explored. Results are shown in Table \ref{Diversity}. The coverage remains superior to 85\% regardless the semantic level, indicating that most of the sounds have been used for each type of scenes.

\subsection{Event density}

The event density is defined as the number of event sound samples used to generate a scene, regardless of the classes.  We can see that the numbers of sound samples vary widely with the subjects. There is no statistical difference between the densities of the i- and ni-scenes ($p=0.14$). This suggests that the global sound events density  does not influence the qualitative evaluation of a urban soundscape.

\section{Class-wise analysis}
For the class-wise analysis, all the statistical analyses are done using a Wilcoxon rank-sum test (Mann-Whitney-Wilcoxon test) at the 5\% significance level. Observations are obtained by averaging the data related to each class.

\subsection{Semantic features: an effective way to evaluate soundscape quality}
\label{Semantic Tag}

In order to describe the scenes in terms of sound sources, the ``\,tags\,''  of the sound classes (event and texture) chosen by the subjects are now investigated. Figures~\ref{tagI} and~\ref{tagNI} display the ``\,tags\,'' of the semantic levels 1 or 2. For ease of reading, the tags \textit{horn}, \textit{fire alarm} and \textit{car alarm} are grouped into a class named \textit{alarm/horn} the classes \textit{bus} and \textit{train} are grouped into a class named \textit{public transportation}, and the classes \textit{truck}, \textit{car}, \textit{scooter}, \textit{motorcycle} and \textit{start passage-way}  are grouped into a class named \textit{traffic}. Classes with a selection percentage less than 2 \% are not displayed.  We only count the selected sound classes and not the number of samples of their respective sequences.

There is considerable difference between the ``\,tags\,'' of the i- and ni-scenes, confirming that sound semantics play an important role in soundscape evaluation. To address this, each simulated scene is represented by a boolean vector of  $n$ dimensions $S_i = (x_1, x_2, \ldots{}, x_{n})$, $i\in[1,80]$. Each dimension corresponds to a sound class (event and texture) of a particular semantic level (for examples $n=44$ classes for the semantic level 1). Thus $x_1 = 1$ indicates that the sound class $x_1$ is present in the scenes $S_i$ (not present if $x_1=0$ resp.). A \textit{Jaccard} distance is then computed between the vectors $S_i$. To quantify how semantic features characterize both the i- and ni-scenes, the precision at rank 5 ($P@5$) metric is used, \textit{i.e.} the average number of items of the same class among the 5 closest items to a given seed item. For each scene $S_i$, $P@5$ is thus obtained by averaging the number of scenes having the same label as $S_i$ (i or ni) among the five scenes $S_j$ ($j=1,...,5$) that are closest to $S_i$. Results are then averaged over the scenes $S_i$ ($i=1, \ldots{}, 80$).

Results are shown in Table~\ref{jaccardtable}. For the semantic level 1, a $P@5$ of 90\% is found, with a random threshold of 58\%. The deeper the hierarchical level, the higher the $P@5$. We find 92.25\% for semantic level 2 and 92.75\% for semantic level 3. This is a strong indicator that semantic values of sounds are good descriptors to distinguish between the i- and ni-scenes. To refine the analysis, the same test is run considering separately the sound textures and the events to describe the scenes (for the semantic level 1, events: $S_i \in  S_i = (x_1, x_2, \ldots{}, x_{31})$ and texture: $S_i \in  S_i = (x_{32},x_{33}, \ldots{}, x_{44})$). The $P@5$ achieved for the semantic level 1 of 91\% for the events and 78.5 \% for the textures demonstrate that sound events contribute more to the perception of soundscape quality than sound textures.

\subsection{Hierarchical analysis of sound categories}

The sound classes showed in the Figures \ref{tagI} and \ref{tagNI} are close to those found by Guastavino \cite{guastavino_ideal_2006} in a psycho-linguistic study also addressing  ideal urban soundscape perception. Classes that suggest human presence and nature are the most present in i-scenes while classes referring to mechanical and construction work sounds are used for ni-scenes. This fact has been observed in other psycho-linguistic studies \cite{dubois2006cognitive, raimbault_urban_2005, guastavino_categorization_2007}. It confirms the \textit{biophilia} hypothesis that ``\,humans are attracted to nature\,'' (Wilson quoted by Guastavino \cite{guastavino_ideal_2006}). However some differences are to be noted. In her study, Guastavino \cite{guastavino_ideal_2006} point out that ``\,public transports\,'' are typical sounds of an ideal urban environment (also mentioned by Dubois et al. \cite{dubois2006cognitive}). She attributes this observation to the fact that urban environment perception is driven by the meaning attributed to the identified sources. As this meaning is influenced by social values, public transport sounds are better accepted than those of private vehicles. Our results tend to qualify this claim  by showing that sounds of public transports (\textit{bus} and \textit{train}) were chosen for both the i-scenes (3.97\% of the selected sound classes) and the ni-scenes (5.1\% of the selected sound classes). Moreover, the sound densities  (i-scenes: 1.1(2.1); ni-scenes: 1.4(2.2);  $p=0.2$) as well as the sample levels (i-scenes(dB): -1.5(3.1); ni-scenes(dB): -1.6(3.3);  $p=0.47$) do not differ significantly. Thus even if the idea of public transport sounds are well accepted due to societal considerations, the sound itself remains similar to that of any vehicles and appears in the ni-scenes more than \textit{car} or \textit{truck} sounds. Although semantic features influence the qualitative evaluation of public transport sounds (\textit{bus} sounds correspond indeed to 3.64\% of the event classes used in the i-scenes, as much as \textit{bicycle} sounds and more than any other vehicle sounds), it seems that, for this sound category, it is not predominant over the role played by the physical attributes of the sound.

\begin{figure}[t!]
\includegraphics[scale=1]{5.eps} 
    \caption{\label{tagI}  Percentage of the number of sound classes of events (\textit{top}) and textures (\textit{bottom}) used by the subjects for the i-scenes at the semantic level 1 or 2 depending of the sounds}
\end{figure}
\begin{figure}[t!]
\includegraphics[scale=1]{6.eps}
    \caption{\label{tagNI}  Percentage of the number of sound classes of events (\textit{top}) and textures (\textit{bottom}) used by the subjects for the ni-scenes at the semantic level 1 or 2 depending of the sounds}
\end{figure}

Figures~\ref{tagI} and~\ref{tagNI}  indicate a counter-intuitive result showing that the event class \textit{traffic} (composed of the event subclasses \textit{truck}, \textit{car}, \textit{scooter}, \textit{motorcycle} and \textit{start passage-way}) is well represented in the i-scenes (10.3\% of the selected sound classes) as well as in the ni-scenes (21.6\%), thus suggesting that traffic sounds are an integral part of an ideal urban environment. If we look at the sample levels, there is no statistical difference between the i- and ni-scenes (i-scenes (dB): -1.8(2.1); ni-scenes (dB): -1.7(3.6); $p=0.17$). The difference occurs only if we look at the sample densities (i-scenes(dB): 2.7(4.5); ni-scenes (dB): 10.4(11.9); $p=3.6*10^{-5}$) which is significantly superior for the ni-scenes. These results tend to indicate that traffic sounds are not absent from the representation of an ideal urban environment provided their densities are not excessive. If we refine the analysis by assessing separately each event subclass of \textit{traffic}, we find significant differences only for the subclasses \textit{scooter} and \textit{motorcycle}. The former has a significantly higher density  ($p=5.6*10^{-4}$) for the ni-scenes, and the latter has both significantly higher sample levels ($p=0.03$) and density ($p=0.02$) for the ni-scenes. Putting these results with the urban problematic of ``\,traffic noise\,'', it appears that 1) it is mostly the sounds of two-wheeled vehicles that are the cause of the annoyance, and  2) that in this case, sound levels are not a relevant indicator of quality compared to sound density.

The texture class \textit{traffic hubbub} is well present in the i-scenes. There is no statistical difference between the i- and ni-scenes for both the sample levels ($p = 0.11$) and the densities ($p = 0.77$). Considering now the subclasses of  \textit{traffic hubbub} (see Figures \ref{tagI} and \ref{tagI}), we see that the \textit{Traffic hubbub of streets} class has been mostly used for the i-scenes (18.75 \% of the textures used in i-scenes) whereas  the \textit{Traffic hubbub of crossroad} class has been only used for the ni-scenes. If we compare those two subclasses, we find again no statistical differences for both the sample levels ($p = 0.19$) and the densities ($p = 0.37$).
These observations  show the importance of context and expectation in soundscape evaluation. The fact that \textit{traffic hubbub} is not depreciated for ideal urban environments shows that ``\,traffic sounds\,'' are understood as being an inherent element of urban environment. This is in contradiction with an intuitive idea that traffic background are indeed ``\,noises\,''. Similar findings are presented by Guastavino \cite{guastavino_etude_2003}: asking subjects to describe urban background, she found that appreciation of the \textit{traffic background} depends on the subject, as it can be reassuring or even appreciated, providing it is not too loud.

\subsection{Markers}

This section investigates the existence of potential event sound markers of an ideal urban environment (resp. non-ideal urban environment), \textit{ie} an event class which has been mostly used in one type of soundscape. To identify markers, the V-test statistical value is used. For each semantic level, considering the population as being  the total number  $n$ of event classes used for both i-scenes and ni-scenes, $n_k$ the number of event classes used for one type of scene $k$ (ni-scenes or i-scenes), $n_j$ the number of event classes $j$ used for both i-scenes and ni-scenes, and $n_{kj}$ the number of event classes $j$ used for one type of scene $k$ (ni-scene or i-scene), the V-test of the event class $j$ (modality) in the scene type $k$ (group) can be computed as follow:

\begin{align*}
\text{V-test}_{jk} = \dfrac{n_{jk} - n_k \frac{n_j}{n}}{\sqrt{n_k\frac{n-n_k}{n-1}\frac{n_j}{n}(1-\frac{n_j}{n})}} 
\end{align*}

The V-test tests the null hypothesis that the $n_k$ individuals of the group $k$ are randomly drawn from the population of $n$ individuals. Usually the V-test value is assessed at the 5\% significance level, that is, if the null hypothesis is true, the V-test value has a 95\% chance to fall within the confidence interval $[-1.96,1.96]$. Thus the V-test is considered as statistically significant if its absolute value remains superior to 2. It is safer to correct the usual 5\% significance level as we are testing many modalities (49 event classes for the third semantic level). To do so, we use the Bonferroni adjustment \cite{perneger1998s}. The Bonferroni adjustment is approximated by dividing the 5\% significance level by the number of modalities $j$ (event classes) to be tested. For example, if we test the 49 event classes of the semantic level 3, the corrected significance level will be 0.001\%, and the V-test value will be statistically significant if its absolute value is superior to 3.29.

Results are shown on the table \ref{table:markers}. Eleven markers are found across all the semantic levels. Confirming again the \textit{Biophilia} hypothesis, sounds of human (\textit{footsteps}, \textit{female laugh}, \textit{male laugh} and \textit{male footsteps concrete}) and nature (\textit{animal}, \textit{birds} and \textit{birds singing} ) are markers of the i-scenes. The presence of \textit{Church Bell} as a strong marker of the i-scenes can be due to the socio-cultural background of the subjects, in great majority French citizens. It confirms Schafer's claim that sounds that are recognized by a community as integral part of its sonic environment are popular \cite{schafer1977tuning}.

For the ni-scenes, the markers  (\textit{Klaxon}, \textit{Siren}, \textit{vehicle work} and \textit{construction work}) are rather intuitive. Interestingly, none of the event classes related to traffic sounds is a marker of the i- or ni-scenes. It confirms the trend observed in \ref{Semantic Tag} that traffic sounds are understood to be intrinsically  part of a urban environment. In other words, although the event densities of traffic sound classes differ between the i- and ni-scenes, traffic sound classes presence is still not a characteristic of a non-ideal urban sound environment.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%    DISCUSSION AND PERSPECTIVES   %%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\subsection{About protocol validity}
One way to validate the protocol is to look at the subject comments. Looking at the database criticisms, 28 subjects stated that they couldn't find one or several desired sounds, with a maximum of 4 sounds for one subject. From all the missing sounds we identified 26 sound classes at different semantic levels. Among those classes, 16 were effectively present in the database, 1 referred to musical sounds which we had chosen to exclude from the database and only 9 were effectively absent. Among the 16 sound classes, they have all been used by at least one subject, except for one sound class (\textit{stroller/trolley}). Among the 9 missing ones, we find very specific sounds as \textit{sport car} or \textit{teenager voice}. We believe that those results show that the database diversity was sufficient for the purpose of the study.
If we look at the interface criticisms, 32.5 \% of all the subject clearly stood that the selection interface was a ``\,straightforward yet effective way\,'' to find a sound whereas  only 10 \% indicated that they encountered difficulties. The remaining 57.5 \%  did not report difficulties.  
Those results tend to confirm that both the data set and the selection interface are well suited for the experiment.

\subsection{Outcomes}

We believe that four main benefits may derive from using the proposed simulation approach: 1) Other studies based on description tasks use full  soundscapes recordings as input. Those recordings are specific exemplars of a very many potential soundscapes which could occur at the same location, at the same time. The re-composition process prevents experimenters from this bias as the simulated soundscape is directly related to the subject representation. Although the simulation is limited by the diversity of the sound data set, we believe that this bias is more controllable, as the expressiveness of the subjects' responses will be less constrained by the size of the data set  than by the exposure to imposed exemplars of a soundscape. 2) The presence of a structured sound data set allows us to focus directly on the sounds chosen by subjects instead of  verbal data only. Thus it may reduce the potential bias of a psycho-linguistic analysis, the first being the mastering of the subject's language by the subject, the second being the lack of definite terms to fully describe sounds or complex sound environments \cite{guastavino_ideal_2006}, and the third being the high inter-subject variability concerning the description of a same sound. 3)  The simulation protocol provides sound signals of simulated urban environments, annotated in term of sound sources, sound intensity, time positioning and qualitative appreciations. It allows us to appreciate new soundscapes descriptors as sample density, which otherwise would be tedious to obtain from recordings.

To some extend, those simulated scenes may be regarded as cognitive summaries of mental representations and may be a useful data set for all Computational Auditory Scene Analysis (CASA) research  or cognitive neuroscience studies investigating 'auditory object' decomposition \cite{nelken2008neurons} or regularities representation impact on high level perceptive processes\cite{winkler2009modeling}.

\section{Conclusions}

In this paper we introduced a new experimental protocol to study soundscape perception based on a simulation paradigm. Application of this protocol allows us to gain knowledge about the perception of urban soundscape by asking human subjects to simulate two urban sound environments: one ideal and the other non-ideal. The results of the presented study show that:

\begin{itemize}
\item The simulation paradigm provides the same high level sound categories as those observed by questionnaire based studies \cite{guastavino_ideal_2006}, and allows us to refine the hierarchical analysis of those sound categories.
\item A semantic characterization of the soundscape in term of presence~/ absence of sound sources is an effective way to characterize ideal or non-ideal urban environment.
\item Global sound level is indeed a good indicator of pleasantness, but not for specific sound categories.
\item Structural features such as sound event density are found to be relevant descriptors to characterize traffic sounds.
\end{itemize}

These results indicate that associated semantic values of events is an effective information to categorize the simulated scenes between ideal and non ideal scenes. More precisely, the study shows that some event categories may be considered as markers of a specific type of soundscape. We believe that those results could be useful to guide the design of computational categorization paradigms based on automatic sound event detection.

We believe that this protocol allows us to directly objectify important aspects of the mental representations of the subjects by being able to look at 1) the sound classes chosen by the subjects to simulate the scenes and 2) their associated quantitative data.  Lastly, the use of a pre-fixed structured data set may facilitate the inter-subjects or inter-studies comparison as all subjects have access to the same decontextualized material to recreate their own vision (contextualisation) of a particular sound environment.

We believe that performing this experiment with subjects of different socio-cultural backgrounds would be interesting for future work as it would allow us to study the cultural impact on the hearing cognitive processes involved.

 \noindent \textbf{Acknowledgements}

 \setlength{\parindent}{0.7cm}

Research project partly funded by ANR-11-JS03-005-01. Thanks to the 44 students of the Ecole Centrale de Nantes for their willing participation.

%----------------------------------------------------------------------------------------
%	REFERENCE LIST
%----------------------------------------------------------------------------------------
\newpage
\bibliographystyle{unsrtnat}
\bibliography{biblio}

\newpage

 \begin{center}

 \large{Tables}

 \end{center}

\begin{table}[h]
\caption{\label{Audio parameters} Description of the audio parameters}
\begin{tabular}{l l} 
Audio Parameters & Description \\
\hline
Sound intensity (dB) & Average and variance \\
Spacing time (sec) & Average and variance \\
(only for event classes)&  \\ 
Position in Scene (sec) & Start and Stop\\
Fade In/out (sec)  & Global \\
Sample fade In/out (sec)  & For each events \\
(only for event classes)&  \\ 
\hline
\end{tabular}
\end{table}

\newpage
\begin{table}[h]
\caption{\label{SNR} Sound levels: mean sound levels in dB averaged over the subjects}
\begin{tabular}{l c c } 
              & Event classes & Texture classes\\
    \hline
	i-scenes & -6.8 (5.4) & -2.6 (3.9) \\
    ni-scenes& -2.4 (3.2) & -1.6 (2.6) \\
    \hline
\end{tabular}
\end{table}

\newpage
\begin{table}[h]
 \caption{\label{Diversity} Diversity and coverage: the coverage is the percentage of the sound classes (events and Textures) of the data set that have been used by all the subjects. The diversity is the mean number of distinct sound classes chosen by each subject, averaged over the subjects}
\begin{tabular}{c  c c c c} 
    Semantic  & \multicolumn{2}{c}{i-scenes} & \multicolumn{2}{c}{ni-scenes} \\
    %\cline{2-5}
     level& Coverage & Diversity & Coverage & Diversity \\
     \hline
    0    & 100 \% & 5.7 (1.4) & 100 \% & 5.7 (1.4) \\
    1    & 95  \% & 7.6 (2.7) & 95  \% & 9.5 (2.5) \\
    2    & 85  \% & 8.1 (3.2) & 89  \% & 9.7 (2.7) \\
    3    & 86  \% & 8.3 (3.4) & 89  \% & 9.7 (2.7) \\
    \hline
\end{tabular}
\end{table}

\newpage
\begin{table}[h]
\caption{\label{density} Density of the sound events: mean number of sound events of each scene averaged over the subjects}
\begin{tabular}{l c} 
          & Density of the sound events \\
    \hline
	i-scenes & 53 (65) \\
    ni-scenes & 63 (64) \\
    \hline
\end{tabular}
\end{table}

\newpage
\begin{table}[h]
\caption{\label{jaccardtable} Precision at rank 5 ($P@5$) computed from the \textit{Jaccard} distances between the scenes for different semantic levels}
\begin{tabular}{c c c c} 
%Semantic   &  \multicolumn{3}{c}{$P@5$} \\
Semantic level & Event and texture & Event & Texture \\
\hline
0  & 81 \%    & 76 \%   & 70 \%    \\
1  & 90 \%    & 91 \%   & 78 \%  \\
2  & 92 \% & 89 \% & 80 \% \\
3  & 93 \% & 91 \% & ---     \\

  \hline
\end{tabular}
\end{table}

\newpage

\begin{table}[h]
 \caption{\label{table:markers} Event classes found to be markers. In each cell, markers are ordered using descending order of V-test values}
\begin{tabular}{c c c c} 
Semantic  level &  \multicolumn{2}{c}{Markers} \\
& i-scenes & ni-scenes \\
\hline
0  &  & construction work  \\
\hline
1  & church bell  & klaxon  \\
   & bicycle bell & siren  \\
   & animal       & vehicle work  \\
   & footsteps    &   \\
   \hline
2  & church bell  & klaxon \\
   & birds        & siren \\
   & bicycle bell & vehicle work \\
   & female laugh & \\
   & male laugh   & \\
   \hline
3  & church bell  & klaxon \\
   & birds singing & siren \\
   & bicycle bell & vehicle work \\
   & female laugh & \\
   & male footsteps concrete &  \\
  \hline
\end{tabular}
\end{table}

\newpage

 \begin{center}

 \large{Figure Captions}

 \end{center}

 \noindent
 Figure 1.  Sketch illustrating the psycho-linguistic analysis paradigm (top) and the proposed approach (bottom).

 \smallskip

 \noindent
 Figure 2.  Sketch illustrating the paradigm of the proposed experimental protocol. \textit{square}: non-generic semantic data; \textit{star}: generic semantic data; \textit{circle}: audio and quantitative data; \textit{Dashed-line}: data not addressed in this paper; \textit{Line}: data addressed in this paper.

  \smallskip

 \noindent
 Figure 3.  Sketch illustrating  the hierarchical structure of both the event and texture data sets. Depending on the considered top class, there could be more than 3 semantic levels.

 \smallskip

 \noindent
 Figure 4.   Sketch illustrating the simulation process. 
 \smallskip

 \noindent
 Figure 5. Percentage of the number of sound classes of events (\textit{top}) and textures (\textit{bottom}) used by the subjects for the i-scenes at the semantic level 1 or 2 depending of the sounds.

 \smallskip

 \noindent
 Figure 6.   Percentage of the number of sound classes of events (\textit{top}) and textures (\textit{bottom}) used by the subjects for the ni-scenes at the semantic level 1 or 2 depending of the sounds.  
 \smallskip

\end{document}

### 2017-03-06 lafaySimulatedScenePerception.tex
rm density and diversity from analysis

--- 2016-08-01 part1/lafaySimulatedScenePerception1.tex
@@
-% Instructions for Authors fuer Acta Acustica
-
 \documentclass[twoside,twocolumn]{article}
@@
 \usepackage{setspace}
+\usepackage{graphicx} 
+\usepackage{subfig}
+\usepackage{caption}
+\usepackage{sectsty}
+
+\allsectionsfont{\bfseries}
+
+\newtheorem{mydef}{Définition}
 
@@
 
-% \footnotemark[1] \Authornotes{\footnotetext[1]{} }
-
 \newcommand{\gl}[1]{\textcolor{blue}{GL : #1}}
@@
 
-\Englishtitle{Investigating soundscapes perception through acoustic scenes simulation. Part I: simulation protocol presentation and case study} \Germanorfrenchtitle{}
+\newcommand{\ie}{\emph{i.\,e.}}
+\newcommand{\Ie}{\emph{I.\,e.}}
+\newcommand{\cf}{cf.}
+\newcommand{\Cf}{Cf.}
+\newcommand{\vs}{\emph{vs.}}
+\newcommand{\Vs}{\emph{Vs.}}
+\newcommand{\al}{\emph{et~al.}}
+\newcommand{\eg}{\emph{e.\,g.}}
+\newcommand{\Eg}{\emph{E.\,g.}} 
+\newcommand{\etc}{\emph{etc.}}
+\newcommand{\myfloatalign}{\centering}
+
+\Englishtitle{Investigating soundscapes perception through acoustic scenes simulation.} \Germanorfrenchtitle{}
 
@@
 \Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process.
-Subjects are asked to create a full soundscape by means of a  software dedicated to sound \gl{edition}, coupled with a structured sound data set.
+Subjects are asked to create a full soundscape by means of a  software dedicated to sound edition, coupled with a structured sound data set.
 This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes.
@@
 %\setstretch{2}
+
 \section{Introduction}
+\label{sec:intro}
 
-%% Original intro
+\gl{[TODO ...]} \\
 
-%% The notion of soundscape has been introduced by Schafer \cite{schafer_new_1969, schafer1977tuning} in the 1970s as the auditory equivalent to landscape.
-Following this paradigm, a sonic environment is described by focusing on the listener's evaluation, rather than only taking into account the acoustic parameters of the sound.
+Comme nous l'avons vu, la recherche sur les paysages sonores a besoin d'outils permettant d'analyser séparément les influences des différentes sources sur les qualités affectives de l'environnement.
+La simulation offre des possibilités intéressantes (\cf~Section~\ref{sec:ch4_modAnaSo}), car elle nous permet d'obtenir des scènes sonores dont nous connaissons tous les paramètres structuraux, en particulier les caractéristiques distinctes des différentes sources.
 
-%% Only during the 1980s did policy-makers started taking into account the link between \textit{noise} and pollution, considering noise as a significant degradation of the quality of life.
-To fight this phenomenon, the first approach consisted in identifying unwanted sounds, and lowering their intensities.
-Thus, in the wake of this realization, several regulations have emerged, often limited to enforcing sound level thresholds.
+Pour démontrer l'intérêt de la simulation en analyse sensorielle, nous choisissons, comme cadre applicatif, le problème de l'agrément perçu dans les environnements sonores urbains.
 
-%% However, it is now widely accepted that acoustic comfort is a complex concept that cannot be described solely with the help of objective acoustical measures such as $LA_{(eq,V)}$, but needs to take into account other perceptive factors \cite{schulte-fortkamp_soundscape_2006, aletta2016soundscape, Yang2005211, kang_semantic_2010}.
-``\,Noise\,''  is a cognitive object which depends upon a listener's appreciation and the context in which it is heard.
-Many urban ``\,noises\,'', such as that of a siren, can annoy as well as warn of a danger, and many town districts are appreciated because of a lively and animated atmosphere which often results in higher sound levels.
-Conventional noise regulations concentrate on high sound levels from transportation and industry, and their scope does not extend beyond unwanted sound sources.
-In other words, they do not try to improve the sonic environment by identifying and reinforcing pleasant or important sounds.
-This latter approach was pioneered by Schafer, who named it the ``\,positive\,'' approach.
+Ce Papier présente les résultats d'une série d'expériences s'appuyant sur la simulation, et visant, chacune, à comprendre comment les différentes sources sonores qui composent une scène influent sur la perception de l'agrément:
 
-%% For two decades, the soundscape paradigm has been considered to study the perception of acoustic scenes in order to provide urban planners with meaningful sound quality indicators (see \cite{dubois2006cognitive} and  \cite{aletta2016soundscape} for extensive reviews).
-Most studies follow the \emph{modus operandi} of cognitive psychology \cite{maffiolo_caracterisation_1999}, by first describing the soundscape using both acoustic measurements and perceptive factors, and then investigating the interactions between those descriptors to understand the cognitive processes involved in the perception of the acoustic scenes.
-Soundscape studies are thus interdisciplinary \cite{davies_perception_2013, aletta2016soundscape}, borrowing tools from various fields such as acoustics, psycho-physics or cognitive psychology, but also psycho-linguistics, sociology and data mining.
+\begin{enumerate}
+\item \emph{expérience 1.a, simulation}: au cours de cette expérience, les sujets simulent les environnements qui serviront de stimuli pour les étapes suivantes;
+\item \emph{expérience 1.b, évaluation de l'agrément}: les sujets doivent évaluer l'agrément des scènes simulées à partir d'une échelle sémantique;
+\item \emph{expérience 2, évaluation de l'agrément après modification des scènes}: comme pour l'expérience précédente, les sujets doivent évaluer l'agrément des scènes simulées à partir d'une échelle sémantique.
+Cependant les scènes ont été modifiées, \ie~privées de certaines classes de sons identifiées comme ayant un impact sur l'agrément perçu;
+\end{enumerate}
 
-%% The soundscape approach appears to be a powerful tool to develop perceptively motivated indicators \cite{schulte-fortkamp_soundscape:_2007}.
-That said, the use of various experimental protocols to assess and measure soundscapes makes the integration of results difficult \cite{davies_perception_2013}.
-Also, there still exist no definite consensus in the soundscape community about the acoustic descriptors to be used \cite{aletta2016soundscape}.
-This issue prevents experimenters to provide  urban planners with generic and clear indicators or models able to predict sound environment quality \cite{schulte-fortkamp_soundscape:_2007, aletta2016soundscape}.
-Recently, several projects such as the European Cooperation in Science and Technology Action (TD0804: soundscape of European Cities and Landscapes) or the Positive Soundscape project \cite{davies_perception_2013} have been undertaken to standardize soundscape assessment and descriptors \cite{schulte2013soundscape}, but those matters are still considered as open problems.
+Le papier comprend 5 sections.
+La première constitue notre introduction.
+La deuxième introduit l'outil de simulation \emph{SimScene}.
+La troisième reprend l'expérience de simulation à proprement parler, ainsi que l'épreuve \emph{évaluation de l'agrément}.
+La quatrième décrit l'épreuve \emph{évaluation de l'agrément après modification des scènes}.
+Nous concluons dans la dernière partie.
 
-%% We believe that soundscapes studies greatly suffer from one important issue: to the best of our knowledge, there is no morphological definition of a soundscape, and therefore no standardized way to decompose it in distinct elements.
-It is thus difficult to study precisely which component of the environment is the cause of potential variations in the perceptive measurements.
-
-%% We propose to tackle this issue by introducing a new experimental protocol, in which  subjects are asked to create a complex sonic environment thanks to a dedicated software tool linked to a dataset of isolated environmental sound samples.
-The simulation provides us with simulated scenes fully annotated in terms of source presence and sound levels.
-Those annotations enable us to ground the analyses of further perceptive descriptors obtained via standard perceptive evaluations of the simulated scenes, as well as to study the contribution of each sound source.
-
-%% As a case study, we focus in this paper on urban soundscape pleasantness.
-We first give an overview of the tools used in soundscape perception studies, and justify the benefits of conducting this kind of simulation experiment as a data generation process.
-We then describe the simulation process, and present the results of the case study.
-The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale.
-
-%% Results show that 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding,  2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterise the appraisal than global descriptors.
-
-%% Version MR 19/05/2016
-
-The study of our sonic environments, both natural and man-made, has gathered growing interest in the past decades, finding applications in fields as varied as context inference and surveillance \cite{heittola13,1540194,park14}, evaluation of urban noise pollution \cite{raimbault_urban_2005} or the emerging area of eco-acoustics \cite{ECOACOUSTICS2014, krause}.
-The notion of \emph{soundscape}, introduced by Schafer \cite{schafer_new_1969, schafer1977tuning} in the 1970s as the auditory equivalent to landscape, is a theoretical framework of choice for that research.
-Following this paradigm, a sonic environment is described by focusing on the listener's evaluation \gl{and perception}, rather than only taking into account the acoustic parameters of the sound scene.
-
-Most soundscape studies follow the \emph{modus operandi} of cognitive psychology \cite{maffiolo_caracterisation_1999}, by first describing the soundscape using both acoustic measurements and perceptual factors, and then investigating the interactions between those descriptors to understand the cognitive processes involved in the perception of the acoustic scenes.
-Soundscape studies are thus interdisciplinary \cite{davies_perception_2013, aletta2016soundscape}, borrowing tools from various fields such as acoustics, psychophysics or cognitive psychology, but also psycho-linguistics, sociology and data mining.
-
-While the soundscape approach appears to be a powerful tool to develop perceptively motivated indicators \cite{schulte-fortkamp_soundscape:_2007}, the use of various experimental protocols to assess and measure soundscapes makes the integration of results difficult \cite{davies_perception_2013}.
-Moreover, there still exist no definite consensus in the soundscape community about the acoustic descriptors to be used \cite{brocolini2012prediction,aletta2016soundscape}.
-We believe that soundscapes studies greatly suffer from one important issue: to the best of our knowledge, there is no morphological definition of a soundscape, and therefore no standardized way to decompose it into distinct elements.
-It is thus difficult to study precisely which component of the environment is the cause of potential variations in the perceptual measurements.
-
-We propose to tackle this issue by introducing a new experimental protocol, in which subjects are asked to create a complex sonic environment thanks to a dedicated software tool linked to a dataset of isolated environmental sound samples.
-The simulation provides us with simulated scenes fully annotated in terms of source presence and sound levels.
-Those annotations enable us to ground the analyses of further perceptual descriptors obtained via standard perceptual evaluations of the simulated scenes, as well as to study the contribution of each sound source.
-
-As a case study to demonstrate the interest of that new approach, we focus in this paper on the issue of urban soundscape pleasantness as tackled in \cite{guastavino_ideal_2006} \nm{dans [14], il est plutot question de ideal / non ideal urban soundscapes .....
-}.
-
-Only during the 1980s did policy-makers start taking into account the link between \gl{``\,community noise\,''} and pollution, considering noise as a significant degradation of the life quality.
-To fight this phenomenon, the first approach consisted in identifying unwanted sounds, and lowering their intensities.
-Thus, in the wake of this realization, several regulations have emerged, often limited to enforcing sound level thresholds.
-
-However, it is now widely accepted that acoustic comfort is a complex concept that cannot be described solely with the help of objective acoustical measures such as $L_{Aeq}$, but needs to take into account other perceptual factors \cite{schulte-fortkamp_soundscape_2006, aletta2016soundscape, Yang2005211, kang_semantic_2010}.
-``\,Noise\,''  is a cognitive object which depends upon a listener's \gl{assessment} and the context in which it is heard.
-Many urban ``\,noises\,'', such as that of a siren, can annoy as well as warn of a danger, and many town districts are appreciated because of a lively and animated atmosphere which often results in higher sound levels.
-Conventional noise regulations concentrate on high sound levels from transportation and industry, and their scope does not extend beyond unwanted sound sources.
-In other words, they do not try to improve the sonic environment by identifying and reinforcing pleasant or important sounds.
-This latter approach was pioneered by Schafer, who named it the ``\,positive\,'' approach.
-
-The soundscape paradigm has been employed for two decades to study the perception of urban acoustic scenes (see \cite{kang2006urban,dubois2006cognitive,aletta2016soundscape} for extensive reviews), however, due to the limitations we have pointed out, experiments have failed to provide  urban planners with generic and clear indicators or models able to predict sound environment quality \cite{schulte-fortkamp_soundscape:_2007, aletta2016soundscape}.
-Recently, several projects such as the European Cooperation in Science and Technology Action (TD0804: soundscape of European Cities and Landscapes) or the Positive Soundscape project \cite{davies_perception_2013} have been undertaken to standardize soundscape assessment and descriptors \cite{schulte2013soundscape}, but those matters are still considered as open problems \cite{ribeiro2013heart}.
-We explore in this paper what additional insights our proposed ``\,generative\,'' approach to soundscape perception analysis can bring to light.
-
-We first give an overview of the tools used in soundscape perception studies, and justify the benefits of conducting this kind of simulation experiment as a data generation process.
-We then describe the simulation process, and present the results of the case study.
-The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale.
-\nm{en revoyant cette description, je me dis qu'il y a pt-etre un petit probleme methodologique dans la mesure ou l'on contruit les scenes selon un attribut semantique (l'idealite) et on les evalue sur une autre echelle (l'agrement) .....
-Je pense qu'il faudra bien argumenter c point plus bas dans le texte}
-
-Results show that \nm{Cette section sur les resultats est etonnemment placee, et a priori, de maniere pas tres orthodoxe.
-Je verrai plutot cette synthese des resultats dans l'abstract ou dans la conclusion ....} 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding, 2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterize the appraisal than global descriptors.
-
-\section{Background and motivations}
-\label{statoftheart}
-
-Two approaches, named in this paper \textit{categorical} and \textit{dimensional}, are commonly used to gain knowledge on soundscape perception \nm{this is not really specific to soundscape ...
-but a global approach of perception mechanisms.
-We can find this distinction also, for instance, in the works about musical timbre}.
-
-The \textit{categorical} approach investigates both the existence of 1) soundscape categories and 2) soundscape components (categories of sound sources) using both sorting \cite{maffiolo_caracterisation_1999, guastavino_categorization_2007} and descriptive \cite{axelsson2005soundscape, raimbault_urban_2005, raimbault2006qualitative, guastavino_ideal_2006} tasks (questionnaires or interviews).
-To identify categories, those methods usually require psycho-linguistic and lexical analyses.
-Their results tend to show that soundscape perception relies upon the identification and assessment of the sound sources which compose the soundscape \cite{dubois2006cognitive}.
-Maffiolo \cite{maffiolo_caracterisation_1999} shows the existence of two generic soundscape categories, namely the ``\,event sequences\,'' and the ``\,amorphous sequences\,''.
-\gl{This distinction is also observed in \cite{guastavino_ideal_2006}}.
-Guastavino \cite{guastavino_categorization_2007} as well as Niessen \emph{et al.} \cite{niessen_categories_2010}  show the prominence of specific sound source categories in soundscape perception: ``\,natural\,'', ``\,human\,'', ``\,traffic\,'' and ``\,mechanical\,''.
-Other studies not using the \textit{categorical} approach reveal that sound source identification underlines the perception of others factors, for example Kuwano \emph{et al.} \cite{kuwano_memory_2003-1} show that overall judgment of loudness of a sound environment sequence is not statistically different from average instantaneous judgment of the recalled sound events of the sequence \nm{Cet exemple n'est pas completement clair pour moi :
-certes, la loudness est bien une dimension continue (et donc en cela rentre bien dans ta terminologie de depart) mais je ne vois pas trop en quoi elle a un rapport avec la question de l'identification des sources ...
-a preciser, je pense.}.
-As reported by \cite{davies_perception_2013}, while the categorical approach does help to understand how a soundscape is composed, it does not provide information on how soundscape components are perceived and influence the qualitative evaluation.
-
-Instead of asking subjects to break soundscapes down into categories, the \textit{dimensional} approach makes use of bipolar semantic scales to describe the soundscape either in a holistic way \cite{jeon2013soundwalk, kang_semantic_2010, cain_development_2013, torija2013application} or by considering the influence of specific sound sources \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}.
-From those descriptors, it is possible to use dimensional analysis techniques such as Principal Component Analysis to find relevant perceptual axes \cite{cain_development_2013}, or to connect those perceptive outcomes with objective acoustics measurements using various statistical analysis or modelling techniques such as multiple linear regression \cite{lavandier2006contribution,ricciardi2015sound} or artificial neural network \cite{yu2009modeling}.
-Several semantic scales show promising results when it comes to modeling perceived affective quality, namely ``\,Calmness\,'' and ``\,Vibrancy\,'' \cite{cain_development_2013, axelsson2010principal}, or ``\,Pleasantness\,'' and ``\,Eventfulness\,'' \cite{axelsson2010principal}; but it has been shown that others context dependent (``\,Appropriateness\,'',``\,Familiarity\,'') or music related (``\,Music-likeness\,'' ) scales provide useful complementary information on soundscape perception \cite{aletta2016soundscape}.
-In addition, several studies have shown that multiple factors such as physical, behavioral, social, demographical, and psychological are involved when evaluating acoustic comfort \cite{marquis2005noise, yu2009modeling}.
-\gl{In particular} \cite{yu2010factors} shows that age and education level significantly influence the sound preference \nm{Je pense qu'il faut etre un peu plus critique sur ces questions la (et celles developpees dans la phrase precedente) et sur ce sujet qu'il faut absolument citer le papier de synthese de C.
-Marquis-Favre :} \gl{Attention \cite{marquis2005noise} est une review ...}.
-
-\gl{Recent} studies investigating the influence of sound sources in soundscape perception show that global sound level (perceived or measured) cannot entirely explain the quality of a sonic environment.
-\gl{voir remarque precedente; sur cette question, je privilegierais plutot un artcile de revue (AAA), voire un bouquin (Kang,2006) qu'un proc.
-de conf.
-...} \cite{guyot2005urban} shows that a rise in soundscape quality between two street soundscapes having a similar time-ratio of traffic sounds and a similar global sound level may result from the addition of a positively connotated sound source, e.g.
-belonging to the ``\,human\,'' category \nm{mais pas que de la categorie human; cf.
-le resultat que tu evoques  plus bas de [13] avec les sources type "public transport" egalement connotes plus positif que des "vehicules legers" du fait de leur caractere ecologique et collectif plus important que la voiture individuelle} \gl{eg means exempli gratia (par exemple), de plus nous montrons que bien  que les bus soient acceptés (presence dans les i-scenes), ils ne sont pas caracteristiques d'un environnement ideal (presence dans les ni-scenes)}.
-Using multiple linear regression to predict unpleasantness from descriptors characterizing global sound level as well as time presence and sound level of each source, \cite{lavandier2006contribution} shows that the percentage of explained variance increases when the source descriptors are added to the global sound level.
-Similar findings are made in \cite{ricciardi2015sound}, showing that a combined use of global and source specific descriptors allows to build more relevant model of sound quality.
-Lastly, \cite{lavandier2006contribution} shows that global sound level is not always appropriate to characterize the appraisal, especially in the case of positively rated soundscapes such as \textit{park} or \textit{market}.
-
-In order to be able to study the relative perceptual importance of each sound source, it is beneficial to first gain knowledge about the presence and level of those sound sources within \nm{inside ?} the scene.
-Acquiring such data may be a serious challenge.
-From most of the studies cited above \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}, subjects are asked to evaluate the time presence and/or the sound level of specific sound source categories using bipolar scales.
-To the best of our knowledge, only \cite{lavandier2006contribution} relies on objective measurement of those indicators by manually annotating the recordings used as stimuli.
-This annotation step is hardly manageable for \emph{in-situ} studies such as \cite{ricciardi2015sound} involving a large number of participants.
-Furthermore, only a pre-imposed restricted set of sound categories are evaluated, preventing experimenters to investigate the influence of more precise or alternative organisations of the sound sources.
-
-There is thus a need to 1) ground the study into a precise taxonomy of sound sources, and 2) generate stimuli that are precisely annotated in terms of sound source presence and level.
-We propose to tackle this issue through the use of a simulation protocol in which subjects are asked to simulate specific complex sound environments thanks to a soundscape simulator linked to a dataset of isolated environmental sound samples.
-This approach leads us to consider the soundscape as a sequence of sound sources \cite{nelken_ear_2013}, following a morphological approach of great potential coined as a ``\,skeleton of events on a bed of texture\,''.
-Experimental subjects are able to create a sound environment by controlling separately each of the sources that make it up.
-Consequently, detailed post-analysis measurement of the impact of each source on the global perception of the simulated sound environment can be made.
-
-Previous works conducted by Bruce \textit{et al.} also use a soundscape simulator to study soundscape perception \cite{bruce2009development,bruce2014effects}.
-However, in their approach, subjects may only add or remove a reduced set of long duration recordings of foreground sounds, and adjust their levels.
-In our approach, subjects are presented with a full dataset of classes of isolated sound samples, that they may finely manipulate in terms of sound intensity, repetition and time positioning.
-Important questions about the representativity of the  dataset are studied in Sections~\ref{sec:SoundDataset} and  \ref{sec:datasetAnalyses}.
+Précisons que l'expérience de simulation a fait l'objet d'une étude pilote dont les résultats sont présentés dans  \cite{lafay2013atiam,lafay2014new}.
+L'outil de simulation \emph{Simscene} est présenté quant à lui en détails dans \cite{rossignol2015simscene,lafay2016JAES}.
+Nous résumons ici ses fonctionnalités d'importance pour notre étude.
 
@@
 
-\section{Simulation protocol}
+\section{L'outil de simulation}
+\label{sec:simscene}
 
-\subsection{Sound dataset}
-\label{sec:SoundDataset}
+\emph{Simscene} est un environnement de travail audio-numérique dont la première version a été développée dans le cadre du projet HOULE\footnote{Projet HOULE : \url{http://houle.ircam.fr/}}.
+Il permet de simuler des paysages sonores à partir d'un corpus de sons.
+Il est prévu pour fonctionner sur les navigateurs internet \emph{Chrome} et \emph{Firefox}.
+L'outil a été développé en javascript à l'aide de la bibliothèque \emph{angular.js}\footnote{\emph{angular.js} : \url{https://angularjs.org/}} et du standard \emph{web-audio}\footnote{\emph{web-audio} : \url{http://www.w3.org/TR/webaudio/}}.
+L'interface de sélection (\cf~Section~\ref{sec:simscene_ssf}) a été développée à l'aide de la bibliothèque \emph{D3.js} \cite{d32011}.
 
-The proposed protocol can be seen as the inverse of that of a description task (see Figure~\ref{psycholing}).
-In contrast with interviews and questionnaires, which both require subjects to describe a soundscape, \textit{i.e.} to decompose it into elements, subjects are here asked to recompose the soundscape from a sound dataset of urban environmental sounds.
+Le fonctionnement de \emph{Simscene} se rapproche de celui d'un séquenceur audio.
+Chaque utilisateur choisit une classe de sons via l'interface de sélection (\cf~Section~\ref{sec:simscene_ssf}).
+Une fois la classe de sons sélectionnée, une piste audio, liée à cette classe, est créée.
+L'utilisateur peut alors modifier certaines propriétés de la piste via un groupe de paramètres de contrôle propre à chacune (\cf~Section~\ref{sec:simscene_parametre}).
+Des champs de texte sont prévus afin de permettre à l'utilisateur 1) de nommer chaque piste, 2) de donner un titre à la scène simulée 3) de commenter la scène simulée.
 
-\gl{To simulate a verisimilar sound environment, we need to provide the subject with a reasonable and representative number of isolated sounds.
-This dataset represents the ``\,sonic world\,'', or more precisely a semantic discretization of it in terms of  ``\,sound sources categories\,''.
-The term ``\,category\,'' refers to a group of objects (in our case sound sources) considered as equivalent.}
+\subsection{Banque de données de samples}
 
-\gl{The design of the dataset is largely inspired by early work by Rosch \cite{roach1978cognition} on the formation of categories of concrete objects.
-According to Rosch, the category system may be viewed as a two dimensional space, where the ``\,vertical\,'' dimension deals with the inclusiveness of the categories (\emph{e.g.} how are ordered ``\,human\,'', ``\,voice\,'' and ``\,laugh\,'') while the ``\,horizontal\,'' dimension concerns the segmentation of categories lying in the same level of inclusiveness (\emph{e.g.} ``\,voice\,'' and ``\,car\,'').
-The ``\,vertical\,'' dimension is a taxonomy, where categories are related to one another by means of class inclusion.
-Rosch formalized three levels of abstraction to structure the ``\,vertical\,'' dimension: superordinate, basic and subordinate.
-The greater the inclusiveness of a category the higher the abstraction level.
-Several studies focusing on sound classification  have already used such hierarchical system \cite{houix_lexical_2012}.
-Following these considerations, our dataset is also structured along two dimensions:}
+\subsubsection*{Typologie}
+\label{sec:simscene_Typologie}
+
+Afin de créer un corpus de classe de sons de référence pour la simulation, nous réalisons une typologie des sons environnementaux urbains.
+
+Pour ce faire, une étude bibliographique est effectuée, afin d'identifier les sources et ambiances sonores les plus souvent citées dans les publications.
+Cette étude porte sur 16 articles ou thèses.
+Chacun d'eux traite de la manière dont nous discriminons les paysages sonores urbains.
+Il ressort que plusieurs approches sont possibles:
 
 \begin{itemize}
-\item object dimension: where we make the difference between distinct sounds sources such as ``\,voice\,'', or ``\,car\,'' and which is closely related to the ``\,horizontal\,'' dimension of Rosh.
-\item abstraction dimension: where we make the difference between distinct level of abstraction use to describe a sound (``\,human\,''>``\,voice\,''>``\,laugh\,''), and which is closely related to the ``\,vertical\,'' dimension of Rosh.
+\item 9 articles abordent le problème par une approche perceptive, soit en identifiant ou répertoriant des catégories de sources sonores, soit en étudiant l'impact de classes de sons spécifiques sur la perception de l'environnement: \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories}
+\item 3 articles proposent une classification morpho-typologique, divisant l’environnement sonore urbain en «~zones sonores~» possédant une identité acoustique forte, selon la configuration et la pratique du site: \cite{maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive}
+\item 2 articles répertorient et classifient les sources sonores d’un point de vue expert: \cite{leobon_analyse_1986,brown2011towards}
 \end{itemize}
 
-Figure~\ref{datasetexample} shows a subset of the described hierarchical organization.
-The top classes of our taxonomy represent concepts such as ``\,urban transport\,'', encompassing classes of sound sources such as ``\,boat\,'' or  ``\,cars\,'', which are themselves split into sub-classes.
-At the leaf level, classes are described using sound source names and actions (\textit{car passing, male-voice yelling}).
-The deeper the class level, the lower the variability between the exemplars lying under the class.
-In this paper the class levels will be referred to as \emph{semantic levels}, the top classes being at semantic level 0, the first subclasses at semantic level 1 and so on.
-Leaf classes are collections of recorded isolated sounds.
+Les réactions à la musique étant trop subjectives, et les jugements esthétiques ne pouvant qu'altérer les données d'évaluation, nous choisissons, dans cette étude, de ne pas considérer les sources musicales de type musiciens de rue, radios de voitures, d'appartements, \etc.
 
-The dataset taxonomy being the same for each subject, it acts as a reference for the analysis, allowing us to get a non ambiguous identification of the sounds populating a scene.
-Most soundscape studies use recordings or \textit{in situ} sound environment as stimuli.
-In both cases, these stimuli remain specific exemplars of many potential soundscapes which could occur at the same location and the same time.
-The re-composition process frees experimenters from this bias as the simulated soundscape is directly related to the subject's mental representation.
-While expressivity in this approach is still limited by the diversity of the available sound dataset, we nonetheless believe that it allows the subjects' responses to be less constrained than by an imposed soundscape exemplar, provided that the sound dataset is carefully designed.
+\subsubsection*{Événements et Textures}
+\label{sec:simscene_eventTexture}
 
-\gl{As stated is section \ref{statoftheart}}, a soundscape  is here understood as being  ``\,a skeleton of events on a bed of texture\,'' \cite{nelken_ear_2013}, where sound events are short isolated sounds such as \textit{car passing} or \textit{baby crying}, whereas sound textures are long stationary  sounds such as \textit{traffic} or \textit{crowd sound}, in which the physical characteristics remain stable over long periods of time~\cite{saint-arnaud_classification_1995}.
-Several studies point out the fact that textures and events drive two distinct cognitive processes \cite{maffiolo_caracterisation_1999, mcdermott2013summary}.
-We thus clearly distinguish between ``\,sound events\,'' and ``\,sound textures\,'' and use two distinct datasets.
+\subsubsection*{Taxonomie}
+\label{sec:simscene_getSample}
 
-For ease of reading, we introduce some terms that will be used throughout the paper:
+Nous nommons sample, un enregistrement d'un son isolé, qu'il s'agisse d'un événement ou d'une texture.
+Chaque classe est alors comprise comme une collection de samples jugés perceptivement équivalents.
 
-\begin{description}
-\item[Sample]: the recording of an isolated sound event or texture.
-\item[Sound class]: either a collection of samples or a collection of sub-classes, depending of the semantic level of the class.
-\item[Track]: a temporal sequence of samples belonging to the same sound class.
-As samples of a texture track are concatenated seamlessly (see section \ref{sec:simulationProcess}), a texture track can be seen as being made of one sample.
-\end{description}
+Les classes de sons sont organisées suivant une structure hiérarchique, afin de former une taxonomie.
+La structure taxonomique s'inspire grandement de l'axe vertical de l'organisation catégorielle proposée par E.
+Rosch \cite{rosch1978cognition}, \ie~plus le niveau d'abstraction de la classe est élevé, plus la description de la classe est précise, et plus les sources sonores incluses dans cette classe sont semblables.
+Si le niveau d'abstraction d'une classe est tel que cette dernière possède des sous-classes, alors sa collection de samples est la somme des collections respectives de chacune des sous-classes.
+
+Nous établissons deux taxonomies, une pour les événements (\cf~Figures~\ref{fig:taxonomieEventa} et~\ref{fig:taxonomieEventb}), une autre pour les textures (\cf~Figures~\ref{fig:taxonomieTexturea} et~\ref{fig:taxonomieTextureb}).
+
+Pour les événements, nous considérons quatre niveaux d'abstraction allant des classes les plus globalisantes (niveau d'abstraction 0), aux classes les plus spécifiques (niveau d'abstraction 3).
+Pour les textures, nous ne considérons que trois niveaux d'abstraction.
+
+Notons que la taxonomie des événements est assez proche de celle de \cite{Salamon14}, introduite ultérieurement à notre étude.
+
+\subsubsection*{Acquisition des samples}
+\label{sec:simscene_getSample}
+
+Sur la base des taxonomies précédemment établies, 483 sons isolés ont été collectés, dont 381 événements, et 102 textures.
+Parmi ces samples, 332 ont été enregistrés, et 151 proviennent de différentes banque de sons (\emph{SoundIdeas}\footnote{\emph{SoundIdeas} : \url{ http://www.sound-ideas.com/}} et \emph{Universal SoundBank}\footnote{\emph{Universal SoundBank} : \url{http://www.universal-soundbank.com/}}).
+
+Tous les enregistrements ont été effectués à l’aide d'un micro canon \emph{AT8035}\footnote{Micro canon \emph{AT8035} : \url{http://eu.audio-technica.com/fr/products/product.asp?catID=1&subID=6&prodID=1845}} relié à un enregistreur \emph{ZOOM H4n}\footnote{Enregistreur \emph{ZOOM H4n} : \url{http://www.zoom.co.jp/english/products/h4n/}}.
+L’utilisation du micro canon nous permet d’isoler les événements sonores du brouhaha urbain.
+Pour les textures, il nous permet d’éviter les événements sonores proches du preneur de son.
+Nous pouvons ainsi pointer des "zones sonores", en nous tenant à une certaine distance de ces dernières, afin de capter uniquement le brouhaha émanant de la zone ciblée.
+
+Tous les sons ont été normalisés au même niveau $RMS$ de $-12$ $dB$ (FS), \ie.
+relativement à un niveau pleine échelle (\emph{relative to Full Scale}).
+Dans notre cas, ce niveau pleine échelle est de 1 Volt.
+
+\subsection{Paramètres}
+\label{sec:simscene_parametre}
+
+L'outil de simulation ne propose pas d’interagir avec un sample en particulier, mais toujours avec une piste, \ie~une séquence de samples.
+Plusieurs paramètres permettent au sujet de contrôler la structure de chaque piste.
+
+\begin{itemize}
+\item \emph{niveau sonore} ($dB$): pour chaque sample, les niveaux sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance;
+\item \emph{espacement inter-onset} (seconde): (piste d'événements seulement) comme pour les niveaux, les espacements sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance;
+\item \emph{début et fin} (seconde): le sujet fixe le début et la fin de chaque piste.
+\end{itemize}
+
+Afin de faciliter la simulation, deux paramètres supplémentaires sont proposés:
+
+\begin{itemize}
+\item \emph{fondu par événement} (seconde): (piste d'événements seulement) le sujet fixe une durée de fondu (entrée et sortie), appliquée à chaque sample d'une piste d'événements;
+\item \emph{fondu global} (seconde): le sujet fixe les durées de fondus pour l'entrée et la sortie de la piste. Ces fondus s'appliquent ainsi à l'ensemble des samples de la piste.
+\end{itemize}
+
+Deux de ces paramètres ne s’appliquent que pour les pistes d'événements (\emph{fondu par événement} et \emph{espacement inter-onset}), les samples des textures étant séquencés sans espacement.
+
+\subsection{Interfaces}
+
+\subsubsection*{Interface de sélection}
+\label{sec:simscene_ssf}
 
 \begin{figure}[t]
-\begin{center}
-\includegraphics[width=.4\paperwidth]{../gfx/1.eps}
-  \caption{\label{psycholing} Psycho-linguistic analysis paradigm (top) and proposed approach (bottom).}
-  \end{center}
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=.8\columnwidth]{gfx/ch_4/SSF}\label{fig:ssf}} \par
+        \subfloat[]
+        {\includegraphics[width=\columnwidth]{gfx/ch_4/simscene}\label{fig:simscene}}
+       \caption{Les interfaces de sélection (a) et de simulation (b) de l'outil \emph{Simscene}.}
 \end{figure}
 
-\begin{figure}[t!]
-\begin{center}
-\includegraphics[width=.4\paperwidth]{../gfx/3.eps}
-\caption{\label{datasetexample} Hierarchical structure of the event and texture data sets. Depending on the considered top class, there could be more or less than 3 semantic levels.}
-\end{center}
+La grande majorité des outils permettant de parcourir une banque de données propose une recherche textuelle sur la base de mots clefs.
+L’efficacité de ce principe repose avant tout sur la structure typologique, et la nomenclature de la base de données.
+Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique?
+
+En effet la description verbale du son, si elle est accessible au sujet, peut potentiellement influencer sa sélection.
+Dans les faits, pour construire une scène environnementale «~calme~», le sujet sélectionne a priori les sons référencés sous le vocable \emph{parc}.
+
+Pour limiter l’influence de l’interface sur le sujet, il nous paraît nécessaire de libérer sa recherche de toute information textuelle.
+Nous proposons à l'utilisateur une interface graphique lui permettant d’explorer la banque de sons exclusivement à partir de l’écoute.
+
+La figure~\ref{fig:ssf} présente l'interface pour la banque de données d'événements sonores.
+Visuellement, les classes du dernier niveau (les seules accessibles par le sujet) sont représentées par des cercles, et positionnées sur un plan.
+La disposition des cercles dans l'espace dépend de l’organisation hiérarchique de la base de données: les sous-classes appartenant à une même classe sont proches les unes des autres, et ainsi de suite jusqu'à atteindre les classes des niveaux d'abstraction élevés.
+Le sujet ne peut accéder qu'aux classes du niveau d'abstraction le plus élevé, classes qui ne possèdent pas de sous-classes, et sont directement liées à une collection de samples.
+
+Chaque classe possède un son prototype.
+Ces sons ont été choisis par les expérimentateurs.
+Lorsqu’on clique sur un cercle, le prototype associé à la classe est joué.
+Le sujet parcourt la banque de sons en cliquant sur les cercles.
+
+Cette organisation visuelle à été pensée afin de:
+
+\begin{enumerate}
+\item faciliter le parcours de la banque de données, les sons similaires (au sens des classes) étant proches les uns des autres.
+\item permettre aux sujets de rapidement appréhender toute l'étendue de la banque de données, \ie~l'ensemble des sons disponibles.
+\end{enumerate}
+
+L'interface de sélection a fait l'objet d'une étude approfondie dont les résultats sont publiés dans \cite{lafay2016JAES}.
+
+\subsubsection*{Interface de simulation}
+
+L'interface de simulation propose un rendu graphique schématisé de la scène en cours de création (\cf~Figure~\ref{fig:simscene}).
+La piste est représentée par une bande possédant un axe temporel.
+Sur cette bande, chaque sample est représenté par un rectangle.
+L'espacement entre les rectangles est relatif à l'espacement entre les samples.
+De même, la hauteur des rectangles est proportionnelle au niveau sonore des samples.
+Dans le cas d'une piste de texture, un unique rectangle apparaît sur toute la longueur de la piste, un son de texture ne pouvant être entrecoupé de silences.
+Les caractéristiques des rectangles évoluent en fonction des changements de paramètres de la piste.
+L'utilisateur a la possibilité, à tout moment, d'écouter la scène simulée.
+
+L'interface de simulation est présentée en détails dans \cite{rossignol2015simscene}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%  Experience 1     %%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Expérience 1}
+\label{sec:xp1}
+
+\subsection{Expérience 1.a : planification}
+\label{sec:ch5_planExpSimu}
+
+\subsubsection*{Procédure}
+
+\begin{figure}[t]
+        \myfloatalign
+        \includegraphics[width=1\linewidth]{gfx/ch_4/4}
+       \caption{Etapes du processus de simulation}\label{fig:etapeSimu}
 \end{figure}
 
-\subsection{Sample Selection process}
+Les sujets doivent simuler deux environnements sonores urbains, chacune des scènes devant durer 1 minute. Pour ces simulations, les sujets doivent se conformer aux consignes suivantes:
 
-The sample selection process is designed to rely only on the listening of sounds themselves, in order not to influence subjects with \textit{a priori} associated semantic values.
-To explore the dataset without any written textual help, we have designed a selection interface where each leaf class is depicted by a point in 2D space.
-\gl{The interface performances are extensively tested in the following study} \cite{lafayJAES}.
-Subjects may only interact with the leaf classes of the datasets but the location of the points in space is constrained by the taxonomy, \textit{i.e.} subclasses belonging to a same class are close to one another.
-This organization is chosen to facilitate the sound class selection process and allow the subject to quickly apprehend the full range of the dataset.
-This may also reduce several potential biases of studies relying on verbal data to describe a soundscape: 1) the variable fluency of the subject in the experiment language, 2) the lack of definite terms to fully describe sounds or complex sound environments \cite{guastavino_ideal_2006}, and 3) the high inter-subject variability in the verbal description of a same sound.
+\begin{itemize}
+\item première simulation: simuler un paysage sonore \textbf{urbain plausible} qui, selon vous, est idéal (où vous aimeriez vivre);
+\item deuxième simulation: simuler un paysage sonore \textbf{urbain plausible} qui, selon vous, est non-idéal (où vous n'aimeriez pas vivre).
+\end{itemize}
 
-\subsection{Simulation process and parameters}
-\label{sec:simulationProcess}
+Tous les sujets commencent par simuler l'environnement idéal. Les sujets ne prennent connaissance de la deuxième consigne qu'à la fin de la première simulation.
 
-Following the above described terminology, a soundscape is a sum of tracks.
-Each track is modeled by a temporal sequence of sound samples belonging to the same sound class.
-Thus, each track is related to one specific leaf sound class (see Section \ref{sec:SoundDataset}).
-To generate a soundscape, subjects may interact with the track parameters, but not with specific samples; for example, if a subject wishes to put a sequence of car sounds, he may specify how often (on average) cars pass by, and how loud they should be on average, but he cannot control each event individually.
-This allows to streamline the creation process, and keep the user focused on a global perception of the soundscape.
+Les sujets sont totalement libres dans le choix des sons, et des paramètres. Ils doivent cependant se soumettre à deux contraintes:
 
-The simulation process involves two steps:
+\begin{itemize}
+\item le sujet doit prendre le point de vue d’un auditeur fixe;
+\item le paysage sonore doit être réaliste, au sens de physiquement plausible.
+Autrement dit, le sujet a tout à fait le droit de placer 10 chiens dans son paysage sonore, mais il n’a pas le droit de placer un chien aboyant toutes les 10 millisecondes.
+\end{itemize}
+
+Ces contraintes font partie de la consigne. Aucun contrôle n'est fait \emph{a priori} dans l'interface de simulation.
+
+Chaque processus de simulation comprend deux parties (\cf~Figure~\ref{fig:etapeSimu}) :
 
 \begin{enumerate}
-\item \textit{Sound Class selection:}  the subject selects a leaf class of sound (``\,car passing\,'', ``\,heavy rain\,'', \ldots{}).
-Once a class is selected, a track made of randomly selected samples belonging to the considered sound class is created.
-For a texture track, samples are sequenced without any gap, using crossfades to guarantee seamless transitions.
-The soundscape simulator guarantees that a texture sample will never transition to itself in order to avoid any unnatural effect due to the repetition of identical sounds, that a human would easily detect \cite{agus_rapid_2010}.
-\item \textit{Sound Class parametrization:} the subject may tune the parameters (time positioning and sound levels) of the track.
+\item la réalisation de la simulation: cette étape peut, elle même, se décomposer en trois actions :
+\begin{itemize}
+\item sélectionner les classes de sons
+\item nommer les classes de sons sélectionnées
+\item paramétrer les pistes (\cf~Section~\ref{sec:simscene_parametre}) relatives aux classes de sons sélectionnées
+\end{itemize}
+\item la production d'un commentaire libre du paysage sonore simulé
 \end{enumerate}
 
-Five audio parameters allow the subject to parametrize a track.
-The parameters always affect all the samples of a track and never a particular isolated sound.
-Two parameters, namely \textit{Inter onset spacing time} and \textit{Sample fade In/Out}, are only used with event tracks.
+En complément, et une fois les deux scènes sonores réalisées, les sujets sont invités à:
 
-\begin{enumerate}
-\item \textit{Sound level} (dB): for each track, the sound levels of individual samples are drawn from a normal distribution.
-Subjects may control the sound level of a track by setting the mean and variance of the distribution.
-\item \textit{Global fade In/Out} (sec):  allows the subject to set a fade effect on the whole sequence of samples. 
-\item \textit{Start and Stop position} (sec): allows the subject to set the beginning and the end times of a temporal sequence.
-\item \textit{Inter onset spacing time} (sec): (event tracks only) similarly to the sound level, inter onset spacing times between events are drawn from a normal distribution that may be parametrized by the subject for each event track.
-\item \textit{Sample fade In/Out} (sec): (event tracks only) allows the subject to set a same fade effect individually on each sample of an event track.
-\end{enumerate}
+\begin{itemize}
+\item indiquer les sources sonores qu'ils voulaient mettre, mais qu'ils n'ont pas trouvées;
+\item commenter l’ergonomie du logiciel de simulation;
+\item commenter l’ergonomie de l'interface de sélection.
+\end{itemize}
 
-Considering a soundscape as a sum of isolated samples is in line with the Auditory Scene Analysis (ASA) theory \cite{bregman1994auditory}, which states that auditory stimuli are decomposed into ``\,auditory objects\,'' called ``\,streams\,'', which may be regarded as sequences of auditory events emitted by putative sound sources \cite{ciocca2007auditory,winkler2009modeling}, and also with recent neuroscience studies suggesting that the primary auditory cortex (A1) produces representations of such ``\,auditory objects\,'' that act as bases for high level cognitive processing \cite{nelken2008neurons}.
-Thus, the track can be seen as the generative counterpart of the stream, which is rather a cognitive object.\nm{ce paragraphe est interessant mais est, a mon avis, mal place; il legitime la structure du simulateur, je le mettrais donc plutot dans le §3.1 en le developpant pt-etre un peu plus notamment les aspects neurosciences ...}
+Avant de commencer la première simulation, un tutoriel de 20 minutes est proposé aux sujets, afin qu'ils se familiarisent avec le logiciel de simulation, et la banque de données.
+L'expérience est prévue pour durer 2h30.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%   EXPERIMENT Plan %%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Dispositif expérimental}
 
-\section{Case Study: urban soundscape pleasantness}
-\label{sec:CaseStudyUrbanSoundscape}
-\subsection{Experimental planning}
+Tous les sujets passent l'expérience sur des machines identiques.
+L'audio est diffusé en monophonie, par le biais de casques audio.
+Pendant le tutoriel, les sujets doivent ajuster le niveau sonore à un volume confortable.
+Ils ne peuvent le modifier par la suite.
 
-We apply the proposed protocol to the field of urban soundscapes appraisal, by asking a panel of subjects to simulate two antagonistic urban soundscapes:  one ``\,ideal\,'', or pleasant (i) and the other ``\,non-ideal\,'' or unpleasant (ni) \nm{voir ma remarque en intro.
-sur la concordance ideal/pleasant ...}.
-The two labels have been chosen to compare the results with those of a previous study addressing ideal urban sound environment representations \cite{guastavino_ideal_2006}.
-The simulation protocol used in this experiment was previously tested and validated thanks to a pilot study~\cite{soundscape3}.
+Tous les sujets réalisent l'expérience simultanément.
+Ils sont répartis de manière égale dans trois pièces identiques, toutes possédant un environnement calme.
+Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
 
-%run with 10 subjects.
+Trois expérimentateurs, un dans chaque pièce, sont présents durant la totalité de l'expérience, afin de contrôler le bon déroulement de cette dernière, et de répondre aux éventuelles questions des sujets.
+\\
 
-To further characterize the simulated scenes, we run a complementary experiment where 10 subjects who did not take part in the first experiment are asked to evaluate the pleasantness of the simulated scenes using a bipolar semantic scale.
-This second experiment has two goals:
+\subsubsection*{Participants}
 
-\begin{enumerate}
-\item gain finer knowledge about the pleasantness of each simulated scene, \nm{trop vague a mon sens : a developper}
-\item detect possible scene outliers.
-As the distinction i/ni \nm{cet encodage n'a  jamais ete defini avant - meme si on devine bien ce qu'il veut dire ....
-;)} \gl{cf.
-plus haut} acts as our reference throughout the analysis, one has to ensure that all the scenes are indeed perceived as ideal or non ideal \nm{il ya confusion entre l'evalution ideal/non ideal et sur l'echelle de pleasantness ....
-est-ce la meme expe.
-?
-a clarifier} scenes without ambiguity, \textit{i.e.} the highest average perceptual rating of the ni-scenes has to remain inferior to the lowest average rating of the i-scenes.
-\end{enumerate}
+44 étudiants (14 femmes) de L’École Centrale de Nantes ont participé à l'expérience.
+Ils ont tous sensiblement le même âge (moyenne: 21.6, écart-type: 2).
+Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
 
-The simulation experiment is referred as experiment-1 and the rating experiment as experiment-2.
-The complete experimental plan is described in Figure~\ref{fig:expLan}.
-\nm{Du coup, cette info; doit etre mise avant dans le paragraphe histoire d'evacuer toutes les confusions (cf.
-mes remarques plus haut)} \gl{non, je pense qu'à moins de deux paragraphes d'intervales, les lecteurs vont s'y retrouver.}
+Sur les 44 sujets, 40 réalisent l'expérience avec succès, produisant au final 80 scènes sonores simulées, dont 40 scènes idéales, et 40 scènes non idéales.
+4 sont éliminés pour non respect ou incompréhension des consignes, d'une part, et dépassement du temps, d'autre part.
 
-The simulated scenes as well as their annotations are available on an Archive Data Repository \footnote{\url{https://archive.org/details/soundSimulatedUrbanScene}}.
-Supplementary information including the experiment webpage and raw data for each experiment can be found on the companion website \footnote{\url{http://www.irccyn.ec-nantes.fr/~lagrange/demonstrations/simScene.html}}.
+\subsection{Expérience 1.b : planification}
+\label{sec:ch5_planExpEvaA}
 
-\begin{figure}[t]
-\begin{center}
-\includegraphics[width=.4\paperwidth]{../gfx/5.eps}
-\caption{\label{fig:expLan} Experimental planning.}
-\end{center}
-\end{figure}
+\subsubsection*{Procédure}
 
-\subsection{Dataset of isolated samples}
+En raison de contraintes temporelles, les sujets n'évaluent que des séquences de 30 secondes des scènes simulées, chacune de ces séquences commençant à la seconde 15, et finissant à le seconde 45 de la scène d'une durée d'origine de 1 minute.
 
-The taxonomy, whose first two levels are depicted on Figure~\ref{fig:taxonomie}, is built following the category names found in previous studies addressing urban sound source categories \cite{maffiolo_caracterisation_1999,dubois2006cognitive,guastavino_categorization_2007, guastavino_ideal_2006, niessen_categories_2010, brown_towards_2011}, and is very similar to the one obtained in an other recent study on urban environmental sounds \cite{Salamon14}.
+L'évaluation s'effectue sur une échelle sémantique bipolaire de 7 points, allant de -3 (non-idéale/très désagréable) à +3 (idéale/très agréable).
+Avant de noter une scène, les sujets doivent obligatoirement écouter les 20 premières secondes de cette dernière.
+Après la notation, ils sont libres de passer à la scène suivante.
 
-483 monophonic urban environmental sounds are collected, including 381 sound events and 102 textures are chosen to build the dataset of isolated sound samples.
-Among them, 260 events and 72 textures are recorded by the authors using a \textit{Audio Technica AT8035} shotgun microphone (to isolate sounds from their background) connected to a \textit{ZOOM H4n} recorder.
-The remaining comes from existing sound banks \nm{preciser lesquelles ?} \gl{non}.
-All sounds are normalized to the same root mean square (RMS) level.
+Pour chaque sujet, les scènes sont présentées dans un ordre aléatoire.
+Les 10 premières scènes permettent au sujet de calibrer ses notes.
+Elles sont obligatoirement composées de 5 scènes idéales, et de 5 scènes non-idéales.
+Ces 10 premières scènes sont rejouées à la fin de l'expérience, et seules les notes données à la deuxième occurrence sont prises en compte.
 
-\begin{figure}[t]
-\begin{center}
-\includegraphics[width=.4\paperwidth]{../gfxHierarchy/taxonomy.eps}
-\caption{\label{fig:taxonomie} Taxonomy of (a) sound events, and (b) sound textures for the semantic levels 0 and 1, as well as an intermediate grouping level denoted $0^{+}$.}
-\end{center}
-\end{figure}
+L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.\\
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%  EXPERIMENT Set Up  %%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Dispositif expérimental}
 
-\subsection{Set-up Experiment-1}
-\subsubsection*{Procedure}
+Tous les sujets passent l'expérience sur des machines identiques.
+L'audio est diffusé en monophonie, par le biais de casques audio semi-ouverts \emph{Beyer-Dynamic DT 990 Pro}.
+Toutes les scènes sonores ont été re-simulées sur la base des partitions obtenues lors de l'expérience de simulation.
+Le niveau sonore de sortie est identique pour tous les sujets.
 
-Subjects are asked to successively create two urban soundscapes, lasting one minute each.
-The first must be ideal (a soundscape in which the subject would like to live), the second non-ideal (one in which they would try to avoid living) \nm{reprendre la formulation exacte de la consigne} \gl{C'est la formulation, bien que cette dernière était transmise oralement}.
-They are asked to mimic a static listener.
-Subjects are not restricted in their design choices, but are forbidden to create physically impossible situations such as ``\,a dog barking every 10 milliseconds\,''.
-At the beginning of the experiment, a 20 minutes tutorial is planned in order to familiarize the subjects with the software environment.
-The experiment is scheduled to last from two to three hours.
+Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
 
-Due to time constraints for the experiment-2, we trim 15 seconds from the start and end of each simulated scene, giving us 30 second scenes.
+Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de celle-ci, et de répondre aux éventuelles questions des sujets.  \\
 
-\subsubsection*{Participants and Apparatus}
+\subsubsection*{Participants}
 
-44 post-graduate students (14 females) of the {\'E}cole Centrale de Nantes (French engineering school) take part in the experiment.
-They are about the same age (M: 21.6, SD: 2).
-All of them have been living in the same large French city (Nantes) for at least two years before the experiment.
-The experiment is run simultaneously with the 44 subjects spread into three identical rooms with a calm environment.
-Subjects are forbidden to talk to one another and asked to adjust volume to a comfortable level once, at the beginning of the experiment.
-Audio is presented diotically to each participant via headphones.
-Three experimenters are always present (one in each room) to give instructions and answer queries if needed.
+10 étudiants (2 femmes) de L’École Centrale de Nantes ont participé à l'expérience.
+Aucun d'entre eux n'a réalisé l'expérience de simulation.
+Tous les sujets ont sensiblement le même âge (moyenne: 23.1, écart-type: 1.8).
+Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
 
-Due to a misunderstanding of the instructions or a failure to proceed within the time limits, four subjects are excluded from the analysis.
-40 subjects completed the experiment successfully, providing 40 ideal urban scenes (i-scenes) and 40 non-ideal urban scenes (ni-scenes).
+Tous les sujets ont réalisé l'expérience avec succès.
 
-\subsection{Set-up experiment-2}
-\label{sec:Experiment-2}
-\subsubsection*{Procedure}
-
-Subjects are asked to evaluate the pleasantness of the 80 simulated scenes using a bipolar scale (also known as Likert scale) of 7 points going from -3 (non ideal, very unpleasant) \nm{a nouveau cet amalgame des 2 termes risque d'etre critiquee et doit etre argumenter et expliciter, je pense.} to +3 (ideal, very pleasant).
-The first ten scenes are used to give the subject an idea of the pleasantness range of the scenes.
-Those ten scenes are repeated at the end of the experiment and the ratings of their first occurrences are not taken into account in the analysis.
-Subjects listen to at least the 20 first seconds of the scenes and are free to skip the last 10 seconds at their convenience.
-Scenes are presented in a randomized order, but the ten calibration scenes are five i-scenes and five ni-scenes.
-\nm{also randomized in the presentation along subjects ?} \gl{a vérifier}
-
-\subsubsection*{Participants and Apparatus}
-
-10 post-graduate students (2 females) of the {\'E}cole Centrale de Nantes took part in the experiment.
-None of them did the first experiment.
-They are about the same age (M: 23.1, SD: 1.8).
-All of them have been living in the same large French city (Nantes) for at least two years.
-The experiment is run simultaneously with the 10 subjects in a calm environment.
-Audio is presented diotically to each participant via \textit{BeyerDynamic DT 990 Pro} semi open headphones at -12 dB (full scale).
-One experimenter is always present to give instructions and answer queries if needed.
-All the  subjects successfully completed the experiment.
-
-\subsection{Data analysis}
-
-As experiment-1 generates a large amount of data suggesting numerous avenues of investigation, we here choose to restrict ourselves to the analysis of 1) selected sound classes, 2) sound event density, and 3) scene sound levels, other aspects being intentionally left for future research.
-
-For ease of reading, we introduce the following terms and concepts that will be used throughout the analysis: 
-\begin{description}
-\item[$Diversity$:] number of distinct sound classes used in a scene.
-This number depends on the considered semantic level.
-For example, considering the two subclasses of the semantic level 2 \textit{car-passing} and \textit{car-starting}, which both belong to the same class \textit{car}, we count 2 distinct sound classes for semantic levels 2 and 3, but only 1 sound class for semantic levels 0 and 1.
-\item[$Density$ ($D$):] logarithm of the number of event samples per chunk of 1 second, averaged over the scene duration.
-Density may be computed considering all sound classes, or only a subset of them.
-If nothing is specified, then the density accounts for all classes.
-\gl{Empirically, we observed that the original distributions were skewed.
-The use of the logarithm compression allowed us to reduce this skewness.}
-\item[$Sound level$ ($L$):]  Decibel of the RMS voltage relative to 1 volt, computed every second and averaged over the scene duration.
-\nm{est-ce que cette reference correspond a une quelconque calibration du microphone
-de prise de son (p.
-ex., X dB a 1 m.) ?
-Si oui, donner plus de details et expliquer comment tu as fait pour les samples issus de base de donnees - dont on ne connait pas a priori ces elements de calibration .....} \gl{non} An A-weighting filtering is performed before the RMS computation.
-Computing other statistical descriptors, namely the minimum, maximum or the 10-90th quantiles, shows that they all present a high correlation with the mean ($r\geq0.85$, $p<0.01$), therefore only the latter is considered.
-\end{description}
-
-We measure significance between the i- and ni-scenes using paired Student's t-test.
-All correlations are computed using the Pearson product-moment correlation coefficient.
-All statistical tests are performed at an $\alpha=5\%$ significance level.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%  Results       %%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Results}
-
-\subsection{Validity of the  dataset and the selection interface}
-\label{sec:datasetAnalyses}
-
-The dataset and selection interface are now assessed by considering the subject comments.
-
-Looking at the database criticisms, 28 subjects stated that they could not find one or several desired sounds, with a maximum of 4 sounds for one subject.
-From all the missing sounds we identified 26 sound classes at different semantic levels.
-Among those classes, 16 were effectively present in the database, 1 referred to musical sounds which we had chosen to exclude from the database, and only 9 were effectively absent.
-Except for one sound class (``\,stroller pushchair\,''), each of the 16 classes which were reported as missing but were present in the database were used by at least one other subject.
-Among the 9 truly missing ones, we find very specific sounds as ``\,sport car\,'' or ``\,teenager voice\,''.
-We believe \nm{n'est-ce pas unpeu faible comme argumentaire ....?
-Pt-etre mettre les resultats en \% pour les rendre plus facilement intrerpretables ?} that those results show that the database diversity was sufficient for the purpose of the study.
-
-If we look at the interface criticisms, 32.5\% of all the subject clearly stated that the selection interface was a ``\,straightforward yet effective way\,'' to find a sound whereas only 10\% indicated that they encountered difficulties.
-The remaining 57.5\% did not report difficulties.
-
-Those results tend to confirm that both the data set and the selection interface are well suited for the experiment.
-
-\subsection{Consistency of the data}
-
-To ensure consistency of the simulated scenes, no ni-scene should be rated in average as more pleasant as any i-scene, and reciprocally.
-Figure~\ref{xp2_1}.a shows the distributions of the averaged ratings considering scenes as observations.
-Four scenes (2 i-scenes and 2 ni-scenes) appear to be ambiguous, and are henceforth removed from the analysis, as well as their antagonist scenes.
-This operation leaves us with 36 i-scenes and 36 ni-scenes.
-Corrected distributions are displayed on Figure~\ref{xp2_1}.b.
-Figure~\ref{xp2_1}.c shows the distributions of the averaged ratings, considering subjects as observations, without the scene outliers.
-It is clear that the pleasantness levels of the scenes, as intended by experiment-1 subjects, are well perceived by experiment-2 subjects ($p<0.01$).
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[width=\columnwidth]{../gfxMatlab/xp2_1.eps}
-\caption{\label{xp2_1} Boxplot of the averaged ratings given by the subject considering (a) scenes as observations, (b) scenes as observations without scene outliers and (c) subjects as observations without scene outliers.
-\nm{les figures sont petites et le chiox des couleurs contribue a leur faible lisibilite} \gl{?}}
-\end{center}
-\end{figure}
-
-\subsection{Density, sound level,  and diversity}
-\label{sec:levelDensityDiversity}
-
-\begin{figure}[t!]
-\centering
-\includegraphics[width=\columnwidth]{../gfxMatlab/xp1_deSoLv_1.eps} 
-\caption{\label{fig:xp1_deSoLv_1}  Boxplot of (a) the scene densities ($D$) and (b) the scene sound levels ($L$). Scatter plot of (c) $D$ \emph{vs} ratings and (d) $L$ \emph{vs} rating}
-\end{figure}
+\subsection{Données et méthodes d'analyses}
 
@@
 \centering
-\begin{tabular}{l c c} 
-                & $L$                          & $D$    \\
+\begin{tabular}{c c} 
+Termes                         & Acronymes              \\   
 \hline                    
-all             & \textbf{-0.76} ($p<0.01$)    & \textbf{-0.35} ($p<0.01$) \\
-i-scenes        & -0.32 ($p=0.06$)             & -0.28 ($p=0.10$) \\
-ni-scenes       & \textbf{-0.66} ($p<0.01$)    & \textbf{-0.47} ($p<0.01$) \\
-
+Niveau                         & $L$                    \\ 
+Niveau (événements)            & $L(E)$                 \\
+Niveau (textures)              & $L(T)$                 \\ 
+Agrément moyen (par scène)     & $\mathcal{A}_{scene}$  \\
+Agrément moyen (par sujet)     & $\mathcal{A}_{sujet}$  \\
+                               &                        \\
+idéale/agréable                & i                      \\
+non-idéale/désagréable         & ni                     \\
+scène idéale/agréable          & i-scène                \\
+scène non-idéale/désagréable   & ni-scène               \\
 \hline
@@
 \vspace{0.5mm}
-\caption{\label{tab:numGlobal} Correlations computed between ratings \emph{vs} sound level ($L$) and event density ($D$)}
+\caption{Acronyme des variables utilisées dans le cadre des expériences sensorielles.}
+\label{tab:acronyme}
 \end{table}
 
-\begin{figure}[t!]
-\centering
-\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_div_1.eps} 
-\caption{\label{fig:analyseDensitySnrGlobal3} Diversity across semantic levels for event and texture classes.
-\nm{il faut mieux differencier les donnees provenant du corpus event et du corpus texture}}
+Chaque scène est décrite par un groupe de descripteurs.
+C'est sur ces descripteurs que nous pratiquons l'analyse.
+Un résumé des descripteurs, ainsi que des acronymes les désignant, est présenté dans le Tableau~\ref{tab:acronyme}.
+Afin de rester cohérent avec l'épreuve d'évaluation, les descripteurs issus des partitions, ou des signaux des scènes, ne sont pas calculés sur la durée totale de celles-ci, mais sur une version réduite de 30 secondes (\cf~Section~\ref{sec:ch5_planExpEvaA}).
+
+Pour chaque scène sonore, trois types de descripteurs sont considérés:
+
+\begin{itemize}
+\item \emph{perceptif}: il s'agit de l'agrément perçu de la scène simulée, évalué sur une échelle sémantique de 7 points.
+Nous notons $\mathcal{A}_{scene}$ l'agrément moyen d'une scène, obtenu en moyennant les notes de tous les sujets.
+De même, nous notons  $\mathcal{A}_{sujet}$ l'agrément par sujet, en moyennant l'ensemble de ses notes.
+Compte tenu du faible nombre de sujets, nous faisons le choix, dans cette étude, de ne pas normaliser les notes d'agrément;
+\item \emph{sémantique}: il s'agit d'un vecteur booléen noté $S=(x_1,x_2,\ldots,x_n)$, indiquant les classes de sons présentes dans la scène.
+Chaque point $x$ de ce vecteur correspond à une classe de sons particulière: $x=1$ si la classe est présente dans la scène, et $x=0$ autrement.
+La dimension $n$ des vecteurs dépend du niveau d'abstraction considéré, \eg~pour le niveau d'abstraction 1, qui comprend $44$ classes de sons, cette dimension sera de $n=44$.
+\item \emph{structurel}: il s'agit du \emph{niveau Sonore}.
+Pour représenter le niveau sonore, nous nous inspirons de la mesure $L_{Aeq}$.
+Dans notre cas, il s'agit d'un scalaire, calculé sur le signal en volts, et non en pression, et donné en décibels, en prenant un référentiel de 1 Volt.
+Le niveau est obtenu en calculant, toutes les secondes, la moyenne quadratique du signal, et en moyennant sur la durée de la scène.
+Un filtrage de type A est opéré avant le calcul des moyennes quadratiques.
+Nous notons $L$, $L(E)$ et $L(T)$, les niveaux calculés en tenant compte respectivement de l'ensemble des samples, des samples d'événements et de textures.
+\end{itemize}
+
+Afin d'évaluer l'impact spécifique des différentes sources sonores sur l'agrément perçu, nous soumettons nos travaux aux 5 tests/études de significativité présentés ci-après:
+
+\begin{itemize}
+\item \emph{Vérification de l'agrément des scènes simulées}: afin de vérifier que la distinction affective imposée entre les i- et ni-scènes se retrouve au niveau de l'agrément perçu, nous observons s'il existe des différences entre les deux types de scènes au niveau de $\mathcal{A}_{scene}$ et $\mathcal{A}_{sujet}$.
+La significativité est évaluée par un test de Student à deux populations indépendantes pour $\mathcal{A}_{scene}$, et par un test de Student à deux populations appariées pour $\mathcal{A}_{sujet}$;
+\item \emph{étude comparative entre les descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels et/ou l'agrément perçu, nous évaluons cette significativité à partir d'un test de Student à deux populations;
+\item \emph{étude de l'influence des descripteurs structurels sur l'agrément perçu}: afin d'évaluer l'impact potentiel des descripteurs structurels sur l'agrément perçu, nous étudions l'existence de corrélations linéaires entre ces deux types de descripteurs.
+Pour mesurer la corrélation, nous utilisons le coefficient de Pearson.
+Nous adoptons ici une méthodologie couramment utilisée dans l'approche dimensionnelle;
+\item \emph{étude comparative entre les descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée a eu un impact sur la composition des scènes en terme de sources sonores, ou, pour être plus précis, s'il existe des classes de sons qui ont été particulièrement utilisées pour simuler un type d'environnement, nous utilisons le V-test.
+Nous vérifions si la présence d'une classe de sons est typique d'un environnement (i ou ni).
+Le test est effectué pour chaque niveau d'abstraction, et séparément, pour les classes d'événements et de textures.
+Pour chaque classe $j$ et chaque type d'environnement $k$ ($k=\lbrace i,ni\rbrace$), la valeur $V_{jk}$ du V-test se calcule comme suit:
+
+\begin{equation*}
+V_{jk}=\dfrac{c_{jk}-c_k\frac{c_j}{c}}{\sqrt{c_k\frac{c-c_k}{c-1}\frac{c_j}{c}(1-\frac{c_j}{c})}}
+\end{equation*}
+
+où $c$ est le nombre de classes utilisées, $c_k$ le nombre de classes utilisées pour un type d'environnement $k$, $c_j$ le nombre de classes $j$ utilisées, et $c_{jk}$ le nombre de classes $j$ utilisées pour un type d'environnement $k$.
+Le V-test teste l'hypothèse nulle que la proportion $\frac{c_{jk}}{c}$ ne diffère pas significativement de la proportion $\frac{c_{jk}}{c_k}$.
+Si pour un environnement $k$, et une classe $j$, l'hypothèse est rejetée, la classe $j$ est alors typique de l'environnement $k$.
+Les classes typiques sont nommées \textbf{marqueurs sonores};
+
+\item \emph{étude des espaces de représentation induits par les descripteurs sémantiques}: afin d'étudier si une représentation basée uniquement sur la présence ou l'absence de classes de sons permet de séparer les deux types d'environnement, nous considérons l'espace induit par les descripteurs sémantiques $S$.
+$S$ étant un vecteur booléen, nous calculons les distances entre les scènes à partir de la distance de Hamming.
+Considérant les deux vecteurs $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, et $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ de dimension $n$, avec $x=\lbrace 0,1\rbrace$, la distance de Hamming $d_{ham}$ mesure le pourcentage de coordonnées qui diffèrent entre les deux vecteurs:
+
+\begin{equation*}
+d_{ham}(S_1,S_2)=\dfrac{1}{n}\sum_{i=1}^{n} (x_{1,i} \bigoplus x_{2,i})
+\end{equation*}
+
+où $\bigoplus$ désigne l'opérateur du \emph{ou-exclusif}.
+Plus la composition des deux scènes est similaire, et plus ces deux scènes sont proches.
+L'utilisation de la distance de Hamming permet de prendre en compte de manière égale les classes présentes et absentes.
+Pour mesurer la capacité intrinsèque de l'espace à séparer les i- et ni-scènes, nous utilisons une métrique de \emph{clustering} nommée précision au rang $k$ ($P@k$).
+La $P@k$ mesure la précision obtenue après que $k$ items ont été retrouvés.
+Formellement, pour chaque scène $s_i$, nous calculons le rapport entre le nombre de scènes $s_j$ prises parmi les $k$ plus proches voisines de $s_i$, et partageant le même label que $s_i$, sur le nombre d'items à retrouver ($k$).
+La $P@k$ est alors la moyenne des rapports pour tous les items;
+
+\item \emph{étude de l'influence spécifique des marqueurs sonores sur l'agrément perçu}: afin d'évaluer les contributions spécifiques de certaines sources sonores, nous évaluons une nouvelle fois l'impact potentiel des descripteurs structurels sur l'agrément perçu, mais en ne tenant compte, cette fois, que des marqueurs sonores pour calculer ces descripteurs.
+\end{itemize}
+
+Excepté le V-test, tous les tests de significativité sont effectués avec un seuil critique $\alpha=0.05$.
+Pour le V-test, étant donné que nous testons beaucoup de classes, une correction de Bonferroni est appliquée.
+Pour les valeurs $p$, dans le cas où $p\geq0.05$, nous indiquons sa valeur.
+Dans le cas où $0.01\leq p<0.05$, nous indiquons seulement $p<0.05$.
+Dans le dernier cas nous indiquons $p<0.01$.
+
+\subsection{Résultats}
+
+\subsubsection*{Vérification de l'agrément des scènes simulées}
+
+Nous analysons ici l'agrément perçu des $80$ scènes sonores simulées. La Figure~\ref{fig:xp2_Aa} affiche l'agrément moyen $\mathcal{A}_{scene}$ pour les i- et ni-scènes.
+
+Dans un premier temps, et afin de garantir la cohérence de nos données, nous voulons nous assurer qu'aucune ni-scène n'ait un $\mathcal{A}_{scene}$ supérieur à celui d'une i-scène.
+Quatre des scènes ne respectent pas la contrainte.
+Elles et leurs correspondantes i ou ni sont retirées.
+36 i-scènes et 36 ni-scènes restent donc dans le champ de l'analyse.
+
+Dans un deuxième temps, nous voulons tester si les sujets ont bien perçu une différence d'agrément entre les i- et ni-scènes.
+Pour ce faire, nous observons l'agrément moyen de chaque sujet $\mathcal{A}_{sujet}$, calculé séparément, pour chaque type d'environnement (\cf~Figure~\ref{fig:xp2_Ab}).
+Il apparaît que les i-scènes ont bien été perçues comme significativement plus agréables ($p<0.01$) que les ni-scènes.
+
+\subsubsection*{Analyse des descripteurs structurels}
+
+En premier lieu, nous nous concentrons sur le niveau sonore.
+Les figures~\ref{fig:soundlevela},~\ref{fig:soundlevelb} et~\ref{fig:soundlevelc} affichent les distributions des niveaux $L$, $L(E)$ et $L(T)$.
+Il existe bien une différence de niveau significative entre les i- et ni-scènes ($L$: $p<0.01$), avec un écart des moyennes de -7 $dB$.
+Cette différence affecte aussi bien les événements ($L(E)$: $p<0.01$, écart moyen: -7 $dB$) que les textures ($L(T)$: $p<0.01$, écart moyen: -6 $dB$).
+
+Nous vérifions, sans surprise, que le niveau des sources sonores est bien un indicateur d'agrément, les ni-scènes ayant tendance à être plus fortes, fait rapporté dans un grand nombre d'études.
+Nous constatons encore que cette différence de niveau s'observe de manière égale pour les événements et les textures sonores.
+
+Il apparaît que ce sont les événements qui impactent le plus le niveau global des scènes, l'écart entre $L$ et $L(E)$ n'étant que de 1 $dB$ pour les i-scènes et les ni-scènes.
+Cette observation fait écho aux résultats obtenus par Kuwano~\al~\cite{kuwano_memory_2003}.
+Au cours de leur expérience, les auteurs demandent à leurs sujets d'évaluer une série d'environnements sonores de manière globale, dans un premier temps, puis d'en évaluer le niveau aux instants où chacun identifie une source sonore.
+L'étude montre qu'il n'y a pas de différences significatives entre les jugements globaux et les moyennes des jugements instantanés.
+Pour en revenir à notre expérience, c'est comme si nos sujets avaient inconsciemment tenu compte de cette réalité perceptive lors de la simulation, en faisant porter le niveau sonore global par des sons courts et bien identifiés, \ie~les événements.
+
+Nous observons, enfin, que le niveau seul ne permet pas de clairement faire la distinction entre les différents types d'environnement.
+En effet, $20\%$ des i-scènes ont un niveau supérieur au niveau minimal des ni-scènes, alors qu'il n'y a pas de recouvrement, si l'on considère l'agrément perçu $\mathcal{A}_{scene}$.
+
+\begin{figure}[t]
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_1}\label{fig:soundlevela}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_3}\label{fig:soundlevelb}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_5}\label{fig:soundlevelc}}\par
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_2}\label{fig:soundleveld}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_4}\label{fig:soundlevele}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_6}\label{fig:soundlevelf}}
+       \caption{Dispersions des descripteurs structurels $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), en fonction du type de scènes (a, b, c) et de l'agrément perçu $\mathcal{A}_{scene}$ de l'expérience 1.b (d, e, f).}
 \end{figure}
 
-The event density ($D$) is slightly higher for ni-scenes than for i-scenes (Figure~\ref{fig:xp1_deSoLv_1}.a).
-The averaged $D$ are $1.2$ for the i-scenes, and $1.6$ for the ni-scenes, with a significant effect of the scene type ($p<0.05$).
-$D$ varies widely from subject to subject.
+\subsubsection*{Influence des descripteurs structurels sur l'agrément perçu}
+\label{sec:ch5_corrDesStruct}
 
-Concerning the sound level (Figure~\ref{fig:xp1_deSoLv_1}.b), $L$ is significantly higher for ni-scenes than for i-scenes ($p<0.01$).
+Dans cette partie, nous analysons les relations fines qui peuvent exister entre les descripteurs structurels, d'une part, et l'agrément perçu, d'autre part.
+Contrairement à la section précédente, où la qualité affective des scènes est représentée de manière binaire (i \vs~ni), nous considérons ici l'agrément moyen $\mathcal{A}_{scene}$ comme descripteur perceptif.
+Il s'agit d'étudier l'existence de potentielles corrélations entre les descripteurs structurels et $\mathcal{A}_{scene}$.
+Les coefficients de corrélations linéaires calculés entre $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ sont présentés dans le tableau~\ref{tab:corrStructA}.
+Les relations entre $\mathcal{A}_{scene}$ et les descripteurs structurels sont illustrées par les figures~\ref{fig:soundleveld},~\ref{fig:soundlevele} et ~\ref{fig:soundlevelf}.
 
-To study the ability of those descriptors to characterize the appraisal, we compute the correlations between the ratings, $D$ and $L$, considering separately the i- and ni-scenes, as wel as all the scenes altogether.
-Results are shown in Table~\ref{tab:numGlobal} and in Figures~\ref{fig:xp1_deSoLv_1}.c and~\ref{fig:xp1_deSoLv_1}.d; \nm{pour rendre compte de cet effet, ne peut-on pas egalement calculer effectuer une MANOVA qui renseignera sur l'interaction de 1er ordre entre la variable dependante (ratings) et les variables independantes / facteurs expe.
-(density et Leq) ?} \gl{NM comments: je ne pense pas qu'une MANOVA s'utilise comme cela, comme toute analyse de variance, la variable independante est categorielle.
-De plus dans une MANOVA, le M de multivarié s'applique aux variales dépendantes, la proposition relève de la two way anova (mais le rating étant continu cela ne marche pas).
-Il faudrait faire une regression multiple avec terme d'interaction P=A+B+A*B, et interpréter A*B...}
+Concernant $L$, on observe une forte corrélation négative ($r=-0.77$, $p<0.01$) avec $\mathcal{A}_{scene}$, indiquant que plus le niveau sonore est élevé, plus la scène est désagréable.
+Cependant, la figure~\ref{fig:soundleveld} suggère que cette relation ne s'opère pas de la même manière pour les i- et ni-scènes.
+En effet, la corrélation entre $L$ et $\mathcal{A}_{scene}$, pour les ni-scènes, reste élevée ($r=-0.78$, $p<0.01$), mais est inexistante pour les i-scènes.
 
-Not surprisingly, these results show that ni-scenes are on average louder than i-scenes.
-The strong correlation of $-76\%$ between $L$ and the ratings indicates that the sound level does influence the pleasantness.
-However, the correlations computed separately for the i- and ni-scenes suggest $L$ does not characterize the notion of pleasantness in a equal manner for the two type of scenes:  there is still a good correlation between $L$ and the ratings for the ni-scenes, but there is no correlation if we take only the i-scenes.
-This is in line with the trend observed by \cite{lavandier2006contribution} that sound level may be a sufficient indicator of comfort for negatively rated acoustics scenes, but not for positively rated ones.
+Cette corrélation élevée, considérant l'ensemble des scènes, résulte du fait que les i-scènes ont tendance à être moins fortes que les ni-scènes, donnant ainsi l'illusion de prolonger la corrélation négative observée pour les ni-scènes.
 
-\gl{Considering the event density, ni-scenes seem to be slightly more populated that i-scenes.
-However, the weak correlation between $D$ and the ratings ($-0.35$) shows that a global measure of $D$ fails to characterize the appraisal if all the scenes are considered altogether.
-If we consider the two types of scenes separately, the correlation remains moderate for the ni-scenes, and is non-existent for the i-scenes.
-This may suggest that either the number of events is an irrelevant indicator, or that only the densities of specific event classes are involved.
-Still, although there is no clear correlation between $D$ and pleasantness for the ni-scenes, $D$ seems to play an indirect role as it is correlated with $L$ ($0.62$, $p<0.01$).
-In other words, $D$ does have an impact on the appraisal only when the addition of new events increases the sound level.}
+Nous en concluons que $L$:
 
-Averaged diversities are shown on Figure~\ref{fig:analyseDensitySnrGlobal3}.
-Except for semantic level 0, the diversity is significantly higher for the event classes of the ni-scenes (levels 1, 2 and 3: $p<0.01$;), revealing that ni-scenes are composed of a larger variety of sound events than i-scenes.
-Diversity remains constant across semantic levels for texture classes of i- and ni-scenes, with no statistical differences.
-\nm{qu'en est-il pour les classes event ?
-Est-ce que les differences observees sur fig.
-7 (a partir de semantic level 1) sont statist.
-siginificatives ?} \gl{cf.
-phrase précédente} The correlations computed between the event diversities and the averaged pleasantness ratings remain weak or moderate ($r<0.45$ \nm{r = 0.45 ?} \gl{NM comments: non, toutes les correlations sont inf à .45}, $p<0.01$), irrespective of the considered semantic level.
-If we look at the two types of scene separately, no correlation is observed.
-\gl{Although the diversity varies between the i- and ni-scenes, this last result shows that it is not a good indicator to precisely characterize the pleasantness.}
+\begin{itemize}
+\item permet bien de faire la distinction entre les i- et ni-scènes,
+\item permet de finement caractériser l'agrément perçu des ni-scènes,
+\item n'est pas un indicateur pertinent de l'agrément perçu pour des environnements a priori agréables.
+\end{itemize}
 
-\subsection{Class selection}
-\label{sec:classSelection}
+Les mêmes observations sont faites concernant $L(E)$ (\cf~\ref{fig:soundlevele}).
+Pour $L(T)$ (\cf~\ref{fig:soundlevelf}), bien que, à considérer l'ensemble des scènes, on observe une corrélation modérée, cela n'est pas vérifié quand on regarde séparément les i-scènes ($r=-0.33$, $p=0.05$) et les ni-scènes ($r=-0.00$, $p=0.99$).
+Là encore on peut penser que la corrélation négative observée pour l'ensemble des scènes est un artefact résultant du fait que le niveau des textures des i-scènes a tendance à être plus bas que celui des ni-scènes.
+Ainsi, si les événements sonores conservent une certaine capacité de prédiction de l'agrément pour les ni-scènes, le niveau des textures n'apporte, lui, que peu d'informations, quel que soit l'environnement.
 
-We investigate the sound class selection by counting the number of times a class has been used to simulate the i- or ni-scenes.
-Figures~\ref{fig:xp1_class_1}.a and~\ref{fig:xp1_class_1}.b  shows the class distributions for the event and texture classes, respectively.
-For ease of reading, we choose an intermediate semantic level, denoted $0^{+}$, laying between semantic levels 0 and 1 (see Figure~\ref{fig:taxonomie}).
+En résumé, en présence d'un environnement désagréable, les niveaux sonores, en particulier ceux des événements, ont un impact négatif sur l'agrément.
+En présence d'un environnement agréable, en revanche, aucun des descripteurs structurels considérés ici ne semble influer sur la perception de l'agrément.
 
-There is a considerable difference between the sound class selections of the i- and ni-scenes, showing that semantics play an important role in soundscape qualitative evaluation.
-The sound classes shown in Figures~\ref{fig:xp1_class_1}.a and~\ref{fig:xp1_class_1}.b are very close to those found by Guastavino \cite{guastavino_ideal_2006}, \textit{i.e.} classes that suggest human presence and nature are the most present in i-scenes while classes referring to mechanical and construction work sounds are used for ni-scenes.
+Ces premiers résultats pourraient montrer qu'il existe deux modes de perception, mobilisant chacun des descripteurs indépendants, modes qui s'activent en fonction de la nature de l'environnement (i ou ni).
 
-\gl{However, some differences have to be reported: \cite{guastavino_ideal_2006} reports that ``\,public transports\,'' are typical sounds of an ideal urban environment.
-This observation is attributed to the fact that urban environment perception is driven by social considerations.
-Public transport sounds are thus better accepted than those of private vehicles, as they are usually positively connotated.
-Results shown in Figure~\ref{fig:xp1_class_1}.c tend to qualify this claim  by showing that sounds of public transports (\textit{bus} and \textit{train}) were chosen by $28\%$ of the subjects for the i-scenes and by $42\%$ of the subjects for the ni-scenes.
-It is important to note that in \cite{guastavino_ideal_2006} subjects are asked to recall an environment in memory, that is, without any sound support.
-Tacking the reality of the sound into account, results show that public transport sounds cannot be considered as peculiar to an ideal urban soundscape.
-The class \textit{public\_transport} does appear in ni-scenes, more than \textit{car} or \textit{truck}.
-Similar findings showing that bus presence may have a negative effect on soundscape appraisal are reported in \cite{lavandier2006contribution}.
-One should note that the data still does not contradict the idea that public transport sounds are well accepted: \textit{bus} sounds were chosen by $25\%$ of the subjects for the i-scenes, circa as much as \textit{bike} sounds and more than any other vehicle sounds.
-}
-
-\begin{figure*}[t!]
-\begin{center}
-\includegraphics[scale=1]{../gfxMatlab/xp1_class_1.eps}
-    \caption{\label{fig:xp1_class_1} (a) Percentage of subjects using (a) event sound classes of the semantic level $0^{+}$, (b) texture sound classes of the semantic level $0^{+}$ and (c) event subclasses of the $0^{+}$ classes \emph{car} and \emph{public transport}, to simulate the i-scenes (left-green) and the ni-scenes (right-red).}
-\end{center}
-\end{figure*}
-
-%\begin{figure}[t!]
-%\begin{center}
-%\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_class_2.eps}
-%    \caption{\label{fig:xp1_class_2} Percentage of subjects using  to simulate the i-scenes (left-green) and the ni-scenes (right-red).}
-%    \end{center}
-%\end{figure}
-
-\subsection{Markers}
-\label{sec:markers}
-
-As semantics is found to play an important role in the qualitative perception of the soundscape, we now investigate the existence of sound markers \nm{est-ce l'accepation de Schafer ?
-d'apres ce qui est ecrit juste apres je ne pense pas; par contre, d'apres ce qui est ecrit un peu plus bas (en citant justement Schafer) ca peut l'etre ....
-Il y a ici une ambiguite sur le terme qu'il serait bon de clarifier.} \gl{Non ce n'est pas  l'interprétation de Schaffer, on expliquait plus en détail avant les termes, mais c'est trop long, et non necessaire pour la compréhension du papier}, \textit{i.e.} sound classes of great importance when considering the pleasantness of a soundscape.
-Here, an event class is considered as a marker if it has been mostly used in one type of soundscape (i or ni).
-To identify markers, a two-tailed V-test statistical test is performed for each semantic level.
-The V-test tests the null hypothesis that the $n_k$ event classes used for the scene type $k$ are randomly drawn from all the $n$ event classes available.
-A rejection of that hypothesis provides evidences that an event class has been significantly used for one type of scene only.
-Since the test concerns many event classes, Bonferroni correction is applied.
-
-Results are shown on Table~\ref{tab:markers}.
-Nine markers are found across all semantic levels.
-As suggested by Figures~\ref{fig:xp1_class_1} and~\ref{fig:xp1_class_2}, sounds evoking human activity (\textit{male footsteps concrete}) and nature (\textit{animal}, \textit{birds} and \textit{birds singing}) are markers of the i-scene.
-The presence of \textit{church bell} and \textit{bell bike} as strong markers of the i-scenes can be due to the socio-cultural background of the subjects, in great majority European citizens.
-It confirms Schafer's claim that sounds that are recognized by a community as integral part of its sonic environment are popular \cite{schafer1977tuning}.
-
-Markers of the ni-scenes are sounds due to construction work (\textit{construction work}), and sounds suggesting dense traffic (\textit{horn}, \textit{siren}).
-Although these markers are rather intuitive, none of the event classes related to motorized vehicles is a marker.
-It seems that subjects chose to illustrate unpleasant traffic sounds using horns and sirens.
+Le fait qu'aucun des descripteurs globaux ne permette de caractériser l'agrément des i-scènes peut nous amener à penser que toutes les sources sonores ne contribuent pas de manière égale à la perception de l'agrément, mais que seules les caractéristiques de certaines d'entre elles ont une réelle influence.
+Afin d'approfondir ce point, nous analysons, dans la section suivante, les scènes d'un point de vue sémantique, \ie~en nous intéressant à la nature des sources qui les composent.
+\\
 
 \begin{table}[t]
- \setlength{\tabcolsep}{0.1pt}
+\centering
+\begin{tabular}{l c c c} 
+               & i-scènes                   & ni-scènes    \\
+\hline
+$L$            & -0.32 ($p=0.06$)           & \textbf{-0.78} ($p<0.01$)\\
+$L(E)$         & -0.20 ($p=0.24$)           & \textbf{-0.75} ($p<0.01$)\\
+$L(T)$         & -0.33 ($p=0.05$)           &  -0.00 ($p=0.99$) \\
+\hline
+\end{tabular}
+\vspace{0.5mm}
+\caption{Coefficients de corrélation linéaire calculés entre l'agrément perçu moyen $\mathcal{A}_{scene}$ de l'expérience 1.b et les descripteurs structurels.}
+\label{tab:corrStructA}
+\end{table}
+
+\subsubsection*{Étude comparative entre les descripteurs sémantiques}
+
+Nous analysons la composition des scènes en comptant le nombre de sujets ayant utilisé une classe de sons pour simuler un type d'environnement.
+Les résultats sont présentés à la figure~\ref{fig:soundsourcea} pour les événements, et à la figure~\ref{fig:soundsourceb} pour les textures.
+Par souci d'espace, nous choisissons un niveau d'abstraction intermédiaire entre les niveaux 0 et 1, noté $0+$, pour représenter les classes (\cf~Figure~\ref{fig:taxonomie}).
+
+Nous observons une différence notable dans le choix des classes entre les i- et ni-scènes.
+La répartition des classes est très proche de celle obtenue dans une étude similaire sur les environnements sonores urbains idéaux \cite{guastavino2006ideal}, \ie~les classes suggérant la présence humaine et la nature sont très présentes dans les i-scènes, a contrario, les classes désignant des sons mécaniques et/ou de travaux sont principalement utilisées pour les ni-scènes.
+
+Ces résultats confirment un fait déjà observé: la nature sémantique des sources sonores joue un rôle prédominant dans l'appréciation de l'environnement \cite{raimbault2005urban,dubois2006cognitive}.
+
+Nous notons quelques différences avec \cite{guastavino2006ideal}: les résultats obtenus par Guastavino montrent que les sons de \emph{transports publics} sont caractéristiques des environnements sonores urbains idéaux.
+Les auteurs attribuent cela au fait que la perception de l'agrément est, entre autre, soumise à un contexte socio-culturel.
+Dans notre représentation du monde, les sons de transports publics sont positivement connotés, et ont ainsi tendance à être mieux acceptés que les sons de véhicules privés.
+
+Dans une certaine mesure, nos résultats contredisent ce fait.
+La figure~\ref{fig:soundsourcea} montre, en effet, que les classes d'événements de \emph{transports publics} (\emph{bus} et \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) ont été utilisées par les sujets, pour des i-scènes, dans $28\%$ des cas, et pour des ni-scènes, dans $42\%$ des cas.
+Les résultats ne remettent pas en question le fait que les sons de \emph{transports publics} soient bien acceptés: $25\%$ des sujets ont utilisé la classe \emph{bus} pour les i-scènes, un chiffre comparable à celui de la classe \emph{Vélo}, et bien supérieur à celui de toute autre classe de véhicules privés.
+Cependant les classes \emph{transports publics} sont également bien présentes dans les ni-scènes, plus que les classes \emph{voiture} ou \emph{camion} par exemple.
+La classe \emph{transports publics} ne peut donc pas être considérée comme typique d'un environnement sonore urbain idéal.
+
+Cette différence peut s'expliquer par la nature des deux protocoles expérimentaux utilisés.
+Comme nous l'avons fait, Guastavino demande à ses sujets de décrire un environnement.
+Mais ces derniers travaillent de mémoire, alors que nos propres sujets disposent de supports sonores.
+Le fait que nos sujets soient confrontés à la réalité acoustique des sons, pour recréer leurs environnements, peut avoir pour effet de diminuer l'impact du contexte socio-culturel.
+D'autres études utilisant des sons comme stimuli montrent que la classe \emph{bus} peut avoir un effet négatif sur l'appréciation de l'environnement \cite{lavandier2006contribution}.
+
+\begin{figure}[t]
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=1\linewidth]{gfx/ch_5/xp1_class_1_cam}\label{fig:soundsourcea}} \par    
+        \subfloat[]
+        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_2_cam}\label{fig:soundsourceb}} 
+        \subfloat[]
+        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_3_cam}\label{fig:soundsourcec}} 
+       \caption[Pourcentage de scènes simulées comportant une classe de son particulière.]{Pourcentage de scènes simulées (i ou ni) comportant une classe de son particulière: (a) classes d'événements du niveau d'abstraction $0+$, (b) classes de textures du niveau d'abstraction $0$, (c) sous classes d'événements du niveau d'abstraction $1$ appartenant aux classes \emph{trafic} et \emph{transport public} du niveau d'abstraction $0$.}\label{fig:soundsource}
+\end{figure}
+
+\subsubsection*{Marqueurs sonores}
+\label{sec:ch5_marqueurXp1}
+
+Nous avons mis en évidence que, qualitativement, la composition des sources sonores des scènes diffère selon les types d'environnement (i ou ni).
+Nous essayons de voir maintenant si, parmi ces classes, certaines sont typiques d'un environnement en particulier.
+Pour ce faire, nous utilisons le V-test (\cf~Section~\ref{sec:ch5_methodoEtStat1}), en considérant séparément chaque niveau d'abstraction.
+Les résultats sont présentés dans le tableau~\ref{tab:markers}.
+
+Concernant les événements sonores, 9 marqueurs sont identifiés sur l'ensemble des niveaux d'abstraction.
+Comme la figure~\ref{fig:soundsource} le laissait présager, les classes relatives à la présence humaine (\emph{pas homme béton}, \emph{sonnette vélo}), et à la nature (\emph{animaux, oiseaux}, \emph{chants d'oiseaux}) sont des marqueurs de i-scènes.
+Nous notons également la présence de la classe \emph{cloche} dans les marqueurs d'un environnement idéal.
+Ce fait est possiblement dû au \emph{background} socio-culturel des sujets, dans leur grande majorité, des citoyens européens.
+En effet, selon Schafer, un son reconnu par un individu comme faisant partie intégrante de son environnement est bien accepté.
+Les marqueurs de ni-scènes sont des classes faisant référence à des sons de travaux (\emph{travaux}), ou suggérant un trafic dense (\emph{klaxon}, \emph{sirène}).
+
+Concernant les textures sonores, 5 marqueurs sont identifiés.
+Pour les i-scènes, il s'agit de classes faisant référence à des ambiances amorphes, calmes, (\emph{cour-intérieur/parc} et \emph{parc}).
+Pour les ni-scènes, il s'agit, comme pour les événements, de classes faisant référence à des bruits de travaux (\emph{travaux} et \emph{véhicule de travaux}), ainsi que d'une classe faisant référence au trafic (\emph{carrefour}).
+
+Bien que l'ensemble des marqueurs identifiés soient intuitifs, aucune des classes d'événements faisant directement référence aux bruits de véhicules motorisés n'est un marqueur, exception faite de la classe de textures \emph{carrefour}.
+Pour représenter un trafic désagréable, les sujets ont porté leurs choix sur les classes \emph{klaxon} et \emph{sirène}.
+On peut supposer que les sons isolés de véhicules sont compris comme faisant partie intégrante de l'environnement urbain, et ne sont donc pas particulièrement associés à un environnement désagréable.
+
+\begin{table}[t]
+ \setlength{\tabcolsep}{0.2pt}
  \centering
@@
 \begin{tabular}{c c c c} 
-Semantic   &  \multicolumn{2}{c}{Markers} \\
-level & i-scenes & ni-scenes \\
+Niveau        & \multicolumn{2}{c}{Marqueurs sonores événements} \\
+d'abstraction & i-scènes & ni-scènes \\
 \hline
-0  &              &  construction work  \\
+0  &                          & travaux (3.78)  \\
 \hline
-1  & church bell  & horn  \\
-   & bell bike    & siren \\
-   & animal       &       \\
+  & cloche  (4.5)             & klaxon  (3.9) \\
+1 & sonnette vélo  (4.3)      & sirène (3.9)\\
+  & animal (4.2)              &       \\
    \hline
-2  & birds         & horn  \\
-   & church bell   & siren \\
-   & bell bike     &       \\
+  & oiseau        (4.8)       & klaxon  (4.0)\\
+2 & cloche  (4.4)             & sirène (4.0)\\
+  & sonnette vélo    (4.2)             &       \\
    \hline
-3  & birds singing & horn  \\
-   & church bell   & siren \\
-   & bell bike     &       \\
-   & male footsteps concrete &  \\
+  & chant oiseau (4.8)        & klaxon  (4.1)\\
+3 & cloche   (4.3)            & sirène (4.0)\\
+  & sonnette vélo     (4.2)   &       \\
+  & pas chaussure  (3.6)      &  \\
+  &                           & \\ 
+  & \multicolumn{2}{c}{Marqueurs sonores textures}      \\
+  & i-scènes & ni-scènes \\
+\hline
+0 &     cour int./parc (4.1) &  travaux (3.9)  \\
+\hline
+1 &     parc (3.65)          &  carrefour (3.6)  \\
+  &                          &  travaux véhicule (3.3)  \\
+\hline
+2 &     parc (3.64)          &  carrefour (3.56)  \\
 \hline
@@
 \vspace{0.5mm}
-\caption{\label{tab:markers} Event classes found to be markers. In each cell, markers are ordered from top to bottom by decreasing V-test values.}
+\caption[Classes d'événements et de textures identifiées comme étant des marqueurs sonores.]{Classes d'événements et de textures identifiées comme étant des marqueurs sonores.
+Dans chaque cellule, les marqueurs sont ordonnés par ordre décroissant de valeur $V$.}
+\label{tab:markers}
 \end{table}
 
-\subsection{Quantifying class selection}
-\label{sec:QuantifyingClassSelection}
+\subsubsection*{Étude des espaces de représentation induits par les descripteurs sémantiques}
+
+Dans cette partie, nous évaluons la capacité d'une représentation sémantique à séparer les deux types d'environnement.
+Pour ce faire, nous calculons une précision au rang 5 ($p@5$) sur l'espace induit par les descripteurs sémantiques $S$, et ce pour chaque niveau d'abstraction (\cf~Section~\ref{sec:ch5_methodoEtStat1}).
+Les vecteurs $S$ sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.
+Nous ne considérons pas les classes de marqueurs de textures, ces dernières étant trop peu nombreuses.
+Pour les mêmes raisons nous ne considérons pas les classes de marqueurs d'événements du niveau d'abstraction $0$.
+Les résultats sont affichés sur la figure~\ref{fig:pa5}.
+
+En ce qui concerne $ET$, la $p@5$ est de $76\%$ pour le niveau d'abstraction 0, et reste supérieure à $86\%$ à partir du niveau d'abstraction 1.
+Ces résultats confirment qu'il est possible de clairement distinguer les deux types d'environnement en se basant seulement sur la présence ou l'absence des classes de sons.
+Nous notons également que, plus le niveau d'abstraction est élevé, plus la capacité de séparer les environnements est importante.
+En d'autres termes, plus nous sommes précis dans notre description de la composition des scènes, plus nous sommes à même d'établir une distinction claire entre les i- et ni-scènes.
+
+En considérant séparément $E$ et $T$, il apparaît 1) que la $p@5$ obtenue avec $E$ est similaire à celle obtenue avec $ET$, 2) que la $p@5$ obtenue avec $T$ est systématiquement inférieure d'environ $10$ à $15\%$ à celle de $E$.
+Ces résultats indiquent que l'information sémantique permettant de séparer les deux environnements est principalement portée par les événements.
+Ces résultats font, par ailleurs, écho aux travaux de  \cite{maffiolo_caracterisation_1999}, qui montrent que nous analysons de manière descriptive (en identifiant les sources) les scènes événementielles, \ie~composées d'événements sonores (\cf~Section~\ref{sec:ch3_catsoundscape}).
+
+Enfin, il apparaît que la $p@5$ obtenue avec $E_{m}$ est égale, voire supérieure à celles obtenues avec $E$ et $ET$, et ce bien qu'une information partielle soit utilisée dans ce cas pour décrire les scènes.
+La dimension des vecteurs de description $S$ pour $E_m$ est en effet inférieure à la dimension des vecteurs $S$ pour $E$, qui est elle même inférieure à celle obtenue dans le cas où toutes les classes sont utilisées ($ET$).
+De plus, dans le cas où les marqueurs ne sont pas pris en compte pour la description ($E_{w/o,m}$), les résultats chutent, passant même en dessous de ceux obtenus en ne considérant que les textures.
+Cela confirme que la majorité de l'information sémantique permettant de faire la distinction entre i-scènes et ni-scènes est incluse dans les marqueurs.
+
+En résumé, nous déduisons de cette analyse les points suivants:
+
+\begin{enumerate}
+\item contrairement à ce que nous avions constaté avec les descripteurs structurels, une description sémantique de la composition des scènes, en terme de présence/absence de sources sonores, permet de bien distinguer les deux types d'environnement (i ou ni);
+\item l'information sémantique est majoritairement portée par les classes d'événements sonores;
+\item parmi les classes d'événements, seule une partie, \ie~les marqueurs sonores, est nécessaire pour faire la distinction entre les i- et ni-scènes.
+\end{enumerate}
+
+Maintenant que nous avons isolé les classes typiques des i- et ni-scènes, et vérifié que la distinction entre ces environnements dépendait de la présence de ces classes, il reste à voir si une description structurelle des scènes, basée uniquement sur ces marqueurs sonores, permet de caractériser l'agrément perçu, mieux qu'une description structurelle globale.
 
 \begin{figure}[t]
-\begin{center}
-\includegraphics[width=.4\paperwidth]{../gfxMatlab/pa5_1.eps}
-\caption{\label{fig:pa5} Precision at rank 5 ($P@5$) computed from the \textit{Hamming} distances between the scenes at different semantic levels, considering the event and texture classes ($ET$), the event classes ($E$), the texture classes ($T$), the event markers ($E_m$) and event classes without event markers ($E_{w/o m}$) .}
-\end{center}
+        \myfloatalign
+        \includegraphics[width=\linewidth]{gfx/ch_5/pa5_1}
+       \caption[$P@5$ obtenues en considérant la matrice de dissimilarité résultant des distances par paires de Hamming calculées entre les vecteurs des descripteurs sémantiques des scènes.]{$P@5$ obtenues en considérant la matrice de dissimilarité résultant des distances par paires de Hamming calculées entre les vecteurs des descripteurs sémantiques des scènes.
+Les vecteurs sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.}\label{fig:pa5}
 \end{figure}
 
-To quantify how class selection characterizes the pleasantness of a urban environment, we now follow a class-based classification of the scenes evaluated with the standard precision at rank 5 ($P@5$) metric.
-
-Each simulated scene $s_i$ is represented by a boolean vector of dimension $n$: $S_i = (x_1, x_2, \ldots{}, x_{n})$, $i\in[1,72]$.
-Each dimension corresponds to a sound class (event and texture) of a particular semantic level (for example: $n=44$ classes for semantic level 1).
-Thus $x_1 = 1$ indicates that the sound class $x_1$ is present in scene $s_i$ (not present if $x_1=0$).
-A \textit{Hamming} distance is computed between all pairs of $S_i$.
-Then, for each vector $S_i$ the number of scenes sharing the same label (i or ni) as $S_i$  among the five scenes $s_j$ ($j=1,...,5$) that have the smallest distance to it is computed.
-The $P@5$ precision is then obtained by averaging the results over all the scenes $s_i$ ($i=1, \ldots{}, 72$).
-
-Results considering both the event and texture classes ($ET$) \nm{symbolisme pas clair} \gl{pk ?} to build the vectors  $S_i$ are shown on Figure~\ref{fig:pa5}.
-The $P@5$ is of 81\% for semantic level 0 and remains above 89\% starting at semantic level 1.
-These result strongly indicates that it is possible to make a clear distinction between the i- and ni-scenes based solely on the class presence/absence.
-We also note that the deeper the hierarchical level, the higher the $P@5$; in other words an increased precision in identifying scene elements allows a better distinction between i- and ni-scenes.
-
-To refine the analysis, only event classes ($E$), texture classes ($T$), event markers ($E_m$) and event classes without event markers ($E_{w/o m}$) are also considered on Figure~\ref{fig:pa5}.
-The constant deviation of circa $10\%$ from the semantic level 1 between $E$ and $T$, and the fact that $P@5$ are similar for $ET$ and $E$ indicate that soundscape quality is mostly conveyed by sound events.
-
-Since similar precision are achieved by $ET$, $E$ and $E_m$, we may conclude that all the event classes are not equivalently important to distinguish between the two scene types.
-Indeed, $E_m$ uses very partial information but still reaches quasi-equivalent results.
-Moreover, the drop in $P@5$ considering $E_{w/o m}$ from the semantic level 1 confirms the fact that soundscape quality mostly relies on the presence of specific sound markers.
-
-\subsection{Marker-wise pleasantness descriptors}
-\label{sec:QuantifyingMarkers}
+\subsubsection*{L'influence spécifique des marqueurs sonores sur l'agrément perçu}
 
 \begin{table}[t]
+\setlength{\tabcolsep}{3pt}
+\centering
+{\renewcommand{\arraystretch}{1}
 \centering
@@
 \hline
-$L_m$         & 0.19  ($p=0.28$)           & \textbf{-0.60} ($p<0.01$) \\
-$L_b$         & \textbf{-0.51} ($p<0.01$)  & -0.16 ($p=0.36$) \\
-$L_{m-b}$     & \textbf{0.69} ($p<0.01$)   & -0.31 ($p=0.06$) \\
-$D_{m}$       & -0.20  ($p=0.24$)          & \textbf{-0.58} ($p<0.01$)  \\
-$D_{b}$       & -0.23  ($p=0.18$)          & -0.30 ($p=0.07$)  \\
+$L_m$              & 0.03  ($p=0.88$)           & \textbf{-0.75} ($p<0.01$) \\
+$L(E)_m$           & 0.08  ($p=0.66$)           & \textbf{-0.71} ($p<0.01$) \\
+$L(T)_m$           & -0.11 ($p=0.66$)           & -0.17 ($p=0.37$) \\
+$L_b$              & \textbf{-0.52} ($p<0.01$)  & -0.32 ($p=0.06$) \\
+$L(E)_b$           & \textbf{-0.51} ($p<0.01$)  & -0.30 ($p=0.07$) \\
+$L(T)_b$           & -0.32 ($p=0.05$)           & \textbf{-0.73} ($p<0.01$) \\
+$L_m-L_b$          & \textbf{0.67} ($p<0.01$)   & -0.31 ($p=0.07$) \\
+$L(E)_m-L(E)_b$    & \textbf{0.66} ($p<0.01$)   & -0.28 ($p=0.10$) \\
+$L(T)_m-L(T)_b$    & 0.16 ($p=0.54$)            & 0.21 ($p=0.28$) \\
 \hline
 \end{tabular}
+}
 \vspace{0.5mm}
-\caption{\label{tab:numMarkers} Correlations computed between rating \emph{vs.} $L_m$, $L_b$, $L_{m-b}$ and $D_m$.}
+\caption{Coefficients de corrélation linéaire calculés entre l'agrément perçu moyen $\mathcal{A}_{scene}$ de l'expérience 1.b et les descripteurs structurels relatifs à la présence des marqueurs sonores.}
+\label{tab:corrMarkers}
 \end{table}
 
+Comme pour la section~\ref{sec:ch5_corrDesStruct}, nous évaluons les corrélations entre $\mathcal{A}_{scene}$ et les descripteurs structurels.
+Pour cette section, les descripteurs structurels sont calculés en tenant compte des marqueurs sonores précédemment identifiés.
+Nous définissons $X_m$ le descripteur $X$ calculé en ne prenant en compte que les sons des marqueurs.
+A l'inverse, nous définissons $X_b$ ($b$: pour «~bruit~») le descripteur $X$ calculé en prenant en compte toutes les classes de sons, excepté les marqueurs.
+Lorsque le descripteur caractérise une i-scène (idem pour une ni-scène), nous ne considérons, pour le calcul, que les marqueurs identifiés pour les i-scènes (ou pour les ni-scènes), que nous nommons i-marqueurs (ou ni-marqueurs).
+Les résultats sont affichés sur le tableau~\ref{tab:corrMarkers}.
+
+Concernant les niveaux sonores (\cf~Figures~\ref{fig:soundlevelMarker}), là encore les mêmes tendances sont observées entre $L_m$, $L(E)_m$ et $L(T)_m$, d'une part, et $L$, $L(E)$ et $L(T)$, d'autre part.
+Que l'on considère uniquement les marqueurs, ou l'ensemble des classes, il s'avère que :
+
+\begin{enumerate}
+\item il existe une différence significative entre les niveaux des i- et ni-scènes ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
+\item le niveau sonore des scènes est majoritairement porté par les événements sonores, comparé aux textures sonores;
+\item le niveau sonore des événements a une influence sur la perception de l'agrément pour les ni-scènes, mais pas pour les i-scènes;
+\item le niveau sonore des textures ne joue aucun rôle dans la perception de l'agrément.
+\end{enumerate}
+
+En conclusion, le niveau des ni-marqueurs a une influence négative sur l'agrément pour les ni-scènes. En revanche le niveau des i-marqueurs n’impacte pas l'agrément perçu pour les i-scènes.
+
+En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoise}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
+C'est la première fois qu'un indicateur objectif nous permet de préciser l'agrément des environnements agréables.
+Ceci nous amène à conclure que le niveau des classes de sons n'étant pas typiques d'un environnement agréable a un impact négatif sur l'agrément.
+
+Par ailleurs, alors que $L(T)$ ne présentait pas de corrélation pour les ni-scènes, une corrélation négative forte est observée pour $L(T)_b$ ($r-0.73$, $p<0.01$).
+Ce fait indique que les niveaux des classes de textures n'étant pas des marqueurs n'affectent pas l'agrément perçu de la même manière pour les i- et ni-scènes.
+Les niveaux semblent avoir un effet négatif pour les ni-scènes, alors que pour les i-scènes, aucun effet n'est relevé.
+
+Pour finir, nous considérons un dernier groupe de descripteurs, nommément $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiff}).
+Ces descripteurs expriment la différence entre les niveaux des marqueurs, et ceux des autres classes de sons.
+Ils traduisent l'émergence des marqueurs par rapport à la mixture sonore.
+
+Pour les i-scènes, une corrélation modérée et positive est observée pour $L_m-L_b$ ($r=0.67$, $p<0.01$) et $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
+Pour les ni-scènes, aucune corrélation n'est observée.
+Dans le cas des i-scènes, ce n'est donc pas le niveau absolu des marqueurs qui importe, mais leur niveau relatif, par rapport aux autres sons qui composent la scène.
+On observe donc pour les environnements idéaux un double mécanisme perceptif:
+
+\begin{itemize}
+\item plus le niveau absolu des sons n'étant pas des i-marqueurs est élevé, plus l'agrément est faible,
+\item plus le niveau relatif des i-marqueurs, par rapport aux autres sons, est élevé, plus l'agrément est élevé.
+\end{itemize}
+
+Pour les ni-scènes, le fait que nous observions des corrélations pour $L_m$ et $L(E)_m$, et aucune pour $L_m-L_b$ et $L(E)_m-L(E)_b$, montre que c'est bien le niveau absolu qui importe.
+
 \begin{figure}[t]
-\begin{center}
-\includegraphics[width=.4\paperwidth]{../gfxMatlab/xp1_deSoLvMar_1.eps}
-  \caption{\label{fig:numMarkers} Averaged pleasantness vs. (a)  $LA_{(eq,V)\_m-b}$, (b) $LA_{(eq,V)\_m}$, (c) $LA_{(eq,V)\_b}$ and (d) $Density_m$.}
-  \end{center}
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_7}\label{fig:soundlevelMarkera}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_9}\label{fig:soundlevelNoisea}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_19}\label{fig:soundlevelMarkerDiffa}}\par
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_8}\label{fig:soundlevelMarkerd}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_10}\label{fig:soundlevelNoised}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}      
+        \caption{Dispersions des descripteurs structurels de niveaux sonores relatifs à la présence des marqueurs $L_m$ (a, d), $L(E)_m$ (b, e) et $L(T)_m$ (c, f), en fonction du type de scènes (a, b, c) et de l'agrément perçu $\mathcal{A}_{scene}$ de l'expérience 1.b (d, e, f).}\label{fig:soundlevelMarker}
 \end{figure}
 
-In order to check if the physical characteristics of the markers alone have an influence on the pleasantness, we now consider the densities and the $LA_{(eq,V)}$ with on one hand the tracks of the markers ($LA_{(eq,V)\_m}$, $Density_{m}$) and on the other hand the remaining sounds ($LA_{(eq,V)\_b}$, $Density_{b}$).
-The marker durations is rather short, and the number of markers in a scene is rather low, a window length of 125 ms is used to compute $LA_{(eq,V)\_m}$ and $LA_{(eq,V)\_b}$.
-For the i-scenes we consider only the markers of the i-scenes (i-markers), and for the ni-scenes only the markers of the ni-scenes (ni-markers).
+\subsection{Discussions}
 
-The difference $LA_{(eq,V)\_m-b}=LA_{(eq,V)\_m}-LA_{(eq,V)\_b}$ is also considered as it illustrates how the markers emerge from the other sounds.
-Correlations between those descriptors and the average perceptual ratings are depicted in Table~\ref{tab:numMarkers} and Figure~\ref{fig:numMarkers}.
+Dans cette expérience, nous notons que les descripteurs structurels ($L$, $L(E)$ et $L(T)$) permettent systématiquement de faire la distinction entre les environnements sonores idéaux et non-idéaux.
 
-Considering $Density_{m}$ (Figure~\ref{fig:numMarkers}.d), results are almost the same than to those found in section~\ref{sec:levelDensityDiversity}.
-There is no correlation for the i-scenes, but the correlation increases for the ni-scenes.
-No correlation is found for $Density_{b}$.
-Sound levels descriptors are much more informative.
-Little to no information is lost by considering only the markers to compute $LA_{(eq,V)\_m}$, \emph{i.e.} there is still a good correlation for the ni-scenes (although inferior to the one observed in Section~\ref{sec:levelDensityDiversity}), and no correlation for the i-scenes.
+Parmi ces indicateurs structurels, seuls $L$ et $LE$ permettent de prédire l'agrément. Nous notons cependant que cette prédiction ne vaut que pour les ni-scènes.
 
-Considering the ni-scenes, the good correlation for $LA_{(eq,V)\_m}$  (Figure~\ref{fig:numMarkers}.b) and the absence of correlation for $LA_{(eq,V)\_b}$  (Figure~\ref{fig:numMarkers}.c) suggests that the sound levels of the ni-markers convey most of the annoyance.
-Also, no correlation is observed for $LA_{(eq,V)\_m-b}$ (Figure~\ref{fig:numMarkers}.a).
-This indicates that the predominant influence of the marker sound levels remains true irrespective of the sound level of the other sounds.
-A loud horn sound is expected to increase the annoyance level of a scene (more so than a soft klaxon sound), whether or not the remaining of the sonic environment is ``\,noisy\,''.
+Nous observons qu'une description sémantique des scènes, basée sur la présence/absence des classes de sons, permet de bien prédire la nature de l'environnement.
+Par ailleurs, il apparaît qu'il est possible d'obtenir une prédiction similaire, voire meilleure, en ne considérant qu'un sous groupe de classes d'événements,\ie~les marqueurs sonores.
 
-That said, this last statement is not valid for the i-scenes.
-The strong uphill correlation found between pleasantness and  $LA_{(eq,V)\_m-b}$ (Figure~\ref{fig:numMarkers}.a) shows that the higher the markers sound levels relative to the other sounds, the higher the pleasantness.
-But the moderate downhill correlation observed for $LA_{(eq,V)\_b}$ (Figure~\ref{fig:numMarkers}.c) shows that the levels of those remaining sounds may be a cause of annoyance if perceived as too loud.
-It is thus the emergence of the i-markers from the others sounds that drive the appraisal.
-Considering for example the sonic environment of a park and following those conclusions, the global appraisal will increase with the sound level of the birds, provided other sounds like human voices or traffic hubbub are not too loud.
+Parmi les descripteurs structurels spécifiques, calculés en tenant compte des marqueurs sonores, plusieurs permettent maintenant de faire la distinction entre les i-scènes et les ni-scènes: $L_m$, $L_{b}$, $L(E)_{m}$, $L(E)_{b}$, $L(T)_{m}$, $L(T)_{b}$, $L_{m}-L_{b}$ et $L(E)_{m}-L(E)_{b}$.
 
-These results are strong indicators that global sound level is not a good indicator to characterize the appraisal of \emph{a priori} positive sound environments such as a park or a quiet street.
-On the contrary, the contribution of each sound source needs to be considered separately, as for certain type of sounds, the pleasantness does increase with their sound levels.
-
-Those observations may also serve as guidelines for urban planner, showing that:
+Parmi ces descripteurs, 7 semblent impacter l'agrément perçu:
 
 \begin{itemize}
-\item in the case of negative environment such as street with public work or busy traffic, where it is often difficult to diminish the global sound level, focusing only on the noise regulation of specific sound sources (jack-hammer or horn) may significantly reduce the overall annoyance;
-\item in the case of positive environment such as park or quiet street, the appraisal may be enhanced by increasing the presence of specific sound sources such as birds singing.
+\item $L_b$ et $L(E)_b$ ont un impact négatif sur les i-scènes;
+\item $L(E)_m-L(E)_b$ et $L_m-L_b$ ont un impact positif sur les i-scènes;
+\item $L_m$, $L(E)_m$ et $L(T)_b$  ont un impact négatif les ni-scènes.
+\end{itemize}
+
+De cette analyse, nous retenons les points suivants:
+
+\begin{itemize}
+\item \emph{distinguer les i- et ni-scènes}: les descripteurs sémantiques, ainsi que certains descripteurs structurels globaux, permettent de faire la distinction entre les i-scènes et les ni-scènes.
+La description sémantique semble être plus performante;
+\item \emph{événements ou textures}: que ce soit pour les descripteurs sémantiques ou structurels, ce sont majoritairement les événements qui permettent de distinguer les deux types d'environnement, les textures n'apportant, au mieux, qu'une information limitée;
+\item \emph{prédire l'agrément}: si l'on considère une description fine de l'agrément, il semble que la manière de percevoir la qualité de l'environnement diffère en fonction de la nature de ce dernier (i ou ni).
+Il n'apparaît pas envisageable de considérer un même jeu de descripteurs pour prédire, à la fois, l'agrément des i-scènes, et l'agrément des ni-scènes.
+Pour les ni-scènes, ce sont le niveau global ($L$ et $L(E)$), ou le niveau des marqueurs sonores ($L_m$ et $L$), qui impactent négativement l'agrément.
+On note ici que prendre en compte les contributions de différentes sources n'améliore pas la capacité de prédiction de l'agrément, par rapport à une analyse holistique de l'environnement.
+Pour les i-scènes, par contre, prédire l'agrément requiert d'étudier, de manière séparée, les caractéristiques des marqueurs sonores, et celles de l'ensemble des autres sons.
+Ainsi, le niveau des marqueurs relatifs au bruit est positivement corrélé à l'agrément, alors que le niveau du bruit est, lui, négativement corrélé.
+\end{itemize}
+
+Le fait que l'agrément des scènes agréables ne se corrèle pas avec des descripteurs physiques holistiques, contrairement à l'agrément des scènes désagréables, a récemment été déjà observé \cite{gozalo2015relationship}.
+
+L'existence de deux modes de perception, mobilisant différents types de descripteurs, et dépendant de la nature (dans notre cas hédonique) des stimuli, est un phénomène qui se rapproche de celui observé pour la perception des textures (\cf~Section~\ref{sec:ch3_eventTexture}).
+Le cerveau adapte sa manière de traiter l'information (résumé statistique pour les textures, description fine pour les événements) suite à une prise de décision antérieure quant à la nature du stimulus (à savoir «~est-ce un événement ou une texture ?~»).
+De la même manière, les indicateurs actifs dans le jugement de l'agrément dépendent, eux aussi, d'une identification préalable de la nature hédonique globale de l'environnement (idéale ou non idéale).
+
+Ces résultats peuvent potentiellement influer sur les stratégies à adopter pour améliorer la qualité de l’environnement sonore:
+
+\begin{itemize}
+\item dans le cadre de scènes non-idéales, il s'agit de diminuer le niveau sonore, soit de manière globale, soit en agissant sur certaines sources (\emph{sirène}, \emph{klaxon});
+\item dans le cadre de scènes idéales, il s'agit 1) d'identifier les sons agréables, \ie~les marqueurs sonores, 2) de baisser le niveau des autres sons, 3) voire, en restant dans la limite du raisonnable, d'augmenter le niveau des marqueurs par rapport aux autres sons.
+\end{itemize}
+
+Nous montrons que les descripteurs à utiliser dépendent de la nature de l'environnement, et que cette nature est elle même dépendante de la composition sémantique, \ie~des sources sonores présentes.
+Dans une certaine mesure, nous pouvons donc dire que les descripteurs dépendent des sources sonores présentes.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%% XP 2 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Expérience 2 : modification de  la composition sémantique}
+\label{sec:xp3}
+
+\subsection{Objectif}
+
+L'expérience précédente a montré que, parmi les classes de sons peuplant le monde sonore, celles regroupant les marqueurs sont caractéristiques de certains types d'environnement.
+Ces marqueurs sonores semblent avoir un impact particulier sur la perception de leurs environnements.
+C'est ce dernier point qui est étudié dans cette expérience.
+
+Afin de vérifier que l'agrément des scènes idéales et non-idéales dépend de la présence des marqueurs, les scènes sonores précédemment simulées sont régénérées, sans les classes de marqueurs.
+Pour les i-scènes, les i-marqueurs sont retirés.
+Pour les ni-scènes, les ni-marqueurs.
+Une épreuve d'évaluation de l'agrément, dont le protocole se rapproche de celui de l'expérience 1.b, est alors conduite.
+
+L'objectif est de vérifier si l'absence des marqueurs a un impact sur l'agrément perçu. Deux hypothèses sont formulées:
+
+\begin{itemize}
+\item \emph{pour les ni-scènes} nous faisons l'hypothèse que l'absence des ni-marqueurs va \textbf{augmenter} la valeur de l'agrément perçu;
+\item \emph{pour les i-scènes} nous faisons l'hypothèse que l'absence des i-marqueurs va \textbf{diminuer} la valeur de l'agrément perçu.
+\end{itemize}
+
+Si la première hypothèse est intuitive, la deuxième l'est moins.
+En effet, il n’apparaît pas évident que la suppression des i-marqueurs, bien que s'agissant de sons positivement connotés, diminue la qualité globale d'un environnement.
+Cette suppression aura, de surcroît, pour effet de diminuer le niveau sonore global de la scène.
+
+Néanmoins, comme nous l'avons vu, le niveau global n'est qu'un indicateur partiel de l'agrément pour les environnements sonores idéaux.
+Qui plus est, cet indicateur, lorsque qu'il décrit le niveau des i-marqueurs, impacte de manière positive la qualité de la scène.
+L'hypothèse mérite donc d'être vérifiée.
+
+\subsection{Planification expérimentale}
+
+\subsubsection*{Stimuli}
+
+La banque de données de stimuli compte 144 séquences de 30 secondes. Ces 144 séquences comprennent:
+
+\begin{itemize}
+\item \emph{72 am-scènes}: les 72 scènes précédemment simulées, incluant les classes de marqueurs (am).
+Nous notons i/am-scènes, les 36 scènes idéales avec marqueurs, et ni/am-scènes les 36 scènes non-idéales avec marqueurs;
+\item \emph{72 sm-scènes}: les 72 scènes précédemment simulées, régénérées sans les classes de marqueurs (sm).
+Nous notons i/sm-scènes, les 36 scènes idéales sans marqueurs, et ni/sm-scènes les 36 scènes non-idéales sans marqueurs.
+\end{itemize}
+
+Nonobstant l'absence des marqueurs, les am- et sm-scènes sont en tout point semblables.
+
+Nombre de am-scènes sont composées, en majorité, de samples de marqueurs.
+Afin de pas abusivement dénaturer ces scènes, en créant notamment des temps de «~vide~», \ie~ne comprenant aucun sample, nous ne supprimons que les marqueurs des classes d'événements du premier niveau d'abstraction (\cf~Tableau~\ref{tab:markers}).
+Ces classes sont:
+
+\begin{itemize}
+\item \emph{cloche}, \emph{sonnette de vélo}, \emph{animaux} pour les i/sm-scènes;
+\item \emph{sirène}, \emph{klaxon} pour les ni/sm-scènes.
+\end{itemize}
+
+Il est important de noter ici que tous les i- et ni-marqueurs ne sont donc pas supprimés dans les sm-scènes.
+
+\subsubsection*{Procédure}
+
+Les sujets évaluent les 144 scènes.
+L'évaluation s'effectue sur une échelle sémantique bipolaire de 11 points allant de -5 (non-idéale/très désagréable) à +5 (idéale/très agréable).
+Avant de noter une scène, les sujets doivent obligatoirement en écouter les 20 premières secondes.
+Après la notation, ils sont libres de passer à la scène suivante.
+
+Pour chaque sujet, les scènes sont présentées dans un ordre aléatoire.
+Les 10 premières scènes permettent au sujet de calibrer ses notes.
+Elles sont obligatoirement composées de 5 i/am-scènes et de 5 ni/am-scènes.
+Ces 10 premières scènes sont rejouées à la fin de l'expérience, et seules les notes données à la deuxième occurrence sont prises en compte.
+
+L'expérience est prévue pour durer 1 heure. Les sujets ne connaissent pas la nature des scènes.
+
+\subsubsection*{Dispositif expérimental}
+
+Tous les sujets passent l'expérience sur des machines identiques.
+L'audio est diffusé en monophonie, par le biais de casques audio semi-ouverts \emph{Beyer-Dynamic DT 990 Pro}.
+Toutes les scènes sonores ont été re-simulées sur la base des partitions obtenues lors de l'expérience de simulation.
+Le niveau sonore de sortie est identique pour tous les sujets.
+
+Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
+
+Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de cette dernière, et de répondre aux éventuelles questions des sujets.
+
+\subsubsection*{Participants}
+
+12 sujets (4 femmes) participent à l'expérience.
+Aucun d'entre eux n'a réalisé l'expérience de simulation, ni la première expérience d'évaluation.
+Les sujets sont âgés de 22 à 61 ans (moyenne: 29.5, écart-type: 14).
+Tous les sujets vivent dans un milieu urbain.
+
+Tous les sujets ont réalisé l'expérience avec succès.
+
+\subsection{Données et outils statistiques}
+
+Les données analysées sont les mêmes que pour la première expérience. Nous invitons le lecteur à se référer à la section~\ref{sec:ch5_dataType1} pour plus de détails.
+
+Il s'agit ici de vérifier que la suppression des i- et ni-marqueurs impacte l'agrément perçu.
+Pour ce faire, nous utilisons l'analyse de variance.
+Nous considérons, comme variable dépendante,  $\mathcal{A}_{sujet}$, et, comme variables indépendantes, le type d'environnement (i/ni), et la présence/absence de marqueurs (am/sm).
+Chaque sujet devant évaluer la totalité des stimuli, une ANOVA à mesures répétées à deux facteurs est utilisée afin vérifier s'il existe des différences significatives d'agrément perçu.
+Les deux variables indépendantes sont considérées comme des facteurs intra-sujet (\emph{within-subject}.
+Les facteurs n'étant composés que de deux niveaux chacun (type: i/ni; marqueur: am/sm), l'hypothèse de sphéricité n'a pas besoin d'être vérifiée.
+Les analyses \emph{post hoc} sont conduites en appliquant la procédure de Tukey-Kramer.
+
+Tous les tests de significativité sont effectués avec un seuil critique $\alpha=0.05$.
+
+\subsection{Résultats}
+
+\subsubsection{Détection de valeurs extrêmes}
+
+\begin{figure}[t]
+        \myfloatalign
+        \subfloat[sujet 1]
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_1}\label{fig:xp4_note_1a}}
+        \subfloat[sujet 2]
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_2}\label{fig:xp4_note_1b}}
+        \subfloat[sujet 3]
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_3}\label{fig:xp4_note_1c}}
+        \subfloat[sujet 4]
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_4}\label{fig:xp4_note_1d}} \par 
+        \subfloat[sujet 5]
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_5}\label{fig:xp4_note_1e}}
+        \subfloat[sujet 6]
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_6}\label{fig:xp4_note_1f}}     
+        \subfloat[sujet 7]
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_7}\label{fig:xp4_note_1g}}
+        \subfloat[sujet 8]
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_8}\label{fig:xp4_note_1h}} \par 
+        \subfloat[sujet 9]
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_9}\label{fig:xp4_note_1i}}
+        \subfloat[sujet 10]
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_10}\label{fig:xp4_note_1j}}
+        \subfloat[sujet 11]
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_11}\label{fig:xp4_note_1h}}
+        \subfloat[sujet 12]
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_12}\label{fig:xp4_note_1l}}       
+        \caption{Dispersions des notes données par les sujets lors de l'expérience 2 aux i/am-scènes (vert) et ni/am-scènes (rouge).}\label{fig:xp4_note_1}
+\end{figure}
+
+Considérons $\mathcal{A}_{sujet}$ pour les am-scènes (\cf~Figure~\ref{fig:xp4_note_2}).
+Il apparaît que les réponses du sujet 7 diffèrent des autres.
+Comme observé sur la figure~\ref{fig:xp4_note_1} (\cf~Figure~\ref{fig:xp4_note_1g} pour le sujet 7), ce dernier a évalué positivement près de la moitié des ni/am-scènes.
+Le sujet 7 a donné à 58\% des ni/am-scènes une note supérieure à 0, contre une moyenne de 11\% pour les autres sujets.
+De plus, le sujet 7 a utilisé l'ambitus maximal (-5 à 5) pour noter à la fois les i/ et ni/am-scènes.
+Ces faits n'ayant pas été observés pour les autres sujets, que l'on considère les expériences 2 ou 1.b, le sujet 7 est éliminé de l'analyse.
+
+\subsubsection{Influence de la présence des marqueurs sur l'agrément perçu}
+\label{sec:ch5_Asujet}
+
+Dans cette section nous étudions comment les sujets ont perçu les différents types de scènes, nommément: i/am-, ni/am-, i/sm- et ni/sm-scène.
+L'ANOVA à mesures répétées pratiquée sur $\mathcal{A}_{sujet}$ (\cf~Figure~\ref{fig:xp4_note_2}) montre un effet significatif du type d'environnement (i/ni: $F[1,10]=175$, $p<0.01$), de la présence/absence des marqueurs (am/sm: $F[1,10]=7$, $p<0.05$), ainsi que de l'interaction entre les deux facteurs ($F[1,10]=67$, $p<0.01$).
+
+L'analyse \emph{post hoc} montre, quant à elle, des différences significatives entre tous les groupes d'observations, notamment entre les i/am- et i/sm-scenes ($p<0.05$) et les ni/am- et ni/sm-scenes ($p<0.01$).
+
+Ces résultats indiquent que la suppression des événements a effectivement modifié la perception des scènes par les sujets. Nos deux hypothèses sont ainsi vérifiées:
+
+\begin{itemize}
+\item la suppression des ni-marqueurs a amélioré les qualités perçues des ni-scènes;
+\item la suppression des i-marqueurs a diminué les qualités perçues des i-scènes.
+\end{itemize}
+
+L'interaction significative montre que l'effet du type d'environnement influe sur l'effet de l'absence/présence des marqueurs.
+En effet la moyenne des écarts entre am- et sm-scènes est plus importante pour les ni-scènes (1.1) que pour les i-scènes (0.5).
+
+\subsection{Discussions}
+
+De cette expérience nous retenons principalement deux points.
+
+Premièrement, il apparaît que la présence, dans une scène, des marqueurs relevés à l'expérience 1 impacte bien l'agrément perçu.
+La suppression des ni-marqueurs a un effet bénéfique sur le ressenti, tandis que, plus surprenant, la suppression des i-marqueurs dégrade légèrement la qualité.
+Ce dernier point est d'autant plus marquant que, du fait de la suppression des marqueurs, le niveau des i/am-scènes est supérieur à celui des i/sm-scènes.
+
+Les i-marqueurs ont donc bien un effet bénéfique sur la perception d'un environnement.
+Le fait que leur suppression diminue $\mathcal{A}_{scene}$ montre clairement qu'il est possible d'améliorer la qualité sonore d'un lieu en ajoutant des sons bien acceptés comme \emph{oiseau}.
+Ces conclusions vont dans le sens de l'approche positive introduite par Schafer \cite{schafer1977tuning} (\cf~Section~\ref{sec:ch3_urbanNoiseSoundscape}).
+
+Deuxièmement, il apparaît que la suppression des marqueurs modifie les liens existant entre niveaux sonores et agrément perçu.
+Pour les i-scènes, ce lien est renforcé, tandis que pour les ni-scènes, il est dégradé.
+Ces résultats vont dans le sens de ceux obtenus lors de l'expérience 1, à savoir que l'influence d'un descripteur sur la perception d'une scène est conditionnée par la qualité hédonique de cette dernière.
+En l'occurrence, nous observons que plus l'agrément d'une scène est élevé, moins les descripteurs structurels influent sur la perception.
+
+Enfin, si l'on considère les représentations mentales, ces observations nous incitent à conjecturer les points suivants:
+
+\begin{itemize}
+\item  il existe un lien entre le concept concret relatif à une source sonore (\emph{oiseau}), et le concept abstrait relatif à une qualité sonore (\emph{agrément});
+\item  il existe un lien entre l'information modale que nous extrayons d'un environnement, et le concept abstrait relatif à sa qualité sonore (\emph{agrément}).
 \end{itemize}
@@
 
-We introduce in this paper a novel experimental protocol to study the perception of soundscapes, where subjects are asked to recreate an auditory environment using a tool specifically developed for that purpose and a structured database of isolated sounds.
-This protocol allows us to generate soundscapes whose composition, in terms of nature, timing and intensity of sound sources, is known with perfect accuracy.
-Using the data thus produced in a series of analyses on the issue of ideal \emph{vs} non-ideal urban soundscape classification and characterization, we show that:
-
-\begin{itemize}
-\item The data obtained with the simulation allows us to reach results that are consistent with previous perceptual studies relying on recorded scenes \cite{lavandier2006contribution} or the experimental subjects' memory \cite{guastavino_ideal_2006}.
-\item From the global descriptors sound level, class diversity and event density, only sound level seems to play a role in pleasantness evaluation.
-\item The ability of sound level to characterize pleasantness relies upon the quality of the scene itself. For positively rated scenes (i-scenes), it is not relevant.
-\item A semantic description of the soundscapes in terms of presence~/ absence of sound sources is an effective criterion to distinguish between ideal and non-ideal urban environments.
-\item A few event markers, appearing at various semantic levels, suffice to perform efficient discrimination between ideal and non-ideal urban environments, although the type of markers found in this study may depend on the socio-cultural background of the subjects.
-\item The sound level computed on the markers only is a better predictor of perceived agreement than the global sound level.
-\end{itemize}
-
-Compared to more traditional survey based methods, the proposed method provides an increased precision and reliability of soundscape description useful for the analysis of subjects' judgments.
-This article was intentionally focused on the analysis of the scene composition; further research shall be dedicated to 1) a thorough study of the influence of the physical characteristics of the sound markers on pleasantness evaluation, as well as categorisation process, and 2) the elaboration of a predictive method based on that analysis, allowing to anticipate the agreement of an acoustic environment from its composition.
-
-In addition, more understanding shall be gained about the actual impact of sound level on soundscape perception: is there a significant difference in evaluation between two scenes having the exact same composition, but different levels?
-Is loudness but a consequence of scene composition ?
-
 \noindent \textbf{Acknowledgements}\\
@@
 
-%----------------------------------------------------------------------------------------
-%	REFERENCE LIST
-%----------------------------------------------------------------------------------------
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%    Bibliography    %%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\References{}{../biblio}{actalit}
+\References{}{Bibliography}{actalit}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%    APPENDICES   %%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\onecolumn
+
+\appendix
+\section{Taxonomies des classes de sons}
+
+\begin{figure}[h]
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_1}\label{fig:taxonomieEventa}}
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_2}\label{fig:taxonomieEventb}}\par
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_1}\label{fig:taxonomieTexturea}}
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_2}\label{fig:taxonomieTextureb}}
+       \caption{Taxonomies des classes de sons utilisées pour la simulation des environnements sonores urbains : (a,b) événements sonores et (c,d) textures sonores.}
+       \label{fig:taxonomie}
+\end{figure}
 

### 2017-03-06 gfxHierarchy/getJsonFromTxt.m

@@ -1,77 +1,0 @@
-clear all
-
-addpath(genpath('~/Dropbox/projets/urbanSoundScapeXP2014/analyse/functions/'));
-hierarchyPath='~/Dropbox/projets/urbanSoundScapeXP2014/analyse/hierarchy/';
-
-json.startup;
-
-type={'E','T'};
-
-for uu=1:length(type)
-    %% get Lvl 0
-    [tags0,~,classes0,~] = getHierarchy( type{uu},0,hierarchyPath);
-    
-    %% get Lvl 1
-    [tags1,~,classes1,classes0p] = getHierarchy( type{uu},1,hierarchyPath);
-    
-    classes0p(cellfun(@(x) isempty(x),classes0p))=classes1(cellfun(@(x) isempty(x),classes0p));
-    
-    classes0=cellfun(@(x) strrep(x,'_',' '),classes0,'UniformOutput',false);
-    classes0p=cellfun(@(x) strrep(x,'_',' '),classes0p,'UniformOutput',false);
-    classes1=cellfun(@(x) strrep(x,'_',' '),classes1,'UniformOutput',false);
-    %% get data
-    switch type{uu}
-        case 'E'
-            data.name='event';
-        case 'T'
-            data.name='texture';
-    end
-    
-    data.children=[];
-    s=1;
-    
-    
-    for jj=1:length(classes0)
-        data.children{jj}.name=classes0{jj};
-        data.children{jj}.children=[];
-        
-        ind1=find(cellfun(@(x) ~isempty(strfind(x,tags0{jj})),tags1));
-        
-        switch type{uu}
-            case 'E'
-                
-                c1Tmp=classes1(ind1);
-                c0pTmp=classes0p(ind1);
-                c0pLabelTmp=unique(c0pTmp);
-                
-                for ii=1:length(c0pLabelTmp)
-                    
-                    ind2=find(strcmp(c0pLabelTmp{ii},c0pTmp));
-                    
-                    data.children{jj}.children{ii}.children={};
-                    
-                    if length(ind2)==1
-                        data.children{jj}.children{ii}.children{1}.name=c0pLabelTmp{ii};
-                        data.children{jj}.children{ii}.children{1}.size=s;
-                        %             data.children{jj}.children{ii}.size=s;
-                    else
-                        data.children{jj}.children{ii}.name=c0pLabelTmp{ii};
-                        for gg=1:length(ind2)
-                            data.children{jj}.children{ii}.children{gg}.name=c1Tmp{ind2(gg)};
[…]

### 2017-03-06 gfxHierarchy/json/data_E.js

@@ -1,1 +1,0 @@
-{"name":"event","children":[{"name":"motorized transport","children":[{"children":[{"name":"air transport","size":1}]},{"children":[{"name":"boats","size":1}]},{"children":[{"name":"garbage truck","size":1}]},{"name":"public transportation","children":[{"name":"train","size":1},{"name":"bus","size":1}]},{"name":"traffic","children":[{"name":"motorcycle","size":1},{"name":"car","size":1},{"name":"truck","size":1},{"name":"scooter","size":1},{"name":"start passage way","size":1}]}]},{"name":"works","children":[{"children":[{"name":"hand tools","size":1}]},{"name":"public work","children":[{"name":"power tools","size":1},{"name":"vehicle work","size":1},{"name":"jackhammer","size":1},{"name":"metal work","size":1}]}]},{"name":"alarm","children":[{"children":[{"name":"bike","size":1}]},{"children":[{"name":"car fire alarm","size":1}]},{"children":[{"name":"church bell","size":1}]},{"children":[{"name":"horn","size":1}]},{"children":[{"name":"siren","size":1}]},{"children":[{"name":"whistle","size":1}]}]},{"name":"voice","children":[{"name":"voice","children":[{"name":"child voice","size":1},{"name":"female voice","size":1},{"name":"male voice","size":1}]}]},{"name":"human movement","children":[{"children":[{"name":"bike","size":1}]},{"children":[{"name":"footstep","size":1}]},{"children":[{"name":"stroller pushchair","size":1}]}]},{"name":"game","children":[{"children":[{"name":"game","size":1}]}]},{"name":"vehicle door","children":[{"children":[{"name":"vehicle door","size":1}]}]},{"name":"nature","children":[{"children":[{"name":"animal","size":1}]},{"children":[{"name":"thunderstorm","size":1}]}]}]}

### 2017-03-06 gfxHierarchy/json/data_T.js

@@ -1,1 +1,0 @@
-{"name":"texture","children":[{"name":"traffic","children":[{"name":"crossroads","size":1},{"name":"square","size":1},{"name":"street","size":1}]},{"name":"work","children":[{"name":"work","size":1},{"name":"vehicle work","size":1},{"name":"ventilation","size":1}]},{"name":"courtyard park","children":[{"name":"park","size":1},{"name":"courtyard","size":1}]},{"name":"nature","children":[{"name":"water","size":1},{"name":"wind","size":1}]},{"name":"human","children":[{"name":"cafe","size":1},{"name":"crowd","size":1},{"name":"market","size":1}]}]}

### 2017-03-07 lafaySimulatedScenePerception.tex
end vf v1

@@
 \newcommand{\gl}[1]{\textcolor{blue}{GL : #1}}
-\newcommand{\nm}[1]{\textcolor{red}{nm : #1}}
-\newcommand{\ml}[1]{\textcolor{magenta}{ML : #1}}
-\newcommand{\mr}[1]{\textcolor{green}{MR : #1}}
+\newcommand{\jlv}[1]{\textcolor{magenta}{#1}}
+\newcommand{\jls}[1]{\textcolor{red}{#1}}
+\newcommand{\jlc}[1]{\textcolor{green}{#1}}
 
@@
 
-\gl{[TODO ...]} \\
-
-Comme nous l'avons vu, la recherche sur les paysages sonores a besoin d'outils permettant d'analyser séparément les influences des différentes sources sur les qualités affectives de l'environnement.
-La simulation offre des possibilités intéressantes (\cf~Section~\ref{sec:ch4_modAnaSo}), car elle nous permet d'obtenir des scènes sonores dont nous connaissons tous les paramètres structuraux, en particulier les caractéristiques distinctes des différentes sources.
-
-Pour démontrer l'intérêt de la simulation en analyse sensorielle, nous choisissons, comme cadre applicatif, le problème de l'agrément perçu dans les environnements sonores urbains.
-
-Ce Papier présente les résultats d'une série d'expériences s'appuyant sur la simulation, et visant, chacune, à comprendre comment les différentes sources sonores qui composent une scène influent sur la perception de l'agrément:
+Un des objectifs premiers des études sur les paysages sonores est d'identifier les informations contenues dans l'environnement qui influent sur les sensations perçues \cite{aletta2016soundscape}.
+Questionner le lien entre sensation (\eg~calme) et environnement (\eg~scène sonore urbaine), revient à objectiver la représentation mentale que se fait un sujet d'une scène sonore en particulier (\eg~une scène sonore urbaine calme).
+
+Si on demande au sujet d'évaluer un environnement donné suivant une échelle perceptive donnée (\eg~clame, agrément) \cite{axelsson2005soundscape,davies2013perception,cain2013development}, le gain d'information de l'expérimentateur est conditionné par la nature des stimuli disponibles, que ces derniers soient des scènes sonores enregistrées (expérience en laboratoire), ou réelles (expérience \emph{in situ}).
+Cependant, le fait que ces stimuli existent lui permet d'en établir une description physique précise.
+
+Si, à l'inverse, on demande au sujet de décrire un environnement donné (\eg~décrivez une scène sonore urbaine «~calme~») \cite{guastavino2006ideal, dubois2006cognitive}, l'expérimentateur obtient bien une information riche, symbolique et sémantique, de la représentation que se fait le sujet de cet environnement.
+Mais, en l'absence de données sonores, il ne peut la caractériser physiquement.
+
+Nous pensons que la simulation permet de faire un lien élégant entre ces deux approches.
+Demander à un sujet de simuler un environnement sonore donné permet de générer une réalisation modale, \ie~définie sur des dimensions physiques (le signal de la scène simulée), de la représentation mentale dudit sujet.
+Cette démarche permet alors d'obtenir des scènes caractérisée à la fois sémantiquement (\eg~sources sonores présentes) et physiquement (\eg~niveaux des sources sonores).
+
+De plus en plus de travaux tendent à montrer que toutes les sources ne participent pas de manière égale à la perception de la scène sonore \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
+szeremeta2009analysis}.
+En conséquence, nombre de recherches se portent maintenant sur l'étude des contributions spécifiques des différentes sources sur les qualités affectives de la scène \cite{gozalo2015relationship,ricciardi2015sound}.
+Ces recherches ont besoin d'outils permettant d'analyser séparément les influences des différentes éléments sonores qui composent l'environnement.
+
+Nous pensons que ces travaux peuvent grandement bénéficier de l'utilisation de scènes simulées, scènes dont les caractéristiques structurelles, en particulier la nature des sources présentes, leurs positions ainsi que leurs signaux séparés, sont directement accessibles par l'expérimentateur.
+Ce qui n'est pas le cas pour des scènes enregistrées.
+
+Afin de répondre à ces problématiques, nous proposons \emph{Simscene}, un outil de simulation permettant à un sujet de simuler des paysages sonores à partir d'un corpus de sons enregistrés.
+Pour montrer l'apport de \emph{Simscene} sur l'étude des paysages sonores, nous choisissons, comme cadre applicatif, le problème de l'agrément perçu dans les environnements sonores urbains.
+Nous présentons les résultats d'une série d'expériences s'appuyant sur la simulation, et visant, chacune, à comprendre comment les différentes sources sonores qui composent une scène influent sur la perception de l'agrément:
 
 \begin{enumerate}
-\item \emph{expérience 1.a, simulation}: au cours de cette expérience, les sujets simulent les environnements qui serviront de stimuli pour les étapes suivantes;
+\item \emph{expérience 1.a, simulation}:  les sujets simulent les environnements qui serviront de stimuli pour les étapes suivantes;
 \item \emph{expérience 1.b, évaluation de l'agrément}: les sujets doivent évaluer l'agrément des scènes simulées à partir d'une échelle sémantique;
 \item \emph{expérience 2, évaluation de l'agrément après modification des scènes}: comme pour l'expérience précédente, les sujets doivent évaluer l'agrément des scènes simulées à partir d'une échelle sémantique.
-Cependant les scènes ont été modifiées, \ie~privées de certaines classes de sons identifiées comme ayant un impact sur l'agrément perçu;
+Cependant les scènes ont été modifiées, \ie~privées de certaines classes de sons identifiées comme ayant un impact sur l'agrément perçu.
 \end{enumerate}
+
+À notre connaissance, seuls Bruce~\al \cite{bruce2009development,bruce2014effects} se sont servis de la simulation afin d'étudier la perception des paysages sonores.
+Les auteurs proposent un système de simulation permettant au sujet d'agir sur un environnement en ajoutant ou supprimant des sources sonores spécifiques.
+Ce système permet par ailleurs de modifier le niveau sonore des sources, et leurs positions spatiales.
+
+À l'aide de cet outil, les auteurs demandent à leurs sujets de manipuler un jeu de sources sonores pré-établies, afin de recréer un environnement urbain.
+Ce faisant, ils montrent que l'inclusion ou l'exclusion des sources dépendent plus de considérations sociales/sémantiques, que de leurs caractéristiques physiques.
+Ils soulignent néanmoins que le manque d'enregistrements disponibles limite l'analyse.
+
+Dans notre démarche, nous faisons le choix d'une restitution simplifiée (monophonique) des scènes, au profit d'un nombre plus important de paramètres et de sources disponibles, ce afin de garantir, à la sortie, des données viables et expressives.
 
@@
 La deuxième introduit l'outil de simulation \emph{SimScene}.
-La troisième reprend l'expérience de simulation à proprement parler, ainsi que l'épreuve \emph{évaluation de l'agrément}.
-La quatrième décrit l'épreuve \emph{évaluation de l'agrément après modification des scènes}.
+La troisième reprend l'expérience 1.a de \emph{simulation}, ainsi que l'épreuve 1.b d'\emph{évaluation de l'agrément}.
+La quatrième décrit l'épreuve 2 d'\emph{évaluation de l'agrément après modification des scènes}.
 Nous concluons dans la dernière partie.
 
-Précisons que l'expérience de simulation a fait l'objet d'une étude pilote dont les résultats sont présentés dans  \cite{lafay2013atiam,lafay2014new}.
-L'outil de simulation \emph{Simscene} est présenté quant à lui en détails dans \cite{rossignol2015simscene,lafay2016JAES}.
-Nous résumons ici ses fonctionnalités d'importance pour notre étude.
+Par souci d'espace, nous ne détaillons pas dans ce papier l'outil de simulation, mais présentons ses fonctionnalités d'importance pour notre étude.
+Pour plus de détails sur \emph{Simscene}, nous invitons le lecteur à se référer à \cite{rossignol2015simscene,lafay2016JAES}.
+Précisons enfin que l'expérience 1.a de simulation a fait l'objet d'une étude pilote dont les résultats sont présentés dans \cite{lafay2013atiam,lafay2014new}.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%   PROTOCOL      %%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   SIMSCENE      %%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\emph{Simscene} est un environnement de travail audio-numérique dont la première version a été développée dans le cadre du projet HOULE\footnote{Projet HOULE : \url{http://houle.ircam.fr/}}.
-Il permet de simuler des paysages sonores à partir d'un corpus de sons.
-Il est prévu pour fonctionner sur les navigateurs internet \emph{Chrome} et \emph{Firefox}.
-L'outil a été développé en javascript à l'aide de la bibliothèque \emph{angular.js}\footnote{\emph{angular.js} : \url{https://angularjs.org/}} et du standard \emph{web-audio}\footnote{\emph{web-audio} : \url{http://www.w3.org/TR/webaudio/}}.
+\emph{Simscene} est un environnement de travail audio-numérique dont la première version a été développée dans le cadre du projet HOULE\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
+Il  Il est prévu pour fonctionner sur les navigateurs internet \emph{Chrome} et \emph{Firefox}.
+L'outil a été développé en javascript à l'aide de la bibliothèque \emph{angular.js}\footnote{\emph{angular.js} : \url{angularjs.org}} et du standard \emph{web-audio}\footnote{\emph{web-audio} : \url{www.w3.org/TR/webaudio}}.
 L'interface de sélection (\cf~Section~\ref{sec:simscene_ssf}) a été développée à l'aide de la bibliothèque \emph{D3.js} \cite{d32011}.
@@
 \subsection{Banque de données de samples}
+\label{sec:simscene_sampleDataSet}
 
 \subsubsection*{Typologie}
-\label{sec:simscene_Typologie}
 
@@
 Pour ce faire, une étude bibliographique est effectuée, afin d'identifier les sources et ambiances sonores les plus souvent citées dans les publications.
-Cette étude porte sur 16 articles ou thèses.
-Chacun d'eux traite de la manière dont nous discriminons les paysages sonores urbains.
-Il ressort que plusieurs approches sont possibles:
-
-\begin{itemize}
-\item 9 articles abordent le problème par une approche perceptive, soit en identifiant ou répertoriant des catégories de sources sonores, soit en étudiant l'impact de classes de sons spécifiques sur la perception de l'environnement: \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories}
-\item 3 articles proposent une classification morpho-typologique, divisant l’environnement sonore urbain en «~zones sonores~» possédant une identité acoustique forte, selon la configuration et la pratique du site: \cite{maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive}
-\item 2 articles répertorient et classifient les sources sonores d’un point de vue expert: \cite{leobon_analyse_1986,brown2011towards}
-\end{itemize}
-
-Les réactions à la musique étant trop subjectives, et les jugements esthétiques ne pouvant qu'altérer les données d'évaluation, nous choisissons, dans cette étude, de ne pas considérer les sources musicales de type musiciens de rue, radios de voitures, d'appartements, \etc.
+Cette étude porte sur 16 articles ou thèses \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards}
+. Chacun d'eux traite de la manière dont nous discriminons les paysages sonores urbains. Notre typologie est établie sur la base des catégories/classes de sons relevées dans cette littérature.
+
+Les réactions à la musique étant jugées trop subjectives, et les jugements esthétiques ne pouvant qu'altérer les données d'évaluation, nous choisissons, dans cette étude, de ne pas considérer les sources musicales de type musiciens de rue, radios de voitures, d'appartements, \etc.
 
 \subsubsection*{Événements et Textures}
-\label{sec:simscene_eventTexture}
+
+Lorsque des événements émergent d'un environnement sonore, le cerveau traite l'information des différentes sources de manière séparée.
+Plusieurs flux auditifs sont ainsi générés \ie~un pour chaque séquence d'événements émis par la même source \cite{carlyon2004brain,winkler2009modeling}.
+Inversement, quand le cerveau ne parvient pas à isoler d'événement, la scène est traitée globalement, tous ces constituants étant agglomérés dans un même flux.
+
+Dans la littérature, quatre types de sons semblent pouvoir être isolés:
+
+\begin{itemize}
+\item {événement sonore}: un son isolé, ponctuel, dont les caractéristiques physiques varient au cours du temps;
+\item {texture sonore}: un son, isolé, long, dont les caractéristiques physiques restent stables au cours du temps, et analysé à partir de statistiques extraites d'une représentation temps-fréquence;
+\item \emph{scène événementielle}: une séquence contenant une information sémantique élevée;
+\item \emph{scène amorphe}: une séquence contenant une information sémantique faible.
+\end{itemize}
+
+Une scène événementielle est une séquence composée soit uniquement d'événements, soit d'événements et de textures, les événements porteurs d'une information plus riche primant quant au choix du processus de traitement à mettre en œuvre.
+
+Les textures et les scènes amorphes, elles, sont traitées de manière holistique, à partir de propriétés acoustiques globales pour les scènes amorphes \cite{dubois2006cognitive,maffiolo_caracterisation_1999}, et sur la base d'une information résumée statistiquement pour les textures \cite{mcdermott2013summary}.
+Toutes deux portent une information limitée \cite{saint1995classification,nelken2013ear}.
+Cependant, les séquences amorphes sont spontanément décrites par les sujets comme des «~fonds sonores~» \cite{maffiolo_caracterisation_1999,guastavino2006ideal}, induisant qu'elles n'existent que grâce à un processus de construction de flux auditifs, alors que les textures sont des objets définis seulement sur la base de leur nature physique.
+Il est cependant possible d'assimiler une scène amorphe à une texture, ses caractéristiques physiques demeurant stables au cours du temps.
+De fait, nombre de scènes amorphes («~brouhaha de rue~», «~brouhaha de trafic~») sont citées comme textures.
+Cependant, l'inverse, considérer une texture comme une scène amorphe, n'est pas forcément vrai.
+Un exemple de texture souvent cité est le son du «~galop~», son qui peut très bien apparaître au premier plan de la scène.
 
 \subsubsection*{Taxonomie}
-\label{sec:simscene_getSample}
+
+\begin{figure}[t]
+        \myfloatalign
+        \includegraphics[width=.8\linewidth]{gfx/ch_4/3}
+       \caption{Organisation hiérarchique de la banque de sons isolés utilisée pour la simulation.}\label{fig:orgDb}
+\end{figure}
 
@@
 
-Les classes de sons sont organisées suivant une structure hiérarchique, afin de former une taxonomie.
+Les classes de sons sont organisées suivant une structure hiérarchique, afin de former une taxonomie (\cf~Figure~\ref{fig:orgDb}).
 La structure taxonomique s'inspire grandement de l'axe vertical de l'organisation catégorielle proposée par E.
@@
 
-Nous établissons deux taxonomies, une pour les événements (\cf~Figures~\ref{fig:taxonomieEventa} et~\ref{fig:taxonomieEventb}), une autre pour les textures (\cf~Figures~\ref{fig:taxonomieTexturea} et~\ref{fig:taxonomieTextureb}).
-
+Nous établissons deux taxonomies, une pour les événements, une autre pour les textures (\cf~Annexe~\ref{app:taxonomie}).
 Pour les événements, nous considérons quatre niveaux d'abstraction allant des classes les plus globalisantes (niveau d'abstraction 0), aux classes les plus spécifiques (niveau d'abstraction 3).
 Pour les textures, nous ne considérons que trois niveaux d'abstraction.
-
 Notons que la taxonomie des événements est assez proche de celle de \cite{Salamon14}, introduite ultérieurement à notre étude.
@@
 \subsubsection*{Acquisition des samples}
-\label{sec:simscene_getSample}
 
 Sur la base des taxonomies précédemment établies, 483 sons isolés ont été collectés, dont 381 événements, et 102 textures.
-Parmi ces samples, 332 ont été enregistrés, et 151 proviennent de différentes banque de sons (\emph{SoundIdeas}\footnote{\emph{SoundIdeas} : \url{ http://www.sound-ideas.com/}} et \emph{Universal SoundBank}\footnote{\emph{Universal SoundBank} : \url{http://www.universal-soundbank.com/}}).
-
-Tous les enregistrements ont été effectués à l’aide d'un micro canon \emph{AT8035}\footnote{Micro canon \emph{AT8035} : \url{http://eu.audio-technica.com/fr/products/product.asp?catID=1&subID=6&prodID=1845}} relié à un enregistreur \emph{ZOOM H4n}\footnote{Enregistreur \emph{ZOOM H4n} : \url{http://www.zoom.co.jp/english/products/h4n/}}.
+Parmi ces samples, 332 ont été enregistrés, et 151 proviennent de différentes banque de sons (\emph{SoundIdeas}\footnote{\emph{SoundIdeas}: \url{www.sound-ideas.com}} et \emph{Universal SoundBank}\footnote{\emph{Universal SoundBank}: \url{www.universal-soundbank.com}}).
+
+Tous les enregistrements ont été effectués à l’aide d'un micro canon \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} relié à un enregistreur \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}}.
 L’utilisation du micro canon nous permet d’isoler les événements sonores du brouhaha urbain.
@@
 
-\subsection{Interfaces}
-
-\subsubsection*{Interface de sélection}
+\subsection{Interface de sélection}
 \label{sec:simscene_ssf}
-
-\begin{figure}[t]
-        \myfloatalign
-        \subfloat[]
-        {\includegraphics[width=.8\columnwidth]{gfx/ch_4/SSF}\label{fig:ssf}} \par
-        \subfloat[]
-        {\includegraphics[width=\columnwidth]{gfx/ch_4/simscene}\label{fig:simscene}}
-       \caption{Les interfaces de sélection (a) et de simulation (b) de l'outil \emph{Simscene}.}
-\end{figure}
 
@@
 L’efficacité de ce principe repose avant tout sur la structure typologique, et la nomenclature de la base de données.
-Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique?
+Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique.
 
@@
 
-L'interface de sélection a fait l'objet d'une étude approfondie dont les résultats sont publiés dans \cite{lafay2016JAES}.
-
-\subsubsection*{Interface de simulation}
+Cette interface a fait l'objet d'une étude approfondie dont les résultats sont publiés dans \cite{lafay2016JAES}.
+
+\begin{figure}[t]
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=.8\columnwidth]{gfx/ch_4/SSF}\label{fig:ssf}} \par
+        \subfloat[]
+        {\includegraphics[width=\columnwidth]{gfx/ch_4/simscene}\label{fig:simscene}}
+       \caption{Les interfaces de sélection (a) et de simulation (b) de l'outil \emph{Simscene}.}
+\end{figure}
+
+\subsection{Interface de simulation}
 
@@
 
-L'interface de simulation est présentée en détails dans \cite{rossignol2015simscene}.
+\emph{Simscene} est présenté en détail dans \cite{rossignol2015simscene}.
 
@@
 
-\subsection{Expérience 1.a : planification}
-\label{sec:ch5_planExpSimu}
-
-\subsubsection*{Procédure}
+\subsection{Objectif}
+
+L'objectif est d'étudier les influences spécifiques des différentes sources sonores qui composent les environnements urbains, sur la perception de l'agrément, en utilisant la simulation.
+Pour ce faire, nous planifions nos deux premières expériences (\cf~Figure~\ref{fig:xp1_2}):
+
+\begin{itemize}
+\item \emph{expérience 1.a de simulation}: au cours de cette expérience, les sujets doivent simuler des environnements sonores urbains, en utilisant l'outil de simulation \emph{Simscene} (\cf~Section~\ref{sec:simscene}).
+Chacun compose deux environnements sonores, le premier idéal/agréable et le deuxième non-idéal/désagréable.
+
+\item \emph{expérience 1.b d'évaluation}: à l'issue de la simulation, nous n'avons, de fait, qu'une connaissance binaire des propriétés affectives des scènes simulées: idéale (i) et non-idéale (ni).
+Cette seconde étape a pour but d'affiner notre connaissance sur l'agrément.
+Pour ce faire, nous demandons à un deuxième groupe de sujets d'évaluer, à partir d'une échelle sémantique, l'agrément de chacune des scènes simulées.
+L'expérience d'évaluation sert deux buts:
+
+\begin{enumerate}
+\item évaluer l'influence respective des différentes sources sur l'agrément, pour chaque type d'environnement (i ou ni);
+\item détecter la présence de cas extrêmes ou ambigus (\emph{outlier}) dans les scènes simulées.
+Pour le reste de notre étude, les qualités hédoniques imposées (i et ni) servent de référence, de vérité terrain.
+Il nous faut donc garantir qu'il n'y ait pas d’ambiguïté entre les cas extrêmes des i- et ni-scènes, \ie~que la note d'agrément la plus basse des i-scènes reste supérieure à la note la plus haute des ni-scènes.
+\end{enumerate}
+
+\end{itemize}
+
+Notre analyse s'appuie sur les données produites par les deux expériences.
 
@@
         \myfloatalign
-        \includegraphics[width=1\linewidth]{gfx/ch_4/4}
-       \caption{Etapes du processus de simulation}\label{fig:etapeSimu}
+        \includegraphics[width=\linewidth]{gfx/ch_5/5}
+        \caption{Planification expérimentale des éxperiences de simulation et d'évaluation de l'agrément}\label{fig:xp1_2}
 \end{figure}
+
+\subsection{Planification  de l'expérience 1.a}
+\label{sec:xp1a_plan}
+
+\subsubsection*{Procédure}
 
@@
 
-Chaque processus de simulation comprend deux parties (\cf~Figure~\ref{fig:etapeSimu}) :
+Chaque processus de simulation comprend deux parties :
 
@@
 Trois expérimentateurs, un dans chaque pièce, sont présents durant la totalité de l'expérience, afin de contrôler le bon déroulement de cette dernière, et de répondre aux éventuelles questions des sujets.
-\\
 
@@
 
-\subsection{Expérience 1.b : planification}
-\label{sec:ch5_planExpEvaA}
+\subsection{Planification  de l'expérience 1.b}
+\label{sec:xp1b_plan}
 
@@
 
-L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.\\
+L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.
 
@@
 
-Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de celle-ci, et de répondre aux éventuelles questions des sujets.  \\
+Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de celle-ci, et de répondre aux éventuelles questions des sujets.
 
@@
 \subsection{Données et méthodes d'analyses}
-
-\begin{table}[t]
-\centering
-\begin{tabular}{c c} 
-Termes                         & Acronymes              \\   
-\hline                    
-Niveau                         & $L$                    \\ 
-Niveau (événements)            & $L(E)$                 \\
-Niveau (textures)              & $L(T)$                 \\ 
-Agrément moyen (par scène)     & $\mathcal{A}_{scene}$  \\
-Agrément moyen (par sujet)     & $\mathcal{A}_{sujet}$  \\
-                               &                        \\
-idéale/agréable                & i                      \\
-non-idéale/désagréable         & ni                     \\
-scène idéale/agréable          & i-scène                \\
-scène non-idéale/désagréable   & ni-scène               \\
-\hline
-\end{tabular}
-\vspace{0.5mm}
-\caption{Acronyme des variables utilisées dans le cadre des expériences sensorielles.}
-\label{tab:acronyme}
-\end{table}
+\label{sec:xp1_dataAna}
 
@@
 Un résumé des descripteurs, ainsi que des acronymes les désignant, est présenté dans le Tableau~\ref{tab:acronyme}.
-Afin de rester cohérent avec l'épreuve d'évaluation, les descripteurs issus des partitions, ou des signaux des scènes, ne sont pas calculés sur la durée totale de celles-ci, mais sur une version réduite de 30 secondes (\cf~Section~\ref{sec:ch5_planExpEvaA}).
+Afin de rester cohérent avec l’expérience 1.b d'évaluation, les descripteurs ne sont pas calculés sur la durée totale des scènes, mais sur une version réduite de 30 secondes (\cf~Section~\ref{sec:xp1b_plan}).
 
@@
 \begin{itemize}
-\item \emph{Vérification de l'agrément des scènes simulées}: afin de vérifier que la distinction affective imposée entre les i- et ni-scènes se retrouve au niveau de l'agrément perçu, nous observons s'il existe des différences entre les deux types de scènes au niveau de $\mathcal{A}_{scene}$ et $\mathcal{A}_{sujet}$.
+\item \emph{vérification de l'agrément des scènes simulées}: afin de vérifier que la distinction affective imposée entre les i- et ni-scènes se retrouve au niveau de l'agrément perçu, nous observons s'il existe des différences entre les deux types de scènes au niveau de $\mathcal{A}_{scene}$ et $\mathcal{A}_{sujet}$.
 La significativité est évaluée par un test de Student à deux populations indépendantes pour $\mathcal{A}_{scene}$, et par un test de Student à deux populations appariées pour $\mathcal{A}_{sujet}$;
-\item \emph{étude comparative entre les descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels et/ou l'agrément perçu, nous évaluons cette significativité à partir d'un test de Student à deux populations;
-\item \emph{étude de l'influence des descripteurs structurels sur l'agrément perçu}: afin d'évaluer l'impact potentiel des descripteurs structurels sur l'agrément perçu, nous étudions l'existence de corrélations linéaires entre ces deux types de descripteurs.
+\item \emph{analyse des descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels et/ou l'agrément perçu, nous évaluons cette significativité à partir d'un test de Student à deux populations;
+\item \emph{influence des descripteurs structurels sur l'agrément perçu}: afin d'évaluer l'impact potentiel des descripteurs structurels sur l'agrément perçu, nous étudions l'existence de corrélations linéaires entre ces deux types de descripteurs.
 Pour mesurer la corrélation, nous utilisons le coefficient de Pearson.
 Nous adoptons ici une méthodologie couramment utilisée dans l'approche dimensionnelle;
-\item \emph{étude comparative entre les descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée a eu un impact sur la composition des scènes en terme de sources sonores, ou, pour être plus précis, s'il existe des classes de sons qui ont été particulièrement utilisées pour simuler un type d'environnement, nous utilisons le V-test.
+\item \emph{analyse des descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée a eu un impact sur la composition des scènes en terme de sources sonores, ou, pour être plus précis, s'il existe des classes de sons qui ont été particulièrement utilisées pour simuler un type d'environnement, nous utilisons le V-test.
 Nous vérifions si la présence d'une classe de sons est typique d'un environnement (i ou ni).
@@
 
-\item \emph{étude des espaces de représentation induits par les descripteurs sémantiques}: afin d'étudier si une représentation basée uniquement sur la présence ou l'absence de classes de sons permet de séparer les deux types d'environnement, nous considérons l'espace induit par les descripteurs sémantiques $S$.
+\item \emph{espaces de représentation induits par les descripteurs sémantiques}: afin d'étudier si une représentation basée uniquement sur la présence ou l'absence de classes de sons permet de séparer les deux types d'environnement, nous considérons l'espace induit par les descripteurs sémantiques $S$.
 $S$ étant un vecteur booléen, nous calculons les distances entre les scènes à partir de la distance de Hamming.
@@
 
-\item \emph{étude de l'influence spécifique des marqueurs sonores sur l'agrément perçu}: afin d'évaluer les contributions spécifiques de certaines sources sonores, nous évaluons une nouvelle fois l'impact potentiel des descripteurs structurels sur l'agrément perçu, mais en ne tenant compte, cette fois, que des marqueurs sonores pour calculer ces descripteurs.
+\item \emph{influence des marqueurs sonores sur l'agrément perçu}: afin d'évaluer les contributions spécifiques de certaines sources sonores, nous évaluons une nouvelle fois l'impact potentiel des descripteurs structurels sur l'agrément perçu, mais en ne tenant compte, cette fois, que des marqueurs sonores pour calculer ces descripteurs.
 \end{itemize}
@@
 
+\begin{table}[t]
+\centering
+\begin{tabular}{c c} 
+Termes                         & Acronymes              \\   
+\hline                    
+Niveau                         & $L$                    \\ 
+Niveau (événements)            & $L(E)$                 \\
+Niveau (textures)              & $L(T)$                 \\ 
+Agrément moyen (par scène)     & $\mathcal{A}_{scene}$  \\
+Agrément moyen (par sujet)     & $\mathcal{A}_{sujet}$  \\
+                               &                        \\
+idéale/agréable                & i                      \\
+non-idéale/désagréable         & ni                     \\
+scène idéale/agréable          & i-scène                \\
+scène non-idéale/désagréable   & ni-scène               \\
+\hline
+\end{tabular}
+\vspace{0.5mm}
+\caption{Acronyme des variables utilisées dans le cadre des expériences sensorielles.}
+\label{tab:acronyme}
+\end{table}
+
 \subsection{Résultats}
@@
 \subsubsection*{Vérification de l'agrément des scènes simulées}
-
-Nous analysons ici l'agrément perçu des $80$ scènes sonores simulées. La Figure~\ref{fig:xp2_Aa} affiche l'agrément moyen $\mathcal{A}_{scene}$ pour les i- et ni-scènes.
 
@@
 Dans un deuxième temps, nous voulons tester si les sujets ont bien perçu une différence d'agrément entre les i- et ni-scènes.
-Pour ce faire, nous observons l'agrément moyen de chaque sujet $\mathcal{A}_{sujet}$, calculé séparément, pour chaque type d'environnement (\cf~Figure~\ref{fig:xp2_Ab}).
+Pour ce faire, nous observons l'agrément moyen de chaque sujet $\mathcal{A}_{sujet}$, calculé séparément, pour chaque type d'environnement.
 Il apparaît que les i-scènes ont bien été perçues comme significativement plus agréables ($p<0.01$) que les ni-scènes.
@@
 \subsubsection*{Influence des descripteurs structurels sur l'agrément perçu}
-\label{sec:ch5_corrDesStruct}
 
@@
 
-Les mêmes observations sont faites concernant $L(E)$ (\cf~\ref{fig:soundlevele}).
-Pour $L(T)$ (\cf~\ref{fig:soundlevelf}), bien que, à considérer l'ensemble des scènes, on observe une corrélation modérée, cela n'est pas vérifié quand on regarde séparément les i-scènes ($r=-0.33$, $p=0.05$) et les ni-scènes ($r=-0.00$, $p=0.99$).
+Les mêmes observations sont faites concernant $L(E)$ (\cf~Figure~\ref{fig:soundlevele}).
+Pour $L(T)$ (\cf~Figure~\ref{fig:soundlevelf}), bien que, à considérer l'ensemble des scènes, on observe une corrélation modérée, cela n'est pas vérifié quand on regarde séparément les i-scènes ($r=-0.33$, $p=0.05$) et les ni-scènes ($r=-0.00$, $p=0.99$).
 Là encore on peut penser que la corrélation négative observée pour l'ensemble des scènes est un artefact résultant du fait que le niveau des textures des i-scènes a tendance à être plus bas que celui des ni-scènes.
@@
 Afin d'approfondir ce point, nous analysons, dans la section suivante, les scènes d'un point de vue sémantique, \ie~en nous intéressant à la nature des sources qui les composent.
-\\
 
@@
 
-\subsubsection*{Étude comparative entre les descripteurs sémantiques}
+\subsubsection*{Analyse des descripteurs sémantiques}
 
@@
         {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_3_cam}\label{fig:soundsourcec}} 
-       \caption[Pourcentage de scènes simulées comportant une classe de son particulière.]{Pourcentage de scènes simulées (i ou ni) comportant une classe de son particulière: (a) classes d'événements du niveau d'abstraction $0+$, (b) classes de textures du niveau d'abstraction $0$, (c) sous classes d'événements du niveau d'abstraction $1$ appartenant aux classes \emph{trafic} et \emph{transport public} du niveau d'abstraction $0$.}\label{fig:soundsource}
+       \caption{Pourcentage de scènes simulées (i ou ni) comportant une classe de son particulière: (a) classes d'événements du niveau d'abstraction $0+$, (b) classes de textures du niveau d'abstraction $0$, (c) sous classes d'événements du niveau d'abstraction $1$ appartenant aux classes \emph{trafic} et \emph{transport public} du niveau d'abstraction $0$.}\label{fig:soundsource}
 \end{figure}
@@
 \subsubsection*{Marqueurs sonores}
-\label{sec:ch5_marqueurXp1}
 
@@
 Nous essayons de voir maintenant si, parmi ces classes, certaines sont typiques d'un environnement en particulier.
-Pour ce faire, nous utilisons le V-test (\cf~Section~\ref{sec:ch5_methodoEtStat1}), en considérant séparément chaque niveau d'abstraction.
+Pour ce faire, nous utilisons le V-test (\cf~Section~\ref{sec:xp1_dataAna}), en considérant séparément chaque niveau d'abstraction.
 Les résultats sont présentés dans le tableau~\ref{tab:markers}.
@@
 \vspace{0.5mm}
-\caption[Classes d'événements et de textures identifiées comme étant des marqueurs sonores.]{Classes d'événements et de textures identifiées comme étant des marqueurs sonores.
-Dans chaque cellule, les marqueurs sont ordonnés par ordre décroissant de valeur $V$.}
+\caption{Classes d'événements et de textures identifiées comme étant des marqueurs sonores. Dans chaque cellule, les marqueurs sont ordonnés par ordre décroissant de valeur $V$.}
 \label{tab:markers}
@@
 
-\subsubsection*{Étude des espaces de représentation induits par les descripteurs sémantiques}
+\subsubsection*{Espaces de représentation induits par les descripteurs sémantiques}
 
 Dans cette partie, nous évaluons la capacité d'une représentation sémantique à séparer les deux types d'environnement.
-Pour ce faire, nous calculons une précision au rang 5 ($p@5$) sur l'espace induit par les descripteurs sémantiques $S$, et ce pour chaque niveau d'abstraction (\cf~Section~\ref{sec:ch5_methodoEtStat1}).
+Pour ce faire, nous calculons une précision au rang 5 ($p@5$) sur l'espace induit par les descripteurs sémantiques $S$, et ce pour chaque niveau d'abstraction (\cf~Section~\ref{sec:xp1_dataAna}).
 Les vecteurs $S$ sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.
@@
 Ces résultats indiquent que l'information sémantique permettant de séparer les deux environnements est principalement portée par les événements.
-Ces résultats font, par ailleurs, écho aux travaux de  \cite{maffiolo_caracterisation_1999}, qui montrent que nous analysons de manière descriptive (en identifiant les sources) les scènes événementielles, \ie~composées d'événements sonores (\cf~Section~\ref{sec:ch3_catsoundscape}).
+Ces résultats font, par ailleurs, écho aux travaux de  \cite{maffiolo_caracterisation_1999}, qui montrent que nous analysons de manière descriptive (en identifiant les sources) les scènes événementielles, \ie~composées d'événements sonores (\cf~Section~\ref{sec:simscene_sampleDataSet}).
 
@@
         \includegraphics[width=\linewidth]{gfx/ch_5/pa5_1}
-       \caption[$P@5$ obtenues en considérant la matrice de dissimilarité résultant des distances par paires de Hamming calculées entre les vecteurs des descripteurs sémantiques des scènes.]{$P@5$ obtenues en considérant la matrice de dissimilarité résultant des distances par paires de Hamming calculées entre les vecteurs des descripteurs sémantiques des scènes.
+       \caption{$P@5$ obtenues en considérant la matrice de dissimilarité résultant des distances par paires de Hamming calculées entre les vecteurs des descripteurs sémantiques des scènes.
 Les vecteurs sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.}\label{fig:pa5}
@@
 
-\subsubsection*{L'influence spécifique des marqueurs sonores sur l'agrément perçu}
+\subsubsection*{Influence des marqueurs sonores sur l'agrément perçu}
+
+Nous évaluons les corrélations entre $\mathcal{A}_{scene}$ et les descripteurs structurels.
+Pour cette section, les descripteurs structurels sont calculés en tenant compte des marqueurs sonores précédemment identifiés.
+Nous définissons $X_m$ le descripteur $X$ calculé en ne prenant en compte que les sons des marqueurs.
+A l'inverse, nous définissons $X_b$ ($b$: pour «~bruit~») le descripteur $X$ calculé en prenant en compte toutes les classes de sons, excepté les marqueurs.
+Lorsque le descripteur caractérise une i-scène (idem pour une ni-scène), nous ne considérons, pour le calcul, que les marqueurs identifiés pour les i-scènes (ou pour les ni-scènes), que nous nommons i-marqueurs (ou ni-marqueurs).
+Les résultats sont affichés sur le tableau~\ref{tab:corrMarkers}.
+
+Concernant les niveaux sonores (\cf~Figures~\ref{fig:soundlevelMarkera} et~\ref{fig:soundlevelMarkerd}), les mêmes tendances sont observées entre $L_m$, $L(E)_m$ et $L(T)_m$, d'une part, et $L$, $L(E)$ et $L(T)$, d'autre part.
+Que l'on considère uniquement les marqueurs, ou l'ensemble des classes, il s'avère que :
+
+\begin{enumerate}
+\item il existe une différence significative entre les niveaux des i- et ni-scènes ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
+\item le niveau sonore des scènes est majoritairement porté par les événements sonores, comparé aux textures sonores;
+\item le niveau sonore des événements a une influence sur la perception de l'agrément pour les ni-scènes, mais pas pour les i-scènes;
+\item le niveau sonore des textures ne joue aucun rôle dans la perception de l'agrément.
+\end{enumerate}
+
+En conclusion, le niveau des ni-marqueurs a une influence négative sur l'agrément pour les ni-scènes. En revanche le niveau des i-marqueurs n’impacte pas l'agrément perçu pour les i-scènes.
+
+En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
+C'est la première fois qu'un indicateur objectif nous permet de préciser l'agrément des environnements agréables.
+Ceci nous amène à conclure que le niveau des classes de sons n'étant pas typiques d'un environnement agréable a un impact négatif sur l'agrément.
+
+Par ailleurs, alors que $L(T)$ ne présentait pas de corrélation pour les ni-scènes, une corrélation négative forte est observée pour $L(T)_b$ ($r-0.73$, $p<0.01$).
+Ce fait indique que les niveaux des classes de textures n'étant pas des marqueurs n'affectent pas l'agrément perçu de la même manière pour les i- et ni-scènes.
+Les niveaux semblent avoir un effet négatif pour les ni-scènes, alors que pour les i-scènes, aucun effet n'est relevé.
+
+Pour finir, nous considérons un dernier groupe de descripteurs, nommément $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiffa} et~\ref{fig:soundlevelMarkerDiffd}).
+Ces descripteurs expriment la différence entre les niveaux des marqueurs, et ceux des autres classes de sons.
+Ils traduisent l'émergence des marqueurs par rapport à la mixture sonore.
+
+Pour les i-scènes, une corrélation modérée et positive est observée pour $L_m-L_b$ ($r=0.67$, $p<0.01$) et $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
+Pour les ni-scènes, aucune corrélation n'est observée.
+Dans le cas des i-scènes, ce n'est donc pas le niveau absolu des marqueurs qui importe, mais leur niveau relatif, par rapport aux autres sons qui composent la scène.
+On observe donc pour les environnements idéaux un double mécanisme perceptif:
+
+\begin{itemize}
+\item plus le niveau absolu des sons n'étant pas des i-marqueurs est élevé, plus l'agrément est faible,
+\item plus le niveau relatif des i-marqueurs, par rapport aux autres sons, est élevé, plus l'agrément est élevé.
+\end{itemize}
+
+Pour les ni-scènes, le fait que nous observions des corrélations pour $L_m$ et $L(E)_m$, et aucune pour $L_m-L_b$ et $L(E)_m-L(E)_b$, montre que c'est bien le niveau absolu qui importe.
 
@@
 
-Comme pour la section~\ref{sec:ch5_corrDesStruct}, nous évaluons les corrélations entre $\mathcal{A}_{scene}$ et les descripteurs structurels.
-Pour cette section, les descripteurs structurels sont calculés en tenant compte des marqueurs sonores précédemment identifiés.
-Nous définissons $X_m$ le descripteur $X$ calculé en ne prenant en compte que les sons des marqueurs.
-A l'inverse, nous définissons $X_b$ ($b$: pour «~bruit~») le descripteur $X$ calculé en prenant en compte toutes les classes de sons, excepté les marqueurs.
-Lorsque le descripteur caractérise une i-scène (idem pour une ni-scène), nous ne considérons, pour le calcul, que les marqueurs identifiés pour les i-scènes (ou pour les ni-scènes), que nous nommons i-marqueurs (ou ni-marqueurs).
-Les résultats sont affichés sur le tableau~\ref{tab:corrMarkers}.
-
-Concernant les niveaux sonores (\cf~Figures~\ref{fig:soundlevelMarker}), là encore les mêmes tendances sont observées entre $L_m$, $L(E)_m$ et $L(T)_m$, d'une part, et $L$, $L(E)$ et $L(T)$, d'autre part.
-Que l'on considère uniquement les marqueurs, ou l'ensemble des classes, il s'avère que :
-
-\begin{enumerate}
-\item il existe une différence significative entre les niveaux des i- et ni-scènes ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
-\item le niveau sonore des scènes est majoritairement porté par les événements sonores, comparé aux textures sonores;
-\item le niveau sonore des événements a une influence sur la perception de l'agrément pour les ni-scènes, mais pas pour les i-scènes;
-\item le niveau sonore des textures ne joue aucun rôle dans la perception de l'agrément.
-\end{enumerate}
-
-En conclusion, le niveau des ni-marqueurs a une influence négative sur l'agrément pour les ni-scènes. En revanche le niveau des i-marqueurs n’impacte pas l'agrément perçu pour les i-scènes.
-
-En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoise}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
-C'est la première fois qu'un indicateur objectif nous permet de préciser l'agrément des environnements agréables.
-Ceci nous amène à conclure que le niveau des classes de sons n'étant pas typiques d'un environnement agréable a un impact négatif sur l'agrément.
-
-Par ailleurs, alors que $L(T)$ ne présentait pas de corrélation pour les ni-scènes, une corrélation négative forte est observée pour $L(T)_b$ ($r-0.73$, $p<0.01$).
-Ce fait indique que les niveaux des classes de textures n'étant pas des marqueurs n'affectent pas l'agrément perçu de la même manière pour les i- et ni-scènes.
-Les niveaux semblent avoir un effet négatif pour les ni-scènes, alors que pour les i-scènes, aucun effet n'est relevé.
-
-Pour finir, nous considérons un dernier groupe de descripteurs, nommément $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiff}).
-Ces descripteurs expriment la différence entre les niveaux des marqueurs, et ceux des autres classes de sons.
-Ils traduisent l'émergence des marqueurs par rapport à la mixture sonore.
-
-Pour les i-scènes, une corrélation modérée et positive est observée pour $L_m-L_b$ ($r=0.67$, $p<0.01$) et $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
-Pour les ni-scènes, aucune corrélation n'est observée.
-Dans le cas des i-scènes, ce n'est donc pas le niveau absolu des marqueurs qui importe, mais leur niveau relatif, par rapport aux autres sons qui composent la scène.
-On observe donc pour les environnements idéaux un double mécanisme perceptif:
-
-\begin{itemize}
-\item plus le niveau absolu des sons n'étant pas des i-marqueurs est élevé, plus l'agrément est faible,
-\item plus le niveau relatif des i-marqueurs, par rapport aux autres sons, est élevé, plus l'agrément est élevé.
-\end{itemize}
-
-Pour les ni-scènes, le fait que nous observions des corrélations pour $L_m$ et $L(E)_m$, et aucune pour $L_m-L_b$ et $L(E)_m-L(E)_b$, montre que c'est bien le niveau absolu qui importe.
-
 \begin{figure}[t]
@@
 
-Dans cette expérience, nous notons que les descripteurs structurels ($L$, $L(E)$ et $L(T)$) permettent systématiquement de faire la distinction entre les environnements sonores idéaux et non-idéaux.
-
-Parmi ces indicateurs structurels, seuls $L$ et $LE$ permettent de prédire l'agrément. Nous notons cependant que cette prédiction ne vaut que pour les ni-scènes.
-
-Nous observons qu'une description sémantique des scènes, basée sur la présence/absence des classes de sons, permet de bien prédire la nature de l'environnement.
-Par ailleurs, il apparaît qu'il est possible d'obtenir une prédiction similaire, voire meilleure, en ne considérant qu'un sous groupe de classes d'événements,\ie~les marqueurs sonores.
-
-Parmi les descripteurs structurels spécifiques, calculés en tenant compte des marqueurs sonores, plusieurs permettent maintenant de faire la distinction entre les i-scènes et les ni-scènes: $L_m$, $L_{b}$, $L(E)_{m}$, $L(E)_{b}$, $L(T)_{m}$, $L(T)_{b}$, $L_{m}-L_{b}$ et $L(E)_{m}-L(E)_{b}$.
-
-Parmi ces descripteurs, 7 semblent impacter l'agrément perçu:
-
-\begin{itemize}
-\item $L_b$ et $L(E)_b$ ont un impact négatif sur les i-scènes;
-\item $L(E)_m-L(E)_b$ et $L_m-L_b$ ont un impact positif sur les i-scènes;
-\item $L_m$, $L(E)_m$ et $L(T)_b$  ont un impact négatif les ni-scènes.
-\end{itemize}
-
 De cette analyse, nous retenons les points suivants:
@@
 \begin{itemize}
-\item \emph{distinguer les i- et ni-scènes}: les descripteurs sémantiques, ainsi que certains descripteurs structurels globaux, permettent de faire la distinction entre les i-scènes et les ni-scènes.
+\item \emph{distinguer les i- et ni-scènes}: les descripteurs sémantiques, ainsi que les descripteurs structurels globaux ($L$, $L(E)$ et $L(T)$), permettent de faire la distinction entre les i-scènes et les ni-scènes.
 La description sémantique semble être plus performante;
 \item \emph{événements ou textures}: que ce soit pour les descripteurs sémantiques ou structurels, ce sont majoritairement les événements qui permettent de distinguer les deux types d'environnement, les textures n'apportant, au mieux, qu'une information limitée;
-\item \emph{prédire l'agrément}: si l'on considère une description fine de l'agrément, il semble que la manière de percevoir la qualité de l'environnement diffère en fonction de la nature de ce dernier (i ou ni).
-Il n'apparaît pas envisageable de considérer un même jeu de descripteurs pour prédire, à la fois, l'agrément des i-scènes, et l'agrément des ni-scènes.
-Pour les ni-scènes, ce sont le niveau global ($L$ et $L(E)$), ou le niveau des marqueurs sonores ($L_m$ et $L$), qui impactent négativement l'agrément.
+\item \emph{prédire l'agrément}: si l'on considère les corrélations entre les descripteurs structurels et l'agrément, il semble que la manière de percevoir la qualité de l'environnement diffère en fonction de la nature de ce dernier (i ou ni).
+Il n'apparaît pas envisageable de considérer un même jeu de descripteurs pour prédire, à la fois, l'agrément des i-scènes, et l'agrément des ni-scènes :
+
+\begin{itemize}
+
+\item \emph{pour les ni-scènes}, ce sont le niveau global ($L$ et $L(E)$), ou le niveau des marqueurs sonores ($L_m$ et $L(E)_{m}$), qui impactent négativement l'agrément.
 On note ici que prendre en compte les contributions de différentes sources n'améliore pas la capacité de prédiction de l'agrément, par rapport à une analyse holistique de l'environnement.
-Pour les i-scènes, par contre, prédire l'agrément requiert d'étudier, de manière séparée, les caractéristiques des marqueurs sonores, et celles de l'ensemble des autres sons.
-Ainsi, le niveau des marqueurs relatifs au bruit est positivement corrélé à l'agrément, alors que le niveau du bruit est, lui, négativement corrélé.
-\end{itemize}
-
-Le fait que l'agrément des scènes agréables ne se corrèle pas avec des descripteurs physiques holistiques, contrairement à l'agrément des scènes désagréables, a récemment été déjà observé \cite{gozalo2015relationship}.
-
-L'existence de deux modes de perception, mobilisant différents types de descripteurs, et dépendant de la nature (dans notre cas hédonique) des stimuli, est un phénomène qui se rapproche de celui observé pour la perception des textures (\cf~Section~\ref{sec:ch3_eventTexture}).
+
+\item \emph{pour les i-scènes}, par contre, prédire l'agrément requiert d'étudier, de manière séparée, les caractéristiques des marqueurs sonores, et celles de l'ensemble des autres sons.
+Ainsi, le niveau des marqueurs pris relativement au bruit ($L(E)_m-L(E)_b$ et $L_m-L_b$) est positivement corrélé à l'agrément, alors que le niveau du bruit ($L_b$ et $L(E)_b$) est, lui, négativement corrélé.
+\end{itemize} 
+\end{itemize}
+
+Le fait que l'agrément des i-scènes ne se corrèle pas avec des descripteurs physiques holistiques, contrairement à l'agrément des ni-scènes, a récemment été déjà observé \cite{gozalo2015relationship}.
+
+L'existence de deux modes de perception, mobilisant différents types de descripteurs, et dépendant de la nature (dans notre cas hédonique) des stimuli, est un phénomène qui se rapproche de celui observé pour la perception des textures (\cf~Section~\ref{sec:simscene_sampleDataSet}).
 Le cerveau adapte sa manière de traiter l'information (résumé statistique pour les textures, description fine pour les événements) suite à une prise de décision antérieure quant à la nature du stimulus (à savoir «~est-ce un événement ou une texture ?~»).
 De la même manière, les indicateurs actifs dans le jugement de l'agrément dépendent, eux aussi, d'une identification préalable de la nature hédonique globale de l'environnement (idéale ou non idéale).
-
-Ces résultats peuvent potentiellement influer sur les stratégies à adopter pour améliorer la qualité de l’environnement sonore:
-
-\begin{itemize}
-\item dans le cadre de scènes non-idéales, il s'agit de diminuer le niveau sonore, soit de manière globale, soit en agissant sur certaines sources (\emph{sirène}, \emph{klaxon});
-\item dans le cadre de scènes idéales, il s'agit 1) d'identifier les sons agréables, \ie~les marqueurs sonores, 2) de baisser le niveau des autres sons, 3) voire, en restant dans la limite du raisonnable, d'augmenter le niveau des marqueurs par rapport aux autres sons.
-\end{itemize}
-
-Nous montrons que les descripteurs à utiliser dépendent de la nature de l'environnement, et que cette nature est elle même dépendante de la composition sémantique, \ie~des sources sonores présentes.
-Dans une certaine mesure, nous pouvons donc dire que les descripteurs dépendent des sources sonores présentes.
 
@@
 
-\subsection{Planification expérimentale}
+\subsection{Planification  de l'expérience 2}
 
@@
 
-\subsection{Données et outils statistiques}
-
-Les données analysées sont les mêmes que pour la première expérience. Nous invitons le lecteur à se référer à la section~\ref{sec:ch5_dataType1} pour plus de détails.
+\subsection{Données et méthodes d'analyses}
+
+Les données analysées sont les mêmes que pour la première expérience. Nous invitons le lecteur à se référer à la section~\ref{sec:xp1_dataAna} pour plus de détails.
 
@@
 
-\subsubsection{Détection de valeurs extrêmes}
-
-\begin{figure}[t]
+\subsubsection{Détection des valeurs extrêmes}
+
+Considérons $\mathcal{A}_{sujet}$ pour les am-scènes.
+Il apparaît que les réponses d'un des sujets diffèrent des autres.
+Ce dernier a évalué positivement près de la moitié des ni/am-scènes (\cf~Annexe~\ref{app:xp2} : sujet 7).
+Le sujet a donné à 58\% des ni/am-scènes une note supérieure à 0, contre une moyenne de 11\% pour les autres sujets.
+De plus, le sujet a utilisé l'ambitus maximal (-5 à 5) pour noter à la fois les i/ et ni/am-scènes.
+Ces faits n'ayant pas été observés pour les autres sujets, que l'on considère les expériences 2 ou 1.b, le sujet 7 est éliminé de l'analyse.
+
+\subsubsection{Influence de la présence des marqueurs sur l'agrément perçu}
+
+Dans cette section nous étudions comment les sujets ont perçu les différents types de scènes, nommément: i/am-, ni/am-, i/sm- et ni/sm-scène.
+L'ANOVA à mesures répétées pratiquée sur $\mathcal{A}_{sujet}$ montre un effet significatif du type d'environnement (i/ni: $F[1,10]=175$, $p<0.01$), de la présence/absence des marqueurs (am/sm: $F[1,10]=7$, $p<0.05$), ainsi que de l'interaction entre les deux facteurs ($F[1,10]=67$, $p<0.01$).
+
+L'analyse \emph{post hoc} montre, quant à elle, des différences significatives entre tous les groupes d'observations, notamment entre les i/am- et i/sm-scenes ($p<0.05$) et les ni/am- et ni/sm-scenes ($p<0.01$).
+
+Ces résultats indiquent que la suppression des événements a effectivement modifié la perception des scènes par les sujets. Nos deux hypothèses sont ainsi vérifiées:
+
+\begin{itemize}
+\item la suppression des ni-marqueurs a amélioré les qualités perçues des ni-scènes;
+\item la suppression des i-marqueurs a diminué les qualités perçues des i-scènes.
+\end{itemize}
+
+L'interaction significative montre que l'effet du type d'environnement influe sur l'effet de l'absence/présence des marqueurs.
+En effet la moyenne des écarts entre am- et sm-scènes est plus importante pour les ni-scènes ($1.1$) que pour les i-scènes ($0.5$).
+
+\subsection{Discussions}
+
+De cette expérience, il apparaît que la présence, dans une scène, des marqueurs relevés à l'expérience 1 impacte bien l'agrément perçu.
+La suppression des ni-marqueurs a un effet bénéfique sur le ressenti, tandis que, plus surprenant, la suppression des i-marqueurs dégrade légèrement la qualité.
+Ce dernier point est d'autant plus marquant que, du fait de la suppression des marqueurs, le niveau des i/am-scènes est supérieur à celui des i/sm-scènes.
+
+Les i-marqueurs ont donc bien un effet bénéfique sur la perception d'un environnement.
+Le fait que leur suppression diminue $\mathcal{A}_{scene}$ montre clairement qu'il est possible d'améliorer la qualité sonore d'un lieu en ajoutant des sons bien acceptés comme \emph{oiseau}.
+Ces conclusions vont dans le sens de l'approche positive introduite par Schafer \cite{schafer1977tuning}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%    DISCUSSION AND PERSPECTIVES   %%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Conclusions et perspectives}
+
+Les expériences ont montré que la majorité des descripteurs utilisés, qu'ils soient sémantiques ou structurels, permettent de faire la distinction entre une scène idéale et une scène non-idéale.
+
+Cependant, nous observons que les caractéristiques physiques corrélées à l'agrément diffèrent clairement suivant la nature hédonique des scènes.
+Dans le cas des scènes idéales, c'est avant tout l'émergence de marqueurs sonores qui détermine la qualité perçue, alors que dans le cas des scènes non-idéales, c'est le niveau sonore global qui influe sur l'agrément.
+
+Ces résultats tendent à confirmer que la perception des qualités d'une scène dépend avant tout des sources sonores qui la composent, les caractéristiques structurelles mobilisées dans le processus perceptif semblant varier d'une source à l'autre, et d'un type d'environnement à l'autre.
+Ce fait montre qu'il est illusoire d'envisager qu'un descripteur physique holistique puisse rendre compte, de manière pertinente, des qualités affectives de tous types d'environnements.
+
+Cet état de fait peut potentiellement influer sur les stratégies à adopter pour améliorer la qualité de l’environnement sonore:
+
+\begin{itemize}
+\item dans le cadre de scènes non-idéales, il s'agit de diminuer le niveau sonore, soit de manière globale, soit en agissant sur certaines sources (\emph{sirène}, \emph{klaxon});
+\item dans le cadre de scènes idéales, il s'agit 1) d'identifier les sons agréables, \ie~les marqueurs sonores, 2) de baisser le niveau des autres sons, 3) voire, en restant dans la limite du raisonnable, d'augmenter le niveau des marqueurs par rapport aux autres sons.
+\end{itemize}
+
+Ces travaux nous permet enfin de conjecturer quant à la nature des représentations mentales des concepts «~environnement sonore urbain agréable~» (EA) et «~environnement sonore urbain désagréable~» (ED).
+
+Premièrement, le fait que les informations sémantiques (sources sonores présentes) et structurelles soient différentes pour les scènes idéales et non-idéales nous porte à croire que ces deux types d'informations caractérisent les concepts EA et ED.
+
+Deuxièmement, le fait que la suppression des marqueurs sonores modifie l'agrément perçu nous porte à croire que le concept abstrait lié à l'agrément (\eg.
+scène agréable) dépend de l'activation d'un réseau de concepts concrets liés aux sources (dans notre cas \emph{oiseau}, \emph{cloche} et \emph{sonnette vélo}).
+
+Troisièmement, le fait que les descripteurs structurels corrélés à l'agrément diffèrent en fonction du type de scènes nous porte à croire que les caractéristiques structurelles permettant de distinguer deux instances d'un même concept abstrait lié à l'agrément ne sont pas les mêmes pour EA et ED.
+
+Au terme de notre étude, nous pensons que la simulation est un outil dont le développement pourrait permettre aux décideurs en matière d'urbanisme d'interroger toute une communauté sur ses représentations propres des environnements sonores auxquels elle est exposée, et, pourquoi pas, sur les représentations des environnements sonores auxquels elle voudrait être exposée.
+
+Dans la continuité des travaux réalisés, il conviendrait de multiplier les expériences de simulation, en faisant varier les qualités affectives (calme, confortable, gênante, \etc), mais aussi en spécifiant des lieux particuliers (parc, place, rue, \etc), afin d'élaborer des corpus entiers de scènes cognitivement renseignées de paysages sonores.
+
+Il serait par ailleurs intéressant d'utiliser la simulation afin d'étudier plus avant les effets provoqués par la modification volontaire d'une caractéristique d'une scène, comme lors de la suppression des marqueurs sonores pratiquée dans l'expérience 2.
+
+Enfin, l'on pourrait encore étudier l'influence des contextes socio-culturels sur la perception.
+Dans les faits, si le son de cloche est le plus souvent un marqueur d'environnement de qualité pour un occidental, cela ne se vérifie pas nécessairement auprès de sujets de culture orientale, moyen-orientale ou autre.
+
+Outre les possibilités déjà évoquées, la simulation présente encore dans ce cas deux avantages:
+
+\begin{itemize}
+\item le simulateur peut être déployé à large échelle via internet;
+\item les scènes simulées peuvent être analysées sans avoir à tenir compte des différentes langues maternelles des sujets, la nature sémantique des classes de sons utilisées étant connue \emph{a priori} par l'expérimentateur.
+\end{itemize}
+
+Bien évidemment, ces approches nécessitent d'accroître la taille des banques de sons isolés disponibles, un effort conséquent, mais nécessaire, qui contribuera grandement aux nombreux domaines de recherche ayant trait aux scènes sonores.
+
+\section*{\normalsize Acknowledgements}
+\setlength{\parindent}{0.7cm} 
+Research project partly funded by ANR-11-JS03-005-01. The authors would like to thank the students of the Ecole Centrale de Nantes for their willing participation.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%    Bibliography    %%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\References{}{Bibliography}{actalit}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%    APPENDICES   %%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\onecolumn
+
+\appendix
+\section{Taxonomies des classes de sons}
+\label{app:taxonomie}
+
+\begin{figure}[h]
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_1}\label{fig:taxonomieEventa}}
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_2}\label{fig:taxonomieEventb}}\par
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_1}\label{fig:taxonomieTexturea}}
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_2}\label{fig:taxonomieTextureb}}
+       \caption{Taxonomies des classes de sons utilisées pour la simulation des environnements sonores urbains : (a,b) événements sonores et (c,d) textures sonores.}
+       \label{fig:taxonomie}
+\end{figure}
+
+\newpage
+\section{Dispersions des notes données par les sujets lors de l'expérience 2}
+\label{app:xp2}
+
+\begin{figure}[h]
         \myfloatalign
@@
 
-Considérons $\mathcal{A}_{sujet}$ pour les am-scènes (\cf~Figure~\ref{fig:xp4_note_2}).
-Il apparaît que les réponses du sujet 7 diffèrent des autres.
-Comme observé sur la figure~\ref{fig:xp4_note_1} (\cf~Figure~\ref{fig:xp4_note_1g} pour le sujet 7), ce dernier a évalué positivement près de la moitié des ni/am-scènes.
-Le sujet 7 a donné à 58\% des ni/am-scènes une note supérieure à 0, contre une moyenne de 11\% pour les autres sujets.
-De plus, le sujet 7 a utilisé l'ambitus maximal (-5 à 5) pour noter à la fois les i/ et ni/am-scènes.
-Ces faits n'ayant pas été observés pour les autres sujets, que l'on considère les expériences 2 ou 1.b, le sujet 7 est éliminé de l'analyse.
-
-\subsubsection{Influence de la présence des marqueurs sur l'agrément perçu}
-\label{sec:ch5_Asujet}
-
-Dans cette section nous étudions comment les sujets ont perçu les différents types de scènes, nommément: i/am-, ni/am-, i/sm- et ni/sm-scène.
-L'ANOVA à mesures répétées pratiquée sur $\mathcal{A}_{sujet}$ (\cf~Figure~\ref{fig:xp4_note_2}) montre un effet significatif du type d'environnement (i/ni: $F[1,10]=175$, $p<0.01$), de la présence/absence des marqueurs (am/sm: $F[1,10]=7$, $p<0.05$), ainsi que de l'interaction entre les deux facteurs ($F[1,10]=67$, $p<0.01$).
-
-L'analyse \emph{post hoc} montre, quant à elle, des différences significatives entre tous les groupes d'observations, notamment entre les i/am- et i/sm-scenes ($p<0.05$) et les ni/am- et ni/sm-scenes ($p<0.01$).
-
-Ces résultats indiquent que la suppression des événements a effectivement modifié la perception des scènes par les sujets. Nos deux hypothèses sont ainsi vérifiées:
-
-\begin{itemize}
-\item la suppression des ni-marqueurs a amélioré les qualités perçues des ni-scènes;
-\item la suppression des i-marqueurs a diminué les qualités perçues des i-scènes.
-\end{itemize}
-
-L'interaction significative montre que l'effet du type d'environnement influe sur l'effet de l'absence/présence des marqueurs.
-En effet la moyenne des écarts entre am- et sm-scènes est plus importante pour les ni-scènes (1.1) que pour les i-scènes (0.5).
-
-\subsection{Discussions}
-
-De cette expérience nous retenons principalement deux points.
-
-Premièrement, il apparaît que la présence, dans une scène, des marqueurs relevés à l'expérience 1 impacte bien l'agrément perçu.
-La suppression des ni-marqueurs a un effet bénéfique sur le ressenti, tandis que, plus surprenant, la suppression des i-marqueurs dégrade légèrement la qualité.
-Ce dernier point est d'autant plus marquant que, du fait de la suppression des marqueurs, le niveau des i/am-scènes est supérieur à celui des i/sm-scènes.
-
-Les i-marqueurs ont donc bien un effet bénéfique sur la perception d'un environnement.
-Le fait que leur suppression diminue $\mathcal{A}_{scene}$ montre clairement qu'il est possible d'améliorer la qualité sonore d'un lieu en ajoutant des sons bien acceptés comme \emph{oiseau}.
-Ces conclusions vont dans le sens de l'approche positive introduite par Schafer \cite{schafer1977tuning} (\cf~Section~\ref{sec:ch3_urbanNoiseSoundscape}).
-
-Deuxièmement, il apparaît que la suppression des marqueurs modifie les liens existant entre niveaux sonores et agrément perçu.
-Pour les i-scènes, ce lien est renforcé, tandis que pour les ni-scènes, il est dégradé.
-Ces résultats vont dans le sens de ceux obtenus lors de l'expérience 1, à savoir que l'influence d'un descripteur sur la perception d'une scène est conditionnée par la qualité hédonique de cette dernière.
-En l'occurrence, nous observons que plus l'agrément d'une scène est élevé, moins les descripteurs structurels influent sur la perception.
-
-Enfin, si l'on considère les représentations mentales, ces observations nous incitent à conjecturer les points suivants:
-
-\begin{itemize}
-\item  il existe un lien entre le concept concret relatif à une source sonore (\emph{oiseau}), et le concept abstrait relatif à une qualité sonore (\emph{agrément});
-\item  il existe un lien entre l'information modale que nous extrayons d'un environnement, et le concept abstrait relatif à sa qualité sonore (\emph{agrément}).
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%    DISCUSSION AND PERSPECTIVES   %%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Discussion and perspectives}
-
-\noindent \textbf{Acknowledgements}\\
-\setlength{\parindent}{0.7cm} 
-Research project partly funded by ANR-11-JS03-005-01. The authors would like to thank the students of the Ecole Centrale de Nantes for their willing participation.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%    Bibliography    %%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\References{}{Bibliography}{actalit}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%    APPENDICES   %%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\onecolumn
-
-\appendix
-\section{Taxonomies des classes de sons}
-
-\begin{figure}[h]
-        \myfloatalign
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_1}\label{fig:taxonomieEventa}}
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_2}\label{fig:taxonomieEventb}}\par
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_1}\label{fig:taxonomieTexturea}}
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_2}\label{fig:taxonomieTextureb}}
-       \caption{Taxonomies des classes de sons utilisées pour la simulation des environnements sonores urbains : (a,b) événements sonores et (c,d) textures sonores.}
-       \label{fig:taxonomie}
-\end{figure}
-
 \end{document}

### 2017-03-07 lafaySimulatedScenePerception.tex
update evet and texture

@@
 \emph{Simscene} est un environnement de travail audio-numérique dont la première version a été développée dans le cadre du projet HOULE\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
-Il  Il est prévu pour fonctionner sur les navigateurs internet \emph{Chrome} et \emph{Firefox}.
+Il est prévu pour fonctionner sur les navigateurs internet \emph{Chrome} et \emph{Firefox}.
 L'outil a été développé en javascript à l'aide de la bibliothèque \emph{angular.js}\footnote{\emph{angular.js} : \url{angularjs.org}} et du standard \emph{web-audio}\footnote{\emph{web-audio} : \url{www.w3.org/TR/webaudio}}.
@@
 
+Dans la littérature, on fait généralement la distinction entre :
+
+\begin{itemize}
+\item {un événement sonore}: un son isolé, ponctuel, dont les caractéristiques physiques varient au cours du temps;
+\item {une texture sonore}: un son, isolé, long, dont les caractéristiques physiques restent stables au cours du temps \cite{saint1995classification}.
+\end{itemize}
+
+\emph{Simscene} tient compte de cette différence morphologique, et considère deux banques de données distinctes : l'une relative aux classes d’événements, l'autre aux classes de textures.
+Les séquences d'événements et les textures profitent chacune d'un processus de simulation particulier.
+
+Cette distinction entre événements et textures sonores a des conséquences sur la perception des scènes. Ces dernières étant importantes pour l'analyse de nos résultats, nous les résumons ici.
+
 Lorsque des événements émergent d'un environnement sonore, le cerveau traite l'information des différentes sources de manière séparée.
-Plusieurs flux auditifs sont ainsi générés \ie~un pour chaque séquence d'événements émis par la même source \cite{carlyon2004brain,winkler2009modeling}.
-Inversement, quand le cerveau ne parvient pas à isoler d'événement, la scène est traitée globalement, tous ces constituants étant agglomérés dans un même flux.
-
-Dans la littérature, quatre types de sons semblent pouvoir être isolés:
-
-\begin{itemize}
-\item {événement sonore}: un son isolé, ponctuel, dont les caractéristiques physiques varient au cours du temps;
-\item {texture sonore}: un son, isolé, long, dont les caractéristiques physiques restent stables au cours du temps, et analysé à partir de statistiques extraites d'une représentation temps-fréquence;
-\item \emph{scène événementielle}: une séquence contenant une information sémantique élevée;
-\item \emph{scène amorphe}: une séquence contenant une information sémantique faible.
-\end{itemize}
-
-Une scène événementielle est une séquence composée soit uniquement d'événements, soit d'événements et de textures, les événements porteurs d'une information plus riche primant quant au choix du processus de traitement à mettre en œuvre.
-
-Les textures et les scènes amorphes, elles, sont traitées de manière holistique, à partir de propriétés acoustiques globales pour les scènes amorphes \cite{dubois2006cognitive,maffiolo_caracterisation_1999}, et sur la base d'une information résumée statistiquement pour les textures \cite{mcdermott2013summary}.
-Toutes deux portent une information limitée \cite{saint1995classification,nelken2013ear}.
-Cependant, les séquences amorphes sont spontanément décrites par les sujets comme des «~fonds sonores~» \cite{maffiolo_caracterisation_1999,guastavino2006ideal}, induisant qu'elles n'existent que grâce à un processus de construction de flux auditifs, alors que les textures sont des objets définis seulement sur la base de leur nature physique.
-Il est cependant possible d'assimiler une scène amorphe à une texture, ses caractéristiques physiques demeurant stables au cours du temps.
-De fait, nombre de scènes amorphes («~brouhaha de rue~», «~brouhaha de trafic~») sont citées comme textures.
-Cependant, l'inverse, considérer une texture comme une scène amorphe, n'est pas forcément vrai.
-Un exemple de texture souvent cité est le son du «~galop~», son qui peut très bien apparaître au premier plan de la scène.
+Plusieurs flux auditifs sont ainsi générés, \ie~un pour chaque séquence d'événements émis par la même source \cite{carlyon2004brain}.
+Inversement, quand le cerveau ne parvient pas à isoler d'événements, les éléments constituants de la scène sont agglomérés dans un même flux.
+
+Maffiolo \cite{maffiolo_caracterisation_1999} met en évidence deux processus de catégorisation distincts, engagés en fonction de la capacité de l'auditeur à identifier des événements sonores.
+Elle montre l'existence de deux catégories cognitives abstraites d'environnements sonores respectivement appelées: «~les séquences événementielles~» et «~les séquences amorphes~».
+
+Les séquences événementielles sont des environnements composés d'événements saillants et identifiables (\emph{démarrage de voiture}, \emph{voix d'homme}).
+Elles profitent d'une analyse descriptive basée sur l'identification des sources sonores.
+Les séquences amorphes sont des environnements dont il est difficile d'isoler des éléments distincts («~brouhaha de rue~», «~brouhaha de trafic~»).
+Elles bénéficient d'une analyse holistique, effectuée à partir d'indicateurs acoustiques (subjectifs) globaux.
+Cette différence de traitement entre scènes événementielles et scènes amorphes est relevée également dans \cite{guastavino2006ideal,dubois2006cognitive}.
+
+Pour ce qui est des textures, \ie~des sons possédants des caractéristiques stables au cours du temps, \cite{mcdermott2011sound,mcdermott2013summary} montrent que le cerveau peut décider sciemment de dégrader l'information physique perçue, en la résumant de manière statistique, faisant fi de toute autre représentation plus détaillée \cite{nelken2013ear}.
+
+Au regard de ces résultats, il apparaît que le processus de traitement de l'information sonore comprend une prise de décision quant à la nature des stimuli \cite{nelken2013ear,mcdermott2013summary}, laquelle va ensuite influer sur la manière d'extraire et d'analyser l'information.
 
@@
 L'utilisateur a la possibilité, à tout moment, d'écouter la scène simulée.
+
+La scène sonore est ainsi vue comme une somme de sources sonores, ou, si l'on admet la distinction opérée entre événements et textures sonores, «~un squelette d'événements sur un lit de textures~» \cite{nelken2013ear}.
 

### 2017-03-20 lafaySimulatedScenePerception.tex
raw correction jl

@@
 
-Si on demande au sujet d'évaluer un environnement donné suivant une échelle perceptive donnée (\eg~clame, agrément) \cite{axelsson2005soundscape,davies2013perception,cain2013development}, le gain d'information de l'expérimentateur est conditionné par la nature des stimuli disponibles, que ces derniers soient des scènes sonores enregistrées (expérience en laboratoire), ou réelles (expérience \emph{in situ}).
-Cependant, le fait que ces stimuli existent lui permet d'en établir une description physique précise.
-
-Si, à l'inverse, on demande au sujet de décrire un environnement donné (\eg~décrivez une scène sonore urbaine «~calme~») \cite{guastavino2006ideal, dubois2006cognitive}, l'expérimentateur obtient bien une information riche, symbolique et sémantique, de la représentation que se fait le sujet de cet environnement.
-Mais, en l'absence de données sonores, il ne peut la caractériser physiquement.
+Si on demande à un sujet d'évaluer un environnement donné suivant une échelle perceptive donnée (\eg~clame, agrément) \cite{axelsson2005soundscape,davies2013perception,cain2013development}, le gain d'information est conditionné par la nature des stimuli disponibles, qu'il s'agisse de scènes sonores enregistrées (expérience en laboratoire), ou réelles (expérience \emph{in situ}).
+Cependant, le fait que ces stimuli existent permet d'en établir une description physique précise.
+
+Si, à l'inverse, on demande à un sujet de décrire un environnement donné (\eg~décrivez une scène sonore urbaine «~calme~») \cite{guastavino2006ideal, dubois2006cognitive}, on collecte une information riche, symbolique et sémantique, de la représentation qu'il se fait de cet environnement.
+Mais, en l'absence de données sonores, on ne peut la caractériser physiquement.
 
 Nous pensons que la simulation permet de faire un lien élégant entre ces deux approches.
-Demander à un sujet de simuler un environnement sonore donné permet de générer une réalisation modale, \ie~définie sur des dimensions physiques (le signal de la scène simulée), de la représentation mentale dudit sujet.
-Cette démarche permet alors d'obtenir des scènes caractérisée à la fois sémantiquement (\eg~sources sonores présentes) et physiquement (\eg~niveaux des sources sonores).
+Demander au sujet de simuler un environnement sonore donné revient à obtenir de lui une version modale, \ie~établie sur des dimensions physiques (le signal de la scène simulée), de sa représentation mentale de l'objet, des scènes caractérisées à la fois sémantiquement (\eg~sources sonores présentes) et physiquement (\eg~niveaux des sources sonores).
 
@@
 szeremeta2009analysis}.
-En conséquence, nombre de recherches se portent maintenant sur l'étude des contributions spécifiques des différentes sources sur les qualités affectives de la scène \cite{gozalo2015relationship,ricciardi2015sound}.
-Ces recherches ont besoin d'outils permettant d'analyser séparément les influences des différentes éléments sonores qui composent l'environnement.
+Les recherches portent désormais sur l'étude des contributions spécifiques des différentes sources sur les qualités affectives de la scène \cite{gozalo2015relationship,ricciardi2015sound}.
+Ces recherches requièrent des outils permettant d'analyser séparément les influences des éléments sonores qui composent l'environnement.
 
 Nous pensons que ces travaux peuvent grandement bénéficier de l'utilisation de scènes simulées, scènes dont les caractéristiques structurelles, en particulier la nature des sources présentes, leurs positions ainsi que leurs signaux séparés, sont directement accessibles par l'expérimentateur.
-Ce qui n'est pas le cas pour des scènes enregistrées.
 
 Afin de répondre à ces problématiques, nous proposons \emph{Simscene}, un outil de simulation permettant à un sujet de simuler des paysages sonores à partir d'un corpus de sons enregistrés.
-Pour montrer l'apport de \emph{Simscene} sur l'étude des paysages sonores, nous choisissons, comme cadre applicatif, le problème de l'agrément perçu dans les environnements sonores urbains.
+Pour montrer l'apport de \emph{Simscene}, nous choisissons, comme cadre applicatif, le problème de l'agrément perçu dans les environnements sonores urbains.
 Nous présentons les résultats d'une série d'expériences s'appuyant sur la simulation, et visant, chacune, à comprendre comment les différentes sources sonores qui composent une scène influent sur la perception de l'agrément:
@@
 
-À notre connaissance, seuls Bruce~\al \cite{bruce2009development,bruce2014effects} se sont servis de la simulation afin d'étudier la perception des paysages sonores.
-Les auteurs proposent un système de simulation permettant au sujet d'agir sur un environnement en ajoutant ou supprimant des sources sonores spécifiques.
-Ce système permet par ailleurs de modifier le niveau sonore des sources, et leurs positions spatiales.
-
-À l'aide de cet outil, les auteurs demandent à leurs sujets de manipuler un jeu de sources sonores pré-établies, afin de recréer un environnement urbain.
-Ce faisant, ils montrent que l'inclusion ou l'exclusion des sources dépendent plus de considérations sociales/sémantiques, que de leurs caractéristiques physiques.
+A notre connaissance, seuls Bruce~\al \cite{bruce2009development,bruce2014effects} se sont servis de la simulation.
+Ils proposent un outil permettant d'agir sur un environnement en ajoutant ou supprimant des sources sonores spécifiques d'une part, en modifiant le niveau sonore des sources, et leurs positions spatiales d'autre part.
+Leurs sujets recréent un un environnement urbain en manipulant un jeu de sources sonores pré-établies.
+Les auteurs montrent que l'inclusion ou l'exclusion des sources dépendent plus de considérations sociales/sémantiques, que de leurs caractéristiques physiques.
 Ils soulignent néanmoins que le manque d'enregistrements disponibles limite l'analyse.
@@
 
-Le papier comprend 5 sections.
-La première constitue notre introduction.
-La deuxième introduit l'outil de simulation \emph{SimScene}.
-La troisième reprend l'expérience 1.a de \emph{simulation}, ainsi que l'épreuve 1.b d'\emph{évaluation de l'agrément}.
-La quatrième décrit l'épreuve 2 d'\emph{évaluation de l'agrément après modification des scènes}.
-Nous concluons dans la dernière partie.
-
-Par souci d'espace, nous ne détaillons pas dans ce papier l'outil de simulation, mais présentons ses fonctionnalités d'importance pour notre étude.
-Pour plus de détails sur \emph{Simscene}, nous invitons le lecteur à se référer à \cite{rossignol2015simscene,lafay2016JAES}.
-Précisons enfin que l'expérience 1.a de simulation a fait l'objet d'une étude pilote dont les résultats sont présentés dans \cite{lafay2013atiam,lafay2014new}.
+Le papier comprend 5 sections: Introduction, L'outil de simulation \emph{SimScene}, Expérience 1 .a \emph{simulation} et .b \emph{évaluation de l'agrément}, Expérience 2 \emph{évaluation de l'agrément après modification des scènes}, et Conclusions et perspectives.
+
+Dans le cadre de cette communication, nous ne présentons de \emph{SimScene} que les fonctionnalités intéressant notre étude.
+Pour plus de détails, se référer à \cite{rossignol2015simscene,lafay2016JAES}.
+Dernière précision, l'expérience 1 .a a fait l'objet d'une étude pilote dont les résultats sont présentés dans \cite{lafay2013atiam,lafay2014new}.
 
@@
 
-Afin de créer un corpus de classe de sons de référence pour la simulation, nous réalisons une typologie des sons environnementaux urbains.
-
-Pour ce faire, une étude bibliographique est effectuée, afin d'identifier les sources et ambiances sonores les plus souvent citées dans les publications.
+Afin de créer un corpus de classes de sons de référence pour la simulation, nous réalisons une typologie des sons environnementaux urbains.
+
+Une étude bibliographique est effectuée, afin d'identifier les sources et ambiances sonores les plus souvent citées dans les publications.
 Cette étude porte sur 16 articles ou thèses \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards}
-. Chacun d'eux traite de la manière dont nous discriminons les paysages sonores urbains. Notre typologie est établie sur la base des catégories/classes de sons relevées dans cette littérature.
+traitant de la manière dont nous discriminons les paysages sonores urbains. Notre typologie est établie sur la base des catégories/classes de sons relevées dans cette littérature.
 
@@
 
-Dans la littérature, on fait généralement la distinction entre :
+\jlv{Dans la littérature, on fait généralement la distinction entre}
+\jls{La littérature fait généralement la distinction entre} :
 
@@
 
-\emph{Simscene} tient compte de cette différence morphologique, et considère deux banques de données distinctes : l'une relative aux classes d’événements, l'autre aux classes de textures.
-Les séquences d'événements et les textures profitent chacune d'un processus de simulation particulier.
-
-Cette distinction entre événements et textures sonores a des conséquences sur la perception des scènes. Ces dernières étant importantes pour l'analyse de nos résultats, nous les résumons ici.
-
-Lorsque des événements émergent d'un environnement sonore, le cerveau traite l'information des différentes sources de manière séparée.
-Plusieurs flux auditifs sont ainsi générés, \ie~un pour chaque séquence d'événements émis par la même source \cite{carlyon2004brain}.
+\jls{ \emph{Simscene} tient compte de cette distinction, et \jlv{considère} \jlc{utilise} deux banques de données : l'une regroupant classes d’événements, l'autre des classes de textures.
+Evénements et textures bénéficient de processus de simulation particuliers.}
+
+\jls{Les différences morphologiques entre événements et textures sonores ont des conséquences sur la perception des scènes.}
+
+\jlc{Je reprends ici ma lecture ce matin 20170314.
+Hier j'ai perdu un peu patience et j'ai, sur les lignes 135 et 137, laissé tomber le système jlv/jlc avant de m'arrêter.
+Mes corrections: s'agissant de 135 = je retire "différences morphologiques" pour réemployer "distinction", par ailleurs je retire de la deuxième phrase "Les séquences de" pour enchaîner directement sur "Evénements et textures..."; s'agissant de 137 = je réemploie "différences morphologiques".
+Le discours gagne en fluidité, et au passage, si j'en ai la possibilité, je réduis le texte.
+C'est dans cet esprit que je continue ma relecture.}
+
+Lorsque des événements émergent d'un environnement sonore, le cerveau traite l'information \jlv{des différentes sources} de manière séparée.
+Plusieurs flux auditifs sont ainsi générés, \jlv{ \ie~un pour chaque séquence d'événements émis par la même source} \cite{carlyon2004brain}.
 Inversement, quand le cerveau ne parvient pas à isoler d'événements, les éléments constituants de la scène sont agglomérés dans un même flux.
+\jlc{on vient de poser clairement la problématique: événements/textures.
+Les notions de 1_ "séquence d'évènements" (j'ai déjà supprimé une locution similaire L135, et j'ajoute par ailleurs que tu abordes dans la phrase juste après le concept de "séquence événementielle selon Maffiolo) 2_ "source" tombent comme un cheveu dans la soupe.
+C'est trop de vocabulaire d'un coup, et ça brouille un peu les pistes...
+Tu vois ce que je veux dire ?
+tu peux reformuler ?}
 
@@
 Les séquences amorphes sont des environnements dont il est difficile d'isoler des éléments distincts («~brouhaha de rue~», «~brouhaha de trafic~»).
-Elles bénéficient d'une analyse holistique, effectuée à partir d'indicateurs acoustiques (subjectifs) globaux.
+Elles profitent, elles, d'une analyse holistique, effectuée à partir d'indicateurs acoustiques (subjectifs) globaux.
 Cette différence de traitement entre scènes événementielles et scènes amorphes est relevée également dans \cite{guastavino2006ideal,dubois2006cognitive}.
@@
 
-Nous nommons sample, un enregistrement d'un son isolé, qu'il s'agisse d'un événement ou d'une texture.
-Chaque classe est alors comprise comme une collection de samples jugés perceptivement équivalents.
-
-Les classes de sons sont organisées suivant une structure hiérarchique, afin de former une taxonomie (\cf~Figure~\ref{fig:orgDb}).
-La structure taxonomique s'inspire grandement de l'axe vertical de l'organisation catégorielle proposée par E.
+\jlv{Nous nommons sample, un enregistrement d'un son isolé, qu'il s'agisse d'un événement ou d'une texture.
+Chaque classe est alors comprise comme une collection de samples jugés perceptivement équivalents} \jls{Nous nommons sample, un enregistrement d'un son isolé, événement ou texture.
+Chaque classe est alors comprise comme une collection de samples jugés perceptivement équivalents}.
+
+\jlv{Les classes de sons sont organisées suivant une structure hiérarchique, afin de former une taxonomie \jlc{Je relis brièvement "psycho-ea" et je m'interroge.
+La taxonomie est-elle la finalité de l'organisation hiérarchique des classes (on emploie d'ailleurs plutôt le terme catégories) ou en est-elle une conséquence ?
+Autrement dit Rosch avait-il pour but de dresser cette taxonomie (comme une nécessité) ou celle-ci est-elle née de sa volonté de hiérarchiser ?
+Je ne sais pas pourquoi, mais au niveau du sens ça me perturbe}(\cf~Figure~\ref{fig:orgDb}).
+La structure taxonomique s'inspire \jlc{Exit: grandement} de l'axe vertical de l'organisation catégorielle proposée par E.
 Rosch \cite{rosch1978cognition}, \ie~plus le niveau d'abstraction de la classe est élevé, plus la description de la classe est précise, et plus les sources sonores incluses dans cette classe sont semblables.
+Si le niveau d'abstraction d'une classe est tel que cette dernière possède des sous-classes, alors sa collection de samples est la somme des collections respectives de chacune des sous-classes.}
+\jls{Les classes de sons sont organisées suivant une structure hiérarchique(\cf~Figure~\ref{fig:orgDb}) \jlc{est-ce que dans ma correction la référence à la figure (\cf~Figure~\ref{fig:orgDb}) est toujours bien placée/pertinente ?}.
+Cette structure hiérarchique est largement inspirée de l'axe vertical de l'organisation catégorielle proposée par E.
+Rosch \cite{rosch1978cognition}, qui dresse une taxonomie des classes ou catégories \ie~plus le niveau d'abstraction de la classe est élevé, plus la description de la classe est précise, et plus les sources sonores incluses dans cette classe sont semblables.
 Si le niveau d'abstraction d'une classe est tel que cette dernière possède des sous-classes, alors sa collection de samples est la somme des collections respectives de chacune des sous-classes.
+\jlc{est-ce que dans ma correction ton "\ie~plus le niveau d'abstraction..." est toujours bien placé/pertinent ?}
 
@@
 Sur la base des taxonomies précédemment établies, 483 sons isolés ont été collectés, dont 381 événements, et 102 textures.
-Parmi ces samples, 332 ont été enregistrés, et 151 proviennent de différentes banque de sons (\emph{SoundIdeas}\footnote{\emph{SoundIdeas}: \url{www.sound-ideas.com}} et \emph{Universal SoundBank}\footnote{\emph{Universal SoundBank}: \url{www.universal-soundbank.com}}).
-
-Tous les enregistrements ont été effectués à l’aide d'un micro canon \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} relié à un enregistreur \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}}.
+Parmi ces samples, 332 ont été enregistrés, et 151 proviennent de différentes banques de sons (\emph{SoundIdeas}\footnote{\emph{SoundIdeas}: \url{www.sound-ideas.com}} et \emph{Universal SoundBank}\footnote{\emph{Universal SoundBank}: \url{www.universal-soundbank.com}}).
+
+\jlv{Tous les enregistrements ont été effectués à l’aide d'un micro canon \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} relié à un enregistreur \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}}.
 L’utilisation du micro canon nous permet d’isoler les événements sonores du brouhaha urbain.
 Pour les textures, il nous permet d’éviter les événements sonores proches du preneur de son.
-Nous pouvons ainsi pointer des "zones sonores", en nous tenant à une certaine distance de ces dernières, afin de capter uniquement le brouhaha émanant de la zone ciblée.
+Nous pouvons ainsi pointer des "zones sonores", en nous tenant à une certaine distance de ces dernières, afin de capter uniquement le brouhaha émanant de la zone ciblée}.
+\jlc{Le fait de préciser le modèle de micro ne nous dispense-t-il pas, dans le cadre de cette communication, d'expliciter encore le pourquoi.
+Désigner l'AT8035 c'est aussi dire ses qualités.
+Je propose...} \jls{Exit: L’utilisation du micro canon nous permet d’isoler les événements sonores du brouhaha urbain.
+Pour les textures, il nous permet d’éviter les événements sonores proches du preneur de son.
+Nous pouvons ainsi pointer des "zones sonores", en nous tenant à une certaine distance de ces dernières, afin de capter uniquement le brouhaha émanant de la zone ciblée}.
 
@@
 
-L'outil de simulation ne propose pas d’interagir avec un sample en particulier, mais toujours avec une piste, \ie~une séquence de samples.
+L'outil de simulation \jlv{ne propose pas} \jls{n'autorise pas} d’interagir avec un sample en particulier, mais toujours avec une piste, \ie~une séquence de samples.
 Plusieurs paramètres permettent au sujet de contrôler la structure de chaque piste.
@@
 \begin{itemize}
-\item \emph{niveau sonore} ($dB$): pour chaque sample, les niveaux sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance;
-\item \emph{espacement inter-onset} (seconde): (piste d'événements seulement) comme pour les niveaux, les espacements sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance;
-\item \emph{début et fin} (seconde): le sujet fixe le début et la fin de chaque piste.
+\item \emph{niveau sonore} ($dB$): \jlv{pour chaque sample, les niveaux sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance} \jls{concerne les samples, les niveaux sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance};
+\item \emph{espacement inter-onset} (seconde): \jlv{(piste d'événements seulement) comme pour les niveaux, les espacements sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance} \jls{concerne les pistes d'événements, les espacements sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance};
+\item \emph{début et fin} (seconde): \jlv{le sujet fixe le début et la fin de chaque piste} \jls{concerne les pistes d'événements, le sujet fixe le début et la fin de chaque séquence}.
 \end{itemize}
@@
 
-\begin{itemize}
-\item \emph{fondu par événement} (seconde): (piste d'événements seulement) le sujet fixe une durée de fondu (entrée et sortie), appliquée à chaque sample d'une piste d'événements;
+\jlc{dans ce deuxième itemize on s'y perd un peu entre les "fondu par événement", "piste d'événements" et "sample" non ?}
+\begin{itemize}
+\item \emph{fondu par événement} (seconde): (piste d'événements seulement) le sujet fixe une durée de fondu (entrée et sortie), appliquée à chaque sample d'une piste d'événements \jlc{"piste d'événements seulement" est contradictoire avec "appliquée à chaque sample d'une piste..." non ?};
 \item \emph{fondu global} (seconde): le sujet fixe les durées de fondus pour l'entrée et la sortie de la piste. Ces fondus s'appliquent ainsi à l'ensemble des samples de la piste.
@@
 
-Deux de ces paramètres ne s’appliquent que pour les pistes d'événements (\emph{fondu par événement} et \emph{espacement inter-onset}), les samples des textures étant séquencés sans espacement.
+\jls{Pour coller au premier itemize (version corrigée) je propose l'itemize suivant... en espérant ne pas faire de faute de sens. Tu valideras.} 
+\begin{itemize}
+\item \emph{fondu par événement} (seconde): concerne les samples, le sujet fixe la durée de fondu (entrée et sortie) de chaque événement};
+\item \emph{fondu global} (seconde): concerne les pistes d'événements, le sujet fixe la durée de fondu (entrée et sortie) de la piste. Ce fondu s'applique à l'ensemble des samples de la séquence.
+\end{itemize}
+
+Deux de ces paramètres ne s’appliquent que pour les pistes d'événements (\emph{fondu par événement} \jlc{je réitère ma remarque.
+Si le \emph{fondu par événement} comme son nom l'indique s'applique au niveau de l'événement, il ne s'applique pas au niveau de la piste.
+Mais peut-être aussi est-ce que je me perds dans tous ces événements/textures/pistes etc.
+Tu dépatouilles ?} et \emph{espacement inter-onset}), les samples des textures étant séquencés sans espacement.
 
@@
 La grande majorité des outils permettant de parcourir une banque de données propose une recherche textuelle sur la base de mots clefs.
-L’efficacité de ce principe repose avant tout sur la structure typologique, et la nomenclature de la base de données.
-Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique.
+\jlc{Exit: L’efficacité de ce principe repose avant tout sur la structure typologique, et la nomenclature de la base de données.} Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique.
 
@@
 
-La figure~\ref{fig:ssf} présente l'interface pour la banque de données d'événements sonores.
+\jlv{La figure~\ref{fig:ssf} présente l'interface pour la banque de données d'événements sonores.
 Visuellement, les classes du dernier niveau (les seules accessibles par le sujet) sont représentées par des cercles, et positionnées sur un plan.
-La disposition des cercles dans l'espace dépend de l’organisation hiérarchique de la base de données: les sous-classes appartenant à une même classe sont proches les unes des autres, et ainsi de suite jusqu'à atteindre les classes des niveaux d'abstraction élevés.
-Le sujet ne peut accéder qu'aux classes du niveau d'abstraction le plus élevé, classes qui ne possèdent pas de sous-classes, et sont directement liées à une collection de samples.
+La disposition des cercles dans l'espace dépend de l’organisation hiérarchique de la base de données: les sous-classes appartenant à une même classe sont proches les unes des autres, et ainsi de suite jusqu'à atteindre les classes de niveaux d'abstraction élevés.
+Le sujet ne peut accéder qu'aux classes du niveau d'abstraction le plus élevé, classes qui ne possèdent pas de sous-classes, et sont directement liées à une collection de samples.}
+\jlc{dans ta parenthèse (les seules accessibles par le sujet) tu précises une chose sur laquelle tu reviens dans la phrase d'après "Le sujet ne peut accéder qu'aux classes du niveau d'abstraction le plus élevé".
+Ta parenthèse n'est peut-être pas nécessaire.
+En lieu et place tu pourrais plutôt préciser ce que sont "les classes du dernier niveau" non ?}.
+\jls{La figure~\ref{fig:ssf} présente l'interface pour la banque de données d'événements sonores.
+Visuellement, les classes du dernier \jlv{niveau} \jlc{ce mot niveau n'est pas le même, je parle de sens, que le le mot niveau de "niveau d'abstraction".
+Ne peut-on utiliser un autre terme cf.
+"dernier palier/échellon etc." ?} (les classes de niveau d'abstraction XXX) sont représentées par des cercles, et positionnées sur un plan.
+La disposition des cercles dans l'espace dépend de l’organisation hiérarchique de la base de données: les sous-classes appartenant à une même classe sont proches les unes des autres, et ainsi de suite jusqu'à atteindre les classes de niveaux d'abstraction élevés, les seules auxquelles le sujet peut accéder, des classes qui ne possèdent pas de sous-classes, et sont directement liées à une collection de samples.}
 
@@
 
+\jlc{Concernant ce qui suit et par quoi tu termines ta "section", je m'interroge sur la nécessité de préciser, dans le cadre toujours de cette communication et compte tenu du format texte qui t'est imposé.
+Ne peut-on élaguer.
+Ou alors ne peut-on aussi préciser en footnote la question étant: les footnotes sont-elles comptabilisées comme texte ou nous offrent-elles une possibilité tricher ;-) ?}
 Cette organisation visuelle à été pensée afin de:
@@
 \begin{enumerate}
-\item faciliter le parcours de la banque de données, les sons similaires (au sens des classes) étant proches les uns des autres.
+\item faciliter le parcours de la banque de données, les sons similaires \jlv{(au sens des classes)} \jls{(au regard des classes)} étant proches les uns des autres.
 \item permettre aux sujets de rapidement appréhender toute l'étendue de la banque de données, \ie~l'ensemble des sons disponibles.
@@
 Sur cette bande, chaque sample est représenté par un rectangle.
-L'espacement entre les rectangles est relatif à l'espacement entre les samples.
+L'espacement entre les rectangles est \jlv{relatif à} \jls{fonction de} l'espacement entre les samples.
 De même, la hauteur des rectangles est proportionnelle au niveau sonore des samples.
@@
 
-L'objectif est d'étudier les influences spécifiques des différentes sources sonores qui composent les environnements urbains, sur la perception de l'agrément, en utilisant la simulation.
+L'objectif est d'étudier les influences spécifiques des différentes sources sonores qui composent les environnements urbains sur la perception de l'agrément, en utilisant la simulation.
 Pour ce faire, nous planifions nos deux premières expériences (\cf~Figure~\ref{fig:xp1_2}):
@@
 
-Les sujets doivent simuler deux environnements sonores urbains, chacune des scènes devant durer 1 minute. Pour ces simulations, les sujets doivent se conformer aux consignes suivantes:
+Les sujets doivent simuler deux environnements sonores urbains, chacune des scènes devant durer 1 minute. Ils doivent se conformer aux consignes suivantes:
 
@@
 
-Tous les sujets commencent par simuler l'environnement idéal. Les sujets ne prennent connaissance de la deuxième consigne qu'à la fin de la première simulation.
+Tous commencent par simuler l'environnement idéal. Ils ne prennent connaissance de la deuxième consigne qu'à la fin de la première simulation.
 
@@
 \item le paysage sonore doit être réaliste, au sens de physiquement plausible.
-Autrement dit, le sujet a tout à fait le droit de placer 10 chiens dans son paysage sonore, mais il n’a pas le droit de placer un chien aboyant toutes les 10 millisecondes.
-\end{itemize}
-
-Ces contraintes font partie de la consigne. Aucun contrôle n'est fait \emph{a priori} dans l'interface de simulation.
+\jlv{Autrement dit, le sujet a tout à fait le droit de placer 10 chiens dans son paysage sonore, mais il n’a pas le droit de placer un chien aboyant toutes les 10 millisecondes} \jls{Le sujet peut très bien réunir 10 chiens dans un paysage sonore.
+Il n'en peut introduire 1 qui aboierait toutes les 10 millisecondes}.
+\end{itemize}
+
+\jlc{Exit: Ces contraintes font partie de la consigne. Aucun contrôle n'est fait \emph{a priori} dans l'interface de simulation.}
 
@@
 \begin{enumerate}
-\item la réalisation de la simulation: cette étape peut, elle même, se décomposer en trois actions :
+\item la réalisation de la simulation: cette étape \jlv{peut} \jlc{il ne s'agit pas d'une possibilité/éventualité, mais d'une réalité} , elle même, se décomposer en trois actions \jls{cette étape se décompose en trois temps}:
 \begin{itemize}
@@
 
-Avant de commencer la première simulation, un tutoriel de 20 minutes est proposé aux sujets, afin qu'ils se familiarisent avec le logiciel de simulation, et la banque de données.
-L'expérience est prévue pour durer 2h30.
+\jlc{Exit: Avant de commencer la première simulation, un tutoriel de 20 minutes est proposé aux sujets, afin qu'ils se familiarisent avec le logiciel de simulation, et la banque de données.
+L'expérience est prévue pour durer 2h30.}
 
 \subsubsection*{Dispositif expérimental} 
-
+\jlc{la subsubsection Dispositif expérimental me paraît pouvoir être retirée}.
 Tous les sujets passent l'expérience sur des machines identiques.
@@
 \subsubsection*{Participants} 
-
+\jlc{la subsubsection Participants me paraît pouvoir être retirée}
 44 étudiants (14 femmes) de L’École Centrale de Nantes ont participé à l'expérience.
@@
 
-En raison de contraintes temporelles, les sujets n'évaluent que des séquences de 30 secondes des scènes simulées, chacune de ces séquences commençant à la seconde 15, et finissant à le seconde 45 de la scène d'une durée d'origine de 1 minute.
+En raison de contraintes temporelles, les sujets n'évaluent que des séquences de 30 secondes des scènes simulées, chacune de ces séquences commençant à la seconde 15, et finissant à le seconde 45 de la scène d'une durée \jlv{d'origine} \jls{initiale} de 1 minute.
 
@@
 
-L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.
+\jlc{Exit: L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.}
 
 \subsubsection*{Dispositif expérimental} 
-
+\jlc{la subsubsection Dispositif expérimental me paraît pouvoir être retirée}
 Tous les sujets passent l'expérience sur des machines identiques.
@@
 \subsubsection*{Participants}
-
+\jlc{la subsubsection Participants me paraît pouvoir être retirée}
 10 étudiants (2 femmes) de L’École Centrale de Nantes ont participé à l'expérience.
@@
 
-Chaque scène est décrite par un groupe de descripteurs.
+Chaque scène est \jlv{décrite par un groupe de descripteurs} \jlc{je sais qu'il n'est plus temps de s'arrêter sur les répétitions.
+Mais bon je ne résiste pas à l'opportunité d'apporter une amélioration.
+Ne peut-on trouver un autre mot que décrite ?} \jls{A chaque scène est associé un groupe de descripteurs}.
 C'est sur ces descripteurs que nous pratiquons l'analyse.
-Un résumé des descripteurs, ainsi que des acronymes les désignant, est présenté dans le Tableau~\ref{tab:acronyme}.
+Un résumé \jlv{des descripteurs} \jls{de ces derniers}, ainsi que des acronymes les désignant, est présenté dans le Tableau~\ref{tab:acronyme}.
 Afin de rester cohérent avec l’expérience 1.b d'évaluation, les descripteurs ne sont pas calculés sur la durée totale des scènes, mais sur une version réduite de 30 secondes (\cf~Section~\ref{sec:xp1b_plan}).
@@
 \begin{itemize}
-\item \emph{perceptif}: il s'agit de l'agrément perçu de la scène simulée, évalué sur une échelle sémantique de 7 points.
+\item \emph{perceptif}: il s'agit de l'agrément perçu de la scène simulée, évalué sur une échelle sémantique de 7 points \jlc{perso, après "évalué" je préciserais ", nous l'avons dit,"...}.
 Nous notons $\mathcal{A}_{scene}$ l'agrément moyen d'une scène, obtenu en moyennant les notes de tous les sujets.
-De même, nous notons  $\mathcal{A}_{sujet}$ l'agrément par sujet, en moyennant l'ensemble de ses notes.
+De même, nous notons $\mathcal{A}_{sujet}$ l'agrément par sujet, en moyennant l'ensemble \jlv{de ses notes} \jls{des notes de chacun}.
 Compte tenu du faible nombre de sujets, nous faisons le choix, dans cette étude, de ne pas normaliser les notes d'agrément;
@@
 Pour représenter le niveau sonore, nous nous inspirons de la mesure $L_{Aeq}$.
-Dans notre cas, il s'agit d'un scalaire, calculé sur le signal en volts, et non en pression, et donné en décibels, en prenant un référentiel de 1 Volt.
+Dans notre cas, il s'agit d'un scalaire calculé sur le signal en volts, \jlc{Exit: et non en pression,} et donné en décibels, en prenant un référentiel de 1 Volt.
 Le niveau est obtenu en calculant, toutes les secondes, la moyenne quadratique du signal, et en moyennant sur la durée de la scène.
 Un filtrage de type A est opéré avant le calcul des moyennes quadratiques.
-Nous notons $L$, $L(E)$ et $L(T)$, les niveaux calculés en tenant compte respectivement de l'ensemble des samples, des samples d'événements et de textures.
-\end{itemize}
-
-Afin d'évaluer l'impact spécifique des différentes sources sonores sur l'agrément perçu, nous soumettons nos travaux aux 5 tests/études de significativité présentés ci-après:
+Nous notons $L$, $L(E)$ et $L(T)$ les niveaux calculés en tenant compte respectivement de l'ensemble des samples, des samples d'événements et de textures.
+\end{itemize}
+
+Pour évaluer l'impact spécifique des différentes sources sonores sur l'agrément perçu, nous soumettons nos travaux aux 5 tests/études de significativité présentés ci-après:
+\jlc{dans l'énumération qui suit, les deux premiers items sont exprimés par des verbes, ce qui est cohérent avec l'idée de tests/études.
+Ne peut-on conserver ce principe jusqu'au bout ?} \jlv{item 3: mesure de l'influence..., item 5: évaluation de espaces..., item 6: mesure de l'influence...}
+\jlc{par ailleurs, concernant les items 2 "analyse des descripteurs structurels" et 3 "influence des DS sur l'agrément perçu", si les problématiques sont distinctes, n'y a-t-il redondance dans tes développements ?
+Pour être précis, est-ce qu'il n'y a pas redite entre "différences significatives entre les descripteurs structurels et/ou l'agrément perçu" et "impact potentiel des descripteurs structurels sur l'agrément perçu" ?
+Mais c'est peut-être moi qui passe à côté du sens !}
 
@@
 \item \emph{analyse des descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels et/ou l'agrément perçu, nous évaluons cette significativité à partir d'un test de Student à deux populations;
-\item \emph{influence des descripteurs structurels sur l'agrément perçu}: afin d'évaluer l'impact potentiel des descripteurs structurels sur l'agrément perçu, nous étudions l'existence de corrélations linéaires entre ces deux types de descripteurs.
+\item \emph{influence des descripteurs structurels sur l'agrément perçu}: afin d'évaluer l'impact potentiel des descripteurs structurels sur l'agrément perçu, nous étudions l'existence de corrélations linéaires entre ces deux types de descripteurs \jlv{le "ces deux types de descripteurs" laisse entendre qu'on en a déjà parlé, et qu'on est à même de les identifier.
+Mais perso je ne sais à quels "deux types" de descripteurs tu fais allusion ?}.
 Pour mesurer la corrélation, nous utilisons le coefficient de Pearson.
 Nous adoptons ici une méthodologie couramment utilisée dans l'approche dimensionnelle;
-\item \emph{analyse des descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée a eu un impact sur la composition des scènes en terme de sources sonores, ou, pour être plus précis, s'il existe des classes de sons qui ont été particulièrement utilisées pour simuler un type d'environnement, nous utilisons le V-test.
-Nous vérifions si la présence d'une classe de sons est typique d'un environnement (i ou ni).
+\item \emph{analyse des descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée \jlv{a eu} \jlc{partout ailleurs tu utilises le présent, et ici tu emploies le passé composé ?} un impact sur la composition des scènes en terme de sources sonores, ou, pour être plus précis, s'il existe des classes de sons qui ont été particulièrement utilisées pour simuler un type d'environnement, nous utilisons le V-test.
+\jlv{Nous vérifions si la présence d'une classe de sons est typique d'un environnement (i ou ni)} \jlc{Les mots changent, mais l'idée est la même.
+N'y a-t-il pas redite entre cette phrase et celle qui précède ?}.
+\jls{\emph{analyse des descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée a un impact sur la composition des scènes en terme de sources sonores, s'il est des classes de sons plus particulièrement utilisées pour simuler un type d'environnement (i ou ni), nous utilisons le V-test}.
 Le test est effectué pour chaque niveau d'abstraction, et séparément, pour les classes d'événements et de textures.
@@
 
-Dans un premier temps, et afin de garantir la cohérence de nos données, nous voulons nous assurer qu'aucune ni-scène n'ait un $\mathcal{A}_{scene}$ supérieur à celui d'une i-scène.
+Dans un premier temps, et afin de garantir la cohérence de nos données, \jlv{nous voulons nous assurer} \jls{nous vérifions} qu'aucune ni-scène n'ait un $\mathcal{A}_{scene}$ supérieur à celui d'une i-scène.
 Quatre des scènes ne respectent pas la contrainte.
@@
 
-Dans un deuxième temps, nous voulons tester si les sujets ont bien perçu une différence d'agrément entre les i- et ni-scènes.
+Dans un deuxième temps, \jlv{nous voulons tester si les sujets ont bien perçu} \jls{nous voulons nous assurer que les sujets ont bien perçu} une différence d'agrément entre les i- et ni-scènes.
 Pour ce faire, nous observons l'agrément moyen de chaque sujet $\mathcal{A}_{sujet}$, calculé séparément, pour chaque type d'environnement.
@@
 
-Il apparaît que ce sont les événements qui impactent le plus le niveau global des scènes, l'écart entre $L$ et $L(E)$ n'étant que de 1 $dB$ pour les i-scènes et les ni-scènes.
+\jlv{Il apparaît que ce sont les événements qui impactent le plus le niveau global des scènes, l'écart entre $L$ et $L(E)$ n'étant que de 1 $dB$ pour les i-scènes et les ni-scènes.
 Cette observation fait écho aux résultats obtenus par Kuwano~\al~\cite{kuwano_memory_2003}.
-Au cours de leur expérience, les auteurs demandent à leurs sujets d'évaluer une série d'environnements sonores de manière globale, dans un premier temps, puis d'en évaluer le niveau aux instants où chacun identifie une source sonore.
-L'étude montre qu'il n'y a pas de différences significatives entre les jugements globaux et les moyennes des jugements instantanés.
-Pour en revenir à notre expérience, c'est comme si nos sujets avaient inconsciemment tenu compte de cette réalité perceptive lors de la simulation, en faisant porter le niveau sonore global par des sons courts et bien identifiés, \ie~les événements.
+\jlc{Exit: Au cours de leur expérience, les auteurs demandent à leurs sujets d'évaluer une série d'environnements sonores de manière globale, dans un premier temps, puis d'en évaluer le niveau aux instants où chacun identifie une source sonore.
+L'étude montre qu'il n'y a pas de différences significatives entre les jugements globaux et les moyennes des jugements instantanés.} Pour en revenir à notre expérience, c'est comme si nos sujets avaient inconsciemment tenu compte de cette réalité perceptive lors de la simulation, en faisant porter le niveau sonore global par des sons courts et bien identifiés, \ie~les événements.}
+\jls{Il apparaît que ce sont les événements qui impactent le plus le niveau global des scènes, l'écart entre $L$ et $L(E)$ n'étant que de 1 $dB$ pour les i-scènes et les ni-scènes.
+C'est comme si nos sujets avaient inconsciemment tenu compte de cette réalité perceptive lors de la simulation, en faisant porter le niveau sonore global par des sons courts et bien identifiés, \ie~les événements.
+\jlc{dans l'enchainement de ces deux phrases il se peut que voulant élaguer j'ai un peu perdu du sens.
+Est-ce que le "C'est comme si..." constitue bien un enchaînement logique entre les deux phrases ?
+Est-ce que "cette réalité perceptive" évoquée dans la seconde est effectivement exprimée dans la seconde ?} Cette observation fait écho aux résultats obtenus par Kuwano~\al~\cite{kuwano_memory_2003}}.
 
@@
 
-Cette corrélation élevée, considérant l'ensemble des scènes, résulte du fait que les i-scènes ont tendance à être moins fortes que les ni-scènes, donnant ainsi l'illusion de prolonger la corrélation négative observée pour les ni-scènes.
+\jlv{Cette corrélation élevée, considérant l'ensemble des scènes, résulte du fait que les i-scènes ont tendance à être moins fortes que les ni-scènes, donnant ainsi l'illusion de prolonger la corrélation négative observée pour les ni-scènes.} \jls{Ce résultat, considérant l'ensemble des scènes, s'explique par le fait que les i-scènes ont tendance à être moins fortes que les ni-scènes, donnant ainsi l'illusion de prolonger la corrélation négative observée pour les ni-scènes.}
 
@@
 
-Les mêmes observations sont faites concernant $L(E)$ (\cf~Figure~\ref{fig:soundlevele}).
+\jlv{Les mêmes observations sont faites \jlc{répétitions} concernant $L(E)$ (\cf~Figure~\ref{fig:soundlevele}).
 Pour $L(T)$ (\cf~Figure~\ref{fig:soundlevelf}), bien que, à considérer l'ensemble des scènes, on observe une corrélation modérée, cela n'est pas vérifié quand on regarde séparément les i-scènes ($r=-0.33$, $p=0.05$) et les ni-scènes ($r=-0.00$, $p=0.99$).
-Là encore on peut penser que la corrélation négative observée pour l'ensemble des scènes est un artefact résultant du fait que le niveau des textures des i-scènes a tendance à être plus bas que celui des ni-scènes.
-Ainsi, si les événements sonores conservent une certaine capacité de prédiction de l'agrément pour les ni-scènes, le niveau des textures n'apporte, lui, que peu d'informations, quel que soit l'environnement.
+\jlc{Exit: Là encore} on peut penser que la corrélation négative observée pour l'ensemble des scènes est un artefact résultant du fait que le niveau des textures des i-scènes a tendance à être plus bas que celui des ni-scènes.
+Ainsi, si les événements sonores conservent une certaine capacité de prédiction de l'agrément pour les ni-scènes, le niveau des textures n'apporte, lui, que peu d'informations, quel que soit l'environnement.}
+\jls{Les mêmes constats sont faits concernant $L(E)$ (\cf~Figure~\ref{fig:soundlevele}).
+Pour $L(T)$ (\cf~Figure~\ref{fig:soundlevelf}), la corrélation modérée observée sur les scènes considérées dans leur globalité n'apparaît plus sur les i-scènes et ni-scènes considérées de façon séparée (i-scènes: $r=-0.33$, $p=0.05$, ni-scènes: $r=-0.00$, $p=0.99$).
+On peut penser que la corrélation négative émergeant au niveau l'ensemble est un artefact résultant du fait que le niveau des textures des i-scènes a tendance à être plus bas que celui des ni-scènes.
+Ainsi, si les événements sonores conservent une certaine capacité de prédiction de l'agrément pour les ni-scènes, le niveau des textures n'apporte, lui, que peu d'informations, quel que soit l'environnement.}
 
@@
 
-Le fait qu'aucun des descripteurs globaux ne permette de caractériser l'agrément des i-scènes peut nous amener à penser que toutes les sources sonores ne contribuent pas de manière égale à la perception de l'agrément, mais que seules les caractéristiques de certaines d'entre elles ont une réelle influence.
+Le fait qu'aucun des \jlv{descripteurs globaux} \jlc{Sauf erreur, c'est la première fois que tu emploies le terme.
+Il fait référence à quoi ?} ne permette de caractériser l'agrément des i-scènes peut nous amener à penser que toutes les sources sonores ne contribuent pas de manière égale à la perception de l'agrément, mais que seules les caractéristiques de certaines d'entre elles ont une réelle influence.
 Afin d'approfondir ce point, nous analysons, dans la section suivante, les scènes d'un point de vue sémantique, \ie~en nous intéressant à la nature des sources qui les composent.
@@
 Concernant les textures sonores, 5 marqueurs sont identifiés.
-Pour les i-scènes, il s'agit de classes faisant référence à des ambiances amorphes, calmes, (\emph{cour-intérieur/parc} et \emph{parc}).
+Pour les i-scènes, il s'agit de classes faisant référence à des ambiances amorphes, calmes, \jlv{(\emph{cour-intérieur/parc}} \jlc{pourquoi un tiré entre cour et intérieur et un slash entre intérieur et parc ?
+intérieur est-il pris comme adjectif se rapportant à cour ou comme nom "int"érieur = maison/appartement ?} et \emph{parc}).
 Pour les ni-scènes, il s'agit, comme pour les événements, de classes faisant référence à des bruits de travaux (\emph{travaux} et \emph{véhicule de travaux}), ainsi que d'une classe faisant référence au trafic (\emph{carrefour}).
@@
         {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}      
-        \caption{Dispersions des descripteurs structurels de niveaux sonores relatifs à la présence des marqueurs $L_m$ (a, d), $L(E)_m$ (b, e) et $L(T)_m$ (c, f), en fonction du type de scènes (a, b, c) et de l'agrément perçu $\mathcal{A}_{scene}$ de l'expérience 1.b (d, e, f).}\label{fig:soundlevelMarker}
+        \caption{\jlv{Dispersions} \jlc{dispersions au plurial ?
+Tu confirmes ?} des descripteurs structurels de niveaux sonores relatifs à la présence des marqueurs $L_m$ (a, d), $L(E)_m$ (b, e) et $L(T)_m$ (c, f), en fonction du type de scènes (a, b, c) et de l'agrément perçu $\mathcal{A}_{scene}$ de l'expérience 1.b (d, e, f).}\label{fig:soundlevelMarker}
 \end{figure}
@@
 \begin{itemize}
-\item \emph{72 am-scènes}: les 72 scènes précédemment simulées, incluant les classes de marqueurs (am).
+\item \emph{72 am-scènes}: les 72 scènes précédemment simulées, \jlv{incluant} \jlc{ici j'emploierais plutôt le mot "avec" qui induit l'abréviation "am".
+Tu n'est pas d'accord ?} les classes de marqueurs (am).
 Nous notons i/am-scènes, les 36 scènes idéales avec marqueurs, et ni/am-scènes les 36 scènes non-idéales avec marqueurs;
@@
 Nombre de am-scènes sont composées, en majorité, de samples de marqueurs.
-Afin de pas abusivement dénaturer ces scènes, en créant notamment des temps de «~vide~», \ie~ne comprenant aucun sample, nous ne supprimons que les marqueurs des classes d'événements du premier niveau d'abstraction (\cf~Tableau~\ref{tab:markers}).
+\jlv{Afin de pas abusivement dénaturer ces scènes} \jls{Afin de ne pas dénaturer abusivement ces scènes}, en créant notamment des temps de «~vide~», \ie~ne comprenant aucun sample, nous ne supprimons que les marqueurs des classes d'événements du premier niveau d'abstraction (\cf~Tableau~\ref{tab:markers}).
 Ces classes sont:
@@
 
-L'expérience est prévue pour durer 1 heure. Les sujets ne connaissent pas la nature des scènes.
+\jlc{Exit: L'expérience est prévue pour durer 1 heure. Les sujets ne connaissent pas la nature des scènes.}
 
 \subsubsection*{Dispositif expérimental} 
-
+\jlc{La subsubsection Dispositif expérimental me paraît pouvoir être retirée.}
 Tous les sujets passent l'expérience sur des machines identiques.
@@
 \subsubsection*{Participants} 
-
+\jlc{La subsubsection Participants me paraît pouvoir être retirée.}
 12 sujets (4 femmes) participent à l'expérience.
@@
 
-De cette expérience, il apparaît que la présence, dans une scène, des marqueurs relevés à l'expérience 1 impacte bien l'agrément perçu.
+\jlv{De cette expérience, il apparaît que} \jls{Cette expérience fait apparaître que} la présence, dans une scène, des marqueurs relevés à l'expérience 1 impacte bien l'agrément perçu.
 La suppression des ni-marqueurs a un effet bénéfique sur le ressenti, tandis que, plus surprenant, la suppression des i-marqueurs dégrade légèrement la qualité.
@@
 Ces résultats tendent à confirmer que la perception des qualités d'une scène dépend avant tout des sources sonores qui la composent, les caractéristiques structurelles mobilisées dans le processus perceptif semblant varier d'une source à l'autre, et d'un type d'environnement à l'autre.
-Ce fait montre qu'il est illusoire d'envisager qu'un descripteur physique holistique puisse rendre compte, de manière pertinente, des qualités affectives de tous types d'environnements.
+Ce fait montre qu'il est illusoire d'envisager qu'un descripteur physique holistique puisse rendre compte, de manière pertinente, des qualités affectives de tous types d'environnement.
 
@@
 
-Ces travaux nous permet enfin de conjecturer quant à la nature des représentations mentales des concepts «~environnement sonore urbain agréable~» (EA) et «~environnement sonore urbain désagréable~» (ED).
+Ces travaux nous permettent enfin de conjecturer quant à la nature des représentations mentales des concepts «~environnement sonore urbain agréable~» (EA) et «~environnement sonore urbain désagréable~» (ED).
 

### 2017-03-20 lafaySimulatedScenePerception.tex
merge jl corrections sections 1 and 2

@@
 
-\jlv{Dans la littérature, on fait généralement la distinction entre}
-\jls{La littérature fait généralement la distinction entre} :
+La littérature fait généralement la distinction entre :
 
@@
 
-\jls{ \emph{Simscene} tient compte de cette distinction, et \jlv{considère} \jlc{utilise} deux banques de données : l'une regroupant classes d’événements, l'autre des classes de textures.
-Evénements et textures bénéficient de processus de simulation particuliers.}
-
-\jls{Les différences morphologiques entre événements et textures sonores ont des conséquences sur la perception des scènes.}
-
-\jlc{Je reprends ici ma lecture ce matin 20170314.
-Hier j'ai perdu un peu patience et j'ai, sur les lignes 135 et 137, laissé tomber le système jlv/jlc avant de m'arrêter.
-Mes corrections: s'agissant de 135 = je retire "différences morphologiques" pour réemployer "distinction", par ailleurs je retire de la deuxième phrase "Les séquences de" pour enchaîner directement sur "Evénements et textures..."; s'agissant de 137 = je réemploie "différences morphologiques".
-Le discours gagne en fluidité, et au passage, si j'en ai la possibilité, je réduis le texte.
-C'est dans cet esprit que je continue ma relecture.}
-
-Lorsque des événements émergent d'un environnement sonore, le cerveau traite l'information \jlv{des différentes sources} de manière séparée.
-Plusieurs flux auditifs sont ainsi générés, \jlv{ \ie~un pour chaque séquence d'événements émis par la même source} \cite{carlyon2004brain}.
+\emph{Simscene} tient compte de cette distinction, et utilise deux banques de données : l'une regroupant classes d’événements, l'autre des classes de textures.
+Événements et textures bénéficient de processus de simulation particuliers.
+
+Les différences morphologiques entre événements et textures sonores ont des conséquences sur la perception des scènes.
+
+Lorsque des événements émergent d'un environnement sonore, le cerveau traite l'information des différentes sources de manière séparée.
+Plusieurs flux auditifs sont ainsi générés, \ie~un pour chaque séquence d'événements émis par la même source \cite{carlyon2004brain}.
 Inversement, quand le cerveau ne parvient pas à isoler d'événements, les éléments constituants de la scène sont agglomérés dans un même flux.
-\jlc{on vient de poser clairement la problématique: événements/textures.
-Les notions de 1_ "séquence d'évènements" (j'ai déjà supprimé une locution similaire L135, et j'ajoute par ailleurs que tu abordes dans la phrase juste après le concept de "séquence événementielle selon Maffiolo) 2_ "source" tombent comme un cheveu dans la soupe.
-C'est trop de vocabulaire d'un coup, et ça brouille un peu les pistes...
-Tu vois ce que je veux dire ?
-tu peux reformuler ?}
 
@@
 
-\jlv{Nous nommons sample, un enregistrement d'un son isolé, qu'il s'agisse d'un événement ou d'une texture.
-Chaque classe est alors comprise comme une collection de samples jugés perceptivement équivalents} \jls{Nous nommons sample, un enregistrement d'un son isolé, événement ou texture.
-Chaque classe est alors comprise comme une collection de samples jugés perceptivement équivalents}.
-
-\jlv{Les classes de sons sont organisées suivant une structure hiérarchique, afin de former une taxonomie \jlc{Je relis brièvement "psycho-ea" et je m'interroge.
-La taxonomie est-elle la finalité de l'organisation hiérarchique des classes (on emploie d'ailleurs plutôt le terme catégories) ou en est-elle une conséquence ?
-Autrement dit Rosch avait-il pour but de dresser cette taxonomie (comme une nécessité) ou celle-ci est-elle née de sa volonté de hiérarchiser ?
-Je ne sais pas pourquoi, mais au niveau du sens ça me perturbe}(\cf~Figure~\ref{fig:orgDb}).
-La structure taxonomique s'inspire \jlc{Exit: grandement} de l'axe vertical de l'organisation catégorielle proposée par E.
-Rosch \cite{rosch1978cognition}, \ie~plus le niveau d'abstraction de la classe est élevé, plus la description de la classe est précise, et plus les sources sonores incluses dans cette classe sont semblables.
-Si le niveau d'abstraction d'une classe est tel que cette dernière possède des sous-classes, alors sa collection de samples est la somme des collections respectives de chacune des sous-classes.}
-\jls{Les classes de sons sont organisées suivant une structure hiérarchique(\cf~Figure~\ref{fig:orgDb}) \jlc{est-ce que dans ma correction la référence à la figure (\cf~Figure~\ref{fig:orgDb}) est toujours bien placée/pertinente ?}.
+Nous nommons sample, un enregistrement d'un son isolé, événement ou texture. Chaque classe est alors comprise comme une collection de samples jugés perceptivement équivalents.
+
+Les classes de sons sont organisées suivant une structure hiérarchique (\cf~Figure~\ref{fig:orgDb}).
 Cette structure hiérarchique est largement inspirée de l'axe vertical de l'organisation catégorielle proposée par E.
-Rosch \cite{rosch1978cognition}, qui dresse une taxonomie des classes ou catégories \ie~plus le niveau d'abstraction de la classe est élevé, plus la description de la classe est précise, et plus les sources sonores incluses dans cette classe sont semblables.
+Rosch \cite{rosch1978cognition}, qui dresse une taxonomie des classes ou catégories \ie~plus le niveau d'abstraction de la classe est faible, plus la description de la classe est précise, et plus les sources sonores incluses dans cette classe sont semblables.
 Si le niveau d'abstraction d'une classe est tel que cette dernière possède des sous-classes, alors sa collection de samples est la somme des collections respectives de chacune des sous-classes.
-\jlc{est-ce que dans ma correction ton "\ie~plus le niveau d'abstraction..." est toujours bien placé/pertinent ?}
 
@@
 
-\jlv{Tous les enregistrements ont été effectués à l’aide d'un micro canon \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} relié à un enregistreur \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}}.
+Tous les enregistrements ont été effectués à l’aide d'un micro canon \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} relié à un enregistreur \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}}.
 L’utilisation du micro canon nous permet d’isoler les événements sonores du brouhaha urbain.
 Pour les textures, il nous permet d’éviter les événements sonores proches du preneur de son.
-Nous pouvons ainsi pointer des "zones sonores", en nous tenant à une certaine distance de ces dernières, afin de capter uniquement le brouhaha émanant de la zone ciblée}.
-\jlc{Le fait de préciser le modèle de micro ne nous dispense-t-il pas, dans le cadre de cette communication, d'expliciter encore le pourquoi.
-Désigner l'AT8035 c'est aussi dire ses qualités.
-Je propose...} \jls{Exit: L’utilisation du micro canon nous permet d’isoler les événements sonores du brouhaha urbain.
-Pour les textures, il nous permet d’éviter les événements sonores proches du preneur de son.
-Nous pouvons ainsi pointer des "zones sonores", en nous tenant à une certaine distance de ces dernières, afin de capter uniquement le brouhaha émanant de la zone ciblée}.
+Nous pouvons ainsi pointer des "zones sonores", en nous tenant à une certaine distance de ces dernières, afin de capter uniquement le brouhaha émanant de la zone ciblée.
 
@@
 
-L'outil de simulation \jlv{ne propose pas} \jls{n'autorise pas} d’interagir avec un sample en particulier, mais toujours avec une piste, \ie~une séquence de samples.
+L'outil de simulation n'autorise pas d’interagir avec un sample en particulier, mais toujours avec une piste, \ie~une séquence de samples.
 Plusieurs paramètres permettent au sujet de contrôler la structure de chaque piste.
@@
 \begin{itemize}
-\item \emph{niveau sonore} ($dB$): \jlv{pour chaque sample, les niveaux sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance} \jls{concerne les samples, les niveaux sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance};
-\item \emph{espacement inter-onset} (seconde): \jlv{(piste d'événements seulement) comme pour les niveaux, les espacements sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance} \jls{concerne les pistes d'événements, les espacements sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance};
-\item \emph{début et fin} (seconde): \jlv{le sujet fixe le début et la fin de chaque piste} \jls{concerne les pistes d'événements, le sujet fixe le début et la fin de chaque séquence}.
+\item \emph{niveau sonore} ($dB$): pour chaque sample, les niveaux sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance;
+\item \emph{espacement inter-onset} (seconde): (piste d'événements seulement) comme pour les niveaux, les espacements sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance;
+\item \emph{début et fin} (seconde): le sujet fixe le début et la fin de chaque piste.
 \end{itemize}
@@
 
-\jlc{dans ce deuxième itemize on s'y perd un peu entre les "fondu par événement", "piste d'événements" et "sample" non ?}
-\begin{itemize}
-\item \emph{fondu par événement} (seconde): (piste d'événements seulement) le sujet fixe une durée de fondu (entrée et sortie), appliquée à chaque sample d'une piste d'événements \jlc{"piste d'événements seulement" est contradictoire avec "appliquée à chaque sample d'une piste..." non ?};
+\begin{itemize}
+\item \emph{fondu par événement} (seconde): (piste d'événements seulement) le sujet fixe une durée de fondu (entrée et sortie) appliquée à chaque sample;
 \item \emph{fondu global} (seconde): le sujet fixe les durées de fondus pour l'entrée et la sortie de la piste. Ces fondus s'appliquent ainsi à l'ensemble des samples de la piste.
@@
 
-\jls{Pour coller au premier itemize (version corrigée) je propose l'itemize suivant... en espérant ne pas faire de faute de sens. Tu valideras.} 
-\begin{itemize}
-\item \emph{fondu par événement} (seconde): concerne les samples, le sujet fixe la durée de fondu (entrée et sortie) de chaque événement};
-\item \emph{fondu global} (seconde): concerne les pistes d'événements, le sujet fixe la durée de fondu (entrée et sortie) de la piste. Ce fondu s'applique à l'ensemble des samples de la séquence.
-\end{itemize}
-
-Deux de ces paramètres ne s’appliquent que pour les pistes d'événements (\emph{fondu par événement} \jlc{je réitère ma remarque.
-Si le \emph{fondu par événement} comme son nom l'indique s'applique au niveau de l'événement, il ne s'applique pas au niveau de la piste.
-Mais peut-être aussi est-ce que je me perds dans tous ces événements/textures/pistes etc.
-Tu dépatouilles ?} et \emph{espacement inter-onset}), les samples des textures étant séquencés sans espacement.
+Deux de ces paramètres ne s’appliquent que pour les pistes d'événements (\emph{fondu par événement}  et \emph{espacement inter-onset}), les samples des textures étant séquencés sans espacement.
 
@@
 La grande majorité des outils permettant de parcourir une banque de données propose une recherche textuelle sur la base de mots clefs.
-\jlc{Exit: L’efficacité de ce principe repose avant tout sur la structure typologique, et la nomenclature de la base de données.} Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique.
+Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique.
 
@@
 
-\jlv{La figure~\ref{fig:ssf} présente l'interface pour la banque de données d'événements sonores.
-Visuellement, les classes du dernier niveau (les seules accessibles par le sujet) sont représentées par des cercles, et positionnées sur un plan.
-La disposition des cercles dans l'espace dépend de l’organisation hiérarchique de la base de données: les sous-classes appartenant à une même classe sont proches les unes des autres, et ainsi de suite jusqu'à atteindre les classes de niveaux d'abstraction élevés.
-Le sujet ne peut accéder qu'aux classes du niveau d'abstraction le plus élevé, classes qui ne possèdent pas de sous-classes, et sont directement liées à une collection de samples.}
-\jlc{dans ta parenthèse (les seules accessibles par le sujet) tu précises une chose sur laquelle tu reviens dans la phrase d'après "Le sujet ne peut accéder qu'aux classes du niveau d'abstraction le plus élevé".
-Ta parenthèse n'est peut-être pas nécessaire.
-En lieu et place tu pourrais plutôt préciser ce que sont "les classes du dernier niveau" non ?}.
-\jls{La figure~\ref{fig:ssf} présente l'interface pour la banque de données d'événements sonores.
-Visuellement, les classes du dernier \jlv{niveau} \jlc{ce mot niveau n'est pas le même, je parle de sens, que le le mot niveau de "niveau d'abstraction".
-Ne peut-on utiliser un autre terme cf.
-"dernier palier/échellon etc." ?} (les classes de niveau d'abstraction XXX) sont représentées par des cercles, et positionnées sur un plan.
-La disposition des cercles dans l'espace dépend de l’organisation hiérarchique de la base de données: les sous-classes appartenant à une même classe sont proches les unes des autres, et ainsi de suite jusqu'à atteindre les classes de niveaux d'abstraction élevés, les seules auxquelles le sujet peut accéder, des classes qui ne possèdent pas de sous-classes, et sont directement liées à une collection de samples.}
+La figure~\ref{fig:ssf} présente l'interface pour la banque de données d'événements sonores.
+Visuellement, les classes du dernier palier (les classes d'un niveau d'abstraction faible) sont représentées par des cercles, et positionnées sur un plan.
+La disposition des cercles dans l'espace dépend de l’organisation hiérarchique de la base de données: les sous-classes appartenant à une même classe sont proches les unes des autres, et ainsi de suite jusqu'à atteindre les classes du dernier palier, les seules auxquelles le sujet peut accéder, des classes qui ne possèdent pas de sous-classes, et sont directement liées à une collection de samples.
 
@@
 
-\jlc{Concernant ce qui suit et par quoi tu termines ta "section", je m'interroge sur la nécessité de préciser, dans le cadre toujours de cette communication et compte tenu du format texte qui t'est imposé.
-Ne peut-on élaguer.
-Ou alors ne peut-on aussi préciser en footnote la question étant: les footnotes sont-elles comptabilisées comme texte ou nous offrent-elles une possibilité tricher ;-) ?}
-Cette organisation visuelle à été pensée afin de:
-
-\begin{enumerate}
-\item faciliter le parcours de la banque de données, les sons similaires \jlv{(au sens des classes)} \jls{(au regard des classes)} étant proches les uns des autres.
-\item permettre aux sujets de rapidement appréhender toute l'étendue de la banque de données, \ie~l'ensemble des sons disponibles.
-\end{enumerate}
-
 Cette interface a fait l'objet d'une étude approfondie dont les résultats sont publiés dans \cite{lafay2016JAES}.
@@
 Sur cette bande, chaque sample est représenté par un rectangle.
-L'espacement entre les rectangles est \jlv{relatif à} \jls{fonction de} l'espacement entre les samples.
+L'espacement entre les rectangles est fonction de l'espacement entre les samples.
 De même, la hauteur des rectangles est proportionnelle au niveau sonore des samples.

### 2017-03-20 lafaySimulatedScenePerception.tex
merge all sections

@@
 Pour les événements, nous considérons quatre niveaux d'abstraction allant des classes les plus globalisantes (niveau d'abstraction 0), aux classes les plus spécifiques (niveau d'abstraction 3).
-Pour les textures, nous ne considérons que trois niveaux d'abstraction.
+Pour les textures, nous ne considérons que trois niveaux d'abstraction .
 Notons que la taxonomie des événements est assez proche de celle de \cite{Salamon14}, introduite ultérieurement à notre étude.
@@
 La figure~\ref{fig:ssf} présente l'interface pour la banque de données d'événements sonores.
-Visuellement, les classes du dernier palier (les classes d'un niveau d'abstraction faible) sont représentées par des cercles, et positionnées sur un plan.
+Visuellement, les classes du dernier palier (niveaux d'abstraction 3 pour les événements, 2 pour les textures) sont représentées par des cercles, et positionnées sur un plan.
 La disposition des cercles dans l'espace dépend de l’organisation hiérarchique de la base de données: les sous-classes appartenant à une même classe sont proches les unes des autres, et ainsi de suite jusqu'à atteindre les classes du dernier palier, les seules auxquelles le sujet peut accéder, des classes qui ne possèdent pas de sous-classes, et sont directement liées à une collection de samples.
 
-Chaque classe possède un son prototype.
+Chaque classe du dernier pallier possède un son prototype.
 Ces sons ont été choisis par les expérimentateurs.
@@
 \item le paysage sonore doit être réaliste, au sens de physiquement plausible.
-\jlv{Autrement dit, le sujet a tout à fait le droit de placer 10 chiens dans son paysage sonore, mais il n’a pas le droit de placer un chien aboyant toutes les 10 millisecondes} \jls{Le sujet peut très bien réunir 10 chiens dans un paysage sonore.
-Il n'en peut introduire 1 qui aboierait toutes les 10 millisecondes}.
-\end{itemize}
-
-\jlc{Exit: Ces contraintes font partie de la consigne. Aucun contrôle n'est fait \emph{a priori} dans l'interface de simulation.}
+Le sujet peut très bien réunir 10 chiens dans un paysage sonore.
+Il n'en peut introduire 1 qui aboierait toutes les 10 millisecondes.
+\end{itemize}
+
+Ces contraintes font partie de la consigne. Aucun contrôle n'est fait \emph{a priori} dans l'interface de simulation.
 
@@
 \begin{enumerate}
-\item la réalisation de la simulation: cette étape \jlv{peut} \jlc{il ne s'agit pas d'une possibilité/éventualité, mais d'une réalité} , elle même, se décomposer en trois actions \jls{cette étape se décompose en trois temps}:
+\item la réalisation de la simulation: cette étape cette étape se décompose en trois temps:
 \begin{itemize}
@@
 
-\jlc{Exit: Avant de commencer la première simulation, un tutoriel de 20 minutes est proposé aux sujets, afin qu'ils se familiarisent avec le logiciel de simulation, et la banque de données.
-L'expérience est prévue pour durer 2h30.}
+Avant de commencer la première simulation, un tutoriel de 20 minutes est proposé aux sujets, afin qu'ils se familiarisent avec le logiciel de simulation, et la banque de données.
+L'expérience est prévue pour durer 2h30.
 
 \subsubsection*{Dispositif expérimental}
-\jlc{la subsubsection Dispositif expérimental me paraît pouvoir être retirée}.
+
 Tous les sujets passent l'expérience sur des machines identiques.
@@
 \subsubsection*{Participants}
-\jlc{la subsubsection Participants me paraît pouvoir être retirée}
+
 44 étudiants (14 femmes) de L’École Centrale de Nantes ont participé à l'expérience.
@@
 
-En raison de contraintes temporelles, les sujets n'évaluent que des séquences de 30 secondes des scènes simulées, chacune de ces séquences commençant à la seconde 15, et finissant à le seconde 45 de la scène d'une durée \jlv{d'origine} \jls{initiale} de 1 minute.
+En raison de contraintes temporelles, les sujets n'évaluent que des séquences de 30 secondes des scènes simulées, chacune de ces séquences commençant à la seconde 15, et finissant à le seconde 45 de la scène d'une durée initiale de 1 minute.
 
@@
 
-\jlc{Exit: L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.}
+L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.
 
 \subsubsection*{Dispositif expérimental}
-\jlc{la subsubsection Dispositif expérimental me paraît pouvoir être retirée}
+
 Tous les sujets passent l'expérience sur des machines identiques.
@@
 \subsubsection*{Participants}
-\jlc{la subsubsection Participants me paraît pouvoir être retirée}
+
 10 étudiants (2 femmes) de L’École Centrale de Nantes ont participé à l'expérience.
@@
 
-Chaque scène est \jlv{décrite par un groupe de descripteurs} \jlc{je sais qu'il n'est plus temps de s'arrêter sur les répétitions.
-Mais bon je ne résiste pas à l'opportunité d'apporter une amélioration.
-Ne peut-on trouver un autre mot que décrite ?} \jls{A chaque scène est associé un groupe de descripteurs}.
+A chaque scène est associé un groupe de descripteurs.
 C'est sur ces descripteurs que nous pratiquons l'analyse.
-Un résumé \jlv{des descripteurs} \jls{de ces derniers}, ainsi que des acronymes les désignant, est présenté dans le Tableau~\ref{tab:acronyme}.
+Un résumé de ces derniers, ainsi que des acronymes les désignant, est présenté dans le Tableau~\ref{tab:acronyme}.
 Afin de rester cohérent avec l’expérience 1.b d'évaluation, les descripteurs ne sont pas calculés sur la durée totale des scènes, mais sur une version réduite de 30 secondes (\cf~Section~\ref{sec:xp1b_plan}).
@@
 \begin{itemize}
-\item \emph{perceptif}: il s'agit de l'agrément perçu de la scène simulée, évalué sur une échelle sémantique de 7 points \jlc{perso, après "évalué" je préciserais ", nous l'avons dit,"...}.
+\item \emph{perceptif}: il s'agit de l'agrément perçu de la scène simulée, évalué sur une échelle sémantique de 7 points.
 Nous notons $\mathcal{A}_{scene}$ l'agrément moyen d'une scène, obtenu en moyennant les notes de tous les sujets.
-De même, nous notons $\mathcal{A}_{sujet}$ l'agrément par sujet, en moyennant l'ensemble \jlv{de ses notes} \jls{des notes de chacun}.
+De même, nous notons $\mathcal{A}_{sujet}$ l'agrément par sujet, en moyennant l'ensemble des notes de chacun.
 Compte tenu du faible nombre de sujets, nous faisons le choix, dans cette étude, de ne pas normaliser les notes d'agrément;
@@
 Pour représenter le niveau sonore, nous nous inspirons de la mesure $L_{Aeq}$.
-Dans notre cas, il s'agit d'un scalaire calculé sur le signal en volts, \jlc{Exit: et non en pression,} et donné en décibels, en prenant un référentiel de 1 Volt.
+Dans notre cas, il s'agit d'un scalaire calculé sur le signal en volts, et non en pression, et donné en décibels, en prenant un référentiel de 1 Volt.
 Le niveau est obtenu en calculant, toutes les secondes, la moyenne quadratique du signal, et en moyennant sur la durée de la scène.
@@
 Pour évaluer l'impact spécifique des différentes sources sonores sur l'agrément perçu, nous soumettons nos travaux aux 5 tests/études de significativité présentés ci-après:
-\jlc{dans l'énumération qui suit, les deux premiers items sont exprimés par des verbes, ce qui est cohérent avec l'idée de tests/études.
-Ne peut-on conserver ce principe jusqu'au bout ?} \jlv{item 3: mesure de l'influence..., item 5: évaluation de espaces..., item 6: mesure de l'influence...}
-\jlc{par ailleurs, concernant les items 2 "analyse des descripteurs structurels" et 3 "influence des DS sur l'agrément perçu", si les problématiques sont distinctes, n'y a-t-il redondance dans tes développements ?
-Pour être précis, est-ce qu'il n'y a pas redite entre "différences significatives entre les descripteurs structurels et/ou l'agrément perçu" et "impact potentiel des descripteurs structurels sur l'agrément perçu" ?
-Mais c'est peut-être moi qui passe à côté du sens !}
 
@@
 \item \emph{analyse des descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels et/ou l'agrément perçu, nous évaluons cette significativité à partir d'un test de Student à deux populations;
-\item \emph{influence des descripteurs structurels sur l'agrément perçu}: afin d'évaluer l'impact potentiel des descripteurs structurels sur l'agrément perçu, nous étudions l'existence de corrélations linéaires entre ces deux types de descripteurs \jlv{le "ces deux types de descripteurs" laisse entendre qu'on en a déjà parlé, et qu'on est à même de les identifier.
-Mais perso je ne sais à quels "deux types" de descripteurs tu fais allusion ?}.
+\item \emph{étude de l'influence des descripteurs structurels sur l'agrément perçu}: afin d'évaluer l'impact potentiel des descripteurs structurels sur l'agrément perçu, nous étudions l'existence de corrélations linéaires entre ces descripteurs et $\mathcal{A}_{scene}$ .
 Pour mesurer la corrélation, nous utilisons le coefficient de Pearson.
 Nous adoptons ici une méthodologie couramment utilisée dans l'approche dimensionnelle;
-\item \emph{analyse des descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée \jlv{a eu} \jlc{partout ailleurs tu utilises le présent, et ici tu emploies le passé composé ?} un impact sur la composition des scènes en terme de sources sonores, ou, pour être plus précis, s'il existe des classes de sons qui ont été particulièrement utilisées pour simuler un type d'environnement, nous utilisons le V-test.
-\jlv{Nous vérifions si la présence d'une classe de sons est typique d'un environnement (i ou ni)} \jlc{Les mots changent, mais l'idée est la même.
-N'y a-t-il pas redite entre cette phrase et celle qui précède ?}.
-\jls{\emph{analyse des descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée a un impact sur la composition des scènes en terme de sources sonores, s'il est des classes de sons plus particulièrement utilisées pour simuler un type d'environnement (i ou ni), nous utilisons le V-test}.
+\item \emph{analyse des descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée a un impact sur la composition des scènes en terme de sources sonores, s'il est des classes de sons plus particulièrement utilisées pour simuler un type d'environnement (i ou ni), nous utilisons le V-test.
 Le test est effectué pour chaque niveau d'abstraction, et séparément, pour les classes d'événements et de textures.
@@
 
-\item \emph{espaces de représentation induits par les descripteurs sémantiques}: afin d'étudier si une représentation basée uniquement sur la présence ou l'absence de classes de sons permet de séparer les deux types d'environnement, nous considérons l'espace induit par les descripteurs sémantiques $S$.
+\item \emph{étude des espaces de représentation induits par les descripteurs sémantiques}: afin d'étudier si une représentation basée uniquement sur la présence ou l'absence de classes de sons permet de séparer les deux types d'environnement, nous considérons l'espace induit par les descripteurs sémantiques $S$.
 $S$ étant un vecteur booléen, nous calculons les distances entre les scènes à partir de la distance de Hamming.
@@
 
-Dans un premier temps, et afin de garantir la cohérence de nos données, \jlv{nous voulons nous assurer} \jls{nous vérifions} qu'aucune ni-scène n'ait un $\mathcal{A}_{scene}$ supérieur à celui d'une i-scène.
+Dans un premier temps, et afin de garantir la cohérence de nos données, nous vérifions qu'aucune ni-scène n'ait un $\mathcal{A}_{scene}$ supérieur à celui d'une i-scène.
 Quatre des scènes ne respectent pas la contrainte.
@@
 
-Dans un deuxième temps, \jlv{nous voulons tester si les sujets ont bien perçu} \jls{nous voulons nous assurer que les sujets ont bien perçu} une différence d'agrément entre les i- et ni-scènes.
+Dans un deuxième temps, nous voulons nous assurer que les sujets ont bien perçu une différence d'agrément entre les i- et ni-scènes.
 Pour ce faire, nous observons l'agrément moyen de chaque sujet $\mathcal{A}_{sujet}$, calculé séparément, pour chaque type d'environnement.
@@
 L'étude montre qu'il n'y a pas de différences significatives entre les jugements globaux et les moyennes des jugements instantanés.} Pour en revenir à notre expérience, c'est comme si nos sujets avaient inconsciemment tenu compte de cette réalité perceptive lors de la simulation, en faisant porter le niveau sonore global par des sons courts et bien identifiés, \ie~les événements.}
+
 \jls{Il apparaît que ce sont les événements qui impactent le plus le niveau global des scènes, l'écart entre $L$ et $L(E)$ n'étant que de 1 $dB$ pour les i-scènes et les ni-scènes.
@@
 
-\jlv{Cette corrélation élevée, considérant l'ensemble des scènes, résulte du fait que les i-scènes ont tendance à être moins fortes que les ni-scènes, donnant ainsi l'illusion de prolonger la corrélation négative observée pour les ni-scènes.} \jls{Ce résultat, considérant l'ensemble des scènes, s'explique par le fait que les i-scènes ont tendance à être moins fortes que les ni-scènes, donnant ainsi l'illusion de prolonger la corrélation négative observée pour les ni-scènes.}
+Ce résultat, considérant l'ensemble des scènes, s'explique par le fait que les i-scènes ont tendance à être moins fortes que les ni-scènes, donnant ainsi l'illusion de prolonger la corrélation négative observée pour les ni-scènes.
 
@@
 
-\jlv{Les mêmes observations sont faites \jlc{répétitions} concernant $L(E)$ (\cf~Figure~\ref{fig:soundlevele}).
-Pour $L(T)$ (\cf~Figure~\ref{fig:soundlevelf}), bien que, à considérer l'ensemble des scènes, on observe une corrélation modérée, cela n'est pas vérifié quand on regarde séparément les i-scènes ($r=-0.33$, $p=0.05$) et les ni-scènes ($r=-0.00$, $p=0.99$).
-\jlc{Exit: Là encore} on peut penser que la corrélation négative observée pour l'ensemble des scènes est un artefact résultant du fait que le niveau des textures des i-scènes a tendance à être plus bas que celui des ni-scènes.
-Ainsi, si les événements sonores conservent une certaine capacité de prédiction de l'agrément pour les ni-scènes, le niveau des textures n'apporte, lui, que peu d'informations, quel que soit l'environnement.}
-\jls{Les mêmes constats sont faits concernant $L(E)$ (\cf~Figure~\ref{fig:soundlevele}).
+Les mêmes constats sont faits concernant $L(E)$ (\cf~Figure~\ref{fig:soundlevele}).
 Pour $L(T)$ (\cf~Figure~\ref{fig:soundlevelf}), la corrélation modérée observée sur les scènes considérées dans leur globalité n'apparaît plus sur les i-scènes et ni-scènes considérées de façon séparée (i-scènes: $r=-0.33$, $p=0.05$, ni-scènes: $r=-0.00$, $p=0.99$).
 On peut penser que la corrélation négative émergeant au niveau l'ensemble est un artefact résultant du fait que le niveau des textures des i-scènes a tendance à être plus bas que celui des ni-scènes.
-Ainsi, si les événements sonores conservent une certaine capacité de prédiction de l'agrément pour les ni-scènes, le niveau des textures n'apporte, lui, que peu d'informations, quel que soit l'environnement.}
+Ainsi, si les événements sonores conservent une certaine capacité de prédiction de l'agrément pour les ni-scènes, le niveau des textures n'apporte, lui, que peu d'informations, quel que soit l'environnement.
 
@@
 
-Le fait qu'aucun des \jlv{descripteurs globaux} \jlc{Sauf erreur, c'est la première fois que tu emploies le terme.
-Il fait référence à quoi ?} ne permette de caractériser l'agrément des i-scènes peut nous amener à penser que toutes les sources sonores ne contribuent pas de manière égale à la perception de l'agrément, mais que seules les caractéristiques de certaines d'entre elles ont une réelle influence.
+Le fait que $L$ ne permet pas de caractériser l'agrément des i-scènes peut nous amener à penser que toutes les sources sonores ne contribuent pas de manière égale à la perception de l'agrément, mais que seules le niveau de certaines d'entre elles a une réelle influence.
 Afin d'approfondir ce point, nous analysons, dans la section suivante, les scènes d'un point de vue sémantique, \ie~en nous intéressant à la nature des sources qui les composent.
@@
 Concernant les textures sonores, 5 marqueurs sont identifiés.
-Pour les i-scènes, il s'agit de classes faisant référence à des ambiances amorphes, calmes, \jlv{(\emph{cour-intérieur/parc}} \jlc{pourquoi un tiré entre cour et intérieur et un slash entre intérieur et parc ?
-intérieur est-il pris comme adjectif se rapportant à cour ou comme nom "int"érieur = maison/appartement ?} et \emph{parc}).
+Pour les i-scènes, il s'agit de classes faisant référence à des ambiances amorphes, calmes, (\emph{cour-intérieur} et \emph{parc}).
 Pour les ni-scènes, il s'agit, comme pour les événements, de classes faisant référence à des bruits de travaux (\emph{travaux} et \emph{véhicule de travaux}), ainsi que d'une classe faisant référence au trafic (\emph{carrefour}).
@@
         {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}      
-        \caption{\jlv{Dispersions} \jlc{dispersions au plurial ?
-Tu confirmes ?} des descripteurs structurels de niveaux sonores relatifs à la présence des marqueurs $L_m$ (a, d), $L(E)_m$ (b, e) et $L(T)_m$ (c, f), en fonction du type de scènes (a, b, c) et de l'agrément perçu $\mathcal{A}_{scene}$ de l'expérience 1.b (d, e, f).}\label{fig:soundlevelMarker}
+        \caption{Dispersions des descripteurs structurels de niveaux sonores relatifs à la présence des marqueurs $L_m$ (a, d), $L(E)_m$ (b, e) et $L(T)_m$ (c, f), en fonction du type de scènes (a, b, c) et de l'agrément perçu $\mathcal{A}_{scene}$ de l'expérience 1.b (d, e, f).}\label{fig:soundlevelMarker}
 \end{figure}
@@
 \begin{itemize}
-\item \emph{72 am-scènes}: les 72 scènes précédemment simulées, \jlv{incluant} \jlc{ici j'emploierais plutôt le mot "avec" qui induit l'abréviation "am".
-Tu n'est pas d'accord ?} les classes de marqueurs (am).
+\item \emph{72 am-scènes}: les 72 scènes précédemment simulées, avec les classes de marqueurs (am).
 Nous notons i/am-scènes, les 36 scènes idéales avec marqueurs, et ni/am-scènes les 36 scènes non-idéales avec marqueurs;
@@
 Nombre de am-scènes sont composées, en majorité, de samples de marqueurs.
-\jlv{Afin de pas abusivement dénaturer ces scènes} \jls{Afin de ne pas dénaturer abusivement ces scènes}, en créant notamment des temps de «~vide~», \ie~ne comprenant aucun sample, nous ne supprimons que les marqueurs des classes d'événements du premier niveau d'abstraction (\cf~Tableau~\ref{tab:markers}).
+Afin de ne pas dénaturer abusivement ces scènes, en créant notamment des temps de «~vide~», \ie~ne comprenant aucun sample, nous ne supprimons que les marqueurs des classes d'événements du premier niveau d'abstraction (\cf~Tableau~\ref{tab:markers}).
 Ces classes sont:
@@
 
-\jlc{Exit: L'expérience est prévue pour durer 1 heure. Les sujets ne connaissent pas la nature des scènes.}
+L'expérience est prévue pour durer 1 heure. Les sujets ne connaissent pas la nature des scènes.
 
 \subsubsection*{Dispositif expérimental}
-\jlc{La subsubsection Dispositif expérimental me paraît pouvoir être retirée.}
+
 Tous les sujets passent l'expérience sur des machines identiques.
@@
 \subsubsection*{Participants}
-\jlc{La subsubsection Participants me paraît pouvoir être retirée.}
+
 12 sujets (4 femmes) participent à l'expérience.
@@
 
-\jlv{De cette expérience, il apparaît que} \jls{Cette expérience fait apparaître que} la présence, dans une scène, des marqueurs relevés à l'expérience 1 impacte bien l'agrément perçu.
+Cette expérience fait apparaître que la présence, dans une scène, des marqueurs relevés à l'expérience 1 impacte bien l'agrément perçu.
 La suppression des ni-marqueurs a un effet bénéfique sur le ressenti, tandis que, plus surprenant, la suppression des i-marqueurs dégrade légèrement la qualité.

### 2017-03-21 lafaySimulatedScenePerception.tex
correction Kuwano

@@
 
-\jlv{Il apparaît que ce sont les événements qui impactent le plus le niveau global des scènes, l'écart entre $L$ et $L(E)$ n'étant que de 1 $dB$ pour les i-scènes et les ni-scènes.
+Il apparaît que ce sont les événements qui impactent le plus le niveau global des scènes, l'écart entre $L$ et $L(E)$ n'étant que de 1 $dB$ pour les i-scènes et les ni-scènes.
 Cette observation fait écho aux résultats obtenus par Kuwano~\al~\cite{kuwano_memory_2003}.
-\jlc{Exit: Au cours de leur expérience, les auteurs demandent à leurs sujets d'évaluer une série d'environnements sonores de manière globale, dans un premier temps, puis d'en évaluer le niveau aux instants où chacun identifie une source sonore.
-L'étude montre qu'il n'y a pas de différences significatives entre les jugements globaux et les moyennes des jugements instantanés.} Pour en revenir à notre expérience, c'est comme si nos sujets avaient inconsciemment tenu compte de cette réalité perceptive lors de la simulation, en faisant porter le niveau sonore global par des sons courts et bien identifiés, \ie~les événements.}
-
-\jls{Il apparaît que ce sont les événements qui impactent le plus le niveau global des scènes, l'écart entre $L$ et $L(E)$ n'étant que de 1 $dB$ pour les i-scènes et les ni-scènes.
-C'est comme si nos sujets avaient inconsciemment tenu compte de cette réalité perceptive lors de la simulation, en faisant porter le niveau sonore global par des sons courts et bien identifiés, \ie~les événements.
-\jlc{dans l'enchainement de ces deux phrases il se peut que voulant élaguer j'ai un peu perdu du sens.
-Est-ce que le "C'est comme si..." constitue bien un enchaînement logique entre les deux phrases ?
-Est-ce que "cette réalité perceptive" évoquée dans la seconde est effectivement exprimée dans la seconde ?} Cette observation fait écho aux résultats obtenus par Kuwano~\al~\cite{kuwano_memory_2003}}.
+Au cours de leur expérience, les auteurs demandent à leurs sujets d'évaluer une série d'environnements sonores de manière globale, dans un premier temps, puis d'en évaluer le niveau aux instants où chacun identifie une source sonore.
+L'étude montre qu'il n'y a pas de différences significatives entre les jugements globaux et les moyennes des jugements instantanés.
+Pour en revenir à notre expérience, c'est comme si nos sujets avaient inconsciemment tenu compte de cette réalité perceptive lors de la simulation, en faisant porter le niveau sonore global par des sons courts et bien identifiés, \ie~les événements.
 

### 2017-03-23 lafaySimulatedScenePerception.tex
fixing intro

@@
 
-Si on demande à un sujet d'évaluer un environnement donné suivant une échelle perceptive donnée (\eg~clame, agrément) \cite{axelsson2005soundscape,davies2013perception,cain2013development}, le gain d'information est conditionné par la nature des stimuli disponibles, qu'il s'agisse de scènes sonores enregistrées (expérience en laboratoire), ou réelles (expérience \emph{in situ}).
+Si on demande à un sujet d'évaluer un environnement donné suivant une échelle perceptive donnée (\eg~calme, agrément) \cite{axelsson2005soundscape,davies2013perception,cain2013development}, le gain d'information est conditionné par la nature des stimuli disponibles, qu'il s'agisse de scènes sonores enregistrées (expérience en laboratoire), ou réelles (expérience \emph{in situ}).
 Cependant, le fait que ces stimuli existent permet d'en établir une description physique précise.
@@
 Si, à l'inverse, on demande à un sujet de décrire un environnement donné (\eg~décrivez une scène sonore urbaine «~calme~») \cite{guastavino2006ideal, dubois2006cognitive}, on collecte une information riche, symbolique et sémantique, de la représentation qu'il se fait de cet environnement.
-Mais, en l'absence de données sonores, on ne peut la caractériser physiquement.
+Néanmoins, en l'absence de données sonores, on ne peut la caractériser physiquement.
 
@@
 
-Nous pensons que ces travaux peuvent grandement bénéficier de l'utilisation de scènes simulées, scènes dont les caractéristiques structurelles, en particulier la nature des sources présentes, leurs positions ainsi que leurs signaux séparés, sont directement accessibles par l'expérimentateur.
-
-Afin de répondre à ces problématiques, nous proposons \emph{Simscene}, un outil de simulation permettant à un sujet de simuler des paysages sonores à partir d'un corpus de sons enregistrés.
-Pour montrer l'apport de \emph{Simscene}, nous choisissons, comme cadre applicatif, le problème de l'agrément perçu dans les environnements sonores urbains.
+Nous pensons que ces efforts de recherche peuvent grandement bénéficier de l'utilisation de scènes simulées, scènes dont les caractéristiques structurelles, en particulier la nature des sources présentes, leur répartition temporelle ainsi que leur forme d'onde respectives, sont directement accessibles par l'expérimentateur.
+
+Afin de répondre à ces problématiques, nous proposons d'utiliser un outil de simulation permettant à un sujet de simuler des paysages sonores à partir d'un corpus de sons enregistrés.
+Pour montrer l'intérêt de l'utilisation d'un tel outil pour questionnner la perception de l'environnement sonore, nous choisissons, comme cadre applicatif, le problème de l'agrément perçu dans les environnements sonores urbains.
 Nous présentons les résultats d'une série d'expériences s'appuyant sur la simulation, et visant, chacune, à comprendre comment les différentes sources sonores qui composent une scène influent sur la perception de l'agrément:
@@
 
-A notre connaissance, seuls Bruce~\al \cite{bruce2009development,bruce2014effects} se sont servis de la simulation.
+A notre connaissance, seuls Bruce~\al \cite{bruce2009development,bruce2014effects} se sont servis de la simulation dans une apporche similaire.
 Ils proposent un outil permettant d'agir sur un environnement en ajoutant ou supprimant des sources sonores spécifiques d'une part, en modifiant le niveau sonore des sources, et leurs positions spatiales d'autre part.
@@
 
-Dans notre démarche, nous faisons le choix d'une restitution simplifiée (monophonique) des scènes, au profit d'un nombre plus important de paramètres et de sources disponibles, ce afin de garantir, à la sortie, des données viables et expressives.
+Dans notre démarche, le simulateur utilisé \foonote{simScene, disponible en ligne \url{http://soundthings.org/research/}} permet une restitution simplifiée (monophonique) des scènes, au profit d'un nombre plus important de paramètres et de sources disponibles, ce afin de garantir, à la sortie, des données viables et expressives.
 
@@
 
-Dans le cadre de cette communication, nous ne présentons de \emph{SimScene} que les fonctionnalités intéressant notre étude.
+Dans le cadre de cette communication, nous ne présentons de \emph{SimScene} que les fonctionnalités pertienentes pour notre étude.
 Pour plus de détails, se référer à \cite{rossignol2015simscene,lafay2016JAES}.
-Dernière précision, l'expérience 1 .a a fait l'objet d'une étude pilote dont les résultats sont présentés dans \cite{lafay2013atiam,lafay2014new}.
+L'expérience 1 .a a fait l'objet d'une étude pilote dont les résultats sont présentés dans \cite{lafay2013atiam,lafay2014new}.
 
@@
 
-\emph{Simscene} tient compte de cette distinction, et utilise deux banques de données : l'une regroupant classes d’événements, l'autre des classes de textures.
+L'outil de simulation \emph{Simscene} tient compte de cette distinction, et utilise deux banques de données distinctes : l'une regroupant classes d’événements, l'autre des classes de textures.
 Événements et textures bénéficient de processus de simulation particuliers.
@@
 Pour les événements, nous considérons quatre niveaux d'abstraction allant des classes les plus globalisantes (niveau d'abstraction 0), aux classes les plus spécifiques (niveau d'abstraction 3).
-Pour les textures, nous ne considérons que trois niveaux d'abstraction .
-Notons que la taxonomie des événements est assez proche de celle de \cite{Salamon14}, introduite ultérieurement à notre étude.
+Pour les textures, nous ne considérons que trois niveaux d'abstraction.
+Notons que la taxonomie des événements est proche de celle de \cite{Salamon14}, introduite postérieurement à notre étude.
 
@@
 
-L'outil de simulation n'autorise pas d’interagir avec un sample en particulier, mais toujours avec une piste, \ie~une séquence de samples.
+Par choix de conception, l'outil de simulation ne permet pas d’interagir avec un sample en particulier, mais toujours avec une piste, \ie~une séquence de samples.
 Plusieurs paramètres permettent au sujet de contrôler la structure de chaque piste.
@@
 
-Chaque classe du dernier pallier possède un son prototype.
+Chaque classe du dernier palier possède un son prototype.
 Ces sons ont été choisis par les expérimentateurs.
@@
 La scène sonore est ainsi vue comme une somme de sources sonores, ou, si l'on admet la distinction opérée entre événements et textures sonores, «~un squelette d'événements sur un lit de textures~» \cite{nelken2013ear}.
-
-\emph{Simscene} est présenté en détail dans \cite{rossignol2015simscene}.
+Cet outil de simulation est présenté en détail dans \cite{rossignol2015simscene}.
 

### 2017-03-27 lafaySimulatedScenePerception.tex
final coorections

@@
 Le sujet peut très bien réunir 10 chiens dans un paysage sonore.
-Il n'en peut introduire 1 qui aboierait toutes les 10 millisecondes.
+Il ne peut en introduire 1 qui aboierait toutes les 10 millisecondes.
 \end{itemize}
@@
 \begin{enumerate}
-\item la réalisation de la simulation: cette étape cette étape se décompose en trois temps:
+\item la réalisation de la simulation: cette étape se décompose en trois temps:
 \begin{itemize}
@@
 Un résumé de ces derniers, ainsi que des acronymes les désignant, est présenté dans le Tableau~\ref{tab:acronyme}.
-Afin de rester cohérent avec l’expérience 1.b d'évaluation, les descripteurs ne sont pas calculés sur la durée totale des scènes, mais sur une version réduite de 30 secondes (\cf~Section~\ref{sec:xp1b_plan}).
+Afin de rester cohérent avec l’expérience 1.b d'évaluation, les descripteurs ne sont pas calculés sur la durée totale des scènes, mais sur la version réduite de 30 secondes (\cf~Section~\ref{sec:xp1b_plan}).
 
@@
 \item \emph{analyse des descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels et/ou l'agrément perçu, nous évaluons cette significativité à partir d'un test de Student à deux populations;
-\item \emph{étude de l'influence des descripteurs structurels sur l'agrément perçu}: afin d'évaluer l'impact potentiel des descripteurs structurels sur l'agrément perçu, nous étudions l'existence de corrélations linéaires entre ces descripteurs et $\mathcal{A}_{scene}$ .
+\item \emph{étude de l'influence des descripteurs structurels sur l'agrément perçu}: afin d'évaluer l'impact potentiel des descripteurs structurels sur l'agrément perçu, nous étudions l'existence de corrélations linéaires entre ces descripteurs et $\mathcal{A}_{scene}$.
 Pour mesurer la corrélation, nous utilisons le coefficient de Pearson.
@@
 L'utilisation de la distance de Hamming permet de prendre en compte de manière égale les classes présentes et absentes.
-Pour mesurer la capacité intrinsèque de l'espace à séparer les i- et ni-scènes, nous utilisons une métrique de \emph{clustering} nommée précision au rang $k$ ($P@k$).
+Pour mesurer la capacité intrinsèque de l'espace à séparer les i- et ni-scènes, nous utilisons une métrique d'\emph{ordre} nommée précision au rang $k$ ($P@k$).
 La $P@k$ mesure la précision obtenue après que $k$ items ont été retrouvés.
@@
 
-Dans un premier temps, et afin de garantir la cohérence de nos données, nous vérifions qu'aucune ni-scène n'ait un $\mathcal{A}_{scene}$ supérieur à celui d'une i-scène.
+Dans un premier temps, et afin de garantir la cohérence de nos données, nous vérifions qu'aucune ni-scène n'a un $\mathcal{A}_{scene}$ supérieur à celui d'une i-scène.
 Quatre des scènes ne respectent pas la contrainte.
@@
 \item le simulateur peut être déployé à large échelle via internet;
-\item les scènes simulées peuvent être analysées sans avoir à tenir compte des différentes langues maternelles des sujets, la nature sémantique des classes de sons utilisées étant connue \emph{a priori} par l'expérimentateur.
+\item les scènes simulées peuvent être analysées sans avoir à tenir compte des différentes langues maternelles des sujets, la nature sémantique des classes de sons utilisées étant connue \emph{a priori} par l'expérimentateur et sans avoir besoin d'analyser la scène sonore pour idientifier les sources, leurs présences dans la scène étant directement exploitable.
 \end{itemize}

### 2017-03-27 lafaySimulatedScenePerception.tex
rename label

@@
 
-Dans notre démarche, le simulateur utilisé \foonote{simScene, disponible en ligne \url{http://soundthings.org/research/}} permet une restitution simplifiée (monophonique) des scènes, au profit d'un nombre plus important de paramètres et de sources disponibles, ce afin de garantir, à la sortie, des données viables et expressives.
+Dans notre démarche, le simulateur utilisé \footnote{simScene, disponible en ligne \url{http://soundthings.org/research/}} permet une restitution simplifiée (monophonique) des scènes, au profit d'un nombre plus important de paramètres et de sources disponibles, ce afin de garantir, à la sortie, des données viables et expressives.
 
@@
         \subfloat[sujet 11]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_11}\label{fig:xp4_note_1h}}
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_11}\label{fig:xp4_note_1z}}
         \subfloat[sujet 12]

### 2017-04-18 vf/lafaySimulatedScenePerception.tex
adressing nm comments`

--- 2017-03-27 lafaySimulatedScenePerception.tex
@@
 
+\graphicspath{{../}}
+
 \allsectionsfont{\bfseries}
@@
 \newcommand{\gl}[1]{\textcolor{blue}{GL : #1}}
-\newcommand{\jlv}[1]{\textcolor{magenta}{#1}}
-\newcommand{\jls}[1]{\textcolor{red}{#1}}
+\newcommand{\nm}[1]{\textcolor{magenta}{#1}}
+\newcommand{\ml}[1]{\textcolor{red}{#1}}
 \newcommand{\jlc}[1]{\textcolor{green}{#1}}
@@
 This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes.
-A rating experiment of the soundscape pleasantness using a 7-points bipolar scale is conducted to further refine the analysis of the simulated urban acoustic scenes.
+A rating experiment of the soundscape pleasantness using a 7-points Likert scale is conducted to further refine the analysis of the simulated urban acoustic scenes.
 Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.}
@@
 
-Si, à l'inverse, on demande à un sujet de décrire un environnement donné (\eg~décrivez une scène sonore urbaine «~calme~») \cite{guastavino2006ideal, dubois2006cognitive}, on collecte une information riche, symbolique et sémantique, de la représentation qu'il se fait de cet environnement.
+Si, à l'inverse, on demande à un sujet de décrire un environnement donné (\eg~décrivez une scène sonore urbaine «~calme~») \cite{guastavino2006ideal, dubois2006cognitive}, on collecte une information riche, à la fois quantitative et sémantique, de la représentation qu'il se fait de cet environnement.
 Néanmoins, en l'absence de données sonores, on ne peut la caractériser physiquement.
 
-Nous pensons que la simulation permet de faire un lien élégant entre ces deux approches.
-Demander au sujet de simuler un environnement sonore donné revient à obtenir de lui une version modale, \ie~établie sur des dimensions physiques (le signal de la scène simulée), de sa représentation mentale de l'objet, des scènes caractérisées à la fois sémantiquement (\eg~sources sonores présentes) et physiquement (\eg~niveaux des sources sonores).
+Nous pensons que la simulation permet de faire un lien élégant et efficace entre ces deux approches.
+Il permet en effet d'obtenir du sujet une version modale, \ie~établie sur des dimensions physiques (le signal de la scène simulée), de sa représentation mentale de l'objet; les scènes générées étant à la fois caractérisées sémantiquement (\eg~sources sonores présentes) et physiquement (\eg~niveaux des sources sonores).
 
@@
 szeremeta2009analysis}.
-Les recherches portent désormais sur l'étude des contributions spécifiques des différentes sources sur les qualités affectives de la scène \cite{gozalo2015relationship,ricciardi2015sound}.
-Ces recherches requièrent des outils permettant d'analyser séparément les influences des éléments sonores qui composent l'environnement.
+Les recherches portent désormais sur l'étude des contributions spécifiques des différentes sources sur la notion de qualité affective de la scène \cite{gozalo2015relationship,ricciardi2015sound}.
+Pour de nombreuses raisons qui seront explicités tout au long de cet article, ces recherches peuvent tirés profit de ce type d'outil qui à l'avantage de permettre d'analyser séparément les influences des éléments sonores qui composent l'environnement.
 
@@
 \begin{enumerate}
-\item \emph{expérience 1.a, simulation}:  les sujets simulent les environnements qui serviront de stimuli pour les étapes suivantes;
-\item \emph{expérience 1.b, évaluation de l'agrément}: les sujets doivent évaluer l'agrément des scènes simulées à partir d'une échelle sémantique;
-\item \emph{expérience 2, évaluation de l'agrément après modification des scènes}: comme pour l'expérience précédente, les sujets doivent évaluer l'agrément des scènes simulées à partir d'une échelle sémantique.
+\item \emph{expérience 1.a, simulation}:  les sujets simulent les environnements idéaux/non idéaux qui serviront de stimuli pour les étapes suivantes;
+\item \emph{expérience 1.b, évaluation de l'agrément}: les sujets doivent évaluer l'agrément des scènes simulées sur une échelle sémantique;
+\item \emph{expérience 2, évaluation de l'agrément après modification des scènes}: comme pour l'expérience précédente, les sujets doivent évaluer l'agrément des scènes simulées sur une échelle sémantique.
 Cependant les scènes ont été modifiées, \ie~privées de certaines classes de sons identifiées comme ayant un impact sur l'agrément perçu.
@@
 Ils proposent un outil permettant d'agir sur un environnement en ajoutant ou supprimant des sources sonores spécifiques d'une part, en modifiant le niveau sonore des sources, et leurs positions spatiales d'autre part.
-Leurs sujets recréent un un environnement urbain en manipulant un jeu de sources sonores pré-établies.
+Leurs sujets recréent un un environnement urbain en manipulant un jeu de sources sonores pré-établi.
 Les auteurs montrent que l'inclusion ou l'exclusion des sources dépendent plus de considérations sociales/sémantiques, que de leurs caractéristiques physiques.
@@
 Le fonctionnement de \emph{Simscene} se rapproche de celui d'un séquenceur audio.
-Chaque utilisateur choisit une classe de sons via l'interface de sélection (\cf~Section~\ref{sec:simscene_ssf}).
+Un utilisateur choisit une classe de sons via l'interface de sélection (\cf~Section~\ref{sec:simscene_ssf}).
 Une fois la classe de sons sélectionnée, une piste audio, liée à cette classe, est créée.
@@
 
-L'outil de simulation \emph{Simscene} tient compte de cette distinction, et utilise deux banques de données distinctes : l'une regroupant classes d’événements, l'autre des classes de textures.
-Événements et textures bénéficient de processus de simulation particuliers.
+L'outil de simulation \emph{Simscene} tient compte de cette distinction, et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures.
+Événements et textures bénéficient de processus de simulation spécifiques.
 
@@
 
-Lorsque des événements émergent d'un environnement sonore, le cerveau traite l'information des différentes sources de manière séparée.
+Lorsque des événements émergent d'un environnement sonore, le cerveau traite l'information des différentes sources de manière séparée \cite{bregman1994auditory}.
 Plusieurs flux auditifs sont ainsi générés, \ie~un pour chaque séquence d'événements émis par la même source \cite{carlyon2004brain}.
-Inversement, quand le cerveau ne parvient pas à isoler d'événements, les éléments constituants de la scène sont agglomérés dans un même flux.
+Inversement, quand le cerveau ne parvient pas à isoler des événements, les éléments constituants de la scène sont agglomérés dans un même flux.
 
@@
 Les séquences amorphes sont des environnements dont il est difficile d'isoler des éléments distincts («~brouhaha de rue~», «~brouhaha de trafic~»).
-Elles profitent, elles, d'une analyse holistique, effectuée à partir d'indicateurs acoustiques (subjectifs) globaux.
-Cette différence de traitement entre scènes événementielles et scènes amorphes est relevée également dans \cite{guastavino2006ideal,dubois2006cognitive}.
+Elles profitent, elles, d'une analyse holistique, effectuée à partir d'indicateurs acoustiques (objectifs) globaux.
+%Cette différence de traitement entre scènes événementielles et scènes amorphes est relevée également dans \cite{guastavino2006ideal,dubois2006cognitive}.
 
@@
 Nous établissons deux taxonomies, une pour les événements, une autre pour les textures (\cf~Annexe~\ref{app:taxonomie}).
-Pour les événements, nous considérons quatre niveaux d'abstraction allant des classes les plus globalisantes (niveau d'abstraction 0), aux classes les plus spécifiques (niveau d'abstraction 3).
+Pour les événements, nous considérons quatre niveaux d'abstraction allant des classes les plus générales (niveau d'abstraction 0), aux classes les plus spécifiques (niveau d'abstraction 3).
 Pour les textures, nous ne considérons que trois niveaux d'abstraction.
@@
 Tous les enregistrements ont été effectués à l’aide d'un micro canon \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} relié à un enregistreur \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}}.
-L’utilisation du micro canon nous permet d’isoler les événements sonores du brouhaha urbain.
+L’utilisation du micro canon nous permet d’isoler les événements sonores du bruit de fond urbain.
 Pour les textures, il nous permet d’éviter les événements sonores proches du preneur de son.
-Nous pouvons ainsi pointer des "zones sonores", en nous tenant à une certaine distance de ces dernières, afin de capter uniquement le brouhaha émanant de la zone ciblée.
+Nous pouvons ainsi pointer des "zones sonores", en nous tenant à une certaine distance de ces dernières, afin de capter uniquement le bruit de fond émanant de la zone ciblée.
 
@@
 \begin{itemize}
-\item \emph{niveau sonore} ($dB$): pour chaque sample, les niveaux sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance;
+\item \emph{niveau sonore} ($dB$): pour chaque sample, les niveaux sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en termes de moyenne et de variance;
 \item \emph{espacement inter-onset} (seconde): (piste d'événements seulement) comme pour les niveaux, les espacements sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance;
@@
 
-La grande majorité des outils permettant de parcourir une banque de données propose une recherche textuelle sur la base de mots clefs.
+La grande majorité des outils permettant de parcourir une banque de données propose une recherche textuelle sur la base de mots-clé.
 Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique.
@@
 Tous les sujets passent l'expérience sur des machines identiques.
-L'audio est diffusé en monophonie, par le biais de casques audio.
+L'audio est diffusé en diotique, par le biais de casques audio.
 Pendant le tutoriel, les sujets doivent ajuster le niveau sonore à un volume confortable.
@@
 Les sujets sont âgés de 22 à 61 ans (moyenne: 29.5, écart-type: 14).
-Tous les sujets vivent dans un milieu urbain.
+Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
 
@@
 Cette expérience fait apparaître que la présence, dans une scène, des marqueurs relevés à l'expérience 1 impacte bien l'agrément perçu.
-La suppression des ni-marqueurs a un effet bénéfique sur le ressenti, tandis que, plus surprenant, la suppression des i-marqueurs dégrade légèrement la qualité.
+La suppression des ni-marqueurs a un effet bénéfique sur le ressenti des ni-scènes, tandis que, plus surprenant, la suppression des i-marqueurs dégrade légèrement la qualité des i-scènes.
 Ce dernier point est d'autant plus marquant que, du fait de la suppression des marqueurs, le niveau des i/am-scènes est supérieur à celui des i/sm-scènes.
@@
 
-Cependant, nous observons que les caractéristiques physiques corrélées à l'agrément diffèrent clairement suivant la nature hédonique des scènes.
+Cependant, nous observons que les caractéristiques physiques corrélées à l'agrément diffèrent clairement suivant le type de scènes.
 Dans le cas des scènes idéales, c'est avant tout l'émergence de marqueurs sonores qui détermine la qualité perçue, alors que dans le cas des scènes non-idéales, c'est le niveau sonore global qui influe sur l'agrément.
@@
 \begin{itemize}
-\item le simulateur peut être déployé à large échelle via internet;
+\item le simulateur peut être déployé à large échelle via internet du fait de l'architecture logicielle choisie pour son développement;
 \item les scènes simulées peuvent être analysées sans avoir à tenir compte des différentes langues maternelles des sujets, la nature sémantique des classes de sons utilisées étant connue \emph{a priori} par l'expérimentateur et sans avoir besoin d'analyser la scène sonore pour idientifier les sources, leurs présences dans la scène étant directement exploitable.
@@
 Bien évidemment, ces approches nécessitent d'accroître la taille des banques de sons isolés disponibles, un effort conséquent, mais nécessaire, qui contribuera grandement aux nombreux domaines de recherche ayant trait aux scènes sonores.
+
+\nm{a ce propos, il n'est pas fait mention des sons manquants pour les participants (cf. protocole expe de 1a);
+ne faudrait -il pas en dire 2 mots ce qui alimenterait aussi la discussion sur la dependance de la qualite (et l'efficacite) de la simulation a  la constitution de la base de donnees de textures et d'evenemetns ?}
 
@@
         \subfloat[sujet 8]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_8}\label{fig:xp4_note_1h}} \par
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_8}\label{}} \par
         \subfloat[sujet 9]

### 2017-04-18 lafaySimulatedScenePerception.tex
set new version

@@
 Cet outil de simulation est présenté en détail dans \cite{rossignol2015simscene}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%  Experience 1     %%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Expérience 1}
-\label{sec:xp1}
-
-\subsection{Objectif}
-
-L'objectif est d'étudier les influences spécifiques des différentes sources sonores qui composent les environnements urbains sur la perception de l'agrément, en utilisant la simulation.
-Pour ce faire, nous planifions nos deux premières expériences (\cf~Figure~\ref{fig:xp1_2}):
-
-\begin{itemize}
-\item \emph{expérience 1.a de simulation}: au cours de cette expérience, les sujets doivent simuler des environnements sonores urbains, en utilisant l'outil de simulation \emph{Simscene} (\cf~Section~\ref{sec:simscene}).
-Chacun compose deux environnements sonores, le premier idéal/agréable et le deuxième non-idéal/désagréable.
-
-\item \emph{expérience 1.b d'évaluation}: à l'issue de la simulation, nous n'avons, de fait, qu'une connaissance binaire des propriétés affectives des scènes simulées: idéale (i) et non-idéale (ni).
-Cette seconde étape a pour but d'affiner notre connaissance sur l'agrément.
-Pour ce faire, nous demandons à un deuxième groupe de sujets d'évaluer, à partir d'une échelle sémantique, l'agrément de chacune des scènes simulées.
-L'expérience d'évaluation sert deux buts:
-
-\begin{enumerate}
-\item évaluer l'influence respective des différentes sources sur l'agrément, pour chaque type d'environnement (i ou ni);
-\item détecter la présence de cas extrêmes ou ambigus (\emph{outlier}) dans les scènes simulées.
-Pour le reste de notre étude, les qualités hédoniques imposées (i et ni) servent de référence, de vérité terrain.
-Il nous faut donc garantir qu'il n'y ait pas d’ambiguïté entre les cas extrêmes des i- et ni-scènes, \ie~que la note d'agrément la plus basse des i-scènes reste supérieure à la note la plus haute des ni-scènes.
-\end{enumerate}
-
-\end{itemize}
-
-Notre analyse s'appuie sur les données produites par les deux expériences.
-
-\begin{figure}[t]
-        \myfloatalign
-        \includegraphics[width=\linewidth]{gfx/ch_5/5}
-        \caption{Planification expérimentale des éxperiences de simulation et d'évaluation de l'agrément}\label{fig:xp1_2}
-\end{figure}
-
-\subsection{Planification de l'expérience 1.a}
-\label{sec:xp1a_plan}
-
-\subsubsection*{Procédure}
-
-Les sujets doivent simuler deux environnements sonores urbains, chacune des scènes devant durer 1 minute. Ils doivent se conformer aux consignes suivantes:
-
-\begin{itemize}
-\item première simulation: simuler un paysage sonore \textbf{urbain plausible} qui, selon vous, est idéal (où vous aimeriez vivre);
-\item deuxième simulation: simuler un paysage sonore \textbf{urbain plausible} qui, selon vous, est non-idéal (où vous n'aimeriez pas vivre).
-\end{itemize}
-
-Tous commencent par simuler l'environnement idéal. Ils ne prennent connaissance de la deuxième consigne qu'à la fin de la première simulation.
-
-Les sujets sont totalement libres dans le choix des sons, et des paramètres. Ils doivent cependant se soumettre à deux contraintes:
-
-\begin{itemize}
-\item le sujet doit prendre le point de vue d’un auditeur fixe;
-\item le paysage sonore doit être réaliste, au sens de physiquement plausible.
-Le sujet peut très bien réunir 10 chiens dans un paysage sonore.
-Il ne peut en introduire 1 qui aboierait toutes les 10 millisecondes.
-\end{itemize}
-
-Ces contraintes font partie de la consigne. Aucun contrôle n'est fait \emph{a priori} dans l'interface de simulation.
-
-Chaque processus de simulation comprend deux parties :
-
-\begin{enumerate}
-\item la réalisation de la simulation: cette étape se décompose en trois temps:
-\begin{itemize}
-\item sélectionner les classes de sons
-\item nommer les classes de sons sélectionnées
-\item paramétrer les pistes (\cf~Section~\ref{sec:simscene_parametre}) relatives aux classes de sons sélectionnées
-\end{itemize}
-\item la production d'un commentaire libre du paysage sonore simulé
-\end{enumerate}
-
-En complément, et une fois les deux scènes sonores réalisées, les sujets sont invités à:
-
-\begin{itemize}
-\item indiquer les sources sonores qu'ils voulaient mettre, mais qu'ils n'ont pas trouvées;
-\item commenter l’ergonomie du logiciel de simulation;
-\item commenter l’ergonomie de l'interface de sélection.
-\end{itemize}
-
-Avant de commencer la première simulation, un tutoriel de 20 minutes est proposé aux sujets, afin qu'ils se familiarisent avec le logiciel de simulation, et la banque de données.
-L'expérience est prévue pour durer 2h30.
-
-\subsubsection*{Dispositif expérimental}
-
-Tous les sujets passent l'expérience sur des machines identiques.
-L'audio est diffusé en monophonie, par le biais de casques audio.
-Pendant le tutoriel, les sujets doivent ajuster le niveau sonore à un volume confortable.
-Ils ne peuvent le modifier par la suite.
-
-Tous les sujets réalisent l'expérience simultanément.
-Ils sont répartis de manière égale dans trois pièces identiques, toutes possédant un environnement calme.
-Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
-
-Trois expérimentateurs, un dans chaque pièce, sont présents durant la totalité de l'expérience, afin de contrôler le bon déroulement de cette dernière, et de répondre aux éventuelles questions des sujets.
-
-\subsubsection*{Participants}
-
-44 étudiants (14 femmes) de L’École Centrale de Nantes ont participé à l'expérience.
-Ils ont tous sensiblement le même âge (moyenne: 21.6, écart-type: 2).
-Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
-
-Sur les 44 sujets, 40 réalisent l'expérience avec succès, produisant au final 80 scènes sonores simulées, dont 40 scènes idéales, et 40 scènes non idéales.
-4 sont éliminés pour non respect ou incompréhension des consignes, d'une part, et dépassement du temps, d'autre part.
-
-\subsection{Planification  de l'expérience 1.b}
-\label{sec:xp1b_plan}
-
-\subsubsection*{Procédure}
-
-En raison de contraintes temporelles, les sujets n'évaluent que des séquences de 30 secondes des scènes simulées, chacune de ces séquences commençant à la seconde 15, et finissant à le seconde 45 de la scène d'une durée initiale de 1 minute.
-
-L'évaluation s'effectue sur une échelle sémantique bipolaire de 7 points, allant de -3 (non-idéale/très désagréable) à +3 (idéale/très agréable).
-Avant de noter une scène, les sujets doivent obligatoirement écouter les 20 premières secondes de cette dernière.
-Après la notation, ils sont libres de passer à la scène suivante.
-
-Pour chaque sujet, les scènes sont présentées dans un ordre aléatoire.
-Les 10 premières scènes permettent au sujet de calibrer ses notes.
-Elles sont obligatoirement composées de 5 scènes idéales, et de 5 scènes non-idéales.
-Ces 10 premières scènes sont rejouées à la fin de l'expérience, et seules les notes données à la deuxième occurrence sont prises en compte.
-
-L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.
-
-\subsubsection*{Dispositif expérimental}
-
-Tous les sujets passent l'expérience sur des machines identiques.
-L'audio est diffusé en monophonie, par le biais de casques audio semi-ouverts \emph{Beyer-Dynamic DT 990 Pro}.
-Toutes les scènes sonores ont été re-simulées sur la base des partitions obtenues lors de l'expérience de simulation.
-Le niveau sonore de sortie est identique pour tous les sujets.
-
-Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
-
-Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de celle-ci, et de répondre aux éventuelles questions des sujets.
-
-\subsubsection*{Participants}
-
-10 étudiants (2 femmes) de L’École Centrale de Nantes ont participé à l'expérience.
-Aucun d'entre eux n'a réalisé l'expérience de simulation.
-Tous les sujets ont sensiblement le même âge (moyenne: 23.1, écart-type: 1.8).
-Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
-
-Tous les sujets ont réalisé l'expérience avec succès.
-
-\subsection{Données et méthodes d'analyses}
-\label{sec:xp1_dataAna}
-
-A chaque scène est associé un groupe de descripteurs.
-C'est sur ces descripteurs que nous pratiquons l'analyse.
-Un résumé de ces derniers, ainsi que des acronymes les désignant, est présenté dans le Tableau~\ref{tab:acronyme}.
-Afin de rester cohérent avec l’expérience 1.b d'évaluation, les descripteurs ne sont pas calculés sur la durée totale des scènes, mais sur la version réduite de 30 secondes (\cf~Section~\ref{sec:xp1b_plan}).
-
-Pour chaque scène sonore, trois types de descripteurs sont considérés:
-
-\begin{itemize}
-\item \emph{perceptif}: il s'agit de l'agrément perçu de la scène simulée, évalué sur une échelle sémantique de 7 points.
-Nous notons $\mathcal{A}_{scene}$ l'agrément moyen d'une scène, obtenu en moyennant les notes de tous les sujets.
-De même, nous notons $\mathcal{A}_{sujet}$ l'agrément par sujet, en moyennant l'ensemble des notes de chacun.
-Compte tenu du faible nombre de sujets, nous faisons le choix, dans cette étude, de ne pas normaliser les notes d'agrément;
-\item \emph{sémantique}: il s'agit d'un vecteur booléen noté $S=(x_1,x_2,\ldots,x_n)$, indiquant les classes de sons présentes dans la scène.
-Chaque point $x$ de ce vecteur correspond à une classe de sons particulière: $x=1$ si la classe est présente dans la scène, et $x=0$ autrement.
-La dimension $n$ des vecteurs dépend du niveau d'abstraction considéré, \eg~pour le niveau d'abstraction 1, qui comprend $44$ classes de sons, cette dimension sera de $n=44$.
-\item \emph{structurel}: il s'agit du \emph{niveau Sonore}.
-Pour représenter le niveau sonore, nous nous inspirons de la mesure $L_{Aeq}$.
-Dans notre cas, il s'agit d'un scalaire calculé sur le signal en volts, et non en pression, et donné en décibels, en prenant un référentiel de 1 Volt.
-Le niveau est obtenu en calculant, toutes les secondes, la moyenne quadratique du signal, et en moyennant sur la durée de la scène.
-Un filtrage de type A est opéré avant le calcul des moyennes quadratiques.
-Nous notons $L$, $L(E)$ et $L(T)$ les niveaux calculés en tenant compte respectivement de l'ensemble des samples, des samples d'événements et de textures.
-\end{itemize}
-
-Pour évaluer l'impact spécifique des différentes sources sonores sur l'agrément perçu, nous soumettons nos travaux aux 5 tests/études de significativité présentés ci-après:
-
-\begin{itemize}
-\item \emph{vérification de l'agrément des scènes simulées}: afin de vérifier que la distinction affective imposée entre les i- et ni-scènes se retrouve au niveau de l'agrément perçu, nous observons s'il existe des différences entre les deux types de scènes au niveau de $\mathcal{A}_{scene}$ et $\mathcal{A}_{sujet}$.
-La significativité est évaluée par un test de Student à deux populations indépendantes pour $\mathcal{A}_{scene}$, et par un test de Student à deux populations appariées pour $\mathcal{A}_{sujet}$;
-\item \emph{analyse des descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels et/ou l'agrément perçu, nous évaluons cette significativité à partir d'un test de Student à deux populations;
-\item \emph{étude de l'influence des descripteurs structurels sur l'agrément perçu}: afin d'évaluer l'impact potentiel des descripteurs structurels sur l'agrément perçu, nous étudions l'existence de corrélations linéaires entre ces descripteurs et $\mathcal{A}_{scene}$.
-Pour mesurer la corrélation, nous utilisons le coefficient de Pearson.
-Nous adoptons ici une méthodologie couramment utilisée dans l'approche dimensionnelle;
-\item \emph{analyse des descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée a un impact sur la composition des scènes en terme de sources sonores, s'il est des classes de sons plus particulièrement utilisées pour simuler un type d'environnement (i ou ni), nous utilisons le V-test.
-Le test est effectué pour chaque niveau d'abstraction, et séparément, pour les classes d'événements et de textures.
-Pour chaque classe $j$ et chaque type d'environnement $k$ ($k=\lbrace i,ni\rbrace$), la valeur $V_{jk}$ du V-test se calcule comme suit:
-
-\begin{equation*}
-V_{jk}=\dfrac{c_{jk}-c_k\frac{c_j}{c}}{\sqrt{c_k\frac{c-c_k}{c-1}\frac{c_j}{c}(1-\frac{c_j}{c})}}
-\end{equation*}
-
-où $c$ est le nombre de classes utilisées, $c_k$ le nombre de classes utilisées pour un type d'environnement $k$, $c_j$ le nombre de classes $j$ utilisées, et $c_{jk}$ le nombre de classes $j$ utilisées pour un type d'environnement $k$.
-Le V-test teste l'hypothèse nulle que la proportion $\frac{c_{jk}}{c}$ ne diffère pas significativement de la proportion $\frac{c_{jk}}{c_k}$.
-Si pour un environnement $k$, et une classe $j$, l'hypothèse est rejetée, la classe $j$ est alors typique de l'environnement $k$.
-Les classes typiques sont nommées \textbf{marqueurs sonores};
-
-\item \emph{étude des espaces de représentation induits par les descripteurs sémantiques}: afin d'étudier si une représentation basée uniquement sur la présence ou l'absence de classes de sons permet de séparer les deux types d'environnement, nous considérons l'espace induit par les descripteurs sémantiques $S$.
-$S$ étant un vecteur booléen, nous calculons les distances entre les scènes à partir de la distance de Hamming.
-Considérant les deux vecteurs $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, et $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ de dimension $n$, avec $x=\lbrace 0,1\rbrace$, la distance de Hamming $d_{ham}$ mesure le pourcentage de coordonnées qui diffèrent entre les deux vecteurs:
-
-\begin{equation*}
-d_{ham}(S_1,S_2)=\dfrac{1}{n}\sum_{i=1}^{n} (x_{1,i} \bigoplus x_{2,i})
-\end{equation*}
-
-où $\bigoplus$ désigne l'opérateur du \emph{ou-exclusif}.
-Plus la composition des deux scènes est similaire, et plus ces deux scènes sont proches.
-L'utilisation de la distance de Hamming permet de prendre en compte de manière égale les classes présentes et absentes.
-Pour mesurer la capacité intrinsèque de l'espace à séparer les i- et ni-scènes, nous utilisons une métrique d'\emph{ordre} nommée précision au rang $k$ ($P@k$).
-La $P@k$ mesure la précision obtenue après que $k$ items ont été retrouvés.
-Formellement, pour chaque scène $s_i$, nous calculons le rapport entre le nombre de scènes $s_j$ prises parmi les $k$ plus proches voisines de $s_i$, et partageant le même label que $s_i$, sur le nombre d'items à retrouver ($k$).
-La $P@k$ est alors la moyenne des rapports pour tous les items;
-
-\item \emph{influence des marqueurs sonores sur l'agrément perçu}: afin d'évaluer les contributions spécifiques de certaines sources sonores, nous évaluons une nouvelle fois l'impact potentiel des descripteurs structurels sur l'agrément perçu, mais en ne tenant compte, cette fois, que des marqueurs sonores pour calculer ces descripteurs.
-\end{itemize}
-
-Excepté le V-test, tous les tests de significativité sont effectués avec un seuil critique $\alpha=0.05$.
-Pour le V-test, étant donné que nous testons beaucoup de classes, une correction de Bonferroni est appliquée.
-Pour les valeurs $p$, dans le cas où $p\geq0.05$, nous indiquons sa valeur.
-Dans le cas où $0.01\leq p<0.05$, nous indiquons seulement $p<0.05$.
-Dans le dernier cas nous indiquons $p<0.01$.
-
-\begin{table}[t]
-\centering
-\begin{tabular}{c c}
-Termes                         & Acronymes              \\
-\hline
-Niveau                         & $L$                    \\
-Niveau (événements)            & $L(E)$                 \\
-Niveau (textures)              & $L(T)$                 \\
-Agrément moyen (par scène)     & $\mathcal{A}_{scene}$  \\
-Agrément moyen (par sujet)     & $\mathcal{A}_{sujet}$  \\
-                               &                        \\
-idéale/agréable                & i                      \\
-non-idéale/désagréable         & ni                     \\
-scène idéale/agréable          & i-scène                \\
-scène non-idéale/désagréable   & ni-scène               \\
-\hline
-\end{tabular}
-\vspace{0.5mm}
-\caption{Acronyme des variables utilisées dans le cadre des expériences sensorielles.}
-\label{tab:acronyme}
-\end{table}
-
-\subsection{Résultats}
-
-\subsubsection*{Vérification de l'agrément des scènes simulées}
-
-Dans un premier temps, et afin de garantir la cohérence de nos données, nous vérifions qu'aucune ni-scène n'a un $\mathcal{A}_{scene}$ supérieur à celui d'une i-scène.
-Quatre des scènes ne respectent pas la contrainte.
-Elles et leurs correspondantes i ou ni sont retirées.
-36 i-scènes et 36 ni-scènes restent donc dans le champ de l'analyse.
-
-Dans un deuxième temps, nous voulons nous assurer que les sujets ont bien perçu une différence d'agrément entre les i- et ni-scènes.
-Pour ce faire, nous observons l'agrément moyen de chaque sujet $\mathcal{A}_{sujet}$, calculé séparément, pour chaque type d'environnement.
-Il apparaît que les i-scènes ont bien été perçues comme significativement plus agréables ($p<0.01$) que les ni-scènes.
-
-\subsubsection*{Analyse des descripteurs structurels}
-
-En premier lieu, nous nous concentrons sur le niveau sonore.
-Les figures~\ref{fig:soundlevela},~\ref{fig:soundlevelb} et~\ref{fig:soundlevelc} affichent les distributions des niveaux $L$, $L(E)$ et $L(T)$.
-Il existe bien une différence de niveau significative entre les i- et ni-scènes ($L$: $p<0.01$), avec un écart des moyennes de -7 $dB$.
-Cette différence affecte aussi bien les événements ($L(E)$: $p<0.01$, écart moyen: -7 $dB$) que les textures ($L(T)$: $p<0.01$, écart moyen: -6 $dB$).
-
-Nous vérifions, sans surprise, que le niveau des sources sonores est bien un indicateur d'agrément, les ni-scènes ayant tendance à être plus fortes, fait rapporté dans un grand nombre d'études.
-Nous constatons encore que cette différence de niveau s'observe de manière égale pour les événements et les textures sonores.
-
-Il apparaît que ce sont les événements qui impactent le plus le niveau global des scènes, l'écart entre $L$ et $L(E)$ n'étant que de 1 $dB$ pour les i-scènes et les ni-scènes.
-Cette observation fait écho aux résultats obtenus par Kuwano~\al~\cite{kuwano_memory_2003}.
-Au cours de leur expérience, les auteurs demandent à leurs sujets d'évaluer une série d'environnements sonores de manière globale, dans un premier temps, puis d'en évaluer le niveau aux instants où chacun identifie une source sonore.
-L'étude montre qu'il n'y a pas de différences significatives entre les jugements globaux et les moyennes des jugements instantanés.
-Pour en revenir à notre expérience, c'est comme si nos sujets avaient inconsciemment tenu compte de cette réalité perceptive lors de la simulation, en faisant porter le niveau sonore global par des sons courts et bien identifiés, \ie~les événements.
-
-Nous observons, enfin, que le niveau seul ne permet pas de clairement faire la distinction entre les différents types d'environnement.
-En effet, $20\%$ des i-scènes ont un niveau supérieur au niveau minimal des ni-scènes, alors qu'il n'y a pas de recouvrement, si l'on considère l'agrément perçu $\mathcal{A}_{scene}$.
-
-\begin{figure}[t]
-        \myfloatalign
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_1}\label{fig:soundlevela}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_3}\label{fig:soundlevelb}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_5}\label{fig:soundlevelc}}\par
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_2}\label{fig:soundleveld}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_4}\label{fig:soundlevele}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_6}\label{fig:soundlevelf}}
-       \caption{Dispersions des descripteurs structurels $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), en fonction du type de scènes (a, b, c) et de l'agrément perçu $\mathcal{A}_{scene}$ de l'expérience 1.b (d, e, f).}
-\end{figure}
-
-\subsubsection*{Influence des descripteurs structurels sur l'agrément perçu}
-
-Dans cette partie, nous analysons les relations fines qui peuvent exister entre les descripteurs structurels, d'une part, et l'agrément perçu, d'autre part.
-Contrairement à la section précédente, où la qualité affective des scènes est représentée de manière binaire (i \vs~ni), nous considérons ici l'agrément moyen $\mathcal{A}_{scene}$ comme descripteur perceptif.
-Il s'agit d'étudier l'existence de potentielles corrélations entre les descripteurs structurels et $\mathcal{A}_{scene}$.
-Les coefficients de corrélations linéaires calculés entre $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ sont présentés dans le tableau~\ref{tab:corrStructA}.
-Les relations entre $\mathcal{A}_{scene}$ et les descripteurs structurels sont illustrées par les figures~\ref{fig:soundleveld},~\ref{fig:soundlevele} et ~\ref{fig:soundlevelf}.
-
-Concernant $L$, on observe une forte corrélation négative ($r=-0.77$, $p<0.01$) avec $\mathcal{A}_{scene}$, indiquant que plus le niveau sonore est élevé, plus la scène est désagréable.
-Cependant, la figure~\ref{fig:soundleveld} suggère que cette relation ne s'opère pas de la même manière pour les i- et ni-scènes.
-En effet, la corrélation entre $L$ et $\mathcal{A}_{scene}$, pour les ni-scènes, reste élevée ($r=-0.78$, $p<0.01$), mais est inexistante pour les i-scènes.
-
-Ce résultat, considérant l'ensemble des scènes, s'explique par le fait que les i-scènes ont tendance à être moins fortes que les ni-scènes, donnant ainsi l'illusion de prolonger la corrélation négative observée pour les ni-scènes.
-
-Nous en concluons que $L$:
-
-\begin{itemize}
-\item permet bien de faire la distinction entre les i- et ni-scènes,
-\item permet de finement caractériser l'agrément perçu des ni-scènes,
-\item n'est pas un indicateur pertinent de l'agrément perçu pour des environnements a priori agréables.
-\end{itemize}
-
-Les mêmes constats sont faits concernant $L(E)$ (\cf~Figure~\ref{fig:soundlevele}).
-Pour $L(T)$ (\cf~Figure~\ref{fig:soundlevelf}), la corrélation modérée observée sur les scènes considérées dans leur globalité n'apparaît plus sur les i-scènes et ni-scènes considérées de façon séparée (i-scènes: $r=-0.33$, $p=0.05$, ni-scènes: $r=-0.00$, $p=0.99$).
-On peut penser que la corrélation négative émergeant au niveau l'ensemble est un artefact résultant du fait que le niveau des textures des i-scènes a tendance à être plus bas que celui des ni-scènes.
-Ainsi, si les événements sonores conservent une certaine capacité de prédiction de l'agrément pour les ni-scènes, le niveau des textures n'apporte, lui, que peu d'informations, quel que soit l'environnement.
-
-En résumé, en présence d'un environnement désagréable, les niveaux sonores, en particulier ceux des événements, ont un impact négatif sur l'agrément.
-En présence d'un environnement agréable, en revanche, aucun des descripteurs structurels considérés ici ne semble influer sur la perception de l'agrément.
-
-Ces premiers résultats pourraient montrer qu'il existe deux modes de perception, mobilisant chacun des descripteurs indépendants, modes qui s'activent en fonction de la nature de l'environnement (i ou ni).
-
-Le fait que $L$ ne permet pas de caractériser l'agrément des i-scènes peut nous amener à penser que toutes les sources sonores ne contribuent pas de manière égale à la perception de l'agrément, mais que seules le niveau de certaines d'entre elles a une réelle influence.
-Afin d'approfondir ce point, nous analysons, dans la section suivante, les scènes d'un point de vue sémantique, \ie~en nous intéressant à la nature des sources qui les composent.
-
-\begin{table}[t]
-\centering
-\begin{tabular}{l c c c}
-               & i-scènes                   & ni-scènes    \\
-\hline
-$L$            & -0.32 ($p=0.06$)           & \textbf{-0.78} ($p<0.01$)\\
-$L(E)$         & -0.20 ($p=0.24$)           & \textbf{-0.75} ($p<0.01$)\\
-$L(T)$         & -0.33 ($p=0.05$)           &  -0.00 ($p=0.99$) \\
-\hline
-\end{tabular}
-\vspace{0.5mm}
-\caption{Coefficients de corrélation linéaire calculés entre l'agrément perçu moyen $\mathcal{A}_{scene}$ de l'expérience 1.b et les descripteurs structurels.}
-\label{tab:corrStructA}
-\end{table}
-
-\subsubsection*{Analyse des descripteurs sémantiques}
-
-Nous analysons la composition des scènes en comptant le nombre de sujets ayant utilisé une classe de sons pour simuler un type d'environnement.
-Les résultats sont présentés à la figure~\ref{fig:soundsourcea} pour les événements, et à la figure~\ref{fig:soundsourceb} pour les textures.
-Par souci d'espace, nous choisissons un niveau d'abstraction intermédiaire entre les niveaux 0 et 1, noté $0+$, pour représenter les classes (\cf~Figure~\ref{fig:taxonomie}).
-
-Nous observons une différence notable dans le choix des classes entre les i- et ni-scènes.
-La répartition des classes est très proche de celle obtenue dans une étude similaire sur les environnements sonores urbains idéaux \cite{guastavino2006ideal}, \ie~les classes suggérant la présence humaine et la nature sont très présentes dans les i-scènes, a contrario, les classes désignant des sons mécaniques et/ou de travaux sont principalement utilisées pour les ni-scènes.
-
-Ces résultats confirment un fait déjà observé: la nature sémantique des sources sonores joue un rôle prédominant dans l'appréciation de l'environnement \cite{raimbault2005urban,dubois2006cognitive}.
-
-Nous notons quelques différences avec \cite{guastavino2006ideal}: les résultats obtenus par Guastavino montrent que les sons de \emph{transports publics} sont caractéristiques des environnements sonores urbains idéaux.
-Les auteurs attribuent cela au fait que la perception de l'agrément est, entre autre, soumise à un contexte socio-culturel.
-Dans notre représentation du monde, les sons de transports publics sont positivement connotés, et ont ainsi tendance à être mieux acceptés que les sons de véhicules privés.
-
-Dans une certaine mesure, nos résultats contredisent ce fait.
-La figure~\ref{fig:soundsourcea} montre, en effet, que les classes d'événements de \emph{transports publics} (\emph{bus} et \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) ont été utilisées par les sujets, pour des i-scènes, dans $28\%$ des cas, et pour des ni-scènes, dans $42\%$ des cas.
-Les résultats ne remettent pas en question le fait que les sons de \emph{transports publics} soient bien acceptés: $25\%$ des sujets ont utilisé la classe \emph{bus} pour les i-scènes, un chiffre comparable à celui de la classe \emph{Vélo}, et bien supérieur à celui de toute autre classe de véhicules privés.
-Cependant les classes \emph{transports publics} sont également bien présentes dans les ni-scènes, plus que les classes \emph{voiture} ou \emph{camion} par exemple.
-La classe \emph{transports publics} ne peut donc pas être considérée comme typique d'un environnement sonore urbain idéal.
-
-Cette différence peut s'expliquer par la nature des deux protocoles expérimentaux utilisés.
-Comme nous l'avons fait, Guastavino demande à ses sujets de décrire un environnement.
-Mais ces derniers travaillent de mémoire, alors que nos propres sujets disposent de supports sonores.
-Le fait que nos sujets soient confrontés à la réalité acoustique des sons, pour recréer leurs environnements, peut avoir pour effet de diminuer l'impact du contexte socio-culturel.
-D'autres études utilisant des sons comme stimuli montrent que la classe \emph{bus} peut avoir un effet négatif sur l'appréciation de l'environnement \cite{lavandier2006contribution}.
-
-\begin{figure}[t]
-        \myfloatalign
-        \subfloat[]
-        {\includegraphics[width=1\linewidth]{gfx/ch_5/xp1_class_1_cam}\label{fig:soundsourcea}} \par
-        \subfloat[]
-        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_2_cam}\label{fig:soundsourceb}}
-        \subfloat[]
-        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_3_cam}\label{fig:soundsourcec}}
-       \caption{Pourcentage de scènes simulées (i ou ni) comportant une classe de son particulière: (a) classes d'événements du niveau d'abstraction $0+$, (b) classes de textures du niveau d'abstraction $0$, (c) sous classes d'événements du niveau d'abstraction $1$ appartenant aux classes \emph{trafic} et \emph{transport public} du niveau d'abstraction $0$.}\label{fig:soundsource}
-\end{figure}
-
-\subsubsection*{Marqueurs sonores}
-
-Nous avons mis en évidence que, qualitativement, la composition des sources sonores des scènes diffère selon les types d'environnement (i ou ni).
-Nous essayons de voir maintenant si, parmi ces classes, certaines sont typiques d'un environnement en particulier.
-Pour ce faire, nous utilisons le V-test (\cf~Section~\ref{sec:xp1_dataAna}), en considérant séparément chaque niveau d'abstraction.
-Les résultats sont présentés dans le tableau~\ref{tab:markers}.
-
-Concernant les événements sonores, 9 marqueurs sont identifiés sur l'ensemble des niveaux d'abstraction.
-Comme la figure~\ref{fig:soundsource} le laissait présager, les classes relatives à la présence humaine (\emph{pas homme béton}, \emph{sonnette vélo}), et à la nature (\emph{animaux, oiseaux}, \emph{chants d'oiseaux}) sont des marqueurs de i-scènes.
-Nous notons également la présence de la classe \emph{cloche} dans les marqueurs d'un environnement idéal.
-Ce fait est possiblement dû au \emph{background} socio-culturel des sujets, dans leur grande majorité, des citoyens européens.
-En effet, selon Schafer, un son reconnu par un individu comme faisant partie intégrante de son environnement est bien accepté.
-Les marqueurs de ni-scènes sont des classes faisant référence à des sons de travaux (\emph{travaux}), ou suggérant un trafic dense (\emph{klaxon}, \emph{sirène}).
-
-Concernant les textures sonores, 5 marqueurs sont identifiés.
-Pour les i-scènes, il s'agit de classes faisant référence à des ambiances amorphes, calmes, (\emph{cour-intérieur} et \emph{parc}).
-Pour les ni-scènes, il s'agit, comme pour les événements, de classes faisant référence à des bruits de travaux (\emph{travaux} et \emph{véhicule de travaux}), ainsi que d'une classe faisant référence au trafic (\emph{carrefour}).
-
-Bien que l'ensemble des marqueurs identifiés soient intuitifs, aucune des classes d'événements faisant directement référence aux bruits de véhicules motorisés n'est un marqueur, exception faite de la classe de textures \emph{carrefour}.
-Pour représenter un trafic désagréable, les sujets ont porté leurs choix sur les classes \emph{klaxon} et \emph{sirène}.
-On peut supposer que les sons isolés de véhicules sont compris comme faisant partie intégrante de l'environnement urbain, et ne sont donc pas particulièrement associés à un environnement désagréable.
-
-\begin{table}[t]
- \setlength{\tabcolsep}{0.2pt}
- \centering
-  {\renewcommand{\arraystretch}{0.9}
-\begin{tabular}{c c c c}
-Niveau        & \multicolumn{2}{c}{Marqueurs sonores événements} \\
-d'abstraction & i-scènes & ni-scènes \\
-\hline
-0  &                          & travaux (3.78)  \\
-\hline
-  & cloche  (4.5)             & klaxon  (3.9) \\
-1 & sonnette vélo  (4.3)      & sirène (3.9)\\
-  & animal (4.2)              &       \\
-   \hline
-  & oiseau        (4.8)       & klaxon  (4.0)\\
-2 & cloche  (4.4)             & sirène (4.0)\\
-  & sonnette vélo    (4.2)             &       \\
-   \hline
-  & chant oiseau (4.8)        & klaxon  (4.1)\\
-3 & cloche   (4.3)            & sirène (4.0)\\
-  & sonnette vélo     (4.2)   &       \\
-  & pas chaussure  (3.6)      &  \\
-  &                           & \\
-  & \multicolumn{2}{c}{Marqueurs sonores textures}      \\
-  & i-scènes & ni-scènes \\
-\hline
-0 &     cour int./parc (4.1) &  travaux (3.9)  \\
-\hline
-1 &     parc (3.65)          &  carrefour (3.6)  \\
-  &                          &  travaux véhicule (3.3)  \\
-\hline
-2 &     parc (3.64)          &  carrefour (3.56)  \\
-\hline
-\end{tabular}
-}
-\vspace{0.5mm}
-\caption{Classes d'événements et de textures identifiées comme étant des marqueurs sonores. Dans chaque cellule, les marqueurs sont ordonnés par ordre décroissant de valeur $V$.}
-\label{tab:markers}
-\end{table}
-
-\subsubsection*{Espaces de représentation induits par les descripteurs sémantiques}
-
-Dans cette partie, nous évaluons la capacité d'une représentation sémantique à séparer les deux types d'environnement.
-Pour ce faire, nous calculons une précision au rang 5 ($p@5$) sur l'espace induit par les descripteurs sémantiques $S$, et ce pour chaque niveau d'abstraction (\cf~Section~\ref{sec:xp1_dataAna}).
-Les vecteurs $S$ sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.
-Nous ne considérons pas les classes de marqueurs de textures, ces dernières étant trop peu nombreuses.
-Pour les mêmes raisons nous ne considérons pas les classes de marqueurs d'événements du niveau d'abstraction $0$.
-Les résultats sont affichés sur la figure~\ref{fig:pa5}.
-
-En ce qui concerne $ET$, la $p@5$ est de $76\%$ pour le niveau d'abstraction 0, et reste supérieure à $86\%$ à partir du niveau d'abstraction 1.
-Ces résultats confirment qu'il est possible de clairement distinguer les deux types d'environnement en se basant seulement sur la présence ou l'absence des classes de sons.
-Nous notons également que, plus le niveau d'abstraction est élevé, plus la capacité de séparer les environnements est importante.
-En d'autres termes, plus nous sommes précis dans notre description de la composition des scènes, plus nous sommes à même d'établir une distinction claire entre les i- et ni-scènes.
-
-En considérant séparément $E$ et $T$, il apparaît 1) que la $p@5$ obtenue avec $E$ est similaire à celle obtenue avec $ET$, 2) que la $p@5$ obtenue avec $T$ est systématiquement inférieure d'environ $10$ à $15\%$ à celle de $E$.
-Ces résultats indiquent que l'information sémantique permettant de séparer les deux environnements est principalement portée par les événements.
-Ces résultats font, par ailleurs, écho aux travaux de  \cite{maffiolo_caracterisation_1999}, qui montrent que nous analysons de manière descriptive (en identifiant les sources) les scènes événementielles, \ie~composées d'événements sonores (\cf~Section~\ref{sec:simscene_sampleDataSet}).
-
-Enfin, il apparaît que la $p@5$ obtenue avec $E_{m}$ est égale, voire supérieure à celles obtenues avec $E$ et $ET$, et ce bien qu'une information partielle soit utilisée dans ce cas pour décrire les scènes.
-La dimension des vecteurs de description $S$ pour $E_m$ est en effet inférieure à la dimension des vecteurs $S$ pour $E$, qui est elle même inférieure à celle obtenue dans le cas où toutes les classes sont utilisées ($ET$).
-De plus, dans le cas où les marqueurs ne sont pas pris en compte pour la description ($E_{w/o,m}$), les résultats chutent, passant même en dessous de ceux obtenus en ne considérant que les textures.
-Cela confirme que la majorité de l'information sémantique permettant de faire la distinction entre i-scènes et ni-scènes est incluse dans les marqueurs.
-
-En résumé, nous déduisons de cette analyse les points suivants:
-
-\begin{enumerate}
-\item contrairement à ce que nous avions constaté avec les descripteurs structurels, une description sémantique de la composition des scènes, en terme de présence/absence de sources sonores, permet de bien distinguer les deux types d'environnement (i ou ni);
-\item l'information sémantique est majoritairement portée par les classes d'événements sonores;
-\item parmi les classes d'événements, seule une partie, \ie~les marqueurs sonores, est nécessaire pour faire la distinction entre les i- et ni-scènes.
-\end{enumerate}
-
-Maintenant que nous avons isolé les classes typiques des i- et ni-scènes, et vérifié que la distinction entre ces environnements dépendait de la présence de ces classes, il reste à voir si une description structurelle des scènes, basée uniquement sur ces marqueurs sonores, permet de caractériser l'agrément perçu, mieux qu'une description structurelle globale.
-
-\begin{figure}[t]
-        \myfloatalign
-        \includegraphics[width=\linewidth]{gfx/ch_5/pa5_1}
-       \caption{$P@5$ obtenues en considérant la matrice de dissimilarité résultant des distances par paires de Hamming calculées entre les vecteurs des descripteurs sémantiques des scènes.
-Les vecteurs sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.}\label{fig:pa5}
-\end{figure}
-
-\subsubsection*{Influence des marqueurs sonores sur l'agrément perçu}
-
-Nous évaluons les corrélations entre $\mathcal{A}_{scene}$ et les descripteurs structurels.
-Pour cette section, les descripteurs structurels sont calculés en tenant compte des marqueurs sonores précédemment identifiés.
-Nous définissons $X_m$ le descripteur $X$ calculé en ne prenant en compte que les sons des marqueurs.
-A l'inverse, nous définissons $X_b$ ($b$: pour «~bruit~») le descripteur $X$ calculé en prenant en compte toutes les classes de sons, excepté les marqueurs.
-Lorsque le descripteur caractérise une i-scène (idem pour une ni-scène), nous ne considérons, pour le calcul, que les marqueurs identifiés pour les i-scènes (ou pour les ni-scènes), que nous nommons i-marqueurs (ou ni-marqueurs).
-Les résultats sont affichés sur le tableau~\ref{tab:corrMarkers}.
-
-Concernant les niveaux sonores (\cf~Figures~\ref{fig:soundlevelMarkera} et~\ref{fig:soundlevelMarkerd}), les mêmes tendances sont observées entre $L_m$, $L(E)_m$ et $L(T)_m$, d'une part, et $L$, $L(E)$ et $L(T)$, d'autre part.
-Que l'on considère uniquement les marqueurs, ou l'ensemble des classes, il s'avère que :
-
-\begin{enumerate}
-\item il existe une différence significative entre les niveaux des i- et ni-scènes ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
-\item le niveau sonore des scènes est majoritairement porté par les événements sonores, comparé aux textures sonores;
-\item le niveau sonore des événements a une influence sur la perception de l'agrément pour les ni-scènes, mais pas pour les i-scènes;
-\item le niveau sonore des textures ne joue aucun rôle dans la perception de l'agrément.
-\end{enumerate}
-
-En conclusion, le niveau des ni-marqueurs a une influence négative sur l'agrément pour les ni-scènes. En revanche le niveau des i-marqueurs n’impacte pas l'agrément perçu pour les i-scènes.
-
-En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
-C'est la première fois qu'un indicateur objectif nous permet de préciser l'agrément des environnements agréables.
-Ceci nous amène à conclure que le niveau des classes de sons n'étant pas typiques d'un environnement agréable a un impact négatif sur l'agrément.
-
-Par ailleurs, alors que $L(T)$ ne présentait pas de corrélation pour les ni-scènes, une corrélation négative forte est observée pour $L(T)_b$ ($r-0.73$, $p<0.01$).
-Ce fait indique que les niveaux des classes de textures n'étant pas des marqueurs n'affectent pas l'agrément perçu de la même manière pour les i- et ni-scènes.
-Les niveaux semblent avoir un effet négatif pour les ni-scènes, alors que pour les i-scènes, aucun effet n'est relevé.
-
-Pour finir, nous considérons un dernier groupe de descripteurs, nommément $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiffa} et~\ref{fig:soundlevelMarkerDiffd}).
-Ces descripteurs expriment la différence entre les niveaux des marqueurs, et ceux des autres classes de sons.
-Ils traduisent l'émergence des marqueurs par rapport à la mixture sonore.
-
-Pour les i-scènes, une corrélation modérée et positive est observée pour $L_m-L_b$ ($r=0.67$, $p<0.01$) et $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
-Pour les ni-scènes, aucune corrélation n'est observée.
-Dans le cas des i-scènes, ce n'est donc pas le niveau absolu des marqueurs qui importe, mais leur niveau relatif, par rapport aux autres sons qui composent la scène.
-On observe donc pour les environnements idéaux un double mécanisme perceptif:
-
-\begin{itemize}
-\item plus le niveau absolu des sons n'étant pas des i-marqueurs est élevé, plus l'agrément est faible,
-\item plus le niveau relatif des i-marqueurs, par rapport aux autres sons, est élevé, plus l'agrément est élevé.
-\end{itemize}
-
-Pour les ni-scènes, le fait que nous observions des corrélations pour $L_m$ et $L(E)_m$, et aucune pour $L_m-L_b$ et $L(E)_m-L(E)_b$, montre que c'est bien le niveau absolu qui importe.
-
-\begin{table}[t]
-\setlength{\tabcolsep}{3pt}
-\centering
-{\renewcommand{\arraystretch}{1}
-\centering
-\begin{tabular}{l r r}
-                  &   i-scenes                  & ni-scenes \\
-\hline
-$L_m$              & 0.03  ($p=0.88$)           & \textbf{-0.75} ($p<0.01$) \\
-$L(E)_m$           & 0.08  ($p=0.66$)           & \textbf{-0.71} ($p<0.01$) \\
-$L(T)_m$           & -0.11 ($p=0.66$)           & -0.17 ($p=0.37$) \\
-$L_b$              & \textbf{-0.52} ($p<0.01$)  & -0.32 ($p=0.06$) \\
-$L(E)_b$           & \textbf{-0.51} ($p<0.01$)  & -0.30 ($p=0.07$) \\
-$L(T)_b$           & -0.32 ($p=0.05$)           & \textbf{-0.73} ($p<0.01$) \\
-$L_m-L_b$          & \textbf{0.67} ($p<0.01$)   & -0.31 ($p=0.07$) \\
-$L(E)_m-L(E)_b$    & \textbf{0.66} ($p<0.01$)   & -0.28 ($p=0.10$) \\
-$L(T)_m-L(T)_b$    & 0.16 ($p=0.54$)            & 0.21 ($p=0.28$) \\
-\hline
-\end{tabular}
-}
-\vspace{0.5mm}
-\caption{Coefficients de corrélation linéaire calculés entre l'agrément perçu moyen $\mathcal{A}_{scene}$ de l'expérience 1.b et les descripteurs structurels relatifs à la présence des marqueurs sonores.}
-\label{tab:corrMarkers}
-\end{table}
-
-\begin{figure}[t]
-        \myfloatalign
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_7}\label{fig:soundlevelMarkera}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_9}\label{fig:soundlevelNoisea}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_19}\label{fig:soundlevelMarkerDiffa}}\par
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_8}\label{fig:soundlevelMarkerd}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_10}\label{fig:soundlevelNoised}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
-        \caption{Dispersions des descripteurs structurels de niveaux sonores relatifs à la présence des marqueurs $L_m$ (a, d), $L(E)_m$ (b, e) et $L(T)_m$ (c, f), en fonction du type de scènes (a, b, c) et de l'agrément perçu $\mathcal{A}_{scene}$ de l'expérience 1.b (d, e, f).}\label{fig:soundlevelMarker}
-\end{figure}
-
-\subsection{Discussions}
-
-De cette analyse, nous retenons les points suivants:
-
-\begin{itemize}
-\item \emph{distinguer les i- et ni-scènes}: les descripteurs sémantiques, ainsi que les descripteurs structurels globaux ($L$, $L(E)$ et $L(T)$), permettent de faire la distinction entre les i-scènes et les ni-scènes.
-La description sémantique semble être plus performante;
-\item \emph{événements ou textures}: que ce soit pour les descripteurs sémantiques ou structurels, ce sont majoritairement les événements qui permettent de distinguer les deux types d'environnement, les textures n'apportant, au mieux, qu'une information limitée;
-\item \emph{prédire l'agrément}: si l'on considère les corrélations entre les descripteurs structurels et l'agrément, il semble que la manière de percevoir la qualité de l'environnement diffère en fonction de la nature de ce dernier (i ou ni).
-Il n'apparaît pas envisageable de considérer un même jeu de descripteurs pour prédire, à la fois, l'agrément des i-scènes, et l'agrément des ni-scènes :
-
-\begin{itemize}
-
-\item \emph{pour les ni-scènes}, ce sont le niveau global ($L$ et $L(E)$), ou le niveau des marqueurs sonores ($L_m$ et $L(E)_{m}$), qui impactent négativement l'agrément.
-On note ici que prendre en compte les contributions de différentes sources n'améliore pas la capacité de prédiction de l'agrément, par rapport à une analyse holistique de l'environnement.
-
-\item \emph{pour les i-scènes}, par contre, prédire l'agrément requiert d'étudier, de manière séparée, les caractéristiques des marqueurs sonores, et celles de l'ensemble des autres sons.
-Ainsi, le niveau des marqueurs pris relativement au bruit ($L(E)_m-L(E)_b$ et $L_m-L_b$) est positivement corrélé à l'agrément, alors que le niveau du bruit ($L_b$ et $L(E)_b$) est, lui, négativement corrélé.
-\end{itemize}
-\end{itemize}
-
-Le fait que l'agrément des i-scènes ne se corrèle pas avec des descripteurs physiques holistiques, contrairement à l'agrément des ni-scènes, a récemment été déjà observé \cite{gozalo2015relationship}.
-
-L'existence de deux modes de perception, mobilisant différents types de descripteurs, et dépendant de la nature (dans notre cas hédonique) des stimuli, est un phénomène qui se rapproche de celui observé pour la perception des textures (\cf~Section~\ref{sec:simscene_sampleDataSet}).
-Le cerveau adapte sa manière de traiter l'information (résumé statistique pour les textures, description fine pour les événements) suite à une prise de décision antérieure quant à la nature du stimulus (à savoir «~est-ce un événement ou une texture ?~»).
-De la même manière, les indicateurs actifs dans le jugement de l'agrément dépendent, eux aussi, d'une identification préalable de la nature hédonique globale de l'environnement (idéale ou non idéale).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%% XP 2 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2017-05-05 lafaySimulatedScenePerception.tex
ml did intro

@@
 \AuthorsI{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J.F. Petiot$^{1)}$}
-\AddressI{$^{1)}$ Institut de Recherche en Communications et  Cybern\'etique de Nantes (IRCCYN), Nantes, France.\\
-\hspace*{8pt}gregoire.lafay@irccyn.ec-nantes.fr\\
+\AddressI{$^{1)}$ Laboratoire des Sciences du Numérique, Nantes, France.\\
+\hspace*{8pt}mathieu.lagrange@cnrs.fr\\
 $^{2)}$ STMS Ircam-CNRS-UPMC Institut de Recherche et Coordination Acoustique/Musique, Paris, France}
@@
 
-Un des objectifs premiers des études sur les paysages sonores est d'identifier les informations contenues dans l'environnement qui influent sur les sensations perçues \cite{aletta2016soundscape}.
+%Un des objectifs premiers des études sur les paysages sonores est d'identifier les informations contenues dans l'environnement qui influent sur les sensations perçues \cite{aletta2016soundscape}.
 Questionner le lien entre sensation (\eg~calme) et environnement (\eg~scène sonore urbaine), revient à objectiver la représentation mentale que se fait un sujet d'une scène sonore en particulier (\eg~une scène sonore urbaine calme).
 
-Si on demande à un sujet d'évaluer un environnement donné suivant une échelle perceptive donnée (\eg~calme, agrément) \cite{axelsson2005soundscape,davies2013perception,cain2013development}, le gain d'information est conditionné par la nature des stimuli disponibles, qu'il s'agisse de scènes sonores enregistrées (expérience en laboratoire), ou réelles (expérience \emph{in situ}).
+One of the main goal of soundscape studies is to identify which components of the soundscape influence human perception \cite{aletta2016soundscape}.
+For example, in order to study the link between an environment (urban sound scene) and the induced human sensation (calmness), one mean is to objectify the mental representation of a specific sound scene, a calm urban sound scene.
+Several means have been considered in the literature to do so.
+
+%Si on demande à un sujet d'évaluer un environnement donné suivant une échelle perceptive donnée (\eg~calme, agrément) \cite{axelsson2005soundscape,davies2013perception,cain2013development}, le gain d'information est conditionné par la nature des stimuli disponibles, qu'il s'agisse de scènes sonores enregistrées (expérience en laboratoire), ou réelles (expérience \emph{in situ}).
 Cependant, le fait que ces stimuli existent permet d'en établir une description physique précise.
+
+First, a subject can be asked to describe a given sound environnment following a given perceptive scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}, the amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}.
+That being said, as those stimuli can be recordered and analysed.
 
@@
 
-Nous pensons que la simulation permet de faire un lien élégant et efficace entre ces deux approches.
+Second, a subject can be asked to describe a given sound environment \cite{guastavino2006ideal, dubois2006cognitive}.
+A large amount of quantitative and semantic information is then collected about the representation of this type of sound environment that the subject has.
+Unfortunately, without any reference to sound data, this representation cannot be caracterized physically.
+
+%Nous pensons que la simulation permet de faire un lien élégant et efficace entre ces deux approches.
 Il permet en effet d'obtenir du sujet une version modale, \ie~établie sur des dimensions physiques (le signal de la scène simulée), de sa représentation mentale de l'objet; les scènes générées étant à la fois caractérisées sémantiquement (\eg~sources sonores présentes) et physiquement (\eg~niveaux des sources sonores).
 
-De plus en plus de travaux tendent à montrer que toutes les sources ne participent pas de manière égale à la perception de la scène sonore \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
-szeremeta2009analysis}.
+We propose in this paper to consider the use of a soundscape simulator that the subject can use to objectify the representation of a given sound environment.
+We believe that the use of such a device allows us to gain the benefits of the two above cited approaches.
+As the subject is saked to produces audio data (the signal of the simulated scene) allows the experimenter to study a modal version of the mental representation of the subject that is both caracterized semantically and physically, respectively in terms of the sound sources and their levels.
+
+%De plus en plus de travaux tendent à montrer que toutes les sources ne participent pas de manière égale à la perception de la scène sonore \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape, szeremeta2009analysis}.
 Les recherches portent désormais sur l'étude des contributions spécifiques des différentes sources sur la notion de qualité affective de la scène \cite{gozalo2015relationship,ricciardi2015sound}.
@@
 
-Nous pensons que ces efforts de recherche peuvent grandement bénéficier de l'utilisation de scènes simulées, scènes dont les caractéristiques structurelles, en particulier la nature des sources présentes, leur répartition temporelle ainsi que leur forme d'onde respectives, sont directement accessibles par l'expérimentateur.
+Recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
+szeremeta2009analysis}.
+Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}.
+For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to very insteresting outcomes as it allows the experimenter to conveniently study separately the influence of the sounds sources that compose a sound environment of interest.
+Indeed, with such a material, not only the type of sound sources is available for study, but also the exact level and audio waveform for each source and also the structural characterestics of the scene, that is the temporal distribution of the events.
 
-Afin de répondre à ces problématiques, nous proposons d'utiliser un outil de simulation permettant à un sujet de simuler des paysages sonores à partir d'un corpus de sons enregistrés.
+%Nous pensons que ces efforts de recherche peuvent grandement bénéficier de l'utilisation de scènes simulées, scènes dont les caractéristiques structurelles, en particulier la nature des sources présentes, leur répartition temporelle ainsi que leur forme d'onde respectives, sont directement accessibles par l'expérimentateur.
+
+%Afin de répondre à ces problématiques, nous proposons d'utiliser un outil de simulation permettant à un sujet de simuler des paysages sonores à partir d'un corpus de sons enregistrés.
 Pour montrer l'intérêt de l'utilisation d'un tel outil pour questionnner la perception de l'environnement sonore, nous choisissons, comme cadre applicatif, le problème de l'agrément perçu dans les environnements sonores urbains.
@@
 
+To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmanelta scenes.
+Results and outcomes of a series of 3 experiments that build on the use of the simulator are studied in order to better comprehend how different sound soucres typically present in a urban scene impact pleasantness:
+
+% \begin{enumerate}
+% \item \emph{expérience 1.a, simulation}:  les sujets simulent les environnements idéaux/non idéaux qui serviront de stimuli pour les étapes suivantes;
+% \item \emph{expérience 1.b, évaluation de l'agrément}: les sujets doivent évaluer l'agrément des scènes simulées sur une échelle sémantique;
+% \item \emph{expérience 2, évaluation de l'agrément après modification des scènes}: comme pour l'expérience précédente, les sujets doivent évaluer l'agrément des scènes simulées sur une échelle sémantique.
+Cependant les scènes ont été modifiées, \ie~privées de certaines classes de sons identifiées comme ayant un impact sur l'agrément perçu.
+% \end{enumerate}
+
 \begin{enumerate}
-\item \emph{expérience 1.a, simulation}:  les sujets simulent les environnements idéaux/non idéaux qui serviront de stimuli pour les étapes suivantes;
-\item \emph{expérience 1.b, évaluation de l'agrément}: les sujets doivent évaluer l'agrément des scènes simulées sur une échelle sémantique;
-\item \emph{expérience 2, évaluation de l'agrément après modification des scènes}: comme pour l'expérience précédente, les sujets doivent évaluer l'agrément des scènes simulées sur une échelle sémantique.
-Cependant les scènes ont été modifiées, \ie~privées de certaines classes de sons identifiées comme ayant un impact sur l'agrément perçu.
+\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non ideal soundscapes that are considered as material for the following experiments
+\item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale
+\item \emph{experiment 2, pleasantness evaluation after modification of the scenes}: the subjects judge the pleasantness of the simulated scenes on a semantic scale as in 1.b, but some scenes are modified beforehand, \ie~some specific sounds classes that are identified as having a significant impact on perceived pleasantness are removed.
 \end{enumerate}
@@
 
-Dans notre démarche, le simulateur utilisé \footnote{simScene, disponible en ligne \url{http://soundthings.org/research/}} permet une restitution simplifiée (monophonique) des scènes, au profit d'un nombre plus important de paramètres et de sources disponibles, ce afin de garantir, à la sortie, des données viables et expressives.
+To the best of our knowledge, only Bruce~\al \cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception.
+They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location.
+The authors shows that the addition or removal of the sources follows more social or semantic matters than their acoustical caracteristics.
+A lack of diversity is nonetheless mentioned by the authors that limit the outcomes of the study.
 
-Le papier comprend 5 sections: Introduction, L'outil de simulation \emph{SimScene}, Expérience 1 .a \emph{simulation} et .b \emph{évaluation de l'agrément}, Expérience 2 \emph{évaluation de l'agrément après modification des scènes}, et Conclusions et perspectives.
+%Dans notre démarche, le simulateur utilisé \footnote{simScene, disponible en ligne \url{http://soundthings.org/research/}} permet une restitution simplifiée (monophonique) des scènes, au profit d'un nombre plus important de paramètres et de sources disponibles, ce afin de garantir, à la sortie, des données viables et expressives.
 
-Dans le cadre de cette communication, nous ne présentons de \emph{SimScene} que les fonctionnalités pertienentes pour notre étude.
+In our approach, the simulator developed for this study\footnote{simScene, available online \url{http://soundthings.org/research/}} gives a monophonic representation of the scene, with the added befenit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are, as much as possible, expressive and useful for analysis.
+
+%Le papier comprend 5 sections: Introduction, L'outil de simulation \emph{SimScene}, Expérience 1 .a \emph{simulation} et .b \emph{évaluation de l'agrément}, Expérience 2 \emph{évaluation de l'agrément après modification des scènes}, et Conclusions et perspectives.
+
+The remaining of the paper is organized as follows: the soundscape simulator \emph{SimScene} is introduced in Section \ref{sec:simulator}.
+The three experiments,  1 .a \emph{simulation}, 1.b \emph{pleasantness evaluation}, and 2 \emph{pleasantness evaluation after modiciation of the scenes} are respectively presented in Sections \ref{sec:simulation}, \ref{sec:evaluation}, and \ref{sec:modification}.
+Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.
+
+%Dans le cadre de cette communication, nous ne présentons de \emph{SimScene} que les fonctionnalités pertienentes pour notre étude.
 Pour plus de détails, se référer à \cite{rossignol2015simscene,lafay2016JAES}.
+
 L'expérience 1 .a a fait l'objet d'une étude pilote dont les résultats sont présentés dans \cite{lafay2013atiam,lafay2014new}.

### 2017-05-05 lafaySimulatedScenePerception.tex
ml start sec 1

@@
 
-\section{L'outil de simulation}
+\section{The simulator}
 \label{sec:simscene}
 
-\emph{Simscene} est un environnement de travail audio-numérique dont la première version a été développée dans le cadre du projet HOULE\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
+%\emph{Simscene} est un environnement de travail audio-numérique dont la première version a été développée dans le cadre du projet HOULE\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
 Il est prévu pour fonctionner sur les navigateurs internet \emph{Chrome} et \emph{Firefox}.
@@
 
-Le fonctionnement de \emph{Simscene} se rapproche de celui d'un séquenceur audio.
+\emph{Simscene} is an online digital audio tool whose first version have been developped as part of the HOULE prokect\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
+It has been designed to run on the popular web browers \emph{Chrome} and \emph{Firefox}.
+It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio} : \url{www.w3.org/TR/webaudio}}.
+The interface for selecting the soud sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.
+
+%Le fonctionnement de \emph{Simscene} se rapproche de celui d'un séquenceur audio.
 Un utilisateur choisit une classe de sons via l'interface de sélection (\cf~Section~\ref{sec:simscene_ssf}).
@@
 
-\subsection{Banque de données de samples}
+\emph{Simscene} is designed to be used as a simplified audio sequencer, with sequencing parameters specifically chosen for the generation of realistic soundscapes.
+To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}.
+A track is then created for this sounds source within the simulator interface.
+The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurences of the sounds source.
+Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.
+
+\subsection{Sound database}
 \label{sec:simscene_sampleDataSet}
 
-\subsubsection*{Typologie}
+In order to provide the user with a sound database that is well organized and covers as much as possible the variety of sound sources that are available in urban areas, a typology of urban sounds is first agreed upon.
 
-Afin de créer un corpus de classes de sons de référence pour la simulation, nous réalisons une typologie des sons environnementaux urbains.
+\subsubsection*{Typology}
 
-Une étude bibliographique est effectuée, afin d'identifier les sources et ambiances sonores les plus souvent citées dans les publications.
-Cette étude porte sur 16 articles ou thèses \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards}
-traitant de la manière dont nous discriminons les paysages sonores urbains. Notre typologie est établie sur la base des catégories/classes de sons relevées dans cette littérature.
+%Afin de créer un corpus de classes de sons de référence pour la simulation, nous réalisons une typologie des sons environnementaux urbains.
 
-Les réactions à la musique étant jugées trop subjectives, et les jugements esthétiques ne pouvant qu'altérer les données d'évaluation, nous choisissons, dans cette étude, de ne pas considérer les sources musicales de type musiciens de rue, radios de voitures, d'appartements, \etc.
+%Une étude bibliographique est effectuée, afin d'identifier les sources et ambiances sonores les plus souvent citées dans les publications.
+Cette étude porte sur 16 articles ou thèses \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} traitant de la manière dont nous discriminons les paysages sonores urbains.
+Notre typologie est établie sur la base des catégories/classes de sons relevées dans cette littérature.
 
-\subsubsection*{Événements et Textures}
+The chosen typology is established based on the category/classes of sounds found while reviewing 16 articles or thesis manuscripts \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} that study how humans discriminate different kind of urban soundscapes.
 
-La littérature fait généralement la distinction entre :
+%Les réactions à la musique étant jugées trop subjectives, et les jugements esthétiques ne pouvant qu'altérer les données d'évaluation, nous choisissons, dans cette étude, de ne pas considérer les sources musicales de type musiciens de rue, radios de voitures, d'appartements, \etc.
 
+We choose not to include any musical content in the database of sounds, as the study of the pleasantness of a given style or genre of music is beyond the scope of the study.
+
+\subsubsection*{Events and Textures}
+
+%La littérature fait généralement la distinction entre :
+% \begin{itemize}
+% \item {un événement sonore}: un son isolé, ponctuel, dont les caractéristiques physiques varient au cours du temps;
+% \item {une texture sonore}: un son, isolé, long, dont les caractéristiques physiques restent stables au cours du temps \cite{saint1995classification}.
+% \end{itemize}
+
+Litterature usually distinguish between:
 \begin{itemize}
-\item {un événement sonore}: un son isolé, ponctuel, dont les caractéristiques physiques varient au cours du temps;
-\item {une texture sonore}: un son, isolé, long, dont les caractéristiques physiques restent stables au cours du temps \cite{saint1995classification}.
+\item {a sound event}: an isolated sound, sparsely ditrubuted whose acoustical characteristics may change with respect to time
+\item {a sound texture}: an isolated sound of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
 \end{itemize}
 
-L'outil de simulation \emph{Simscene} tient compte de cette distinction, et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures.
+The simulation tool \emph{Simscene} follows this distinction and consider two disctint sound databases: one with the classes of events only and the other with classes of textures.
+These two types have specific simulation procedures.% et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures.
 Événements et textures bénéficient de processus de simulation spécifiques.
 
-Les différences morphologiques entre événements et textures sonores ont des conséquences sur la perception des scènes.
+%Les différences morphologiques entre événements et textures sonores ont des conséquences sur la perception des scènes.
+
+We do so, because we assume that the morphological differences between those two classes have some important consequences on the percpetion of the scenes.
 

### 2017-05-10 lafaySimulatedScenePerception.tex
ml done with simscene presentation

@@
 
-Lorsque des événements émergent d'un environnement sonore, le cerveau traite l'information des différentes sources de manière séparée \cite{bregman1994auditory}.
+%Lorsque des événements émergent d'un environnement sonore, le cerveau traite l'information des différentes sources de manière séparée \cite{bregman1994auditory}.
 Plusieurs flux auditifs sont ainsi générés, \ie~un pour chaque séquence d'événements émis par la même source \cite{carlyon2004brain}.
@@
 
-Maffiolo \cite{maffiolo_caracterisation_1999} met en évidence deux processus de catégorisation distincts, engagés en fonction de la capacité de l'auditeur à identifier des événements sonores.
+According to the Auditory Scene Analysis (ASA) theory, the brain processes sperately events emerging from different sources of an sound environment \cite{bregman1994auditory}.
+Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}.
+In this case, the source is a perceptual entity.
+Indeed, events that are not isolated are grouped into a stream.
+
+%Maffiolo \cite{maffiolo_caracterisation_1999} met en évidence deux processus de catégorisation distincts, engagés en fonction de la capacité de l'auditeur à identifier des événements sonores.
 Elle montre l'existence de deux catégories cognitives abstraites d'environnements sonores respectivement appelées: «~les séquences événementielles~» et «~les séquences amorphes~».
 
-Les séquences événementielles sont des environnements composés d'événements saillants et identifiables (\emph{démarrage de voiture}, \emph{voix d'homme}).
+Maffiolo \cite{maffiolo_caracterisation_1999} distinguish two distinct categorisation processes, that differentiate themselves with the capacity of the listener to identify sound events.
+She shows that those processes leads to two abstract cognitive category respectively termed ""~the event sequences ~" and "~ the amorphous sequences".
+
+%Les séquences événementielles sont des environnements composés d'événements saillants et identifiables (\emph{démarrage de voiture}, \emph{voix d'homme}).
 Elles profitent d'une analyse descriptive basée sur l'identification des sources sonores.
@@
 
-Pour ce qui est des textures, \ie~des sons possédants des caractéristiques stables au cours du temps, \cite{mcdermott2011sound,mcdermott2013summary} montrent que le cerveau peut décider sciemment de dégrader l'information physique perçue, en la résumant de manière statistique, faisant fi de toute autre représentation plus détaillée \cite{nelken2013ear}.
+The event sequences are composed of salient and identifiable events such as \emph{car start}, \emph{mal speech}.
+They arise from a descriptive analysis based on the identification of the sound sources.
+On contrary, the amorphous sequences are environments where distinct events can hardly be found, such as \emph{traffic hubhub}.
+They build upon an holistic analysis based on global acoustic features.
+
+%Pour ce qui est des textures, \ie~des sons possédants des caractéristiques stables au cours du temps, \cite{mcdermott2011sound,mcdermott2013summary} montrent que le cerveau peut décider sciemment de dégrader l'information physique perçue, en la résumant de manière statistique, faisant fi de toute autre représentation plus détaillée \cite{nelken2013ear}.
+
+Concerning sound textures, \ie~sounds that have stable caracteristics along time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrate that the human brain can decide not to abstract the physical information available using some of the summary statistics discarding lower level information \cite{nelken2013ear}.
 
@@
 
-\subsubsection*{Taxonomie}
+At the light of those results, it appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli \cite{nelken2013ear,mcdermott2013summary}.
+We thus consider in this study the soundscape as "skeleton of events on a bed of textures" as coined in \cite{nelken2013ear}.
+
+\subsubsection*{Taxonomy}
 
@@
         \includegraphics[width=.8\linewidth]{gfx/ch_4/3}
-       \caption{Organisation hiérarchique de la banque de sons isolés utilisée pour la simulation.}\label{fig:orgDb}
+       \caption{Hierachical organisation of the isolated sounds used ofr the simulation.}\label{fig:orgDb}
 \end{figure}
 
-Nous nommons sample, un enregistrement d'un son isolé, événement ou texture. Chaque classe est alors comprise comme une collection de samples jugés perceptivement équivalents.
+%Nous nommons sample, un enregistrement d'un son isolé, événement ou texture. Chaque classe est alors comprise comme une collection de samples jugés perceptivement équivalents.
 
-Les classes de sons sont organisées suivant une structure hiérarchique (\cf~Figure~\ref{fig:orgDb}).
+We term sample, a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples percieved equivalently.
+
+%Les classes de sons sont organisées suivant une structure hiérarchique (\cf~Figure~\ref{fig:orgDb}).
 Cette structure hiérarchique est largement inspirée de l'axe vertical de l'organisation catégorielle proposée par E.
@@
 
-Nous établissons deux taxonomies, une pour les événements, une autre pour les textures (\cf~Annexe~\ref{app:taxonomie}).
+The sound classes are organized hierachically (\cf~Figure~\ref{fig:orgDb}) according to a structure close to the one of the vertical axis of the categorical organization proposed by E.
+Rosch \cite{rosch1978cognition}.
+In this taxonomy, the lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources.
+For class with high level of abstraction with sub classes, its collection of samples is the union of the collections of the sub classes.
+
+%Nous établissons deux taxonomies, une pour les événements, une autre pour les textures (\cf~Annexe~\ref{app:taxonomie}).
 Pour les événements, nous considérons quatre niveaux d'abstraction allant des classes les plus générales (niveau d'abstraction 0), aux classes les plus spécifiques (niveau d'abstraction 3).
@@
 
-\subsubsection*{Acquisition des samples}
+Accounting for the perceptual matters detailed previously, two taxonomies are built, one for event sounds and one for texture sounds.
+Four levels of abstraction are considered from the more generic one (level 0), to the more specific classe (level 3), leading to a taxonomy close to the one used in \cite{Salamon14}.
+Only three levels of abstraction are considered for the texture sounds.
 
-Sur la base des taxonomies précédemment établies, 483 sons isolés ont été collectés, dont 381 événements, et 102 textures.
+\subsubsection*{Sound samples collection}
+
+%Sur la base des taxonomies précédemment établies, 483 sons isolés ont été collectés, dont 381 événements, et 102 textures.
 Parmi ces samples, 332 ont été enregistrés, et 151 proviennent de différentes banques de sons (\emph{SoundIdeas}\footnote{\emph{SoundIdeas}: \url{www.sound-ideas.com}} et \emph{Universal SoundBank}\footnote{\emph{Universal SoundBank}: \url{www.universal-soundbank.com}}).
 
-Tous les enregistrements ont été effectués à l’aide d'un micro canon \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} relié à un enregistreur \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}}.
+483 isolated sounds are collected and organized with the two above discussed typologies, 381 events and 102 textures.
+Among those samples, 332 have been recorded and 151 have been taken from two sound libraries: \emph{SoundIdeas}\footnote{\emph{SoundIdeas}: \url{www.sound-ideas.com}} and \emph{Universal SoundBank}\footnote{\emph{Universal SoundBank}: \url{www.universal-soundbank.com}}.
+
+%Tous les enregistrements ont été effectués à l’aide d'un micro canon \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} relié à un enregistreur \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}}.
 L’utilisation du micro canon nous permet d’isoler les événements sonores du bruit de fond urbain.
@@
 
-Tous les sons ont été normalisés au même niveau $RMS$ de $-12$ $dB$ (FS), \ie.
+Orignal sounds have been recorded using a shotgun microphone \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} relié à un enregistreur \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}}.
+The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background.
+Conversely, it allowed us to avoid dominant sound sources while recording texture sounds by pointing the microphone to distant eareas with no dominant sound sources.
+
+%Tous les sons ont été normalisés au même niveau $RMS$ de $-12$ $dB$ (FS), \ie.
 relativement à un niveau pleine échelle (\emph{relative to Full Scale}).
@@
 
-\subsection{Paramètres}
+All samples are normalised to the same $RMS$ level of $-12$ $dB$ FS, \ie. relative to Full Scale. In our case, the full scale level is of 1 Volt.
+
+\subsection{Parameters}
 \label{sec:simscene_parametre}
 
-Par choix de conception, l'outil de simulation ne permet pas d’interagir avec un sample en particulier, mais toujours avec une piste, \ie~une séquence de samples.
+%Par choix de conception, l'outil de simulation ne permet pas d’interagir avec un sample en particulier, mais toujours avec une piste, \ie~une séquence de samples.
 Plusieurs paramètres permettent au sujet de contrôler la structure de chaque piste.
 
+By design, the simulation tool do not allow the user to interect and control a specific sample.
+Interaction is done at the track level, a track being a sequence of samples.
+Several paramters are available to the subject to control the track:
+
+% \begin{itemize}
+% \item \emph{niveau sonore} ($dB$): pour chaque sample, les niveaux sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en termes de moyenne et de variance;
+% \item \emph{espacement inter-onset} (seconde): (piste d'événements seulement) comme pour les niveaux, les espacements sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance;
+% \item \emph{début et fin} (seconde): le sujet fixe le début et la fin de chaque piste.
+% \end{itemize}
+
 \begin{itemize}
-\item \emph{niveau sonore} ($dB$): pour chaque sample, les niveaux sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en termes de moyenne et de variance;
-\item \emph{espacement inter-onset} (seconde): (piste d'événements seulement) comme pour les niveaux, les espacements sont tirés aléatoirement à partir d'une distribution normale, paramétrée par le sujet en terme de moyenne et de variance;
-\item \emph{début et fin} (seconde): le sujet fixe le début et la fin de chaque piste.
+\item \emph{level} ($dB$): the levels are  randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
+\item \emph{inter-onset spacing} (second): for the event's track only, the spacings arerandomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
+\item \emph{start and end time} (second): the subject set the first and last occurence of a sample of the selected class.
 \end{itemize}
 
-Afin de faciliter la simulation, deux paramètres supplémentaires sont proposés:
+%Afin de faciliter la simulation, deux paramètres supplémentaires sont proposés:
+
+To ease the simulation, two parameters are also proposed:
+
+% \begin{itemize}
+% \item \emph{fondu par événement} (seconde): (piste d'événements seulement) le sujet fixe une durée de fondu (entrée et sortie) appliquée à chaque sample;
+% \item \emph{fondu global} (seconde): le sujet fixe les durées de fondus pour l'entrée et la sortie de la piste. Ces fondus s'appliquent ainsi à l'ensemble des samples de la piste.
+% \end{itemize}
 
 \begin{itemize}
-\item \emph{fondu par événement} (seconde): (piste d'événements seulement) le sujet fixe une durée de fondu (entrée et sortie) appliquée à chaque sample;
-\item \emph{fondu global} (seconde): le sujet fixe les durées de fondus pour l'entrée et la sortie de la piste. Ces fondus s'appliquent ainsi à l'ensemble des samples de la piste.
+\item \emph{event's fades} (seconds): for the event's track only, the subject can set a fade in / fade out duration applied to each samples;
+\item \emph{global fades} (seconds): the subject set the duration of the fade in /fade out at the beginning and the end of the track.
 \end{itemize}
 
-Deux de ces paramètres ne s’appliquent que pour les pistes d'événements (\emph{fondu par événement}  et \emph{espacement inter-onset}), les samples des textures étant séquencés sans espacement.
+%Deux de ces paramètres ne s’appliquent que pour les pistes d'événements (\emph{fondu par événement}  et \emph{espacement inter-onset}), les samples des textures étant séquencés sans espacement.
 
-\subsection{Interface de sélection}
+The samples of texture are sequenced without time spacing, therefore the parameters \emph{event's fade} and \emph{inter-onset spacing} are not available ofr this kind of track.
+
+\subsection{Selection interface}
 \label{sec:simscene_ssf}
 
-La grande majorité des outils permettant de parcourir une banque de données propose une recherche textuelle sur la base de mots-clé.
+%La grande majorité des outils permettant de parcourir une banque de données propose une recherche textuelle sur la base de mots-clé.
 Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique.
 
-En effet la description verbale du son, si elle est accessible au sujet, peut potentiellement influencer sa sélection.
+%En effet la description verbale du son, si elle est accessible au sujet, peut potentiellement influencer sa sélection.
 Dans les faits, pour construire une scène environnementale «~calme~», le sujet sélectionne a priori les sons référencés sous le vocable \emph{parc}.
 
-Pour limiter l’influence de l’interface sur le sujet, il nous paraît nécessaire de libérer sa recherche de toute information textuelle.
+Most of browsing tools are based on keyword indexing.
+For sensory experiment studying the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject.
+Indeed, the subject will select \textit{a priori} sound referenced belonging to \emph{park} to build a \emph{calm} soundscape.
+
+%Pour limiter l’influence de l’interface sur le sujet, il nous paraît nécessaire de libérer sa recherche de toute information textuelle.
 Nous proposons à l'utilisateur une interface graphique lui permettant d’explorer la banque de sons exclusivement à partir de l’écoute.
 
-La figure~\ref{fig:ssf} présente l'interface pour la banque de données d'événements sonores.
+Therefore, the selection interface is text free and composed of spatially organized circles on a two dimensional plane.
+
+%La figure~\ref{fig:ssf} présente l'interface pour la banque de données d'événements sonores.
 Visuellement, les classes du dernier palier (niveaux d'abstraction 3 pour les événements, 2 pour les textures) sont représentées par des cercles, et positionnées sur un plan.
@@
 
-Chaque classe du dernier palier possède un son prototype.
+The Figure~\ref{fig:ssf} shows the interface used for the selection of events.
+The classes of the higher level of abstraction are plotted with circles.
+The spatial location of the circles is chosen according to the hierarchical organization of the sound database: sub-classes belonging to the same class are close to each others, and so on until the user reaches the lower level classes displayed in grey which are directly linked to a collection of samples.
+
+%Chaque classe du dernier palier possède un son prototype.
 Ces sons ont été choisis par les expérimentateurs.
@@
 Le sujet parcourt la banque de sons en cliquant sur les cercles.
+Cette interface a fait l'objet d'une étude approfondie dont les résultats sont publiés dans \cite{lafay2016JAES}.
 
-Cette interface a fait l'objet d'une étude approfondie dont les résultats sont publiés dans \cite{lafay2016JAES}.
+Each of those classes have a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle.
+The subject can browse the database by listening to those prototype sound.
+The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed in \cite{lafay2016JAES}.
 
@@
         {\includegraphics[width=\columnwidth]{gfx/ch_4/simscene}\label{fig:simscene}}
-       \caption{Les interfaces de sélection (a) et de simulation (b) de l'outil \emph{Simscene}.}
+       \caption{Visual interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{Simscene} tool.}
 \end{figure}
 
-\subsection{Interface de simulation}
+\subsection{Simulation interface}
 
-L'interface de simulation propose un rendu graphique schématisé de la scène en cours de création (\cf~Figure~\ref{fig:simscene}).
+%L'interface de simulation propose un rendu graphique schématisé de la scène en cours de création (\cf~Figure~\ref{fig:simscene}).
 La piste est représentée par une bande possédant un axe temporel.
@@
 
-La scène sonore est ainsi vue comme une somme de sources sonores, ou, si l'on admet la distinction opérée entre événements et textures sonores, «~un squelette d'événements sur un lit de textures~» \cite{nelken2013ear}.
+As shown on Figure~\ref{fig:simscene}, the simulation interface displays a schematic of the scene under creation.
+Each track is represented as an horizontal band with a temporal axis.
+Each sample of this track is displayed by a rectangle whose height is proportional to the amplitude of the sample.
+For event's tracks, the horizontal spacing between those rectangles is a function of the time delay between their onsets.
+For tracks of texture, a unique rectangle is displayed as this kind of sounds do not allow spacing with silence.
+As actual amplitude and spacing values are drawn from random variables, each time the subject changes the value of a parameter, the location and height of the rectangles are updated to reflect the changes in the sequencing of the samples.
+The subject can listen to the resulting sound scene.
+
+%La scène sonore est ainsi vue comme une somme de sources sonores, ou, si l'on admet la distinction opérée entre événements et textures sonores, «~un squelette d'événements sur un lit de textures~» \cite{nelken2013ear}.
 Cet outil de simulation est présenté en détail dans \cite{rossignol2015simscene}.
+
+As such, the underlying model of the scene is a sum of sound sources, more precisely a "skeleton of events on a bed of textures" \cite{nelken2013ear}.
+Further details about the simulation interface can be found in \cite{rossignol2015simscene}.
 

### 2017-05-10 lafaySimulatedScenePerception.tex
ml starts xp 2

@@
 
-\section{Expérience 2 : modification de la composition sémantique}
+\section{Experiment 2: modification of the semantic content}
 \label{sec:xp3}
 
-\subsection{Objectif}
+\subsection{Objective}
 
-L'expérience précédente a montré que, parmi les classes de sons peuplant le monde sonore, celles regroupant les marqueurs sont caractéristiques de certains types d'environnement.
+%L'expérience précédente a montré que, parmi les classes de sons peuplant le monde sonore, celles regroupant les marqueurs sont caractéristiques de certains types d'environnement.
 Ces marqueurs sonores semblent avoir un impact particulier sur la perception de leurs environnements.
@@
 
-Afin de vérifier que l'agrément des scènes idéales et non-idéales dépend de la présence des marqueurs, les scènes sonores précédemment simulées sont régénérées, sans les classes de marqueurs.
+The previous experiment demonstrated that, among the classes of sounds occuring in urban soundscapes, those with markers are specific of some of environments.
+Those sound markers seem to have a great impact on perception.
+This impact is studied in more detail using an added benefit of the simulation paradigm proposed in this paper, the ability to manipulate and modify the generated scenes.
+
+%Afin de vérifier que l'agrément des scènes idéales et non-idéales dépend de la présence des marqueurs, les scènes sonores précédemment simulées sont régénérées, sans les classes de marqueurs.
 Pour les i-scènes, les i-marqueurs sont retirés.
@@
 
-L'objectif est de vérifier si l'absence des marqueurs a un impact sur l'agrément perçu. Deux hypothèses sont formulées:
+In order to verify that the pleasantness of the ideal and non ideal scenes if affected by the sound markers, the audio waveforms of the simulated scenes are regenerated without considering the classes identified as markers.
+For the i-scenes, the i-markers are removed.
+For the ni-scenes, the ni-markers are removed.
+A perceptive test evaluating the pleasantness is conducted with a protocol close to the one considered in experiment 1.b.
+
+%L'objectif est de vérifier si l'absence des marqueurs a un impact sur l'agrément perçu. Deux hypothèses sont formulées:
+
+Thus the objective of this experiment is to study if the removal of the previously identified markers have an impact on the percieved pleasantness. Two hypothesis are formulated:
+
+%\begin{itemize}
+% \item \emph{pour les ni-scènes} nous faisons l'hypothèse que l'absence des ni-marqueurs va \textbf{augmenter} la valeur de l'agrément perçu;
+% \item \emph{pour les i-scènes} nous faisons l'hypothèse que l'absence des i-marqueurs va \textbf{diminuer} la valeur de l'agrément perçu.
+% \end{itemize}
 
 \begin{itemize}
-\item \emph{pour les ni-scènes} nous faisons l'hypothèse que l'absence des ni-marqueurs va \textbf{augmenter} la valeur de l'agrément perçu;
-\item \emph{pour les i-scènes} nous faisons l'hypothèse que l'absence des i-marqueurs va \textbf{diminuer} la valeur de l'agrément perçu.
+\item \emph{for the ni-scenes} we hypothesize that the absence of ni-markers will \textbf{increase} the pleasantness score;
+\item \emph{for the i-scenes} we hypothesize that the absence of the i-markers will \textbf{decrease} the pleasantness score.
 \end{itemize}
 
-Si la première hypothèse est intuitive, la deuxième l'est moins.
+%Si la première hypothèse est intuitive, la deuxième l'est moins.
 En effet, il n’apparaît pas évident que la suppression des i-marqueurs, bien que s'agissant de sons positivement connotés, diminue la qualité globale d'un environnement.
@@
 
-Néanmoins, comme nous l'avons vu, le niveau global n'est qu'un indicateur partiel de l'agrément pour les environnements sonores idéaux.
+If the first hypothesis is rather intuitive, the second is less.
+Indeed, it is not obvious that the removal of the i-markers, though perceptively connoted, decreases the global quality of a soundscape.
+This removal also decreases the global level of the scene.
+
+%Néanmoins, comme nous l'avons vu, le niveau global n'est qu'un indicateur partiel de l'agrément pour les environnements sonores idéaux.
 Qui plus est, cet indicateur, lorsque qu'il décrit le niveau des i-marqueurs, impacte de manière positive la qualité de la scène.
@@
 
-\subsection{Planification de l'expérience 2}
+Though, as discussed before, the global amplitude level can only be considered as a partial indicator of pleasantness of the ideal soundscapes.
+Furthermore, the level of i-markers positively impact the pleasantness.
+For those reasons, the validation of the second hypothesis is of high interest.
+
+\subsection{Planification of experiment 2}
 

### 2017-05-11 lafaySimulatedScenePerception.tex
ml reads more of xp 2

@@
 
-La banque de données de stimuli compte 144 séquences de 30 secondes. Ces 144 séquences comprennent:
+%La banque de données de stimuli compte 144 séquences de 30 secondes. Ces 144 séquences comprennent:
+% \begin{itemize}
+% \item \emph{72 am-scenes}: les 72 scènes précédemment simulées, avec les classes de marqueurs (am).
+Nous notons i/am-scènes, les 36 scènes idéales avec marqueurs, et ni/am-scènes les 36 scènes non-idéales avec marqueurs;
+% \item \emph{72 sm-scènes}: les 72 scènes précédemment simulées, régénérées sans les classes de marqueurs (sm).
+Nous notons i/sm-scènes, les 36 scènes idéales sans marqueurs, et ni/sm-scènes les 36 scènes non-idéales sans marqueurs.
+% \end{itemize}
+
+There are 144 stimuli of 30 seconds durations. More precisely:
 
 \begin{itemize}
-\item \emph{72 am-scènes}: les 72 scènes précédemment simulées, avec les classes de marqueurs (am).
-Nous notons i/am-scènes, les 36 scènes idéales avec marqueurs, et ni/am-scènes les 36 scènes non-idéales avec marqueurs;
-\item \emph{72 sm-scènes}: les 72 scènes précédemment simulées, régénérées sans les classes de marqueurs (sm).
-Nous notons i/sm-scènes, les 36 scènes idéales sans marqueurs, et ni/sm-scènes les 36 scènes non-idéales sans marqueurs.
+\item \emph{72 km-scenes}: the 72 scenes originally simulated by the subject of experiment 1.a where the sound classes identified as markers are kept.
+The 36 ideal scenes with markers are noted i/km-scenes, and the 36 non ideal scenes with markers are noted ni/km-scenes.
+\item \emph{72 rm-scènes}:  72 scenes where the sound classes identified as markers are removed.
+The 36 ideal scenes without markers are noted i/rm-scenes, and the 36 non ideal scenes without markers are noted ni/rm-scenes.
 \end{itemize}
 
-Nonobstant l'absence des marqueurs, les am- et sm-scènes sont en tout point semblables.
+%Nonobstant l'absence des marqueurs, les am- et sm-scènes sont en tout point semblables.
 
-Nombre de am-scènes sont composées, en majorité, de samples de marqueurs.
+Notwithstanding the presence or absence of markers, the km-scenes and rm-scenes are exactely the same.
+
+%Nombre de am-scènes sont composées, en majorité, de samples de marqueurs.
 Afin de ne pas dénaturer abusivement ces scènes, en créant notamment des temps de «~vide~», \ie~ne comprenant aucun sample, nous ne supprimons que les marqueurs des classes d'événements du premier niveau d'abstraction (\cf~Tableau~\ref{tab:markers}).
@@
 
+In order to create km-scenes with still some sound diversity and no long duration absence of sound activity within the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}.
+Those classes are:
+
+% \begin{itemize}
+% \item \emph{cloche}, \emph{sonnette de vélo}, \emph{animaux} pour les i/sm-scènes;
+% \item \emph{sirène}, \emph{klaxon} pour les ni/sm-scènes.
+% \end{itemize}
+
 \begin{itemize}
-\item \emph{cloche}, \emph{sonnette de vélo}, \emph{animaux} pour les i/sm-scènes;
-\item \emph{sirène}, \emph{klaxon} pour les ni/sm-scènes.
+\item \emph{Church bell}, \emph{bicycle bell}, \emph{animals} pour les i/sm-scènes;
+\item \emph{siren}, \emph{car horn} pour les ni/sm-scènes.
 \end{itemize}
 
-Il est important de noter ici que tous les i- et ni-marqueurs ne sont donc pas supprimés dans les sm-scènes.
+%Il est important de noter ici que tous les i- et ni-marqueurs ne sont donc pas supprimés dans les sm-scènes.
 
-\subsubsection*{Procédure}
+Thus, only a fraction of the i-markers and ni-markers are removed in the rm-scenes.
+
+\subsubsection*{Experiment}
 
@@
 
-\subsubsection*{Dispositif expérimental}
+\subsubsection*{Experimental protocol}
 
-Tous les sujets passent l'expérience sur des machines identiques.
+%Tous les sujets passent l'expérience sur des machines identiques.
 L'audio est diffusé en monophonie, par le biais de casques audio semi-ouverts \emph{Beyer-Dynamic DT 990 Pro}.
@@
 
-Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
+All the subjects perform the test on the same type of computers . Thy listen to the stimuli with semi-open \emph{Beyer-Dynamic DT 990 Pro} headphones. The sound level is the same for all subjects.
 
-Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de cette dernière, et de répondre aux éventuelles questions des sujets.
+%Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
+
+All the subjects perform the test at the same time, in a quiet environment. They are not allowed to talk to each other during the experiment.
+
+%Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de cette dernière, et de répondre aux éventuelles questions des sujets.
+
+A supervisor is available during the whole duration of the experiment to ensure a smooth running of the latter and answer any questions the subjects may have.
 
@@
 
-12 sujets (4 femmes) participent à l'expérience.
+%12 sujets (4 femmes) participent à l'expérience.
 Aucun d'entre eux n'a réalisé l'expérience de simulation, ni la première expérience d'évaluation.
 Les sujets sont âgés de 22 à 61 ans (moyenne: 29.5, écart-type: 14).
-Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
+Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.Tous les sujets ont réalisé l'expérience avec succès.
 
-Tous les sujets ont réalisé l'expérience avec succès.
+12 subjects (4 womens) perform the test.
+None of them have performed experiments 1.a and 1.b.
+The subjects are 22 to 61 years old (average: 29.5, standard deviation: 14).
+All the subjects live in Nantes, France, for at least two years.
+All the subjects performed the test succesfully.
 
-\subsection{Données et méthodes d'analyses}
+\subsection{Data and analysis method}
 
-Les données analysées sont les mêmes que pour la première expérience. Nous invitons le lecteur à se référer à la section~\ref{sec:xp1_dataAna} pour plus de détails.
+%Les données analysées sont les mêmes que pour la première expérience. Nous invitons le lecteur à se référer à la section~\ref{sec:xp1_dataAna} pour plus de détails.
+
+The type of data analysed in this experiment have been considered for experiment 1.a, see Section~\ref{sec:xp1_dataAna} for details.
 
@@
 
-Tous les tests de significativité sont effectués avec un seuil critique $\alpha=0.05$.
+To do so, we perform a variance analysis.
+We consider $\mathcal{A}_{sujet}$ as a dependant variable, and as independant variables, the type of envorinment (i/ni), and the presence/absence of markers (km/rm).
+As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used in order to verify the hypothesis that there exist a significant difference of percieved pleasantness between km-scenes and rm-scenes.
 
-\subsection{Résultats}
+(\emph{within-subject}.
 
-\subsubsection{Détection des valeurs extrêmes}
+\emph{post hoc} are done by following the Tukey-Kramer procedure.
+
+%Tous les tests de significativité sont effectués avec un seuil critique $\alpha=0.05$.
+
+All the significance tests are done using a critical threshold $\alpha=0.05$.
+
+\subsection{Results}
+
+\subsubsection{Outliers detection}
 

### 2017-05-12 lafaySimulatedScenePerception.tex
ml done with xp 2

@@
 
-Il s'agit ici de vérifier que la suppression des i- et ni-marqueurs impacte l'agrément perçu.
+%Il s'agit ici de vérifier que la suppression des i- et ni-marqueurs impacte l'agrément perçu.
 Pour ce faire, nous utilisons l'analyse de variance.
@@
 
+The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness.
 To do so, we perform a variance analysis.
@@
 As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used in order to verify the hypothesis that there exist a significant difference of percieved pleasantness between km-scenes and rm-scenes.
-
-(\emph{within-subject}.
-
-\emph{post hoc} are done by following the Tukey-Kramer procedure.
+The two independant variables are cosidered as \emph{within-subject} factors.
+The factors being of only two levels each (type: i/ni; marker: km/rm), the sphericity impothesis does not need to be checked.
+\emph{Post hoc} analysis are done by following the Tukey-Kramer procedure.
 
@@
 
-Considérons $\mathcal{A}_{sujet}$ pour les am-scènes.
+%Considérons $\mathcal{A}_{sujet}$ pour les am-scènes.
 Il apparaît que les réponses d'un des sujets diffèrent des autres.
@@
 
-\subsubsection{Influence de la présence des marqueurs sur l'agrément perçu}
+Let us consider $\mathcal{A}_{sujet}$ for the km-scenes.
+Close inspection that the answers given by one subject strongly differs with the ones of others.
+This subject evaluated positively close to half of the ni/km-scenes, see Annex~\ref{app:xp2}, subject 7.
+This subject gave a grade above 0 for 58\% of the ni/km-scenes, contrary to an average of 11\% for the remaining of the subjects.
 
-Dans cette section nous étudions comment les sujets ont perçu les différents types de scènes, nommément: i/am-, ni/am-, i/sm- et ni/sm-scène.
+Furthermore, this subject used the whole scale (from -5 to 5) to grade both the i-scenes and the ni-scenes.
+Those behaviors strongly differing with the ones of the others subjects be they from this experiment or the previous ones, we decide to discard the subject 7 from the analysis.
+
+\subsubsection{Influence of the markers on perceived pleasantness}
+
+%Dans cette section nous étudions comment les sujets ont perçu les différents types de scènes, nommément: i/am-, ni/am-, i/sm- et ni/sm-scène.
 L'ANOVA à mesures répétées pratiquée sur $\mathcal{A}_{sujet}$ montre un effet significatif du type d'environnement (i/ni: $F[1,10]=175$, $p<0.01$), de la présence/absence des marqueurs (am/sm: $F[1,10]=7$, $p<0.05$), ainsi que de l'interaction entre les deux facteurs ($F[1,10]=67$, $p<0.01$).
 
-L'analyse \emph{post hoc} montre, quant à elle, des différences significatives entre tous les groupes d'observations, notamment entre les i/am- et i/sm-scenes ($p<0.05$) et les ni/am- et ni/sm-scenes ($p<0.01$).
+In this section, we study the grades given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes et ni/rm-scenes.
+The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ exhibit a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/abscence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
 
-Ces résultats indiquent que la suppression des événements a effectivement modifié la perception des scènes par les sujets. Nos deux hypothèses sont ainsi vérifiées:
+%The \emph{post hoc} montre, quant à elle, des différences significatives entre tous les groupes d'observations, notamment entre les i/am- et i/sm-scenes ($p<0.05$) et les ni/am- et ni/sm-scenes ($p<0.01$).
+
+The \emph{post hoc} analysis exhibit significative differences between all the groups of observations, notably between the i/km-scenes ($p<0.05$) and the i/rm-scenes ($p<0.01$).
+
+%Ces résultats indiquent que la suppression des événements a effectivement modifié la perception des scènes par les sujets. Nos deux hypothèses sont ainsi vérifiées:
+
+These results indicate that the removal of the markers did indeed modify the perception of the scenes by the subjects. Our two hypotheses are thus verified:
+
+% \begin{itemize}
+% \item la suppression des ni-marqueurs a amélioré les qualités perçues des ni-scènes;
+% \item la suppression des i-marqueurs a diminué les qualités perçues des i-scènes.
+% \end{itemize}
 
 \begin{itemize}
-\item la suppression des ni-marqueurs a amélioré les qualités perçues des ni-scènes;
-\item la suppression des i-marqueurs a diminué les qualités perçues des i-scènes.
+\item the removal of the ni-markers improved the pleasantness of the ni-scenes;
+\item the removal of the i-markers reduced the pleasantness of the i-scenes.
 \end{itemize}
 
-L'interaction significative montre que l'effet du type d'environnement influe sur l'effet de l'absence/présence des marqueurs.
+%L'interaction significative montre que l'effet du type d'environnement influe sur l'effet de l'absence/présence des marqueurs.
 En effet la moyenne des écarts entre am- et sm-scènes est plus importante pour les ni-scènes ($1.1$) que pour les i-scènes ($0.5$).
 
-\subsection{Discussions}
+The significative interaction shows that the effect of the type of environment influences the effect of the presence/absence of the markers.
+Indeed, the average of the differences between km-scenes and rm-scenes is larger for the ni-scenes ($1.1$) than for the i-scenes ($0.5$).
 
-Cette expérience fait apparaître que la présence, dans une scène, des marqueurs relevés à l'expérience 1 impacte bien l'agrément perçu.
+\subsection{Discussion}
+
+%Cette expérience fait apparaître que la présence, dans une scène, des marqueurs relevés à l'expérience 1 impacte bien l'agrément perçu.
 La suppression des ni-marqueurs a un effet bénéfique sur le ressenti des ni-scènes, tandis que, plus surprenant, la suppression des i-marqueurs dégrade légèrement la qualité des i-scènes.
@@
 
-Les i-marqueurs ont donc bien un effet bénéfique sur la perception d'un environnement.
+This experiment shows that the presence of markers identified during the analysis of experiment 1 does have an impact on the percieved pleasantness.
+The removal of the ni-markers has a positive effect on the perception of the ni-scenes.
+Perhaps more surprisingly the removal of the i-markers slightly decreases the perception of the i-scenes.
+This is even more striking because, due to the removal of the markers, the acoustic pressure level of the i/km-scenes is higher than the one of the i/rm-scenes.
+
+%Les i-marqueurs ont donc bien un effet bénéfique sur la perception d'un environnement.
 Le fait que leur suppression diminue $\mathcal{A}_{scene}$ montre clairement qu'il est possible d'améliorer la qualité sonore d'un lieu en ajoutant des sons bien acceptés comme \emph{oiseau}.
 Ces conclusions vont dans le sens de l'approche positive introduite par Schafer \cite{schafer1977tuning}.
+
+So the i-markers do have a positive impact on the perception of an environment.
+The fact that their removal decreases $\mathcal{A}_{scene}$ clearly shows that it is possible to improve the perceived pleasantnes of a given place by the addition of sounds commonly considered as pleasant such as \emph{brid calls}.
+Those conclusions are in line with the positive approacj introduced by Schafer in \cite{schafer1977tuning}.
 

### 2017-05-12 lafaySimulatedScenePerception.tex
ml starts conclusion

@@
 
-\section{Conclusions et perspectives}
+\section{Outcomes for sounscape perception}
 
-Les expériences ont montré que la majorité des descripteurs utilisés, qu'ils soient sémantiques ou structurels, permettent de faire la distinction entre une scène idéale et une scène non-idéale.
+%Les expériences ont montré que la majorité des descripteurs utilisés, qu'ils soient sémantiques ou structurels, permettent de faire la distinction entre une scène idéale et une scène non-idéale.
 
-Cependant, nous observons que les caractéristiques physiques corrélées à l'agrément diffèrent clairement suivant le type de scènes.
+This series of three experiments showed that most of the descriptors used in this study, be they semantic or compositional, allows us to distinguish between a said ideal scene et a non ideal one.
+
+%Cependant, nous observons que les caractéristiques physiques corrélées à l'agrément diffèrent clairement suivant le type de scènes.
 Dans le cas des scènes idéales, c'est avant tout l'émergence de marqueurs sonores qui détermine la qualité perçue, alors que dans le cas des scènes non-idéales, c'est le niveau sonore global qui influe sur l'agrément.
 
-Ces résultats tendent à confirmer que la perception des qualités d'une scène dépend avant tout des sources sonores qui la composent, les caractéristiques structurelles mobilisées dans le processus perceptif semblant varier d'une source à l'autre, et d'un type d'environnement à l'autre.
+That being said, we observe that the physical caracteristics correlated with the perceived pleasantness clearly differ depending on the type of scenes.
+In the case of ideal scenes, it is above all the emergence of sound markers that determines the perceived quality, whereas in the case of non-ideal scenes, it is the overall sound level that proeminently influences the perceived quality.
+
+%Ces résultats tendent à confirmer que la perception des qualités d'une scène dépend avant tout des sources sonores qui la composent, les caractéristiques structurelles mobilisées dans le processus perceptif semblant varier d'une source à l'autre, et d'un type d'environnement à l'autre.
 Ce fait montre qu'il est illusoire d'envisager qu'un descripteur physique holistique puisse rendre compte, de manière pertinente, des qualités affectives de tous types d'environnement.
 
-Cet état de fait peut potentiellement influer sur les stratégies à adopter pour améliorer la qualité de l’environnement sonore:
+These results tend to confirm that the perception of the qualities of a scene depends above all on the sound sources that compose it, the compostional characteristics taken into accound during the perceptual process appearing to vary from one source to the other, from one type of environment to the other.
+This fact lead the authors to conclude that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.
+
+%Cet état de fait peut potentiellement influer sur les stratégies à adopter pour améliorer la qualité de l’environnement sonore:
+
+Those results may have an impact on the relevant strategies to adopt while trying to improve the quality of a sound environment:
+
+% \begin{itemize}
+% \item dans le cadre de scènes non-idéales, il s'agit de diminuer le niveau sonore, soit de manière globale, soit en agissant sur certaines sources (\emph{sirène}, \emph{klaxon});
+% \item dans le cadre de scènes idéales, il s'agit 1) d'identifier les sons agréables, \ie~les marqueurs sonores, 2) de baisser le niveau des autres sons, 3) voire, en restant dans la limite du raisonnable, d'augmenter le niveau des marqueurs par rapport aux autres sons.
+% \end{itemize}
 
 \begin{itemize}
-\item dans le cadre de scènes non-idéales, il s'agit de diminuer le niveau sonore, soit de manière globale, soit en agissant sur certaines sources (\emph{sirène}, \emph{klaxon});
-\item dans le cadre de scènes idéales, il s'agit 1) d'identifier les sons agréables, \ie~les marqueurs sonores, 2) de baisser le niveau des autres sons, 3) voire, en restant dans la limite du raisonnable, d'augmenter le niveau des marqueurs par rapport aux autres sons.
+\item in the case of non ideal scenes, one such focus on reducing the acoustic pressure level, wether globally, or by discarding specific sources such as \emph{sirens} or \emph{car horns}
+\item in the case of ideal scenes, one should first idenity which sources are pleasant to the targeted community, second lower the volume of the other sound sources, and, if possible, raise the contribution or add positive sound markers.
 \end{itemize}
 
-Ces travaux nous permettent enfin de conjecturer quant à la nature des représentations mentales des concepts «~environnement sonore urbain agréable~» (EA) et «~environnement sonore urbain désagréable~» (ED).
+%Ces travaux nous permettent enfin de conjecturer quant à la nature des représentations mentales des concepts «~environnement sonore urbain agréable~» (EA) et «~environnement sonore urbain désagréable~» (ED).
 
-Premièrement, le fait que les informations sémantiques (sources sonores présentes) et structurelles soient différentes pour les scènes idéales et non-idéales nous porte à croire que ces deux types d'informations caractérisent les concepts EA et ED.
+Finally, these works allow us to conjecture as to the nature of the mental representations of the concepts "pleasant urban sound environment" (EA) and "unpleasant urban sound environment" (ED).
 
-Deuxièmement, le fait que la suppression des marqueurs sonores modifie l'agrément perçu nous porte à croire que le concept abstrait lié à l'agrément (\eg.
+%Premièrement, le fait que les informations sémantiques (sources sonores présentes) et structurelles soient différentes pour les scènes idéales et non-idéales nous porte à croire que ces deux types d'informations caractérisent les concepts EA et ED.
+
+First, the fact that the semantic information (in terms of which sound sources are present) and compositional information are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the EA and ED concepts.
+
+%Deuxièmement, le fait que la suppression des marqueurs sonores modifie l'agrément perçu nous porte à croire que le concept abstrait lié à l'agrément (\eg.
 scène agréable) dépend de l'activation d'un réseau de concepts concrets liés aux sources (dans notre cas \emph{oiseau}, \emph{cloche} et \emph{sonnette vélo}).
 
-Troisièmement, le fait que les descripteurs structurels corrélés à l'agrément diffèrent en fonction du type de scènes nous porte à croire que les caractéristiques structurelles permettant de distinguer deux instances d'un même concept abstrait lié à l'agrément ne sont pas les mêmes pour EA et ED.
+Secondly, the fact that the removal of sound markers modifies the perceived pleasantness leads us to believe that the abstract concept related to pleasantness depends on the activation of a network of concrete concepts strongly linked to the sources which are in the case of this study: \emph{bird}, \emph{church bell} and \emph{bicycle bell}).
 
-Au terme de notre étude, nous pensons que la simulation est un outil dont le développement pourrait permettre aux décideurs en matière d'urbanisme d'interroger toute une communauté sur ses représentations propres des environnements sonores auxquels elle est exposée, et, pourquoi pas, sur les représentations des environnements sonores auxquels elle voudrait être exposée.
+%Troisièmement, le fait que les descripteurs structurels corrélés à l'agrément diffèrent en fonction du type de scènes nous porte à croire que les caractéristiques structurelles permettant de distinguer deux instances d'un même concept abstrait lié à l'agrément ne sont pas les mêmes pour EA et ED.
+
+Third, the fact that the compositional descriptors correlated with the percieved pleasantness change according to the type of scenes leads us to believe that the compositional characteristics allowing to distinguish two instances of the same abstract concept linked to the pleasantness are not the same for EA and ED.
+
+\section{Conclusion}
+
+%Au terme de notre étude, nous pensons que la simulation est un outil dont le développement pourrait permettre aux décideurs en matière d'urbanisme d'interroger toute une communauté sur ses représentations propres des environnements sonores auxquels elle est exposée, et, pourquoi pas, sur les représentations des environnements sonores auxquels elle voudrait être exposée.
+
+In the opiion of the authors, the study described in this paper demonstrate strongly the usefulness of considering simulation tools such as \emph{simScene} in order to scientifically question  the perception of soundscapes.
+We also believe that its wider usage could enable urban planning decision-makers to question an entire community about its own representations of the sound environment to which it is exposed, and about the representations of the sound environments to which it would like to be exposed.
 
@@
 
+Une des pistes privilégiée dans ce cadre serait d'étudier l'impact de l'ajout de marqueurs sonores positivement appréciés à une scène non idéal pour étudier l'hypothèse que ce type d'ajout améliorerait la qualité perçue de la scène.
+
 Enfin, l'on pourrait encore étudier l'influence des contextes socio-culturels sur la perception.
@@
 
-Outre les possibilités déjà évoquées, la simulation présente encore dans ce cas deux avantages:
+Outre les possibilités déjà évoquées, le protocole de simulation présenté ici ainsi que son implémentation apporte encore dans ce cas deux avantages:
 
@@
 
-Bien évidemment, ces approches nécessitent d'accroître la taille des banques de sons isolés disponibles, un effort conséquent, mais nécessaire, qui contribuera grandement aux nombreux domaines de recherche ayant trait aux scènes sonores.
+%Bien évidemment, ces approches nécessitent d'accroître la taille des banques de sons isolés disponibles, un effort conséquent, mais nécessaire, qui contribuera grandement aux nombreux domaines de recherche ayant trait aux scènes sonores.
 

### 2017-05-12 lafaySimulatedScenePerception.tex
ml is done for first pass

@@
 
-In the opiion of the authors, the study described in this paper demonstrate strongly the usefulness of considering simulation tools such as \emph{simScene} in order to scientifically question  the perception of soundscapes.
+In the opinion of the authors, the study described in this paper demonstrate the usefulness of considering simulation tools such as \emph{simScene} in order to scientifically question the perception of soundscapes.
 We also believe that its wider usage could enable urban planning decision-makers to question an entire community about its own representations of the sound environment to which it is exposed, and about the representations of the sound environments to which it would like to be exposed.
 
-Dans la continuité des travaux réalisés, il conviendrait de multiplier les expériences de simulation, en faisant varier les qualités affectives (calme, confortable, gênante, \etc), mais aussi en spécifiant des lieux particuliers (parc, place, rue, \etc), afin d'élaborer des corpus entiers de scènes cognitivement renseignées de paysages sonores.
+%Dans la continuité des travaux réalisés, il conviendrait de multiplier les expériences de simulation, en faisant varier les qualités affectives (calme, confortable, gênante, \etc), mais aussi en spécifiant des lieux particuliers (parc, place, rue, \etc), afin d'élaborer des corpus entiers de scènes cognitivement renseignées de paysages sonores.
 
-Il serait par ailleurs intéressant d'utiliser la simulation afin d'étudier plus avant les effets provoqués par la modification volontaire d'une caractéristique d'une scène, comme lors de la suppression des marqueurs sonores pratiquée dans l'expérience 2.
+Future work should consider several other simulation experiments by varying the emotional qualities (quiet, comfortable, troublesome, etc.), but also by specifying particular places (park, square, street, etc.), in order to provide to the scientific community entire corpora of cognitively informed soundscapes.
 
-Une des pistes privilégiée dans ce cadre serait d'étudier l'impact de l'ajout de marqueurs sonores positivement appréciés à une scène non idéal pour étudier l'hypothèse que ce type d'ajout améliorerait la qualité perçue de la scène.
+%Il serait par ailleurs intéressant d'utiliser la simulation afin d'étudier plus avant les effets provoqués par la modification volontaire d'une caractéristique d'une scène, comme lors de la suppression des marqueurs sonores pratiquée dans l'expérience 2.
 
-Enfin, l'on pourrait encore étudier l'influence des contextes socio-culturels sur la perception.
+It would also be interesting to use simulation in order to study further the effects caused by the voluntary modification of a characteristic of a scene, as during the suppression of the sound markers practiced in experiment 2.
+
+%Une des pistes privilégiée dans ce cadre serait d'étudier l'impact de l'ajout de marqueurs sonores positivement appréciés à une scène non idéal pour étudier l'hypothèse que ce type d'ajout améliorerait la qualité perçue de la scène.
+
+One interesting avenue in this vein would be to study the impact of adding positively appreciated sound markers to a non ideal scene in order to study the hypothesis that this type of addition would improve the perceived quality of the scene.
+
+%Enfin, l'on pourrait encore étudier l'influence des contextes socio-culturels sur la perception.
 Dans les faits, si le son de cloche est le plus souvent un marqueur d'environnement de qualité pour un occidental, cela ne se vérifie pas nécessairement auprès de sujets de culture orientale, moyen-orientale ou autre.
 
-Outre les possibilités déjà évoquées, le protocole de simulation présenté ici ainsi que son implémentation apporte encore dans ce cas deux avantages:
+Finally, one should study the influence of socio-cultural contexts on perception.
+Indeed, if the sound of the church bell is most often a quality environment marker for a Westerner, this is not necessarily true with subjects of Eastern, Middle Eastern or other cultures.
+
+%Outre les possibilités déjà évoquées, le protocole de simulation présenté ici ainsi que son implémentation apporte encore dans ce cas deux avantages:
+
+Once again, besides the interesting possibilities already mentioned, the simulation protocol presented here as well as its implementation brings in this case two important advantages:
+
+%\begin{itemize}
+% \item le simulateur peut être déployé à large échelle via internet du fait de l'architecture logicielle choisie pour son développement;
+% \item les scènes simulées peuvent être analysées sans avoir à tenir compte des différentes langues maternelles des sujets, la nature sémantique des classes de sons utilisées étant connue \emph{a priori} par l'expérimentateur et sans avoir besoin d'analyser la scène sonore pour idientifier les sources, leurs présences dans la scène étant directement exploitable.
+% \end{itemize}
 
 \begin{itemize}
-\item le simulateur peut être déployé à large échelle via internet du fait de l'architecture logicielle choisie pour son développement;
-\item les scènes simulées peuvent être analysées sans avoir à tenir compte des différentes langues maternelles des sujets, la nature sémantique des classes de sons utilisées étant connue \emph{a priori} par l'expérimentateur et sans avoir besoin d'analyser la scène sonore pour idientifier les sources, leurs présences dans la scène étant directement exploitable.
+\item The simulator can be deployed on a large scale via the Internet thanks to the chosen software architecture;
+\item simulated scenes can be analyzed without having to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurences in the simulated scene being directly available.
 \end{itemize}
@@
 
-\nm{a ce propos, il n'est pas fait mention des sons manquants pour les participants (cf. protocole expe de 1a);
-ne faudrait -il pas en dire 2 mots ce qui alimenterait aussi la discussion sur la dependance de la qualite (et l'efficacite) de la simulation a  la constitution de la base de donnees de textures et d'evenemetns ?}
+%\nm{a ce propos, il n'est pas fait mention des sons manquants pour les participants (cf. protocole expe de 1a);
+%ne faudrait -il pas en dire 2 mots ce qui alimenterait aussi la discussion sur la dependance de la qualite (et l'efficacite) de la simulation a  la constitution de la base de donnees de textures et d'evenemetns ?}
 

### 2017-05-12 lafaySimulatedScenePerception.tex
ml spell check done

@@
 
-Si, à l'inverse, on demande à un sujet de décrire un environnement donné (\eg~décrivez une scène sonore urbaine «~calme~») \cite{guastavino2006ideal, dubois2006cognitive}, on collecte une information riche, à la fois quantitative et sémantique, de la représentation qu'il se fait de cet environnement.
+%Si, à l'inverse, on demande à un sujet de décrire un environnement donné (\eg~décrivez une scène sonore urbaine «~calme~») \cite{guastavino2006ideal, dubois2006cognitive}, on collecte une information riche, à la fois quantitative et sémantique, de la représentation qu'il se fait de cet environnement.
 Néanmoins, en l'absence de données sonores, on ne peut la caractériser physiquement.
@@
 A large amount of quantitative and semantic information is then collected about the representation of this type of sound environment that the subject has.
-Unfortunately, without any reference to sound data, this representation cannot be caracterized physically.
+Unfortunately, without any reference to sound data, this representation cannot be characterized physically.
 
@@
 We believe that the use of such a device allows us to gain the benefits of the two above cited approaches.
-As the subject is saked to produces audio data (the signal of the simulated scene) allows the experimenter to study a modal version of the mental representation of the subject that is both caracterized semantically and physically, respectively in terms of the sound sources and their levels.
+As the subject is asked to produces audio data (the signal of the simulated scene) allows the experimenter to study a modal version of the mental representation of the subject that is both characterized semantically and physically, respectively in terms of the sound sources and their levels.
 
@@
 Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}.
-For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to very insteresting outcomes as it allows the experimenter to conveniently study separately the influence of the sounds sources that compose a sound environment of interest.
-Indeed, with such a material, not only the type of sound sources is available for study, but also the exact level and audio waveform for each source and also the structural characterestics of the scene, that is the temporal distribution of the events.
+For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to very interesting outcomes as it allows the experimenter to conveniently study separately the influence of the sounds sources that compose a sound environment of interest.
+Indeed, with such a material, not only the type of sound sources is available for study, but also the exact level and audio waveform for each source and also the structural characteristics of the scene, that is the temporal distribution of the events.
 
@@
 
-To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmanelta scenes.
-Results and outcomes of a series of 3 experiments that build on the use of the simulator are studied in order to better comprehend how different sound soucres typically present in a urban scene impact pleasantness:
+To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmental scenes.
+Results and outcomes of a series of 3 experiments that build on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:
 
@@
 
-A notre connaissance, seuls Bruce~\al \cite{bruce2009development,bruce2014effects} se sont servis de la simulation dans une apporche similaire.
+%A notre connaissance, seuls Bruce~\al \cite{bruce2009development,bruce2014effects} se sont servis de la simulation dans une apporche similaire.
 Ils proposent un outil permettant d'agir sur un environnement en ajoutant ou supprimant des sources sonores spécifiques d'une part, en modifiant le niveau sonore des sources, et leurs positions spatiales d'autre part.
@@
 They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location.
-The authors shows that the addition or removal of the sources follows more social or semantic matters than their acoustical caracteristics.
+The authors shows that the addition or removal of the sources follows more social or semantic matters than their acoustical characteristics.
 A lack of diversity is nonetheless mentioned by the authors that limit the outcomes of the study.
@@
 
-In our approach, the simulator developed for this study\footnote{simScene, available online \url{http://soundthings.org/research/}} gives a monophonic representation of the scene, with the added befenit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are, as much as possible, expressive and useful for analysis.
+In our approach, the simulator developed for this study\footnote{simScene, available online \url{http://soundthings.org/research/}} gives a monophonic representation of the scene, with the added benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are, as much as possible, expressive and useful for analysis.
 
@@
 
-The remaining of the paper is organized as follows: the soundscape simulator \emph{SimScene} is introduced in Section \ref{sec:simulator}.
-The three experiments,  1 .a \emph{simulation}, 1.b \emph{pleasantness evaluation}, and 2 \emph{pleasantness evaluation after modiciation of the scenes} are respectively presented in Sections \ref{sec:simulation}, \ref{sec:evaluation}, and \ref{sec:modification}.
+The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}.
+The three experiments,  1 .a \emph{simulation}, 1.b \emph{pleasantness evaluation}, and 2 \emph{pleasantness evaluation after modification of the scenes} are respectively presented in Sections \ref{sec:simulation}, \ref{sec:evaluation}, and \ref{sec:modification}.
 Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.
@@
 A track is then created for this sounds source within the simulator interface.
-The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurences of the sounds source.
+The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sounds source.
 Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.
@@
 
-Litterature usually distinguish between:
+Literature usually distinguish between:
 \begin{itemize}
-\item {a sound event}: an isolated sound, sparsely ditrubuted whose acoustical characteristics may change with respect to time
+\item {a sound event}: an isolated sound, sparsely distributed whose acoustical characteristics may change with respect to time
 \item {a sound texture}: an isolated sound of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
@@
 
-The simulation tool \emph{Simscene} follows this distinction and consider two disctint sound databases: one with the classes of events only and the other with classes of textures.
+The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures.
 These two types have specific simulation procedures.% et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures.
@@
 
-We do so, because we assume that the morphological differences between those two classes have some important consequences on the percpetion of the scenes.
+We do so, because we assume that the morphological differences between those two classes have some important consequences on the perception of the scenes.
 
@@
 
-Maffiolo \cite{maffiolo_caracterisation_1999} distinguish two distinct categorisation processes, that differentiate themselves with the capacity of the listener to identify sound events.
+Maffiolo \cite{maffiolo_caracterisation_1999} distinguish two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events.
 She shows that those processes leads to two abstract cognitive category respectively termed ""~the event sequences ~" and "~ the amorphous sequences".
@@
 
-We term sample, a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples percieved equivalently.
+We term sample, a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples perceived equivalently.
 
@@
 
-Orignal sounds have been recorded using a shotgun microphone \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} relié à un enregistreur \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}}.
+Original sounds have been recorded using a shotgun microphone \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} plugged to a \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}} recorder.
 The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background.
-Conversely, it allowed us to avoid dominant sound sources while recording texture sounds by pointing the microphone to distant eareas with no dominant sound sources.
+Conversely, it allowed us to avoid dominant sound sources while recording texture sounds by pointing the microphone to distant areas with no dominant sound sources.
 
@@
 
-All samples are normalised to the same $RMS$ level of $-12$ $dB$ FS, \ie. relative to Full Scale. In our case, the full scale level is of 1 Volt.
+All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie. relative to Full Scale. In our case, the full scale level is of 1 Volt.
 
@@
 
-By design, the simulation tool do not allow the user to interect and control a specific sample.
+By design, the simulation tool do not allow the user to interact and control a specific sample.
 Interaction is done at the track level, a track being a sequence of samples.
-Several paramters are available to the subject to control the track:
+Several parameters are available to the subject to control the track:
 
@@
 \item \emph{level} ($dB$): the levels are  randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
-\item \emph{inter-onset spacing} (second): for the event's track only, the spacings arerandomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
-\item \emph{start and end time} (second): the subject set the first and last occurence of a sample of the selected class.
+\item \emph{inter-onset spacing} (second): for the event's track only, the spacings are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
+\item \emph{start and end time} (second): the subject set the first and last occurrence of a sample of the selected class.
 \end{itemize}
@@
 
-The samples of texture are sequenced without time spacing, therefore the parameters \emph{event's fade} and \emph{inter-onset spacing} are not available ofr this kind of track.
+The samples of texture are sequenced without time spacing, therefore the parameters \emph{event's fade} and \emph{inter-onset spacing} are not available for this kind of track.
 
@@
 
-The previous experiment demonstrated that, among the classes of sounds occuring in urban soundscapes, those with markers are specific of some of environments.
+The previous experiment demonstrated that, among the classes of sounds occurring in urban soundscapes, those with markers are specific of some of environments.
 Those sound markers seem to have a great impact on perception.
@@
 
-In order to verify that the pleasantness of the ideal and non ideal scenes if affected by the sound markers, the audio waveforms of the simulated scenes are regenerated without considering the classes identified as markers.
+In order to verify that the pleasantness of the ideal and non ideal scenes if affected by the sound markers, the audio waveform of the simulated scenes are regenerated without considering the classes identified as markers.
 For the i-scenes, the i-markers are removed.
@@
 
-Thus the objective of this experiment is to study if the removal of the previously identified markers have an impact on the percieved pleasantness. Two hypothesis are formulated:
+Thus the objective of this experiment is to study if the removal of the previously identified markers have an impact on the perceived pleasantness. Two hypothesis are formulated:
 
@@
 
-\subsection{Planification of experiment 2}
+\subsection{Planning of experiment 2}
 
@@
 
-There are 144 stimuli of 30 seconds durations. More precisely:
+There are 144 stimuli of 30 seconds duration. More precisely:
 
@@
 The 36 ideal scenes with markers are noted i/km-scenes, and the 36 non ideal scenes with markers are noted ni/km-scenes.
-\item \emph{72 rm-scènes}:  72 scenes where the sound classes identified as markers are removed.
+\item \emph{72 rm-scenes}:  72 scenes where the sound classes identified as markers are removed.
 The 36 ideal scenes without markers are noted i/rm-scenes, and the 36 non ideal scenes without markers are noted ni/rm-scenes.
@@
 
-Notwithstanding the presence or absence of markers, the km-scenes and rm-scenes are exactely the same.
+Notwithstanding the presence or absence of markers, the km-scenes and rm-scenes are exactly the same.
 
@@
 \begin{itemize}
-\item \emph{Church bell}, \emph{bicycle bell}, \emph{animals} pour les i/sm-scènes;
-\item \emph{siren}, \emph{car horn} pour les ni/sm-scènes.
+\item \emph{Church bell}, \emph{bicycle bell}, \emph{animals} for the i/sm-scenes;
+\item \emph{siren}, \emph{car horn} for the ni/sm-scenes.
 \end{itemize}
@@
 
-Les sujets évaluent les 144 scènes.
+%Les sujets évaluent les 144 scènes.
 L'évaluation s'effectue sur une échelle sémantique bipolaire de 11 points allant de -5 (non-idéale/très désagréable) à +5 (idéale/très agréable).
@@
 
-Pour chaque sujet, les scènes sont présentées dans un ordre aléatoire.
+%Pour chaque sujet, les scènes sont présentées dans un ordre aléatoire.
 Les 10 premières scènes permettent au sujet de calibrer ses notes.
@@
 
-L'expérience est prévue pour durer 1 heure. Les sujets ne connaissent pas la nature des scènes.
+%L'expérience est prévue pour durer 1 heure. Les sujets ne connaissent pas la nature des scènes.
+
+The subjects evaluate the 144 scenes.
+The evaluation is done on an 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant).
+Before grading a scene, subjects must listen to the first 20 seconds.
+After scoring, they are free to move on to the next scene.
+
+For each subject, the scenes are presented in random order.
+The first 10 scenes allow the subject to calibrate his/her grades.
+Those calibration scenes are  5 i/km-scenes and 5 ni/km-scenes.
+These first 10 scenes are replayed at the end of the experiment, and only the notes given at the second occurrence are taken into account.
+
+The experiment is scheduled to last 1 hour. The subjects do not know the nature of the scenes.
 
@@
 
-12 subjects (4 womens) perform the test.
+12 subjects (4 women) perform the test.
 None of them have performed experiments 1.a and 1.b.
@@
 All the subjects live in Nantes, France, for at least two years.
-All the subjects performed the test succesfully.
+All the subjects performed the test successfully.
 
@@
 
-The type of data analysed in this experiment have been considered for experiment 1.a, see Section~\ref{sec:xp1_dataAna} for details.
+The type of data analyzed in this experiment have been considered for experiment 1.a, see Section~\ref{sec:xp1_dataAna} for details.
 
@@
 To do so, we perform a variance analysis.
-We consider $\mathcal{A}_{sujet}$ as a dependant variable, and as independant variables, the type of envorinment (i/ni), and the presence/absence of markers (km/rm).
-As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used in order to verify the hypothesis that there exist a significant difference of percieved pleasantness between km-scenes and rm-scenes.
-The two independant variables are cosidered as \emph{within-subject} factors.
-The factors being of only two levels each (type: i/ni; marker: km/rm), the sphericity impothesis does not need to be checked.
+We consider $\mathcal{A}_{sujet}$ as a dependent variable, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm).
+As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used in order to verify the hypothesis that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes.
+The two independent variables are considered as \emph{within-subject} factors.
+The factors being of only two levels each (type: i/ni; marker: km/rm), the sphericity hypothesis does not need to be checked.
 \emph{Post hoc} analysis are done by following the Tukey-Kramer procedure.
@@
 In this section, we study the grades given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes et ni/rm-scenes.
-The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ exhibit a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/abscence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
+The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ exhibit a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
 
@@
 
-This experiment shows that the presence of markers identified during the analysis of experiment 1 does have an impact on the percieved pleasantness.
+This experiment shows that the presence of markers identified during the analysis of experiment 1 does have an impact on the perceived pleasantness.
 The removal of the ni-markers has a positive effect on the perception of the ni-scenes.
@@
 
-This series of three experiments showed that most of the descriptors used in this study, be they semantic or compositional, allows us to distinguish between a said ideal scene et a non ideal one.
+This series of three experiments showed that most of the descriptors used in this study, be they semantic or compositional, allows us to distinguish between a said ideal scene and a non ideal one.
 
@@
 
-That being said, we observe that the physical caracteristics correlated with the perceived pleasantness clearly differ depending on the type of scenes.
-In the case of ideal scenes, it is above all the emergence of sound markers that determines the perceived quality, whereas in the case of non-ideal scenes, it is the overall sound level that proeminently influences the perceived quality.
+That being said, we observe that the physical characteristics correlated with the perceived pleasantness clearly differ depending on the type of scenes.
+In the case of ideal scenes, it is above all the emergence of sound markers that determines the perceived quality, whereas in the case of non-ideal scenes, it is the overall sound level that prominently influences the perceived quality.
 
@@
 
-These results tend to confirm that the perception of the qualities of a scene depends above all on the sound sources that compose it, the compostional characteristics taken into accound during the perceptual process appearing to vary from one source to the other, from one type of environment to the other.
+These results tend to confirm that the perception of the qualities of a scene depends above all on the sound sources that compose it, the compostional characteristics taken into account during the perceptual process appearing to vary from one source to the other, from one type of environment to the other.
 This fact lead the authors to conclude that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.
@@
 \begin{itemize}
-\item in the case of non ideal scenes, one such focus on reducing the acoustic pressure level, wether globally, or by discarding specific sources such as \emph{sirens} or \emph{car horns}
-\item in the case of ideal scenes, one should first idenity which sources are pleasant to the targeted community, second lower the volume of the other sound sources, and, if possible, raise the contribution or add positive sound markers.
+\item in the case of non ideal scenes, one such focus on reducing the acoustic pressure level, whether globally, or by discarding specific sources such as \emph{sirens} or \emph{car horns}
+\item in the case of ideal scenes, one should first identity which sources are pleasant to the targeted community, second lower the volume of the other sound sources, and, if possible, raise the contribution or add positive sound markers.
 \end{itemize}
@@
 
-Third, the fact that the compositional descriptors correlated with the percieved pleasantness change according to the type of scenes leads us to believe that the compositional characteristics allowing to distinguish two instances of the same abstract concept linked to the pleasantness are not the same for EA and ED.
+Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics allowing to distinguish two instances of the same abstract concept linked to the pleasantness are not the same for EA and ED.
 
@@
 \item The simulator can be deployed on a large scale via the Internet thanks to the chosen software architecture;
-\item simulated scenes can be analyzed without having to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurences in the simulated scene being directly available.
+\item simulated scenes can be analyzed without having to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
 \end{itemize}

### 2017-05-15 lafaySimulatedScenePerception.tex
ml is done

@@
 \Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process.
-Subjects are asked to create a full soundscape by means of a software dedicated to sound edition, coupled with a structured sound data set.
+Subjects are asked to create a full soundscape by means of a software tool dedicated to sound edition, coupled with a structured sound data set.
 This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes.
@@
 
-First, a subject can be asked to describe a given sound environnment following a given perceptive scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}, the amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}.
-That being said, as those stimuli can be recordered and analysed.
+First, a subject can be asked to describe a given sound environnment following a given perceptive scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}.
+The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}.
+That being said, as those stimuli can be recordered and analysed, it allow the experimenter to gather a coarse grain physical description of the stimuli.
 
@@
 
-We propose in this paper to consider the use of a soundscape simulator that the subject can use to objectify the representation of a given sound environment.
+We propose in this paper to consider the use of a soundscape simulator that the subject can use to objectify her representation of a given sound environment.
 We believe that the use of such a device allows us to gain the benefits of the two above cited approaches.
-As the subject is asked to produces audio data (the signal of the simulated scene) allows the experimenter to study a modal version of the mental representation of the subject that is both characterized semantically and physically, respectively in terms of the sound sources and their levels.
+As the subject is asked to produces audio data (the signal of the simulated scene), it allows the experimenter to study a precise modal version of the mental representation of the subject that is both characterized semantically and physically, respectively in terms of the sound sources and their levels.
 
@@
 
-Recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
+We believe that the availability of a fine grain description of the sound stimuli is of great insterest, beceause recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
 szeremeta2009analysis}.
 Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}.
-For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to very interesting outcomes as it allows the experimenter to conveniently study separately the influence of the sounds sources that compose a sound environment of interest.
+
+For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to very interesting outcomes as it allows the experimenter to separately study  the influence of the sounds sources that compose a sound environment of interest.
 Indeed, with such a material, not only the type of sound sources is available for study, but also the exact level and audio waveform for each source and also the structural characteristics of the scene, that is the temporal distribution of the events.
@@
 To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmental scenes.
-Results and outcomes of a series of 3 experiments that build on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:
+Results and outcomes of a series of three experiments that build on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:
 
@@
 They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location.
-The authors shows that the addition or removal of the sources follows more social or semantic matters than their acoustical characteristics.
+The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics.
 A lack of diversity is nonetheless mentioned by the authors that limit the outcomes of the study.
@@
 The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}.
-The three experiments,  1 .a \emph{simulation}, 1.b \emph{pleasantness evaluation}, and 2 \emph{pleasantness evaluation after modification of the scenes} are respectively presented in Sections \ref{sec:simulation}, \ref{sec:evaluation}, and \ref{sec:modification}.
+The three experiments,  1.a \emph{simulation}, 1.b \emph{pleasantness evaluation}, and 2 \emph{pleasantness evaluation after modification of the scenes} are respectively presented in Sections \ref{sec:simulation}, \ref{sec:evaluation}, and \ref{sec:modification}.
 Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.
+Experiment 1.a is based on a pilot study described in \cite{lafay2014new}.
 
@@
 
-L'expérience 1 .a a fait l'objet d'une étude pilote dont les résultats sont présentés dans \cite{lafay2013atiam,lafay2014new}.
+%L'expérience 1.a a fait l'objet d'une étude pilote dont les résultats sont présentés dans \cite{lafay2013atiam,lafay2014new}.
 
@@
 It has been designed to run on the popular web browers \emph{Chrome} and \emph{Firefox}.
-It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio} : \url{www.w3.org/TR/webaudio}}.
-The interface for selecting the soud sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.
+It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}.
+The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.
 
@@
 
-\emph{Simscene} is designed to be used as a simplified audio sequencer, with sequencing parameters specifically chosen for the generation of realistic soundscapes.
+\emph{Simscene} is designed as a simplified audio sequencer, with sequencing parameters specifically chosen for the generation of realistic soundscapes.
 To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}.
@@
 
-In order to provide the user with a sound database that is well organized and covers as much as possible the variety of sound sources that are available in urban areas, a typology of urban sounds is first agreed upon.
+In order to provide the user with a sound database that is well organized and covers as much as possible the variety of sound sources that are available in urban areas, a typology of urban sounds is first defined.
 
@@
 The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures.
-These two types have specific simulation procedures.% et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures.
-Événements et textures bénéficient de processus de simulation spécifiques.
+These two types have specific simulation procedures.
+
+% et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures. Événements et textures bénéficient de processus de simulation spécifiques.
 
@@
 
-We do so, because we assume that the morphological differences between those two classes have some important consequences on the perception of the scenes.
+We do so because we assume that the morphological differences between those two classes have some important consequences on the perception of the scenes.
 
@@
 
-According to the Auditory Scene Analysis (ASA) theory, the brain processes sperately events emerging from different sources of an sound environment \cite{bregman1994auditory}.
+According to the Auditory Scene Analysis (ASA) theory, the brain processes seperately events emerging from different sources of a sound environment \cite{bregman1994auditory}.
 Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}.
-In this case, the source is a perceptual entity.
-Indeed, events that are not isolated are grouped into a stream.
+In this case, the source is a perceptual entity as events that are not isolated are grouped into a stream.
 
@@
 Maffiolo \cite{maffiolo_caracterisation_1999} distinguish two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events.
-She shows that those processes leads to two abstract cognitive category respectively termed ""~the event sequences ~" and "~ the amorphous sequences".
+She shows that those processes leads to two abstract cognitive category respectively termed "the event sequences" and "the amorphous sequences".
 
@@
 
-The event sequences are composed of salient and identifiable events such as \emph{car start}, \emph{mal speech}.
+The event sequences are composed of salient and identifiable events such as \emph{car start}, \emph{male speech}.
 They arise from a descriptive analysis based on the identification of the sound sources.
@@
 
-Concerning sound textures, \ie~sounds that have stable caracteristics along time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrate that the human brain can decide not to abstract the physical information available using some of the summary statistics discarding lower level information \cite{nelken2013ear}.
+Concerning sound textures, \ie~sounds that have stable caracteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can decide not to abstract the physical information available using some of the summary statistics discarding lower level information \cite{nelken2013ear}.
 
-Au regard de ces résultats, il apparaît que le processus de traitement de l'information sonore comprend une prise de décision quant à la nature des stimuli \cite{nelken2013ear,mcdermott2013summary}, laquelle va ensuite influer sur la manière d'extraire et d'analyser l'information.
+%Au regard de ces résultats, il apparaît que le processus de traitement de l'information sonore comprend une prise de décision quant à la nature des stimuli \cite{nelken2013ear,mcdermott2013summary}, laquelle va ensuite influer sur la manière d'extraire et d'analyser l'information.
 
 At the light of those results, it appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli \cite{nelken2013ear,mcdermott2013summary}.
-We thus consider in this study the soundscape as "skeleton of events on a bed of textures" as coined in \cite{nelken2013ear}.
+We thus consider in this study the soundscape as a "skeleton of events on a bed of textures" as coined in \cite{nelken2013ear}.
 
@@
 
-The sound classes are organized hierachically (\cf~Figure~\ref{fig:orgDb}) according to a structure close to the one of the vertical axis of the categorical organization proposed by E.
+The sound classes are organized hierachically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E.
 Rosch \cite{rosch1978cognition}.
-In this taxonomy, the lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources.
-For class with high level of abstraction with sub classes, its collection of samples is the union of the collections of the sub classes.
+The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources.
+For classes with high level of abstraction with sub classes, its collection of samples is the union of the collections of the sub classes.
 
@@
 
-483 isolated sounds are collected and organized with the two above discussed typologies, 381 events and 102 textures.
+483 isolated sounds are collected and organized with the two typologies discussed above, 381 events and 102 textures.
 Among those samples, 332 have been recorded and 151 have been taken from two sound libraries: \emph{SoundIdeas}\footnote{\emph{SoundIdeas}: \url{www.sound-ideas.com}} and \emph{Universal SoundBank}\footnote{\emph{Universal SoundBank}: \url{www.universal-soundbank.com}}.
@@
 The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background.
-Conversely, it allowed us to avoid dominant sound sources while recording texture sounds by pointing the microphone to distant areas with no dominant sound sources.
+It also allowed us to avoid dominant sound sources while recording texture sounds by aiming the microphone to distant areas with no dominant sound sources.
 
@@
 \item \emph{event's fades} (seconds): for the event's track only, the subject can set a fade in / fade out duration applied to each samples;
-\item \emph{global fades} (seconds): the subject set the duration of the fade in /fade out at the beginning and the end of the track.
+\item \emph{global fades} (seconds): the subject set the duration of the fade in / fade out at the beginning and the end of the track.
 \end{itemize}
@@
 
+Once the typology and the set sound are available, an important design question that remains is to find a suitable way of presenting the data to the user.
 Most of browsing tools are based on keyword indexing.
-For sensory experiment studying the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject.
-Indeed, the subject will select \textit{a priori} sound referenced belonging to \emph{park} to build a \emph{calm} soundscape.
+However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject.
+Indeed, the subject may select \textit{a priori} sound referenced belonging to \emph{park} to build a \emph{calm} soundscape.
 
@@
 
-Therefore, the selection interface is text free and composed of spatially organized circles on a two dimensional plane.
+Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a two dimensional plane.
 
@@
         {\includegraphics[width=\columnwidth]{gfx/ch_4/simscene}\label{fig:simscene}}
-       \caption{Visual interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{Simscene} tool.}
+       \caption{Graphical interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{simScene} tool.}
 \end{figure}
@@
 As actual amplitude and spacing values are drawn from random variables, each time the subject changes the value of a parameter, the location and height of the rectangles are updated to reflect the changes in the sequencing of the samples.
-The subject can listen to the resulting sound scene.
+The subject can listen to the resulting sound scene at any time.
 
@@
 
-The previous experiment demonstrated that, among the classes of sounds occurring in urban soundscapes, those with markers are specific of some of environments.
+The previous experiment demonstrated that, among the classes of sounds occurring in urban soundscapes, those with markers are specific to some of environments.
 Those sound markers seem to have a great impact on perception.
-This impact is studied in more detail using an added benefit of the simulation paradigm proposed in this paper, the ability to manipulate and modify the generated scenes.
+This impact is studied here in more detail using an added benefit of the simulation paradigm proposed in this paper, the ability to manipulate and modify the generated scenes.
 
@@
 
-In order to verify that the pleasantness of the ideal and non ideal scenes if affected by the sound markers, the audio waveform of the simulated scenes are regenerated without considering the classes identified as markers.
-For the i-scenes, the i-markers are removed.
-For the ni-scenes, the ni-markers are removed.
-A perceptive test evaluating the pleasantness is conducted with a protocol close to the one considered in experiment 1.b.
+In order to verify that the pleasantness of the ideal and non ideal scenes if affected by the sound markers, the audio waveform of the simulated scenes are regenerated with or without considering the classes identified as markers.
+For the i-scenes, some of the i-markers are removed.
+For the ni-scenes, some of the ni-markers are removed.
+A perceptive test evaluating the pleasantness is then conducted with a protocol close to the one considered in experiment 1.b.
 
@@
 
-Thus the objective of this experiment is to study if the removal of the previously identified markers have an impact on the perceived pleasantness. Two hypothesis are formulated:
+Thus the objective of this experiment is to study if the removal of the previously identified markers have an impact on the perceived pleasantness. Two hypothesis are thus formulated:
 
@@
 \begin{itemize}
-\item \emph{Church bell}, \emph{bicycle bell}, \emph{animals} for the i/sm-scenes;
-\item \emph{siren}, \emph{car horn} for the ni/sm-scenes.
+\item \emph{Church bell}, \emph{bicycle bell}, and \emph{animals} for the i/rm-scenes;
+\item \emph{siren}, \emph{car horn} for the ni/rm-scenes.
 \end{itemize}
@@
 The subjects evaluate the 144 scenes.
-The evaluation is done on an 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant).
-Before grading a scene, subjects must listen to the first 20 seconds.
+The evaluation is done on a 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant).
+Before scoring a scene, subjects must listen to the first 20 seconds.
 After scoring, they are free to move on to the next scene.
 
-For each subject, the scenes are presented in random order.
-The first 10 scenes allow the subject to calibrate his/her grades.
+For each subject, the scenes are presented in a random order.
+The first 10 scenes allow the subject to calibrate their scores.
 Those calibration scenes are  5 i/km-scenes and 5 ni/km-scenes.
@@
 We consider $\mathcal{A}_{sujet}$ as a dependent variable, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm).
-As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used in order to verify the hypothesis that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes.
+As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to check the validity of the hypothesis that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes.
 The two independent variables are considered as \emph{within-subject} factors.
@@
 This subject evaluated positively close to half of the ni/km-scenes, see Annex~\ref{app:xp2}, subject 7.
-This subject gave a grade above 0 for 58\% of the ni/km-scenes, contrary to an average of 11\% for the remaining of the subjects.
+This subject also gave a score above 0 for 58\% of the ni/km-scenes, contrary to an average of 11\% for the remaining of the subjects.
 
-Furthermore, this subject used the whole scale (from -5 to 5) to grade both the i-scenes and the ni-scenes.
-Those behaviors strongly differing with the ones of the others subjects be they from this experiment or the previous ones, we decide to discard the subject 7 from the analysis.
+Furthermore, this subject used the whole scale (from -5 to 5) to score both the i-scenes and the ni-scenes.
+Those behaviors strongly differing with the ones of the others subjects be they from this experiment or the previous ones, subject 7 is discarded from the analysis.
 
@@
 
-In this section, we study the grades given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes et ni/rm-scenes.
-The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ exhibit a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
+In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes et ni/rm-scenes.
+The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ show a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
 
@@
 
-So the i-markers do have a positive impact on the perception of an environment.
-The fact that their removal decreases $\mathcal{A}_{scene}$ clearly shows that it is possible to improve the perceived pleasantnes of a given place by the addition of sounds commonly considered as pleasant such as \emph{brid calls}.
+The i-markers do have a positive impact on the perception of an environment.
+The fact that their removal decreases $\mathcal{A}_{scene}$ clearly shows that it is possible to improve the perceived pleasantnes of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
 Those conclusions are in line with the positive approacj introduced by Schafer in \cite{schafer1977tuning}.
@@
 
-These results tend to confirm that the perception of the qualities of a scene depends above all on the sound sources that compose it, the compostional characteristics taken into account during the perceptual process appearing to vary from one source to the other, from one type of environment to the other.
+These results tend to confirm that the perception of the qualities of a scene indeed depends above all on the sound sources that compose it.
+The compositional characteristics that are taken into account during the perceptual process appear to vary from one source to the other, from one type of environment to the other.
 This fact lead the authors to conclude that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.
@@
 \begin{itemize}
-\item in the case of non ideal scenes, one such focus on reducing the acoustic pressure level, whether globally, or by discarding specific sources such as \emph{sirens} or \emph{car horns}
+\item in the case of non ideal scenes, one should focus on reducing the acoustic pressure level, whether globally, or by discarding specific sources such as \emph{sirens} or \emph{car horns}
 \item in the case of ideal scenes, one should first identity which sources are pleasant to the targeted community, second lower the volume of the other sound sources, and, if possible, raise the contribution or add positive sound markers.
@@
 
-First, the fact that the semantic information (in terms of which sound sources are present) and compositional information are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the EA and ED concepts.
+First, the fact that the semantic information (in terms of which sound source is present) and compositional information are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the EA and ED concepts.
 
@@
 
-Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics allowing to distinguish two instances of the same abstract concept linked to the pleasantness are not the same for EA and ED.
+Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics that allow us to distinguish between two instances of the same abstract concept linked to the pleasantness are not the same for EA and ED.
 
@@
 
-In the opinion of the authors, the study described in this paper demonstrate the usefulness of considering simulation tools such as \emph{simScene} in order to scientifically question the perception of soundscapes.
+The outcomes of the three experiments described in this paper demonstrate the usefulness of considering simulation tools such as \emph{simScene} in order to scientifically question the perception of soundscapes.
 We also believe that its wider usage could enable urban planning decision-makers to question an entire community about its own representations of the sound environment to which it is exposed, and about the representations of the sound environments to which it would like to be exposed.
@@
 
-It would also be interesting to use simulation in order to study further the effects caused by the voluntary modification of a characteristic of a scene, as during the suppression of the sound markers practiced in experiment 2.
+It would also be interesting to use the alteration capabilities of the proposed paradigm in order to study further the effects caused by the voluntary modification of a characteristic of a scene, as during the suppression of the sound markers practiced in experiment 2.
 
@@
 
-One interesting avenue in this vein would be to study the impact of adding positively appreciated sound markers to a non ideal scene in order to study the hypothesis that this type of addition would improve the perceived quality of the scene.
+One interesting avenue in this direction would be to study the impact of adding positively appreciated sound markers to a non ideal scene in order to study the hypothesis that this type of addition would improve the perceived quality of the scene.
 

### 2017-07-04 NM.tex
mid way to inlcuding NM translation

3. Experiment 1
3.1 Goals
Experiment 1 aims at using the simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness. For that, two first experiments are planned (Figure 3): • experiment 1.a (simulation): during this experiment, participants have to create urban sound environments, using the simulation tool Simscene (see sect. 2). Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant. • experiment 2.a (evaluation): actually, after the simulation phase, we only have a binary information on the emotional properties of the respective scenes: ideal (i) and non ideal (ni). This second experimental step aims at investigating our knowledge on pleasantness. For that, a second goup of participants are asked to evaluate the pleasantness of each simulated scene, on a semantic scale. This evaluation protocol has two goals: 1. to evaluate the respective influence on the pleasantness of the various sources composing the scenes (i or ni) 2. to detect the presence of outliers or ambiguous samples. For the remainder of the study, the defined hedonic properties (i and ni) are used as ground truth. Then, we must have to be sure that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene, in terms of pleasantness rating.
Our analysis is based on the data collected by these two experiments (1.a, 1.b)
[FIGURE 3] Figure 3: experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b)
3.2 Experiment 1.a
Procedure The participants have to simulate two urban sound environments of 1 minute each. They have to follow the instructions below: • 1st simulation: create a plausible urban soundscape which is ideal, according to you (where you would like to live); • 2nd simulation: create a plausible urban soundscape which is non ideal, according to you (where you would not like to live);
All the participants start by working on the ideal environment; they read the second set of instruction at the end of the first experiment. Participants are completely free of their choice concerning sounds and synthesis parameters. They must nevertheless respect the two following constraints : • the listening point of view is the one of a fixed listener; • the soundscape must be realistic, i.e. physically plausible. For instance, the participant can insert ten dogs in the soundscape but he/she can’t insert one dog barking every 10 milliseconds. These constraints are notified in the instructions; no control is done a priori in the simulation user interface
Each simulation process contains two parts : 1. composition of the simulation: this step is split in three steps • select the class of sounds • give a name to the selected class of sounds • set the parameters of the tracks (see section 2.2) related to the selected class of sounds. 2. production of a free verbalization of the composed soundscape
In addition, once the two sound scenes have been realized, participants are invited to • point out the sound sources that were missing for the composition • give a comment about the ergonomics of the simulation environment • give a comment about the ergonomics of the selection tool
Before starting the first simulation, a 20-minute tutorial is given to the participants in order to familiarize them with the simulation environment and the sound database. The experiment is planned to last about two hours and a half, including breaks that could be taken by the participants.
Apparatus All the participants do the experiment on similar computers. The audio files are played in diotic conditions by headphones. During the tutorial, participants are asked to adjust the sound level in order to be comfortable; they can’t modify it in the following of the experiment. All the participants do the experiment simultaneously. They are equally split in three identical quiet rooms. They can’t talk together during the experiment. Three experimenters (one in each room) are present during the whole duration of the experiment in order to check that everything goes well and possibly to answer questions.
Participants 44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in Nantes downtown since 2 years or more. Among the 44 participants, 40 succeeded in doing the experiment, producing in the end, 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes). 4 participants were excluded from the process for not having respect or understand the instructions, or for having exceed the allowed time.
3.3 Experiment 1.b
Procedure Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes composed in Exp. 1.a (from sec. 15 to sec. 45). The assessment is done in a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant). Before evaluating a scene, participants must listen to the first 2à seconds of the scene. After the evaluation, they are free to switch to the next scene. For each participant, sound scenes are played in a random order. The first 10 scenes allow the participants to calibrate their marks; they imperatively gather 5 ideal scenes and 5 non ideal scenes. These first 10 scenes are played back again at the end of the experiment and the corresponding evaluations are the ones that are only taken into account in the results. The experiment is planned to last 30 minutes. The participants don’t know anything beforehand about the nature of the sound scenes.
Apparatus All the participants do the experiment on similar computers. The audio files are played in diotic conditions by semi-open headphones (Beyerdynamic DT 990 Pro). All the sound scenes have been re-composed on the basis of the scores obtained in the simulation experiment (Exp. 1.a). The output sound level is the same for all the participants. All the participants do the experiment simultaneously. They can’t talk together during the experiment. An experimenter is present during the whole duration of the experiment in order to check that everything goes well and possibly to answer questions.
Participants 10 students (2 females - mean age 23.1, SD = 1.8) from Ecole Centrale de Nantes took part to the experiment. They are all living in Nantes since 2 years or more. None of them took part to the previous simulation experiment (Exp. 1.a). All the participants succeeded in doing the experiment.
3.4 Data and analysis methods
A set of features is attached to each sound scene. The analysis is done on the basis of these features. A summary of these features (and the corresponding acronyms) is presented in Table I. In order to stay coherent with the evaluation experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version (see Section 3.3). For each sound scene, three types of features are considered : • perceptual: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale. $\mathcal{A}_{scene}$ is the mean pleasantness for one scene, as the average of the marks of all the participants. In the same idea, $\mathcal{A}_{subject}$ is the mean pleasantness for each participant, as the average of all the marks of each participant. Given the low volume of participants in Exp. 1.b, we decide not to normalize the values of pleasantness marks. • semantic: boolean vector S = (x1, x2, …, xn) that indicates the class of sounds involved in the scene. Each point x of this vector corresponds to a specifi class of sounds: x = 1 if the class is present in the scene, and x = 0 otherwise. The vector dimension (n) depends on the level of abstraction that is considered. For instance, for the abstraction level 1, including 44 classes of sounds, n = 44. • structural: related to sound level. To figure out the sound level, we got inspired from the $L_{Aeq}$ feature. In our case, it is a scalar computed from the signal (in Volt and not in Pascal), and converted in deciBels, with a reference of 1 Volt. The level is obtained by computing every second the quadratic mean of the signal and by averaging over the total duration of the scene. An A-filtering module is also processed before the quadratic means are computed. We note L, $L(E)$ and $L(T)$ the computed levels by taking into account respectively the whole set of samples, only the set of event samples and only the set of texture samples.
In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following tests of significativity: • check of the pleasantness of the simulated sound scene: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with regards to the two types of scene. the significativity is evaluated by a Student test on two independant populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$ ; • analysis of the structural features: in order to evaluate wether the emotional difference forced to the i- and ni- scenes has a significative influence – i.e., if significative differences exist between structural features and perceived pleasantness –, we measure this significativity with a Student test on two populations; • study of the influence of the structural features on the perceived pleasantness: in order to evaluate a possible impact of the structural features on the perceived pleasantness, we look at linear correlations between these features and $\mathcal{A}_{scene}$. For that, we use the Pearson correlation coefficient, on the basis of the standard methodology used in multidimensional studies; • analysis of the semantic features: in order to determine if the forced emotional difference has an influence on the scenes composition in terms of sound sources, or if specific souns classes are more frequently used to simulate a given environment (i or ni), we apply a V-test. This test is done at every level of abstraction, and separately, for texture and event classes. For each class j, and each environment type k (k = {i, ni}), the V-test variable Vjk is computed as follows:
[EQUATION]
where c is the amount of classes, ck is the amount of classes for a type of environment k, cj the amount of j-classes, and Cjk the amount of j-classes used for an environment k. The V-test tests the null hypothesis that the ratio cjk / c is not significatively different from the ratio cjk / ck. For a given environment k and class j, if the hypothesis is rejected, the class j can then be considered typical with regards to the environment k. The typical classes are called sound markers. • study of the representation spaces induced by the semantic features: in order to determine if a representation only based on the presence / absence of sound sources allows to split the two types of environment, we investigate the space induced by the semantic features S. S is a boolean vector and the distance between two scenes is computed with the Hamming distance. Considering two n-dimension vectors S1 = (x1,1, x1,2, …, x1,n) and S2 = (x2,1, x2,2, …, x2,n) where x = {0,1}, the Hamming distance dham measures the proportion of coordinates that differ between two vectors, as follows:
[EQUATION]
where ✇ designate the exclusive-or operator. The more the two scenes are similar, the closer they are. The use of Hamming distance allows to take into account equally the present and absent classes. For measuring the intrinsic ability of the space to split the i- and ni- scene, we use a ranking metric named k- precision ($P@k$) . The $P@k$ computes the precision obtained after k items have been found. Formally, for each si scene, we compute the ratio between the number of sj scenes taken among the k nearest neighbors of si and sharing the same label than si with regards to the k items to be found. the $P@k$ is then the mean of this ratio along all the items; • influence of the sound markers on the perceived pleasantness: in order to assess the specific contributions of some sound sources, we again estimate the possible impact of the structural features on the perceived pleasantness, but this time, by only taking into account the sound markers for the computation of these features.
Except for the V-test, all the statistical test of significativity are done with a critical threshold ⍺ = 0.05. For the V-test, considering that a great number of classes is tested, a Bonferroni correction is applied. For the p-value, when p > 0.05, the value is mentioned. When 0.01 < p < 0.05, we only mention p < 0.05, otherwise we mention p < 0.01.
3.5 Results I Cheking of the pleasantness of the synthesized scenes First, in order to ensure the coherence of the data, wr check that none of the ni- scenes gets a $\mathcal{A}_{scene}$ higher than one of a i- scene. Four ni- scenes don’t respect that rule: they are removed, together with their corresponding i- scenes. 36 i -scenes and 36 ni -scenes remain for analysis. Secondly, we want to be sure that participants really perceived a difference in terms of pleasantness between i- and ni -scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It appears that the i- scenes were perceived significatively more pleasant (p < 0.01) than the ni- scenes.
Analysis of the structural features First, we focus on the sound level. Figures 4a, 4b, and 4c depict the distributions of levels L, $L(E)$, $L(T)$. There is a significative difference in terms of sound level between i- and ni -scenes (L: p < 0.01) with a standard deviation of -7 $dB$. This difference is either siginificant for events ($L(E)$: p < 0.01), STD: -7 $dB$) than on textures ($L(T)$: p < 0.01), STD: -6 $dB$). As expected, we verify that the sound sources sound level is indeed a pleasantness indicator, the ni -scenes tending to be louder - what also comes out from a large number of related works. We also notice that this difference of sound level is equally true for events or textures.
[FIGURE 4] Figure 4. Distributions of the structural features L (a, d), $L(E)$ (B, e) and $L(T)$ (c, f) with regards to the type of scene (a, b, c) and the perceived pleasantness $\mathcal{A}_{scene}$ resulting from experiment 1.b (d, e, f)
It appears that the biggest influence on the global sound level comes from the events, the gap between L and $L(E)$ being only 1 $dB$ for i- and ni- scenes. This observation is in agreement with the results obtained by Kuwano et al. [37]. During their experiment, the authors ask to their participants to assess a set of soundscapes at a global level and then to do the same judgement at the moments when each of them detect a sound source. The study shows that there is no significative difference between global and averaged instantaneous judgements. In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, i. e. the events.
Finally, we notice that the level alone is not enough to differentiate the different types of environment. In fact, 20% of the i- scenes hast got a sound level higher than the nominal level of the ni- scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
Influence of the structural features on the perceived pleasantness In this section, we analyse the deep relationships that could exist between structural features and perceived pleasantness. Contrary to the previous section, we don’t consider the emotional property of a scene on a binary mode (i vs. ni): here, we consider the mean pleasantness $\mathcal{A}_{scene}$ as a perceptual feature. The aim is to investigate the level of correlation between structural features and $\mathcal{A}_{scene}$. The linear correlation coefficients computed between $\mathcal{A}_{scene}$ and L, $L(E)$, and $L(T)$ are presented in Table II. Relationships between $\mathcal{A}_{scene}$ and the structural features are depicted in Figure 4d, 4e, and 4f.
Concerning L, we can observe a strong negative correlation (r = 0.77, p <0.01) avec $\mathcal{A}_{scene}$, indicating that the higher the sound level is, the more unpleasant the scene is perceived. Nevertheless, Figure 4d suggests that this relationship doesn’t occurr in the same way for i- and ni- scenes. In fact, the correlation between L and $\mathcal{A}_{scene}$ is high for ni- scenes (r = -0.78, p <0.01), but is quasi null for i- scenes. When considering the whole set of scenes, this result can be explained by the fact the i- scenes tends to be softer than the ni- scenes, deluding that the negative correlation observed for the ni- scenes is extended.
We can conclude that L: • allows to differentiate between i- and ni- scenes, • allows a fine characterization of the perceived pleasantness for ni- scenes, • is not a relevant feature for the perceived pleasantness for a priori pleasant soundscapes (i- scene)
The same report can be done concerning $L(E)$ (cf. Figure 4e). For $L(T)$ (cf. Figure 4f), the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i- scenes, r = -0.33, p = 0.05; ni- scenes, r = -0.00, p = 0.99). One can suppose that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer tant ni - scenes). Thus, the sound events keep a relative ability to predict the pleasantness of the ni- scenes, but the level of the textures doesn’t bring a lot of informations whatever the type of environment is.
To sum up, for an unpleasant environment, sound levels - especially those of the events - negatively influence the perceived pleasantness. On the contrary, for a pleasant environment, none of the structural features considered in the study seem to influence the perceived pleasantness. Those first outcomes could show that it exists two modes of perception depending on the nature of the environment (i- or ni-), and each involving independent features. The fact that L is not able to characterize the pleasantness of the i- scenes can lead us to think that all the sound sources don’t equally contribute to the perception of pleasantness, but that only the level of some of them can influence this percept. In order to go deeper in that direction, in the next section, we analyse these scenes on a semantic point of view, i.e. we take an interest in the nature of the sources they are made of.
[TABLE II] Table II. Linear correlation coefficients computed between mean perceived pleasantness (epx. 1.b) and structural features.
Analysis of the semantic features We analyse the composition of the scenes by counting the number of participants who used a given class of sounds to simulate the environment. Results are shown on Figure 5a for the events, and on Figure 5b for the textures. For sake of clarity, we chose a transitional level of abstraction between level 0 and 1, named 0+, to represent the classes. We observe a significant difference between i- and ni- scenes for the choise of classes. The distribution of the classes is very similar to to the one obtained in a related work on ideal urban soundscapes [5]: classes involving human presence and nature are prevailing in the i- scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni- scenes. These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment [23, 6] Nevertheless, we notice some differences with [5]: Guastavino’s results show that sounds of public transportation are specific of ideal urban soundscapes. The authors interpret this by the fact that the perception of pleasantness is, among other things, due to socio-cultural factors. In our representation of the world, sounds of public transportation are positively connoted and tend to be more accepted than sounds of personal vehicles. To a certain extent, our results are in contradiction with this assumption. In fact, Figure 5a shows public transportation classes (bus and train - see Figure 5c) have been used by the participants for i- scenes (28%) and ni - scenes (42%). the results don’t re-evaluate the fact that sounds of public transportation are well accepted: 25% of the participants have used the bus class for the i- scenes, that is comparable to the bicycle class and much higher than all the personal vehicles classes. This being, public transportation classes are also strongly present in the ni- scenes, for instance more than light vehicle or trucks classes. On the basis of our results, the public transportation class can’t be considered as typical of an ideal urban soundscape.
[FIGURE 5] Figure 5. Proportion of synthesized scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level 0+, (b) texture classes at an abstraction level 0, (c) event sub-classes at an abstraction level 1 belonging to traffic and public transportation classes of the abstraction level 0.
This difference can be explained by the nature of the experimental protocol used in the two studies. As in our case, Guastavino asks the participants for describing an environment; but in that case, they work from their memories whereas, in our case, participants do that task using sound samples. The fact that participants in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact. Other works using sounds as stimuli have shown that the bus class can have a negative influence on the assessment of the environment [8].

///

Sound markers We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs regarding the type of environment (i or ni). We’re now trying to to investigate if some of the sound classes are specific to a given environment. For that, we use the V-test (see section 3.4) and consider separately each abstraction level. Results are presented in Table III.
[TABLE III] Table III. Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of V value
Regarding the sound events, 9 markers are identified at all the abstraction levels. As Figure 5 could let it predict, classes related to human presence (male footsteps on concrete, bicycle ring) are i- scenes markers; we also notice the bell class as an i- scenes marker. This latter result may be due to the socio-cultural background of the participants who are mostly european citizens. In fact, according to Schafer, a sound that is identified by a person as being an element of his/her environment, is well accepted. Sound markers of ni- scenes are classes related to construction site (public works), or suggesting a strong traffic (horn, siren). Regarding the textures, 5 markers are identified. For the i- scenes, they are classes related to passive or quiet ambiances (courtyard, park). for the ni- scenes, as for the events, they are classes related to construction site (public work, construction vehicle), together with a class related to traffic (crossroad). Although the whode identified markers are intuitive, none of the event classes related to the noise of the motor vehicles is a marker, except for the texture class crossroad. To simulate an unpleasant traffic, participants chose the classes horn or siren. One can suppose that isolated motor vehicle sounds are understood as taking fully part of the urban environment, and thus are not especially linked to an unpleasant soundscape.
Representation space induced by semantic features In this section, we evaluate the ability of a semantic representation to split the two types of environment. For that, we compute a rank 5-precision (p@5) on the space induced by the semantic features S, and for each abstraction level (see section 3.4). The vectors S are built by using all the classes (ET), only the event classes (E), only the texture classes (T), only the event classes corresponding to sound markers (Em), or only the event classes excluding sound markers (Ew/o,m). We don’t consider the texture classes corresponding to sound markers, these latter being not enough numerous. for the same reasons, we don’t consider the event classes corresponding to sound markers at an abstraction level 0. Results are shown on Figure 6. Concerning ET, the p@5 is 76% for the abstraction level 0, and stay above 86% from the abstraction level 1. These results confirm that it is possible to make a clear distinction between the two types of environment on the unique basis of presence / absence of sound classes. We also notice that the higher the abstraction level is, the more important is the ability to make this distinction. In other words, the more one is precise in the scene description, the more one is able to clearly differentiate between i- and ni- scenes. Considering E and T separately, ir appears: 1) that the p@5 obtained with E is similar to the one obtained with ET ; 2) that the p@5 obtained with T is systematically lower than the one obtained with E, by 10% or 15%. Those results indicate that the semantic information allowing to split the two environments is mainly contained in the events. Moreover, these results are in line with the work done in [20] showing that we analyse in a causal manner (i.e. by identifying the sources) the event scenes which are composed of several sound events (see section 2.1). Finally, it appears that the p@5 obtained with Em is equal – or above – to the ones obtained with E or ET, although a partial information is used in that case to describe the scenes. The dimension of the vectors S for Em is lower than the dimension of vector S for E, itself lower than the dimension of vector S obtained when all the classes are considered (ET). Moreover, in the case when the sound markers are not taken into account for the description (Ew/o,m), the results fall even going under those obtained with only the textures. This confirms that most of the semantic information allowing to differentiate between i- and ni- scenes is included in the markers. To sum up, we deduce the following points from this analysis:
[FIGURE 6] Figure 6. P@5 obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors. The vectors are built by using all the classes (ET), only the event classes (E), only the texture classes (T), only the event classes corresponding to sound markers (Em), or only the event classes excluding sound markers (Ew/o,m).
1. unlike what we outlined with the structural features, a semantic description of the scenes constitution in terms of presence / absence of sound sources, allows to differentiate the two types of environments (i or ni); 2. the semantic information is mainly contained in the event sound classes; 3. only a part of the event classes, i.e. the sound markers, are useful to differentiate the i- and ni- scenes.
Since we have extracted the typical classes of the i- and ni- scenes and verified that the distinction between them was dependant to the presence of these classes, we have now to investigate wether a structural description of the scenes only based on sound markers, allows to characterize the perceived pleasantness, better than a global structural description.
Influence of the sound markers on the perceived pleasantness We evaluate the correlations between $\mathcal{A}_{scene}$ and the structural features. In this section, the structural features are computed by taking into account of the sound markers previoulsy identified. We define Xm, the X feature computed by taking into account only the sound markers. On the opposite, we define Xb (b for bruit / noise), the X feature computed by taking into account all the sound classes, except the sound markers. When the feature characterize an i- scene (idem for a ni- scene), we consider for the computation only the markers identified for the i- scenes (or the ni- scenes); we named them, i- markers (or ni- markers). Results are shown in Table IV. Concerning the sound levels (see Figure 7a, 7d), the same trends are observed between Lm, $L(E)$m and $L(T)$m, on one hand, and L, $L(E)$ and $L(T)$, on the other hand. Considering all the classes or only the markers, it appears that: 1. it exists a significative difference between levels of i- and ni- scenes (Lm, $L(E)$m and $L(T)$m: p < 0.01); 2. the sound level of scenes is mainly related to the sound events, compared to the textures; 3. the sound level of events has an influence on the perception of pleasantness for ni- scenes, but not for i- scenes; 4. the sound level of textures doesn’t play any role in the perception of the pleasantness.
In conclusion, the level of ni- markers has a negative influence on the pleasantness for the ni- scenes. On the other hand, the level of i- markers doesn’t influence the perceived pleasantness for the i- scenes.
Considering the non markers classes (see Figure 7b, 7e), we can observe on the i- scenes results, a weak negative correlation for Lb (r = -0.52, p < 0.01) and $L(E)$b (r = -0.51, p < 0.01). It is the first time that an objective feature allows us to define the pleasantness of the pleasant environments. This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.
Besides, whereas $L(T)$ didn’t show any significative correlation for ni- scenes, a strong negative correlation is observed for $L(T)$b (r = -0.73, p < 0.01). This indicates that the level of non-markers texture classes doesn’t influence the perceived pleasantness in the same way for i- and ni- scenes. Sound level seems to have a negative effect for the ni- scenes, and no significant effect for the i- scenes.
To end, we consider a last group of features, namely Lm - Lb, $L(E)$m - $L(E)$b, $L(T)$m - $L(T)$b (see Figures 7c and 7f). These features describe the difference between the markers level and those of the other sound classes. They express the saliency of the markers with regards to the sound mixture.
For the i- scenes, a moderate positive correlation is observed for Lm - Lb (r = 0.67, p < 0.01) and $L(E)$m - $L(E)$b (r = 0.66, p < 0.01). for the ni- scenes no correlation is observed. Thus, for the i- scenes, it is not the absolute markers level that is important, but their relative level with regards to the other sounds composing the scene. Then, we can induce a double perceptual mechanism for the ideal environments:
[TABLE IV] Table IV. Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ (Exp. 1.b) and structural features related to the presence of sound markers.
[FIGURE 7] Figure 7. Distribution of the relative sound levels structural features related to the presence of markers Lm (A, d), $L(E)$m (b, e) and $L(T)$m (c, f) with regards to the type of scenes (a, b, c) and the perceived pleasantness measured in Exp. 1.b (d, e, f).
• the higher the absolute level of sounds not being i- markers, the weaker the pleasantness is, • the higher the relative level of i- markers is, with regards to the other sounds, the higher the pleasantness is.
For the ni -scenes, the fact that we observe significative correlations for Lm and $L(E)$m and no correlation for Lm - Lb and $L(E)$m - $L(E)$b shows that this the absolute level that is important.
3.6 Discussions
From this analysis, we can outline the following points: • differentiate the i- and ni- scenes: the semantic features, together with the global structural features (L, $L(E)$, and $L(T)$) allow to differentiate between i- and ni- scenes. The semantic description seems to be more powerful; • events or textures: whatever semantic or structural features, events are mainly useful to differentiate the two types of events; textures bring a limited amount of informations; • pleasantness prediction: if we consider the correlations between structural features and pleasantness, it seems that the way to perceive the quality of a given environment depends on its nature (i or ni). It is not conceivable to consider a same set of features to predict the pleasantness of i- and ni- scenes in the same time: - for the ni- scenes, global levels (L and $L(E)$), or level of sound markers (Lm and $L(E)$m) have a negative influence on pleasantness. We can notice here taht taking into account the contribution of each the different sources of the scene doesn’t improve the prediction performance, wit regards to a global analysis of the environment. - for the i- scenes, on the contrary, the pleasantness prediction needs to study separately the sound markers characteristics and those of the other sounds. Thus, the markers level, relative to the noise ($L(E)$m - $L(E)$b and Lm - Lb) is positively correlated to the pleasantness, whereas the noise level (Lb and $L(E)$b) is negatively correlated. The fact that the pleasantness of the i- scenes is not correlated to global physical features, contrary to pleasantness of ni- scenes has been recently studied [11].
The existence of two perceptual modes involving different types of features, and depending on the nature (in our case hedonic) of the stimuli, is a phenomenon similar to the one observed for the perception of textures (see section 2.1). the brain adapts its way to encode the information (statistic summary for textures, fine description for events) following a previous decision making with regards to the nature of the stimuli (i.e. is it an event or a texture ?). Similarly, features that are active in the pleasantness judgment also depends on a preliminary judgment of the global hedonic nature of the environment (ideal or non ideal).

### 2017-07-04 lafaySimulatedScenePerception.tex
mid way to inlcuding NM translation

@@
 \newcommand{\jlc}[1]{\textcolor{green}{#1}}
-
 \newcommand{\ie}{\emph{i.\,e.}}
@@
 Further details about the simulation interface can be found in \cite{rossignol2015simscene}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%  Experience 1     %%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Expérience 1}
+\label{sec:xp1}
+
+\subsection{Objectif}
+
+% L'objectif est d'étudier les influences spécifiques des différentes sources sonores qui composent les environnements urbains sur la perception de l'agrément, en utilisant la simulation.
+Pour ce faire, nous planifions nos deux premières expériences (\cf~Figure~\ref{fig:xp1_2}):
+%
+% \begin{itemize}
+% \item \emph{expérience 1.a de simulation}: au cours de cette expérience, les sujets doivent simuler des environnements sonores urbains, en utilisant l'outil de simulation \emph{Simscene} (\cf~Section~\ref{sec:simscene}).
+Chacun compose deux environnements sonores, le premier idéal/agréable et le deuxième non-idéal/désagréable.
+%
+% \item \emph{expérience 1.b d'évaluation}: à l'issue de la simulation, nous n'avons, de fait, qu'une connaissance binaire des propriétés affectives des scènes simulées: idéale (i) et non-idéale (ni).
+Cette seconde étape a pour but d'affiner notre connaissance sur l'agrément.
+Pour ce faire, nous demandons à un deuxième groupe de sujets d'évaluer, à partir d'une échelle sémantique, l'agrément de chacune des scènes simulées.
+L'expérience d'évaluation sert deux buts:
+%
+% \begin{enumerate}
+% \item évaluer l'influence respective des différentes sources sur l'agrément, pour chaque type d'environnement (i ou ni);
+% \item détecter la présence de cas extrêmes ou ambigus (\emph{outlier}) dans les scènes simulées.
+Pour le reste de notre étude, les qualités hédoniques imposées (i et ni) servent de référence, de vérité terrain.
+Il nous faut donc garantir qu'il n'y ait pas d’ambiguïté entre les cas extrêmes des i- et ni-scènes, \ie~que la note d'agrément la plus basse des i-scènes reste supérieure à la note la plus haute des ni-scènes.
+% \end{enumerate}
+%
+% \end{itemize}
+%
+% Notre analyse s'appuie sur les données produites par les deux expériences.
+
+Experiment 1 aims at using the simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness.
+For that, two first experiments are planned (\cf~Figure~\ref{fig:xp1_2}):
+\begin{itemize}
+\item \emph{experiment 1.a (simulation)}: during this experiment, participants have to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simscene}.
+Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
+\item \emph{experiment 2.a (evaluation)}: actually, after the simulation phase, we only have a binary information on the emotional properties of the respective scenes: ideal (i) and non ideal (ni).
+This second experimental step aims at investigating our knowledge on pleasantness.
+For that, a second goup of participants are asked to evaluate the pleasantness of each simulated scene, on a semantic scale.
+This evaluation protocol has two goals:
+\begin{enumerate}
+\item to evaluate the respective influence on the pleasantness of the various sources composing the scenes (i or ni)
+\item to detect the presence of \emph{outliers} or ambiguous samples.
+For the remainder of the study, the defined hedonic properties (i and ni) are used as ground truth.
+Then, we must have to be sure that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of pleasantness rating.
+\end{enumerate}
+\end{itemize}
+
+Our analysis is based on the data collected by these two experiments (1.a, 1.b)
+
+\begin{figure}[t]
+        \myfloatalign
+        \includegraphics[width=\linewidth]{gfx/ch_5/5}
+        \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b)}\label{fig:xp1_2}
+\end{figure}
+
+\subsection{Planning of Experiment 1.a}
+\label{sec:xp1a_plan}
+
+\subsubsection*{Procedure}
+
+% Les sujets doivent simuler deux environnements sonores urbains, chacune des scènes devant durer 1 minute. Ils doivent se conformer aux consignes suivantes:
+%
+% \begin{itemize}
+% \item première simulation: simuler un paysage sonore \textbf{urbain plausible} qui, selon vous, est idéal (où vous aimeriez vivre);
+% \item deuxième simulation: simuler un paysage sonore \textbf{urbain plausible} qui, selon vous, est non-idéal (où vous n'aimeriez pas vivre).
+% \end{itemize}
+%
+% Tous commencent par simuler l'environnement idéal. Ils ne prennent connaissance de la deuxième consigne qu'à la fin de la première simulation.
+%
+% Les sujets sont totalement libres dans le choix des sons, et des paramètres. Ils doivent cependant se soumettre à deux contraintes:
+%
+% \begin{itemize}
+% \item le sujet doit prendre le point de vue d’un auditeur fixe;
+% \item le paysage sonore doit être réaliste, au sens de physiquement plausible.
+Le sujet peut très bien réunir 10 chiens dans un paysage sonore.
+Il ne peut en introduire 1 qui aboierait toutes les 10 millisecondes.
+% \end{itemize}
+%
+% Ces contraintes font partie de la consigne. Aucun contrôle n'est fait \emph{a priori} dans l'interface de simulation.
+
+The participants have to simulate two urban sound environments of 1 minute each. They have to follow the instructions below:
+\begin{itemize}
+\item  1st simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
+\item 2nd simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
+\end{itemize}
+
+All the participants start by working on the ideal environment; they read the second set of instruction at the end of the first experiment.
+Participants are completely free of their choice concerning sounds and synthesis parameters.
+They must nevertheless respect the two following constraints :
+\begin{itemize}
+\item the listening point of view is the one of a fixed listener;
+\item  the soundscape must be realistic, i.e. physically plausible.
+\end{itemize}
+For instance, the participant can insert ten dogs in the soundscape but he/she can’t insert one dog barking every 10 milliseconds.
+These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation user interface.
+
+% Chaque processus de simulation comprend deux parties :
+%
+% \begin{enumerate}
+% \item la réalisation de la simulation: cette étape se décompose en trois temps:
+% \begin{itemize}
+% \item sélectionner les classes de sons
+% \item nommer les classes de sons sélectionnées
+% \item paramétrer les pistes (\cf~Section~\ref{sec:simscene_parametre}) relatives aux classes de sons sélectionnées
+% \end{itemize}
+% \item la production d'un commentaire libre du paysage sonore simulé
+% \end{enumerate}
+%
+% En complément, et une fois les deux scènes sonores réalisées, les sujets sont invités à:
+%
+% \begin{itemize}
+% \item indiquer les sources sonores qu'ils voulaient mettre, mais qu'ils n'ont pas trouvées;
+% \item commenter l’ergonomie du logiciel de simulation;
+% \item commenter l’ergonomie de l'interface de sélection.
+% \end{itemize}
+%
+% Avant de commencer la première simulation, un tutoriel de 20 minutes est proposé aux sujets, afin qu'ils se familiarisent avec le logiciel de simulation, et la banque de données.
+L'expérience est prévue pour durer 2h30.
+
+Each simulation process contains two parts :
+\begin{enumerate}
+\item composition of the simulation: this step is split in three steps
+\begin{itemize}
+  \item  select the class of sounds
+	\item  give a name to the selected class of sounds
+	\item  set the parameters of the tracks related to the selected class of sounds, see Section~\ref{sec:simscene_parametre}.
+\end{itemize}
+\item production of a free verbalization of the composed soundscape
+\end{enumerate}
+
+In addition, once the two sound scenes have been realized, participants are invited to
+\begin{itemize}
+\item  point out the sound sources that were missing for the composition
+\item  give a comment about the ergonomics of the simulation environment
+\item   give a comment about the ergonomics of the selection tool
+\end{itemize}
+
+Before starting the first simulation, a 20-minute tutorial is given to the participants in order to familiarize them with the simulation environment and the sound database.
+The experiment is planned to last about two hours and a half, including breaks that could be taken by the participants.
+
+\subsubsection*{Apparatus}
+
+% Tous les sujets passent l'expérience sur des machines identiques.
+L'audio est diffusé en diotique, par le biais de casques audio.
+Pendant le tutoriel, les sujets doivent ajuster le niveau sonore à un volume confortable.
+Ils ne peuvent le modifier par la suite.
+%
+% Tous les sujets réalisent l'expérience simultanément.
+Ils sont répartis de manière égale dans trois pièces identiques, toutes possédant un environnement calme.
+Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
+%
+% Trois expérimentateurs, un dans chaque pièce, sont présents durant la totalité de l'expérience, afin de contrôler le bon déroulement de cette dernière, et de répondre aux éventuelles questions des sujets.
+
+All the participants do the experiment on similar computers.
+The audio files are played in diotic conditions by headphones.
+During the tutorial, participants are asked to adjust the sound level in order to be comfortable; they can’t modify it in the following of the experiment.
+All the participants do the experiment simultaneously. They are equally split in three identical quiet rooms. They can’t talk together during the experiment.
+Three experimenters (one in each room) are present during the whole duration of the experiment in order to check that everything goes well and possibly to answer questions.
+
+\subsubsection*{Participants}
+
+% 44 étudiants (14 femmes) de L’École Centrale de Nantes ont participé à l'expérience.
+Ils ont tous sensiblement le même âge (moyenne: 21.6, écart-type: 2).
+Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
+%
+% Sur les 44 sujets, 40 réalisent l'expérience avec succès, produisant au final 80 scènes sonores simulées, dont 40 scènes idéales, et 40 scènes non idéales.
+4 sont éliminés pour non respect ou incompréhension des consignes, d'une part, et dépassement du temps, d'autre part.
+
+44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in Nantes downtown since 2 years or more.
+Among the 44 participants, 40 succeeded in doing the experiment, producing in the end, 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
+4 participants were excluded from the process for not having respect or understand the instructions, or for exceeding the maximum duration of the experiment.
+
+\subsection{Planning of experiment 1.b}
+\label{sec:xp1b_plan}
+
+\subsubsection*{Procedure}
+
+% En raison de contraintes temporelles, les sujets n'évaluent que des séquences de 30 secondes des scènes simulées, chacune de ces séquences commençant à la seconde 15, et finissant à le seconde 45 de la scène d'une durée initiale de 1 minute.
+%
+% L'évaluation s'effectue sur une échelle sémantique bipolaire de 7 points, allant de -3 (non-idéale/très désagréable) à +3 (idéale/très agréable).
+Avant de noter une scène, les sujets doivent obligatoirement écouter les 20 premières secondes de cette dernière.
+Après la notation, ils sont libres de passer à la scène suivante.
+%
+% Pour chaque sujet, les scènes sont présentées dans un ordre aléatoire.
+Les 10 premières scènes permettent au sujet de calibrer ses notes.
+Elles sont obligatoirement composées de 5 scènes idéales, et de 5 scènes non-idéales.
+Ces 10 premières scènes sont rejouées à la fin de l'expérience, et seules les notes données à la deuxième occurrence sont prises en compte.
+%
+% L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.
+
+Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes composed in Exp. 1.a (from sec. 15 to sec. 45).
+The assessment is done in a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
+Before evaluating a scene, participants must listen to the first 2à seconds of the scene.
+After the evaluation, they are free to switch to the next scene.
+For each participant, sound scenes are played in a random order.
+The first 10 scenes allow the participants to calibrate their marks; they imperatively gather 5 ideal scenes and 5 non ideal scenes.
+These first 10 scenes are played back again at the end of the experiment and the corresponding evaluations are the ones that are only taken into account in the results.
+The experiment is planned to last 30 minutes. The participants don’t know anything beforehand about the nature of the sound scenes.
+
+\subsubsection*{Apparatus}
+
+% Tous les sujets passent l'expérience sur des machines identiques.
+L'audio est diffusé en monophonie, par le biais de casques audio semi-ouverts \emph{Beyer-Dynamic DT 990 Pro}.
+Toutes les scènes sonores ont été re-simulées sur la base des partitions obtenues lors de l'expérience de simulation.
+Le niveau sonore de sortie est identique pour tous les sujets.
+%
+% Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
+%
+% Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de celle-ci, et de répondre aux éventuelles questions des sujets.
+
+All the participants do the experiment on similar computers.
+The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}.
+All the sound scenes have been re-composed on the basis of the scores obtained in the simulation experiment (Exp.
+1.a).
+The output sound level is the same for all the participants.
+All the participants do the experiment simultaneously. They can’t talk together during the experiment.
+An experimenter is present during the whole duration of the experiment in order to check that everything goes well and possibly to answer questions.
+
+\subsubsection*{Participants}
+
+% 10 étudiants (2 femmes) de L’École Centrale de Nantes ont participé à l'expérience.
+Aucun d'entre eux n'a réalisé l'expérience de simulation.
+Tous les sujets ont sensiblement le même âge (moyenne: 23.1, écart-type: 1.8).
+Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
+%
+% Tous les sujets ont réalisé l'expérience avec succès.
+
+10 students (2 females - mean age 23.1, SD = 1.8) from Ecole Centrale de Nantes took part to the experiment.
+They are all living in Nantes since 2 years or more.
+None of them took part to the previous simulation experiment (Exp.
+1.a).
+All the participants succeeded in doing the experiment.
+
+\subsection{Data and analysis methods}
+\label{sec:xp1_dataAna}
+
+% A chaque scène est associé un groupe de descripteurs.
+C'est sur ces descripteurs que nous pratiquons l'analyse.
+Un résumé de ces derniers, ainsi que des acronymes les désignant, est présenté dans le Tableau~\ref{tab:acronyme}.
+Afin de rester cohérent avec l’expérience 1.b d'évaluation, les descripteurs ne sont pas calculés sur la durée totale des scènes, mais sur la version réduite de 30 secondes (\cf~Section~\ref{sec:xp1b_plan}).
+%
+% Pour chaque scène sonore, trois types de descripteurs sont considérés:
+%
+% \begin{itemize}
+% \item \emph{perceptif}: il s'agit de l'agrément perçu de la scène simulée, évalué sur une échelle sémantique de 7 points.
+Nous notons $\mathcal{A}_{scene}$ l'agrément moyen d'une scène, obtenu en moyennant les notes de tous les sujets.
+De même, nous notons $\mathcal{A}_{sujet}$ l'agrément par sujet, en moyennant l'ensemble des notes de chacun.
+Compte tenu du faible nombre de sujets, nous faisons le choix, dans cette étude, de ne pas normaliser les notes d'agrément;
+% \item \emph{sémantique}: il s'agit d'un vecteur booléen noté $S=(x_1,x_2,\ldots,x_n)$, indiquant les classes de sons présentes dans la scène.
+Chaque point $x$ de ce vecteur correspond à une classe de sons particulière: $x=1$ si la classe est présente dans la scène, et $x=0$ autrement.
+La dimension $n$ des vecteurs dépend du niveau d'abstraction considéré, \eg~pour le niveau d'abstraction 1, qui comprend $44$ classes de sons, cette dimension sera de $n=44$.
+% \item \emph{structurel}: il s'agit du \emph{niveau Sonore}.
+Pour représenter le niveau sonore, nous nous inspirons de la mesure $L_{Aeq}$.
+Dans notre cas, il s'agit d'un scalaire calculé sur le signal en volts, et non en pression, et donné en décibels, en prenant un référentiel de 1 Volt.
+Le niveau est obtenu en calculant, toutes les secondes, la moyenne quadratique du signal, et en moyennant sur la durée de la scène.
+Un filtrage de type A est opéré avant le calcul des moyennes quadratiques.
+Nous notons $L$, $L(E)$ et $L(T)$ les niveaux calculés en tenant compte respectivement de l'ensemble des samples, des samples d'événements et de textures.
+% \end{itemize}
+
+A set of features is attached to each sound scene.
+The analysis is done on the basis of these features.
+A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}.
+In order to stay coherent with the evaluation experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version, see Section \ref{sec:xp1b_plan}.
+For each sound scene, three types of features are considered :
+\begin{itemize}
+\item \emph{perceptual}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale.
+$\mathcal{A}_{scene}$ is the mean pleasantness for one scene, as the average of the marks of all the participants.
+In the same idea, $\mathcal{A}_{subject}$ is the mean pleasantness for each participant, as the average of all the marks of each participant.
+Given the low number of participants in Exp.
+1.b, we decide not to normalize the values of pleasantness marks.
+\item \emph{semantic}: boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene.
+Each point $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise.
+The vector dimension ($n$) depends on the level of abstraction that is considered.
+For instance, for the abstraction level 1, including 44 classes of sounds, $n = 44$.
+\item \emph{structural}: related to \emph{sound level}.
+To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference.
+In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in deciBels, with a reference of 1 Volt.
+The level is obtained by computing every second the quadratic mean of the signal and by averaging over the total duration of the scene.
+An A-filtering module processes the data before the quadratic means are computed.
+We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples and only the set of texture samples.
+\end{itemize}
+
+In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following tests of significativity:
+
+\begin{itemize}
+\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with regards to the two types of scene.
+the significativity is evaluated by a Student test on two independant populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$ ;
+\item \emph{analysis of the structural features}: in order to evaluate wether the emotional difference forced to the i- and ni- scenes has a significative influence \ie, if significative differences exist between structural features and perceived pleasantness, we measure this significativity with a Student test on two populations;
+\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, we look at linear correlations between these features and $\mathcal{A}_{scene}$.
+For that, we use the Pearson correlation coefficient, on the basis of the standard methodology used in multidimensional studies;
+\item \emph{analysis of the semantic features}: in order to determine if the forced emotional difference has an influence on the scenes composition in terms of sound sources, or if specific souns classes are more frequently used to simulate a given environment (i or ni), we apply a V-test.
+This test is done at every level of abstraction, and separately, for texture and event classes.
+For each class $j$ , and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:
+
+\begin{equation*}
+V_{jk}=\dfrac{c_{jk}-c_k\frac{c_j}{c}}{\sqrt{c_k\frac{c-c_k}{c-1}\frac{c_j}{c}(1-\frac{c_j}{c})}}
+\end{equation*}
+
+where $c$ is the number of available classes, $c_k$ the number of classes for a given type of environment $k$, $c_j$ the number of  $j$ classes, and $c_{jk}$ the number of $j$ classes used for an environment $k$.
+The V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significatively different from the ratio $\frac{c_{jk}}{c_k}$.
+For a given environment $k$, and class $j$, if the hypothesis is rejected, the class $j$ is said to be typical with respect to environment $k$.
+The typical classes are called \textbf{sound markers};
+
+\item \emph{Study of the representation spaces induced by the semantic features}: in order to determine if a representation only based on the presence / absence of sound sources allows to split the two types of environment, we investigate the space induced by the semantic features $S$.
+$S$ is a boolean vector and the distance between two scenes is computed with the Hamming distance.
+Considering two n-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, where $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ is of cardinality $n$, with $x=\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between two vectors, as follows:
+
+\begin{equation*}
+d_{ham}(S_1,S_2)=\dfrac{1}{n}\sum_{i=1}^{n} (x_{1,i} \bigoplus x_{2,i})
+\end{equation*}
+
+where $\bigoplus$ is the \emph{exclusive-or} The more the two scenes are similar, the closer they are.
+The use of Hamming distance allows to take into account equally the present and absent classes.
+For measuring the intrinsic ability of the space to split the i- and ni- scene, we use a ranking metric named $k$- precision ($P@k$) .
+The $P@k$ computes the precision obtained after $k$ items have been found.
+Formally, for each $s_i$ scene, we compute the ratio between the number of $s_j$ scenes taken among the k nearest neighbors of $s_i$ and sharing the same label than $s_i$ with regards to the $k$ items to be found.
+the $P@k$ is then the mean of this ratio along all the items;
+
+\item \emph{influence of the sound markers on the perceived pleasantness}: in order to assess the specific contributions of some sound sources, we again estimate the possible impact of the structural features on the perceived pleasantness, but this time, by only taking into account the sound markers for the computation of these features.
+\end{itemize}
+
+Except for the V-test, all the statistical test of significativity are done with a critical threshold of $\alpha=0.05$.
+For the V-test, considering that a great number of classes is tested, a Bonferroni correction is applied.
+For the $p$-value, if $p\geq0.05$, the value is reported.
+If $0.01\leq p<0.05$, we only report $p<0.05$, otherwise we report $p<0.01$.
+
+\begin{table}[t]
+\centering
+\begin{tabular}{c c}
+Termes                         & Acronyms              \\
+\hline
+Niveau                         & $L$                    \\
+Niveau (événements)            & $L(E)$                 \\
+Niveau (textures)              & $L(T)$                 \\
+Agrément moyen (par scène)     & $\mathcal{A}_{scene}$  \\
+Agrément moyen (par sujet)     & $\mathcal{A}_{sujet}$  \\
+                               &                        \\
+idéale/agréable                & i                      \\
+non-idéale/désagréable         & ni                     \\
+scène idéale/agréable          & i-scene                \\
+scène non-idéale/désagréable   & ni-scene               \\
+\hline
+\end{tabular}
+\vspace{0.5mm}
+\caption{Acronyme des variables utilisées dans le cadre des expériences sensorielles.}
+\label{tab:acronyme}
+\end{table}
+
+\subsection{Results}
+
+\subsubsection*{Cheking of the pleasantness of the synthesized scenes}
+
+First, in order to ensure the coherence of the data, wr check that none of the ni- scenes gets a $\mathcal{A}_{scene}$ higher than one of a i- scene.
+Four ni- scenes don’t respect that rule: they are removed, together with their corresponding i- scenes.
+36 i -scenes and 36 ni -scenes remain for analysis.
+Secondly, we want to be sure that participants really perceived a difference in terms of pleasantness between i- and ni -scenes.
+For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment.
+It appears that the i- scenes were perceived significatively more pleasant ($p < 0.01$) than the ni- scenes.
+
+\subsubsection*{Analysis of the structural features}
+
+First, we focus on the sound level.
+Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} depict the levels $L$, $L(E)$ and $L(T)$.
+There is a significative difference in terms of sound level between i- and ni -scenes ($L$: $p<0.01$), with a standard deviation of -7 $dB$.
+This difference is siginificant for events ($L(E)$: $p<0.01$, STD: -7 $dB$) and textures ($L(T)$: $p<0.01$, STD: -6 $dB$).
+
+As expected, we verify that the sound sources sound level is indeed a pleasantness indicator, the ni -scenes tending to be louder - what also comes out from a large number of related works.
+We also notice that this difference of sound level is equally true for events or textures.
+
+It appears that the biggest influence on the global sound level comes from the events, the gap between $L$ and $L(E)$ being only 1 $dB$ for i- and ni- scenes.
+This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}.
+During their experiment, the authors ask to their participants to assess a set of soundscapes at a global level and then to do the same judgement at the moments when each of them detect a sound source.
+The study shows that there is no significative difference between global and averaged instantaneous judgements.
+In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, \ie the events.
+
+The level alone is not enough to differentiate the different types of environment.
+In fact, $20\%$ of the i- scenes hast got a sound level higher than the nominal level of the ni- scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
+
+\begin{figure}[t]
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_1}\label{fig:soundlevela}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_3}\label{fig:soundlevelb}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_5}\label{fig:soundlevelc}}\par
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_2}\label{fig:soundleveld}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_4}\label{fig:soundlevele}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_6}\label{fig:soundlevelf}}
+       \caption{Distributions of the structural features $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
+\end{figure}
+
+\subsubsection*{Influence of the structural features on the perceived pleasantness}
+
+In this section, we analyse the deep relationships that could exist between structural features and perceived pleasantness.
+Contrary to the previous section, we don’t consider the emotional property of a scene on a binary mode (i vs.
+ni): here, we consider the mean pleasantness $\mathcal{A}_{scene}$ as a perceptual feature.
+The aim is to investigate the level of correlation between structural features and $\mathcal{A}_{scene}$.
+The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown on Table~\ref{tab:corrStructA}.
+Relationships between $\mathcal{A}_{scene}$ and the structural features are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and ~\ref{fig:soundlevelf}.
+
+COncerning $L$, we can observe a strong negative correlation ($r=-0.77$, $p<0.01$) with $\mathcal{A}_{scene}$, indicating that the higher the sound level is, the more unpleasant the scene is perceived.
+Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship doesn’t occurr in the same way for i- and ni- scenes.
+In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni- scenes ($r=-0.78$, $p<0.01$), but is quasi null for i-scenes.
+
+When considering the whole set of scenes, this result can be explained by the fact the i- scenes tends to be softer than the ni- scenes, deluding that the negative correlation observed for the ni- scenes is extended.
+
+We thus conclude that $L$:
+
+\begin{itemize}
+\item allows to differentiate between i- and ni- scenes,
+\item allows a fine characterization of the perceived pleasantness for ni- scenes,
+\item is not a relevant feature for the perceived pleasantness for a priori pleasant soundscapes (i- scene)
+\end{itemize}
+
+The same conclusions can be drawn for $L(E)$ (\cf~Figure~\ref{fig:soundlevele}).
+For $L(T)$ (\cf~Figure~\ref{fig:soundlevelf}), the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$).
+One can suppose that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer than ni - scenes).
+Thus, the sound events keep a relative ability to predict the pleasantness of the ni- scenes, but the level of the textures doesn’t bring a lot of information whatever the type of environment is.
+
+To sum up, for an unpleasant environment, sound levels - especially those of the events - negatively influence the perceived pleasantness.
+On the contrary, for a pleasant environment, none of the structural features considered in the study seem to influence the perceived pleasantness.
+
+Those first outcomes tend to show that two modes of perception exists depending on the nature of the environment (i- or ni-), and each involve independent features.
+The fact that $L$ is not able to characterize the pleasantness of the i- scenes can lead us to think that all the sound sources don’t equally contribute to the perception of pleasantness, but that only the level of some of them can influence this percept.
+In order to go deeper in that direction, in the next section, we analyse these scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.
+
+\begin{table}[t]
+\centering
+\begin{tabular}{l c c c}
+               & i-scenes                   & ni-scenes    \\
+\hline
+$L$            & -0.32 ($p=0.06$)           & \textbf{-0.78} ($p<0.01$)\\
+$L(E)$         & -0.20 ($p=0.24$)           & \textbf{-0.75} ($p<0.01$)\\
+$L(T)$         & -0.33 ($p=0.05$)           &  -0.00 ($p=0.99$) \\
+\hline
+\end{tabular}
+\vspace{0.5mm}
+\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and structural features.}
+\label{tab:corrStructA}
+\end{table}
+
+\subsubsection*{Analysis of the semantic features}
+
+We analyse the composition of the scenes by counting the number of participants who used a given class of sounds to simulate the environment.
+Results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
+For sake of clarity, we chose a transitional level of abstraction between level 0 and 1, named $0+$, to depict classes, see ~Figure~\ref{fig:taxonomie}.
+
+We observe a significant difference between i- and ni- scenes for the choise of classes.
+The distribution of the classes is very similar to to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~lasses involving human presence and nature are prevailing in the i- scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni- scenes.
+These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.
+
+Nevertheless, we notice some differences with \cite{guastavino2006ideal}: Guastavino’s results show that sounds of \emph{public transportation} are specific of ideal urban soundscapes.
+The authors interpret this by the fact that the perception of pleasantness is, among other things, due to socio-cultural factors.
+In our representation of the world, sounds of public transportation are positively connoted and tend to be more accepted than sounds of personal vehicles.
+
+To a certain extent, our results are in contradiction with this assumption.
+In fact,  Figure~\ref{fig:soundsourcea} show \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the participants for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes.
+the results don’t re-evaluate the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} pour les i-scenes, a level that is comparable to the of the \emph{bike} class, and much higher than all the personal vehicles classes.
+Though, \emph{public transport} classes are also strongly present in the ni-scenes, for instance more than \emph{light vehicle} or \emph{trucks} classes.
+On the basis of our results, the public transportation class can’t be considered as typical of an ideal urban soundscape.
+
+This difference can be explained by the nature of the experimental protocol used in the two studies.
+As in our case, Guastavino asks the participants for describing an environment; but in that case, they work from their memories whereas, in our case, participants do that task using sound samples.
+The fact that participants in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact.
+Other works using sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment \cite{lavandier2006contribution}.
+
+\begin{figure}[t]
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=1\linewidth]{gfx/ch_5/xp1_class_1_cam}\label{fig:soundsourcea}} \par
+        \subfloat[]
+        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_2_cam}\label{fig:soundsourceb}}
+        \subfloat[]
+        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_3_cam}\label{fig:soundsourcec}}
+       \caption{Proportion of synthesized scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.}\label{fig:soundsource}
+\end{figure}
+
+\subsubsection*{Marqueurs sonores}
+
+Nous avons mis en évidence que, qualitativement, la composition des sources sonores des scènes diffère selon les types d'environnement (i ou ni).
+Nous essayons de voir maintenant si, parmi ces classes, certaines sont typiques d'un environnement en particulier.
+Pour ce faire, nous utilisons le V-test (\cf~Section~\ref{sec:xp1_dataAna}), en considérant séparément chaque niveau d'abstraction.
+Les résultats sont présentés dans le tableau~\ref{tab:markers}.
+
+Concernant les événements sonores, 9 marqueurs sont identifiés sur l'ensemble des niveaux d'abstraction.
+Comme la figure~\ref{fig:soundsource} le laissait présager, les classes relatives à la présence humaine (\emph{pas homme béton}, \emph{sonnette vélo}), et à la nature (\emph{animaux, oiseaux}, \emph{chants d'oiseaux}) sont des marqueurs de i-scènes.
+Nous notons également la présence de la classe \emph{cloche} dans les marqueurs d'un environnement idéal.
+Ce fait est possiblement dû au \emph{background} socio-culturel des sujets, dans leur grande majorité, des citoyens européens.
+En effet, selon Schafer, un son reconnu par un individu comme faisant partie intégrante de son environnement est bien accepté.
+Les marqueurs de ni-scènes sont des classes faisant référence à des sons de travaux (\emph{travaux}), ou suggérant un trafic dense (\emph{klaxon}, \emph{sirène}).
+
+Concernant les textures sonores, 5 marqueurs sont identifiés.
+Pour les i-scènes, il s'agit de classes faisant référence à des ambiances amorphes, calmes, (\emph{cour-intérieur} et \emph{parc}).
+Pour les ni-scènes, il s'agit, comme pour les événements, de classes faisant référence à des bruits de travaux (\emph{travaux} et \emph{véhicule de travaux}), ainsi que d'une classe faisant référence au trafic (\emph{carrefour}).
+
+Bien que l'ensemble des marqueurs identifiés soient intuitifs, aucune des classes d'événements faisant directement référence aux bruits de véhicules motorisés n'est un marqueur, exception faite de la classe de textures \emph{carrefour}.
+Pour représenter un trafic désagréable, les sujets ont porté leurs choix sur les classes \emph{klaxon} et \emph{sirène}.
+On peut supposer que les sons isolés de véhicules sont compris comme faisant partie intégrante de l'environnement urbain, et ne sont donc pas particulièrement associés à un environnement désagréable.
+
+\begin{table}[t]
+ \setlength{\tabcolsep}{0.2pt}
+ \centering
+  {\renewcommand{\arraystretch}{0.9}
+\begin{tabular}{c c c c}
+Niveau        & \multicolumn{2}{c}{Marqueurs sonores événements} \\
+d'abstraction & i-scènes & ni-scènes \\
+\hline
+0  &                          & travaux (3.78)  \\
+\hline
+  & cloche  (4.5)             & klaxon  (3.9) \\
+1 & sonnette vélo  (4.3)      & sirène (3.9)\\
+  & animal (4.2)              &       \\
+   \hline
+  & oiseau        (4.8)       & klaxon  (4.0)\\
+2 & cloche  (4.4)             & sirène (4.0)\\
+  & sonnette vélo    (4.2)             &       \\
+   \hline
+  & chant oiseau (4.8)        & klaxon  (4.1)\\
+3 & cloche   (4.3)            & sirène (4.0)\\
+  & sonnette vélo     (4.2)   &       \\
+  & pas chaussure  (3.6)      &  \\
+  &                           & \\
+  & \multicolumn{2}{c}{Marqueurs sonores textures}      \\
+  & i-scènes & ni-scènes \\
+\hline
+0 &     cour int./parc (4.1) &  travaux (3.9)  \\
+\hline
+1 &     parc (3.65)          &  carrefour (3.6)  \\
+  &                          &  travaux véhicule (3.3)  \\
+\hline
+2 &     parc (3.64)          &  carrefour (3.56)  \\
+\hline
+\end{tabular}
+}
+\vspace{0.5mm}
+\caption{Classes d'événements et de textures identifiées comme étant des marqueurs sonores. Dans chaque cellule, les marqueurs sont ordonnés par ordre décroissant de valeur $V$.}
+\label{tab:markers}
+\end{table}
+
+\subsubsection*{Espaces de représentation induits par les descripteurs sémantiques}
+
+Dans cette partie, nous évaluons la capacité d'une représentation sémantique à séparer les deux types d'environnement.
+Pour ce faire, nous calculons une précision au rang 5 ($p@5$) sur l'espace induit par les descripteurs sémantiques $S$, et ce pour chaque niveau d'abstraction (\cf~Section~\ref{sec:xp1_dataAna}).
+Les vecteurs $S$ sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.
+Nous ne considérons pas les classes de marqueurs de textures, ces dernières étant trop peu nombreuses.
+Pour les mêmes raisons nous ne considérons pas les classes de marqueurs d'événements du niveau d'abstraction $0$.
+Les résultats sont affichés sur la figure~\ref{fig:pa5}.
+
+En ce qui concerne $ET$, la $p@5$ est de $76\%$ pour le niveau d'abstraction 0, et reste supérieure à $86\%$ à partir du niveau d'abstraction 1.
+Ces résultats confirment qu'il est possible de clairement distinguer les deux types d'environnement en se basant seulement sur la présence ou l'absence des classes de sons.
+Nous notons également que, plus le niveau d'abstraction est élevé, plus la capacité de séparer les environnements est importante.
+En d'autres termes, plus nous sommes précis dans notre description de la composition des scènes, plus nous sommes à même d'établir une distinction claire entre les i- et ni-scènes.
+
+En considérant séparément $E$ et $T$, il apparaît 1) que la $p@5$ obtenue avec $E$ est similaire à celle obtenue avec $ET$, 2) que la $p@5$ obtenue avec $T$ est systématiquement inférieure d'environ $10$ à $15\%$ à celle de $E$.
+Ces résultats indiquent que l'information sémantique permettant de séparer les deux environnements est principalement portée par les événements.
+Ces résultats font, par ailleurs, écho aux travaux de  \cite{maffiolo_caracterisation_1999}, qui montrent que nous analysons de manière descriptive (en identifiant les sources) les scènes événementielles, \ie~composées d'événements sonores (\cf~Section~\ref{sec:simscene_sampleDataSet}).
+
+Enfin, il apparaît que la $p@5$ obtenue avec $E_{m}$ est égale, voire supérieure à celles obtenues avec $E$ et $ET$, et ce bien qu'une information partielle soit utilisée dans ce cas pour décrire les scènes.
+La dimension des vecteurs de description $S$ pour $E_m$ est en effet inférieure à la dimension des vecteurs $S$ pour $E$, qui est elle même inférieure à celle obtenue dans le cas où toutes les classes sont utilisées ($ET$).
+De plus, dans le cas où les marqueurs ne sont pas pris en compte pour la description ($E_{w/o,m}$), les résultats chutent, passant même en dessous de ceux obtenus en ne considérant que les textures.
+Cela confirme que la majorité de l'information sémantique permettant de faire la distinction entre i-scènes et ni-scènes est incluse dans les marqueurs.
+
+En résumé, nous déduisons de cette analyse les points suivants:
+
+\begin{enumerate}
+\item contrairement à ce que nous avions constaté avec les descripteurs structurels, une description sémantique de la composition des scènes, en terme de présence/absence de sources sonores, permet de bien distinguer les deux types d'environnement (i ou ni);
+\item l'information sémantique est majoritairement portée par les classes d'événements sonores;
+\item parmi les classes d'événements, seule une partie, \ie~les marqueurs sonores, est nécessaire pour faire la distinction entre les i- et ni-scènes.
+\end{enumerate}
+
+Maintenant que nous avons isolé les classes typiques des i- et ni-scènes, et vérifié que la distinction entre ces environnements dépendait de la présence de ces classes, il reste à voir si une description structurelle des scènes, basée uniquement sur ces marqueurs sonores, permet de caractériser l'agrément perçu, mieux qu'une description structurelle globale.
+
+\begin{figure}[t]
+        \myfloatalign
+        \includegraphics[width=\linewidth]{gfx/ch_5/pa5_1}
+       \caption{$P@5$ obtenues en considérant la matrice de dissimilarité résultant des distances par paires de Hamming calculées entre les vecteurs des descripteurs sémantiques des scènes.
+Les vecteurs sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.}\label{fig:pa5}
+\end{figure}
+
+\subsubsection*{Influence des marqueurs sonores sur l'agrément perçu}
+
+Nous évaluons les corrélations entre $\mathcal{A}_{scene}$ et les descripteurs structurels.
+Pour cette section, les descripteurs structurels sont calculés en tenant compte des marqueurs sonores précédemment identifiés.
+Nous définissons $X_m$ le descripteur $X$ calculé en ne prenant en compte que les sons des marqueurs.
+A l'inverse, nous définissons $X_b$ ($b$: pour «~bruit~») le descripteur $X$ calculé en prenant en compte toutes les classes de sons, excepté les marqueurs.
+Lorsque le descripteur caractérise une i-scène (idem pour une ni-scène), nous ne considérons, pour le calcul, que les marqueurs identifiés pour les i-scènes (ou pour les ni-scènes), que nous nommons i-marqueurs (ou ni-marqueurs).
+Les résultats sont affichés sur le tableau~\ref{tab:corrMarkers}.
+
+Concernant les niveaux sonores (\cf~Figures~\ref{fig:soundlevelMarkera} et~\ref{fig:soundlevelMarkerd}), les mêmes tendances sont observées entre $L_m$, $L(E)_m$ et $L(T)_m$, d'une part, et $L$, $L(E)$ et $L(T)$, d'autre part.
+Que l'on considère uniquement les marqueurs, ou l'ensemble des classes, il s'avère que :
+
+\begin{enumerate}
+\item il existe une différence significative entre les niveaux des i- et ni-scènes ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
+\item le niveau sonore des scènes est majoritairement porté par les événements sonores, comparé aux textures sonores;
+\item le niveau sonore des événements a une influence sur la perception de l'agrément pour les ni-scènes, mais pas pour les i-scènes;
+\item le niveau sonore des textures ne joue aucun rôle dans la perception de l'agrément.
+\end{enumerate}
+
+En conclusion, le niveau des ni-marqueurs a une influence négative sur l'agrément pour les ni-scènes. En revanche le niveau des i-marqueurs n’impacte pas l'agrément perçu pour les i-scènes.
+
+En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
+C'est la première fois qu'un indicateur objectif nous permet de préciser l'agrément des environnements agréables.
+Ceci nous amène à conclure que le niveau des classes de sons n'étant pas typiques d'un environnement agréable a un impact négatif sur l'agrément.
+
+Par ailleurs, alors que $L(T)$ ne présentait pas de corrélation pour les ni-scènes, une corrélation négative forte est observée pour $L(T)_b$ ($r-0.73$, $p<0.01$).
+Ce fait indique que les niveaux des classes de textures n'étant pas des marqueurs n'affectent pas l'agrément perçu de la même manière pour les i- et ni-scènes.
+Les niveaux semblent avoir un effet négatif pour les ni-scènes, alors que pour les i-scènes, aucun effet n'est relevé.
+
+Pour finir, nous considérons un dernier groupe de descripteurs, nommément $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiffa} et~\ref{fig:soundlevelMarkerDiffd}).
+Ces descripteurs expriment la différence entre les niveaux des marqueurs, et ceux des autres classes de sons.
+Ils traduisent l'émergence des marqueurs par rapport à la mixture sonore.
+
+Pour les i-scènes, une corrélation modérée et positive est observée pour $L_m-L_b$ ($r=0.67$, $p<0.01$) et $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
+Pour les ni-scènes, aucune corrélation n'est observée.
+Dans le cas des i-scènes, ce n'est donc pas le niveau absolu des marqueurs qui importe, mais leur niveau relatif, par rapport aux autres sons qui composent la scène.
+On observe donc pour les environnements idéaux un double mécanisme perceptif:
+
+\begin{itemize}
+\item plus le niveau absolu des sons n'étant pas des i-marqueurs est élevé, plus l'agrément est faible,
+\item plus le niveau relatif des i-marqueurs, par rapport aux autres sons, est élevé, plus l'agrément est élevé.
+\end{itemize}
+
+Pour les ni-scènes, le fait que nous observions des corrélations pour $L_m$ et $L(E)_m$, et aucune pour $L_m-L_b$ et $L(E)_m-L(E)_b$, montre que c'est bien le niveau absolu qui importe.
+
+\begin{table}[t]
+\setlength{\tabcolsep}{3pt}
+\centering
+{\renewcommand{\arraystretch}{1}
+\centering
+\begin{tabular}{l r r}
+                  &   i-scenes                  & ni-scenes \\
+\hline
+$L_m$              & 0.03  ($p=0.88$)           & \textbf{-0.75} ($p<0.01$) \\
+$L(E)_m$           & 0.08  ($p=0.66$)           & \textbf{-0.71} ($p<0.01$) \\
+$L(T)_m$           & -0.11 ($p=0.66$)           & -0.17 ($p=0.37$) \\
+$L_b$              & \textbf{-0.52} ($p<0.01$)  & -0.32 ($p=0.06$) \\
+$L(E)_b$           & \textbf{-0.51} ($p<0.01$)  & -0.30 ($p=0.07$) \\
+$L(T)_b$           & -0.32 ($p=0.05$)           & \textbf{-0.73} ($p<0.01$) \\
+$L_m-L_b$          & \textbf{0.67} ($p<0.01$)   & -0.31 ($p=0.07$) \\
+$L(E)_m-L(E)_b$    & \textbf{0.66} ($p<0.01$)   & -0.28 ($p=0.10$) \\
+$L(T)_m-L(T)_b$    & 0.16 ($p=0.54$)            & 0.21 ($p=0.28$) \\
+\hline
+\end{tabular}
+}
+\vspace{0.5mm}
+\caption{Coefficients de corrélation linéaire calculés entre l'agrément perçu moyen $\mathcal{A}_{scene}$ de l'expérience 1.b et les descripteurs structurels relatifs à la présence des marqueurs sonores.}
+\label{tab:corrMarkers}
+\end{table}
+
+\begin{figure}[t]
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_7}\label{fig:soundlevelMarkera}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_9}\label{fig:soundlevelNoisea}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_19}\label{fig:soundlevelMarkerDiffa}}\par
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_8}\label{fig:soundlevelMarkerd}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_10}\label{fig:soundlevelNoised}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
+        \caption{Dispersions des descripteurs structurels de niveaux sonores relatifs à la présence des marqueurs $L_m$ (a, d), $L(E)_m$ (b, e) et $L(T)_m$ (c, f), en fonction du type de scènes (a, b, c) et de l'agrément perçu $\mathcal{A}_{scene}$ de l'expérience 1.b (d, e, f).}\label{fig:soundlevelMarker}
+\end{figure}
+
+\subsection{Discussions}
+
+De cette analyse, nous retenons les points suivants:
+
+\begin{itemize}
+\item \emph{distinguer les i- et ni-scènes}: les descripteurs sémantiques, ainsi que les descripteurs structurels globaux ($L$, $L(E)$ et $L(T)$), permettent de faire la distinction entre les i-scènes et les ni-scènes.
+La description sémantique semble être plus performante;
+\item \emph{événements ou textures}: que ce soit pour les descripteurs sémantiques ou structurels, ce sont majoritairement les événements qui permettent de distinguer les deux types d'environnement, les textures n'apportant, au mieux, qu'une information limitée;
+\item \emph{prédire l'agrément}: si l'on considère les corrélations entre les descripteurs structurels et l'agrément, il semble que la manière de percevoir la qualité de l'environnement diffère en fonction de la nature de ce dernier (i ou ni).
+Il n'apparaît pas envisageable de considérer un même jeu de descripteurs pour prédire, à la fois, l'agrément des i-scènes, et l'agrément des ni-scènes :
+
+\begin{itemize}
+
+\item \emph{pour les ni-scènes}, ce sont le niveau global ($L$ et $L(E)$), ou le niveau des marqueurs sonores ($L_m$ et $L(E)_{m}$), qui impactent négativement l'agrément.
+On note ici que prendre en compte les contributions de différentes sources n'améliore pas la capacité de prédiction de l'agrément, par rapport à une analyse holistique de l'environnement.
+
+\item \emph{pour les i-scènes}, par contre, prédire l'agrément requiert d'étudier, de manière séparée, les caractéristiques des marqueurs sonores, et celles de l'ensemble des autres sons.
+Ainsi, le niveau des marqueurs pris relativement au bruit ($L(E)_m-L(E)_b$ et $L_m-L_b$) est positivement corrélé à l'agrément, alors que le niveau du bruit ($L_b$ et $L(E)_b$) est, lui, négativement corrélé.
+\end{itemize}
+\end{itemize}
+
+Le fait que l'agrément des i-scènes ne se corrèle pas avec des descripteurs physiques holistiques, contrairement à l'agrément des ni-scènes, a récemment été déjà observé \cite{gozalo2015relationship}.
+
+L'existence de deux modes de perception, mobilisant différents types de descripteurs, et dépendant de la nature (dans notre cas hédonique) des stimuli, est un phénomène qui se rapproche de celui observé pour la perception des textures (\cf~Section~\ref{sec:simscene_sampleDataSet}).
+Le cerveau adapte sa manière de traiter l'information (résumé statistique pour les textures, description fine pour les événements) suite à une prise de décision antérieure quant à la nature du stimulus (à savoir «~est-ce un événement ou une texture ?~»).
+De la même manière, les indicateurs actifs dans le jugement de l'agrément dépendent, eux aussi, d'une identification préalable de la nature hédonique globale de l'environnement (idéale ou non idéale).
 

### 2017-07-05 lafaySimulatedScenePerception.tex
one step further

@@
 
-\subsubsection*{Marqueurs sonores}
-
-Nous avons mis en évidence que, qualitativement, la composition des sources sonores des scènes diffère selon les types d'environnement (i ou ni).
-Nous essayons de voir maintenant si, parmi ces classes, certaines sont typiques d'un environnement en particulier.
-Pour ce faire, nous utilisons le V-test (\cf~Section~\ref{sec:xp1_dataAna}), en considérant séparément chaque niveau d'abstraction.
-Les résultats sont présentés dans le tableau~\ref{tab:markers}.
-
-Concernant les événements sonores, 9 marqueurs sont identifiés sur l'ensemble des niveaux d'abstraction.
-Comme la figure~\ref{fig:soundsource} le laissait présager, les classes relatives à la présence humaine (\emph{pas homme béton}, \emph{sonnette vélo}), et à la nature (\emph{animaux, oiseaux}, \emph{chants d'oiseaux}) sont des marqueurs de i-scènes.
-Nous notons également la présence de la classe \emph{cloche} dans les marqueurs d'un environnement idéal.
-Ce fait est possiblement dû au \emph{background} socio-culturel des sujets, dans leur grande majorité, des citoyens européens.
-En effet, selon Schafer, un son reconnu par un individu comme faisant partie intégrante de son environnement est bien accepté.
-Les marqueurs de ni-scènes sont des classes faisant référence à des sons de travaux (\emph{travaux}), ou suggérant un trafic dense (\emph{klaxon}, \emph{sirène}).
-
-Concernant les textures sonores, 5 marqueurs sont identifiés.
-Pour les i-scènes, il s'agit de classes faisant référence à des ambiances amorphes, calmes, (\emph{cour-intérieur} et \emph{parc}).
-Pour les ni-scènes, il s'agit, comme pour les événements, de classes faisant référence à des bruits de travaux (\emph{travaux} et \emph{véhicule de travaux}), ainsi que d'une classe faisant référence au trafic (\emph{carrefour}).
-
-Bien que l'ensemble des marqueurs identifiés soient intuitifs, aucune des classes d'événements faisant directement référence aux bruits de véhicules motorisés n'est un marqueur, exception faite de la classe de textures \emph{carrefour}.
-Pour représenter un trafic désagréable, les sujets ont porté leurs choix sur les classes \emph{klaxon} et \emph{sirène}.
-On peut supposer que les sons isolés de véhicules sont compris comme faisant partie intégrante de l'environnement urbain, et ne sont donc pas particulièrement associés à un environnement désagréable.
+\subsubsection*{Sound markers}
+
+We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs regarding the type of environment (i or ni).
+We’re now trying to to investigate if some of the sound classes are specific to a given environment.
+For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna}, while considering separately each abstraction level.
+Results are presented in Table~\ref{tab:markers}.
+
+Regarding the sound events, 9 markers are identified at all the abstraction levels.
+As Figure~\ref{fig:soundsource} could let it predict, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i- scenes markers; we also notice the \emph{church bell} class as an i- scenes marker.
+This latter result may be due to the socio-cultural background of the participants who are mostly european citizens.
+In fact, according to Schafer, a sound that is identified by a person as being an element of his/her environment, is well accepted.
+Sound markers of ni- scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
+
+Regarding the textures, 5 markers are identified.
+For the i- scenes, they are classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}).
+for the ni- scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).
+
+Although the whode identified markers are intuitive, none of the event classes related to the noise of the motor vehicles is a marker, except for the texture class \emph{crossroad}.
+To simulate an unpleasant traffic, participants chose the classes \emph{horn} or \emph{siren}.
+One can suppose that isolated motor vehicle sounds are understood as taking fully part of the urban environment, and thus are not especially linked to an unpleasant soundscape.
 
@@
 \begin{tabular}{c c c c}
-Niveau        & \multicolumn{2}{c}{Marqueurs sonores événements} \\
-d'abstraction & i-scènes & ni-scènes \\
+Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
+level & i-scenes & ni-scenes \\
 \hline
-0  &                          & travaux (3.78)  \\
+0  &                          & construction work (3.78)  \\
 \hline
-  & cloche  (4.5)             & klaxon  (3.9) \\
-1 & sonnette vélo  (4.3)      & sirène (3.9)\\
+  & church bell  (4.5)             & car horn  (3.9) \\
+1 & bicycle bell  (4.3)      & siren (3.9)\\
   & animal (4.2)              &       \\
    \hline
-  & oiseau        (4.8)       & klaxon  (4.0)\\
-2 & cloche  (4.4)             & sirène (4.0)\\
-  & sonnette vélo    (4.2)             &       \\
+  & bird        (4.8)       & car horn  (4.0)\\
+2 & church bell  (4.4)             & siren (4.0)\\
+  & bicycle bell     (4.2)             &       \\
    \hline
-  & chant oiseau (4.8)        & klaxon  (4.1)\\
-3 & cloche   (4.3)            & sirène (4.0)\\
-  & sonnette vélo     (4.2)   &       \\
-  & pas chaussure  (3.6)      &  \\
+  & bird song (4.8)        & klaxon  (4.1)\\
+3 & church bell   (4.3)            & siren (4.0)\\
+  & bicycle bell      (4.2)   &       \\
+  & foot steps  (3.6)      &  \\
   &                           & \\
-  & \multicolumn{2}{c}{Marqueurs sonores textures}      \\
-  & i-scènes & ni-scènes \\
+  & \multicolumn{2}{c}{Texture sound markers}      \\
+  & i-scenes & ni-scenes \\
 \hline
-0 &     cour int./parc (4.1) &  travaux (3.9)  \\
+0 &     courtyard / park (4.1) &  construction (3.9)  \\
 \hline
-1 &     parc (3.65)          &  carrefour (3.6)  \\
-  &                          &  travaux véhicule (3.3)  \\
+1 &     park (3.65)          &  crossroad (3.6)  \\
+  &                          &  construction vehicule (3.3)  \\
 \hline
-2 &     parc (3.64)          &  carrefour (3.56)  \\
+2 &     park (3.64)          &  crossroad (3.56)  \\
 \hline
@@
 \vspace{0.5mm}
-\caption{Classes d'événements et de textures identifiées comme étant des marqueurs sonores. Dans chaque cellule, les marqueurs sont ordonnés par ordre décroissant de valeur $V$.}
+\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of $V$ value.}
 \label{tab:markers}
@@
 
-\subsubsection*{Espaces de représentation induits par les descripteurs sémantiques}
-
-Dans cette partie, nous évaluons la capacité d'une représentation sémantique à séparer les deux types d'environnement.
-Pour ce faire, nous calculons une précision au rang 5 ($p@5$) sur l'espace induit par les descripteurs sémantiques $S$, et ce pour chaque niveau d'abstraction (\cf~Section~\ref{sec:xp1_dataAna}).
-Les vecteurs $S$ sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.
-Nous ne considérons pas les classes de marqueurs de textures, ces dernières étant trop peu nombreuses.
-Pour les mêmes raisons nous ne considérons pas les classes de marqueurs d'événements du niveau d'abstraction $0$.
-Les résultats sont affichés sur la figure~\ref{fig:pa5}.
-
-En ce qui concerne $ET$, la $p@5$ est de $76\%$ pour le niveau d'abstraction 0, et reste supérieure à $86\%$ à partir du niveau d'abstraction 1.
-Ces résultats confirment qu'il est possible de clairement distinguer les deux types d'environnement en se basant seulement sur la présence ou l'absence des classes de sons.
-Nous notons également que, plus le niveau d'abstraction est élevé, plus la capacité de séparer les environnements est importante.
-En d'autres termes, plus nous sommes précis dans notre description de la composition des scènes, plus nous sommes à même d'établir une distinction claire entre les i- et ni-scènes.
-
-En considérant séparément $E$ et $T$, il apparaît 1) que la $p@5$ obtenue avec $E$ est similaire à celle obtenue avec $ET$, 2) que la $p@5$ obtenue avec $T$ est systématiquement inférieure d'environ $10$ à $15\%$ à celle de $E$.
-Ces résultats indiquent que l'information sémantique permettant de séparer les deux environnements est principalement portée par les événements.
-Ces résultats font, par ailleurs, écho aux travaux de  \cite{maffiolo_caracterisation_1999}, qui montrent que nous analysons de manière descriptive (en identifiant les sources) les scènes événementielles, \ie~composées d'événements sonores (\cf~Section~\ref{sec:simscene_sampleDataSet}).
-
-Enfin, il apparaît que la $p@5$ obtenue avec $E_{m}$ est égale, voire supérieure à celles obtenues avec $E$ et $ET$, et ce bien qu'une information partielle soit utilisée dans ce cas pour décrire les scènes.
-La dimension des vecteurs de description $S$ pour $E_m$ est en effet inférieure à la dimension des vecteurs $S$ pour $E$, qui est elle même inférieure à celle obtenue dans le cas où toutes les classes sont utilisées ($ET$).
-De plus, dans le cas où les marqueurs ne sont pas pris en compte pour la description ($E_{w/o,m}$), les résultats chutent, passant même en dessous de ceux obtenus en ne considérant que les textures.
-Cela confirme que la majorité de l'information sémantique permettant de faire la distinction entre i-scènes et ni-scènes est incluse dans les marqueurs.
-
-En résumé, nous déduisons de cette analyse les points suivants:
+\subsubsection*{Representation space induced by semantic features}
+
+In this section, we evaluate the ability of a semantic representation to split the two types of environment.
+For that, we compute a rank 5-precision ($p@5$) on the space induced by the semantic features $S$, and for each abstraction level, see Section~\ref{sec:xp1_dataAna}.
+The vectors $S$ are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers ($E_{w/o,m}$).
+Texture classes corresponding to sound markers are not considered as these latter are not numerous enough.
+For the same reasons, event classes corresponding to sound markers at abstraction level $0$ are also discarded.
+Results are shown on Figure~\ref{fig:pa5}.
+
+Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and stay above $86\%$ from the abstraction level $1$.
+These results confirm that it is possible to make a clear distinction between the two types of environment on the unique basis of presence / absence of sound classes.
+We also notice that the higher the abstraction level is, the more important is the ability to make this distinction.
+In other words, the more one is precise in the scene description, the more one is able to clearly differentiate between i- and ni- scenes.
+
+Considering $E$ and $T$ separately, ir appears: 1) that the $p@5$ obtained with $E$ is similar to the one obtained with $ET$ ; 2) that the $p@5$ obtained with $T$ is systematically lower than the one obtained with $E$, by $10\%$ or $15\%$.
+Those results indicate that the semantic information allowing to split the two environments is mainly contained in the events.
+Moreover, these results are in line with the work done in \cite{maffiolo_caracterisation_1999} showing that we analyse in a causal manner (i.e.
+by identifying the sources) the event scenes which are composed of several sound events, see Section~\ref{sec:simscene_sampleDataSet}.
+
+Finally, it appears that the $p@5$ obtained with $E_m$ is equal – or above – to the ones obtained with $E$ or $ET$, although a partial information is used in that case to describe the scenes.
+The dimension of the vectors $S$ for $E_m$ is lower than the dimension of vector $S$ for $E$, itself lower than the dimension of vector S obtained when all the classes are considered ($ET$).
+Moreover, in the case when the sound markers are not taken into account for the description ($E_{w/o,m}$), the results fall even going under those obtained with only the textures.
+This confirms that most of the semantic information allowing to differentiate between i- and ni- scenes is included in the markers.
+
+To sum up, we deduce the following points from this analysis:
 
@@
 
-Maintenant que nous avons isolé les classes typiques des i- et ni-scènes, et vérifié que la distinction entre ces environnements dépendait de la présence de ces classes, il reste à voir si une description structurelle des scènes, basée uniquement sur ces marqueurs sonores, permet de caractériser l'agrément perçu, mieux qu'une description structurelle globale.
+Since we have extracted the typical classes of the i- and ni- scenes and verified that the distinction between them was dependant to the presence of these classes, we have now to investigate wether a structural description of the scenes only based on sound markers could allow us to characterize the perceived pleasantness, perhaps better than a global structural description.
 
@@
         \includegraphics[width=\linewidth]{gfx/ch_5/pa5_1}
-       \caption{$P@5$ obtenues en considérant la matrice de dissimilarité résultant des distances par paires de Hamming calculées entre les vecteurs des descripteurs sémantiques des scènes.
-Les vecteurs sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.}\label{fig:pa5}
+       \caption{$P@5$ obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors.
+The vectors are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers $E_{w/o,m}$.}\label{fig:pa5}
 \end{figure}
 
-\subsubsection*{Influence des marqueurs sonores sur l'agrément perçu}
-
-Nous évaluons les corrélations entre $\mathcal{A}_{scene}$ et les descripteurs structurels.
-Pour cette section, les descripteurs structurels sont calculés en tenant compte des marqueurs sonores précédemment identifiés.
-Nous définissons $X_m$ le descripteur $X$ calculé en ne prenant en compte que les sons des marqueurs.
-A l'inverse, nous définissons $X_b$ ($b$: pour «~bruit~») le descripteur $X$ calculé en prenant en compte toutes les classes de sons, excepté les marqueurs.
-Lorsque le descripteur caractérise une i-scène (idem pour une ni-scène), nous ne considérons, pour le calcul, que les marqueurs identifiés pour les i-scènes (ou pour les ni-scènes), que nous nommons i-marqueurs (ou ni-marqueurs).
-Les résultats sont affichés sur le tableau~\ref{tab:corrMarkers}.
-
-Concernant les niveaux sonores (\cf~Figures~\ref{fig:soundlevelMarkera} et~\ref{fig:soundlevelMarkerd}), les mêmes tendances sont observées entre $L_m$, $L(E)_m$ et $L(T)_m$, d'une part, et $L$, $L(E)$ et $L(T)$, d'autre part.
-Que l'on considère uniquement les marqueurs, ou l'ensemble des classes, il s'avère que :
+\subsubsection*{Influence of the sound markers on the perceived pleasantness}
+
+We evaluate the correlations between $\mathcal{A}_{scene}$ and the structural features.
+In this section, the structural features are computed by taking into account of the sound markers previoulsy identified.
+We define $Xm$, the $X$ feature computed by taking into account only the sound markers.
+On the opposite, we define $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
+When the feature characterize an i- scene (idem for a ni- scene), we consider for the computation only the markers identified for the i- scenes (or the ni- scenes); we named them, i- markers (or ni- markers).
+Results are shown in Table~\ref{tab:corrMarkers}.
+
+Concerning the sound levels (\cf~Figures~\ref{fig:soundlevelMarkera} et~\ref{fig:soundlevelMarkerd}), the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ et $L(T)$ on the other side.
+Whether all the classes or only the markers ae considered, it appears that:
 
 \begin{enumerate}
-\item il existe une différence significative entre les niveaux des i- et ni-scènes ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
-\item le niveau sonore des scènes est majoritairement porté par les événements sonores, comparé aux textures sonores;
-\item le niveau sonore des événements a une influence sur la perception de l'agrément pour les ni-scènes, mais pas pour les i-scènes;
-\item le niveau sonore des textures ne joue aucun rôle dans la perception de l'agrément.
+\item a significative difference between levels of i- and ni- scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
+\item the sound level of scenes is mainly related to the sound events, compared to the textures;
+\item the sound level of events has an influence on the perception of pleasantness for ni- scenes, but not for i- scenes;
+\item the sound level of textures doesn’t play any role in the perception of the pleasantness.
 \end{enumerate}
 
-En conclusion, le niveau des ni-marqueurs a une influence négative sur l'agrément pour les ni-scènes. En revanche le niveau des i-marqueurs n’impacte pas l'agrément perçu pour les i-scènes.
+In conclusion, the level of ni- markers has a negative influence on the pleasantness for the ni- scenes.
+On the other hand, the level of i- markers doesn’t influence the perceived pleasantness for the i- scenes.
 

### 2017-07-05 lafaySimulatedScenePerception.tex
NM insertion done

@@
 
-En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
+% En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
 C'est la première fois qu'un indicateur objectif nous permet de préciser l'agrément des environnements agréables.
@@
 
-Par ailleurs, alors que $L(T)$ ne présentait pas de corrélation pour les ni-scènes, une corrélation négative forte est observée pour $L(T)_b$ ($r-0.73$, $p<0.01$).
+Considering the non markers classes (see~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), we can observe on the i- scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
+It is the first time that an objective feature allows us to define the pleasantness of the pleasant environments.
+This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.
+
+% Par ailleurs, alors que $L(T)$ ne présentait pas de corrélation pour les ni-scènes, une corrélation négative forte est observée pour $L(T)_b$ ($r-0.73$, $p<0.01$).
 Ce fait indique que les niveaux des classes de textures n'étant pas des marqueurs n'affectent pas l'agrément perçu de la même manière pour les i- et ni-scènes.
@@
 
-Pour finir, nous considérons un dernier groupe de descripteurs, nommément $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiffa} et~\ref{fig:soundlevelMarkerDiffd}).
+Besides, whereas $L(T)$ didn’t show any significative correlation for ni- scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
+This indicates that the level of non-markers texture classes doesn’t influence the perceived pleasantness in the same way for i- and ni- scenes.
+Sound level seems to have a negative effect for the ni- scenes, and no significant effect for the i- scenes.
+
+% Pour finir, nous considérons un dernier groupe de descripteurs, nommément $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiffa} et~\ref{fig:soundlevelMarkerDiffd}).
 Ces descripteurs expriment la différence entre les niveaux des marqueurs, et ceux des autres classes de sons.
@@
 
-Pour les i-scènes, une corrélation modérée et positive est observée pour $L_m-L_b$ ($r=0.67$, $p<0.01$) et $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
+We consider a last group of features, namely $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiffa} et~\ref{fig:soundlevelMarkerDiffd}).
+These features describe the difference between the markers level and those of the other sound classes.
+They express the saliency of the markers with regards to the sound mixture.
+
+% Pour les i-scènes, une corrélation modérée et positive est observée pour $L_m-L_b$ ($r=0.67$, $p<0.01$) et $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
 Pour les ni-scènes, aucune corrélation n'est observée.
@@
 
-\begin{itemize}
-\item plus le niveau absolu des sons n'étant pas des i-marqueurs est élevé, plus l'agrément est faible,
-\item plus le niveau relatif des i-marqueurs, par rapport aux autres sons, est élevé, plus l'agrément est élevé.
-\end{itemize}
-
-Pour les ni-scènes, le fait que nous observions des corrélations pour $L_m$ et $L(E)_m$, et aucune pour $L_m-L_b$ et $L(E)_m-L(E)_b$, montre que c'est bien le niveau absolu qui importe.
+For the i- scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
+for the ni- scenes no correlation is observed.
+Thus, for the i- scenes, it is not the absolute markers level that is important, but their relative level with regards to the other sounds composing the scene.
+Then, we can induce a double perceptual mechanism for the ideal environments:
+
+\begin{itemize}
+\item the higher the absolute level of sounds not being i- markers, the weaker the pleasantness,
+\item the higher the relative level of i- markers is, with regards to the other sounds, the higher the pleasantness.
+\end{itemize}
+
+For the ni -scenes, the fact that we observe significative correlations for Lm and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.
 
@@
 \vspace{0.5mm}
-\caption{Coefficients de corrélation linéaire calculés entre l'agrément perçu moyen $\mathcal{A}_{scene}$ de l'expérience 1.b et les descripteurs structurels relatifs à la présence des marqueurs sonores.}
+\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ (Exp. 1.b) and structural features related to the presence of sound markers.}
 \label{tab:corrMarkers}
@@
         {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
-        \caption{Dispersions des descripteurs structurels de niveaux sonores relatifs à la présence des marqueurs $L_m$ (a, d), $L(E)_m$ (b, e) et $L(T)_m$ (c, f), en fonction du type de scènes (a, b, c) et de l'agrément perçu $\mathcal{A}_{scene}$ de l'expérience 1.b (d, e, f).}\label{fig:soundlevelMarker}
+        \caption{Distribution of the relative sound levels structural features related to the presence of markers $L_m$ (a, d), $L(E)_m$ (b, e) and $L(T)_m$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
 \end{figure}
 
-\subsection{Discussions}
-
-De cette analyse, nous retenons les points suivants:
-
-\begin{itemize}
-\item \emph{distinguer les i- et ni-scènes}: les descripteurs sémantiques, ainsi que les descripteurs structurels globaux ($L$, $L(E)$ et $L(T)$), permettent de faire la distinction entre les i-scènes et les ni-scènes.
-La description sémantique semble être plus performante;
-\item \emph{événements ou textures}: que ce soit pour les descripteurs sémantiques ou structurels, ce sont majoritairement les événements qui permettent de distinguer les deux types d'environnement, les textures n'apportant, au mieux, qu'une information limitée;
-\item \emph{prédire l'agrément}: si l'on considère les corrélations entre les descripteurs structurels et l'agrément, il semble que la manière de percevoir la qualité de l'environnement diffère en fonction de la nature de ce dernier (i ou ni).
-Il n'apparaît pas envisageable de considérer un même jeu de descripteurs pour prédire, à la fois, l'agrément des i-scènes, et l'agrément des ni-scènes :
-
-\begin{itemize}
-
-\item \emph{pour les ni-scènes}, ce sont le niveau global ($L$ et $L(E)$), ou le niveau des marqueurs sonores ($L_m$ et $L(E)_{m}$), qui impactent négativement l'agrément.
-On note ici que prendre en compte les contributions de différentes sources n'améliore pas la capacité de prédiction de l'agrément, par rapport à une analyse holistique de l'environnement.
-
-\item \emph{pour les i-scènes}, par contre, prédire l'agrément requiert d'étudier, de manière séparée, les caractéristiques des marqueurs sonores, et celles de l'ensemble des autres sons.
-Ainsi, le niveau des marqueurs pris relativement au bruit ($L(E)_m-L(E)_b$ et $L_m-L_b$) est positivement corrélé à l'agrément, alors que le niveau du bruit ($L_b$ et $L(E)_b$) est, lui, négativement corrélé.
-\end{itemize}
-\end{itemize}
-
-Le fait que l'agrément des i-scènes ne se corrèle pas avec des descripteurs physiques holistiques, contrairement à l'agrément des ni-scènes, a récemment été déjà observé \cite{gozalo2015relationship}.
-
-L'existence de deux modes de perception, mobilisant différents types de descripteurs, et dépendant de la nature (dans notre cas hédonique) des stimuli, est un phénomène qui se rapproche de celui observé pour la perception des textures (\cf~Section~\ref{sec:simscene_sampleDataSet}).
-Le cerveau adapte sa manière de traiter l'information (résumé statistique pour les textures, description fine pour les événements) suite à une prise de décision antérieure quant à la nature du stimulus (à savoir «~est-ce un événement ou une texture ?~»).
-De la même manière, les indicateurs actifs dans le jugement de l'agrément dépendent, eux aussi, d'une identification préalable de la nature hédonique globale de l'environnement (idéale ou non idéale).
+\subsection{Discussion}
+
+From this analysis, we can outline the following points:
+
+\begin{itemize}
+\item \emph{diferentiating the i- and ni-scenes}: the semantic features, and the global structural features ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni- scenes.
+The semantic description seems to be more powerful;
+\item \emph{events or textures}: whatever semantic or structural features, events are mainly useful to differentiate the two types of events; textures bring a limited amount of information;
+\item \emph{pleasantness prediction}: if we consider the correlations between structural features and pleasantness, it seems that the way to perceive the quality of a given environment depends on its nature (i or ni).
+It is not conceivable to consider a same set of features to predict the pleasantness of i- and ni- scenes in the same time:
+
+\begin{itemize}
+
+\item \emph{for ni-scenes}, global levelsl ($L$ et $L(E)$), or level of sound markers ($L_m$ et $L(E)_{m}$), have a negative influence on pleasantness.
+We can notice here taht taking into account the contribution of each the different sources of the scene doesn’t improve the prediction performance, wit regards to a global analysis of the environment.
+
+\item \emph{for i-scenes}, on the contrary, the pleasantness prediction needs to study separately the sound markers characteristics and those of the other sounds.
+Thus, the markers level, relative to the noise ($L(E)_m-L(E)_b$ et $L_m-L_b$) is positively correlated to the pleasantness, whereas the noise level ($L_b$ et $L(E)_b$)is negatively correlated.
+\end{itemize}
+\end{itemize}
+
+The fact that the pleasantness of the i- scenes is not correlated to global physical features, contrary to pleasantness of ni- scenes has been studied recently \cite{gozalo2015relationship}.
+
+We thus observe two perceptual modes that involve different types of features, and that rely on the hedonic nature  of the stimuli.
+A phenomenon similar is observed for the perception of textures, see Section~\ref{sec:simscene_sampleDataSet}.
+It seems that the brain adapts its way to encode the information (statistic summary for textures, fine description for events) following a previous decision making with regards to the nature of the stimuli, (()\ie is it an event or a texture ?).
+Similarly, features that are considered for the pleasantness judgment also depends on a preliminary judgment of the global hedonic nature of the environment (ideal or non ideal).
 

### 2017-07-05 lafaySimulatedScenePerception.tex
typos

@@
 
-First, a subject can be asked to describe a given sound environnment following a given perceptive scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}.
+First, a subject can be asked to describe a given sound  following a given perceptive scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}.
 The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}.
-That being said, as those stimuli can be recordered and analysed, it allow the experimenter to gather a coarse grain physical description of the stimuli.
+That being said, as those stimuli can be recorded and analysed, it allow the experimenter to gather a coarse grain physical description of the stimuli.
 
@@
 
-We believe that the availability of a fine grain description of the sound stimuli is of great insterest, beceause recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
+We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
 szeremeta2009analysis}.
@@
 
-\emph{Simscene} is an online digital audio tool whose first version have been developped as part of the HOULE prokect\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
+\emph{Simscene} is an online digital audio tool whose first version have been developed as part of the HOULE prokect\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
 It has been designed to run on the popular web browers \emph{Chrome} and \emph{Firefox}.
@@
 
-According to the Auditory Scene Analysis (ASA) theory, the brain processes seperately events emerging from different sources of a sound environment \cite{bregman1994auditory}.
+According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from different sources of a sound environment \cite{bregman1994auditory}.
 Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}.
@@
 
-Concerning sound textures, \ie~sounds that have stable caracteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can decide not to abstract the physical information available using some of the summary statistics discarding lower level information \cite{nelken2013ear}.
+Concerning sound textures, \ie~sounds that have stable characteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can decide not to abstract the physical information available using some of the summary statistics discarding lower level information \cite{nelken2013ear}.
 
@@
 
-The sound classes are organized hierachically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E.
+The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E.
 Rosch \cite{rosch1978cognition}.
@@
 
-All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie. relative to Full Scale. In our case, the full scale level is of 1 Volt.
+All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie relative to Full Scale. In our case, the full scale level is of 1 Volt.
 
@@
 This second experimental step aims at investigating our knowledge on pleasantness.
-For that, a second goup of participants are asked to evaluate the pleasantness of each simulated scene, on a semantic scale.
+For that, a second group of participants are asked to evaluate the pleasantness of each simulated scene, on a semantic scale.
 This evaluation protocol has two goals:
@@
 \end{itemize}
-For instance, the participant can insert ten dogs in the soundscape but he/she can’t insert one dog barking every 10 milliseconds.
+For instance, the participant can insert ten dogs in the soundscape but he/she cannot insert one dog barking every 10 milliseconds.
 These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation user interface.
@@
 During the tutorial, participants are asked to adjust the sound level in order to be comfortable; they can’t modify it in the following of the experiment.
-All the participants do the experiment simultaneously. They are equally split in three identical quiet rooms. They can’t talk together during the experiment.
+All the participants do the experiment simultaneously. They are equally split in three identical quiet rooms. Theyare not allowed to talk to each other during the experiment.
 Three experimenters (one in each room) are present during the whole duration of the experiment in order to check that everything goes well and possibly to answer questions.
@@
 The assessment is done in a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
-Before evaluating a scene, participants must listen to the first 2à seconds of the scene.
+Before evaluating a scene, participants must listen to the first 20 seconds of the scene.
 After the evaluation, they are free to switch to the next scene.
@@
 These first 10 scenes are played back again at the end of the experiment and the corresponding evaluations are the ones that are only taken into account in the results.
-The experiment is planned to last 30 minutes. The participants don’t know anything beforehand about the nature of the sound scenes.
+The experiment is planned to last 30 minutes. The participants do not know anything beforehand about the nature of the sound scenes.
 
@@
 The output sound level is the same for all the participants.
-All the participants do the experiment simultaneously. They can’t talk together during the experiment.
+All the participants do the experiment simultaneously. They are not allowed to talk to each other during the experiment.
 An experimenter is present during the whole duration of the experiment in order to check that everything goes well and possibly to answer questions.
@@
 To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference.
-In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in deciBels, with a reference of 1 Volt.
+In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt.
 The level is obtained by computing every second the quadratic mean of the signal and by averaging over the total duration of the scene.
@@
 \item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with regards to the two types of scene.
-the significativity is evaluated by a Student test on two independant populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$ ;
-\item \emph{analysis of the structural features}: in order to evaluate wether the emotional difference forced to the i- and ni- scenes has a significative influence \ie, if significative differences exist between structural features and perceived pleasantness, we measure this significativity with a Student test on two populations;
+the significativity is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$ ;
+\item \emph{analysis of the structural features}: in order to evaluate whether the emotional difference forced to the i- and ni- scenes has a significative influence \ie, if significative differences exist between structural features and perceived pleasantness, we measure this significativity with a Student test on two populations;
 \item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, we look at linear correlations between these features and $\mathcal{A}_{scene}$.
 For that, we use the Pearson correlation coefficient, on the basis of the standard methodology used in multidimensional studies;
-\item \emph{analysis of the semantic features}: in order to determine if the forced emotional difference has an influence on the scenes composition in terms of sound sources, or if specific souns classes are more frequently used to simulate a given environment (i or ni), we apply a V-test.
+\item \emph{analysis of the semantic features}: in order to determine if the forced emotional difference has an influence on the scenes composition in terms of sound sources, or if specific  classes are more frequently used to simulate a given environment (i or ni), we apply a V-test.
 This test is done at every level of abstraction, and separately, for texture and event classes.
@@
 
-First, in order to ensure the coherence of the data, wr check that none of the ni- scenes gets a $\mathcal{A}_{scene}$ higher than one of a i- scene.
+First, in order to ensure the coherence of the data, we check that none of the ni- scenes gets a $\mathcal{A}_{scene}$ higher than one of a i- scene.
 Four ni- scenes don’t respect that rule: they are removed, together with their corresponding i- scenes.
@@
 This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}.
-During their experiment, the authors ask to their participants to assess a set of soundscapes at a global level and then to do the same judgement at the moments when each of them detect a sound source.
-The study shows that there is no significative difference between global and averaged instantaneous judgements.
+During their experiment, the authors ask to their participants to assess a set of soundscapes at a global level and then to do the same judgment at the moments when each of them detect a sound source.
+The study shows that there is no significative difference between global and averaged instantaneous judgments.
 In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, \ie the events.
@@
 
-In this section, we analyse the deep relationships that could exist between structural features and perceived pleasantness.
-Contrary to the previous section, we don’t consider the emotional property of a scene on a binary mode (i vs.
+In this section, relationships that could exist between structural features and perceived pleasantness are investigated.
+Contrary to the previous section, we do not consider the emotional property of a scene on a binary mode (i vs.
 ni): here, we consider the mean pleasantness $\mathcal{A}_{scene}$ as a perceptual feature.
@@
 
-COncerning $L$, we can observe a strong negative correlation ($r=-0.77$, $p<0.01$) with $\mathcal{A}_{scene}$, indicating that the higher the sound level is, the more unpleasant the scene is perceived.
-Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship doesn’t occurr in the same way for i- and ni- scenes.
+Concerning $L$, we can observe a strong negative correlation ($r=-0.77$, $p<0.01$) with $\mathcal{A}_{scene}$, indicating that the higher the sound level is, the more unpleasant the scene is perceived.
+Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship do not occur in the same way for i- and ni- scenes.
 In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni- scenes ($r=-0.78$, $p<0.01$), but is quasi null for i-scenes.
@@
 One can suppose that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer than ni - scenes).
-Thus, the sound events keep a relative ability to predict the pleasantness of the ni- scenes, but the level of the textures doesn’t bring a lot of information whatever the type of environment is.
+Thus, the sound events keep a relative ability to predict the pleasantness of the ni- scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.
 
@@
 Those first outcomes tend to show that two modes of perception exists depending on the nature of the environment (i- or ni-), and each involve independent features.
-The fact that $L$ is not able to characterize the pleasantness of the i- scenes can lead us to think that all the sound sources don’t equally contribute to the perception of pleasantness, but that only the level of some of them can influence this percept.
-In order to go deeper in that direction, in the next section, we analyse these scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.
+The fact that $L$ is not able to characterize the pleasantness of the i- scenes can lead us to think that all the sound sources do not equally contribute to the perception of pleasantness, but that only the level of some of them can influence this percept.
+In order to go deeper in that direction, in the next section, we analyze these scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.
 
@@
 
-We observe a significant difference between i- and ni- scenes for the choise of classes.
+We observe a significant difference between i- and ni- scenes for the classes choice.
 The distribution of the classes is very similar to to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~lasses involving human presence and nature are prevailing in the i- scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni- scenes.
@@
 In fact,  Figure~\ref{fig:soundsourcea} show \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the participants for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes.
-the results don’t re-evaluate the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} pour les i-scenes, a level that is comparable to the of the \emph{bike} class, and much higher than all the personal vehicles classes.
+the results do not re-evaluate the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} pour les i-scenes, a level that is comparable to the of the \emph{bike} class, and much higher than all the personal vehicles classes.
 Though, \emph{public transport} classes are also strongly present in the ni-scenes, for instance more than \emph{light vehicle} or \emph{trucks} classes.
-On the basis of our results, the public transportation class can’t be considered as typical of an ideal urban soundscape.
+On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.
 
@@
 We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs regarding the type of environment (i or ni).
-We’re now trying to to investigate if some of the sound classes are specific to a given environment.
+We are now trying to to investigate if some of the sound classes are specific to a given environment.
 For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna}, while considering separately each abstraction level.
@@
 As Figure~\ref{fig:soundsource} could let it predict, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i- scenes markers; we also notice the \emph{church bell} class as an i- scenes marker.
-This latter result may be due to the socio-cultural background of the participants who are mostly european citizens.
+This latter result may be due to the socio-cultural background of the participants who are mostly European citizens.
 In fact, according to Schafer, a sound that is identified by a person as being an element of his/her environment, is well accepted.
@@
 
-Although the whode identified markers are intuitive, none of the event classes related to the noise of the motor vehicles is a marker, except for the texture class \emph{crossroad}.
+Although the whole set of identified markers are intuitive, none of the event classes related to the noise of the motor vehicles is a marker, except for the texture class \emph{crossroad}.
 To simulate an unpleasant traffic, participants chose the classes \emph{horn} or \emph{siren}.
@@
 Those results indicate that the semantic information allowing to split the two environments is mainly contained in the events.
-Moreover, these results are in line with the work done in \cite{maffiolo_caracterisation_1999} showing that we analyse in a causal manner (i.e.
+Moreover, these results are in line with the work done in \cite{maffiolo_caracterisation_1999} showing that we analyze in a causal manner (i.e.
 by identifying the sources) the event scenes which are composed of several sound events, see Section~\ref{sec:simscene_sampleDataSet}.
@@
 
-Since we have extracted the typical classes of the i- and ni- scenes and verified that the distinction between them was dependant to the presence of these classes, we have now to investigate wether a structural description of the scenes only based on sound markers could allow us to characterize the perceived pleasantness, perhaps better than a global structural description.
+Since we have extracted the typical classes of the i- and ni- scenes and verified that the distinction between them was dependent to the presence of these classes, we have now to investigate whether a structural description of the scenes only based on sound markers could allow us to characterize the perceived pleasantness, perhaps better than a global structural description.
 
@@
 We evaluate the correlations between $\mathcal{A}_{scene}$ and the structural features.
-In this section, the structural features are computed by taking into account of the sound markers previoulsy identified.
+In this section, the structural features are computed by taking into account of the sound markers previously identified.
 We define $Xm$, the $X$ feature computed by taking into account only the sound markers.
@@
 \item the sound level of events has an influence on the perception of pleasantness for ni- scenes, but not for i- scenes;
-\item the sound level of textures doesn’t play any role in the perception of the pleasantness.
+\item the sound level of textures does not play any role in the perception of the pleasantness.
 \end{enumerate}
@@
 In conclusion, the level of ni- markers has a negative influence on the pleasantness for the ni- scenes.
-On the other hand, the level of i- markers doesn’t influence the perceived pleasantness for the i- scenes.
+On the other hand, the level of i- markers does not influence the perceived pleasantness for the i- scenes.
 
@@
 
-Besides, whereas $L(T)$ didn’t show any significative correlation for ni- scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
-This indicates that the level of non-markers texture classes doesn’t influence the perceived pleasantness in the same way for i- and ni- scenes.
+Besides, whereas $L(T)$ did not show any significative correlation for ni- scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
+This indicates that the level of non-markers texture classes does not influence the perceived pleasantness in the same way for i- and ni- scenes.
 Sound level seems to have a negative effect for the ni- scenes, and no significant effect for the i- scenes.
@@
 \begin{itemize}
-\item \emph{diferentiating the i- and ni-scenes}: the semantic features, and the global structural features ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni- scenes.
+\item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global structural features ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni- scenes.
 The semantic description seems to be more powerful;
@@
 
-\item \emph{for ni-scenes}, global levelsl ($L$ et $L(E)$), or level of sound markers ($L_m$ et $L(E)_{m}$), have a negative influence on pleasantness.
-We can notice here taht taking into account the contribution of each the different sources of the scene doesn’t improve the prediction performance, wit regards to a global analysis of the environment.
+\item \emph{for ni-scenes}, global levels ($L$ et $L(E)$), or level of sound markers ($L_m$ et $L(E)_{m}$), have a negative influence on pleasantness.
+We can notice here that taking into account the contribution of each the different sources of the scene does not improve the prediction performance, wit regards to a global analysis of the environment.
 

### 2017-07-06 lafaySimulatedScenePerception.tex
fixing references

@@
 The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}.
-The three experiments,  1.a \emph{simulation}, 1.b \emph{pleasantness evaluation}, and 2 \emph{pleasantness evaluation after modification of the scenes} are respectively presented in Sections \ref{sec:simulation}, \ref{sec:evaluation}, and \ref{sec:modification}.
+The experiments  1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section \ref{sec:simulation}.
+Experiment 2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section \ref{sec:modification}.
 Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.
@@
 \section{The simulator}
-\label{sec:simscene}
+\label{sec:simulator}
 
@@
 
-\section{Expérience 1}
-\label{sec:xp1}
-
-\subsection{Objectif}
+\section{Experiment 1}
+\label{sec:simulation}
+
+\subsection{Objective}
 
@@
 \begin{itemize}
-\item \emph{experiment 1.a (simulation)}: during this experiment, participants have to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simscene}.
+\item \emph{experiment 1.a (simulation)}: during this experiment, participants have to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}.
 Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
@@
 \begin{tabular}{c c}
-Termes                         & Acronyms              \\
+Terms                         & Acronyms              \\
 \hline
-Niveau                         & $L$                    \\
-Niveau (événements)            & $L(E)$                 \\
-Niveau (textures)              & $L(T)$                 \\
-Agrément moyen (par scène)     & $\mathcal{A}_{scene}$  \\
-Agrément moyen (par sujet)     & $\mathcal{A}_{sujet}$  \\
+Level                        & $L$                    \\
+Level (events)            & $L(E)$                 \\
+Level (textures)              & $L(T)$                 \\
+Average pleasantness (per scene)     & $\mathcal{A}_{scene}$  \\
+Average pleasantness (per subject)     & $\mathcal{A}_{sujet}$  \\
                                &                        \\
-idéale/agréable                & i                      \\
-non-idéale/désagréable         & ni                     \\
-scène idéale/agréable          & i-scene                \\
-scène non-idéale/désagréable   & ni-scene               \\
+ideal/pleasant                & i                      \\
+non-ideal/unpleasant         & ni                     \\
+scene ideal/pleasant          & i-scene                \\
+scene non-ideal/unpleasant   & ni-scene               \\
 \hline
@@
 \vspace{0.5mm}
-\caption{Acronyme des variables utilisées dans le cadre des expériences sensorielles.}
+\caption{Acronyms of features used in the analysis of the sensory experiments.}
 \label{tab:acronyme}
@@
 
-Finally, it appears that the $p@5$ obtained with $E_m$ is equal – or above – to the ones obtained with $E$ or $ET$, although a partial information is used in that case to describe the scenes.
+Lastly, it appears that the $p@5$ obtained with $E_m$ is equal – or above – to the ones obtained with $E$ or $ET$, although a partial information is used in that case to describe the scenes.
 The dimension of the vectors $S$ for $E_m$ is lower than the dimension of vector $S$ for $E$, itself lower than the dimension of vector S obtained when all the classes are considered ($ET$).
@@
 \begin{enumerate}
-\item contrairement à ce que nous avions constaté avec les descripteurs structurels, une description sémantique de la composition des scènes, en terme de présence/absence de sources sonores, permet de bien distinguer les deux types d'environnement (i ou ni);
-\item l'information sémantique est majoritairement portée par les classes d'événements sonores;
-\item parmi les classes d'événements, seule une partie, \ie~les marqueurs sonores, est nécessaire pour faire la distinction entre les i- et ni-scènes.
+\item unlike what we outlined with the structural features, a semantic description of the scenes constitution in terms of presence / absence of sound sources, allows to differentiate the two types of environments (i or ni);
+\item the semantic information is mainly contained in the event sound classes;
+\item only a part of the event classes, i.e. the sound markers, are useful to differentiate the i- and ni- scenes.
 \end{enumerate}
@@
 \section{Experiment 2: modification of the semantic content}
-\label{sec:xp3}
+\label{sec:modification}
 
@@
 
-\section{Outcomes for sounscape perception}
-
+\section{Outcomes for soundscape perception}
+\label{sec:conclusion}
 %Les expériences ont montré que la majorité des descripteurs utilisés, qu'ils soient sémantiques ou structurels, permettent de faire la distinction entre une scène idéale et une scène non-idéale.
@@
 \appendix
-\section{Taxonomies des classes de sons}
+\section{Taxonomy of sound classes}
 \label{app:taxonomie}
@@
 \newpage
-\section{Dispersions des notes données par les sujets lors de l'expérience 2}
+\section{Distribution of scores given par the subjects during experiment 2}
 \label{app:xp2}

### 2017-07-06 lafaySimulatedScenePerception.tex
translation of figure

@@
         \myfloatalign
-        \includegraphics[width=.8\linewidth]{gfx/ch_4/3}
+        \includegraphics[width=.8\linewidth]{gfx/ch_4/3.png}
        \caption{Hierachical organisation of the isolated sounds used ofr the simulation.}\label{fig:orgDb}
@@
         \myfloatalign
-        \includegraphics[width=\linewidth]{gfx/ch_5/5}
+        \includegraphics[width=\linewidth]{gfx/ch_5/5.png}
         \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b)}\label{fig:xp1_2}
@@
 
-\begin{figure}[t]
+\begin{figure*}[t]
         \myfloatalign
@@
        \caption{Distributions of the structural features $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
-\end{figure}
+\end{figure*}
 
@@
         \subfloat[]
-        {\includegraphics[width=1\linewidth]{gfx/ch_5/xp1_class_1_cam}\label{fig:soundsourcea}} \par
-        \subfloat[]
-        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_2_cam}\label{fig:soundsourceb}}
-        \subfloat[]
-        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_3_cam}\label{fig:soundsourcec}}
+        {\includegraphics[width=1\linewidth]{gfx/ch_5/xp1_class_1}\label{fig:soundsourcea}} \par
+        \subfloat[]
+        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_2}\label{fig:soundsourceb}}
+        \subfloat[]
+        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_3}\label{fig:soundsourcec}}
        \caption{Proportion of synthesized scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.}\label{fig:soundsource}
@@
 
-\begin{table}[t]
+\begin{table*}[t]
  \setlength{\tabcolsep}{0.2pt}
@@
 \label{tab:markers}
-\end{table}
+\end{table*}
 
@@
         \myfloatalign
-        \includegraphics[width=\linewidth]{gfx/ch_5/pa5_1}
-       \caption{$P@5$ obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors.
+        \includegraphics[width=\linewidth]{gfx/ch_5/pa5_1_en}
+       \caption{$P@5$ obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors as a function of the abstraction level.
 The vectors are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers $E_{w/o,m}$.}\label{fig:pa5}
@@
         \caption{Distribution of the relative sound levels structural features related to the presence of markers $L_m$ (a, d), $L(E)_m$ (b, e) and $L(T)_m$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
-\end{figure}
+\end{figure*}
 
@@
         {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_2}\label{fig:taxonomieTextureb}}
-       \caption{Taxonomies des classes de sons utilisées pour la simulation des environnements sonores urbains : (a,b) événements sonores et (c,d) textures sonores.}
+       \caption{Taxonomy of sound classes used for the simulation urban soundscapes: events (a,b) and textures (c,d).}
        \label{fig:taxonomie}
@@
         \myfloatalign
-        \subfloat[sujet 1]
+        \subfloat[subject 1]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_1}\label{fig:xp4_note_1a}}
-        \subfloat[sujet 2]
+        \subfloat[subject 2]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_2}\label{fig:xp4_note_1b}}
-        \subfloat[sujet 3]
+        \subfloat[subject 3]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_3}\label{fig:xp4_note_1c}}
-        \subfloat[sujet 4]
+        \subfloat[subject 4]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_4}\label{fig:xp4_note_1d}} \par
-        \subfloat[sujet 5]
+        \subfloat[subject 5]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_5}\label{fig:xp4_note_1e}}
-        \subfloat[sujet 6]
+        \subfloat[subject 6]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_6}\label{fig:xp4_note_1f}}
-        \subfloat[sujet 7]
+        \subfloat[subject 7]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_7}\label{fig:xp4_note_1g}}
-        \subfloat[sujet 8]
+        \subfloat[subject 8]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_8}\label{}} \par
-        \subfloat[sujet 9]
+        \subfloat[subject 9]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_9}\label{fig:xp4_note_1i}}
-        \subfloat[sujet 10]
+        \subfloat[subject 10]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_10}\label{fig:xp4_note_1j}}
-        \subfloat[sujet 11]
+        \subfloat[subject 11]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_11}\label{fig:xp4_note_1z}}
-        \subfloat[sujet 12]
+        \subfloat[subject 12]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_12}\label{fig:xp4_note_1l}}
-        \caption{Dispersions des notes données par les sujets lors de l'expérience 2 aux i/am-scènes (vert) et ni/am-scènes (rouge).}\label{fig:xp4_note_1}
+        \caption{Distribution of scores given par the subjects during experiment 2 for i/am-scenes (green) et ni/am-scenes (red).}\label{fig:xp4_note_1}
 \end{figure}

### 2017-07-06 lafaySimulatedScenePerception.tex
ml proof read 1

@@
 \Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process.
-Subjects are asked to create a full soundscape by means of a software tool dedicated to sound edition, coupled with a structured sound data set.
+Subjects are asked to create a full soundscape by means of a dedicated software tool, coupled with a structured sound data set.
 This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes.
@@
 The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}.
-That being said, as those stimuli can be recorded and analysed, it allow the experimenter to gather a coarse grain physical description of the stimuli.
+That being said, those stimuli respectively are or can be recorded and analysed.
+It thus allows the experimenter to gather a coarse grain physical description of the stimuli.
 
@@
 A large amount of quantitative and semantic information is then collected about the representation of this type of sound environment that the subject has.
-Unfortunately, without any reference to sound data, this representation cannot be characterized physically.
+Unfortunately, without any reference to sound data, this representation can hardly be characterized physically.
 
@@
 
-For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to very interesting outcomes as it allows the experimenter to separately study  the influence of the sounds sources that compose a sound environment of interest.
-Indeed, with such a material, not only the type of sound sources is available for study, but also the exact level and audio waveform for each source and also the structural characteristics of the scene, that is the temporal distribution of the events.
+For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sounds sources that compose a sound environment of interest.
+With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the events.
 
@@
 To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmental scenes.
-Results and outcomes of a series of three experiments that build on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:
+Results and outcomes of a series of three experiments that foster on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:
 
@@
 The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics.
-A lack of diversity is nonetheless mentioned by the authors that limit the outcomes of the study.
+A lack of diversity is nonetheless mentioned by the authors that limit the strenght of the outcomes given in this study.
 
@@
 
-In our approach, the simulator developed for this study\footnote{simScene, available online \url{http://soundthings.org/research/}} gives a monophonic representation of the scene, with the added benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are, as much as possible, expressive and useful for analysis.
+In our approach, the simulator developed for this study\footnote{simScene, available online \url{https://bitbucket.org/mlagrange/simscene}} gives a monophonic representation of the scene, with the added benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are both expressive and useful for analysis.
 
@@
 To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}.
-A track is then created for this sounds source within the simulator interface.
+A track is then created for this sound source within the simulator interface.
 The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sounds source.
@@
 
-The chosen typology is established based on the category/classes of sounds found while reviewing 16 articles or thesis manuscripts \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} that study how humans discriminate different kind of urban soundscapes.
+The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} that study how humans discriminate different kind of urban soundscapes.
 
@@
 
-Literature usually distinguish between:
+The literature usually distinguish between:
 \begin{itemize}
@@
 
-The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures.
-These two types have specific simulation procedures.
+The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures only.
+These two types also have specific scheduling procedures during the sumulation process.
 
@@
 
-The event sequences are composed of salient and identifiable events such as \emph{car start}, \emph{male speech}.
+The event sequences are composed of salient events that can easily be recognized such as \emph{car start} or \emph{male speech}.
 They arise from a descriptive analysis based on the identification of the sound sources.
-On contrary, the amorphous sequences are environments where distinct events can hardly be found, such as \emph{traffic hubhub}.
-They build upon an holistic analysis based on global acoustic features.
+On contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}.
+They build upon a holistic analysis based on global acoustic features.
 
@@
 Rosch \cite{rosch1978cognition}.
-The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources.
-For classes with high level of abstraction with sub classes, its collection of samples is the union of the collections of the sub classes.
+The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources is.
+For classes with high level of abstraction that have sub classes, its collection of samples is the union of the collections of the sub classes.
 
@@
 Accounting for the perceptual matters detailed previously, two taxonomies are built, one for event sounds and one for texture sounds.
-Four levels of abstraction are considered from the more generic one (level 0), to the more specific classe (level 3), leading to a taxonomy close to the one used in \cite{Salamon14}.
+Four levels of abstraction are considered from the more generic one (level 0), to the more specific classes (level 3), leading to a taxonomy close to the one used in \cite{Salamon14}.
 Only three levels of abstraction are considered for the texture sounds.
@@
 The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background.
-It also allowed us to avoid dominant sound sources while recording texture sounds by aiming the microphone to distant areas with no dominant sound sources.
+It also allowed us to avoid dominant sound sources while recording texture sounds by targeting distant areas with no dominant sound sources.
 

### 2017-07-07 lafaySimulatedScenePerception.tex
ml proof read 2

@@
 
-By design, the simulation tool do not allow the user to interact and control a specific sample.
+By a volountary design choice, the simulation tool do not allow the user to interact and control a specific sample.
 Interaction is done at the track level, a track being a sequence of samples.
@@
 
-To ease the simulation, two parameters are also proposed:
+To improve simulation quality, two parameters are also proposed:
 
@@
 
-Once the typology and the set sound are available, an important design question that remains is to find a suitable way of presenting the data to the user.
-Most of browsing tools are based on keyword indexing.
-However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject.
-Indeed, the subject may select \textit{a priori} sound referenced belonging to \emph{park} to build a \emph{calm} soundscape.
+Once the typology and the set of sounds are available, an important design issue os the need of a suitable way to display the sound dataset to the user.
+Most browsing tools are based on keyword indexing.
+However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potenitally induce biases in the analysis conclusions.
+Indeed, the subject may select \textit{a priori} sounds referenced as belonging to \emph{park} to build a \emph{calm} soundscape.
 
@@
 Each of those classes have a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle.
-The subject can browse the database by listening to those prototype sound.
+The subject can browse the database by listening to those prototype sounds.
 The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed in \cite{lafay2016JAES}.
@@
 
-Experiment 1 aims at using the simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness.
-For that, two first experiments are planned (\cf~Figure~\ref{fig:xp1_2}):
-\begin{itemize}
-\item \emph{experiment 1.a (simulation)}: during this experiment, participants have to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}.
+Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness.
+For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
+\begin{itemize}
+\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}.
 Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
-\item \emph{experiment 2.a (evaluation)}: actually, after the simulation phase, we only have a binary information on the emotional properties of the respective scenes: ideal (i) and non ideal (ni).
-This second experimental step aims at investigating our knowledge on pleasantness.
+\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes: respectively ideal (i) or non ideal (ni).
+Furthermore, this information is given by the creator of the scene.
+The second experimental step aims at investigating more deeply and more broadly our knowledge on pleasantness.
 For that, a second group of participants are asked to evaluate the pleasantness of each simulated scene, on a semantic scale.
-This evaluation protocol has two goals:
+This experiment has two goals:
 \begin{enumerate}
-\item to evaluate the respective influence on the pleasantness of the various sources composing the scenes (i or ni)
-\item to detect the presence of \emph{outliers} or ambiguous samples.
-For the remainder of the study, the defined hedonic properties (i and ni) are used as ground truth.
-Then, we must have to be sure that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of pleasantness rating.
+\item evaluate the respective influence on the pleasantness of the various sources composing the scenes (i or ni)
+\item detect the presence of \emph{outliers} or ambiguous samples.
+For the remainder of the study, the defined hedonic properties (i and ni) are used as references.
+We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of percieved pleasantness.
 \end{enumerate}
@@
 
-Our analysis is based on the data collected by these two experiments (1.a, 1.b)
+Our analysis is based on the data collected by these two experiments: 1.a and 1.b.
 
@@
 
-The participants have to simulate two urban sound environments of 1 minute each. They have to follow the instructions below:
-\begin{itemize}
-\item  1st simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
-\item 2nd simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
-\end{itemize}
-
-All the participants start by working on the ideal environment; they read the second set of instruction at the end of the first experiment.
+The participants have to simulate two urban sound environments of one minute each. They follow the instructions below:
+\begin{itemize}
+\item  first simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
+\item second simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
+\end{itemize}
+
+All the participants start by desiging the ideal environment; they read the second set of instruction at the end of the first experiment.
 Participants are completely free of their choice concerning sounds and synthesis parameters.
@@
 \end{itemize}
-For instance, the participant can insert ten dogs in the soundscape but he/she cannot insert one dog barking every 10 milliseconds.
+For instance, the participant can insert ten dogs in the soundscape but she cannot insert one dog barking every 10 milliseconds.
 These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation user interface.
@@
 
-Each simulation process contains two parts :
+Each simulation process contains two parts:
 \begin{enumerate}
-\item composition of the simulation: this step is split in three steps
+\item composition of the simulation: this step is split in three steps:
 \begin{itemize}
@@
 
-In addition, once the two sound scenes have been realized, participants are invited to
+In addition, once the two sound scenes are done, participants are invited to
 \begin{itemize}
@@
 
-All the participants do the experiment on similar computers.
+All the participants do the experiment on desktop computers with the same hardware and software configurations.
 The audio files are played in diotic conditions by headphones.
-During the tutorial, participants are asked to adjust the sound level in order to be comfortable; they can’t modify it in the following of the experiment.
-All the participants do the experiment simultaneously. They are equally split in three identical quiet rooms. Theyare not allowed to talk to each other during the experiment.
-Three experimenters (one in each room) are present during the whole duration of the experiment in order to check that everything goes well and possibly to answer questions.
+During the tutorial, participants are asked to adjust the sound level in order to be comfortable.
+Once set, they are not allowed to modify it in the remaining of the experiment.
+All the participants do the experiment at the same time. They are equally split in three identical quiet rooms. They are not allowed to talk to each other during the experiment.
+Three experimenters (one in each room) are present during the whole duration of the experiment in order to assist participants with potential hardware and software issues and to answer questions.
 
@@
 
-44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in Nantes downtown since 2 years or more.
-Among the 44 participants, 40 succeeded in doing the experiment, producing in the end, 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
-4 participants were excluded from the process for not having respect or understand the instructions, or for exceeding the maximum duration of the experiment.
+44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in downtown Nantes since 2 years or more.
+Among the 44 participants, 40 succeeded, producing in the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
+the 4 other participants were excluded from the process due to a lack of respect or understanding of the instructions, or for exceeding the maximum duration of the experiment.
 
@@
 
-Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes composed in Exp. 1.a (from sec. 15 to sec. 45).
-The assessment is done in a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
-Before evaluating a scene, participants must listen to the first 20 seconds of the scene.
+Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes composed in Experiment 1.a (from sec. 15 to sec. 45).
+The assessment is done with a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
+Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli.
 After the evaluation, they are free to switch to the next scene.
 For each participant, sound scenes are played in a random order.
-The first 10 scenes allow the participants to calibrate their marks; they imperatively gather 5 ideal scenes and 5 non ideal scenes.
-These first 10 scenes are played back again at the end of the experiment and the corresponding evaluations are the ones that are only taken into account in the results.
+The first 10 scenes allow the participants to calibrate their scores; as they are always composed of 5 ideal scenes and 5 non ideal scenes.
+These first 10 scenes are played back again at the end of the experiment and the last evaluations are the ones that are only taken into account in the results.
 The experiment is planned to last 30 minutes. The participants do not know anything beforehand about the nature of the sound scenes.
@@
 
-All the participants do the experiment on similar computers.
+All the participants do the experiment on computers with equal hardware and software configurations.
 The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}.
@@
 All the participants do the experiment simultaneously. They are not allowed to talk to each other during the experiment.
-An experimenter is present during the whole duration of the experiment in order to check that everything goes well and possibly to answer questions.
+An experimenter is present during the whole duration of the experiment in order to assist participants and possibly answer questions.
 
@@
 10 students (2 females - mean age 23.1, SD = 1.8) from Ecole Centrale de Nantes took part to the experiment.
-They are all living in Nantes since 2 years or more.
+All are living in downtown Nantes for 2 years or more.
 None of them took part to the previous simulation experiment (Exp.

### 2017-07-07 lafaySimulatedScenePerception.tex
ml proof read 2

@@
 A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}.
-In order to stay coherent with the evaluation experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version, see Section \ref{sec:xp1b_plan}.
+In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as timuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.
 For each sound scene, three types of features are considered :
@@
 \item \emph{perceptual}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale.
-$\mathcal{A}_{scene}$ is the mean pleasantness for one scene, as the average of the marks of all the participants.
-In the same idea, $\mathcal{A}_{subject}$ is the mean pleasantness for each participant, as the average of all the marks of each participant.
+$\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants.
+$\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant.
 Given the low number of participants in Exp.
-1.b, we decide not to normalize the values of pleasantness marks.
-\item \emph{semantic}: boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene.
-Each point $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise.
-The vector dimension ($n$) depends on the level of abstraction that is considered.
-For instance, for the abstraction level 1, including 44 classes of sounds, $n = 44$.
-\item \emph{structural}: related to \emph{sound level}.
+1.b, we choose not to normalize the pleasantness scores.
+\item \emph{semantic}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene.
+Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise.
+The vector dimension ($n$) depends on the level of abstraction that is considered for analysis.
+For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.
+\item \emph{structural}: features related to \emph{sound level}.
 To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference.
 In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt.
-The level is obtained by computing every second the quadratic mean of the signal and by averaging over the total duration of the scene.
+The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene.
 An A-filtering module processes the data before the quadratic means are computed.
@@
 In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following tests of significativity:
+
+///
 

### 2017-07-07 lafaySimulatedScenePerception.tex
ml proof read 3

@@
 
-///
-
-\begin{itemize}
-\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with regards to the two types of scene.
-the significativity is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$ ;
-\item \emph{analysis of the structural features}: in order to evaluate whether the emotional difference forced to the i- and ni- scenes has a significative influence \ie, if significative differences exist between structural features and perceived pleasantness, we measure this significativity with a Student test on two populations;
-\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, we look at linear correlations between these features and $\mathcal{A}_{scene}$.
-For that, we use the Pearson correlation coefficient, on the basis of the standard methodology used in multidimensional studies;
-\item \emph{analysis of the semantic features}: in order to determine if the forced emotional difference has an influence on the scenes composition in terms of sound sources, or if specific  classes are more frequently used to simulate a given environment (i or ni), we apply a V-test.
-This test is done at every level of abstraction, and separately, for texture and event classes.
-For each class $j$ , and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:
+\begin{itemize}
+\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with respect to the two types of scenes.
+The significativity is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
+\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni- scenes has a significative influence  on the nature of the scenes \ie if significative differences exist between structural features and perceived pleasantness.
+The significativity is measured with a Student test on two populations;
+\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, linear correlations between these features and $\mathcal{A}_{scene}$ are considered.
+The Pearson correlation coefficient is used for that purpose, on the basis of the standard methodology used in multidimensional studies;
+\item \emph{analysis of the semantic features}: in order to determine if the affective difference between  the i- and ni- scenes has an influence on the scenes composition in terms of sound sources, or if specific classes are more frequently used to simulate a given environment (i or ni), a V-test is considered.
+Testing is done at every level of abstraction, and separately for texture and event classes.
+For each class $j$, and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:
 
@@
 
-\item \emph{Study of the representation spaces induced by the semantic features}: in order to determine if a representation only based on the presence / absence of sound sources allows to split the two types of environment, we investigate the space induced by the semantic features $S$.
+\item \emph{Study of the representation spaces induced by the semantic features}: in order to determine if a representation only based on the presence / absence of sound sources allows us to distinguish between the two types of environment, we investigate the space induced by the semantic features $S$.
 $S$ is a boolean vector and the distance between two scenes is computed with the Hamming distance.
-Considering two n-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, where $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ is of cardinality $n$, with $x=\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between two vectors, as follows:
+Considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, where $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ is of cardinality $n$, with $x=\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between two vectors, ans is defined as follows:
 
@@
 
-where $\bigoplus$ is the \emph{exclusive-or} The more the two scenes are similar, the closer they are.
-The use of Hamming distance allows to take into account equally the present and absent classes.
-For measuring the intrinsic ability of the space to split the i- and ni- scene, we use a ranking metric named $k$- precision ($P@k$) .
-The $P@k$ computes the precision obtained after $k$ items have been found.
-Formally, for each $s_i$ scene, we compute the ratio between the number of $s_j$ scenes taken among the k nearest neighbors of $s_i$ and sharing the same label than $s_i$ with regards to the $k$ items to be found.
-the $P@k$ is then the mean of this ratio along all the items;
-
-\item \emph{influence of the sound markers on the perceived pleasantness}: in order to assess the specific contributions of some sound sources, we again estimate the possible impact of the structural features on the perceived pleasantness, but this time, by only taking into account the sound markers for the computation of these features.
-\end{itemize}
-
-Except for the V-test, all the statistical test of significativity are done with a critical threshold of $\alpha=0.05$.
+where $\bigoplus$ is the \emph{exclusive-or}.
+The more the comosition of the two scenes are similar, the closer they are in this space.
+The use of Hamming distance allows us to take into account equally the presence and absence of classes.
+For measuring the intrinsic ability of the space to split the i- and ni- scene, we use a ranking metric named the precision at rank $k$ ($P@k$) .
+The $P@k$ computes the precision obtained after the $k$ closest items of a given seed item have been found.
+Formally, for each $s_i$ scene (considered as seed), we compute the ratio between the number of $s_j$ scenes taken among the $k$ nearest neighbors of $s_i$ and sharing the same label than $s_i$ versus $k$.
+the $P@k$ is then the average of this ratio for all the items considered as search seeds;
+
+\item \emph{influence of the sound markers on the perceived pleasantness}: in order to assess the specific contributions of some sound sources, we again estimate the impact of the structural features on the perceived pleasantness by taking into account only the sound markers for the computation of those features.
+\end{itemize}
+
+Except for the V-test, the statistical test of significativity are done with a critical threshold of $\alpha=0.05$.
 For the V-test, considering that a great number of classes is tested, a Bonferroni correction is applied.
@@
 
+///
+
 \subsection{Results}
 
-\subsubsection*{Cheking of the pleasantness of the synthesized scenes}
+\subsubsection*{Checking of the pleasantness of the synthesized scenes}
 

### 2017-07-17 lafaySimulatedScenePerception.tex
ml proof read 4

@@
 
-///
-
 \subsection{Results}
@@
 First, in order to ensure the coherence of the data, we check that none of the ni- scenes gets a $\mathcal{A}_{scene}$ higher than one of a i- scene.
-Four ni- scenes don’t respect that rule: they are removed, together with their corresponding i- scenes.
-36 i -scenes and 36 ni -scenes remain for analysis.
-Secondly, we want to be sure that participants really perceived a difference in terms of pleasantness between i- and ni -scenes.
+Four ni- scenes don’t fullfill that constraint: they are thus removed, together with their corresponding i- scenes.
+As a consequence, 36 i -scenes and 36 ni -scenes remain for analysis.
+Second, we verify that participants really perceived a difference in terms of pleasantness between i- and ni -scenes.
 For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment.
@@
 
-First, we focus on the sound level.
-Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} depict the levels $L$, $L(E)$ and $L(T)$.
+First, our analyses focuses on the sound level.
+Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the levels $L$, $L(E)$ and $L(T)$.
 There is a significative difference in terms of sound level between i- and ni -scenes ($L$: $p<0.01$), with a standard deviation of -7 $dB$.
-This difference is siginificant for events ($L(E)$: $p<0.01$, STD: -7 $dB$) and textures ($L(T)$: $p<0.01$, STD: -6 $dB$).
-
-As expected, we verify that the sound sources sound level is indeed a pleasantness indicator, the ni -scenes tending to be louder - what also comes out from a large number of related works.
+This difference is siginificant for events ($L(E)$: $p<0.01$, STD: -7 $dB$) and for textures ($L(T)$: $p<0.01$, STD: -6 $dB$).
+
+As expected, the sound sources sound level is indeed a pleasantness indicator, as the ni -scenes tend to be louder, an outcome of a large number of related works.
 We also notice that this difference of sound level is equally true for events or textures.
@@
 This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}.
-During their experiment, the authors ask to their participants to assess a set of soundscapes at a global level and then to do the same judgment at the moments when each of them detect a sound source.
+During their experiment, the authors ask to their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source.
 The study shows that there is no significative difference between global and averaged instantaneous judgments.
@@
 
-The level alone is not enough to differentiate the different types of environment.
-In fact, $20\%$ of the i- scenes hast got a sound level higher than the nominal level of the ni- scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
+Though, the level alone is not sufficient to differentiate the type of environment.
+In fact, $20\%$ of the i- scenes have a sound level higher than the nominal level of the ni- scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
 
@@
 In this section, relationships that could exist between structural features and perceived pleasantness are investigated.
-Contrary to the previous section, we do not consider the emotional property of a scene on a binary mode (i vs.
-ni): here, we consider the mean pleasantness $\mathcal{A}_{scene}$ as a perceptual feature.
+Contrary to the previous section, we do not consider the pleasantness property of a scene on a binary mode (i vs.
+ni): this is the mean pleasantness $\mathcal{A}_{scene}$ that is considered here as a perceptual feature.
 The aim is to investigate the level of correlation between structural features and $\mathcal{A}_{scene}$.
@@
 
-When considering the whole set of scenes, this result can be explained by the fact the i- scenes tends to be softer than the ni- scenes, deluding that the negative correlation observed for the ni- scenes is extended.
+When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the i- scenes tend to be softer than the ni- scenes, thus allowing us to extend to the i-scenes the negative correlation observed for the ni- scenes.
 
@@
 \begin{itemize}
-\item allows to differentiate between i- and ni- scenes,
-\item allows a fine characterization of the perceived pleasantness for ni- scenes,
-\item is not a relevant feature for the perceived pleasantness for a priori pleasant soundscapes (i- scene)
-\end{itemize}
-
-The same conclusions can be drawn for $L(E)$ (\cf~Figure~\ref{fig:soundlevele}).
-For $L(T)$ (\cf~Figure~\ref{fig:soundlevelf}), the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$).
-One can suppose that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer than ni - scenes).
+\item allows us to differentiate between i- and ni- scenes,
+\item characterizes precisely the perceived pleasantness for ni- scenes,
+\item is not a relevant feature for modeling the perceived pleasantness of \textit{a priori} pleasant soundscapes (i- scenes)
+\end{itemize}
+
+The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}.
+For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$).
+Again, we believe that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer than ni - scenes).
 Thus, the sound events keep a relative ability to predict the pleasantness of the ni- scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.
@@
 
-Those first outcomes tend to show that two modes of perception exists depending on the nature of the environment (i- or ni-), and each involve independent features.
-The fact that $L$ is not able to characterize the pleasantness of the i- scenes can lead us to think that all the sound sources do not equally contribute to the perception of pleasantness, but that only the level of some of them can influence this percept.
-In order to go deeper in that direction, in the next section, we analyze these scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.
+Those first outcomes tend to show that 1) two modes of perception exists depending on the very nature of the environment (i- or ni-), and 2) each involve independent features.
+The fact that $L$ is not sufficient to characterize the pleasantness of the i- scenes can lead us to conclude that all the sound sources do not equally contribute to the perception of pleasantness.
+Thus, we think that only the level of some of them can influence this percept.
+In order to investigate further in that direction, we analyze in the next section the scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.
 
@@
 
-We analyse the composition of the scenes by counting the number of participants who used a given class of sounds to simulate the environment.
+We study the composition of the scenes by counting the number of participants who used a given class of sounds to simulate the environment.
 Results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
-For sake of clarity, we chose a transitional level of abstraction between level 0 and 1, named $0+$, to depict classes, see ~Figure~\ref{fig:taxonomie}.
+For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$ is used to depict classes, see ~Figure~\ref{fig:taxonomie}.
 
 We observe a significant difference between i- and ni- scenes for the classes choice.
-The distribution of the classes is very similar to to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~lasses involving human presence and nature are prevailing in the i- scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni- scenes.
+The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~classes involving human presence and nature are prevailing in the i- scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni- scenes.
 These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.
@@
 The authors interpret this by the fact that the perception of pleasantness is, among other things, due to socio-cultural factors.
-In our representation of the world, sounds of public transportation are positively connoted and tend to be more accepted than sounds of personal vehicles.
+Thus, in our representation of the world, sounds of public transportation are positively connoted and tend to be more accepted than sounds of personal vehicles.
 
 To a certain extent, our results are in contradiction with this assumption.
-In fact,  Figure~\ref{fig:soundsourcea} show \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the participants for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes.
+In fact,  Figure~\ref{fig:soundsourcea} show that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the participants for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes.
 the results do not re-evaluate the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} pour les i-scenes, a level that is comparable to the of the \emph{bike} class, and much higher than all the personal vehicles classes.
@@
 
-This difference can be explained by the nature of the experimental protocol used in the two studies.
-As in our case, Guastavino asks the participants for describing an environment; but in that case, they work from their memories whereas, in our case, participants do that task using sound samples.
+This difference may be explained by the nature of the experimental protocol used in the two studies.
+As in our study, Guastavino asks the participants for describing an environment starting from their memories whereas, in our case, participants perform the same task using actual sound samples.
 The fact that participants in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact.
@@
 
-We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs regarding the type of environment (i or ni).
-We are now trying to to investigate if some of the sound classes are specific to a given environment.
-For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna}, while considering separately each abstraction level.
+We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs with respect to the type of environment (i or ni).
+Let us now investigate whether some of the sound classes are specific to a given environment.
+For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna} separately for each abstraction level.
 Results are presented in Table~\ref{tab:markers}.
 
-Regarding the sound events, 9 markers are identified at all the abstraction levels.
-As Figure~\ref{fig:soundsource} could let it predict, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i- scenes markers; we also notice the \emph{church bell} class as an i- scenes marker.
+Regarding the sound events, 9 markers are identified for all abstraction levels.
+As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i- scenes markers as well as the \emph{church bell} class.
 This latter result may be due to the socio-cultural background of the participants who are mostly European citizens.
@@
 For the i- scenes, they are classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}).
-for the ni- scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).
-
-Although the whole set of identified markers are intuitive, none of the event classes related to the noise of the motor vehicles is a marker, except for the texture class \emph{crossroad}.
+For the ni- scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).
+
+Although the whole set of identified markers are rather intuitive, none of the event classes related to the noise of the motor vehicles are identified as markers, except for the texture class \emph{crossroad}.
 To simulate an unpleasant traffic, participants chose the classes \emph{horn} or \emph{siren}.
-One can suppose that isolated motor vehicle sounds are understood as taking fully part of the urban environment, and thus are not especially linked to an unpleasant soundscape.
+We thus conclude that isolated motor vehicle sounds are understood as being part of the urban environment, and thus their nature is not necessarily linked to an unpleasant soundscape.
 
@@
 
-In this section, we evaluate the ability of a semantic representation to split the two types of environment.
-For that, we compute a rank 5-precision ($p@5$) on the space induced by the semantic features $S$, and for each abstraction level, see Section~\ref{sec:xp1_dataAna}.
+In this section, we evaluate the ability of a semantic representation of the scenes to discriminate between the two types of environment.
+For that, we consider a rank 5-precision ($p@5$) on the space induced by the semantic features $S$, and for each abstraction level, see Section~\ref{sec:xp1_dataAna}.
 The vectors $S$ are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers ($E_{w/o,m}$).
-Texture classes corresponding to sound markers are not considered as these latter are not numerous enough.
+Texture classes corresponding to sound markers are not considered as these latter are too few to reliably compute the metric.
 For the same reasons, event classes corresponding to sound markers at abstraction level $0$ are also discarded.
@@
 Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and stay above $86\%$ from the abstraction level $1$.
-These results confirm that it is possible to make a clear distinction between the two types of environment on the unique basis of presence / absence of sound classes.
-We also notice that the higher the abstraction level is, the more important is the ability to make this distinction.
-In other words, the more one is precise in the scene description, the more one is able to clearly differentiate between i- and ni- scenes.
-
-Considering $E$ and $T$ separately, ir appears: 1) that the $p@5$ obtained with $E$ is similar to the one obtained with $ET$ ; 2) that the $p@5$ obtained with $T$ is systematically lower than the one obtained with $E$, by $10\%$ or $15\%$.
-Those results indicate that the semantic information allowing to split the two environments is mainly contained in the events.
-Moreover, these results are in line with the work done in \cite{maffiolo_caracterisation_1999} showing that we analyze in a causal manner (i.e.
-by identifying the sources) the event scenes which are composed of several sound events, see Section~\ref{sec:simscene_sampleDataSet}.
+These results confirm that it is possible to make a clear distinction between the two types of environment on the sole basis of presence / absence of sound classes.
+We also notice that the higher the abstraction level is, the more important the ability to make this distinction is.
+In other words, the more precise in the scene description is, the more clear the discrimination between i- and ni- scenes is.
+
+Considering $E$ and $T$ separately, it appears that: 1) the $p@5$ obtained with $E$ is similar to the one obtained with $ET$ ; 2) the $p@5$ obtained with $T$ is systematically lower than the one obtained with $E$, by $10\%$ or $15\%$.
+Those results indicate that the discriminative semantic information is mainly contained in the events.
+These results are in line with the work done in \cite{maffiolo_caracterisation_1999} showing that humans analyze the event scenes which are composed of several sound events in a causal manner, \ie by identifying the sources, as discussed in Section~\ref{sec:simscene_sampleDataSet}.
 

### 2017-07-17 lafaySimulatedScenePerception.tex
ml proof read 5

@@
 
-Lastly, it appears that the $p@5$ obtained with $E_m$ is equal – or above – to the ones obtained with $E$ or $ET$, although a partial information is used in that case to describe the scenes.
-The dimension of the vectors $S$ for $E_m$ is lower than the dimension of vector $S$ for $E$, itself lower than the dimension of vector S obtained when all the classes are considered ($ET$).
-Moreover, in the case when the sound markers are not taken into account for the description ($E_{w/o,m}$), the results fall even going under those obtained with only the textures.
+The dimension of the vectors $S$ for $E_m$ is lower than the dimension of vector $S$ for $E$, itself lower than the dimension of vector $S$ obtained when all the classes are considered ($ET$).
+Despite this, it appears that the $p@5$ obtained with $E_m$ is equal – or above – to the ones obtained with $E$ or $ET$, although a partial information is used in that case to describe the scenes.
+Moreover, in the case when the sound markers are not taken into account for the description ($E_{w/o,m}$), the performance is under those obtained with only the textures as features.
 This confirms that most of the semantic information allowing to differentiate between i- and ni- scenes is included in the markers.
@@
 \begin{enumerate}
-\item unlike what we outlined with the structural features, a semantic description of the scenes constitution in terms of presence / absence of sound sources, allows to differentiate the two types of environments (i or ni);
+\item unlike what we outlined with the structural features, a semantic description of the scenes constitution in terms of presence / absence of sound sources, allows us to differentiate the two types of environments (i or ni);
 \item the semantic information is mainly contained in the event sound classes;
-\item only a part of the event classes, i.e. the sound markers, are useful to differentiate the i- and ni- scenes.
+\item only a part of the event classes, i.e. the sound markers, are useful to differentiate between the i- and ni- scenes.
 \end{enumerate}
 
-Since we have extracted the typical classes of the i- and ni- scenes and verified that the distinction between them was dependent to the presence of these classes, we have now to investigate whether a structural description of the scenes only based on sound markers could allow us to characterize the perceived pleasantness, perhaps better than a global structural description.
+Since we have extracted the typical classes of the i- and ni- scenes and verified that the distinction between them was largely dependent on the presence of these classes, we shall now investigate whether a structural description of the scenes only based on sound markers could characterize the perceived pleasantness, perhaps better than a global structural description.
 
@@
 We evaluate the correlations between $\mathcal{A}_{scene}$ and the structural features.
-In this section, the structural features are computed by taking into account of the sound markers previously identified.
+In this section, the structural features are computed by taking into account only the previously identified sound markers.
 We define $Xm$, the $X$ feature computed by taking into account only the sound markers.
 On the opposite, we define $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
-When the feature characterize an i- scene (idem for a ni- scene), we consider for the computation only the markers identified for the i- scenes (or the ni- scenes); we named them, i- markers (or ni- markers).
+When the feature characterizes an i- scene (idem for a ni- scene),  only the markers identified for the i- scenes (or the ni- scenes) are considered; we named them, i- markers (or ni- markers).
 Results are shown in Table~\ref{tab:corrMarkers}.
 
-Concerning the sound levels (\cf~Figures~\ref{fig:soundlevelMarkera} et~\ref{fig:soundlevelMarkerd}), the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ et $L(T)$ on the other side.
-Whether all the classes or only the markers ae considered, it appears that:
+Concerning the sound levels, the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ et $L(T)$ on the other side, see~Figures~\ref{fig:soundlevelMarkera} and~\ref{fig:soundlevelMarkerd}, respectively.
+No matter whether all the classes or only the markers are considered, it appears that:
 
@@
 
-Considering the non markers classes (see~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), we can observe on the i- scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
+Considering the non markers classes, we can observe on the i- scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}.
 It is the first time that an objective feature allows us to define the pleasantness of the pleasant environments.
@@
 
-We consider a last group of features, namely $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiffa} et~\ref{fig:soundlevelMarkerDiffd}).
-These features describe the difference between the markers level and those of the other sound classes.
-They express the saliency of the markers with regards to the sound mixture.
+We consider a last group of features, namely $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$.
+These features describe the difference between the markers level and those of the other sound classes, see ~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}).
+They express the saliency of the markers with respect to the sound mixture.
 
@@
 
-For the ni -scenes, the fact that we observe significative correlations for Lm and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.
+For the ni -scenes, the fact that we observe significative correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.
 
@@
 
-From this analysis, we can outline the following points:
+From this analysis, the following points can be outlined:
 
@@
 The semantic description seems to be more powerful;
-\item \emph{events or textures}: whatever semantic or structural features, events are mainly useful to differentiate the two types of events; textures bring a limited amount of information;
-\item \emph{pleasantness prediction}: if we consider the correlations between structural features and pleasantness, it seems that the way to perceive the quality of a given environment depends on its nature (i or ni).
-It is not conceivable to consider a same set of features to predict the pleasantness of i- and ni- scenes in the same time:
+\item \emph{events or textures}: whatever the feature type, be it semantic or structural, events are mainly useful to differentiate the two types of events; textures bring a limited amount of information;
+\item \emph{pleasantness prediction}: considering the correlation between structural features and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its nature (i or ni).
+From the data gathred in those experiments, it is not conceivable to consider the same set of features to predict the pleasantness of i- and ni- scenes in the same time:
 
@@
 \item \emph{for ni-scenes}, global levels ($L$ et $L(E)$), or level of sound markers ($L_m$ et $L(E)_{m}$), have a negative influence on pleasantness.
-We can notice here that taking into account the contribution of each the different sources of the scene does not improve the prediction performance, wit regards to a global analysis of the environment.
-
-\item \emph{for i-scenes}, on the contrary, the pleasantness prediction needs to study separately the sound markers characteristics and those of the other sounds.
-Thus, the markers level, relative to the noise ($L(E)_m-L(E)_b$ et $L_m-L_b$) is positively correlated to the pleasantness, whereas the noise level ($L_b$ et $L(E)_b$)is negatively correlated.
+We can notice here that taking into account the contribution of each of the different sources of the scene does not improve the prediction performance compared to a global analysis of the environment.
+
+\item \emph{for i-scenes}, on contrary, the pleasantness prediction needs to consider separately the sound markers characteristics and those of the other sounds.
+Thus, the markers level, relative to the noise ($L(E)_m-L(E)_b$ and $L_m-L_b$) is positively correlated to the pleasantness, whereas the noise level ($L_b$ and $L(E)_b$) is negatively correlated.
 \end{itemize}
@@
 
-We thus observe two perceptual modes that involve different types of features, and that rely on the hedonic nature  of the stimuli.
+We thus observe two perceptual modes of operation that involve different types of features, and that rely on the hedonic nature  of the stimuli.
 A phenomenon similar is observed for the perception of textures, see Section~\ref{sec:simscene_sampleDataSet}.
-It seems that the brain adapts its way to encode the information (statistic summary for textures, fine description for events) following a previous decision making with regards to the nature of the stimuli, (()\ie is it an event or a texture ?).
-Similarly, features that are considered for the pleasantness judgment also depends on a preliminary judgment of the global hedonic nature of the environment (ideal or non ideal).
+It seems that the brain adapts its way to encode the information (statistic summary for textures, fine description for events) following a previous decision making with regards to the nature of the stimuli, \ie is it an event or a texture ?.
+Similarly, in the light of this study, it appears that the features considered for the pleasantness judgment also depends on a preliminary judgment of the global hedonic nature of the environment (ideal or non ideal).
 

### 2017-07-17 lafaySimulatedScenePerception.tex

@@
 
-The previous experiment demonstrated that, among the classes of sounds occurring in urban soundscapes, those with markers are specific to some of environments.
+The previous experiment demonstrated that, among the classes of sounds occurring in urban soundscapes, those with markers are specific to some environments.
 Those sound markers seem to have a great impact on perception.
@@
 
-In order to verify that the pleasantness of the ideal and non ideal scenes if affected by the sound markers, the audio waveform of the simulated scenes are regenerated with or without considering the classes identified as markers.
+In order to verify that the pleasantness of the ideal and non ideal scenes is affected by the sound markers, the audio waveform of the simulated scenes are regenerated with or without using the classes identified as markers.
 For the i-scenes, some of the i-markers are removed.
 For the ni-scenes, some of the ni-markers are removed.
-A perceptive test evaluating the pleasantness is then conducted with a protocol close to the one considered in experiment 1.b.
+To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptive test is then conducted with a protocol close to the one considered in experiment 1.b.
 
@@
 
-Thus the objective of this experiment is to study if the removal of the previously identified markers have an impact on the perceived pleasantness. Two hypothesis are thus formulated:
+The objective of this experiment is to study if the removal of the previously identified markers have an impact on the perceived pleasantness. Two hypothesis are thus formulated:
 
@@
 If the first hypothesis is rather intuitive, the second is less.
-Indeed, it is not obvious that the removal of the i-markers, though perceptively connoted, decreases the global quality of a soundscape.
-This removal also decreases the global level of the scene.
+Indeed, it is not obvious that the removal of the i-markers, though perceptively positively connoted, decreases the global quality of a soundscape.
+Indeed, this removal also decreases the global level of the scene.
 
@@
 \item \emph{72 km-scenes}: the 72 scenes originally simulated by the subject of experiment 1.a where the sound classes identified as markers are kept.
-The 36 ideal scenes with markers are noted i/km-scenes, and the 36 non ideal scenes with markers are noted ni/km-scenes.
+The 36 ideal scenes with markers are noted i/km-scenes, and the 36 non ideal scenes with markers are noted ni/km-scenes (k for kept).
 \item \emph{72 rm-scenes}:  72 scenes where the sound classes identified as markers are removed.
-The 36 ideal scenes without markers are noted i/rm-scenes, and the 36 non ideal scenes without markers are noted ni/rm-scenes.
+The 36 ideal scenes without markers are noted i/rm-scenes, and the 36 non ideal scenes without markers are noted ni/rm-scenes (r for removed).
 \end{itemize}
@@
 
-In order to create km-scenes with still some sound diversity and no long duration absence of sound activity within the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}.
+In order to create km-scenes with still some sound diversity and no  absence of sound activity within a long period of the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}.
 Those classes are:
@@
 Those calibration scenes are  5 i/km-scenes and 5 ni/km-scenes.
-These first 10 scenes are replayed at the end of the experiment, and only the notes given at the second occurrence are taken into account.
+These first 10 scenes are replayed at the end of the experiment, and only the scores given at the second occurrence are taken into account.
 
@@
 
-All the subjects perform the test on the same type of computers . Thy listen to the stimuli with semi-open \emph{Beyer-Dynamic DT 990 Pro} headphones. The sound level is the same for all subjects.
+All the subjects perform the test on the same type of computers. They listen to the stimuli with semi-open \emph{Beyer-Dynamic DT 990 Pro} headphones. The sound level is the same for all subjects.
 
@@
 
-A supervisor is available during the whole duration of the experiment to ensure a smooth running of the latter and answer any questions the subjects may have.
+A supervisor is available during the whole duration of the experiment to monitor the experiment and answer any questions the subjects may have.
 
@@
 The two independent variables are considered as \emph{within-subject} factors.
-The factors being of only two levels each (type: i/ni; marker: km/rm), the sphericity hypothesis does not need to be checked.
-\emph{Post hoc} analysis are done by following the Tukey-Kramer procedure.
+The factors being of only two levels each (type: i/ni,  marker: km/rm), the sphericity hypothesis does not need to be checked.
+\emph{Post hoc} analyses are done using the Tukey-Kramer procedure.
 
@@
 Let us consider $\mathcal{A}_{sujet}$ for the km-scenes.
-Close inspection that the answers given by one subject strongly differs with the ones of others.
+Close inspection of the answers show that one subject strongly differs with the ones of others.
 This subject evaluated positively close to half of the ni/km-scenes, see Annex~\ref{app:xp2}, subject 7.
@@
 Furthermore, this subject used the whole scale (from -5 to 5) to score both the i-scenes and the ni-scenes.
-Those behaviors strongly differing with the ones of the others subjects be they from this experiment or the previous ones, subject 7 is discarded from the analysis.
+Those behaviors strongly differ with the ones of the others subjects be they from this experiment or the previous ones.
+Subject 7 is thus discarded from the analysis.
 
@@
 
-In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes et ni/rm-scenes.
+In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes and ni/rm-scenes.
 The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ show a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
@@
 
-The \emph{post hoc} analysis exhibit significative differences between all the groups of observations, notably between the i/km-scenes ($p<0.05$) and the i/rm-scenes ($p<0.01$).
+The \emph{post hoc} analysis exhibits significative differences between all the groups of observations, notably between the i/km-scenes ($p<0.05$) and the i/rm-scenes ($p<0.01$).
 
@@
 The fact that their removal decreases $\mathcal{A}_{scene}$ clearly shows that it is possible to improve the perceived pleasantnes of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
-Those conclusions are in line with the positive approacj introduced by Schafer in \cite{schafer1977tuning}.
+Those conclusions are in line with the positive approach introduced by Schafer in \cite{schafer1977tuning}.
 
@@
 The compositional characteristics that are taken into account during the perceptual process appear to vary from one source to the other, from one type of environment to the other.
-This fact lead the authors to conclude that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.
+This fact lead the authors to believe that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.
 
@@
 \item in the case of non ideal scenes, one should focus on reducing the acoustic pressure level, whether globally, or by discarding specific sources such as \emph{sirens} or \emph{car horns}
-\item in the case of ideal scenes, one should first identity which sources are pleasant to the targeted community, second lower the volume of the other sound sources, and, if possible, raise the contribution or add positive sound markers.
+\item in the case of ideal scenes, one should first identify which sources are pleasant to the targeted community, second lower the volume of the other sound sources, and, if possible, raise the contribution or add positive sound markers.
 \end{itemize}
@@
 
-The outcomes of the three experiments described in this paper demonstrate the usefulness of considering simulation tools such as \emph{simScene} in order to scientifically question the perception of soundscapes.
-We also believe that its wider usage could enable urban planning decision-makers to question an entire community about its own representations of the sound environment to which it is exposed, and about the representations of the sound environments to which it would like to be exposed.
+The outcomes of the three experiments described in this paper demonstrate the usefulness of considering a dedicated simulation tool such as \emph{simScene} in order to scientifically question the perception of soundscapes in an innovative way.
+We also believe that its wider usage could enable urban planning decision-makers to question an entire community about its own representations of the sound environment to which they are exposed, and about the representations of the sound environments to which they would like to be exposed.
 
@@
 
-Future work should consider several other simulation experiments by varying the emotional qualities (quiet, comfortable, troublesome, etc.), but also by specifying particular places (park, square, street, etc.), in order to provide to the scientific community entire corpora of cognitively informed soundscapes.
+Future work should consider several other simulation experiments by varying the emotional qualities (quiet, comfortable, troublesome, etc.), but also by specifying specific urban places (park, square, street, etc.), in order to provide to the scientific community an entire corpora of cognitively informed soundscapes.
 
@@
 Finally, one should study the influence of socio-cultural contexts on perception.
-Indeed, if the sound of the church bell is most often a quality environment marker for a Westerner, this is not necessarily true with subjects of Eastern, Middle Eastern or other cultures.
+Indeed, if the sound of the church bell is most often a quality environment marker for a Westerner, this do not necessarily hold true for subjects of Eastern, Middle Eastern or other cultures.
 
@@
 \item The simulator can be deployed on a large scale via the Internet thanks to the chosen software architecture;
-\item simulated scenes can be analyzed without having to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
+\item simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
 \end{itemize}

### 2017-07-25 lafaySimulatedScenePerception.tex
new annex figures

@@
         \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_1}\label{fig:taxonomieEventa}}
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_2}\label{fig:taxonomieEventb}}\par
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_1}\label{fig:taxonomieTexturea}}
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_2}\label{fig:taxonomieTextureb}}
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_1_en}\label{fig:taxonomieEventa}}
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_2_en}\label{fig:taxonomieEventb}}\par
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_1_en}\label{fig:taxonomieTexturea}}
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_2_en}\label{fig:taxonomieTextureb}}
        \caption{Taxonomy of sound classes used for the simulation urban soundscapes: events (a,b) and textures (c,d).}

### 2017-07-31 lafaySimulatedScenePerception.tex
moving to elsevier style

@@
-\documentclass[twoside,twocolumn]{article}
-\usepackage{acta}
+\documentclass[preprint,12pt]{elsarticle}
 \usepackage{url}
@@
 \usepackage{sectsty}
+\usepackage{amsmath}
 
@@
 \newtheorem{mydef}{Définition}
-
-\begin{document}
-\Eingang{00}{00}{0000} \Annahme{00}{00}{0000} \Sachgebiet{1}
-\PACS{43.50.Lj, 43.50.Rq, 43.50.Qp, 43.66.Lj, 43.66.Ba}
-\Band{0} \Jahr{0000} \Heft{} \Ersteseite{1} \Letzteseite{}
-
-\AuthorsI{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J.F. Petiot$^{1)}$}
-\AddressI{$^{1)}$ Laboratoire des Sciences du Numérique, Nantes, France.\\
-\hspace*{8pt}mathieu.lagrange@cnrs.fr\\
-$^{2)}$ STMS Ircam-CNRS-UPMC Institut de Recherche et Coordination Acoustique/Musique, Paris, France}
 
@@
 
-\Englishtitle{Investigating soundscapes perception through acoustic scenes simulation.} \Germanorfrenchtitle{}
-
-\Kolumnentitel{Lafay et al. : Approaching mental representations}
-
-\Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process.
+\journal{Nuclear Physics}
+
+\begin{document}
+
+\begin{frontmatter}
+
+    \author{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J.F. Petiot$^{1)}$}
+
+    \address{$^{1)}$ Laboratoire des Sciences du Numérique, Nantes, France.\\
+    \hspace*{8pt}mathieu.lagrange@cnrs.fr\\
+    $^{2)}$ STMS Ircam-CNRS-UPMC Institut de Recherche et Coordination Acoustique/Musique, Paris, France}
+
+    \title{Investigating soundscapes perception through acoustic scenes simulation.}
+
+  \begin{abstract}
+    This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process.
 Subjects are asked to create a full soundscape by means of a dedicated software tool, coupled with a structured sound data set.
@@
 A rating experiment of the soundscape pleasantness using a 7-points Likert scale is conducted to further refine the analysis of the simulated urban acoustic scenes.
-Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.}
-\Germanorfrenchabstract{}
-
-\ScientificPaper
+Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.
+  \end{abstract}
+
+  \begin{keyword}
+  %% keywords here, in the form: keyword \sep keyword
+  toto \sep toto
+  %% PACS codes here, in the form: \PACS code \sep code
+
+  %% MSC codes here, in the form: \MSC code \sep code
+  %% or \MSC[2008] code \sep code (2000 is the default)
+
+  \end{keyword}
+
+\end{frontmatter}
 
@@
 
-\begin{equation*}
+\begin{equation}
 V_{jk}=\dfrac{c_{jk}-c_k\frac{c_j}{c}}{\sqrt{c_k\frac{c-c_k}{c-1}\frac{c_j}{c}(1-\frac{c_j}{c})}}
-\end{equation*}
+\end{equation}
 
@@
 
-\begin{equation*}
+\begin{equation}
 d_{ham}(S_1,S_2)=\dfrac{1}{n}\sum_{i=1}^{n} (x_{1,i} \bigoplus x_{2,i})
-\end{equation*}
+\end{equation}
 
@@
 
-\References{}{Bibliography}{actalit}
-
+\bibliographystyle{elsarticle-num}
+\bibliography{Bibliography}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-07-31 lafaySimulatedScenePerception.tex
done with NM comments

@@
   %% keywords here, in the form: keyword \sep keyword
-  toto \sep toto
+ cognitive psychology \sep soundscape assessment
   %% PACS codes here, in the form: \PACS code \sep code
@@
 
-First, a subject can be asked to describe a given sound following a given perceptive scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}.
+First, a subject can be asked to describe a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}.
 The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}.
@@
 
-To the best of our knowledge, only Bruce~\al \cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception.
+To the best of our knowledge, only Bruce~\cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception.
 They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location.
@@
 
-\emph{Simscene} is an online digital audio tool whose first version have been developed as part of the HOULE prokect\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
-It has been designed to run on the popular web browers \emph{Chrome} and \emph{Firefox}.
+\emph{Simscene} is an online digital audio tool whose first version have been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
+It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}.
 It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}.
@@
 
-We choose not to include any musical content in the database of sounds, as the study of the pleasantness of a given style or genre of music is beyond the scope of the study.
+We choose not to include any musical content in the sound database, as the study of the pleasantness of a given style or genre of music is beyond the scope of the study.
 
@@
 The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures only.
-These two types also have specific scheduling procedures during the sumulation process.
+These two types also have specific scheduling procedures during the simulation process.
 
@@
 On contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}.
-They build upon a holistic analysis based on global acoustic features.
+They result from a holistic analysis based on global acoustic features.
 
@@
         \includegraphics[width=.8\linewidth]{gfx/ch_4/3.png}
-       \caption{Hierachical organisation of the isolated sounds used ofr the simulation.}\label{fig:orgDb}
+       \caption{Hierachical organisation of the isolated sounds used of the simulation.}\label{fig:orgDb}
 \end{figure}
@@
 
-Once the typology and the set of sounds are available, an important design issue os the need of a suitable way to display the sound dataset to the user.
+Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user.
 Most browsing tools are based on keyword indexing.
@@
 
-Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a two dimensional plane.
+Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a plane.
 
@@
 
-As such, the underlying model of the scene is a sum of sound sources, more precisely a "skeleton of events on a bed of textures" \cite{nelken2013ear}.
-Further details about the simulation interface can be found in \cite{rossignol2015simscene}.
+As such, the underlying model of the scene is a sum of sound sources. Further details about the simulation interface can be found in \cite{rossignol2015simscene}.
 
@@
 For the remainder of the study, the defined hedonic properties (i and ni) are used as references.
-We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of percieved pleasantness.
+We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of perceived pleasantness.
 \end{enumerate}
@@
 
-All the participants start by desiging the ideal environment; they read the second set of instruction at the end of the first experiment.
+All the participants start by designing the ideal environment; they read the second set of instruction at the end of the first experiment.
 Participants are completely free of their choice concerning sounds and synthesis parameters.
@@
 A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}.
-In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as timuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.
+In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.
 For each sound scene, three types of features are considered :
@@
 The significativity is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
-\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni- scenes has a significative influence  on the nature of the scenes \ie if significative differences exist between structural features and perceived pleasantness.
+\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni- scenes has a significative influence  on the nature of the scenes \ie~ if significative differences exist between structural features and perceived pleasantness.
 The significativity is measured with a Student test on two populations;
@@
 where $\bigoplus$ is the \emph{exclusive-or}.
-The more the comosition of the two scenes are similar, the closer they are in this space.
+The more the composition of the two scenes are similar, the closer they are in this space.
 The use of Hamming distance allows us to take into account equally the presence and absence of classes.
@@
 This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}.
-During their experiment, the authors ask to their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source.
+During their experiment, the authors ask their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source.
 The study shows that there is no significative difference between global and averaged instantaneous judgments.
@@
 For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$).
-Again, we believe that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer than ni - scenes).
+Again, we believe that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer than ni- scenes).
 Thus, the sound events keep a relative ability to predict the pleasantness of the ni- scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.
@@
 
-Those first outcomes tend to show that 1) two modes of perception exists depending on the very nature of the environment (i- or ni-), and 2) each involve independent features.
+Those first outcomes tend to show that 1) two modes of perception exist depending on the very nature of the environment (i- or ni-), and 2) each involve independent features.
 The fact that $L$ is not sufficient to characterize the pleasantness of the i- scenes can lead us to conclude that all the sound sources do not equally contribute to the perception of pleasantness.
@@
 Results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
-For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$ is used to depict classes, see ~Figure~\ref{fig:taxonomie}.
+For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$ is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.
 
@@
 In fact,  Figure~\ref{fig:soundsourcea} show that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the participants for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes.
-the results do not re-evaluate the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} pour les i-scenes, a level that is comparable to the of the \emph{bike} class, and much higher than all the personal vehicles classes.
+the results do not re-evaluate the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the i-scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes.
 Though, \emph{public transport} classes are also strongly present in the ni-scenes, for instance more than \emph{light vehicle} or \emph{trucks} classes.
@@
 
-Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and stay above $86\%$ from the abstraction level $1$.
+Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and stay above $86\%$ for the abstraction level $1$.
 These results confirm that it is possible to make a clear distinction between the two types of environment on the sole basis of presence / absence of sound classes.
 We also notice that the higher the abstraction level is, the more important the ability to make this distinction is.
-In other words, the more precise in the scene description is, the more clear the discrimination between i- and ni- scenes is.
+In other words, the more precise the scene description is, the more clear the discrimination between i- and ni- scenes is.
 
@@
 In this section, the structural features are computed by taking into account only the previously identified sound markers.
-We define $Xm$, the $X$ feature computed by taking into account only the sound markers.
+We define $X_m$, the $X$ feature computed by taking into account only the sound markers.
 On the opposite, we define $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
@@
 Considering the non markers classes, we can observe on the i- scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}.
-It is the first time that an objective feature allows us to define the pleasantness of the pleasant environments.
+It is the first time that an objective feature allows us to define the pleasantness of i- scenes.
 This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.
@@
 We consider a last group of features, namely $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$.
-These features describe the difference between the markers level and those of the other sound classes, see ~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}).
+These features describe the difference between the markers level and those of the other sound classes, see ~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}.
 They express the saliency of the markers with respect to the sound mixture.
@@
 The semantic description seems to be more powerful;
-\item \emph{events or textures}: whatever the feature type, be it semantic or structural, events are mainly useful to differentiate the two types of events; textures bring a limited amount of information;
+\item \emph{events or textures}: whatever the feature type, be it semantic or structural, events are mainly useful to differentiate the two types of scenes; textures bring a limited amount of information;
 \item \emph{pleasantness prediction}: considering the correlation between structural features and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its nature (i or ni).
-From the data gathred in those experiments, it is not conceivable to consider the same set of features to predict the pleasantness of i- and ni- scenes in the same time:
+From the data gathered in those experiments, it is not conceivable to consider the same set of features to predict the pleasantness of i- and ni- scenes in the same time:
 
@@
 
-The fact that the pleasantness of the i- scenes is not correlated to global physical features, contrary to pleasantness of ni- scenes has been studied recently \cite{gozalo2015relationship}.
-
-We thus observe two perceptual modes of operation that involve different types of features, and that rely on the hedonic nature  of the stimuli.
-A phenomenon similar is observed for the perception of textures, see Section~\ref{sec:simscene_sampleDataSet}.
-It seems that the brain adapts its way to encode the information (statistic summary for textures, fine description for events) following a previous decision making with regards to the nature of the stimuli, \ie is it an event or a texture ?.
+The fact that the pleasantness of the i- scenes is not correlated to global physical features, contrary to pleasantness of ni- scenes, has been studied recently \cite{gozalo2015relationship}.
+
+We observe two perceptual modes of operation that involve different types of features, and that rely on the hedonic nature  of the stimuli.
+A similar phenomenon is observed for the perception of textures, see Section~\ref{sec:simscene_sampleDataSet}.
+It seems that the brain adapts its way to encode the information (statistic summary for textures, fine description for events) following a previous decision making with regards to the nature of the stimuli, \ie~is it an event or a texture ?
 Similarly, in the light of this study, it appears that the features considered for the pleasantness judgment also depends on a preliminary judgment of the global hedonic nature of the environment (ideal or non ideal).
@@
 
-In order to verify that the pleasantness of the ideal and non ideal scenes is affected by the sound markers, the audio waveform of the simulated scenes are regenerated with or without using the classes identified as markers.
+In order to verify that the pleasantness of the ideal and non ideal scenes is affected by the sound markers, the audio waveform of the simulated scenes are regenerated with or without the classes identified as markers.
 For the i-scenes, some of the i-markers are removed.
 For the ni-scenes, some of the ni-markers are removed.
-To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptive test is then conducted with a protocol close to the one considered in experiment 1.b.
+To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptual test is then conducted with a protocol close to the one considered in experiment 1.b.
 
@@
 
-Finally, these works allow us to conjecture as to the nature of the mental representations of the concepts "pleasant urban sound environment" (EA) and "unpleasant urban sound environment" (ED).
+Finally, these works allow us to conjecture as to the nature of the mental representations of the concepts "pleasant (urban sound) environment" (PE) and "unpleasant (urban sound) environment" (UE).
 
@@
 
-First, the fact that the semantic information (in terms of which sound source is present) and compositional information are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the EA and ED concepts.
+First, the fact that the semantic information (in terms of which sound source is present) and compositional information are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the PE and UE concepts.
 
@@
 
-Secondly, the fact that the removal of sound markers modifies the perceived pleasantness leads us to believe that the abstract concept related to pleasantness depends on the activation of a network of concrete concepts strongly linked to the sources which are in the case of this study: \emph{bird}, \emph{church bell} and \emph{bicycle bell}).
+Secondly, the fact that the removal of sound markers modifies the perceived pleasantness leads us to believe that the abstract concept related to pleasantness depends on the activation of a network of concrete concepts strongly linked to the sources which are in the case of this study: \emph{bird}, \emph{church bell} and \emph{bicycle bell}.
 
@@
 
-Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics that allow us to distinguish between two instances of the same abstract concept linked to the pleasantness are not the same for EA and ED.
+Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics that allow us to distinguish between two instances of the same abstract concept linked to the pleasantness are not the same for PE and UE.
 
@@
 
-Future work should consider several other simulation experiments by varying the emotional qualities (quiet, comfortable, troublesome, etc.), but also by specifying specific urban places (park, square, street, etc.), in order to provide to the scientific community an entire corpora of cognitively informed soundscapes.
+Future work should consider several other simulation experiments by varying the emotional qualities (quiet, comfortable, troublesome, etc.), but also by specifying specific urban locations (park, square, street, etc.), in order to provide to the scientific community an entire corpora of cognitively informed soundscapes.
 
@@
 \begin{figure}[h]
-        \myfloatalign
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_1_en}\label{fig:taxonomieEventa}}
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_2_en}\label{fig:taxonomieEventb}}\par
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_1_en}\label{fig:taxonomieTexturea}}
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_2_en}\label{fig:taxonomieTextureb}}
-       \caption{Taxonomy of sound classes used for the simulation urban soundscapes: events (a,b) and textures (c,d).}
-       \label{fig:taxonomie}
+    \centering
+        {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.7\columnwidth]{gfx/ch_5/event_1_en}\label{fig:taxonomieEventa}} \par
+         \caption{Taxonomy of sound classes of mechanical events used for the simulation urban soundscapes  with level of abstraction from left to right.}
+       \label{fig:taxonomieEa}
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+        {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.7\columnwidth]{gfx/ch_5/event_2_en}\label{fig:taxonomieEventb}}
+       \caption{Taxonomy of sound classes of non-mechanical events used for the simulation urban soundscapes with level of abstraction from left to right.}
+       \label{fig:taxonomieEb}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+        {\includegraphics[trim={ 0 0 7cm 0},clip,width=.7\columnwidth]{gfx/ch_5/texture_1_en}\label{fig:taxonomieTexturea}} \par
+
+        {\includegraphics[trim={ 0 0 7cm 1cm},clip,width=.7\columnwidth]{gfx/ch_5/texture_2_en}\label{fig:taxonomieTextureb}}
+       \caption{Taxonomy of sound classes of textures used for the simulation urban soundscapes with level of abstraction from left to right.}
+       \label{fig:taxonomieT}
 \end{figure}
@@
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_12}\label{fig:xp4_note_1l}}
-        \caption{Distribution of scores given par the subjects during experiment 2 for i/am-scenes (green) et ni/am-scenes (red).}\label{fig:xp4_note_1}
+        \caption{Distribution of scores given par the subjects during experiment 2 for i/am-scenes (green) et ni/am-scenes (red).
+The horizontal axis is, from left to right, the 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant)}\label{fig:xp4_note_1}
 \end{figure}

### 2017-08-23 lafaySimulatedScenePerception.tex
jfp comments

--- 2017-07-25 lafaySimulatedScenePerception.tex
@@
 
-\AuthorsI{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J.F. Petiot$^{1)}$}
+\AuthorsI{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J-F. Petiot$^{1)}$}
 \AddressI{$^{1)}$ Laboratoire des Sciences du Numérique, Nantes, France.\\
@@
 \Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process.
-Subjects are asked to create a full soundscape by means of a dedicated software tool, coupled with a structured sound data set.
+Subjects are asked to create a full soundscape by means of a software tool dedicated to sound edition, coupled with a structured sound data set.
 This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes.
@@
 For example, in order to study the link between an environment (urban sound scene) and the induced human sensation (calmness), one mean is to objectify the mental representation of a specific sound scene, a calm urban sound scene.
-Several means have been considered in the literature to do so.
+Several means have been considered in the literature to perform this task.
 
@@
 
-First, a subject can be asked to describe a given sound following a given perceptive scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}.
+First, a subject can be asked to assess a given sound environnment following a given perceptive scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}.
 The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}.
-That being said, those stimuli respectively are or can be recorded and analysed.
-It thus allows the experimenter to gather a coarse grain physical description of the stimuli.
+That being said, those recorded stimuli can be analysed, which allows the experimenter to gather a coarse
+grain physical description of them.
 
@@
 Second, a subject can be asked to describe a given sound environment \cite{guastavino2006ideal, dubois2006cognitive}.
-A large amount of quantitative and semantic information is then collected about the representation of this type of sound environment that the subject has.
-Unfortunately, without any reference to sound data, this representation can hardly be characterized physically.
+A large amount of quantitative and semantic information is then collected about the subject's representation of this type of sound environment.
+Unfortunately, without any reference to sound data, this representation cannot be characterized physically.
 
@@
 
-We propose in this paper to consider the use of a soundscape simulator that the subject can use to objectify her representation of a given sound environment.
+We propose in this paper to consider the use of a soundscape simulator that the subject can manage to objectify his/her representation of a given sound environment.
 We believe that the use of such a device allows us to gain the benefits of the two above cited approaches.
@@
 
-We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
+We believe that the availability of a fine grain description of the sound stimuli is of great insterest, beceause recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
 szeremeta2009analysis}.
@@
 
-For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sounds sources that compose a sound environment of interest.
-With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the events.
+For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to very interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment of interest.
+Indeed, with such a material, not only the type of sound sources is available for study, but also the exact level and audio waveform for each source and also the structural characteristics of the scene, that is the temporal distribution of the events.
 
@@
 To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmental scenes.
-Results and outcomes of a series of three experiments that foster on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:
+Results and outcomes of a series of three experiments that build on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:
 
@@
 The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics.
-A lack of diversity is nonetheless mentioned by the authors that limit the strenght of the outcomes given in this study.
+A lack of diversity is nonetheless mentioned by the authors that limit the outcomes of the study.
 
@@
 
-In our approach, the simulator developed for this study\footnote{simScene, available online \url{https://bitbucket.org/mlagrange/simscene}} gives a monophonic representation of the scene, with the added benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are both expressive and useful for analysis.
+In our approach, the simulator developed for this study\footnote{simScene, available online \url{http://soundthings.org/research/}} gives a monophonic representation of the scene, with the added benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are, as much as possible, expressive and useful for analysis.
 
@@
 The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}.
-The experiments  1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section \ref{sec:simulation}.
-Experiment 2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section \ref{sec:modification}.
+The three experiments,  1.a \emph{simulation}, 1.b \emph{pleasantness evaluation}, and 2 \emph{pleasantness evaluation after modification of the scenes} are respectively presented in Sections \ref{sec:simulation}, \ref{sec:evaluation}, and \ref{sec:modification}.
 Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.
-Experiment 1.a is based on a pilot study described in \cite{lafay2014new}.
-
 %Dans le cadre de cette communication, nous ne présentons de \emph{SimScene} que les fonctionnalités pertienentes pour notre étude.
@@
 \section{The simulator}
-\label{sec:simulator}
+\label{sec:simscene}
 
@@
 
-\emph{Simscene} is an online digital audio tool whose first version have been developed as part of the HOULE prokect\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
+\emph{Simscene} is an online digital audio tool whose first version has been developped as part of the HOULE prokect\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
 It has been designed to run on the popular web browers \emph{Chrome} and \emph{Firefox}.
@@
 To do so, the user first selects a sound source using a non verbal selection interface presented in Section~\ref{sec:simscene_ssf}.
-A track is then created for this sound source within the simulator interface.
-The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sounds source.
+A track is then created for this sound sources within the simulator interface.
+The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sound sources.
 Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.
@@
 
-In order to provide the user with a sound database that is well organized and covers as much as possible the variety of sound sources that are available in urban areas, a typology of urban sounds is first defined.
+In order to provide the user with a sound database that is well organized and covers as much as possible the variety of sound sources that are present in urban areas, a typology of urban sounds is first defined.
 
@@
 
-The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} that study how humans discriminate different kind of urban soundscapes.
+The chosen typology is established based on the category/classes of sounds found while reviewing 16 articles or thesis manuscripts \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} that study how humans discriminate different kind of urban soundscapes.
 
@@
 
-The literature usually distinguish between:
+Literature usually distinguishes between:
 \begin{itemize}
@@
 
-The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures only.
-These two types also have specific scheduling procedures during the sumulation process.
+The simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with classes of events only and the other with classes of textures.
+These two types have specific simulation procedures.
 
@@
 
-According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from different sources of a sound environment \cite{bregman1994auditory}.
+According to the Auditory Scene Analysis (ASA) theory, the brain processes seperately events emerging from different sources of a sound environment \cite{bregman1994auditory}.
 Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}.
@@
 
-Maffiolo \cite{maffiolo_caracterisation_1999} distinguish two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events.
+Maffiolo \cite{maffiolo_caracterisation_1999} distinguishes between two different categorization processes, that differentiate themselves with the capacity of the listener to identify sound events.
 She shows that those processes leads to two abstract cognitive category respectively termed "the event sequences" and "the amorphous sequences".
@@
 
-The event sequences are composed of salient events that can easily be recognized such as \emph{car start} or \emph{male speech}.
+The event sequences are composed of salient and identifiable events such as \emph{car start}, \emph{male speech}.
 They arise from a descriptive analysis based on the identification of the sound sources.
-On contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}.
-They build upon a holistic analysis based on global acoustic features.
+On the contrary, the amorphous sequences are environments where distinct events can hardly be found, such as \emph{traffic hubhub}.
+They build upon an holistic analysis based on global acoustic features.
 
@@
 
-Concerning sound textures, \ie~sounds that have stable characteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can decide not to abstract the physical information available using some of the summary statistics discarding lower level information \cite{nelken2013ear}.
+Concerning sound textures, \ie~sounds that have stable caracteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can decide not to abstract the physical information available using some of the summary statistics discarding lower level information \cite{nelken2013ear}.
 
@@
         \myfloatalign
-        \includegraphics[width=.8\linewidth]{gfx/ch_4/3.png}
-       \caption{Hierachical organisation of the isolated sounds used ofr the simulation.}\label{fig:orgDb}
+        \includegraphics[trim={5cm 0 0 0},clip,width=.8\linewidth]{gfx/ch_4/3}
+       \caption{Hierarchical organisation of the isolated sounds used in the simulation.}\label{fig:orgDb}
 \end{figure}
@@
 
-We term sample, a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples perceived equivalently.
+We term "sample", a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples perceived equivalently.
 
@@
 
-The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E.
+The sound classes are organized hierachically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E.
 Rosch \cite{rosch1978cognition}.
-The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources is.
-For classes with high level of abstraction that have sub classes, its collection of samples is the union of the collections of the sub classes.
+The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources.
+For classes with high level of abstraction with sub classes, their collection of samples is the union of the collections of the sub classes.
 
@@
 Accounting for the perceptual matters detailed previously, two taxonomies are built, one for event sounds and one for texture sounds.
-Four levels of abstraction are considered from the more generic one (level 0), to the more specific classes (level 3), leading to a taxonomy close to the one used in \cite{Salamon14}.
+Four levels of abstraction are considered from the more generic one (level 0), to the more specific classe (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}.
 Only three levels of abstraction are considered for the texture sounds.
@@
 The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background.
-It also allowed us to avoid dominant sound sources while recording texture sounds by targeting distant areas with no dominant sound sources.
+It also allowed us to avoid dominant sound sources while recording texture sounds by aiming the microphone to distant areas with no dominant sound sources.
 
@@
 
-All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie relative to Full Scale. In our case, the full scale level is of 1 Volt.
+All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie. relative to Full Scale. In our case, the full scale level is of 1 Volt.
 
@@
 
-By a volountary design choice, the simulation tool do not allow the user to interact and control a specific sample.
+By design, the simulation tool does not allow the user to interact and control directly a specific sample.
 Interaction is done at the track level, a track being a sequence of samples.
@@
 
-To improve simulation quality, two parameters are also proposed:
+To ease the simulation, two parameters are also proposed:
 
@@
 
-Once the typology and the set of sounds are available, an important design issue os the need of a suitable way to display the sound dataset to the user.
-Most browsing tools are based on keyword indexing.
-However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potenitally induce biases in the analysis conclusions.
-Indeed, the subject may select \textit{a priori} sounds referenced as belonging to \emph{park} to build a \emph{calm} soundscape.
+Once the typology and the set sound are available, an important design question that remains is to find a suitable way of presenting the data to the user.
+Most of browsing tools are based on keyword indexing.
+However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject.
+Indeed, the subject may select \textit{a priori} sound referenced belonging to \emph{park} to build a \emph{calm} soundscape.
 
@@
 
-Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a two dimensional plane.
+Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a two dimensional plane, each circle reprensenting a level 2 class.
 
@@
 Each of those classes have a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle.
-The subject can browse the database by listening to those prototype sounds.
+The subject can browse the database by listening to those prototype sound.
 The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed in \cite{lafay2016JAES}.
@@
 For event's tracks, the horizontal spacing between those rectangles is a function of the time delay between their onsets.
-For tracks of texture, a unique rectangle is displayed as this kind of sounds do not allow spacing with silence.
+For tracks of texture, a unique rectangle is displayed as this kind of sounds does not allow spacing with silence.
 As actual amplitude and spacing values are drawn from random variables, each time the subject changes the value of a parameter, the location and height of the rectangles are updated to reflect the changes in the sequencing of the samples.
@@
 
-\section{Experiment 1}
-\label{sec:simulation}
-
-\subsection{Objective}
+\section{Expérience 1}
+\label{sec:xp1}
+
+\subsection{Objectif}
 
@@
 
-Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness.
-For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
-\begin{itemize}
-\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}.
+Experiment 1 aims at using the simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness.
+For that, two first experiments are planned (\cf~Figure~\ref{fig:xp1_2}):
+\begin{itemize}
+\item \emph{experiment 1.a (simulation)}: during this experiment, participants have to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simscene}.
 Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
-\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes: respectively ideal (i) or non ideal (ni).
-Furthermore, this information is given by the creator of the scene.
-The second experimental step aims at investigating more deeply and more broadly our knowledge on pleasantness.
-For that, a second group of participants are asked to evaluate the pleasantness of each simulated scene, on a semantic scale.
-This experiment has two goals:
+\item \emph{experiment 2.a (evaluation)}: actually, after the simulation phase, only a binary information on the emotional properties of the respective scenes is available: ideal (i) and non ideal (ni).
+This second experimental step aims at investigating our knowledge on pleasantness.
+For that, a second goup of participants is asked to evaluate the pleasantness of each simulated scene, on a semantic scale.
+This evaluation protocol has two goals:
 \begin{enumerate}
-\item evaluate the respective influence on the pleasantness of the various sources composing the scenes (i or ni)
-\item detect the presence of \emph{outliers} or ambiguous samples.
-For the remainder of the study, the defined hedonic properties (i and ni) are used as references.
-We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of percieved pleasantness.
+\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene.
+\item to detect the presence of \emph{outliers} or ambiguous samples.
+For the remainder of the study, the defined hedonic properties (i and ni) are used as ground truth.
+Then, we must have to be sure that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of pleasantness rating.
 \end{enumerate}
@@
 
-Our analysis is based on the data collected by these two experiments: 1.a and 1.b.
+Our analysis is based on the data collected by these two experiments (1.a, 1.b)
 
@@
         \myfloatalign
-        \includegraphics[width=\linewidth]{gfx/ch_5/5.png}
+        \includegraphics[width=\linewidth]{gfx/ch_5/5}
         \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b)}\label{fig:xp1_2}
@@
 \label{sec:xp1a_plan}
+
+In order to validate the design of this experiment, a pilot study has been performed and is described in \cite{lafay2014new}.
 
@@
 
-The participants have to simulate two urban sound environments of one minute each. They follow the instructions below:
-\begin{itemize}
-\item  first simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
-\item second simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
-\end{itemize}
-
-All the participants start by desiging the ideal environment; they read the second set of instruction at the end of the first experiment.
-Participants are completely free of their choice concerning sounds and synthesis parameters.
+The participants have to simulate two urban sound environments of 1 minute each. They have to follow the instructions below:
+\begin{itemize}
+\item  1st simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
+\item 2nd simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
+\end{itemize}
+
+All the participants start by working on the ideal environment; they read the second set of instruction at the end of the first experiment.
+Participants are completely free of their choices concerning sounds and synthesis parameters.
 They must nevertheless respect the two following constraints :
@@
 \item the listening point of view is the one of a fixed listener;
-\item the soundscape must be realistic, i.e. physically plausible.
-\end{itemize}
-For instance, the participant can insert ten dogs in the soundscape but she cannot insert one dog barking every 10 milliseconds.
+\item  the soundscape must be realistic, i.e. physically plausible. For instance, the participant can insert ten dogs in the soundscape but he/she can’t insert one dog barking every 10 milliseconds.
+\end{itemize}
 These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation user interface.
@@
 
-Each simulation process contains two parts:
+Each simulation process contains two parts :
 \begin{enumerate}
-\item composition of the simulation: this step is split in three steps:
+\item composition of the simulation: this step is split in three steps
 \begin{itemize}
@@
 
-In addition, once the two sound scenes are done, participants are invited to
+In addition, once the two sound scenes have been realized, participants are invited to
 \begin{itemize}
@@
 
-All the participants do the experiment on desktop computers with the same hardware and software configurations.
+All the participants do the experiment on similar computers.
 The audio files are played in diotic conditions by headphones.
-During the tutorial, participants are asked to adjust the sound level in order to be comfortable.
-Once set, they are not allowed to modify it in the remaining of the experiment.
-All the participants do the experiment at the same time. They are equally split in three identical quiet rooms. They are not allowed to talk to each other during the experiment.
-Three experimenters (one in each room) are present during the whole duration of the experiment in order to assist participants with potential hardware and software issues and to answer questions.
+During the tutorial, participants are asked to adjust the sound level in order to be comfortable; they can’t modify it in the following of the experiment.
+All the participants do the experiment simultaneously. They are equally split in three identical quiet rooms. They can’t talk together during the experiment.
+Three experimenters (one in each room) are present during the whole duration of the experiment in order to check that everything goes well and possibly to answer questions.
 
@@
 
-44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in downtown Nantes since 2 years or more.
-Among the 44 participants, 40 succeeded, producing in the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
-the 4 other participants were excluded from the process due to a lack of respect or understanding of the instructions, or for exceeding the maximum duration of the experiment.
+44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in Nantes downtown since 2 years or more.
+Among the 44 participants, 40 succeeded in doing the experiment, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
+4 participants were excluded from the process for a lack of understanding of the instructions, or for exceeding the maximum duration of the experiment.
 
@@
 
-Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes composed in Experiment 1.a (from sec. 15 to sec. 45).
-The assessment is done with a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
-Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli.
+Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes composed in Exp. 1.a (from sec. 15 to sec. 45).
+The assessment is done in a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
+Before evaluating a scene, participants must listen to the first 2à seconds of the scene.
 After the evaluation, they are free to switch to the next scene.
 For each participant, sound scenes are played in a random order.
-The first 10 scenes allow the participants to calibrate their scores; as they are always composed of 5 ideal scenes and 5 non ideal scenes.
-These first 10 scenes are played back again at the end of the experiment and the last evaluations are the ones that are only taken into account in the results.
-The experiment is planned to last 30 minutes. The participants do not know anything beforehand about the nature of the sound scenes.
+The first 10 scenes allow the participants to calibrate their marks; they imperatively gather 5 ideal scenes and 5 non ideal scenes.
+These first 10 scenes are played back again at the end of the experiment and the corresponding evaluations are the ones that are only taken into account in the results.
+The experiment is planned to last 30 minutes. The participants don’t know anything beforehand about the nature of the sound scenes.
 
@@
 
-All the participants do the experiment on computers with equal hardware and software configurations.
+All the participants do the experiment on similar computers.
 The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}.
-All the sound scenes have been re-composed on the basis of the scores obtained in the simulation experiment (Exp.
+The stimuli are the sound scenes obtained in the simulation experiment (Exp.
 1.a).
 The output sound level is the same for all the participants.
-All the participants do the experiment simultaneously. They are not allowed to talk to each other during the experiment.
-An experimenter is present during the whole duration of the experiment in order to assist participants and possibly answer questions.
+All the participants do the experiment simultaneously. They can’t talk together during the experiment.
+An experimenter is present during the whole duration of the experiment in order to check that everything goes well and possibly to answer questions.
 
@@
 10 students (2 females - mean age 23.1, SD = 1.8) from Ecole Centrale de Nantes took part to the experiment.
-All are living in downtown Nantes for 2 years or more.
+They are all living in Nantes since 2 years or more.
 None of them took part to the previous simulation experiment (Exp.
@@
 A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}.
-In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as timuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.
+In order to stay coherent with the evaluation experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version, see Section \ref{sec:xp1b_plan}.
 For each sound scene, three types of features are considered :
@@
 \item \emph{perceptual}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale.
-$\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants.
-$\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant.
+$\mathcal{A}_{scene}$ is the mean pleasantness for one scene, as the average of the marks of all the participants.
+In the same idea, $\mathcal{A}_{subject}$ is the mean pleasantness for each participant, as the average of all the marks of each participant.
 Given the low number of participants in Exp.
-1.b, we choose not to normalize the pleasantness scores.
-\item \emph{semantic}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene.
-Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise.
-The vector dimension ($n$) depends on the level of abstraction that is considered for analysis.
-For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.
-\item \emph{structural}: features related to \emph{sound level}.
+1.b, we decide not to normalize the values of pleasantness marks.
+\item \emph{semantic}: boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene.
+Each point $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise.
+The vector dimension ($n$) depends on the level of abstraction that is considered.
+For instance, for the abstraction level 1, including 44 classes of sounds, $n = 44$.
+\item \emph{structural}: related to \emph{sound level}.
 To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference.
-In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt.
-The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene.
+In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in deciBels, with a reference of 1 Volt (full scale).
+The level is obtained by computing every second the quadratic mean of the signal and by averaging over the total duration of the scene.
 An A-filtering module processes the data before the quadratic means are computed.
@@
 
-In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following tests of significativity:
-
-\begin{itemize}
-\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with respect to the two types of scenes.
-The significativity is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
-\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni- scenes has a significative influence  on the nature of the scenes \ie if significative differences exist between structural features and perceived pleasantness.
-The significativity is measured with a Student test on two populations;
-\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, linear correlations between these features and $\mathcal{A}_{scene}$ are considered.
-The Pearson correlation coefficient is used for that purpose, on the basis of the standard methodology used in multidimensional studies;
-\item \emph{analysis of the semantic features}: in order to determine if the affective difference between  the i- and ni- scenes has an influence on the scenes composition in terms of sound sources, or if specific classes are more frequently used to simulate a given environment (i or ni), a V-test is considered.
-Testing is done at every level of abstraction, and separately for texture and event classes.
-For each class $j$, and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:
+In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following significance tests:
+
+\begin{itemize}
+\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with regards to the two types of scene.
+the significance is evaluated by a Student test on two independant populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$ ;
+\item \emph{analysis of the structural features}: in order to evaluate wether the emotional difference forced to the i- and ni- scenes has a significative influence \ie, if significative differences exist between structural features and perceived pleasantness, we measure this significance with a Student test on two populations;
+\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, we look at linear correlations between these features and $\mathcal{A}_{scene}$.
+For that, we use the Pearson correlation coefficient, on the basis of the standard methodology used in multidimensional studies;
+\item \emph{analysis of the semantic features}: in order to determine if the forced emotional difference has an influence on the scenes composition in terms of sound sources, or if specific souns classes are more frequently used to simulate a given environment (i or ni), we apply a V-test.
+This test is done at every level of abstraction, and separately, for texture and event classes.
+For each class $j$ , and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:
 
@@
 
-\item \emph{Study of the representation spaces induced by the semantic features}: in order to determine if a representation only based on the presence / absence of sound sources allows us to distinguish between the two types of environment, we investigate the space induced by the semantic features $S$.
+\item \emph{Study of the representation spaces induced by the semantic features}: in order to determine if a representation only based on the presence / absence of sound sources allows to split the two types of environment, we investigate the space induced by the semantic features $S$.
 $S$ is a boolean vector and the distance between two scenes is computed with the Hamming distance.
-Considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, where $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ is of cardinality $n$, with $x=\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between two vectors, ans is defined as follows:
+Considering two n-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, where $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ is of cardinality $n$, with $x=\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between two vectors, as follows:
 
@@
 
-where $\bigoplus$ is the \emph{exclusive-or}.
-The more the comosition of the two scenes are similar, the closer they are in this space.
-The use of Hamming distance allows us to take into account equally the presence and absence of classes.
-For measuring the intrinsic ability of the space to split the i- and ni- scene, we use a ranking metric named the precision at rank $k$ ($P@k$) .
-The $P@k$ computes the precision obtained after the $k$ closest items of a given seed item have been found.
-Formally, for each $s_i$ scene (considered as seed), we compute the ratio between the number of $s_j$ scenes taken among the $k$ nearest neighbors of $s_i$ and sharing the same label than $s_i$ versus $k$.
-the $P@k$ is then the average of this ratio for all the items considered as search seeds;
-
-\item \emph{influence of the sound markers on the perceived pleasantness}: in order to assess the specific contributions of some sound sources, we again estimate the impact of the structural features on the perceived pleasantness by taking into account only the sound markers for the computation of those features.
-\end{itemize}
-
-Except for the V-test, the statistical test of significativity are done with a critical threshold of $\alpha=0.05$.
+where $\bigoplus$ is the \emph{exclusive-or} The more the two scenes are similar, the closer they are.
+The use of Hamming distance allows to take into account equally the present and absent classes.
+For measuring the intrinsic ability of the space to split the i- and ni- scene, we use a ranking metric named $k$- precision ($P@k$) .
+The $P@k$ computes the precision obtained after $k$ items have been found.
+Formally, for each $s_i$ scene, we compute the ratio between the number of $s_j$ scenes taken among the k nearest neighbors of $s_i$ and sharing the same label than $s_i$ with regards to the $k$ items to be found.
+the $P@k$ is then the mean of this ratio along all the items;
+
+\item \emph{influence of the sound markers on the perceived pleasantness}: in order to assess the specific contributions of some sound sources, we again estimate the possible impact of the structural features on the perceived pleasantness, but this time, by only taking into account the sound markers for the computation of these features.
+\end{itemize}
+
+Except for the V-test, all the statistical test of significance are done with a critical threshold of $\alpha=0.05$.
 For the V-test, considering that a great number of classes is tested, a Bonferroni correction is applied.
@@
 \begin{tabular}{c c}
-Terms                         & Acronyms              \\
+Termes                         & Acronyms              \\
 \hline
-Level                        & $L$                    \\
-Level (events)            & $L(E)$                 \\
-Level (textures)              & $L(T)$                 \\
-Average pleasantness (per scene)     & $\mathcal{A}_{scene}$  \\
-Average pleasantness (per subject)     & $\mathcal{A}_{sujet}$  \\
+Niveau                         & $L$                    \\
+Niveau (événements)            & $L(E)$                 \\
+Niveau (textures)              & $L(T)$                 \\
+Agrément moyen (par scène)     & $\mathcal{A}_{scene}$  \\
+Agrément moyen (par sujet)     & $\mathcal{A}_{sujet}$  \\
                                &                        \\
-ideal/pleasant                & i                      \\
-non-ideal/unpleasant         & ni                     \\
-scene ideal/pleasant          & i-scene                \\
-scene non-ideal/unpleasant   & ni-scene               \\
+idéale/agréable                & i                      \\
+non-idéale/désagréable         & ni                     \\
+scène idéale/agréable          & i-scene                \\
+scène non-idéale/désagréable   & ni-scene               \\
 \hline
@@
 \vspace{0.5mm}
-\caption{Acronyms of features used in the analysis of the sensory experiments.}
+\caption{Acronyme des variables utilisées dans le cadre des expériences sensorielles.}
 \label{tab:acronyme}
@@
 
-\subsubsection*{Checking of the pleasantness of the synthesized scenes}
-
-First, in order to ensure the coherence of the data, we check that none of the ni- scenes gets a $\mathcal{A}_{scene}$ higher than one of a i- scene.
-Four ni- scenes don’t fullfill that constraint: they are thus removed, together with their corresponding i- scenes.
-As a consequence, 36 i -scenes and 36 ni -scenes remain for analysis.
-Second, we verify that participants really perceived a difference in terms of pleasantness between i- and ni -scenes.
+\subsubsection*{Cheking of the pleasantness of the synthesized scenes}
+
+First, in order to ensure the coherence of the data, wr check that none of the ni- scenes gets a $\mathcal{A}_{scene}$ higher than one of a i- scene.
+Four ni- scenes don’t respect that rule: they are removed, together with their corresponding i- scenes.
+36 i -scenes and 36 ni -scenes remain for analysis.
+Secondly, we want to be sure that participants really perceived a difference in terms of pleasantness between i- and ni -scenes.
 For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment.
@@
 
-First, our analyses focuses on the sound level.
-Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the levels $L$, $L(E)$ and $L(T)$.
+First, we focus on the sound level.
+Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} depict the levels $L$, $L(E)$ and $L(T)$.
 There is a significative difference in terms of sound level between i- and ni -scenes ($L$: $p<0.01$), with a standard deviation of -7 $dB$.
-This difference is siginificant for events ($L(E)$: $p<0.01$, STD: -7 $dB$) and for textures ($L(T)$: $p<0.01$, STD: -6 $dB$).
-
-As expected, the sound sources sound level is indeed a pleasantness indicator, as the ni -scenes tend to be louder, an outcome of a large number of related works.
+This difference is siginificant for events ($L(E)$: $p<0.01$, STD: -7 $dB$) and textures ($L(T)$: $p<0.01$, STD: -6 $dB$).
+
+As expected, we verify that the sound sources sound level is indeed a pleasantness indicator, the ni -scenes tending to be louder - what also comes out from a large number of related works.
 We also notice that this difference of sound level is equally true for events or textures.
@@
 This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}.
-During their experiment, the authors ask to their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source.
-The study shows that there is no significative difference between global and averaged instantaneous judgments.
+During their experiment, the authors ask to their participants to assess a set of soundscapes at a global level and then to do the same judgement at the moments when each of them detect a sound source.
+The study shows that there is no significative difference between global and averaged instantaneous judgements.
 In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, \ie the events.
 
-Though, the level alone is not sufficient to differentiate the type of environment.
-In fact, $20\%$ of the i- scenes have a sound level higher than the nominal level of the ni- scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
-
-\begin{figure*}[t]
+The level alone is not enough to differentiate the different types of environment.
+In fact, $20\%$ of the i- scenes hast got a sound level higher than the nominal level of the ni- scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
+
+\begin{figure}[t]
         \myfloatalign
@@
        \caption{Distributions of the structural features $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
-\end{figure*}
+\end{figure}
 
@@
 
-In this section, relationships that could exist between structural features and perceived pleasantness are investigated.
-Contrary to the previous section, we do not consider the pleasantness property of a scene on a binary mode (i vs.
-ni): this is the mean pleasantness $\mathcal{A}_{scene}$ that is considered here as a perceptual feature.
+In this section, we analyse the deep relationships that could exist between structural features and perceived pleasantness.
+Contrary to the previous section, we don’t consider the emotional property of a scene on a binary mode (i vs.
+ni): here, we consider the mean pleasantness $\mathcal{A}_{scene}$ as a perceptual feature.
 The aim is to investigate the level of correlation between structural features and $\mathcal{A}_{scene}$.
@@
 
-Concerning $L$, we can observe a strong negative correlation ($r=-0.77$, $p<0.01$) with $\mathcal{A}_{scene}$, indicating that the higher the sound level is, the more unpleasant the scene is perceived.
-Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship do not occur in the same way for i- and ni- scenes.
+COncerning $L$, we can observe a strong negative correlation ($r=-0.77$, $p<0.01$) with $\mathcal{A}_{scene}$, indicating that the higher the sound level is, the more unpleasant the scene is perceived.
+Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship doesn’t occurr in the same way for i- and ni- scenes.
 In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni- scenes ($r=-0.78$, $p<0.01$), but is quasi null for i-scenes.
 
-When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the i- scenes tend to be softer than the ni- scenes, thus allowing us to extend to the i-scenes the negative correlation observed for the ni- scenes.
+When considering the whole set of scenes, this result can be explained by the fact the i- scenes tends to be softer than the ni- scenes, deluding that the negative correlation observed for the ni- scenes is extended.
 
@@
 \begin{itemize}
-\item allows us to differentiate between i- and ni- scenes,
-\item characterizes precisely the perceived pleasantness for ni- scenes,
-\item is not a relevant feature for modeling the perceived pleasantness of \textit{a priori} pleasant soundscapes (i- scenes)
-\end{itemize}
-
-The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}.
-For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$).
-Again, we believe that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer than ni - scenes).
-Thus, the sound events keep a relative ability to predict the pleasantness of the ni- scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.
+\item allows to differentiate between i- and ni- scenes,
+\item allows a fine characterization of the perceived pleasantness for ni- scenes,
+\item is not a relevant feature for the perceived pleasantness for a priori pleasant soundscapes (i- scene)
+\end{itemize}
+
+The same conclusions can be drawn for $L(E)$ (\cf~Figure~\ref{fig:soundlevele}).
+For $L(T)$ (\cf~Figure~\ref{fig:soundlevelf}), the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$).
+One can suppose that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer than ni - scenes).
+Thus, the sound events keep a relative ability to predict the pleasantness of the ni- scenes, but the level of the textures doesn’t bring a lot of information whatever the type of environment is.
 
@@
 
-Those first outcomes tend to show that 1) two modes of perception exists depending on the very nature of the environment (i- or ni-), and 2) each involve independent features.
-The fact that $L$ is not sufficient to characterize the pleasantness of the i- scenes can lead us to conclude that all the sound sources do not equally contribute to the perception of pleasantness.
-Thus, we think that only the level of some of them can influence this percept.
-In order to investigate further in that direction, we analyze in the next section the scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.
+Those first outcomes tend to show that two modes of perception exists depending on the nature of the environment (i- or ni-), and each involve independent features.
+The fact that $L$ is not able to characterize the pleasantness of the i- scenes can lead us to think that all the sound sources don’t equally contribute to the perception of pleasantness, but that only the level of some of them can influence this percept.
+In order to go deeper in that direction, in the next section, we analyse these scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.
 
@@
 
-We study the composition of the scenes by counting the number of participants who used a given class of sounds to simulate the environment.
+We analyse the composition of the scenes by counting the number of participants who used a given class of sounds to simulate the environment.
 Results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
-For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$ is used to depict classes, see ~Figure~\ref{fig:taxonomie}.
-
-We observe a significant difference between i- and ni- scenes for the classes choice.
-The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~classes involving human presence and nature are prevailing in the i- scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni- scenes.
+For sake of clarity, we chose a transitional level of abstraction between level 0 and 1, named $0+$, to depict classes, see ~Figure~\ref{fig:taxonomie}.
+
+We observe a significant difference between i- and ni- scenes for the choise of classes.
+The distribution of the classes is very similar to to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~lasses involving human presence and nature are prevailing in the i- scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni- scenes.
 These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.
@@
 The authors interpret this by the fact that the perception of pleasantness is, among other things, due to socio-cultural factors.
-Thus, in our representation of the world, sounds of public transportation are positively connoted and tend to be more accepted than sounds of personal vehicles.
+In our representation of the world, sounds of public transportation are positively connoted and tend to be more accepted than sounds of personal vehicles.
 
 To a certain extent, our results are in contradiction with this assumption.
-In fact,  Figure~\ref{fig:soundsourcea} show that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the participants for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes.
-the results do not re-evaluate the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} pour les i-scenes, a level that is comparable to the of the \emph{bike} class, and much higher than all the personal vehicles classes.
+In fact,  Figure~\ref{fig:soundsourcea} show \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the participants for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes.
+the results don’t re-evaluate the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} pour les i-scenes, a level that is comparable to the of the \emph{bike} class, and much higher than all the personal vehicles classes.
 Though, \emph{public transport} classes are also strongly present in the ni-scenes, for instance more than \emph{light vehicle} or \emph{trucks} classes.
-On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.
-
-This difference may be explained by the nature of the experimental protocol used in the two studies.
-As in our study, Guastavino asks the participants for describing an environment starting from their memories whereas, in our case, participants perform the same task using actual sound samples.
+On the basis of our results, the public transportation class can’t be considered as typical of an ideal urban soundscape.
+
+This difference can be explained by the nature of the experimental protocol used in the two studies.
+As in our case, Guastavino asks the participants for describing an environment; but in that case, they work from their memories whereas, in our case, participants do that task using sound samples.
 The fact that participants in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact.
@@
         \subfloat[]
-        {\includegraphics[width=1\linewidth]{gfx/ch_5/xp1_class_1}\label{fig:soundsourcea}} \par
-        \subfloat[]
-        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_2}\label{fig:soundsourceb}}
-        \subfloat[]
-        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_3}\label{fig:soundsourcec}}
+        {\includegraphics[width=1\linewidth]{gfx/ch_5/xp1_class_1_cam}\label{fig:soundsourcea}} \par
+        \subfloat[]
+        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_2_cam}\label{fig:soundsourceb}}
+        \subfloat[]
+        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_3_cam}\label{fig:soundsourcec}}
        \caption{Proportion of synthesized scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.}\label{fig:soundsource}
@@
 
-\subsubsection*{Sound markers}
-
-We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs with respect to the type of environment (i or ni).
-Let us now investigate whether some of the sound classes are specific to a given environment.
-For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna} separately for each abstraction level.
-Results are presented in Table~\ref{tab:markers}.
-
-Regarding the sound events, 9 markers are identified for all abstraction levels.
-As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i- scenes markers as well as the \emph{church bell} class.
-This latter result may be due to the socio-cultural background of the participants who are mostly European citizens.
-In fact, according to Schafer, a sound that is identified by a person as being an element of his/her environment, is well accepted.
-Sound markers of ni- scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
-
-Regarding the textures, 5 markers are identified.
-For the i- scenes, they are classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}).
-For the ni- scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).
-
-Although the whole set of identified markers are rather intuitive, none of the event classes related to the noise of the motor vehicles are identified as markers, except for the texture class \emph{crossroad}.
-To simulate an unpleasant traffic, participants chose the classes \emph{horn} or \emph{siren}.
-We thus conclude that isolated motor vehicle sounds are understood as being part of the urban environment, and thus their nature is not necessarily linked to an unpleasant soundscape.
-
-\begin{table*}[t]
+\subsubsection*{Marqueurs sonores}
+
+Nous avons mis en évidence que, qualitativement, la composition des sources sonores des scènes diffère selon les types d'environnement (i ou ni).
+Nous essayons de voir maintenant si, parmi ces classes, certaines sont typiques d'un environnement en particulier.
+Pour ce faire, nous utilisons le V-test (\cf~Section~\ref{sec:xp1_dataAna}), en considérant séparément chaque niveau d'abstraction.
+Les résultats sont présentés dans le tableau~\ref{tab:markers}.
+
+Concernant les événements sonores, 9 marqueurs sont identifiés sur l'ensemble des niveaux d'abstraction.
+Comme la figure~\ref{fig:soundsource} le laissait présager, les classes relatives à la présence humaine (\emph{pas homme béton}, \emph{sonnette vélo}), et à la nature (\emph{animaux, oiseaux}, \emph{chants d'oiseaux}) sont des marqueurs de i-scènes.
+Nous notons également la présence de la classe \emph{cloche} dans les marqueurs d'un environnement idéal.
+Ce fait est possiblement dû au \emph{background} socio-culturel des sujets, dans leur grande majorité, des citoyens européens.
+En effet, selon Schafer, un son reconnu par un individu comme faisant partie intégrante de son environnement est bien accepté.
+Les marqueurs de ni-scènes sont des classes faisant référence à des sons de travaux (\emph{travaux}), ou suggérant un trafic dense (\emph{klaxon}, \emph{sirène}).
+
+Concernant les textures sonores, 5 marqueurs sont identifiés.
+Pour les i-scènes, il s'agit de classes faisant référence à des ambiances amorphes, calmes, (\emph{cour-intérieur} et \emph{parc}).
+Pour les ni-scènes, il s'agit, comme pour les événements, de classes faisant référence à des bruits de travaux (\emph{travaux} et \emph{véhicule de travaux}), ainsi que d'une classe faisant référence au trafic (\emph{carrefour}).
+
+Bien que l'ensemble des marqueurs identifiés soient intuitifs, aucune des classes d'événements faisant directement référence aux bruits de véhicules motorisés n'est un marqueur, exception faite de la classe de textures \emph{carrefour}.
+Pour représenter un trafic désagréable, les sujets ont porté leurs choix sur les classes \emph{klaxon} et \emph{sirène}.
+On peut supposer que les sons isolés de véhicules sont compris comme faisant partie intégrante de l'environnement urbain, et ne sont donc pas particulièrement associés à un environnement désagréable.
+
+\begin{table}[t]
  \setlength{\tabcolsep}{0.2pt}
@@
 \begin{tabular}{c c c c}
-Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
-level & i-scenes & ni-scenes \\
+Niveau        & \multicolumn{2}{c}{Marqueurs sonores événements} \\
+d'abstraction & i-scènes & ni-scènes \\
 \hline
-0  &                          & construction work (3.78)  \\
+0  &                          & travaux (3.78)  \\
 \hline
-  & church bell  (4.5)             & car horn  (3.9) \\
-1 & bicycle bell  (4.3)      & siren (3.9)\\
+  & cloche  (4.5)             & klaxon  (3.9) \\
+1 & sonnette vélo  (4.3)      & sirène (3.9)\\
   & animal (4.2)              &       \\
    \hline
-  & bird        (4.8)       & car horn  (4.0)\\
-2 & church bell  (4.4)             & siren (4.0)\\
-  & bicycle bell     (4.2)             &       \\
+  & oiseau        (4.8)       & klaxon  (4.0)\\
+2 & cloche  (4.4)             & sirène (4.0)\\
+  & sonnette vélo    (4.2)             &       \\
    \hline
-  & bird song (4.8)        & klaxon  (4.1)\\
-3 & church bell   (4.3)            & siren (4.0)\\
-  & bicycle bell      (4.2)   &       \\
-  & foot steps  (3.6)      &  \\
+  & chant oiseau (4.8)        & klaxon  (4.1)\\
+3 & cloche   (4.3)            & sirène (4.0)\\
+  & sonnette vélo     (4.2)   &       \\
+  & pas chaussure  (3.6)      &  \\
   &                           & \\
-  & \multicolumn{2}{c}{Texture sound markers}      \\
-  & i-scenes & ni-scenes \\
+  & \multicolumn{2}{c}{Marqueurs sonores textures}      \\
+  & i-scènes & ni-scènes \\
 \hline
-0 &     courtyard / park (4.1) &  construction (3.9)  \\
+0 &     cour int./parc (4.1) &  travaux (3.9)  \\
 \hline
-1 &     park (3.65)          &  crossroad (3.6)  \\
-  &                          &  construction vehicule (3.3)  \\
+1 &     parc (3.65)          &  carrefour (3.6)  \\
+  &                          &  travaux véhicule (3.3)  \\
 \hline
-2 &     park (3.64)          &  crossroad (3.56)  \\
+2 &     parc (3.64)          &  carrefour (3.56)  \\
 \hline
@@
 \vspace{0.5mm}
-\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of $V$ value.}
+\caption{Classes d'événements et de textures identifiées comme étant des marqueurs sonores. Dans chaque cellule, les marqueurs sont ordonnés par ordre décroissant de valeur $V$.}
 \label{tab:markers}
-\end{table*}
-
-\subsubsection*{Representation space induced by semantic features}
-
-In this section, we evaluate the ability of a semantic representation of the scenes to discriminate between the two types of environment.
-For that, we consider a rank 5-precision ($p@5$) on the space induced by the semantic features $S$, and for each abstraction level, see Section~\ref{sec:xp1_dataAna}.
-The vectors $S$ are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers ($E_{w/o,m}$).
-Texture classes corresponding to sound markers are not considered as these latter are too few to reliably compute the metric.
-For the same reasons, event classes corresponding to sound markers at abstraction level $0$ are also discarded.
-Results are shown on Figure~\ref{fig:pa5}.
-
-Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and stay above $86\%$ from the abstraction level $1$.
-These results confirm that it is possible to make a clear distinction between the two types of environment on the sole basis of presence / absence of sound classes.
-We also notice that the higher the abstraction level is, the more important the ability to make this distinction is.
-In other words, the more precise in the scene description is, the more clear the discrimination between i- and ni- scenes is.
-
-Considering $E$ and $T$ separately, it appears that: 1) the $p@5$ obtained with $E$ is similar to the one obtained with $ET$ ; 2) the $p@5$ obtained with $T$ is systematically lower than the one obtained with $E$, by $10\%$ or $15\%$.
-Those results indicate that the discriminative semantic information is mainly contained in the events.
-These results are in line with the work done in \cite{maffiolo_caracterisation_1999} showing that humans analyze the event scenes which are composed of several sound events in a causal manner, \ie by identifying the sources, as discussed in Section~\ref{sec:simscene_sampleDataSet}.
-
-The dimension of the vectors $S$ for $E_m$ is lower than the dimension of vector $S$ for $E$, itself lower than the dimension of vector $S$ obtained when all the classes are considered ($ET$).
-Despite this, it appears that the $p@5$ obtained with $E_m$ is equal – or above – to the ones obtained with $E$ or $ET$, although a partial information is used in that case to describe the scenes.
-Moreover, in the case when the sound markers are not taken into account for the description ($E_{w/o,m}$), the performance is under those obtained with only the textures as features.
-This confirms that most of the semantic information allowing to differentiate between i- and ni- scenes is included in the markers.
-
-To sum up, we deduce the following points from this analysis:
+\end{table}
+
+\subsubsection*{Espaces de représentation induits par les descripteurs sémantiques}
+
+Dans cette partie, nous évaluons la capacité d'une représentation sémantique à séparer les deux types d'environnement.
+Pour ce faire, nous calculons une précision au rang 5 ($p@5$) sur l'espace induit par les descripteurs sémantiques $S$, et ce pour chaque niveau d'abstraction (\cf~Section~\ref{sec:xp1_dataAna}).
+Les vecteurs $S$ sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.
+Nous ne considérons pas les classes de marqueurs de textures, ces dernières étant trop peu nombreuses.
+Pour les mêmes raisons nous ne considérons pas les classes de marqueurs d'événements du niveau d'abstraction $0$.
+Les résultats sont affichés sur la figure~\ref{fig:pa5}.
+
+En ce qui concerne $ET$, la $p@5$ est de $76\%$ pour le niveau d'abstraction 0, et reste supérieure à $86\%$ à partir du niveau d'abstraction 1.
+Ces résultats confirment qu'il est possible de clairement distinguer les deux types d'environnement en se basant seulement sur la présence ou l'absence des classes de sons.
+Nous notons également que, plus le niveau d'abstraction est élevé, plus la capacité de séparer les environnements est importante.
+En d'autres termes, plus nous sommes précis dans notre description de la composition des scènes, plus nous sommes à même d'établir une distinction claire entre les i- et ni-scènes.
+
+En considérant séparément $E$ et $T$, il apparaît 1) que la $p@5$ obtenue avec $E$ est similaire à celle obtenue avec $ET$, 2) que la $p@5$ obtenue avec $T$ est systématiquement inférieure d'environ $10$ à $15\%$ à celle de $E$.
+Ces résultats indiquent que l'information sémantique permettant de séparer les deux environnements est principalement portée par les événements.
+Ces résultats font, par ailleurs, écho aux travaux de  \cite{maffiolo_caracterisation_1999}, qui montrent que nous analysons de manière descriptive (en identifiant les sources) les scènes événementielles, \ie~composées d'événements sonores (\cf~Section~\ref{sec:simscene_sampleDataSet}).
+
+Enfin, il apparaît que la $p@5$ obtenue avec $E_{m}$ est égale, voire supérieure à celles obtenues avec $E$ et $ET$, et ce bien qu'une information partielle soit utilisée dans ce cas pour décrire les scènes.
+La dimension des vecteurs de description $S$ pour $E_m$ est en effet inférieure à la dimension des vecteurs $S$ pour $E$, qui est elle même inférieure à celle obtenue dans le cas où toutes les classes sont utilisées ($ET$).
+De plus, dans le cas où les marqueurs ne sont pas pris en compte pour la description ($E_{w/o,m}$), les résultats chutent, passant même en dessous de ceux obtenus en ne considérant que les textures.
+Cela confirme que la majorité de l'information sémantique permettant de faire la distinction entre i-scènes et ni-scènes est incluse dans les marqueurs.
+
+En résumé, nous déduisons de cette analyse les points suivants:
 
 \begin{enumerate}
-\item unlike what we outlined with the structural features, a semantic description of the scenes constitution in terms of presence / absence of sound sources, allows us to differentiate the two types of environments (i or ni);
-\item the semantic information is mainly contained in the event sound classes;
-\item only a part of the event classes, i.e. the sound markers, are useful to differentiate between the i- and ni- scenes.
+\item contrairement à ce que nous avions constaté avec les descripteurs structurels, une description sémantique de la composition des scènes, en terme de présence/absence de sources sonores, permet de bien distinguer les deux types d'environnement (i ou ni);
+\item l'information sémantique est majoritairement portée par les classes d'événements sonores;
+\item parmi les classes d'événements, seule une partie, \ie~les marqueurs sonores, est nécessaire pour faire la distinction entre les i- et ni-scènes.
 \end{enumerate}
 
-Since we have extracted the typical classes of the i- and ni- scenes and verified that the distinction between them was largely dependent on the presence of these classes, we shall now investigate whether a structural description of the scenes only based on sound markers could characterize the perceived pleasantness, perhaps better than a global structural description.
+Maintenant que nous avons isolé les classes typiques des i- et ni-scènes, et vérifié que la distinction entre ces environnements dépendait de la présence de ces classes, il reste à voir si une description structurelle des scènes, basée uniquement sur ces marqueurs sonores, permet de caractériser l'agrément perçu, mieux qu'une description structurelle globale.
 
@@
         \myfloatalign
-        \includegraphics[width=\linewidth]{gfx/ch_5/pa5_1_en}
-       \caption{$P@5$ obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors as a function of the abstraction level.
-The vectors are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers $E_{w/o,m}$.}\label{fig:pa5}
+        \includegraphics[width=\linewidth]{gfx/ch_5/pa5_1}
+       \caption{$P@5$ obtenues en considérant la matrice de dissimilarité résultant des distances par paires de Hamming calculées entre les vecteurs des descripteurs sémantiques des scènes.
+Les vecteurs sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.}\label{fig:pa5}
 \end{figure}
 
-\subsubsection*{Influence of the sound markers on the perceived pleasantness}
-
-We evaluate the correlations between $\mathcal{A}_{scene}$ and the structural features.
-In this section, the structural features are computed by taking into account only the previously identified sound markers.
-We define $Xm$, the $X$ feature computed by taking into account only the sound markers.
-On the opposite, we define $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
-When the feature characterizes an i- scene (idem for a ni- scene),  only the markers identified for the i- scenes (or the ni- scenes) are considered; we named them, i- markers (or ni- markers).
-Results are shown in Table~\ref{tab:corrMarkers}.
-
-Concerning the sound levels, the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ et $L(T)$ on the other side, see~Figures~\ref{fig:soundlevelMarkera} and~\ref{fig:soundlevelMarkerd}, respectively.
-No matter whether all the classes or only the markers are considered, it appears that:
+\subsubsection*{Influence des marqueurs sonores sur l'agrément perçu}
+
+Nous évaluons les corrélations entre $\mathcal{A}_{scene}$ et les descripteurs structurels.
+Pour cette section, les descripteurs structurels sont calculés en tenant compte des marqueurs sonores précédemment identifiés.
+Nous définissons $X_m$ le descripteur $X$ calculé en ne prenant en compte que les sons des marqueurs.
+A l'inverse, nous définissons $X_b$ ($b$: pour «~bruit~») le descripteur $X$ calculé en prenant en compte toutes les classes de sons, excepté les marqueurs.
+Lorsque le descripteur caractérise une i-scène (idem pour une ni-scène), nous ne considérons, pour le calcul, que les marqueurs identifiés pour les i-scènes (ou pour les ni-scènes), que nous nommons i-marqueurs (ou ni-marqueurs).
+Les résultats sont affichés sur le tableau~\ref{tab:corrMarkers}.
+
+Concernant les niveaux sonores (\cf~Figures~\ref{fig:soundlevelMarkera} et~\ref{fig:soundlevelMarkerd}), les mêmes tendances sont observées entre $L_m$, $L(E)_m$ et $L(T)_m$, d'une part, et $L$, $L(E)$ et $L(T)$, d'autre part.
+Que l'on considère uniquement les marqueurs, ou l'ensemble des classes, il s'avère que :
 
 \begin{enumerate}
-\item a significative difference between levels of i- and ni- scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
-\item the sound level of scenes is mainly related to the sound events, compared to the textures;
-\item the sound level of events has an influence on the perception of pleasantness for ni- scenes, but not for i- scenes;
-\item the sound level of textures does not play any role in the perception of the pleasantness.
+\item il existe une différence significative entre les niveaux des i- et ni-scènes ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
+\item le niveau sonore des scènes est majoritairement porté par les événements sonores, comparé aux textures sonores;
+\item le niveau sonore des événements a une influence sur la perception de l'agrément pour les ni-scènes, mais pas pour les i-scènes;
+\item le niveau sonore des textures ne joue aucun rôle dans la perception de l'agrément.
 \end{enumerate}
 
-In conclusion, the level of ni- markers has a negative influence on the pleasantness for the ni- scenes.
-On the other hand, the level of i- markers does not influence the perceived pleasantness for the i- scenes.
-
-% En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
+En conclusion, le niveau des ni-marqueurs a une influence négative sur l'agrément pour les ni-scènes. En revanche le niveau des i-marqueurs n’impacte pas l'agrément perçu pour les i-scènes.
+
+En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
 C'est la première fois qu'un indicateur objectif nous permet de préciser l'agrément des environnements agréables.
@@
 
-Considering the non markers classes, we can observe on the i- scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}.
-It is the first time that an objective feature allows us to define the pleasantness of the pleasant environments.
-This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.
-
-% Par ailleurs, alors que $L(T)$ ne présentait pas de corrélation pour les ni-scènes, une corrélation négative forte est observée pour $L(T)_b$ ($r-0.73$, $p<0.01$).
+Par ailleurs, alors que $L(T)$ ne présentait pas de corrélation pour les ni-scènes, une corrélation négative forte est observée pour $L(T)_b$ ($r-0.73$, $p<0.01$).
 Ce fait indique que les niveaux des classes de textures n'étant pas des marqueurs n'affectent pas l'agrément perçu de la même manière pour les i- et ni-scènes.
@@
 
-Besides, whereas $L(T)$ did not show any significative correlation for ni- scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
-This indicates that the level of non-markers texture classes does not influence the perceived pleasantness in the same way for i- and ni- scenes.
-Sound level seems to have a negative effect for the ni- scenes, and no significant effect for the i- scenes.
-
-% Pour finir, nous considérons un dernier groupe de descripteurs, nommément $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiffa} et~\ref{fig:soundlevelMarkerDiffd}).
+Pour finir, nous considérons un dernier groupe de descripteurs, nommément $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiffa} et~\ref{fig:soundlevelMarkerDiffd}).
 Ces descripteurs expriment la différence entre les niveaux des marqueurs, et ceux des autres classes de sons.
@@
 
-We consider a last group of features, namely $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$.
-These features describe the difference between the markers level and those of the other sound classes, see ~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}).
-They express the saliency of the markers with respect to the sound mixture.
-
-% Pour les i-scènes, une corrélation modérée et positive est observée pour $L_m-L_b$ ($r=0.67$, $p<0.01$) et $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
+Pour les i-scènes, une corrélation modérée et positive est observée pour $L_m-L_b$ ($r=0.67$, $p<0.01$) et $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
 Pour les ni-scènes, aucune corrélation n'est observée.
@@
 
-For the i- scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
-for the ni- scenes no correlation is observed.
-Thus, for the i- scenes, it is not the absolute markers level that is important, but their relative level with regards to the other sounds composing the scene.
-Then, we can induce a double perceptual mechanism for the ideal environments:
-
-\begin{itemize}
-\item the higher the absolute level of sounds not being i- markers, the weaker the pleasantness,
-\item the higher the relative level of i- markers is, with regards to the other sounds, the higher the pleasantness.
-\end{itemize}
-
-For the ni -scenes, the fact that we observe significative correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.
+\begin{itemize}
+\item plus le niveau absolu des sons n'étant pas des i-marqueurs est élevé, plus l'agrément est faible,
+\item plus le niveau relatif des i-marqueurs, par rapport aux autres sons, est élevé, plus l'agrément est élevé.
+\end{itemize}
+
+Pour les ni-scènes, le fait que nous observions des corrélations pour $L_m$ et $L(E)_m$, et aucune pour $L_m-L_b$ et $L(E)_m-L(E)_b$, montre que c'est bien le niveau absolu qui importe.
 
@@
 \vspace{0.5mm}
-\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ (Exp. 1.b) and structural features related to the presence of sound markers.}
+\caption{Coefficients de corrélation linéaire calculés entre l'agrément perçu moyen $\mathcal{A}_{scene}$ de l'expérience 1.b et les descripteurs structurels relatifs à la présence des marqueurs sonores.}
 \label{tab:corrMarkers}
@@
 
-\begin{figure*}[t]
+\begin{figure}[t]
         \myfloatalign
@@
         {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
-        \caption{Distribution of the relative sound levels structural features related to the presence of markers $L_m$ (a, d), $L(E)_m$ (b, e) and $L(T)_m$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
-\end{figure*}
-
-\subsection{Discussion}
-
-From this analysis, the following points can be outlined:
-
-\begin{itemize}
-\item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global structural features ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni- scenes.
-The semantic description seems to be more powerful;
-\item \emph{events or textures}: whatever the feature type, be it semantic or structural, events are mainly useful to differentiate the two types of events; textures bring a limited amount of information;
-\item \emph{pleasantness prediction}: considering the correlation between structural features and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its nature (i or ni).
-From the data gathred in those experiments, it is not conceivable to consider the same set of features to predict the pleasantness of i- and ni- scenes in the same time:
-
-\begin{itemize}
-
-\item \emph{for ni-scenes}, global levels ($L$ et $L(E)$), or level of sound markers ($L_m$ et $L(E)_{m}$), have a negative influence on pleasantness.
-We can notice here that taking into account the contribution of each of the different sources of the scene does not improve the prediction performance compared to a global analysis of the environment.
-
-\item \emph{for i-scenes}, on contrary, the pleasantness prediction needs to consider separately the sound markers characteristics and those of the other sounds.
-Thus, the markers level, relative to the noise ($L(E)_m-L(E)_b$ and $L_m-L_b$) is positively correlated to the pleasantness, whereas the noise level ($L_b$ and $L(E)_b$) is negatively correlated.
-\end{itemize}
-\end{itemize}
-
-The fact that the pleasantness of the i- scenes is not correlated to global physical features, contrary to pleasantness of ni- scenes has been studied recently \cite{gozalo2015relationship}.
-
-We thus observe two perceptual modes of operation that involve different types of features, and that rely on the hedonic nature  of the stimuli.
-A phenomenon similar is observed for the perception of textures, see Section~\ref{sec:simscene_sampleDataSet}.
-It seems that the brain adapts its way to encode the information (statistic summary for textures, fine description for events) following a previous decision making with regards to the nature of the stimuli, \ie is it an event or a texture ?.
-Similarly, in the light of this study, it appears that the features considered for the pleasantness judgment also depends on a preliminary judgment of the global hedonic nature of the environment (ideal or non ideal).
+        \caption{Dispersions des descripteurs structurels de niveaux sonores relatifs à la présence des marqueurs $L_m$ (a, d), $L(E)_m$ (b, e) et $L(T)_m$ (c, f), en fonction du type de scènes (a, b, c) et de l'agrément perçu $\mathcal{A}_{scene}$ de l'expérience 1.b (d, e, f).}\label{fig:soundlevelMarker}
+\end{figure}
+
+\subsection{Discussions}
+
+De cette analyse, nous retenons les points suivants:
+
+\begin{itemize}
+\item \emph{distinguer les i- et ni-scènes}: les descripteurs sémantiques, ainsi que les descripteurs structurels globaux ($L$, $L(E)$ et $L(T)$), permettent de faire la distinction entre les i-scènes et les ni-scènes.
+La description sémantique semble être plus performante;
+\item \emph{événements ou textures}: que ce soit pour les descripteurs sémantiques ou structurels, ce sont majoritairement les événements qui permettent de distinguer les deux types d'environnement, les textures n'apportant, au mieux, qu'une information limitée;
+\item \emph{prédire l'agrément}: si l'on considère les corrélations entre les descripteurs structurels et l'agrément, il semble que la manière de percevoir la qualité de l'environnement diffère en fonction de la nature de ce dernier (i ou ni).
+Il n'apparaît pas envisageable de considérer un même jeu de descripteurs pour prédire, à la fois, l'agrément des i-scènes, et l'agrément des ni-scènes :
+
+\begin{itemize}
+
+\item \emph{pour les ni-scènes}, ce sont le niveau global ($L$ et $L(E)$), ou le niveau des marqueurs sonores ($L_m$ et $L(E)_{m}$), qui impactent négativement l'agrément.
+On note ici que prendre en compte les contributions de différentes sources n'améliore pas la capacité de prédiction de l'agrément, par rapport à une analyse holistique de l'environnement.
+
+\item \emph{pour les i-scènes}, par contre, prédire l'agrément requiert d'étudier, de manière séparée, les caractéristiques des marqueurs sonores, et celles de l'ensemble des autres sons.
+Ainsi, le niveau des marqueurs pris relativement au bruit ($L(E)_m-L(E)_b$ et $L_m-L_b$) est positivement corrélé à l'agrément, alors que le niveau du bruit ($L_b$ et $L(E)_b$) est, lui, négativement corrélé.
+\end{itemize}
+\end{itemize}
+
+Le fait que l'agrément des i-scènes ne se corrèle pas avec des descripteurs physiques holistiques, contrairement à l'agrément des ni-scènes, a récemment été déjà observé \cite{gozalo2015relationship}.
+
+L'existence de deux modes de perception, mobilisant différents types de descripteurs, et dépendant de la nature (dans notre cas hédonique) des stimuli, est un phénomène qui se rapproche de celui observé pour la perception des textures (\cf~Section~\ref{sec:simscene_sampleDataSet}).
+Le cerveau adapte sa manière de traiter l'information (résumé statistique pour les textures, description fine pour les événements) suite à une prise de décision antérieure quant à la nature du stimulus (à savoir «~est-ce un événement ou une texture ?~»).
+De la même manière, les indicateurs actifs dans le jugement de l'agrément dépendent, eux aussi, d'une identification préalable de la nature hédonique globale de l'environnement (idéale ou non idéale).
 
 \section{Experiment 2: modification of the semantic content}
-\label{sec:modification}
+\label{sec:xp3}
 
@@
 
-The previous experiment demonstrated that, among the classes of sounds occurring in urban soundscapes, those with markers are specific to some environments.
+The previous experiment demonstrated that, among the classes of sounds occurring in urban soundscapes, those with markers are specific to some of environments.
 Those sound markers seem to have a great impact on perception.
@@
 
-In order to verify that the pleasantness of the ideal and non ideal scenes is affected by the sound markers, the audio waveform of the simulated scenes are regenerated with or without using the classes identified as markers.
+In order to verify that the pleasantness of the ideal and non ideal scenes if affected by the sound markers, the audio waveform of the simulated scenes are regenerated with or without considering the classes identified as markers.
 For the i-scenes, some of the i-markers are removed.
 For the ni-scenes, some of the ni-markers are removed.
-To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptive test is then conducted with a protocol close to the one considered in experiment 1.b.
+A perceptive test evaluating the pleasantness is then conducted with a protocol close to the one considered in experiment 1.b.
 
@@
 
-The objective of this experiment is to study if the removal of the previously identified markers have an impact on the perceived pleasantness. Two hypothesis are thus formulated:
+Thus the objective of this experiment is to study if the removal of the previously identified markers have an impact on the perceived pleasantness. Two hypothesis are thus formulated:
 
@@
 If the first hypothesis is rather intuitive, the second is less.
-Indeed, it is not obvious that the removal of the i-markers, though perceptively positively connoted, decreases the global quality of a soundscape.
-Indeed, this removal also decreases the global level of the scene.
+Indeed, it is not obvious that the removal of the i-markers, though perceptively connoted, decreases the global quality of a soundscape.
+This removal also decreases the global level of the scene.
 
@@
 \item \emph{72 km-scenes}: the 72 scenes originally simulated by the subject of experiment 1.a where the sound classes identified as markers are kept.
-The 36 ideal scenes with markers are noted i/km-scenes, and the 36 non ideal scenes with markers are noted ni/km-scenes (k for kept).
+The 36 ideal scenes with markers are noted i/km-scenes, and the 36 non ideal scenes with markers are noted ni/km-scenes.
 \item \emph{72 rm-scenes}:  72 scenes where the sound classes identified as markers are removed.
-The 36 ideal scenes without markers are noted i/rm-scenes, and the 36 non ideal scenes without markers are noted ni/rm-scenes (r for removed).
+The 36 ideal scenes without markers are noted i/rm-scenes, and the 36 non ideal scenes without markers are noted ni/rm-scenes.
 \end{itemize}
@@
 
-In order to create km-scenes with still some sound diversity and no  absence of sound activity within a long period of the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}.
+In order to create km-scenes with still some sound diversity and no long duration absence of sound activity within the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}.
 Those classes are:
@@
 Those calibration scenes are  5 i/km-scenes and 5 ni/km-scenes.
-These first 10 scenes are replayed at the end of the experiment, and only the scores given at the second occurrence are taken into account.
+These first 10 scenes are replayed at the end of the experiment, and only the notes given at the second occurrence are taken into account.
 
@@
 
-All the subjects perform the test on the same type of computers. They listen to the stimuli with semi-open \emph{Beyer-Dynamic DT 990 Pro} headphones. The sound level is the same for all subjects.
+All the subjects perform the test on the same type of computers. Thy listen to the stimuli with semi-open \emph{Beyer-Dynamic DT 990 Pro} headphones. The sound level is the same for all subjects.
 
@@
 
-A supervisor is available during the whole duration of the experiment to monitor the experiment and answer any questions the subjects may have.
+A supervisor is available during the whole duration of the experiment to ensure a smooth running of the latter and answer any questions the subjects may have.
 
@@
 The two independent variables are considered as \emph{within-subject} factors.
-The factors being of only two levels each (type: i/ni,  marker: km/rm), the sphericity hypothesis does not need to be checked.
-\emph{Post hoc} analyses are done using the Tukey-Kramer procedure.
+The factors being of only two levels each (type: i/ni; marker: km/rm), the sphericity hypothesis does not need to be checked.
+\emph{Post hoc} analysis are done by following the Tukey-Kramer procedure.
 
@@
 Let us consider $\mathcal{A}_{sujet}$ for the km-scenes.
-Close inspection of the answers show that one subject strongly differs with the ones of others.
+Close inspection that the answers given by one subject strongly differs with the ones of others.
 This subject evaluated positively close to half of the ni/km-scenes, see Annex~\ref{app:xp2}, subject 7.
@@
 Furthermore, this subject used the whole scale (from -5 to 5) to score both the i-scenes and the ni-scenes.
-Those behaviors strongly differ with the ones of the others subjects be they from this experiment or the previous ones.
-Subject 7 is thus discarded from the analysis.
+Those behaviors strongly differing with the ones of the others subjects be they from this experiment or the previous ones, subject 7 is discarded from the analysis.
 
@@
 
-In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes and ni/rm-scenes.
+In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes et ni/rm-scenes.
 The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ show a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
@@
 
-The \emph{post hoc} analysis exhibits significative differences between all the groups of observations, notably between the i/km-scenes ($p<0.05$) and the i/rm-scenes ($p<0.01$).
+The \emph{post hoc} analysis exhibit significative differences between all the groups of observations, notably between the i/km-scenes ($p<0.05$) and the i/rm-scenes ($p<0.01$).
 
@@
 The fact that their removal decreases $\mathcal{A}_{scene}$ clearly shows that it is possible to improve the perceived pleasantnes of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
-Those conclusions are in line with the positive approach introduced by Schafer in \cite{schafer1977tuning}.
+Those conclusions are in line with the positive approacj introduced by Schafer in \cite{schafer1977tuning}.
 
@@
 
-\section{Outcomes for soundscape perception}
-\label{sec:conclusion}
+\section{Outcomes for sounscape perception}
+
 %Les expériences ont montré que la majorité des descripteurs utilisés, qu'ils soient sémantiques ou structurels, permettent de faire la distinction entre une scène idéale et une scène non-idéale.
@@
 The compositional characteristics that are taken into account during the perceptual process appear to vary from one source to the other, from one type of environment to the other.
-This fact lead the authors to believe that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.
+This fact lead the authors to conclude that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.
 
@@
 \item in the case of non ideal scenes, one should focus on reducing the acoustic pressure level, whether globally, or by discarding specific sources such as \emph{sirens} or \emph{car horns}
-\item in the case of ideal scenes, one should first identify which sources are pleasant to the targeted community, second lower the volume of the other sound sources, and, if possible, raise the contribution or add positive sound markers.
+\item in the case of ideal scenes, one should first identity which sources are pleasant to the targeted community, second lower the volume of the other sound sources, and, if possible, raise the contribution or add positive sound markers.
 \end{itemize}
@@
 
-The outcomes of the three experiments described in this paper demonstrate the usefulness of considering a dedicated simulation tool such as \emph{simScene} in order to scientifically question the perception of soundscapes in an innovative way.
-We also believe that its wider usage could enable urban planning decision-makers to question an entire community about its own representations of the sound environment to which they are exposed, and about the representations of the sound environments to which they would like to be exposed.
+The outcomes of the three experiments described in this paper demonstrate the usefulness of considering simulation tools such as \emph{simScene} in order to scientifically question the perception of soundscapes.
+We also believe that its wider usage could enable urban planning decision-makers to question an entire community about its own representations of the sound environment to which it is exposed, and about the representations of the sound environments to which it would like to be exposed.
 
@@
 
-Future work should consider several other simulation experiments by varying the emotional qualities (quiet, comfortable, troublesome, etc.), but also by specifying specific urban places (park, square, street, etc.), in order to provide to the scientific community an entire corpora of cognitively informed soundscapes.
+Future work should consider several other simulation experiments by varying the emotional qualities (quiet, comfortable, troublesome, etc.), but also by specifying particular places (park, square, street, etc.), in order to provide to the scientific community entire corpora of cognitively informed soundscapes.
 
@@
 Finally, one should study the influence of socio-cultural contexts on perception.
-Indeed, if the sound of the church bell is most often a quality environment marker for a Westerner, this do not necessarily hold true for subjects of Eastern, Middle Eastern or other cultures.
+Indeed, if the sound of the church bell is most often a quality environment marker for a Westerner, this is not necessarily true with subjects of Eastern, Middle Eastern or other cultures.
 
@@
 \item The simulator can be deployed on a large scale via the Internet thanks to the chosen software architecture;
-\item simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
+\item simulated scenes can be analyzed without having to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
 \end{itemize}
@@
 \appendix
-\section{Taxonomy of sound classes}
+\section{Taxonomies des classes de sons}
 \label{app:taxonomie}
@@
         \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_1_en}\label{fig:taxonomieEventa}}
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_2_en}\label{fig:taxonomieEventb}}\par
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_1_en}\label{fig:taxonomieTexturea}}
-        \subfloat[]
-        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_2_en}\label{fig:taxonomieTextureb}}
-       \caption{Taxonomy of sound classes used for the simulation urban soundscapes: events (a,b) and textures (c,d).}
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_1}\label{fig:taxonomieEventa}}
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/event_2}\label{fig:taxonomieEventb}}\par
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_1}\label{fig:taxonomieTexturea}}
+        \subfloat[]
+        {\includegraphics[width=.5\columnwidth]{gfx/ch_5/texture_2}\label{fig:taxonomieTextureb}}
+       \caption{Taxonomies des classes de sons utilisées pour la simulation des environnements sonores urbains : (a,b) événements sonores et (c,d) textures sonores.}
        \label{fig:taxonomie}
@@
 \newpage
-\section{Distribution of scores given par the subjects during experiment 2}
+\section{Dispersions des notes données par les sujets lors de l'expérience 2}
 \label{app:xp2}
@@
         \myfloatalign
-        \subfloat[subject 1]
+        \subfloat[sujet 1]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_1}\label{fig:xp4_note_1a}}
-        \subfloat[subject 2]
+        \subfloat[sujet 2]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_2}\label{fig:xp4_note_1b}}
-        \subfloat[subject 3]
+        \subfloat[sujet 3]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_3}\label{fig:xp4_note_1c}}
-        \subfloat[subject 4]
+        \subfloat[sujet 4]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_4}\label{fig:xp4_note_1d}} \par
-        \subfloat[subject 5]
+        \subfloat[sujet 5]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_5}\label{fig:xp4_note_1e}}
-        \subfloat[subject 6]
+        \subfloat[sujet 6]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_6}\label{fig:xp4_note_1f}}
-        \subfloat[subject 7]
+        \subfloat[sujet 7]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_7}\label{fig:xp4_note_1g}}
-        \subfloat[subject 8]
+        \subfloat[sujet 8]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_8}\label{}} \par
-        \subfloat[subject 9]
+        \subfloat[sujet 9]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_9}\label{fig:xp4_note_1i}}
-        \subfloat[subject 10]
+        \subfloat[sujet 10]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_10}\label{fig:xp4_note_1j}}
-        \subfloat[subject 11]
+        \subfloat[sujet 11]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_11}\label{fig:xp4_note_1z}}
-        \subfloat[subject 12]
+        \subfloat[sujet 12]
         {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_12}\label{fig:xp4_note_1l}}
-        \caption{Distribution of scores given par the subjects during experiment 2 for i/am-scenes (green) et ni/am-scenes (red).}\label{fig:xp4_note_1}
+        \caption{Dispersions des notes données par les sujets lors de l'expérience 2 aux i/am-scènes (vert) et ni/am-scènes (rouge).}\label{fig:xp4_note_1}
 \end{figure}

### 2017-08-23 lafaySimulatedScenePerception.tex
on track

--- 2017-07-31 lafaySimulatedScenePerception.tex
@@
 \newcommand{\ml}[1]{\textcolor{red}{#1}}
-\newcommand{\jlc}[1]{\textcolor{green}{#1}}
+\newcommand{\jfp}[1]{\textcolor{green}{JFP : #1}}
 \newcommand{\ie}{\emph{i.\,e.}}
@@
 
-    \author{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J.F. Petiot$^{1)}$}
+    \author{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J-F. Petiot$^{1)}$}
 
@@
 
-First, a subject can be asked to describe a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}.
+First, a subject can be asked to assess a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}.
 The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}.
-That being said, those stimuli respectively are or can be recorded and analysed.
-It thus allows the experimenter to gather a coarse grain physical description of the stimuli.
+That being said, those recorded stimuli  can be analysed, what allows the experimenter to gather a coarse
+grain physical description of them.
 
@@
 Second, a subject can be asked to describe a given sound environment \cite{guastavino2006ideal, dubois2006cognitive}.
-A large amount of quantitative and semantic information is then collected about the representation of this type of sound environment that the subject has.
+A large amount of quantitative and semantic information is then collected about the subject's representation of this type of sound environment.
 Unfortunately, without any reference to sound data, this representation can hardly be characterized physically.
@@
 
-For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sounds sources that compose a sound environment of interest.
+For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment of interest.
 With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the events.
@@
 Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.
-Experiment 1.a is based on a pilot study described in \cite{lafay2014new}.
 
@@
 
-\emph{Simscene} is an online digital audio tool whose first version have been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
+\emph{Simscene} is an online digital audio tool whose first version has been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
 It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}.
@@
 A track is then created for this sound source within the simulator interface.
-The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sounds source.
+The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sound sources.
 Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.
@@
 
-The literature usually distinguish between:
+The literature usually distinguishes between:
 \begin{itemize}
@@
 
-The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures only.
+The simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with the classes of events only and the other with classes of textures only.
 These two types also have specific scheduling procedures during the simulation process.
@@
 
-Maffiolo \cite{maffiolo_caracterisation_1999} distinguish two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events.
+Maffiolo \cite{maffiolo_caracterisation_1999} distinguishes two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events.
 She shows that those processes leads to two abstract cognitive category respectively termed "the event sequences" and "the amorphous sequences".
@@
 They arise from a descriptive analysis based on the identification of the sound sources.
-On contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}.
+On the contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}.
 They result from a holistic analysis based on global acoustic features.
@@
 
+\ml{fix trim and check why figures goes offline}
+
 \begin{figure}[t]
         \myfloatalign
-        \includegraphics[width=.8\linewidth]{gfx/ch_4/3.png}
+        \includegraphics[trim={5cm 0 0 0},clip,width=.8\linewidth]{gfx/ch_4/3.png}
        \caption{Hierachical organisation of the isolated sounds used of the simulation.}\label{fig:orgDb}
@@
 \label{sec:xp1a_plan}
+
+The design of this experiment has been validated with a pilot study described in \cite{lafay2014new}.
 
@@
 The significativity is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
+\jfp{ce qu'est Ascene dans ce paragraphe n'est pas clair :
+A scene est défini comme la moyenne des scores d'une scène donnée pour tous les participants.
+
+le test, lui, porte sur la moyenne des Ascene pour les deux catégories
+même chose pour les sujets}
 \item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni- scenes has a significative influence  on the nature of the scenes \ie~ if significative differences exist between structural features and perceived pleasantness.
-The significativity is measured with a Student test on two populations;
+The significativity is measured with a Student test on two populations \jfp{je ne comprends pas bien ce test}
 \item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, linear correlations between these features and $\mathcal{A}_{scene}$ are considered.
@@
 where $c$ is the number of available classes, $c_k$ the number of classes for a given type of environment $k$, $c_j$ the number of  $j$ classes, and $c_{jk}$ the number of $j$ classes used for an environment $k$.
-The V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significatively different from the ratio $\frac{c_{jk}}{c_k}$.
+\jfp{il faudrait donner une référence pour ce test : bouquin Lebart, Morineau, Piron, STATISTIQUE EXPLORATOIRE MULTIDIMENSIONNELLE
+peut être donner les bornes de variations de Vjk pour aider le lecteur à comprendre cet indicateur} \ml{y pas une ref en Anglais ?} The V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significatively different from the ratio $\frac{c_{jk}}{c_k}$.
 For a given environment $k$, and class $j$, if the hypothesis is rejected, the class $j$ is said to be typical with respect to environment $k$.
-The typical classes are called \textbf{sound markers};
+The typical classes are called \textbf{sound markers};\jfp{on ne comprend pas bien l'hypothèse alternative du test.
+
+VOIR BOUQUIN LEBART page 184
+ou bien article de Morineau sur le sujet
+
+Vjk est une estimation approchée de la valeur test
+}
 
@@
 Relationships between $\mathcal{A}_{scene}$ and the structural features are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and ~\ref{fig:soundlevelf}.
+\jfp{en table 2, il devrait y avoir les coefficients de corrélation pour toutes les données, puis pour les deux cas ni et i.
+}
 
@@
 \vspace{0.5mm}
-\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and structural features.}
+\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and structural features.\jfp{rajouter une colonne corélation globale}}
 \label{tab:corrStructA}

### 2017-08-23 lafaySimulatedScenePerception.tex
done with JFP comments

@@
         \includegraphics[trim={5cm 0 0 0},clip,width=.8\linewidth]{gfx/ch_4/3.png}
-       \caption{Hierachical organisation of the isolated sounds used of the simulation.}\label{fig:orgDb}
+       \caption{Hierarchical organisation of the isolated sounds used in the simulation.}\label{fig:orgDb}
 \end{figure}
@@
 The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources is.
-For classes with high level of abstraction that have sub classes, its collection of samples is the union of the collections of the sub classes.
+For classes with high level of abstraction that have sub classes, their collection of samples is the union of the collections of the sub classes.
 
@@
 Accounting for the perceptual matters detailed previously, two taxonomies are built, one for event sounds and one for texture sounds.
-Four levels of abstraction are considered from the more generic one (level 0), to the more specific classes (level 3), leading to a taxonomy close to the one used in \cite{Salamon14}.
+Four levels of abstraction are considered from the more generic one (level 0), to the more specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}.
 Only three levels of abstraction are considered for the texture sounds.
@@
 
-All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie relative to Full Scale. In our case, the full scale level is of 1 Volt.
+All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie relative to Full Scale. In our case, the full scale level is set arbitrarily to 1 Volt.
 
@@
 
-By a volountary design choice, the simulation tool do not allow the user to interact and control a specific sample.
+By a volountary design choice, the simulation tool does not allow the user to interact and control directly a specific sample.
 Interaction is done at the track level, a track being a sequence of samples.
@@
 Most browsing tools are based on keyword indexing.
-However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potenitally induce biases in the analysis conclusions.
+However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potentially induce biases in the analysis conclusions.
 Indeed, the subject may select \textit{a priori} sounds referenced as belonging to \emph{park} to build a \emph{calm} soundscape.
@@
 
-Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a plane.
+Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a plane, each representing one leaf class.
 
@@
 Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
-\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes: respectively ideal (i) or non ideal (ni).
+\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni).
 Furthermore, this information is given by the creator of the scene.
 The second experimental step aims at investigating more deeply and more broadly our knowledge on pleasantness.
-For that, a second group of participants are asked to evaluate the pleasantness of each simulated scene, on a semantic scale.
+For that, a second group of participants is asked to evaluate the pleasantness of each simulated scene, on a semantic scale.
 This experiment has two goals:
 \begin{enumerate}
-\item evaluate the respective influence on the pleasantness of the various sources composing the scenes (i or ni)
-\item detect the presence of \emph{outliers} or ambiguous samples.
+\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene.
+\item to detect the presence of \emph{outliers} or ambiguous samples.
 For the remainder of the study, the defined hedonic properties (i and ni) are used as references.
@@
 All the participants start by designing the ideal environment; they read the second set of instruction at the end of the first experiment.
-Participants are completely free of their choice concerning sounds and synthesis parameters.
+Participants are completely free of their choices concerning sounds and synthesis parameters.
 They must nevertheless respect the two following constraints :
@@
 \item the listening point of view is the one of a fixed listener;
-\item the soundscape must be realistic, i.e. physically plausible.
-\end{itemize}
-For instance, the participant can insert ten dogs in the soundscape but she cannot insert one dog barking every 10 milliseconds.
+\item the soundscape must be realistic, i.e. physically plausible. For instance, the participant can insert ten dogs in the soundscape but she cannot insert one dog barking every 10 milliseconds.
+\end{itemize}
+
 These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation user interface.
@@
 44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in downtown Nantes since 2 years or more.
-Among the 44 participants, 40 succeeded, producing in the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
+Among the 44 participants, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
 the 4 other participants were excluded from the process due to a lack of respect or understanding of the instructions, or for exceeding the maximum duration of the experiment.
@@
 The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}.
-All the sound scenes have been re-composed on the basis of the scores obtained in the simulation experiment (Exp.
-1.a).
+The stimuli are the scenes obtained by the simulation experiment of Exp.
+1.a.
 The output sound level is the same for all the participants.
@@
 To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference.
-In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt.
+In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale).
 The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene.
@@
 
-In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following tests of significativity:
-
-\begin{itemize}
-\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with respect to the two types of scenes.
-The significativity is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
+In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following significance tests:
+
+\begin{itemize}
+\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni-scenes is also present in terms of perceived pleasantness, we look at the existence of differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ for each type of scenes (i or ni).
+The significance is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
 \jfp{ce qu'est Ascene dans ce paragraphe n'est pas clair :
@@
 même chose pour les sujets}
-\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni- scenes has a significative influence  on the nature of the scenes \ie~ if significative differences exist between structural features and perceived pleasantness.
-The significativity is measured with a Student test on two populations \jfp{je ne comprends pas bien ce test}
+\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni-scenes has a significative influence  on the structural features of the scenes \ie~ if significative differences exist between structural features and perceived pleasantness.
+The significance is measured with a Student test on two populations \jfp{je ne comprends pas bien ce test}
 \item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, linear correlations between these features and $\mathcal{A}_{scene}$ are considered.
 The Pearson correlation coefficient is used for that purpose, on the basis of the standard methodology used in multidimensional studies;
-\item \emph{analysis of the semantic features}: in order to determine if the affective difference between  the i- and ni- scenes has an influence on the scenes composition in terms of sound sources, or if specific classes are more frequently used to simulate a given environment (i or ni), a V-test is considered.
+\item \emph{analysis of the semantic features}: in order to determine if the affective difference between  the i- and ni-scenes has an influence on the scenes composition in terms of sound sources, or if specific classes are more frequently used to simulate a given environment (i or ni), a V-test is considered.
 Testing is done at every level of abstraction, and separately for texture and event classes.
@@
 $S$ is a boolean vector and the distance between two scenes is computed with the Hamming distance.
-Considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, where $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ is of cardinality $n$, with $x=\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between two vectors, ans is defined as follows:
+Considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, where $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ is of cardinality $n$, with $x \in\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between two vectors, ans is defined as follows:
 
@@
 The use of Hamming distance allows us to take into account equally the presence and absence of classes.
-For measuring the intrinsic ability of the space to split the i- and ni- scene, we use a ranking metric named the precision at rank $k$ ($P@k$) .
+For measuring the intrinsic ability of the space to split the i- and ni-scene, we use a ranking metric named the precision at rank $k$ ($P@k$) .
 The $P@k$ computes the precision obtained after the $k$ closest items of a given seed item have been found.
@@
 
-Except for the V-test, the statistical test of significativity are done with a critical threshold of $\alpha=0.05$.
+Except for the V-test, the statistical tests of significance are done with a critical threshold of $\alpha=0.05$.
 For the V-test, considering that a great number of classes is tested, a Bonferroni correction is applied.
@@
 
-First, in order to ensure the coherence of the data, we check that none of the ni- scenes gets a $\mathcal{A}_{scene}$ higher than one of a i- scene.
-Four ni- scenes don’t fullfill that constraint: they are thus removed, together with their corresponding i- scenes.
-As a consequence, 36 i -scenes and 36 ni -scenes remain for analysis.
-Second, we verify that participants really perceived a difference in terms of pleasantness between i- and ni -scenes.
+First, in order to ensure the coherence of the data, we check that none of the ni-scenes gets a $\mathcal{A}_{scene}$ higher than one of a i-scene.
+Four ni-scenes don’t fullfill that constraint: they are thus removed, together with their corresponding i-scenes.
+As a consequence, 36 i-scenes and 36 ni-scenes remain for analysis.
+Second, we verify that participants really perceived a difference in terms of pleasantness between i- and ni-scenes.
 For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment.
-It appears that the i- scenes were perceived significatively more pleasant ($p < 0.01$) than the ni- scenes.
+It appears that the i-scenes were perceived significatively more pleasant ($p < 0.01$) than the ni-scenes.
 
@@
 Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the levels $L$, $L(E)$ and $L(T)$.
-There is a significative difference in terms of sound level between i- and ni -scenes ($L$: $p<0.01$), with a standard deviation of -7 $dB$.
-This difference is siginificant for events ($L(E)$: $p<0.01$, STD: -7 $dB$) and for textures ($L(T)$: $p<0.01$, STD: -6 $dB$).
-
-As expected, the sound sources sound level is indeed a pleasantness indicator, as the ni -scenes tend to be louder, an outcome of a large number of related works.
-We also notice that this difference of sound level is equally true for events or textures.
-
-It appears that the biggest influence on the global sound level comes from the events, the gap between $L$ and $L(E)$ being only 1 $dB$ for i- and ni- scenes.
+There is a significative difference in terms of sound level between i- and ni-scenes ($L$: $p<0.01$).
+This difference is siginificant for events ($L(E)$: $p<0.01$) and for textures ($L(T)$: $p<0.01$).
+%, with a standard deviation of -7 $dB$
+%, STD: -7 $dB$
+%, STD: -6 $dB$
+
+As expected, the sound sources sound level is indeed a pleasantness indicator, as the ni-scenes tend to be louder.
+This results is also an outcome of a large number of related works.
+We also notice that this difference of sound level is both significant for events or textures.
+
+It appears that the biggest influence on the global sound level comes from the events, the gap between $L$ and $L(E)$ being only 1 $dB$ for i- and ni-scenes.
 This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}.
@@
 Though, the level alone is not sufficient to differentiate the type of environment.
-In fact, $20\%$ of the i- scenes have a sound level higher than the nominal level of the ni- scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
+In fact, $20\%$ of the i-scenes have a sound level higher than the nominal level of the ni-scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
 
@@
 Concerning $L$, we can observe a strong negative correlation ($r=-0.77$, $p<0.01$) with $\mathcal{A}_{scene}$, indicating that the higher the sound level is, the more unpleasant the scene is perceived.
-Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship do not occur in the same way for i- and ni- scenes.
-In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni- scenes ($r=-0.78$, $p<0.01$), but is quasi null for i-scenes.
-
-When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the i- scenes tend to be softer than the ni- scenes, thus allowing us to extend to the i-scenes the negative correlation observed for the ni- scenes.
+Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship do not occur in the same way for i- and ni-scenes.
+In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni-scenes ($r=-0.78$, $p<0.01$), but is quasi null for i-scenes.
+
+When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the i-scenes tend to be softer than the ni-scenes, thus allowing us to extend to the i-scenes the negative correlation observed for the ni-scenes.
 
@@
 \begin{itemize}
-\item allows us to differentiate between i- and ni- scenes,
-\item characterizes precisely the perceived pleasantness for ni- scenes,
-\item is not a relevant feature for modeling the perceived pleasantness of \textit{a priori} pleasant soundscapes (i- scenes)
+\item allows us to differentiate between i- and ni-scenes,
+\item characterizes precisely the perceived pleasantness for ni-scenes,
+\item is not a relevant feature for modeling the perceived pleasantness of \textit{a priori} pleasant soundscapes (i-scenes)
 \end{itemize}
@@
 For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$).
-Again, we believe that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer than ni- scenes).
-Thus, the sound events keep a relative ability to predict the pleasantness of the ni- scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.
+Again, we believe that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i-scenes tend to be softer than ni-scenes).
+Thus, the sound events keep a relative ability to predict the pleasantness of the ni-scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.
 
@@
 
-Those first outcomes tend to show that 1) two modes of perception exist depending on the very nature of the environment (i- or ni-), and 2) each involve independent features.
-The fact that $L$ is not sufficient to characterize the pleasantness of the i- scenes can lead us to conclude that all the sound sources do not equally contribute to the perception of pleasantness.
+Those first outcomes tend to show that 1) two modes of perception exist depending on the very nature of the environment (i- or ni-), and 2) each involves independent features.
+The fact that $L$ is not sufficient to characterize the pleasantness of the i-scenes can lead us to conclude that all the sound sources do not equally contribute to the perception of pleasantness.
 Thus, we think that only the level of some of them can influence this percept.
@@
 
-We observe a significant difference between i- and ni- scenes for the classes choice.
-The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~classes involving human presence and nature are prevailing in the i- scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni- scenes.
+We observe a significant difference between i- and ni-scenes for the classes choice.
+The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~classes involving human presence and nature are prevailing in the i-scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni-scenes.
 These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.
@@
 Results are presented in Table~\ref{tab:markers}.
+\jfp{le V-test est un test statistique : on peut mettre la p-value dans le tableau}
 
 Regarding the sound events, 9 markers are identified for all abstraction levels.
-As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i- scenes markers as well as the \emph{church bell} class.
+As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i-scenes markers as well as the \emph{church bell} class.
 This latter result may be due to the socio-cultural background of the participants who are mostly European citizens.
 In fact, according to Schafer, a sound that is identified by a person as being an element of his/her environment, is well accepted.
-Sound markers of ni- scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
+Sound markers of ni-scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
 
 Regarding the textures, 5 markers are identified.
-For the i- scenes, they are classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}).
-For the ni- scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).
+For the i-scenes, they are classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}).
+For the ni-scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).
 
@@
 \vspace{0.5mm}
-\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of $V$ value.}
+\caption{Event and texture classes identified as sound markers.
+In each cell, markers are ranked in the decreasing order of V-test value.\jfp{a partir de quelle valeur de V-test le marker est présent dans ce tableau ?
+
+mettre la p-value , non ? }}
 \label{tab:markers}
@@
 We also notice that the higher the abstraction level is, the more important the ability to make this distinction is.
-In other words, the more precise the scene description is, the more clear the discrimination between i- and ni- scenes is.
+In other words, the more precise the scene description is, the more clear the discrimination between i- and ni-scenes is.
 
@@
 Moreover, in the case when the sound markers are not taken into account for the description ($E_{w/o,m}$), the performance is under those obtained with only the textures as features.
-This confirms that most of the semantic information allowing to differentiate between i- and ni- scenes is included in the markers.
+This confirms that most of the semantic information allowing to differentiate between i- and ni-scenes is included in the markers.
 
@@
 \item the semantic information is mainly contained in the event sound classes;
-\item only a part of the event classes, i.e. the sound markers, are useful to differentiate between the i- and ni- scenes.
+\item only a part of the event classes, i.e. the sound markers, are useful to differentiate between the i- and ni-scenes.
 \end{enumerate}
 
-Since we have extracted the typical classes of the i- and ni- scenes and verified that the distinction between them was largely dependent on the presence of these classes, we shall now investigate whether a structural description of the scenes only based on sound markers could characterize the perceived pleasantness, perhaps better than a global structural description.
+Since we have extracted the typical classes of the i- and ni-scenes and verified that the distinction between them was largely dependent on the presence of these classes, we shall now investigate whether a structural description of the scenes only based on sound markers could characterize the perceived pleasantness, perhaps better than a global structural description.
 
@@
 On the opposite, we define $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
-When the feature characterizes an i- scene (idem for a ni- scene),  only the markers identified for the i- scenes (or the ni- scenes) are considered; we named them, i- markers (or ni- markers).
+When the feature characterizes an i-scene (idem for a ni-scene),  only the markers identified for the i-scenes (or the ni-scenes) are considered; we named them, i- markers (or ni- markers).
 Results are shown in Table~\ref{tab:corrMarkers}.
 
-Concerning the sound levels, the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ et $L(T)$ on the other side, see~Figures~\ref{fig:soundlevelMarkera} and~\ref{fig:soundlevelMarkerd}, respectively.
+Concerning the sound levels, the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ and $L(T)$ on the other side, see~Figures~\ref{fig:soundlevelMarkera} and~\ref{fig:soundlevelMarkerd}, respectively.
 No matter whether all the classes or only the markers are considered, it appears that:
@@
 \begin{enumerate}
-\item a significative difference between levels of i- and ni- scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
+\item a significative difference between levels of i- and ni-scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
 \item the sound level of scenes is mainly related to the sound events, compared to the textures;
-\item the sound level of events has an influence on the perception of pleasantness for ni- scenes, but not for i- scenes;
+\item the sound level of events has an influence on the perception of pleasantness for ni-scenes, but not for i-scenes;
 \item the sound level of textures does not play any role in the perception of the pleasantness.
@@
 
-In conclusion, the level of ni- markers has a negative influence on the pleasantness for the ni- scenes.
-On the other hand, the level of i- markers does not influence the perceived pleasantness for the i- scenes.
+In conclusion, the level of ni- markers has a negative influence on the pleasantness for the ni-scenes.
+On the other hand, the level of i- markers does not influence the perceived pleasantness for the i-scenes.
 
@@
 
-Considering the non markers classes, we can observe on the i- scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}.
-It is the first time that an objective feature allows us to define the pleasantness of i- scenes.
+Considering the non markers classes, we can observe on the i-scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}.
+It is the first time that an objective feature allows us to define the pleasantness of i-scenes.
 This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.
@@
 
-Besides, whereas $L(T)$ did not show any significative correlation for ni- scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
-This indicates that the level of non-markers texture classes does not influence the perceived pleasantness in the same way for i- and ni- scenes.
-Sound level seems to have a negative effect for the ni- scenes, and no significant effect for the i- scenes.
+Besides, whereas $L(T)$ did not show any significative correlation for ni-scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
+This indicates that the level of non-markers texture classes does not influence the perceived pleasantness in the same way for i- and ni-scenes.
+Sound level seems to have a negative effect for the ni-scenes, and no significant effect for the i-scenes.
 
@@
 
-For the i- scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
-for the ni- scenes no correlation is observed.
-Thus, for the i- scenes, it is not the absolute markers level that is important, but their relative level with regards to the other sounds composing the scene.
+For the i-scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
+for the ni-scenes no correlation is observed.
+Thus, for the i-scenes, it is not the absolute markers level that is important, but their relative level with regards to the other sounds composing the scene.
 Then, we can induce a double perceptual mechanism for the ideal environments:
@@
 
-For the ni -scenes, the fact that we observe significative correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.
+For the ni-scenes, the fact that we observe significative correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.
 
@@
         {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
-        \caption{Distribution of the relative sound levels structural features related to the presence of markers $L_m$ (a, d), $L(E)_m$ (b, e) and $L(T)_m$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
+        \caption{Distribution of the relative sound levels structural features related to the presence of markers $L_m$ (a, d), $L_b$ (b, e) and $L_m-L_b$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
 \end{figure*}
@@
 \begin{itemize}
-\item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global structural features ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni- scenes.
+\item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global structural features ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni-scenes.
 The semantic description seems to be more powerful;
@@
 \item \emph{pleasantness prediction}: considering the correlation between structural features and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its nature (i or ni).
-From the data gathered in those experiments, it is not conceivable to consider the same set of features to predict the pleasantness of i- and ni- scenes in the same time:
+From the data gathered in those experiments, it is not conceivable to consider the same set of features to predict the pleasantness of i- and ni-scenes in the same time:
 
@@
 
-The fact that the pleasantness of the i- scenes is not correlated to global physical features, contrary to pleasantness of ni- scenes, has been studied recently \cite{gozalo2015relationship}.
+The fact that the pleasantness of the i-scenes is not correlated to global physical features, contrary to pleasantness of ni-scenes, has been studied recently \cite{gozalo2015relationship}.
 
@@
 \item \emph{72 km-scenes}: the 72 scenes originally simulated by the subject of experiment 1.a where the sound classes identified as markers are kept.
-The 36 ideal scenes with markers are noted i/km-scenes, and the 36 non ideal scenes with markers are noted ni/km-scenes (k for kept).
+The 36 ideal scenes with markers are noted i/km-scenes, and the 36 non ideal scenes with markers are noted ni/km-scenes (km for kept markers).
 \item \emph{72 rm-scenes}:  72 scenes where the sound classes identified as markers are removed.
-The 36 ideal scenes without markers are noted i/rm-scenes, and the 36 non ideal scenes without markers are noted ni/rm-scenes (r for removed).
+The 36 ideal scenes without markers are noted i/rm-scenes, and the 36 non ideal scenes without markers are noted ni/rm-scenes (r for removed markers).
 \end{itemize}
@@
 
-In order to create km-scenes with still some sound diversity and no  absence of sound activity within a long period of the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}.
+In order to create rm-scenes with still some sound diversity and no  absence of sound activity within a long period of the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}.
 Those classes are:
@@
 After scoring, they are free to move on to the next scene.
+\jfp{on se demande pourquoi on va de -5 a +5 alors qu'avant on allait de -3 à +3}
 
@@
 The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness.
-To do so, we perform a variance analysis.
-We consider $\mathcal{A}_{sujet}$ as a dependent variable, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm).
+To do so, we perform a an Analysis of Variance.
+We consider $\mathcal{A}_{subject}$ as a dependent variable \jfp{je crois que la défintion de la variable dépendante, Asubject, n'est pas la bonne
+
+A subject a a été définit avant comme "Asubject is
+the average pleasantness for each participant"
+l'ANOVA à deux facteurs à mesures répétées ne porte pas sur la valeur moyenne mais sur la valeur de l'évaluation pour une scene donnée
+
+il faut donc définir une notation appropriée
+}, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm).
 As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to check the validity of the hypothesis that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes.
@@
 Let us consider $\mathcal{A}_{sujet}$ for the km-scenes.
-Close inspection of the answers show that one subject strongly differs with the ones of others.
+Close inspection of the answers show that one subject strongly differs from the ones of others.
 This subject evaluated positively close to half of the ni/km-scenes, see Annex~\ref{app:xp2}, subject 7.
@@
 The \emph{post hoc} analysis exhibits significative differences between all the groups of observations, notably between the i/km-scenes ($p<0.05$) and the i/rm-scenes ($p<0.01$).
+\jfp{je ne comprends pas les 2 p-value
+
+il ne peut y a voir qu'une seule p-value lorsqu'on compare deux conditions expérimentales
+}
 
@@
 
-These results indicate that the removal of the markers did indeed modify the perception of the scenes by the subjects. Our two hypotheses are thus verified:
+These results indicate that the removal of the markers did indeed modify the perception of the scenes by the subjects.
+Our two hypotheses are thus verified \jfp{avant cela il faut dire un mot sur le sens des effets : ils vont bien dans le sens souhaité : ce n'est pas le test qui le dit}:
 
@@
 The i-markers do have a positive impact on the perception of an environment.
-The fact that their removal decreases $\mathcal{A}_{scene}$ clearly shows that it is possible to improve the perceived pleasantnes of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
+The fact that their removal decreases $\mathcal{A}_{scene}$ clearly shows that it is possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
 Those conclusions are in line with the positive approach introduced by Schafer in \cite{schafer1977tuning}.
@@
 
-These results tend to confirm that the perception of the qualities of a scene indeed depends above all on the sound sources that compose it.
+These results tend to confirm that the perception of the qualities of a scene indeed depends primarily on the sound sources that compose it.
 The compositional characteristics that are taken into account during the perceptual process appear to vary from one source to the other, from one type of environment to the other.
-This fact lead the authors to believe that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.
+This fact leads the authors to believe that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.
 
@@
 \newpage
-\section{Distribution of scores given par the subjects during experiment 2}
+\section{Distribution of scores given by the subjects during experiment 2}
 \label{app:xp2}

### 2017-08-24 lafaySimulatedScenePerception.tex
toto

@@
 \usepackage{amsmath}
+\usepackage{color}
 
@@
 
-\newcommand{\gl}[1]{\textcolor{blue}{GL : #1}}
+\newcommand{\gl}[1]{\textcolor{magenta}{GL : #1}}
 \newcommand{\nm}[1]{\textcolor{magenta}{#1}}
 \newcommand{\ml}[1]{\textcolor{red}{#1}}
-\newcommand{\jfp}[1]{\textcolor{green}{JFP : #1}}
+\newcommand{\jfp}[1]{\textcolor{blue}{JFP : #1}}
 \newcommand{\ie}{\emph{i.\,e.}}
@@
 A scene est défini comme la moyenne des scores d'une scène donnée pour tous les participants.
-
 le test, lui, porte sur la moyenne des Ascene pour les deux catégories
-même chose pour les sujets}
+meme chose pour les sujets}
 \item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni-scenes has a significative influence  on the structural features of the scenes \ie~ if significative differences exist between structural features and perceived pleasantness.
@@
 The typical classes are called \textbf{sound markers};\jfp{on ne comprend pas bien l'hypothèse alternative du test.
-
-VOIR BOUQUIN LEBART page 184
-ou bien article de Morineau sur le sujet
-
-Vjk est une estimation approchée de la valeur test
-}
+VOIR BOUQUIN LEBART page 184 ou bien article de Morineau sur le sujet Vjk est une estimation approchée de la valeur test}
 
@@
 In each cell, markers are ranked in the decreasing order of V-test value.\jfp{a partir de quelle valeur de V-test le marker est présent dans ce tableau ?
-
-mettre la p-value , non ? }}
+mettre la p-value , non ?
+}}
 \label{tab:markers}
@@
 To do so, we perform a an Analysis of Variance.
-We consider $\mathcal{A}_{subject}$ as a dependent variable \jfp{je crois que la défintion de la variable dépendante, Asubject, n'est pas la bonne
-
+We consider $\mathcal{A}_{subject}$ as a dependent variable \jfp{je crois que la défintion de la variable dépendante, Asubject, n'est pas la bonne.
 A subject a a été définit avant comme "Asubject is
-the average pleasantness for each participant"
-l'ANOVA à deux facteurs à mesures répétées ne porte pas sur la valeur moyenne mais sur la valeur de l'évaluation pour une scene donnée
-
-il faut donc définir une notation appropriée
-}, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm).
+the average pleasantness for each participant".
+l'ANOVA à deux facteurs à mesures répétées ne porte pas sur la valeur moyenne mais sur la valeur de l'évaluation pour une scene donnée.
+il faut donc définir une notation appropriée}, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm).
 As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to check the validity of the hypothesis that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes.
@@
 The \emph{post hoc} analysis exhibits significative differences between all the groups of observations, notably between the i/km-scenes ($p<0.05$) and the i/rm-scenes ($p<0.01$).
-\jfp{je ne comprends pas les 2 p-value
-
+\jfp{je ne comprends pas les 2 p-value.
 il ne peut y a voir qu'une seule p-value lorsqu'on compare deux conditions expérimentales

### 2017-08-24 lafaySimulatedScenePerception.tex
fixed figure layout

@@
-\documentclass[preprint,12pt]{elsarticle}
+\documentclass[12pt]{elsarticle}
 \usepackage{url}
@@
 
-\journal{Nuclear Physics}
+\journal{Behavior Research Methods}
 
@@
 
-\ml{fix trim and check why figures goes offline}
-
-\begin{figure}[t]
+\begin{figure}
         \myfloatalign
-        \includegraphics[trim={5cm 0 0 0},clip,width=.8\linewidth]{gfx/ch_4/3.png}
+        \includegraphics[trim={0 0 0 2.5cm},clip,width=.8\linewidth]{gfx/ch_4/3.png}
        \caption{Hierarchical organisation of the isolated sounds used in the simulation.}\label{fig:orgDb}
@@
 
-\begin{figure}[t]
-        \myfloatalign
+\begin{figure}
         \subfloat[]
-        {\includegraphics[width=.8\columnwidth]{gfx/ch_4/SSF}\label{fig:ssf}} \par
+        {\includegraphics[width=.8\linewidth]{gfx/ch_4/SSF}\label{fig:ssf}} \par
         \subfloat[]
-        {\includegraphics[width=\columnwidth]{gfx/ch_4/simscene}\label{fig:simscene}}
+        {\includegraphics[width=\linewidth]{gfx/ch_4/simscene}\label{fig:simscene}}
        \caption{Graphical interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{simScene} tool.}
@@
 
-\begin{figure}[t]
-        \myfloatalign
+\begin{figure}
         \includegraphics[width=\linewidth]{gfx/ch_5/5.png}
@@
 
-\begin{figure*}[t]
-        \myfloatalign
+\begin{figure*}
         \subfloat[]
@@
 
-\begin{figure}[t]
-        \myfloatalign
+\begin{figure}
         \subfloat[]
@@
         \subfloat[]
-        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_2}\label{fig:soundsourceb}}
+        {\includegraphics[width=.5\linewidth]{gfx/ch_5/xp1_class_2}\label{fig:soundsourceb}}
         \subfloat[]
-        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_3}\label{fig:soundsourcec}}
+        {\includegraphics[width=.5\linewidth]{gfx/ch_5/xp1_class_3}\label{fig:soundsourcec}}
        \caption{Proportion of synthesized scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.}\label{fig:soundsource}

### 2017-09-01 lafaySimulatedScenePerception.tex
improve introduction

@@
   %% keywords here, in the form: keyword \sep keyword
- cognitive psychology \sep soundscape assessment
+ cognitive psychology \sep soundscape perception \ soundscape simulator
   %% PACS codes here, in the form: \PACS code \sep code
@@
 One of the main goal of soundscape studies is to identify which components of the soundscape influence human perception \cite{aletta2016soundscape}.
-For example, in order to study the link between an environment (urban sound scene) and the induced human sensation (calmness), one mean is to objectify the mental representation of a specific sound scene, a calm urban sound scene.
+For example, in order to study the link between an environment (urban sound scene) and the induced human sensation (calmness), one mean is to objectify the mental representation of a specific sound scene (a calm urban sound scene).
 Several means have been considered in the literature to perform this task.
@@
 The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}.
-That being said, those recorded stimuli  can be analysed, what allows the experimenter to gather a coarse
+That being said, those recorded stimuli  can be analyzed, what allows the experimenter to gather a coarse
 grain physical description of them.
@@
 
+%Nous pensons que ces efforts de recherche peuvent grandement bénéficier de l'utilisation de scènes simulées, scènes dont les caractéristiques structurelles, en particulier la nature des sources présentes, leur répartition temporelle ainsi que leur forme d'onde respectives, sont directement accessibles par l'expérimentateur.
+
 For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment of interest.
-With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the events.
-
-%Nous pensons que ces efforts de recherche peuvent grandement bénéficier de l'utilisation de scènes simulées, scènes dont les caractéristiques structurelles, en particulier la nature des sources présentes, leur répartition temporelle ainsi que leur forme d'onde respectives, sont directement accessibles par l'expérimentateur.
+With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the sound events.
 
@@
 The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics.
-A lack of diversity is nonetheless mentioned by the authors that limit the strenght of the outcomes given in this study.
+A lack of diversity in terms of sound sources is nonetheless mentioned by the authors that limit the strength of the outcomes given in this study.
 
@@
 It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}.
-It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}.
+It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digital audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}.
 The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.
@@
 
+%Afin de créer un corpus de classes de sons de référence pour la simulation, nous réalisons une typologie des sons environnementaux urbains.
+
 In order to provide the user with a sound database that is well organized and covers as much as possible the variety of sound sources that are present in urban areas, a typology of urban sounds is first defined.
@@
 \subsubsection*{Typology}
-
-%Afin de créer un corpus de classes de sons de référence pour la simulation, nous réalisons une typologie des sons environnementaux urbains.
 
@@
 The literature usually distinguishes between:
+
 \begin{itemize}
@@
 
-The simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with the classes of events only and the other with classes of textures only.
-These two types also have specific scheduling procedures during the simulation process.
-
 % et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures. Événements et textures bénéficient de processus de simulation spécifiques.
+
+The simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with the classes of events only and the other with the classes of textures only.
+These two types of sound classes also have specific scheduling procedures during the simulation process.
 
@@
 
-According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from different sources of a sound environment \cite{bregman1994auditory}.
+According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment \cite{bregman1994auditory}.
 Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}.
@@
 Rosch \cite{rosch1978cognition}.
-The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources is.
+The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources.
 For classes with high level of abstraction that have sub classes, their collection of samples is the union of the collections of the sub classes.
@@
 
-By a volountary design choice, the simulation tool does not allow the user to interact and control directly a specific sample.
+By a voluntary design choice, the simulation tool does not allow the user to interact and control directly a specific sample.
 Interaction is done at the track level, a track being a sequence of samples.
@@
 \begin{itemize}
-\item \emph{level} ($dB$): the levels are  randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
-\item \emph{inter-onset spacing} (second): for the event's track only, the spacings are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
+\item \emph{sound level} ($dB$): for each sample, the sound levels are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
+\item \emph{inter-onset spacing} (second): for the event's track only, and for each sound event sample, the inter-onset spacings are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
 \item \emph{start and end time} (second): the subject set the first and last occurrence of a sample of the selected class.
@@
 \begin{itemize}
-\item \emph{event's fades} (seconds): for the event's track only, the subject can set a fade in / fade out duration applied to each samples;
-\item \emph{global fades} (seconds): the subject set the duration of the fade in / fade out at the beginning and the end of the track.
+\item \emph{event's fades} (seconds): for the event's track only, the subject can set a fade in / fade out duration applied to each sample;
+\item \emph{global fades} (seconds): the subject can set global fade in / fade out durations applied to the entire track.
 \end{itemize}
@@
 
-%La grande majorité des outils permettant de parcourir une banque de données propose une recherche textuelle sur la base de mots-clé.
-Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique.
-
-%En effet la description verbale du son, si elle est accessible au sujet, peut potentiellement influencer sa sélection.
-Dans les faits, pour construire une scène environnementale «~calme~», le sujet sélectionne a priori les sons référencés sous le vocable \emph{parc}.
-
-Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user.
-Most browsing tools are based on keyword indexing.
-However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potentially induce biases in the analysis conclusions.
-Indeed, the subject may select \textit{a priori} sounds referenced as belonging to \emph{park} to build a \emph{calm} soundscape.
-
-%Pour limiter l’influence de l’interface sur le sujet, il nous paraît nécessaire de libérer sa recherche de toute information textuelle.
-Nous proposons à l'utilisateur une interface graphique lui permettant d’explorer la banque de sons exclusivement à partir de l’écoute.
-
-Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a plane, each representing one leaf class.
-
-%La figure~\ref{fig:ssf} présente l'interface pour la banque de données d'événements sonores.
-Visuellement, les classes du dernier palier (niveaux d'abstraction 3 pour les événements, 2 pour les textures) sont représentées par des cercles, et positionnées sur un plan.
-La disposition des cercles dans l'espace dépend de l’organisation hiérarchique de la base de données: les sous-classes appartenant à une même classe sont proches les unes des autres, et ainsi de suite jusqu'à atteindre les classes du dernier palier, les seules auxquelles le sujet peut accéder, des classes qui ne possèdent pas de sous-classes, et sont directement liées à une collection de samples.
-
-The Figure~\ref{fig:ssf} shows the interface used for the selection of events.
-The classes of the higher level of abstraction are plotted with circles.
-The spatial location of the circles is chosen according to the hierarchical organization of the sound database: sub-classes belonging to the same class are close to each others, and so on until the user reaches the lower level classes displayed in grey which are directly linked to a collection of samples.
-
-%Chaque classe du dernier palier possède un son prototype.
-Ces sons ont été choisis par les expérimentateurs.
-Lorsqu’on clique sur un cercle, le prototype associé à la classe est joué.
-Le sujet parcourt la banque de sons en cliquant sur les cercles.
-Cette interface a fait l'objet d'une étude approfondie dont les résultats sont publiés dans \cite{lafay2016JAES}.
-
-Each of those classes have a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle.
-The subject can browse the database by listening to those prototype sounds.
-The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed in \cite{lafay2016JAES}.
-
 \begin{figure}
+		\centering
         \subfloat[]
@@
 \end{figure}
+
+%La grande majorité des outils permettant de parcourir une banque de données propose une recherche textuelle sur la base de mots-clé.
+Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique.
+
+%En effet la description verbale du son, si elle est accessible au sujet, peut potentiellement influencer sa sélection.
+Dans les faits, pour construire une scène environnementale «~calme~», le sujet sélectionne a priori les sons référencés sous le vocable \emph{parc}.
+
+Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user.
+Most browsing tools are based on keyword indexing.
+However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potentially induce biases in the analysis conclusions.
+Indeed, the subject may select \textit{a priori} sounds referenced as belonging to \emph{park} to build a \emph{calm} soundscape.
+
+%Pour limiter l’influence de l’interface sur le sujet, il nous paraît nécessaire de libérer sa recherche de toute information textuelle.
+Nous proposons à l'utilisateur une interface graphique lui permettant d’explorer la banque de sons exclusivement à partir de l’écoute.
+
+Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a plane, each representing one leaf class.
+
+%La figure~\ref{fig:ssf} présente l'interface pour la banque de données d'événements sonores.
+Visuellement, les classes du dernier palier (niveaux d'abstraction 3 pour les événements, 2 pour les textures) sont représentées par des cercles, et positionnées sur un plan.
+La disposition des cercles dans l'espace dépend de l’organisation hiérarchique de la base de données: les sous-classes appartenant à une même classe sont proches les unes des autres, et ainsi de suite jusqu'à atteindre les classes du dernier palier, les seules auxquelles le sujet peut accéder, des classes qui ne possèdent pas de sous-classes, et sont directement liées à une collection de samples.
+
+The Figure~\ref{fig:ssf} shows the interface used for the selection of events.
+The classes of the higher level of abstraction are plotted with circles.
+The spatial location of the circles is chosen according to the hierarchical organization of the sound database: sub-classes belonging to the same class are close to each others, and so on until the user reaches the lower level classes displayed in grey which are directly linked to a collection of samples.
+
+%Chaque classe du dernier palier possède un son prototype.
+Ces sons ont été choisis par les expérimentateurs.
+Lorsqu’on clique sur un cercle, le prototype associé à la classe est joué.
+Le sujet parcourt la banque de sons en cliquant sur les cercles.
+Cette interface a fait l'objet d'une étude approfondie dont les résultats sont publiés dans \cite{lafay2016JAES}.
+
+Each of those classes have a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle.
+The subject can browse the database by listening to those prototype sounds.
+The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed in \cite{lafay2016JAES}.
 
@@
 In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.
+
 For each sound scene, three types of features are considered :
+
 \begin{itemize}
@@
 \begin{itemize}
-\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni-scenes is also present in terms of perceived pleasantness, we look at the existence of differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ for each type of scenes (i or ni).
-The significance is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
-\jfp{ce qu'est Ascene dans ce paragraphe n'est pas clair :
-A scene est défini comme la moyenne des scores d'une scène donnée pour tous les participants.
-le test, lui, porte sur la moyenne des Ascene pour les deux catégories
-meme chose pour les sujets}
-\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni-scenes has a significative influence  on the structural features of the scenes \ie~ if significative differences exist between structural features and perceived pleasantness.
-The significance is measured with a Student test on two populations \jfp{je ne comprends pas bien ce test}
-\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, linear correlations between these features and $\mathcal{A}_{scene}$ are considered.
-The Pearson correlation coefficient is used for that purpose, on the basis of the standard methodology used in multidimensional studies;
-\item \emph{analysis of the semantic features}: in order to determine if the affective difference between  the i- and ni-scenes has an influence on the scenes composition in terms of sound sources, or if specific classes are more frequently used to simulate a given environment (i or ni), a V-test is considered.
-Testing is done at every level of abstraction, and separately for texture and event classes.
-For each class $j$, and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:
+\item \emph{analysis of the perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness differ between the i- and ni-scenes.
+We look at the existence of differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ for each type of scenes (i or ni).
+The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$;
+
+%\jfp{ce qu'est Ascene dans ce paragraphe n'est pas clair :A scene est défini comme la moyenne des scores d'une scène donnée pour tous les participants.
+le test, lui, porte sur la moyenne des Ascene pour les deux catégories meme chose pour les sujets}
+
+\item \emph{analysis of the structural features}: the goal is to evaluate whether the structural features differ between the i- and ni-scenes.
+The significance is measured with a two-sample Student test;
+
+\item \emph{influence of the structural features on the perceived pleasantness}: the goal is to evaluate if structural features impact the perceived pleasantness.
+We look at linear correlations between these features and $\mathcal{A}_{scene}$.
+The Pearson correlation coefficient is used for that purpose;
+
+\item \emph{analysis of the semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni).
+A V-test is considered \citep{lebart2000statistique} \ml{y pas une ref en Anglais ?}.
+With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of time a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significatively different from the ratio $\frac{c_{jk}}{c_k}$.
+For each class $j$, and each environment type $k$, an approximation of the statistic value $V_{jk}$ is computed as follows:
 
@@
 
-where $c$ is the number of available classes, $c_k$ the number of classes for a given type of environment $k$, $c_j$ the number of  $j$ classes, and $c_{jk}$ the number of $j$ classes used for an environment $k$.
-\jfp{il faudrait donner une référence pour ce test : bouquin Lebart, Morineau, Piron, STATISTIQUE EXPLORATOIRE MULTIDIMENSIONNELLE
-peut être donner les bornes de variations de Vjk pour aider le lecteur à comprendre cet indicateur} \ml{y pas une ref en Anglais ?} The V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significatively different from the ratio $\frac{c_{jk}}{c_k}$.
-For a given environment $k$, and class $j$, if the hypothesis is rejected, the class $j$ is said to be typical with respect to environment $k$.
-The typical classes are called \textbf{sound markers};\jfp{on ne comprend pas bien l'hypothèse alternative du test.
+If the hypothesis is rejected, the class $j$ is said to be typical with respect to the type of scene $k$.
+The typical classes are called \textbf{sound markers}.
+Testing is done for each class, at every level of abstraction, and separately for texture and event classes;
+
+%\jfp{il faudrait donner une référence pour ce test : bouquin Lebart, Morineau, Piron, STATISTIQUE EXPLORATOIRE MULTIDIMENSIONNELLE peut être donner les bornes de variations de Vjk pour aider le lecteur à comprendre cet indicateur}  \jfp{on ne comprend pas bien l'hypothèse alternative du test.
 VOIR BOUQUIN LEBART page 184 ou bien article de Morineau sur le sujet Vjk est une estimation approchée de la valeur test}
 
-\item \emph{Study of the representation spaces induced by the semantic features}: in order to determine if a representation only based on the presence / absence of sound sources allows us to distinguish between the two types of environment, we investigate the space induced by the semantic features $S$.
-$S$ is a boolean vector and the distance between two scenes is computed with the Hamming distance.
-Considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, where $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ is of cardinality $n$, with $x \in\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between two vectors, ans is defined as follows:
+\item \emph{representation space induced by the semantic features}: the goal is to determine if a representation space of the scenes solely based on the presence / absence of sound sources allows us to distinguish between the two types of scene.
+Considering $S_i$ the semantic feature of the scene $i$, we compute the distances between all the vectors $S_i$.
+A Hamming distance is used.
+Considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$ and $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$, with $x \in\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between the two vectors.
+It is defined as follows:
 
@@
 where $\bigoplus$ is the \emph{exclusive-or}.
-The more the composition of the two scenes are similar, the closer they are in this space.
+Two scenes having similar source compositions will be close in such space.
 The use of Hamming distance allows us to take into account equally the presence and absence of classes.
-For measuring the intrinsic ability of the space to split the i- and ni-scene, we use a ranking metric named the precision at rank $k$ ($P@k$) .
+For measuring the intrinsic ability of the space to split the i- and ni-scene, we use a ranking metric named the precision at rank $k$ ($P@k$).
 The $P@k$ computes the precision obtained after the $k$ closest items of a given seed item have been found.
@@
 
-Except for the V-test, the statistical tests of significance are done with a critical threshold of $\alpha=0.05$.
+The statistical tests of significance are done with a critical threshold of $\alpha=0.05$.
 For the V-test, considering that a great number of classes is tested, a Bonferroni correction is applied.
@@
 
-\subsubsection*{Checking of the pleasantness of the synthesized scenes}
+\subsubsection*{Analysis of the perceived pleasantness}
 
@@
 \vspace{0.5mm}
-\caption{Event and texture classes identified as sound markers.
-In each cell, markers are ranked in the decreasing order of V-test value.\jfp{a partir de quelle valeur de V-test le marker est présent dans ce tableau ?
-mettre la p-value , non ?
-}}
+\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of V-test value. $p\leq0.01$ for all sound markers.}
 \label{tab:markers}
@@
 After scoring, they are free to move on to the next scene.
-\jfp{on se demande pourquoi on va de -5 a +5 alors qu'avant on allait de -3 à +3}
+
+%\jfp{on se demande pourquoi on va de -5 a +5 alors qu'avant on allait de -3 à +3}
 
@@
 The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness.
-To do so, we perform a an Analysis of Variance.
-We consider $\mathcal{A}_{subject}$ as a dependent variable \jfp{je crois que la défintion de la variable dépendante, Asubject, n'est pas la bonne.
-A subject a a été définit avant comme "Asubject is
-the average pleasantness for each participant".
-l'ANOVA à deux facteurs à mesures répétées ne porte pas sur la valeur moyenne mais sur la valeur de l'évaluation pour une scene donnée.
-il faut donc définir une notation appropriée}, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm).
+To do so, we perform an Analysis of Variance.
+We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm).
 As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to check the validity of the hypothesis that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes.
@@
 \emph{Post hoc} analyses are done using the Tukey-Kramer procedure.
+
+%\jfp{je crois que la défintion de la variable dépendante, Asubject, n'est pas la bonne.
+A subject a a été définit avant comme "Asubject is the average pleasantness for each participant".
+l'ANOVA à deux facteurs à mesures répétées ne porte pas sur la valeur moyenne mais sur la valeur de l'évaluation pour une scene donnée.
+il faut donc définir une notation appropriée}
 
@@
 The \emph{post hoc} analysis exhibits significative differences between all the groups of observations, notably between the i/km-scenes ($p<0.05$) and the i/rm-scenes ($p<0.01$).
-\jfp{je ne comprends pas les 2 p-value.
-il ne peut y a voir qu'une seule p-value lorsqu'on compare deux conditions expérimentales
-}
+
+%\jfp{je ne comprends pas les 2 p-value. il ne peut y a voir qu'une seule p-value lorsqu'on compare deux conditions expérimentales}
 

### 2017-09-01 lafaySimulatedScenePerception.tex
improve simulator sec and planning sec

@@
 
+\begin{figure}
+        \includegraphics[width=\linewidth]{gfx/ch_5/5.png}
+        \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b)}\label{fig:xp1_2}
+\end{figure}
+
 % L'objectif est d'étudier les influences spécifiques des différentes sources sonores qui composent les environnements urbains sur la perception de l'agrément, en utilisant la simulation.
 Pour ce faire, nous planifions nos deux premières expériences (\cf~Figure~\ref{fig:xp1_2}):
-%
+
+Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness.
+For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
+
 % \begin{itemize}
@@
 Chacun compose deux environnements sonores, le premier idéal/agréable et le deuxième non-idéal/désagréable.
-%
 % \item \emph{expérience 1.b d'évaluation}: à l'issue de la simulation, nous n'avons, de fait, qu'une connaissance binaire des propriétés affectives des scènes simulées: idéale (i) et non-idéale (ni).
@@
 L'expérience d'évaluation sert deux buts:
-%
 % \begin{enumerate}
@@
 % \end{enumerate}
-%
 % \end{itemize}
-%
-% Notre analyse s'appuie sur les données produites par les deux expériences.
-
-Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness.
-For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
-\begin{itemize}
-\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}.
-Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
+
+\begin{itemize}
+\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to simulate urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}.
+Each of them has to create two sound environments: one ideal / pleasant and the other one, non ideal / unpleasant.
 \item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni).
 Furthermore, this information is given by the creator of the scene.
-The second experimental step aims at investigating more deeply and more broadly our knowledge on pleasantness.
-For that, a second group of participants is asked to evaluate the pleasantness of each simulated scene, on a semantic scale.
+The second experimental step aims at investigating more deeply and more broadly our knowledge on the pleasantness of the simulated scenes.
+For that, a second group of participants is asked to evaluate the pleasantness of each scene, on a semantic scale.
 This experiment has two goals:
@@
 \item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene.
-\item to detect the presence of \emph{outliers} or ambiguous samples.
-For the remainder of the study, the defined hedonic properties (i and ni) are used as references.
-We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of perceived pleasantness.
+\item to detect the presence of \emph{outliers} or ambiguous scenes.
+Indeed, throughout our analyses, the predefined hedonic properties of the scenes (i and ni) are used as groundtruth.
+We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes, \ie~that the less pleasant i-scene remains above the more pleasant ni-scene.
 \end{enumerate}
@@
 
-Our analysis is based on the data collected by these two experiments: 1.a and 1.b.
-
-\begin{figure}
-        \includegraphics[width=\linewidth]{gfx/ch_5/5.png}
-        \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b)}\label{fig:xp1_2}
-\end{figure}
+% Notre analyse s'appuie sur les données produites par les deux expériences.
+
+The data collected by these two experiments (1.a and 1.b) are analyzed conjointly.
 
@@
 % Les sujets doivent simuler deux environnements sonores urbains, chacune des scènes devant durer 1 minute. Ils doivent se conformer aux consignes suivantes:
-%
+
+The participants have to simulate two urban sound environments of one minute each. They follow the instructions below:
+
 % \begin{itemize}
@@
 % \end{itemize}
-%
+
+\begin{itemize}
+\item  first simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
+\item second simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
+\end{itemize}
+
 % Tous commencent par simuler l'environnement idéal. Ils ne prennent connaissance de la deuxième consigne qu'à la fin de la première simulation.
-%
+
 % Les sujets sont totalement libres dans le choix des sons, et des paramètres. Ils doivent cependant se soumettre à deux contraintes:
-%
+
+All the participants start by designing the ideal environment; they read the second set of instruction at the end of the first experiment.
+Participants are completely free of their choices concerning sounds and synthesis parameters.
+They must nevertheless respect the two following constraints :
+
 % \begin{itemize}
@@
 % \end{itemize}
-%
+
+\begin{itemize}
+\item the listening point of view is the one of a fixed listener;
+\item the soundscape must be realistic, \ie~physically plausible. For instance, participants are free to insert ten dogs in the soundscape but they cannot insert one dog barking every 10 milliseconds.
+\end{itemize}
+
 % Ces contraintes font partie de la consigne. Aucun contrôle n'est fait \emph{a priori} dans l'interface de simulation.
-
-The participants have to simulate two urban sound environments of one minute each. They follow the instructions below:
-\begin{itemize}
-\item  first simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
-\item second simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
-\end{itemize}
-
-All the participants start by designing the ideal environment; they read the second set of instruction at the end of the first experiment.
-Participants are completely free of their choices concerning sounds and synthesis parameters.
-They must nevertheless respect the two following constraints :
-\begin{itemize}
-\item the listening point of view is the one of a fixed listener;
-\item the soundscape must be realistic, i.e. physically plausible. For instance, the participant can insert ten dogs in the soundscape but she cannot insert one dog barking every 10 milliseconds.
-\end{itemize}
 
@@
 % \end{enumerate}
-%
+
+Each simulation process comprises two parts:
+
+\begin{enumerate}
+\item the simulation: this step is split in three steps:
+\begin{itemize}
+  \item  select the class of sounds
+	\item  give a name to the selected class of sounds
+	\item  set the parameters of the tracks related to the selected class of sounds, see Section~\ref{sec:simscene_parametre}.
+\end{itemize}
+\item the production of a free comment of the composed soundscape
+\end{enumerate}
+
 % En complément, et une fois les deux scènes sonores réalisées, les sujets sont invités à:
@@
 % \end{itemize}
-%
-% Avant de commencer la première simulation, un tutoriel de 20 minutes est proposé aux sujets, afin qu'ils se familiarisent avec le logiciel de simulation, et la banque de données.
-L'expérience est prévue pour durer 2h30.
-
-Each simulation process contains two parts:
-\begin{enumerate}
-\item composition of the simulation: this step is split in three steps:
-\begin{itemize}
-  \item  select the class of sounds
-	\item  give a name to the selected class of sounds
-	\item  set the parameters of the tracks related to the selected class of sounds, see Section~\ref{sec:simscene_parametre}.
-\end{itemize}
-\item production of a free verbalization of the composed soundscape
-\end{enumerate}
-
-In addition, once the two sound scenes are done, participants are invited to
+
+In addition, once the two sound scenes are done, participants are invited to:
+
 \begin{itemize}
@@
 
-Before starting the first simulation, a 20-minute tutorial is given to the participants in order to familiarize them with the simulation environment and the sound database.
+% Avant de commencer la première simulation, un tutoriel de 20 minutes est proposé aux sujets, afin qu'ils se familiarisent avec le logiciel de simulation, et la banque de données.
+L'expérience est prévue pour durer 2h30.
+
+Before starting the first simulation, a 20-minute tutorial is given to the participants in order to familiarize them with the simulation interface and the sound database.
 The experiment is planned to last about two hours and a half, including breaks that could be taken by the participants.
@@
 Ils ne peuvent le modifier par la suite.
-%
+
+All the participants do the experiment on desktop computers with the same hardware and software configurations.
+The audio files are played in diotic conditions by headphones.
+During the tutorial, participants are asked to adjust the global sound volume to a comfortable level.
+Once set, they are not allowed to modify it in the remaining of the experiment.
+
 % Tous les sujets réalisent l'expérience simultanément.
@@
 Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
-%
+
+All the participants do the experiment at the same time. They are equally split in three identical quiet rooms. They are not allowed to talk to each other during the experiment.
+
 % Trois expérimentateurs, un dans chaque pièce, sont présents durant la totalité de l'expérience, afin de contrôler le bon déroulement de cette dernière, et de répondre aux éventuelles questions des sujets.
 
-All the participants do the experiment on desktop computers with the same hardware and software configurations.
-The audio files are played in diotic conditions by headphones.
-During the tutorial, participants are asked to adjust the sound level in order to be comfortable.
-Once set, they are not allowed to modify it in the remaining of the experiment.
-All the participants do the experiment at the same time. They are equally split in three identical quiet rooms. They are not allowed to talk to each other during the experiment.
 Three experimenters (one in each room) are present during the whole duration of the experiment in order to assist participants with potential hardware and software issues and to answer questions.
@@
 Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
-%
+
+44 students (14 females; averaging 21.6 years of age, s.d.
+of 2.0 years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment.
+They are all living in downtown Nantes (a french city) since 2 years or more.
+
 % Sur les 44 sujets, 40 réalisent l'expérience avec succès, produisant au final 80 scènes sonores simulées, dont 40 scènes idéales, et 40 scènes non idéales.
@@
 
-44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in downtown Nantes since 2 years or more.
 Among the 44 participants, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
@@
 % En raison de contraintes temporelles, les sujets n'évaluent que des séquences de 30 secondes des scènes simulées, chacune de ces séquences commençant à la seconde 15, et finissant à le seconde 45 de la scène d'une durée initiale de 1 minute.
-%
+
+Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from sec. 15 to sec. 45).
+
 % L'évaluation s'effectue sur une échelle sémantique bipolaire de 7 points, allant de -3 (non-idéale/très désagréable) à +3 (idéale/très agréable).
@@
 Après la notation, ils sont libres de passer à la scène suivante.
-%
+
+The assessment is done with a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
+Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli.
+After the evaluation, they are free to switch to the next scene.
+
 % Pour chaque sujet, les scènes sont présentées dans un ordre aléatoire.
@@
 Ces 10 premières scènes sont rejouées à la fin de l'expérience, et seules les notes données à la deuxième occurrence sont prises en compte.
-%
-% L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.
-
-Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes composed in Experiment 1.a (from sec. 15 to sec. 45).
-The assessment is done with a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
-Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli.
-After the evaluation, they are free to switch to the next scene.
+
 For each participant, sound scenes are played in a random order.
@@
 These first 10 scenes are played back again at the end of the experiment and the last evaluations are the ones that are only taken into account in the results.
+
+% L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.
+
 The experiment is planned to last 30 minutes. The participants do not know anything beforehand about the nature of the sound scenes.
@@
 Le niveau sonore de sortie est identique pour tous les sujets.
-%
-% Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
-%
-% Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de celle-ci, et de répondre aux éventuelles questions des sujets.
 
@@
 The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}.
-The stimuli are the scenes obtained by the simulation experiment of Exp.
-1.a.
+The stimuli are the scenes obtained by the simulation experiment of experiment 1.a.
 The output sound level is the same for all the participants.
+
+% Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
+
 All the participants do the experiment simultaneously. They are not allowed to talk to each other during the experiment.
+
+% Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de celle-ci, et de répondre aux éventuelles questions des sujets.
+
 An experimenter is present during the whole duration of the experiment in order to assist participants and possibly answer questions.
@@
 Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
-%
-% Tous les sujets ont réalisé l'expérience avec succès.
-
-10 students (2 females - mean age 23.1, SD = 1.8) from Ecole Centrale de Nantes took part to the experiment.
-All are living in downtown Nantes for 2 years or more.
+
+10 students (2 females; averaging 23.1 years of age, s.d.
+of 1.8 years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment.
+They are living in downtown Nantes (a french city) since 2 years or more.
 None of them took part to the previous simulation experiment (Exp.
 1.a).
+
+% Tous les sujets ont réalisé l'expérience avec succès.
+
 All the participants succeeded in doing the experiment.
 
-\subsection{Data and analysis methods}
+\subsection{Data and statistical analysis}
 \label{sec:xp1_dataAna}
+
+\begin{table}[t]
+\centering
+\begin{tabular}{c c}
+Terms                         & Acronyms              \\
+\hline
+sound Level                        & $L$                    \\
+sound Level (events)            & $L(E)$                 \\
+sound Level (textures)              & $L(T)$                 \\
+Average pleasantness (per scene)     & $\mathcal{A}_{scene}$  \\
+Average pleasantness (per subject)     & $\mathcal{A}_{sujet}$  \\
+                               &                        \\
+ideal/pleasant                & i                      \\
+non-ideal/unpleasant         & ni                     \\
+scene ideal/pleasant          & i-scene                \\
+scene non-ideal/unpleasant   & ni-scene               \\
+\hline
+\end{tabular}
+\vspace{0.5mm}
+\caption{Acronyms of features used in the analysis of the sensory experiments.}
+\label{tab:acronyme}
+\end{table}
 
@@
 Afin de rester cohérent avec l’expérience 1.b d'évaluation, les descripteurs ne sont pas calculés sur la durée totale des scènes, mais sur la version réduite de 30 secondes (\cf~Section~\ref{sec:xp1b_plan}).
-%
+
+A set of features is attached to each sound scene.
+The analysis is done on the basis of these features.
+A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}.
+In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section~\ref{sec:xp1b_plan}.
+
 % Pour chaque scène sonore, trois types de descripteurs sont considérés:
-%
+
+For each sound scene, three types of features are considered :
+
 % \begin{itemize}
@@
 
-A set of features is attached to each sound scene.
-The analysis is done on the basis of these features.
-A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}.
-In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.
-
-For each sound scene, three types of features are considered :
-
-\begin{itemize}
-\item \emph{perceptual}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale.
+\begin{itemize}
+\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale.
 $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants.
 $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant.
-Given the low number of participants in Exp.
-1.b, we choose not to normalize the pleasantness scores.
-\item \emph{semantic}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene.
+Given the low number of participants in experiment 1.b, we choose not to normalize the pleasantness scores.
+
+\item \emph{semantic features}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene.
 Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise.
@@
 For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.
-\item \emph{structural}: features related to \emph{sound level}.
+
+\item \emph{structural features}: features related to the \emph{sound level}.
 To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference.
@@
 If $0.01\leq p<0.05$, we only report $p<0.05$, otherwise we report $p<0.01$.
-
-\begin{table}[t]
-\centering
-\begin{tabular}{c c}
-Terms                         & Acronyms              \\
-\hline
-Level                        & $L$                    \\
-Level (events)            & $L(E)$                 \\
-Level (textures)              & $L(T)$                 \\
-Average pleasantness (per scene)     & $\mathcal{A}_{scene}$  \\
-Average pleasantness (per subject)     & $\mathcal{A}_{sujet}$  \\
-                               &                        \\
-ideal/pleasant                & i                      \\
-non-ideal/unpleasant         & ni                     \\
-scene ideal/pleasant          & i-scene                \\
-scene non-ideal/unpleasant   & ni-scene               \\
-\hline
-\end{tabular}
-\vspace{0.5mm}
-\caption{Acronyms of features used in the analysis of the sensory experiments.}
-\label{tab:acronyme}
-\end{table}
 

### 2017-09-10 lafaySimulatedScenePerception.tex
add new corr value, add figure XP2, samll correction

@@
 \newcommand{\gl}[1]{\textcolor{magenta}{GL : #1}}
-\newcommand{\nm}[1]{\textcolor{magenta}{#1}}
-\newcommand{\ml}[1]{\textcolor{red}{#1}}
+\newcommand{\mr}[1]{\textcolor{green}{MR : #1}}
+\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
 \newcommand{\jfp}[1]{\textcolor{blue}{JFP : #1}}
@@
 This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes.
-A rating experiment of the soundscape pleasantness using a 7-points Likert scale is conducted to further refine the analysis of the simulated urban acoustic scenes.
+A rating experiment of the soundscape pleasantness using a 7-point bipolar semantic scale is conducted to further refine the analysis of the simulated urban acoustic scenes.
 Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.
@@
   %% keywords here, in the form: keyword \sep keyword
- cognitive psychology \sep soundscape perception \ soundscape simulator
+ cognitive psychology \sep soundscape perception \sep soundscape simulator
   %% PACS codes here, in the form: \PACS code \sep code
@@
 
-\begin{figure}
+\begin{figure}[t]
         \myfloatalign
@@
 
-\begin{figure}
-		\centering
+\begin{figure}[!tp]
+		\myfloatalign
         \subfloat[]
@@
 
-\begin{figure}
+\begin{figure}[t]
         \includegraphics[width=\linewidth]{gfx/ch_5/5.png}
@@
 
-44 students (14 females; averaging 21.6 years of age, s.d.
-of 2.0 years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment.
-They are all living in downtown Nantes (a french city) since 2 years or more.
+44 students (14 females; averaging $21.6$ years of age, s.d.
+of $2.0$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment.
+All the subjects live in Nantes, France, for at least two years
 
@@
 
-The assessment is done with a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
+The assessment is done with a 7-point bipolar semantic scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
 Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli.
@@
 
-10 students (2 females; averaging 23.1 years of age, s.d.
-of 1.8 years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment.
-They are living in downtown Nantes (a french city) since 2 years or more.
-None of them took part to the previous simulation experiment (Exp.
-1.a).
+10 students (2 females; averaging $23.1$ years of age, s.d.
+of $1.8$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment.
+All the subjects live in Nantes, France, for at least two years.
+None of them took part to the previous simulation experiment (experiment 1.a).
 
@@
 \hline
-sound Level                        & $L$                    \\
-sound Level (events)            & $L(E)$                 \\
-sound Level (textures)              & $L(T)$                 \\
+Sound level                        & $L$                        \\
+Sound level (events)               & $L(E)$                     \\
+Sound level (textures)             & $L(T)$                     \\
 Average pleasantness (per scene)   & $\mathcal{A}_{scene}$      \\
-Average pleasantness (per subject)     & $\mathcal{A}_{sujet}$  \\
+Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
                                    &                            \\
@@
 \begin{itemize}
-\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale.
+\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale.
 $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants.
@@
 
-\item \emph{semantic features}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene.
+%\jfp{ce qu'est Ascene dans ce paragraphe n'est pas clair :A scene est défini comme la moyenne des scores d'une scène donnée pour tous les participants.
+le test, lui, porte sur la moyenne des Ascene pour les deux catégories meme chose pour les sujets}
+
+\item \emph{semantic features}: the sound classes that are present / absent from the scene.
+We use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene.
 Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise.
@@
 
-\item \emph{structural features}: features related to the \emph{sound level}.
-To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference.
+\item \emph{structural features}: the \emph{sound level}.
+To figure out the sound level, we draw inspiration from the $L_{Aeq}$ feature.
 In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale).
@@
 
-%\jfp{ce qu'est Ascene dans ce paragraphe n'est pas clair :A scene est défini comme la moyenne des scores d'une scène donnée pour tous les participants.
-le test, lui, porte sur la moyenne des Ascene pour les deux catégories meme chose pour les sujets}
-
 \item \emph{analysis of the structural features}: the goal is to evaluate whether the structural features differ between the i- and ni-scenes.
@@
 \item \emph{analysis of the semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni).
-A V-test is considered \citep{lebart2000statistique} \ml{y pas une ref en Anglais ?}.
-With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of time a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significatively different from the ratio $\frac{c_{jk}}{c_k}$.
+A V-test is considered \citep{lebart2000statistique} \ml{ref en Anglais ?}.
+With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of time a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
 For each class $j$, and each environment type $k$, an approximation of the statistic value $V_{jk}$ is computed as follows:
@@
 
-If the hypothesis is rejected, the class $j$ is said to be typical with respect to the type of scene $k$.
+If the null hypothesis is rejected, the class $j$ is said to be typical with respect to the type of scene $k$.
 The typical classes are called \textbf{sound markers}.
@@
 First, in order to ensure the coherence of the data, we check that none of the ni-scenes gets a $\mathcal{A}_{scene}$ higher than one of a i-scene.
-Four ni-scenes don’t fullfill that constraint: they are thus removed, together with their corresponding i-scenes.
+4 ni-scenes don't fulfill that constraint: they are thus removed, together with their corresponding i-scenes.
 As a consequence, 36 i-scenes and 36 ni-scenes remain for analysis.
@@
 For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment.
-It appears that the i-scenes were perceived significatively more pleasant ($p < 0.01$) than the ni-scenes.
+It appears that the i-scenes were perceived significantly more pleasant ($p < 0.01$) than the ni-scenes.
 
 \subsubsection*{Analysis of the structural features}
+
+\begin{figure}[t]
+		\myfloatalign
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_1}\label{fig:soundlevela}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_3}\label{fig:soundlevelb}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_5}\label{fig:soundlevelc}}\par
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_2}\label{fig:soundleveld}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_4}\label{fig:soundlevele}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_6}\label{fig:soundlevelf}}
+       \caption{Distributions of the structural features $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
+\end{figure}
 
@@
 Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the levels $L$, $L(E)$ and $L(T)$.
-There is a significative difference in terms of sound level between i- and ni-scenes ($L$: $p<0.01$).
+There is a significant difference in terms of sound level between i- and ni-scenes ($L$: $p<0.01$).
 This difference is siginificant for events ($L(E)$: $p<0.01$) and for textures ($L(T)$: $p<0.01$).
@@
 During their experiment, the authors ask their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source.
-The study shows that there is no significative difference between global and averaged instantaneous judgments.
+The study shows that there is no significant difference between global and averaged instantaneous judgments.
 In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, \ie the events.
@@
 
-\begin{figure*}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_1}\label{fig:soundlevela}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_3}\label{fig:soundlevelb}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_5}\label{fig:soundlevelc}}\par
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_2}\label{fig:soundleveld}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_4}\label{fig:soundlevele}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_6}\label{fig:soundlevelf}}
-       \caption{Distributions of the structural features $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
-\end{figure*}
-
 \subsubsection*{Influence of the structural features on the perceived pleasantness}
+
+\begin{table}[t]
+\centering
+\begin{tabular}{l c c c}
+               & all scenes                & i-scenes          & ni-scenes    \\
+\hline
+$L$            & \textbf{-0.77} ($p<0.01$) & -0.32 ($p=0.06$)  & \textbf{-0.78} ($p<0.01$)\\
+$L(E)$         & \textbf{-0.75} ($p<0.01$) & -0.20 ($p=0.24$)  & \textbf{-0.75} ($p<0.01$)\\
+$L(T)$         & \textbf{-0.53} ($p<0.01$) & -0.33 ($p=0.05$)  &  -0.00 ($p=0.99$) \\
+\hline
+\end{tabular}
+\vspace{0.5mm}
+\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and structural features.}
+\label{tab:corrStructA}
+\end{table}
 
@@
 The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown on Table~\ref{tab:corrStructA}.
-Relationships between $\mathcal{A}_{scene}$ and the structural features are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and ~\ref{fig:soundlevelf}.
-\jfp{en table 2, il devrait y avoir les coefficients de corrélation pour toutes les données, puis pour les deux cas ni et i.
-}
+Relationships between $\mathcal{A}_{scene}$ and the structural features are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.
 
@@
 
-\begin{table}[t]
-\centering
-\begin{tabular}{l c c c}
-               & i-scenes                   & ni-scenes    \\
-\hline
-$L$            & -0.32 ($p=0.06$)           & \textbf{-0.78} ($p<0.01$)\\
-$L(E)$         & -0.20 ($p=0.24$)           & \textbf{-0.75} ($p<0.01$)\\
-$L(T)$         & -0.33 ($p=0.05$)           &  -0.00 ($p=0.99$) \\
-\hline
-\end{tabular}
-\vspace{0.5mm}
-\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and structural features.\jfp{rajouter une colonne corélation globale}}
-\label{tab:corrStructA}
-\end{table}
-
 \subsubsection*{Analysis of the semantic features}
+
+\begin{figure}[!tp]
+		\myfloatalign
+        \subfloat[]
+        {\includegraphics[width=1\linewidth]{gfx/ch_5/xp1_class_1}\label{fig:soundsourcea}}\par
+        \subfloat[]
+        {\includegraphics[width=.5\linewidth]{gfx/ch_5/xp1_class_2}\label{fig:soundsourceb}}
+        \subfloat[]
+        {\includegraphics[width=.5\linewidth]{gfx/ch_5/xp1_class_3}\label{fig:soundsourcec}}
+       \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.}\label{fig:soundsource}
+\end{figure}
 
@@
 
-\begin{figure}
-        \subfloat[]
-        {\includegraphics[width=1\linewidth]{gfx/ch_5/xp1_class_1}\label{fig:soundsourcea}} \par
-        \subfloat[]
-        {\includegraphics[width=.5\linewidth]{gfx/ch_5/xp1_class_2}\label{fig:soundsourceb}}
-        \subfloat[]
-        {\includegraphics[width=.5\linewidth]{gfx/ch_5/xp1_class_3}\label{fig:soundsourcec}}
-       \caption{Proportion of synthesized scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.}\label{fig:soundsource}
-\end{figure}
-
 \subsubsection*{Sound markers}
 
-We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs with respect to the type of environment (i or ni).
-Let us now investigate whether some of the sound classes are specific to a given environment.
-For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna} separately for each abstraction level.
-Results are presented in Table~\ref{tab:markers}.
-\jfp{le V-test est un test statistique : on peut mettre la p-value dans le tableau}
-
-Regarding the sound events, 9 markers are identified for all abstraction levels.
-As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i-scenes markers as well as the \emph{church bell} class.
-This latter result may be due to the socio-cultural background of the participants who are mostly European citizens.
-In fact, according to Schafer, a sound that is identified by a person as being an element of his/her environment, is well accepted.
-Sound markers of ni-scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
-
-Regarding the textures, 5 markers are identified.
-For the i-scenes, they are classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}).
-For the ni-scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).
-
-Although the whole set of identified markers are rather intuitive, none of the event classes related to the noise of the motor vehicles are identified as markers, except for the texture class \emph{crossroad}.
-To simulate an unpleasant traffic, participants chose the classes \emph{horn} or \emph{siren}.
-We thus conclude that isolated motor vehicle sounds are understood as being part of the urban environment, and thus their nature is not necessarily linked to an unpleasant soundscape.
-
-\begin{table*}[t]
+\begin{table}[t]
  \setlength{\tabcolsep}{0.2pt}
@@
 \label{tab:markers}
-\end{table*}
+\end{table}
+
+We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs with respect to the type of environment (i or ni).
+Let us now investigate whether some of the sound classes are specific to a given environment.
+For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna} separately for each abstraction level.
+Results are presented in Table~\ref{tab:markers}.
+
+Regarding the sound events, 9 markers are identified for all abstraction levels.
+As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i-scenes markers as well as the \emph{church bell} class.
+This latter result may be due to the socio-cultural background of the participants who are mostly European citizens.
+In fact, according to Schafer, a sound that is identified by a person as being an element of his/her environment, is well accepted.
+Sound markers of ni-scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
+
+Regarding the textures, 5 markers are identified.
+For the i-scenes, they are classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}).
+For the ni-scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).
+
+Although the whole set of identified markers are rather intuitive, none of the event classes related to the noise of the motor vehicles are identified as markers, except for the texture class \emph{crossroad}.
+To simulate an unpleasant traffic, participants chose the classes \emph{horn} or \emph{siren}.
+We thus conclude that isolated motor vehicle sounds are understood as being part of the urban environment, and thus their nature is not necessarily linked to an unpleasant soundscape.
 
 \subsubsection*{Representation space induced by semantic features}
+
+\begin{figure}[t]
+        \myfloatalign
+        \includegraphics[width=\linewidth]{gfx/ch_5/pa5_1_en}
+       \caption{$P@5$ obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors as a function of the abstraction level.
+The vectors are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers $E_{w/o,m}$.}\label{fig:pa5}
+\end{figure}
 
@@
 
-\begin{figure}[t]
-        \myfloatalign
-        \includegraphics[width=\linewidth]{gfx/ch_5/pa5_1_en}
-       \caption{$P@5$ obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors as a function of the abstraction level.
-The vectors are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers $E_{w/o,m}$.}\label{fig:pa5}
-\end{figure}
-
 \subsubsection*{Influence of the sound markers on the perceived pleasantness}
-
-We evaluate the correlations between $\mathcal{A}_{scene}$ and the structural features.
-In this section, the structural features are computed by taking into account only the previously identified sound markers.
-We define $X_m$, the $X$ feature computed by taking into account only the sound markers.
-On the opposite, we define $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
-When the feature characterizes an i-scene (idem for a ni-scene),  only the markers identified for the i-scenes (or the ni-scenes) are considered; we named them, i- markers (or ni- markers).
-Results are shown in Table~\ref{tab:corrMarkers}.
-
-Concerning the sound levels, the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ and $L(T)$ on the other side, see~Figures~\ref{fig:soundlevelMarkera} and~\ref{fig:soundlevelMarkerd}, respectively.
-No matter whether all the classes or only the markers are considered, it appears that:
-
-\begin{enumerate}
-\item a significative difference between levels of i- and ni-scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
-\item the sound level of scenes is mainly related to the sound events, compared to the textures;
-\item the sound level of events has an influence on the perception of pleasantness for ni-scenes, but not for i-scenes;
-\item the sound level of textures does not play any role in the perception of the pleasantness.
-\end{enumerate}
-
-In conclusion, the level of ni- markers has a negative influence on the pleasantness for the ni-scenes.
-On the other hand, the level of i- markers does not influence the perceived pleasantness for the i-scenes.
-
-% En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
-C'est la première fois qu'un indicateur objectif nous permet de préciser l'agrément des environnements agréables.
-Ceci nous amène à conclure que le niveau des classes de sons n'étant pas typiques d'un environnement agréable a un impact négatif sur l'agrément.
-
-Considering the non markers classes, we can observe on the i-scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}.
-It is the first time that an objective feature allows us to define the pleasantness of i-scenes.
-This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.
-
-% Par ailleurs, alors que $L(T)$ ne présentait pas de corrélation pour les ni-scènes, une corrélation négative forte est observée pour $L(T)_b$ ($r-0.73$, $p<0.01$).
-Ce fait indique que les niveaux des classes de textures n'étant pas des marqueurs n'affectent pas l'agrément perçu de la même manière pour les i- et ni-scènes.
-Les niveaux semblent avoir un effet négatif pour les ni-scènes, alors que pour les i-scènes, aucun effet n'est relevé.
-
-Besides, whereas $L(T)$ did not show any significative correlation for ni-scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
-This indicates that the level of non-markers texture classes does not influence the perceived pleasantness in the same way for i- and ni-scenes.
-Sound level seems to have a negative effect for the ni-scenes, and no significant effect for the i-scenes.
-
-% Pour finir, nous considérons un dernier groupe de descripteurs, nommément $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiffa} et~\ref{fig:soundlevelMarkerDiffd}).
-Ces descripteurs expriment la différence entre les niveaux des marqueurs, et ceux des autres classes de sons.
-Ils traduisent l'émergence des marqueurs par rapport à la mixture sonore.
-
-We consider a last group of features, namely $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$.
-These features describe the difference between the markers level and those of the other sound classes, see ~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}.
-They express the saliency of the markers with respect to the sound mixture.
-
-% Pour les i-scènes, une corrélation modérée et positive est observée pour $L_m-L_b$ ($r=0.67$, $p<0.01$) et $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
-Pour les ni-scènes, aucune corrélation n'est observée.
-Dans le cas des i-scènes, ce n'est donc pas le niveau absolu des marqueurs qui importe, mais leur niveau relatif, par rapport aux autres sons qui composent la scène.
-On observe donc pour les environnements idéaux un double mécanisme perceptif:
-
-For the i-scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
-for the ni-scenes no correlation is observed.
-Thus, for the i-scenes, it is not the absolute markers level that is important, but their relative level with regards to the other sounds composing the scene.
-Then, we can induce a double perceptual mechanism for the ideal environments:
-
-\begin{itemize}
-\item the higher the absolute level of sounds not being i- markers, the weaker the pleasantness,
-\item the higher the relative level of i- markers is, with regards to the other sounds, the higher the pleasantness.
-\end{itemize}
-
-For the ni-scenes, the fact that we observe significative correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.
 
@@
         \caption{Distribution of the relative sound levels structural features related to the presence of markers $L_m$ (a, d), $L_b$ (b, e) and $L_m-L_b$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
-\end{figure*}
+\end{figure}
+
+We evaluate the correlations between $\mathcal{A}_{scene}$ and the structural features.
+In this section, the structural features are computed by taking into account only the previously identified sound markers.
+We define $X_m$, the $X$ feature computed by taking into account only the sound markers.
+On the opposite, we define $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
+When the feature characterizes an i-scene (idem for a ni-scene),  only the markers identified for the i-scenes (or the ni-scenes) are considered; we named them, i- markers (or ni- markers).
+Results are shown in Table~\ref{tab:corrMarkers}.
+
+Concerning the sound levels, the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ and $L(T)$ on the other side, see~Figures~\ref{fig:soundlevelMarkera} and~\ref{fig:soundlevelMarkerd}, respectively.
+No matter whether all the classes or only the markers are considered, it appears that:
+
+\begin{enumerate}
+\item a significant difference between levels of i- and ni-scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
+\item the sound level of scenes is mainly related to the sound events, compared to the textures;
+\item the sound level of events has an influence on the perception of pleasantness for ni-scenes, but not for i-scenes;
+\item the sound level of textures does not play any role in the perception of the pleasantness.
+\end{enumerate}
+
+In conclusion, the level of ni- markers has a negative influence on the pleasantness for the ni-scenes.
+On the other hand, the level of i- markers does not influence the perceived pleasantness for the i-scenes.
+
+% En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
+C'est la première fois qu'un indicateur objectif nous permet de préciser l'agrément des environnements agréables.
+Ceci nous amène à conclure que le niveau des classes de sons n'étant pas typiques d'un environnement agréable a un impact négatif sur l'agrément.
+
+Considering the non markers classes, we can observe on the i-scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}.
+It is the first time that an objective feature allows us to define the pleasantness of i-scenes.
+This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.
+
+% Par ailleurs, alors que $L(T)$ ne présentait pas de corrélation pour les ni-scènes, une corrélation négative forte est observée pour $L(T)_b$ ($r-0.73$, $p<0.01$).
+Ce fait indique que les niveaux des classes de textures n'étant pas des marqueurs n'affectent pas l'agrément perçu de la même manière pour les i- et ni-scènes.
+Les niveaux semblent avoir un effet négatif pour les ni-scènes, alors que pour les i-scènes, aucun effet n'est relevé.
+
+Besides, whereas $L(T)$ did not show any significant correlation for ni-scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
+This indicates that the level of non-markers texture classes does not influence the perceived pleasantness in the same way for i- and ni-scenes.
+Sound level seems to have a negative effect for the ni-scenes, and no significant effect for the i-scenes.
+
+% Pour finir, nous considérons un dernier groupe de descripteurs, nommément $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiffa} et~\ref{fig:soundlevelMarkerDiffd}).
+Ces descripteurs expriment la différence entre les niveaux des marqueurs, et ceux des autres classes de sons.
+Ils traduisent l'émergence des marqueurs par rapport à la mixture sonore.
+
+We consider a last group of features, namely $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$.
+These features describe the difference between the markers level and those of the other sound classes, see ~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}.
+They express the saliency of the markers with respect to the sound mixture.
+
+% Pour les i-scènes, une corrélation modérée et positive est observée pour $L_m-L_b$ ($r=0.67$, $p<0.01$) et $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
+Pour les ni-scènes, aucune corrélation n'est observée.
+Dans le cas des i-scènes, ce n'est donc pas le niveau absolu des marqueurs qui importe, mais leur niveau relatif, par rapport aux autres sons qui composent la scène.
+On observe donc pour les environnements idéaux un double mécanisme perceptif:
+
+For the i-scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
+for the ni-scenes no correlation is observed.
+Thus, for the i-scenes, it is not the absolute markers level that is important, but their relative level with regards to the other sounds composing the scene.
+Then, we can induce a double perceptual mechanism for the ideal environments:
+
+\begin{itemize}
+\item the higher the absolute level of sounds not being i- markers, the weaker the pleasantness,
+\item the higher the relative level of i- markers is, with regards to the other sounds, the higher the pleasantness.
+\end{itemize}
+
+For the ni-scenes, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.
 
@@
 
-%\jfp{on se demande pourquoi on va de -5 a +5 alors qu'avant on allait de -3 à +3}
-
 For each subject, the scenes are presented in a random order.
@@
 
-12 subjects (4 women) perform the test.
+12 subjects perform the test (4 females; averaging $29.5$ years of age, s.d.
+of $14$ years).
 None of them have performed experiments 1.a and 1.b.
-The subjects are 22 to 61 years old (average: 29.5, standard deviation: 14).
 All the subjects live in Nantes, France, for at least two years.
@@
 
+\begin{figure}[t]
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp4_note_13}\label{fig:xp2_Asubject}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp4_note_15}\label{fig:xp2_Ascene}}
+        \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness  per scenes$\mathcal{A}_{scene}$ (b) versus scene type.
+Black stars in subfigure (a) stand for the detected outlier,~\ie~subject 7.}\label{fig:xp2A}
+\end{figure}
+
 %Considérons $\mathcal{A}_{sujet}$ pour les am-scènes.
@@
 
-Let us consider $\mathcal{A}_{sujet}$ for the km-scenes.
+Let us consider $\mathcal{A}_{subject}$ for the km-scenes, see Figure~\ref{fig:xp2_Asubject}.
 Close inspection of the answers show that one subject strongly differs from the ones of others.
@@
 
-In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes and ni/rm-scenes.
-The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ show a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
+In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes and ni/rm-scenes, see Figure~\ref{fig:xp2_Ascene}.
+The repeated measure ANOVA applied to $\mathcal{A}_{subject}$ show a significant effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
 
@@
 
-The \emph{post hoc} analysis exhibits significative differences between all the groups of observations, notably between the i/km-scenes ($p<0.05$) and the i/rm-scenes ($p<0.01$).
-
-%\jfp{je ne comprends pas les 2 p-value. il ne peut y a voir qu'une seule p-value lorsqu'on compare deux conditions expérimentales}
+The \emph{post hoc} analysis exhibits significant differences between all the groups of observations, notably between the i/km- and the i/rm scenes ($p<0.05$) as well as between the ni/km- and the ni/rm-scenes ($p<0.01$).
 
@@
 
-These results indicate that the removal of the markers did indeed modify the perception of the scenes by the subjects.
-Our two hypotheses are thus verified \jfp{avant cela il faut dire un mot sur le sens des effets : ils vont bien dans le sens souhaité : ce n'est pas le test qui le dit}:
+These results indicate that the removal of the markers did indeed modify the perception of the scenes by the subjects. Our two hypotheses are thus verified :
 
@@
 
-The significative interaction shows that the effect of the type of environment influences the effect of the presence/absence of the markers.
+The significant interaction shows that the effect of the type of environment influences the effect of the presence/absence of the markers.
 Indeed, the average of the differences between km-scenes and rm-scenes is larger for the ni-scenes ($1.1$) than for the i-scenes ($0.5$).
@@
 \bibliography{Bibliography}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\onecolumn
-
 \appendix
+\newpage
 \section{Taxonomy of sound classes}
@@
 
-\begin{figure}[h]
+\begin{figure}[hp]
     \centering
@@
 
-\begin{figure}[h]
+\begin{figure}[hp]
   \centering
@@
 \newpage
-\section{Distribution of scores given by the subjects during experiment 2}
+\section{Distribution of pleasantness scores given by the subjects during experiment 2}
 \label{app:xp2}

### 2017-12-13 lafaySimulatedScenePerception.tex
Mise à jour

@@
         \myfloatalign
-        \includegraphics[trim={0 0 0 2.5cm},clip,width=.8\linewidth]{gfx/ch_4/3.png}
-       \caption{Hierarchical organisation of the isolated sounds used in the simulation.}\label{fig:orgDb}
+        \includegraphics[width=.8\linewidth]{gfx/3.pdf}
+       \caption{Hierarchical organisation of the isolated sounds used in the simulation.\gl{TODO : translate and remove subject box}}\label{fig:orgDb}
 \end{figure}
@@
         \subfloat[]
-        {\includegraphics[width=.8\linewidth]{gfx/ch_4/SSF}\label{fig:ssf}} \par
+        {\includegraphics[width=.8\linewidth]{gfx/SSF}\label{fig:ssf}} \par
         \subfloat[]
-        {\includegraphics[width=\linewidth]{gfx/ch_4/simscene}\label{fig:simscene}}
+        {\includegraphics[width=\linewidth]{gfx/simscene}\label{fig:simscene}}
        \caption{Graphical interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{simScene} tool.}
@@
 \begin{figure}[t]
-        \includegraphics[width=\linewidth]{gfx/ch_5/5.png}
-        \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b)}\label{fig:xp1_2}
+        \includegraphics[width=\linewidth]{gfx/5.pdf}
+        \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b).\gl{TODO : translate}}\label{fig:xp1_2}
 \end{figure}
@@
 
-%\jfp{ce qu'est Ascene dans ce paragraphe n'est pas clair :A scene est défini comme la moyenne des scores d'une scène donnée pour tous les participants.
-le test, lui, porte sur la moyenne des Ascene pour les deux catégories meme chose pour les sujets}
-
 \item \emph{semantic features}: the sound classes that are present / absent from the scene.
@@
 
+%Pour évaluer l'impact spécifique des différentes sources sonores sur l'agrément perçu, nous soumettons nos travaux aux 5 tests/études de significativité présentés ci-après:
+
 In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following significance tests:
@@
 \begin{itemize}
+
+%\item \emph{vérification de l'agrément des scènes simulées}: afin de vérifier que la distinction affective imposée entre les i- et ni-scènes se retrouve au niveau de l'agrément perçu, nous observons s'il existe des différences entre les deux types de scènes au niveau de $\mathcal{A}_{scene}$ et $\mathcal{A}_{sujet}$.
+La significativité est évaluée par un test de Student à deux populations indépendantes pour $\mathcal{A}_{scene}$, et par un test de Student à deux populations appariées pour $\mathcal{A}_{sujet}$;
+
 \item \emph{analysis of the perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness differ between the i- and ni-scenes.
@@
 
+%\item \emph{analyse des descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels et/ou l'agrément perçu, nous évaluons cette significativité à partir d'un test de Student à deux populations;
+
 \item \emph{analysis of the structural features}: the goal is to evaluate whether the structural features differ between the i- and ni-scenes.
 The significance is measured with a two-sample Student test;
+
+%\item \emph{étude de l'influence des descripteurs structurels sur l'agrément perçu}: afin d'évaluer l'impact potentiel des descripteurs structurels sur l'agrément perçu, nous étudions l'existence de corrélations linéaires entre ces descripteurs et $\mathcal{A}_{scene}$.
+Pour mesurer la corrélation, nous utilisons le coefficient de Pearson.
+Nous adoptons ici une méthodologie couramment utilisée dans l'approche dimensionnelle;
 
@@
 The Pearson correlation coefficient is used for that purpose;
+
+%\item \emph{analyse des descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée a un impact sur la composition des scènes en terme de sources sonores, s'il est des classes de sons plus particulièrement utilisées pour simuler un type d'environnement (i ou ni), nous utilisons le V-test.
+Le test est effectué pour chaque niveau d'abstraction, et séparément, pour les classes d'événements et de textures.
+Pour chaque classe $j$ et chaque type d'environnement $k$ ($k=\lbrace i,ni\rbrace$), la valeur $V_{jk}$ du V-test se calcule comme suit:
+
+%où $c$ est le nombre de classes utilisées, $c_k$ le nombre de classes utilisées pour un type d'environnement $k$, $c_j$ le nombre de classes $j$ utilisées, et $c_{jk}$ le nombre de classes $j$ utilisées pour un type d'environnement $k$.
+Le V-test teste l'hypothèse nulle que la proportion $\frac{c_{jk}}{c}$ ne diffère pas significativement de la proportion $\frac{c_{jk}}{c_k}$.
+Si pour un environnement $k$, et une classe $j$, l'hypothèse est rejetée, la classe $j$ est alors typique de l'environnement $k$.
+Les classes typiques sont nommées \textbf{marqueurs sonores};
 
@@
 
-%\jfp{il faudrait donner une référence pour ce test : bouquin Lebart, Morineau, Piron, STATISTIQUE EXPLORATOIRE MULTIDIMENSIONNELLE peut être donner les bornes de variations de Vjk pour aider le lecteur à comprendre cet indicateur}  \jfp{on ne comprend pas bien l'hypothèse alternative du test.
-VOIR BOUQUIN LEBART page 184 ou bien article de Morineau sur le sujet Vjk est une estimation approchée de la valeur test}
+%\item \emph{étude des espaces de représentation induits par les descripteurs sémantiques}: afin d'étudier si une représentation basée uniquement sur la présence ou l'absence de classes de sons permet de séparer les deux types d'environnement, nous considérons l'espace induit par les descripteurs sémantiques $S$.
+$S$ étant un vecteur booléen, nous calculons les distances entre les scènes à partir de la distance de Hamming.
+Considérant les deux vecteurs $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, et $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ de dimension $n$, avec $x=\lbrace 0,1\rbrace$, la distance de Hamming $d_{ham}$ mesure le pourcentage de coordonnées qui diffèrent entre les deux vecteurs:
 
@@
 \end{equation}
+
+%où $\bigoplus$ désigne l'opérateur du \emph{ou-exclusif}.
+Plus la composition des deux scènes est similaire, et plus ces deux scènes sont proches.
+L'utilisation de la distance de Hamming permet de prendre en compte de manière égale les classes présentes et absentes.
+Pour mesurer la capacité intrinsèque de l'espace à séparer les i- et ni-scènes, nous utilisons une métrique d'\emph{ordre} nommée précision au rang $k$ ($P@k$).
+La $P@k$ mesure la précision obtenue après que $k$ items ont été retrouvés.
+Formellement, pour chaque scène $s_i$, nous calculons le rapport entre le nombre de scènes $s_j$ prises parmi les $k$ plus proches voisines de $s_i$, et partageant le même label que $s_i$, sur le nombre d'items à retrouver ($k$).
+La $P@k$ est alors la moyenne des rapports pour tous les items;
 
@@
 
+%\item \emph{influence des marqueurs sonores sur l'agrément perçu}: afin d'évaluer les contributions spécifiques de certaines sources sonores, nous évaluons une nouvelle fois l'impact potentiel des descripteurs structurels sur l'agrément perçu, mais en ne tenant compte, cette fois, que des marqueurs sonores pour calculer ces descripteurs.
+%\end{itemize}
+
 \item \emph{influence of the sound markers on the perceived pleasantness}: in order to assess the specific contributions of some sound sources, we again estimate the impact of the structural features on the perceived pleasantness by taking into account only the sound markers for the computation of those features.
 \end{itemize}
+
+%tous les tests de significativité sont effectués avec un seuil critique $\alpha=0.05$.
+Pour le V-test, étant donné que nous testons beaucoup de classes, une correction de Bonferroni est appliquée.
+Pour les valeurs $p$, dans le cas où $p\geq0.05$, nous indiquons sa valeur.
+Dans le cas où $0.01\leq p<0.05$, nous indiquons seulement $p<0.05$.
+Dans le dernier cas nous indiquons $p<0.01$.
 
@@
 \subsubsection*{Analysis of the perceived pleasantness}
+
+%Dans un premier temps, et afin de garantir la cohérence de nos données, nous vérifions qu'aucune ni-scène n'a un $\mathcal{A}_{scene}$ supérieur à celui d'une i-scène.
+Quatre des scènes ne respectent pas la contrainte.
+Elles et leurs correspondantes i ou ni sont retirées.
+36 i-scènes et 36 ni-scènes restent donc dans le champ de l'analyse.
+Dans un deuxième temps, nous voulons nous assurer que les sujets ont bien perçu une différence d'agrément entre les i- et ni-scènes.
+Pour ce faire, nous observons l'agrément moyen de chaque sujet $\mathcal{A}_{sujet}$, calculé séparément, pour chaque type d'environnement.
+Il apparaît que les i-scènes ont bien été perçues comme significativement plus agréables ($p<0.01$) que les ni-scènes.
 
@@
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_1}\label{fig:soundlevela}}
+        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_1}\label{fig:soundlevela}}
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_3}\label{fig:soundlevelb}}
+        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_3}\label{fig:soundlevelb}}
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_5}\label{fig:soundlevelc}}\par
+        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_5}\label{fig:soundlevelc}}\par
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_2}\label{fig:soundleveld}}
+        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_2}\label{fig:soundleveld}}
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_4}\label{fig:soundlevele}}
+        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_4}\label{fig:soundlevele}}
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_6}\label{fig:soundlevelf}}
+        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_6}\label{fig:soundlevelf}}
        \caption{Distributions of the structural features $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
 \end{figure}
+
+%En premier lieu, nous nous concentrons sur le niveau sonore.
+Les figures~\ref{fig:soundlevela},~\ref{fig:soundlevelb} et~\ref{fig:soundlevelc} affichent les distributions des niveaux $L$, $L(E)$ et $L(T)$.
+Il existe bien une différence de niveau significative entre les i- et ni-scènes ($L$: $p<0.01$), avec un écart des moyennes de -7 $dB$.
+Cette différence affecte aussi bien les événements ($L(E)$: $p<0.01$, écart moyen: -7 $dB$) que les textures ($L(T)$: $p<0.01$, écart moyen: -6 $dB$).
 
@@
 This difference is siginificant for events ($L(E)$: $p<0.01$) and for textures ($L(T)$: $p<0.01$).
-%, with a standard deviation of -7 $dB$
-%, STD: -7 $dB$
-%, STD: -6 $dB$
+
+%Nous vérifions, sans surprise, que le niveau des sources sonores est bien un indicateur d'agrément, les ni-scènes ayant tendance à être plus fortes, fait rapporté dans un grand nombre d'études.
+Nous constatons encore que cette différence de niveau s'observe de manière égale pour les événements et les textures sonores.
 
@@
 We also notice that this difference of sound level is both significant for events or textures.
+
+%Il apparaît que ce sont les événements qui impactent le plus le niveau global des scènes, l'écart entre $L$ et $L(E)$ n'étant que de 1 $dB$ pour les i-scènes et les ni-scènes.
+Cette observation fait écho aux résultats obtenus par Kuwano~\al~\cite{kuwano_memory_2003}.
+Au cours de leur expérience, les auteurs demandent à leurs sujets d'évaluer une série d'environnements sonores de manière globale, dans un premier temps, puis d'en évaluer le niveau aux instants où chacun identifie une source sonore.
+L'étude montre qu'il n'y a pas de différences significatives entre les jugements globaux et les moyennes des jugements instantanés.
+Pour en revenir à notre expérience, c'est comme si nos sujets avaient inconsciemment tenu compte de cette réalité perceptive lors de la simulation, en faisant porter le niveau sonore global par des sons courts et bien identifiés, \ie~les événements.
 
@@
 In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, \ie the events.
+
+%Nous observons, enfin, que le niveau seul ne permet pas de clairement faire la distinction entre les différents types d'environnement.
+En effet, $20\%$ des i-scènes ont un niveau supérieur au niveau minimal des ni-scènes, alors qu'il n'y a pas de recouvrement, si l'on considère l'agrément perçu $\mathcal{A}_{scene}$.
 
@@
 
+%Dans cette partie, nous analysons les relations fines qui peuvent exister entre les descripteurs structurels, d'une part, et l'agrément perçu, d'autre part.
+Contrairement à la section précédente, où la qualité affective des scènes est représentée de manière binaire (i \vs~ni), nous considérons ici l'agrément moyen $\mathcal{A}_{scene}$ comme descripteur perceptif.
+Il s'agit d'étudier l'existence de potentielles corrélations entre les descripteurs structurels et $\mathcal{A}_{scene}$.
+Les coefficients de corrélations linéaires calculés entre $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ sont présentés dans le tableau~\ref{tab:corrStructA}.
+Les relations entre $\mathcal{A}_{scene}$ et les descripteurs structurels sont illustrées par les figures~\ref{fig:soundleveld},~\ref{fig:soundlevele} et ~\ref{fig:soundlevelf}.
+
 In this section, relationships that could exist between structural features and perceived pleasantness are investigated.
@@
 
+%Concernant $L$, on observe une forte corrélation négative ($r=-0.77$, $p<0.01$) avec $\mathcal{A}_{scene}$, indiquant que plus le niveau sonore est élevé, plus la scène est désagréable.
+Cependant, la figure~\ref{fig:soundleveld} suggère que cette relation ne s'opère pas de la même manière pour les i- et ni-scènes.
+En effet, la corrélation entre $L$ et $\mathcal{A}_{scene}$, pour les ni-scènes, reste élevée ($r=-0.78$, $p<0.01$), mais est inexistante pour les i-scènes.
+
 Concerning $L$, we can observe a strong negative correlation ($r=-0.77$, $p<0.01$) with $\mathcal{A}_{scene}$, indicating that the higher the sound level is, the more unpleasant the scene is perceived.
@@
 
+%Ce résultat, considérant l'ensemble des scènes, s'explique par le fait que les i-scènes ont tendance à être moins fortes que les ni-scènes, donnant ainsi l'illusion de prolonger la corrélation négative observée pour les ni-scènes.
+
 When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the i-scenes tend to be softer than the ni-scenes, thus allowing us to extend to the i-scenes the negative correlation observed for the ni-scenes.
+
+%Nous en concluons que $L$:
+%
+%\begin{itemize}
+%\item permet bien de faire la distinction entre les i- et ni-scènes,
+%\item permet de finement caractériser l'agrément perçu des ni-scènes,
+%\item n'est pas un indicateur pertinent de l'agrément perçu pour des environnements a priori agréables.
+%\end{itemize}
 
@@
 \end{itemize}
+
+%Les mêmes constats sont faits concernant $L(E)$ (\cf~Figure~\ref{fig:soundlevele}).
+Pour $L(T)$ (\cf~Figure~\ref{fig:soundlevelf}), la corrélation modérée observée sur les scènes considérées dans leur globalité n'apparaît plus sur les i-scènes et ni-scènes considérées de façon séparée (i-scènes: $r=-0.33$, $p=0.05$, ni-scènes: $r=-0.00$, $p=0.99$).
+On peut penser que la corrélation négative émergeant au niveau l'ensemble est un artefact résultant du fait que le niveau des textures des i-scènes a tendance à être plus bas que celui des ni-scènes.
+Ainsi, si les événements sonores conservent une certaine capacité de prédiction de l'agrément pour les ni-scènes, le niveau des textures n'apporte, lui, que peu d'informations, quel que soit l'environnement.
 
@@
 
+%En résumé, en présence d'un environnement désagréable, les niveaux sonores, en particulier ceux des événements, ont un impact négatif sur l'agrément.
+En présence d'un environnement agréable, en revanche, aucun des descripteurs structurels considérés ici ne semble influer sur la perception de l'agrément.
+
 To sum up, for an unpleasant environment, sound levels - especially those of the events - negatively influence the perceived pleasantness.
 On the contrary, for a pleasant environment, none of the structural features considered in the study seem to influence the perceived pleasantness.
+
+%Ces premiers résultats pourraient montrer qu'il existe deux modes de perception, mobilisant chacun des descripteurs indépendants, modes qui s'activent en fonction de la nature de l'environnement (i ou ni).
+%
+%Le fait que $L$ ne permet pas de caractériser l'agrément des i-scènes peut nous amener à penser que toutes les sources sonores ne contribuent pas de manière égale à la perception de l'agrément, mais que seules le niveau de certaines d'entre elles a une réelle influence.
+Afin d'approfondir ce point, nous analysons, dans la section suivante, les scènes d'un point de vue sémantique, \ie~en nous intéressant à la nature des sources qui les composent.
 
@@
         \subfloat[]
-        {\includegraphics[width=1\linewidth]{gfx/ch_5/xp1_class_1}\label{fig:soundsourcea}}\par
+        {\includegraphics[width=1\linewidth]{gfx/xp1_class_1}\label{fig:soundsourcea}}\par
         \subfloat[]
-        {\includegraphics[width=.5\linewidth]{gfx/ch_5/xp1_class_2}\label{fig:soundsourceb}}
+        {\includegraphics[width=.5\linewidth]{gfx/xp1_class_2}\label{fig:soundsourceb}}
         \subfloat[]
-        {\includegraphics[width=.5\linewidth]{gfx/ch_5/xp1_class_3}\label{fig:soundsourcec}}
-       \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.}\label{fig:soundsource}
+        {\includegraphics[width=.5\linewidth]{gfx/xp1_class_3}\label{fig:soundsourcec}}
+       \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.\gl{TODO : align}}\label{fig:soundsource}
 \end{figure}
+
+%Nous analysons la composition des scènes en comptant le nombre de sujets ayant utilisé une classe de sons pour simuler un type d'environnement.
+Les résultats sont présentés à la figure~\ref{fig:soundsourcea} pour les événements, et à la figure~\ref{fig:soundsourceb} pour les textures.
+Par souci d'espace, nous choisissons un niveau d'abstraction intermédiaire entre les niveaux 0 et 1, noté $0+$, pour représenter les classes (\cf~Figure~\ref{fig:taxonomie}).
 
@@
 
+%Nous observons une différence notable dans le choix des classes entre les i- et ni-scènes.
+La répartition des classes est très proche de celle obtenue dans une étude similaire sur les environnements sonores urbains idéaux \cite{guastavino2006ideal}, \ie~les classes suggérant la présence humaine et la nature sont très présentes dans les i-scènes, a contrario, les classes désignant des sons mécaniques et/ou de travaux sont principalement utilisées pour les ni-scènes.
+
 We observe a significant difference between i- and ni-scenes for the classes choice.
@@
 These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.
+
+%Dans une certaine mesure, nos résultats contredisent ce fait.
+La figure~\ref{fig:soundsourcea} montre, en effet, que les classes d'événements de \emph{transports publics} (\emph{bus} et \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) ont été utilisées par les sujets, pour des i-scènes, dans $28\%$ des cas, et pour des ni-scènes, dans $42\%$ des cas.
+Les résultats ne remettent pas en question le fait que les sons de \emph{transports publics} soient bien acceptés: $25\%$ des sujets ont utilisé la classe \emph{bus} pour les i-scènes, un chiffre comparable à celui de la classe \emph{Vélo}, et bien supérieur à celui de toute autre classe de véhicules privés.
+Cependant les classes \emph{transports publics} sont également bien présentes dans les ni-scènes, plus que les classes \emph{voiture} ou \emph{camion} par exemple.
+La classe \emph{transports publics} ne peut donc pas être considérée comme typique d'un environnement sonore urbain idéal.
 
@@
 On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.
+
+%Cette différence peut s'expliquer par la nature des deux protocoles expérimentaux utilisés.
+Comme nous l'avons fait, Guastavino demande à ses sujets de décrire un environnement.
+Mais ces derniers travaillent de mémoire, alors que nos propres sujets disposent de supports sonores.
+Le fait que nos sujets soient confrontés à la réalité acoustique des sons, pour recréer leurs environnements, peut avoir pour effet de diminuer l'impact du contexte socio-culturel.
+D'autres études utilisant des sons comme stimuli montrent que la classe \emph{bus} peut avoir un effet négatif sur l'appréciation de l'environnement \cite{lavandier2006contribution}.
 
@@
 
+%Nous avons mis en évidence que, qualitativement, la composition des sources sonores des scènes diffère selon les types d'environnement (i ou ni).
+Nous essayons de voir maintenant si, parmi ces classes, certaines sont typiques d'un environnement en particulier.
+Pour ce faire, nous utilisons le V-test (\cf~Section~\ref{sec:xp1_dataAna}), en considérant séparément chaque niveau d'abstraction.
+Les résultats sont présentés dans le tableau~\ref{tab:markers}.
+
 We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs with respect to the type of environment (i or ni).
@@
 Results are presented in Table~\ref{tab:markers}.
+
+%Concernant les événements sonores, 9 marqueurs sont identifiés sur l'ensemble des niveaux d'abstraction.
+Comme la figure~\ref{fig:soundsource} le laissait présager, les classes relatives à la présence humaine (\emph{pas homme béton}, \emph{sonnette vélo}), et à la nature (\emph{animaux, oiseaux}, \emph{chants d'oiseaux}) sont des marqueurs de i-scènes.
+Nous notons également la présence de la classe \emph{cloche} dans les marqueurs d'un environnement idéal.
+Ce fait est possiblement dû au \emph{background} socio-culturel des sujets, dans leur grande majorité, des citoyens européens.
+En effet, selon Schafer, un son reconnu par un individu comme faisant partie intégrante de son environnement est bien accepté.
+Les marqueurs de ni-scènes sont des classes faisant référence à des sons de travaux (\emph{travaux}), ou suggérant un trafic dense (\emph{klaxon}, \emph{sirène}).
 
@@
 
+%Concernant les textures sonores, 5 marqueurs sont identifiés.
+Pour les i-scènes, il s'agit de classes faisant référence à des ambiances amorphes, calmes, (\emph{cour-intérieur} et \emph{parc}).
+Pour les ni-scènes, il s'agit, comme pour les événements, de classes faisant référence à des bruits de travaux (\emph{travaux} et \emph{véhicule de travaux}), ainsi que d'une classe faisant référence au trafic (\emph{carrefour}).
+
 Regarding the textures, 5 markers are identified.
@@
 
+%Bien que l'ensemble des marqueurs identifiés soient intuitifs, aucune des classes d'événements faisant directement référence aux bruits de véhicules motorisés n'est un marqueur, exception faite de la classe de textures \emph{carrefour}.
+Pour représenter un trafic désagréable, les sujets ont porté leurs choix sur les classes \emph{klaxon} et \emph{sirène}.
+On peut supposer que les sons isolés de véhicules sont compris comme faisant partie intégrante de l'environnement urbain, et ne sont donc pas particulièrement associés à un environnement désagréable.
+
 Although the whole set of identified markers are rather intuitive, none of the event classes related to the noise of the motor vehicles are identified as markers, except for the texture class \emph{crossroad}.
@@
         \myfloatalign
-        \includegraphics[width=\linewidth]{gfx/ch_5/pa5_1_en}
+        \includegraphics[width=\linewidth]{gfx/pa5_1_en}
        \caption{$P@5$ obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors as a function of the abstraction level.
@@
 \end{figure}
+
+%Dans cette partie, nous évaluons la capacité d'une représentation sémantique à séparer les deux types d'environnement.
+Pour ce faire, nous calculons une précision au rang 5 ($p@5$) sur l'espace induit par les descripteurs sémantiques $S$, et ce pour chaque niveau d'abstraction (\cf~Section~\ref{sec:xp1_dataAna}).
+Les vecteurs $S$ sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.
+Nous ne considérons pas les classes de marqueurs de textures, ces dernières étant trop peu nombreuses.
+Pour les mêmes raisons nous ne considérons pas les classes de marqueurs d'événements du niveau d'abstraction $0$.
+Les résultats sont affichés sur la figure~\ref{fig:pa5}.
 
@@
 
+%En ce qui concerne $ET$, la $p@5$ est de $76\%$ pour le niveau d'abstraction 0, et reste supérieure à $86\%$ à partir du niveau d'abstraction 1.
+Ces résultats confirment qu'il est possible de clairement distinguer les deux types d'environnement en se basant seulement sur la présence ou l'absence des classes de sons.
+Nous notons également que, plus le niveau d'abstraction est élevé, plus la capacité de séparer les environnements est importante.
+En d'autres termes, plus nous sommes précis dans notre description de la composition des scènes, plus nous sommes à même d'établir une distinction claire entre les i- et ni-scènes.
+
 Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and stay above $86\%$ for the abstraction level $1$.
@@
 
+%En considérant séparément $E$ et $T$, il apparaît 1) que la $p@5$ obtenue avec $E$ est similaire à celle obtenue avec $ET$, 2) que la $p@5$ obtenue avec $T$ est systématiquement inférieure d'environ $10$ à $15\%$ à celle de $E$.
+Ces résultats indiquent que l'information sémantique permettant de séparer les deux environnements est principalement portée par les événements.
+Ces résultats font, par ailleurs, écho aux travaux de  \cite{maffiolo_caracterisation_1999}, qui montrent que nous analysons de manière descriptive (en identifiant les sources) les scènes événementielles, \ie~composées d'événements sonores (\cf~Section~\ref{sec:simscene_sampleDataSet}).
+
 Considering $E$ and $T$ separately, it appears that: 1) the $p@5$ obtained with $E$ is similar to the one obtained with $ET$ ; 2) the $p@5$ obtained with $T$ is systematically lower than the one obtained with $E$, by $10\%$ or $15\%$.
@@
 These results are in line with the work done in \cite{maffiolo_caracterisation_1999} showing that humans analyze the event scenes which are composed of several sound events in a causal manner, \ie by identifying the sources, as discussed in Section~\ref{sec:simscene_sampleDataSet}.
+
+%Enfin, il apparaît que la $p@5$ obtenue avec $E_{m}$ est égale, voire supérieure à celles obtenues avec $E$ et $ET$, et ce bien qu'une information partielle soit utilisée dans ce cas pour décrire les scènes.
+La dimension des vecteurs de description $S$ pour $E_m$ est en effet inférieure à la dimension des vecteurs $S$ pour $E$, qui est elle même inférieure à celle obtenue dans le cas où toutes les classes sont utilisées ($ET$).
+De plus, dans le cas où les marqueurs ne sont pas pris en compte pour la description ($E_{w/o,m}$), les résultats chutent, passant même en dessous de ceux obtenus en ne considérant que les textures.
+Cela confirme que la majorité de l'information sémantique permettant de faire la distinction entre i-scènes et ni-scènes est incluse dans les marqueurs.
 
@@
 This confirms that most of the semantic information allowing to differentiate between i- and ni-scenes is included in the markers.
+
+%En résumé, nous déduisons de cette analyse les points suivants:
+%
+%\begin{enumerate}
+%\item contrairement à ce que nous avions constaté avec les descripteurs structurels, une description sémantique de la composition des scènes, en terme de présence/absence de sources sonores, permet de bien distinguer les deux types d'environnement (i ou ni);
+%\item l'information sémantique est majoritairement portée par les classes d'événements sonores;
+%\item parmi les classes d'événements, seule une partie, \ie~les marqueurs sonores, est nécessaire pour faire la distinction entre les i- et ni-scènes.
+%\end{enumerate}
 
@@
 \end{enumerate}
+
+%Maintenant que nous avons isolé les classes typiques des i- et ni-scènes, et vérifié que la distinction entre ces environnements dépendait de la présence de ces classes, il reste à voir si une description structurelle des scènes, basée uniquement sur ces marqueurs sonores, permet de caractériser l'agrément perçu, mieux qu'une description structurelle globale.
 
@@
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_7}\label{fig:soundlevelMarkera}}
+        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_7}\label{fig:soundlevelMarkera}}
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_9}\label{fig:soundlevelNoisea}}
+        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_9}\label{fig:soundlevelNoisea}}
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_19}\label{fig:soundlevelMarkerDiffa}}\par
+        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_19}\label{fig:soundlevelMarkerDiffa}}\par
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_8}\label{fig:soundlevelMarkerd}}
+        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_8}\label{fig:soundlevelMarkerd}}
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_10}\label{fig:soundlevelNoised}}
+        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_10}\label{fig:soundlevelNoised}}
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
+        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
         \caption{Distribution of the relative sound levels structural features related to the presence of markers $L_m$ (a, d), $L_b$ (b, e) and $L_m-L_b$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
 \end{figure}
+
+%Nous évaluons les corrélations entre $\mathcal{A}_{scene}$ et les descripteurs structurels.
+Pour cette section, les descripteurs structurels sont calculés en tenant compte des marqueurs sonores précédemment identifiés.
+Nous définissons $X_m$ le descripteur $X$ calculé en ne prenant en compte que les sons des marqueurs.
+A l'inverse, nous définissons $X_b$ ($b$: pour «~bruit~») le descripteur $X$ calculé en prenant en compte toutes les classes de sons, excepté les marqueurs.
+Lorsque le descripteur caractérise une i-scène (idem pour une ni-scène), nous ne considérons, pour le calcul, que les marqueurs identifiés pour les i-scènes (ou pour les ni-scènes), que nous nommons i-marqueurs (ou ni-marqueurs).
+Les résultats sont affichés sur le tableau~\ref{tab:corrMarkers}.
 
@@
 
+%Concernant les niveaux sonores (\cf~Figures~\ref{fig:soundlevelMarkera} et~\ref{fig:soundlevelMarkerd}), les mêmes tendances sont observées entre $L_m$, $L(E)_m$ et $L(T)_m$, d'une part, et $L$, $L(E)$ et $L(T)$, d'autre part.
+Que l'on considère uniquement les marqueurs, ou l'ensemble des classes, il s'avère que :
+
 Concerning the sound levels, the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ and $L(T)$ on the other side, see~Figures~\ref{fig:soundlevelMarkera} and~\ref{fig:soundlevelMarkerd}, respectively.
 No matter whether all the classes or only the markers are considered, it appears that:
+
+%\begin{enumerate}
+%\item il existe une différence significative entre les niveaux des i- et ni-scènes ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
+%\item le niveau sonore des scènes est majoritairement porté par les événements sonores, comparé aux textures sonores;
+%\item le niveau sonore des événements a une influence sur la perception de l'agrément pour les ni-scènes, mais pas pour les i-scènes;
+%\item le niveau sonore des textures ne joue aucun rôle dans la perception de l'agrément.
+%\end{enumerate}
 
@@
 
+%En conclusion, le niveau des ni-marqueurs a une influence négative sur l'agrément pour les ni-scènes. En revanche le niveau des i-marqueurs n’impacte pas l'agrément perçu pour les i-scènes.
+
 In conclusion, the level of ni- markers has a negative influence on the pleasantness for the ni-scenes.
@@
 
+%\begin{itemize}
+%\item plus le niveau absolu des sons n'étant pas des i-marqueurs est élevé, plus l'agrément est faible,
+%\item plus le niveau relatif des i-marqueurs, par rapport aux autres sons, est élevé, plus l'agrément est élevé.
+%\end{itemize}
+
 \begin{itemize}
@@
 
+%Pour les ni-scènes, le fait que nous observions des corrélations pour $L_m$ et $L(E)_m$, et aucune pour $L_m-L_b$ et $L(E)_m-L(E)_b$, montre que c'est bien le niveau absolu qui importe.
+
 For the ni-scenes, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.
@@
 
+%De cette analyse, nous retenons les points suivants:
+
 From this analysis, the following points can be outlined:
@@
 \begin{itemize}
+
+%\item \emph{distinguer les i- et ni-scènes}: les descripteurs sémantiques, ainsi que les descripteurs structurels globaux ($L$, $L(E)$ et $L(T)$), permettent de faire la distinction entre les i-scènes et les ni-scènes.
+La description sémantique semble être plus performante;
+%\item \emph{événements ou textures}: que ce soit pour les descripteurs sémantiques ou structurels, ce sont majoritairement les événements qui permettent de distinguer les deux types d'environnement, les textures n'apportant, au mieux, qu'une information limitée;
+%\item \emph{prédire l'agrément}: si l'on considère les corrélations entre les descripteurs structurels et l'agrément, il semble que la manière de percevoir la qualité de l'environnement diffère en fonction de la nature de ce dernier (i ou ni).
+Il n'apparaît pas envisageable de considérer un même jeu de descripteurs pour prédire, à la fois, l'agrément des i-scènes, et l'agrément des ni-scènes :
+
 \item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global structural features ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni-scenes.
@@
 
+%\item \emph{pour les ni-scènes}, ce sont le niveau global ($L$ et $L(E)$), ou le niveau des marqueurs sonores ($L_m$ et $L(E)_{m}$), qui impactent négativement l'agrément.
+On note ici que prendre en compte les contributions de différentes sources n'améliore pas la capacité de prédiction de l'agrément, par rapport à une analyse holistique de l'environnement.
+
 \item \emph{for ni-scenes}, global levels ($L$ et $L(E)$), or level of sound markers ($L_m$ et $L(E)_{m}$), have a negative influence on pleasantness.
@@
 
+%\item \emph{pour les i-scènes}, par contre, prédire l'agrément requiert d'étudier, de manière séparée, les caractéristiques des marqueurs sonores, et celles de l'ensemble des autres sons.
+Ainsi, le niveau des marqueurs pris relativement au bruit ($L(E)_m-L(E)_b$ et $L_m-L_b$) est positivement corrélé à l'agrément, alors que le niveau du bruit ($L_b$ et $L(E)_b$) est, lui, négativement corrélé.
+
 \item \emph{for i-scenes}, on contrary, the pleasantness prediction needs to consider separately the sound markers characteristics and those of the other sounds.
 Thus, the markers level, relative to the noise ($L(E)_m-L(E)_b$ and $L_m-L_b$) is positively correlated to the pleasantness, whereas the noise level ($L_b$ and $L(E)_b$) is negatively correlated.
-\end{itemize}
-\end{itemize}
+
+\end{itemize}
+\end{itemize}
+
+%Le fait que l'agrément des i-scènes ne se corrèle pas avec des descripteurs physiques holistiques, contrairement à l'agrément des ni-scènes, a récemment été déjà observé \cite{gozalo2015relationship}.
 
 The fact that the pleasantness of the i-scenes is not correlated to global physical features, contrary to pleasantness of ni-scenes, has been studied recently \cite{gozalo2015relationship}.
+
+%L'existence de deux modes de perception, mobilisant différents types de descripteurs, et dépendant de la nature (dans notre cas hédonique) des stimuli, est un phénomène qui se rapproche de celui observé pour la perception des textures (\cf~Section~\ref{sec:simscene_sampleDataSet}).
+Le cerveau adapte sa manière de traiter l'information (résumé statistique pour les textures, description fine pour les événements) suite à une prise de décision antérieure quant à la nature du stimulus (à savoir «~est-ce un événement ou une texture ?~»).
+De la même manière, les indicateurs actifs dans le jugement de l'agrément dépendent, eux aussi, d'une identification préalable de la nature hédonique globale de l'environnement (idéale ou non idéale).
 
@@
 
-%\jfp{je crois que la défintion de la variable dépendante, Asubject, n'est pas la bonne.
-A subject a a été définit avant comme "Asubject is the average pleasantness for each participant".
-l'ANOVA à deux facteurs à mesures répétées ne porte pas sur la valeur moyenne mais sur la valeur de l'évaluation pour une scene donnée.
-il faut donc définir une notation appropriée}
-
 %Tous les tests de significativité sont effectués avec un seuil critique $\alpha=0.05$.
@@
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp4_note_13}\label{fig:xp2_Asubject}}
+        {\includegraphics[width=.33\linewidth]{gfx/xp4_note_13}\label{fig:xp2_Asubject}}
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp4_note_15}\label{fig:xp2_Ascene}}
+        {\includegraphics[width=.33\linewidth]{gfx/xp4_note_15}\label{fig:xp2_Ascene}}
         \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness  per scenes$\mathcal{A}_{scene}$ (b) versus scene type.
-Black stars in subfigure (a) stand for the detected outlier,~\ie~subject 7.}\label{fig:xp2A}
+Black stars in subfigure (a) stand for the detected outlier, \ie~subject 7.}\label{fig:xp2A}
 \end{figure}
@@
 
-\begin{figure}[hp]
+\begin{figure}[!hp]
     \centering
-        {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.7\columnwidth]{gfx/ch_5/event_1_en}\label{fig:taxonomieEventa}} \par
+        {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.9\columnwidth]{gfx/appendix/event_1_en}\label{fig:taxonomieEventa}} \par
          \caption{Taxonomy of sound classes of mechanical events used for the simulation urban soundscapes  with level of abstraction from left to right.}
@@
     \centering
-        {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.7\columnwidth]{gfx/ch_5/event_2_en}\label{fig:taxonomieEventb}}
+        {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.9\columnwidth]{gfx/appendix/event_2_en}\label{fig:taxonomieEventb}}
        \caption{Taxonomy of sound classes of non-mechanical events used for the simulation urban soundscapes with level of abstraction from left to right.}
@@
   \centering
-        {\includegraphics[trim={ 0 0 7cm 0},clip,width=.7\columnwidth]{gfx/ch_5/texture_1_en}\label{fig:taxonomieTexturea}} \par
-
-        {\includegraphics[trim={ 0 0 7cm 1cm},clip,width=.7\columnwidth]{gfx/ch_5/texture_2_en}\label{fig:taxonomieTextureb}}
+        {\includegraphics[trim={ 0 0 7cm 0},clip,width=.7\columnwidth]{gfx/appendix/texture_1_en}\label{fig:taxonomieTexturea}} \par
+
+        {\includegraphics[trim={ 0 0 7cm 1cm},clip,width=.7\columnwidth]{gfx/appendix/texture_2_en}\label{fig:taxonomieTextureb}}
        \caption{Taxonomy of sound classes of textures used for the simulation urban soundscapes with level of abstraction from left to right.}
@@
         \subfloat[subject 1]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_1}\label{fig:xp4_note_1a}}
+        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_1}\label{fig:xp4_note_1a}}
         \subfloat[subject 2]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_2}\label{fig:xp4_note_1b}}
+        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_2}\label{fig:xp4_note_1b}}
         \subfloat[subject 3]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_3}\label{fig:xp4_note_1c}}
+        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_3}\label{fig:xp4_note_1c}}
         \subfloat[subject 4]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_4}\label{fig:xp4_note_1d}} \par
+        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_4}\label{fig:xp4_note_1d}} \par
         \subfloat[subject 5]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_5}\label{fig:xp4_note_1e}}
+        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_5}\label{fig:xp4_note_1e}}
         \subfloat[subject 6]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_6}\label{fig:xp4_note_1f}}
+        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_6}\label{fig:xp4_note_1f}}
         \subfloat[subject 7]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_7}\label{fig:xp4_note_1g}}
+        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_7}\label{fig:xp4_note_1g}}
         \subfloat[subject 8]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_8}\label{}} \par
+        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_8}\label{}} \par
         \subfloat[subject 9]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_9}\label{fig:xp4_note_1i}}
+        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_9}\label{fig:xp4_note_1i}}
         \subfloat[subject 10]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_10}\label{fig:xp4_note_1j}}
+        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_10}\label{fig:xp4_note_1j}}
         \subfloat[subject 11]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_11}\label{fig:xp4_note_1z}}
+        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_11}\label{fig:xp4_note_1z}}
         \subfloat[subject 12]
-        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_12}\label{fig:xp4_note_1l}}
+        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_12}\label{fig:xp4_note_1l}}
         \caption{Distribution of scores given par the subjects during experiment 2 for i/am-scenes (green) et ni/am-scenes (red).
-The horizontal axis is, from left to right, the 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant)}\label{fig:xp4_note_1}
+The horizontal axis is, from left to right, the 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant).}\label{fig:xp4_note_1}
 \end{figure}

### 2017-12-13 lafaySimulatedScenePerception.tex
up until results

@@
 
-All the participants start by designing the ideal environment; they read the second set of instruction at the end of the first experiment.
+All the participants start by designing the ideal environment; they read the second set of instructions at the end of the first experiment.
 Participants are completely free of their choices concerning sounds and synthesis parameters.
-They must nevertheless respect the two following constraints :
+The created sound environments must nevertheless fullfill the two following constraints :
 
@@
 
-These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation user interface.
+These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation interface.
 
@@
 
-Each simulation process comprises two parts:
+Each simulation process is decomposed in several steps:
 
 \begin{enumerate}
-\item the simulation: this step is split in three steps:
+\item the simulation: the user is asked to:
 \begin{itemize}
@@
 \end{itemize}
-\item the production of a free comment of the composed soundscape
+\item the feedback : writing of a free form comment about the composed soundscape
 \end{enumerate}
@@
 
-In addition, once the two sound scenes are done, participants are invited to:
+In addition, once the two sound scenes are completed, participants are invited to:
 
@@
 Before starting the first simulation, a 20-minute tutorial is given to the participants in order to familiarize them with the simulation interface and the sound database.
-The experiment is planned to last about two hours and a half, including breaks that could be taken by the participants.
+The experiment is planned to last about two hours and a half, including breaks that the subjects are allowed to take.
 
@@
 
-All the participants do the experiment on desktop computers with the same hardware and software configurations.
-The audio files are played in diotic conditions by headphones.
-During the tutorial, participants are asked to adjust the global sound volume to a comfortable level.
-Once set, they are not allowed to modify it in the remaining of the experiment.
+All the participants perform the experiment on standard desktop computers with the same hardware and software configurations.
+The audio files are played in diotic conditions using headphones.
+During the tutorial, participants are asked to adjust the sound volume to a comfortable level.
+Once set, they are not allowed to modify it during the remaining of the experiment.
 
@@
 
-All the participants do the experiment at the same time. They are equally split in three identical quiet rooms. They are not allowed to talk to each other during the experiment.
+All the participants performed the experiment at the same time. They are equally distributed in three identical quiet rooms. They are not allowed to talk to each other during the experiment.
 
@@
 
-Three experimenters (one in each room) are present during the whole duration of the experiment in order to assist participants with potential hardware and software issues and to answer questions.
+Three experimenters (one in each room) are available during the whole duration of the experiment in order to assist participants with potential hardware and software issues and to answer questions.
 
@@
 of $2.0$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment.
-All the subjects live in Nantes, France, for at least two years
+All the subjects live in Nantes, France, for at least two years.
 
@@
 Among the 44 participants, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
-the 4 other participants were excluded from the process due to a lack of respect or understanding of the instructions, or for exceeding the maximum duration of the experiment.
+the 4 other participants were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.
 
@@
 Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli.
-After the evaluation, they are free to switch to the next scene.
+After the evaluation, they are free to continue to the next scene.
 
@@
 
-For each participant, sound scenes are played in a random order.
-The first 10 scenes allow the participants to calibrate their scores; as they are always composed of 5 ideal scenes and 5 non ideal scenes.
-These first 10 scenes are played back again at the end of the experiment and the last evaluations are the ones that are only taken into account in the results.
+For each participant, sound scenes are played in a quasi random order.
+5 ideal scenes and 5 non ideal scenes are first sequenced to allow the participants to calibrate their scores.
+These first 10 scenes are played back again at the end of the experiment.
+Only the last evaluations are taken into account.
 
@@
 
-All the participants do the experiment on computers with equal hardware and software configurations.
+All the participants perform the experiment on computers with equivalent hardware and software configurations.
 The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}.
@@
 
-All the participants do the experiment simultaneously. They are not allowed to talk to each other during the experiment.
+All the participants perfrom the experiment simultaneously. They are not allowed to talk to each other during the experiment.
 
@@
 
-An experimenter is present during the whole duration of the experiment in order to assist participants and possibly answer questions.
+An experimenter is available during the whole duration of the experiment in order to assist participants and answer questions any questions the participants may have.
 
@@
 A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}.
-In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section~\ref{sec:xp1b_plan}.
+In order to be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section~\ref{sec:xp1b_plan}.
 
@@
 
-\item \emph{semantic features}: the sound classes that are present / absent from the scene.
-We use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene.
+\item \emph{semantic features}: We use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene, \textit{i.e.} the sound classes that are present / absent from the scene.
 Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise.
-The vector dimension ($n$) depends on the level of abstraction that is considered for analysis.
+The vector dimension ($n$) depends on the level of abstraction that is considered for the analysis.
 For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.
@@
 An A-filtering module processes the data before the quadratic means are computed.
-We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples and only the set of texture samples.
+We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples, and only the set of texture samples.
 \end{itemize}
@@
 
-\item \emph{analysis of the perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness differ between the i- and ni-scenes.
-We look at the existence of differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ for each type of scenes (i or ni).
+\item \emph{analysis of the perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and ni-scenes durint the experiment 1.a.
+To do so, we consider the differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ for each type of scenes (i or ni).
 The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$;

### 2017-12-13 lafaySimulatedScenePerception.tex
up until results

@@
 \item \emph{analysis of the perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and ni-scenes durint the experiment 1.a.
-To do so, we consider the differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ for each type of scenes (i or ni).
+To do so, we consider if there exist significative differences for the i- and ni-scenes of the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$.
 The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$;
+
+% GL comprends pas en francais
 
@@
 
-\item \emph{influence of the structural features on the perceived pleasantness}: the goal is to evaluate if structural features impact the perceived pleasantness.
-We look at linear correlations between these features and $\mathcal{A}_{scene}$.
+\item \emph{influence of the structural features on the perceived pleasantness}: the goal is to evaluate if the structural features impact the perceived pleasantness.
+To do so, we consider linear correlations between these features and $\mathcal{A}_{scene}$.
 The Pearson correlation coefficient is used for that purpose;
@@
 \item \emph{analysis of the semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni).
-A V-test is considered \citep{lebart2000statistique} \ml{ref en Anglais ?}.
-With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of time a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
+To do so, a V-test is considered \citep{lebart2000statistique} \ml{ref en Anglais ?}.
+With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of time a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
 For each class $j$, and each environment type $k$, an approximation of the statistic value $V_{jk}$ is computed as follows:
@@
 The typical classes are called \textbf{sound markers}.
-Testing is done for each class, at every level of abstraction, and separately for texture and event classes;
+Testing is done for each class, at each level of abstraction, and separately for texture and event classes;
 
@@
 The use of Hamming distance allows us to take into account equally the presence and absence of classes.
-For measuring the intrinsic ability of the space to split the i- and ni-scene, we use a ranking metric named the precision at rank $k$ ($P@k$).
-The $P@k$ computes the precision obtained after the $k$ closest items of a given seed item have been found.
+For measuring the intrinsic ability of the space to discriminate between the i- and ni-scenes, we use a ranking metric named the precision at rank $k$ ($P@k$).
+The $P@k$ computes the precision obtained after the $k$ closest items with respect to a given seed item have been found.
 Formally, for each $s_i$ scene (considered as seed), we compute the ratio between the number of $s_j$ scenes taken among the $k$ nearest neighbors of $s_i$ and sharing the same label than $s_i$ versus $k$.
@@
 The statistical tests of significance are done with a critical threshold of $\alpha=0.05$.
-For the V-test, considering that a great number of classes is tested, a Bonferroni correction is applied.
+For the V-test, considering that a large number of classes is tested, a Bonferroni correction is applied.
 For the $p$-value, if $p\geq0.05$, the value is reported.

### 2017-12-14 lafaySimulatedScenePerception.tex
improving translation

@@
 \item \emph{event's fades} (seconds): for the event's track only, the subject can set a fade in / fade out duration applied to each sample;
-\item \emph{global fades} (seconds): the subject can set global fade in / fade out durations applied to the entire track.
+\item \emph{global fades} (seconds): the subject can set global fade in / fade out duration applied to the entire track.
 \end{itemize}
@@
 \item to detect the presence of \emph{outliers} or ambiguous scenes.
-Indeed, throughout our analyses, the predefined hedonic properties of the scenes (i and ni) are used as groundtruth.
+Indeed, throughout our analyses, the predefined hedonic properties of the scenes (i and ni) are used as reference.
 We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes, \ie~that the less pleasant i-scene remains above the more pleasant ni-scene.
@@
 Participants are completely free of their choices concerning sounds and synthesis parameters.
-The created sound environments must nevertheless fullfill the two following constraints :
+The created sound environments must nevertheless fulfill the two following constraints :
 
@@
 
-All the participants perfrom the experiment simultaneously. They are not allowed to talk to each other during the experiment.
+All the participants perform the experiment simultaneously. They are not allowed to talk to each other during the experiment.
 
@@
 
-\item \emph{semantic features}: We use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene, \textit{i.e.} the sound classes that are present / absent from the scene.
+\item \emph{semantic features}: we use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene, \textit{i.e.} the sound classes that are present / absent from the scene.
 Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise.
@@
 
-\item \emph{structural features}: the \emph{sound level}.
-To figure out the sound level, we draw inspiration from the $L_{Aeq}$ feature.
+\item \emph{sound levels}: to figure out the sound levels, we draw inspiration from the $L_{Aeq}$.
 In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale).
@@
 
-\item \emph{analysis of the perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and ni-scenes durint the experiment 1.a.
+\item \emph{analysis of the perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and ni-scenes during the experiment 1.a.
 To do so, we consider if there exist significative differences for the i- and ni-scenes of the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$.
@@
 
-%\item \emph{analyse des descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels et/ou l'agrément perçu, nous évaluons cette significativité à partir d'un test de Student à deux populations;
-
-\item \emph{analysis of the structural features}: the goal is to evaluate whether the structural features differ between the i- and ni-scenes.
-The significance is measured with a two-sample Student test;
+%\item \emph{analyse des descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels, nous évaluons cette significativité à partir d'un test de Student à deux populations;
+
+\item \emph{analysis of the sound levels}: the goal is to evaluate whether the sound levels differ between the i- and ni-scenes. The significance is measured with a two-sample Student test;
 
@@
 
-\item \emph{influence of the structural features on the perceived pleasantness}: the goal is to evaluate if the structural features impact the perceived pleasantness.
+\item \emph{influence of the sound levels on the perceived pleasantness}: the goal is to evaluate if the sound levels impact the perceived pleasantness.
 To do so, we consider linear correlations between these features and $\mathcal{A}_{scene}$.
@@
 
-\item \emph{influence of the sound markers on the perceived pleasantness}: in order to assess the specific contributions of some sound sources, we again estimate the impact of the structural features on the perceived pleasantness by taking into account only the sound markers for the computation of those features.
+\item \emph{influence of the sound markers on the perceived pleasantness}: in order to assess the specific contributions of some sound sources, we again estimate the impact of the sound levels on the perceived pleasantness by taking into account only the sound markers for the computation of those features.
 \end{itemize}
@@
 For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment.
-It appears that the i-scenes were perceived significantly more pleasant ($p < 0.01$) than the ni-scenes.
-
-\subsubsection*{Analysis of the structural features}
+It indeed appears that the i-scenes were perceived significantly more pleasant ($p < 0.01$) than the ni-scenes.
+
+\subsubsection*{Analysis of the sound levels}
 
@@
         {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_6}\label{fig:soundlevelf}}
-       \caption{Distributions of the structural features $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
+       \caption{Distributions of the sound levels $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
 \end{figure}
@@
 
-First, our analyses focuses on the sound level.
-Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the levels $L$, $L(E)$ and $L(T)$.
-There is a significant difference in terms of sound level between i- and ni-scenes ($L$: $p<0.01$).
-This difference is siginificant for events ($L(E)$: $p<0.01$) and for textures ($L(T)$: $p<0.01$).
+First, our analyses focuses on the sound levels.
+Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the distribution of the levels $L$, $L(E)$ and $L(T)$.
+There is a significant difference in terms of sound levels between i- and ni-scenes ($L$: $p<0.01$).
+This difference is significant for events ($L(E)$: $p<0.01$, mean deviation: -7 $dB$) and for textures ($L(T)$: $p<0.01$, mean deviation: -6 $dB$).
 
@@
 
-As expected, the sound sources sound level is indeed a pleasantness indicator, as the ni-scenes tend to be louder.
+As expected, the sound level of the sources is indeed a pleasantness indicator, as the ni-scenes tend to be louder.
 This results is also an outcome of a large number of related works.
-We also notice that this difference of sound level is both significant for events or textures.
+We also notice that this difference of sound levels is both significant for events or textures.
 
@@
 
-It appears that the biggest influence on the global sound level comes from the events, the gap between $L$ and $L(E)$ being only 1 $dB$ for i- and ni-scenes.
+It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for i- and ni-scenes.
 This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}.
@@
 The study shows that there is no significant difference between global and averaged instantaneous judgments.
-In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, \ie the events.
+In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie the events.
 
@@
 
-Though, the level alone is not sufficient to differentiate the type of environment.
-In fact, $20\%$ of the i-scenes have a sound level higher than the nominal level of the ni-scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
-
-\subsubsection*{Influence of the structural features on the perceived pleasantness}
+Though, the level alone is not sufficient to differentiate completely between i- and ni-scenes.
+In fact, $20\%$ of the i-scenes have a sound levels higher than the nominal level of the ni-scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
+
+\subsubsection*{Influence of the sound levels on the perceived pleasantness}
 
@@
 \vspace{0.5mm}
-\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and structural features.}
+\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and sound levels.}
 \label{tab:corrStructA}
@@
 
-In this section, relationships that could exist between structural features and perceived pleasantness are investigated.
+In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated.
 Contrary to the previous section, we do not consider the pleasantness property of a scene on a binary mode (i vs.
-ni): this is the mean pleasantness $\mathcal{A}_{scene}$ that is considered here as a perceptual feature.
-The aim is to investigate the level of correlation between structural features and $\mathcal{A}_{scene}$.
+ni): we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature.
+The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$.
 The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown on Table~\ref{tab:corrStructA}.
-Relationships between $\mathcal{A}_{scene}$ and the structural features are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.
+Relationships between $\mathcal{A}_{scene}$ and the sound levels are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.
 
@@
 
-Concerning $L$, we can observe a strong negative correlation ($r=-0.77$, $p<0.01$) with $\mathcal{A}_{scene}$, indicating that the higher the sound level is, the more unpleasant the scene is perceived.
+Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived.
 Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship do not occur in the same way for i- and ni-scenes.
@@
 
-When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the i-scenes tend to be softer than the ni-scenes, thus allowing us to extend to the i-scenes the negative correlation observed for the ni-scenes.
+When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the i-scenes tend to be softer than the ni-scenes, thus allowing us to extend artificially to the i-scenes the negative correlation observed for the ni-scenes.
 
@@
 \item characterizes precisely the perceived pleasantness for ni-scenes,
-\item is not a relevant feature for modeling the perceived pleasantness of \textit{a priori} pleasant soundscapes (i-scenes)
+\item is not a relevant feature for modeling the perceived pleasantness of an \textit{a priori} pleasant soundscape (i-scene).
 \end{itemize}
@@
 For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$).
-Again, we believe that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i-scenes tend to be softer than ni-scenes).
+Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (i-scenes tend to be softer than ni-scenes).
 Thus, the sound events keep a relative ability to predict the pleasantness of the ni-scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.
@@
 To sum up, for an unpleasant environment, sound levels - especially those of the events - negatively influence the perceived pleasantness.
-On the contrary, for a pleasant environment, none of the structural features considered in the study seem to influence the perceived pleasantness.
+On the contrary, for a pleasant environment, none of the sound levels considered in the study seem to influence the perceived pleasantness.
 
@@
 \begin{enumerate}
-\item unlike what we outlined with the structural features, a semantic description of the scenes constitution in terms of presence / absence of sound sources, allows us to differentiate the two types of environments (i or ni);
+\item unlike what we outlined with the sound levels, a semantic description of the scenes constitution in terms of presence / absence of sound sources, allows us to differentiate the two types of environments (i or ni);
 \item the semantic information is mainly contained in the event sound classes;
@@
 \vspace{0.5mm}
-\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ (Exp. 1.b) and structural features related to the presence of sound markers.}
+\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ (Exp. 1.b) and sound levels related to the presence of sound markers.}
 \label{tab:corrMarkers}
@@
         {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
-        \caption{Distribution of the relative sound levels structural features related to the presence of markers $L_m$ (a, d), $L_b$ (b, e) and $L_m-L_b$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
+        \caption{Distribution of the relative sound levels  related to the presence of markers $L_m$ (a, d), $L_b$ (b, e) and $L_m-L_b$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
 \end{figure}
@@
 
-We evaluate the correlations between $\mathcal{A}_{scene}$ and the structural features.
-In this section, the structural features are computed by taking into account only the previously identified sound markers.
+We evaluate the correlations between $\mathcal{A}_{scene}$ and the sound levels.
+In this section, the sound levels are computed by taking into account only the previously identified sound markers.
 We define $X_m$, the $X$ feature computed by taking into account only the sound markers.
@@
 
-\item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global structural features ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni-scenes.
+\item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni-scenes.
 The semantic description seems to be more powerful;
 \item \emph{events or textures}: whatever the feature type, be it semantic or structural, events are mainly useful to differentiate the two types of scenes; textures bring a limited amount of information;
-\item \emph{pleasantness prediction}: considering the correlation between structural features and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its nature (i or ni).
+\item \emph{pleasantness prediction}: considering the correlation between sound levels and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its nature (i or ni).
 From the data gathered in those experiments, it is not conceivable to consider the same set of features to predict the pleasantness of i- and ni-scenes in the same time:

### 2017-12-18 lafaySimulatedScenePerception.tex
more

@@
 \begin{itemize}
-\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to simulate urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}.
+\item \emph{experiment 1.a (simulation)}: during this experiment, subjects are asked to simulate urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}.
 Each of them has to create two sound environments: one ideal / pleasant and the other one, non ideal / unpleasant.
@@
 The second experimental step aims at investigating more deeply and more broadly our knowledge on the pleasantness of the simulated scenes.
-For that, a second group of participants is asked to evaluate the pleasantness of each scene, on a semantic scale.
+For that, a second group of subjects is asked to evaluate the pleasantness of each scene, on a semantic scale.
 This experiment has two goals:
@@
 
-The participants have to simulate two urban sound environments of one minute each. They follow the instructions below:
+The subjects have to simulate two urban sound environments of one minute each. They follow the instructions below:
 
@@
 
-All the participants start by designing the ideal environment; they read the second set of instructions at the end of the first experiment.
-Participants are completely free of their choices concerning sounds and synthesis parameters.
+All the subjects start by designing the ideal environment; they read the second set of instructions at the end of the first experiment.
+subjects are completely free of their choices concerning sounds and synthesis parameters.
 The created sound environments must nevertheless fulfill the two following constraints :
@@
 \item the listening point of view is the one of a fixed listener;
-\item the soundscape must be realistic, \ie~physically plausible. For instance, participants are free to insert ten dogs in the soundscape but they cannot insert one dog barking every 10 milliseconds.
+\item the soundscape must be realistic, \ie~physically plausible. For instance, subjects are free to insert ten dogs in the soundscape but they cannot insert one dog barking every 10 milliseconds.
 \end{itemize}
@@
 
-In addition, once the two sound scenes are completed, participants are invited to:
+In addition, once the two sound scenes are completed, subjects are invited to:
 
@@
 
-Before starting the first simulation, a 20-minute tutorial is given to the participants in order to familiarize them with the simulation interface and the sound database.
+Before starting the first simulation, a 20-minute tutorial is given to the subjects in order to familiarize them with the simulation interface and the sound database.
 The experiment is planned to last about two hours and a half, including breaks that the subjects are allowed to take.
@@
 
-All the participants perform the experiment on standard desktop computers with the same hardware and software configurations.
+All the subjects perform the experiment on standard desktop computers with the same hardware and software configurations.
 The audio files are played in diotic conditions using headphones.
-During the tutorial, participants are asked to adjust the sound volume to a comfortable level.
+During the tutorial, subjects are asked to adjust the sound volume to a comfortable level.
 Once set, they are not allowed to modify it during the remaining of the experiment.
@@
 
-All the participants performed the experiment at the same time. They are equally distributed in three identical quiet rooms. They are not allowed to talk to each other during the experiment.
+All the subjects performed the experiment at the same time. They are equally distributed in three identical quiet rooms. They are not allowed to talk to each other during the experiment.
 
@@
 
-Three experimenters (one in each room) are available during the whole duration of the experiment in order to assist participants with potential hardware and software issues and to answer questions.
-
-\subsubsection*{Participants}
+Three experimenters (one in each room) are available during the whole duration of the experiment in order to assist subjects with potential hardware and software issues and to answer questions.
+
+\subsubsection*{subjects}
 
@@
 
-Among the 44 participants, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
-the 4 other participants were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.
+Among the 44 subjects, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
+the 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.
 
@@
 
-Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from sec. 15 to sec. 45).
+Due to temporal constraints, subjects only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from sec. 15 to sec. 45).
 
@@
 The assessment is done with a 7-point bipolar semantic scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
-Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli.
+Before evaluating a scene, subjects must listen to the first 20 seconds of the stimuli.
 After the evaluation, they are free to continue to the next scene.
@@
 For each participant, sound scenes are played in a quasi random order.
-5 ideal scenes and 5 non ideal scenes are first sequenced to allow the participants to calibrate their scores.
+5 ideal scenes and 5 non ideal scenes are first sequenced to allow the subjects to calibrate their scores.
 These first 10 scenes are played back again at the end of the experiment.
@@
 
-The experiment is planned to last 30 minutes. The participants do not know anything beforehand about the nature of the sound scenes.
+The experiment is planned to last 30 minutes. The subjects do not know anything beforehand about the nature of the sound scenes.
 
@@
 
-All the participants perform the experiment on computers with equivalent hardware and software configurations.
+All the subjects perform the experiment on computers with equivalent hardware and software configurations.
 The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}.
 The stimuli are the scenes obtained by the simulation experiment of experiment 1.a.
-The output sound level is the same for all the participants.
+The output sound level is the same for all the subjects.
 
@@
 
-All the participants perform the experiment simultaneously. They are not allowed to talk to each other during the experiment.
+All the subjects perform the experiment simultaneously. They are not allowed to talk to each other during the experiment.
 
@@
 
-An experimenter is available during the whole duration of the experiment in order to assist participants and answer questions any questions the participants may have.
-
-\subsubsection*{Participants}
+An experimenter is available during the whole duration of the experiment in order to assist subjects and answer questions any questions the subjects may have.
+
+\subsubsection*{subjects}
 
@@
 
-All the participants succeeded in doing the experiment.
+All the subjects succeeded in doing the experiment.
 
@@
 \item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale.
-$\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants.
+$\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the subjects.
 $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant.
-Given the low number of participants in experiment 1.b, we choose not to normalize the pleasantness scores.
+Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores.
 
@@
 As a consequence, 36 i-scenes and 36 ni-scenes remain for analysis.
-Second, we verify that participants really perceived a difference in terms of pleasantness between i- and ni-scenes.
+Second, we verify that subjects really perceived a difference in terms of pleasantness between i- and ni-scenes.
 For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment.
@@
 This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}.
-During their experiment, the authors ask their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source.
+During their experiment, the authors ask their subjects to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source.
 The study shows that there is no significant difference between global and averaged instantaneous judgments.
-In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie the events.
+In our case, the result can be interpreted as if the subjects had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie the events.
 
@@
 
-We study the composition of the scenes by counting the number of participants who used a given class of sounds to simulate the environment.
+We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate the environment.
 Results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
@@
 
-We observe a significant difference between i- and ni-scenes for the classes choice.
-The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~classes involving human presence and nature are prevailing in the i-scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni-scenes.
+We observe a noticable difference in terms of class choices between the i- and ni-scenes.
+The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~ on one hand, classes involving human presence and nature are prevailing in the i-scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the ni-scenes.
 These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.
@@
 
-To a certain extent, our results are in contradiction with this assumption.
-In fact,  Figure~\ref{fig:soundsourcea} show that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the participants for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes.
-the results do not re-evaluate the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the i-scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes.
+To a certain extent, our results contradict this result.
+In fact,  Figure~\ref{fig:soundsourcea} show that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the subjects for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes.
+Those results do not state the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the i-scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes.
 Though, \emph{public transport} classes are also strongly present in the ni-scenes, for instance more than \emph{light vehicle} or \emph{trucks} classes.
@@
 This difference may be explained by the nature of the experimental protocol used in the two studies.
-As in our study, Guastavino asks the participants for describing an environment starting from their memories whereas, in our case, participants perform the same task using actual sound samples.
-The fact that participants in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact.
-Other works using sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment \cite{lavandier2006contribution}.
+As in our study, Guastavino asks the subjects to describe an environment.
+She asks them to perform this task starting only from their memories whereas, in our case, subjects perform the same task using actual sound samples that they can listen to.
+The fact that subjects in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact.
+Other studies that considered sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment \cite{lavandier2006contribution}.
 
@@
 \vspace{0.5mm}
-\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of V-test value. $p\leq0.01$ for all sound markers.}
+\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked as decreasing order of V-test value. $p\leq0.01$ for all sound markers.}
 \label{tab:markers}
@@
 
-We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs with respect to the type of environment (i or ni).
-Let us now investigate whether some of the sound classes are specific to a given environment.
-For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna} separately for each abstraction level.
+We have shown that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs i- or ni-scenes.
+We now investigate whether some of the sound classes are specific to a given environment.
+For that purpose, the V-test detailed in  Section~\ref{sec:xp1_dataAna} is considered separately for each abstraction level.
 Results are presented in Table~\ref{tab:markers}.
@@
 As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i-scenes markers as well as the \emph{church bell} class.
-This latter result may be due to the socio-cultural background of the participants who are mostly European citizens.
-In fact, according to Schafer, a sound that is identified by a person as being an element of his/her environment, is well accepted.
+This latter result may be due to the socio-cultural background of the subjects who are mostly European citizens.
+In fact, according to Schafer, a sound that is identified by a person as being an important element of his/her environment, is well accepted.
 Sound markers of ni-scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
@@
 Regarding the textures, 5 markers are identified.
-For the i-scenes, they are classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}).
-For the ni-scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).
+For the i-scenes, those classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}).
+The marker classes for the ni-scenes are, as for the events, related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).
 
@@
 Although the whole set of identified markers are rather intuitive, none of the event classes related to the noise of the motor vehicles are identified as markers, except for the texture class \emph{crossroad}.
-To simulate an unpleasant traffic, participants chose the classes \emph{horn} or \emph{siren}.
+To generate an unpleasant traffic, subjects chose the classes \emph{horn} or \emph{siren}.
 We thus conclude that isolated motor vehicle sounds are understood as being part of the urban environment, and thus their nature is not necessarily linked to an unpleasant soundscape.
@@
 
-In this section, we evaluate the ability of a semantic representation of the scenes to discriminate between the two types of environment.
-For that, we consider a rank 5-precision ($p@5$) on the space induced by the semantic features $S$, and for each abstraction level, see Section~\ref{sec:xp1_dataAna}.
+In this section, we evaluate at which level a semantic representation of the scenes allows us to discriminate between the two types of environment.
+For this purpose, a rank 5-precision ($p@5$) is computed on the space induced by the semantic features $S$, and for each abstraction level, see Section~\ref{sec:xp1_dataAna}.
 The vectors $S$ are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers ($E_{w/o,m}$).
-Texture classes corresponding to sound markers are not considered as these latter are too few to reliably compute the metric.
+Texture classes corresponding to sound markers are not numerous enough to reliably compute the metric.
+They are thus not considered.
 For the same reasons, event classes corresponding to sound markers at abstraction level $0$ are also discarded.
@@
 
-Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and stay above $86\%$ for the abstraction level $1$.
-These results confirm that it is possible to make a clear distinction between the two types of environment on the sole basis of presence / absence of sound classes.
-We also notice that the higher the abstraction level is, the more important the ability to make this distinction is.
-In other words, the more precise the scene description is, the more clear the discrimination between i- and ni-scenes is.
+Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and remains above $86\%$ for higher abstraction levels.
+Considering only the presence / absence of sound classes thus allows us to nicely discriminate between the two types of environments.
+We also notice that the higher the abstraction level is, thus the more precise the description, the more efective the description is.
 
@@
 
-\subsubsection*{Participants}
+\subsubsection*{subjects}
 
@@
 
-%\nm{a ce propos, il n'est pas fait mention des sons manquants pour les participants (cf. protocole expe de 1a);
+%\nm{a ce propos, il n'est pas fait mention des sons manquants pour les subjects (cf. protocole expe de 1a);
 %ne faudrait -il pas en dire 2 mots ce qui alimenterait aussi la discussion sur la dependance de la qualite (et l'efficacite) de la simulation a  la constitution de la base de donnees de textures et d'evenemetns ?}

### 2017-12-19 lafaySimulatedScenePerception.tex
more

@@
 
-Considering $E$ and $T$ separately, it appears that: 1) the $p@5$ obtained with $E$ is similar to the one obtained with $ET$ ; 2) the $p@5$ obtained with $T$ is systematically lower than the one obtained with $E$, by $10\%$ or $15\%$.
-Those results indicate that the discriminative semantic information is mainly contained in the events.
-These results are in line with the work done in \cite{maffiolo_caracterisation_1999} showing that humans analyze the event scenes which are composed of several sound events in a causal manner, \ie by identifying the sources, as discussed in Section~\ref{sec:simscene_sampleDataSet}.
+Considering $E$ and $T$ separately, it appears that: 1) the $p@5$ obtained with $E$ is similar to the one obtained with $ET$ ; 2) the $p@5$ obtained with $T$ is always lower than the one obtained with $E$, by $10\%$ to $15\%$.
+Those results indicate that the semantic information that is discriminative is mostly carried by the events.
+Those results are in line with \cite{maffiolo_caracterisation_1999}.
+As discussed in Section~\ref{sec:simscene_sampleDataSet}, it seems that humans analyze the event scenes which are composed of several sound events in a causal manner, \ie by identifying the sources.
 
@@
 The dimension of the vectors $S$ for $E_m$ is lower than the dimension of vector $S$ for $E$, itself lower than the dimension of vector $S$ obtained when all the classes are considered ($ET$).
-Despite this, it appears that the $p@5$ obtained with $E_m$ is equal – or above – to the ones obtained with $E$ or $ET$, although a partial information is used in that case to describe the scenes.
-Moreover, in the case when the sound markers are not taken into account for the description ($E_{w/o,m}$), the performance is under those obtained with only the textures as features.
-This confirms that most of the semantic information allowing to differentiate between i- and ni-scenes is included in the markers.
+$S$ being a boolean vector, the smaller the dimension, the lower the amount of information it can carry.
+Despite this, it appears that the $p@5$ obtained with $E_m$ is equal – or above – to the ones obtained with $E$ or $ET$, although only a partial information is used in that case to describe the scenes.
+On the opposite, if the sound markers are not taken into account for the description ($E_{w/o,m}$), the performance is below the one achieved when considering only the textures as features.
+Thus, most of the semantic information allowing to differentiate between i- and ni-scenes is included in the markers.
 
@@
 
-To sum up, we deduce the following points from this analysis:
+The outcomes of this analysis are:
 
 \begin{enumerate}
-\item unlike what we outlined with the sound levels, a semantic description of the scenes constitution in terms of presence / absence of sound sources, allows us to differentiate the two types of environments (i or ni);
+\item unlike what we outlined with the sound levels, a semantic description of the scenes composition in terms of presence / absence of sound sources, allows us to differentiate reliably the two types of environments (i or ni);
 \item the semantic information is mainly contained in the event sound classes;
@@
 
-Since we have extracted the typical classes of the i- and ni-scenes and verified that the distinction between them was largely dependent on the presence of these classes, we shall now investigate whether a structural description of the scenes only based on sound markers could characterize the perceived pleasantness, perhaps better than a global structural description.
+Since we have extracted the typical classes of the i- and ni-scenes and verified that the distinction between them was largely dependent on the presence of these classes, we shall now investigate whether a description of the scenes only based on the sound pressure level of sound markers could characterize the perceived pleasantness, perhaps better than a globally computed sound level.
 
@@
 
-We evaluate the correlations between $\mathcal{A}_{scene}$ and the sound levels.
+To do so, the correlations between $\mathcal{A}_{scene}$ and the sound levels are evaluated.
 In this section, the sound levels are computed by taking into account only the previously identified sound markers.
@@
 On the opposite, we define $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
-When the feature characterizes an i-scene (idem for a ni-scene),  only the markers identified for the i-scenes (or the ni-scenes) are considered; we named them, i- markers (or ni- markers).
+When the feature characterizes an i-scene (idem for a ni-scene),  only the markers identified for the i-scenes (or the ni-scenes) are considered; we named them, i-markers (or ni- markers).
 Results are shown in Table~\ref{tab:corrMarkers}.
@@
 
-In conclusion, the level of ni- markers has a negative influence on the pleasantness for the ni-scenes.
+To conclude, the level of ni- markers has a negative influence on the pleasantness for the ni-scenes.
 On the other hand, the level of i- markers does not influence the perceived pleasantness for the i-scenes.
@@
 
-Besides, whereas $L(T)$ did not show any significant correlation for ni-scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
+Alos, whereas $L(T)$ did not show any significant correlation for ni-scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
 This indicates that the level of non-markers texture classes does not influence the perceived pleasantness in the same way for i- and ni-scenes.
@@
 
-We consider a last group of features, namely $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$.
+A last group of features is now considered, namely $L_m-L_b$, $L(E)_m-L(E)_b$ and $L(T)_m-L(T)_b$.
 These features describe the difference between the markers level and those of the other sound classes, see ~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}.
@@
 for the ni-scenes no correlation is observed.
-Thus, for the i-scenes, it is not the absolute markers level that is important, but their relative level with regards to the other sounds composing the scene.
-Then, we can induce a double perceptual mechanism for the ideal environments:
+Thus, for the i-scenes, it is not the absolute markers level that is important, but their relative level with respect to the other sounds composing the scene.
+A double perceptual mechanism for the ideal environments: can thus be observed:
 
@@
 \begin{itemize}
-\item the higher the absolute level of sounds not being i- markers, the weaker the pleasantness,
-\item the higher the relative level of i- markers is, with regards to the other sounds, the higher the pleasantness.
+\item the higher the absolute level of sounds not being i-markers, the weaker the pleasantness,
+\item the higher the relative level of i-markers is, with respect to the remaining sounds, the higher the pleasantness.
 \end{itemize}
@@
 
-For the ni-scenes, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.
+On contrary, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ for the ni-scenes, shows that this is indeed the absolute level that matters.
 
@@
 The semantic description seems to be more powerful;
-\item \emph{events or textures}: whatever the feature type, be it semantic or structural, events are mainly useful to differentiate the two types of scenes; textures bring a limited amount of information;
+\item \emph{events or textures}: whatever the feature type, be it semantic or related to the sound pressure level, events are mainly useful to differentiate the two types of scenes; textures bring a limited amount of information;
 \item \emph{pleasantness prediction}: considering the correlation between sound levels and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its nature (i or ni).

### 2017-12-20 lafaySimulatedScenePerception.tex
vtest ref

--- 2017-12-14 lafaySimulatedScenePerception.tex
@@
-\documentclass[12pt]{elsarticle}
+\documentclass[preprint,12pt]{elsarticle}
 \usepackage{url}
@@
 \usepackage{amsmath}
-\usepackage{color}
 
@@
 
-\newcommand{\gl}[1]{\textcolor{magenta}{GL : #1}}
-\newcommand{\mr}[1]{\textcolor{green}{MR : #1}}
-\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
-\newcommand{\jfp}[1]{\textcolor{blue}{JFP : #1}}
+\newcommand{\gl}[1]{\textcolor{blue}{GL : #1}}
+\newcommand{\nm}[1]{\textcolor{magenta}{#1}}
+\newcommand{\ml}[1]{\textcolor{red}{#1}}
+\newcommand{\jlc}[1]{\textcolor{green}{#1}}
 \newcommand{\ie}{\emph{i.\,e.}}
@@
 
-\journal{Behavior Research Methods}
+\journal{Nuclear Physics}
 
@@
 
-    \author{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J-F. Petiot$^{1)}$}
+    \author{G. Lafay$^{1)}$, M. Rossignol$^{2)}$, N. Misdariis$^{2)}$, M. Lagrange$^{1)}$, J.F. Petiot$^{1)}$}
 
@@
 This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes.
-A rating experiment of the soundscape pleasantness using a 7-point bipolar semantic scale is conducted to further refine the analysis of the simulated urban acoustic scenes.
+A rating experiment of the soundscape pleasantness using a 7-points Likert scale is conducted to further refine the analysis of the simulated urban acoustic scenes.
 Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.
@@
   %% keywords here, in the form: keyword \sep keyword
- cognitive psychology \sep soundscape perception \sep soundscape simulator
+ cognitive psychology \sep soundscape assessment
   %% PACS codes here, in the form: \PACS code \sep code
@@
 One of the main goal of soundscape studies is to identify which components of the soundscape influence human perception \cite{aletta2016soundscape}.
-For example, in order to study the link between an environment (urban sound scene) and the induced human sensation (calmness), one mean is to objectify the mental representation of a specific sound scene (a calm urban sound scene).
-Several means have been considered in the literature to perform this task.
+For example, in order to study the link between an environment (urban sound scene) and the induced human sensation (calmness), one mean is to objectify the mental representation of a specific sound scene, a calm urban sound scene.
+Several means have been considered in the literature to do so.
 
@@
 
-First, a subject can be asked to assess a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}.
+First, a subject can be asked to describe a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}.
 The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}.
-That being said, those recorded stimuli  can be analyzed, what allows the experimenter to gather a coarse
-grain physical description of them.
+That being said, those stimuli respectively are or can be recorded and analysed.
+It thus allows the experimenter to gather a coarse grain physical description of the stimuli.
 
@@
 Second, a subject can be asked to describe a given sound environment \cite{guastavino2006ideal, dubois2006cognitive}.
-A large amount of quantitative and semantic information is then collected about the subject's representation of this type of sound environment.
+A large amount of quantitative and semantic information is then collected about the representation of this type of sound environment that the subject has.
 Unfortunately, without any reference to sound data, this representation can hardly be characterized physically.
@@
 
-We propose in this paper to consider the use of a soundscape simulator that the subject can manage to objectify his/her representation of a given sound environment.
+We propose in this paper to consider the use of a soundscape simulator that the subject can use to objectify her representation of a given sound environment.
 We believe that the use of such a device allows us to gain the benefits of the two above cited approaches.
@@
 
+For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sounds sources that compose a sound environment of interest.
+With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the events.
+
 %Nous pensons que ces efforts de recherche peuvent grandement bénéficier de l'utilisation de scènes simulées, scènes dont les caractéristiques structurelles, en particulier la nature des sources présentes, leur répartition temporelle ainsi que leur forme d'onde respectives, sont directement accessibles par l'expérimentateur.
-
-For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment of interest.
-With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the sound events.
 
@@
 The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics.
-A lack of diversity in terms of sound sources is nonetheless mentioned by the authors that limit the strength of the outcomes given in this study.
+A lack of diversity is nonetheless mentioned by the authors that limit the strenght of the outcomes given in this study.
 
@@
 Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.
+Experiment 1.a is based on a pilot study described in \cite{lafay2014new}.
 
@@
 
-\emph{Simscene} is an online digital audio tool whose first version has been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
+\emph{Simscene} is an online digital audio tool whose first version have been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
 It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}.
-It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digital audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}.
+It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digita audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}.
 The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.
@@
 A track is then created for this sound source within the simulator interface.
-The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sound sources.
+The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sounds source.
 Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.
@@
 
+In order to provide the user with a sound database that is well organized and covers as much as possible the variety of sound sources that are available in urban areas, a typology of urban sounds is first defined.
+
+\subsubsection*{Typology}
+
 %Afin de créer un corpus de classes de sons de référence pour la simulation, nous réalisons une typologie des sons environnementaux urbains.
-
-In order to provide the user with a sound database that is well organized and covers as much as possible the variety of sound sources that are present in urban areas, a typology of urban sounds is first defined.
-
-\subsubsection*{Typology}
 
@@
 
-The literature usually distinguishes between:
-
+The literature usually distinguish between:
 \begin{itemize}
@@
 
+The simulation tool \emph{Simscene} follows this distinction and consider two distinct sound databases: one with the classes of events only and the other with classes of textures only.
+These two types also have specific scheduling procedures during the simulation process.
+
 % et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures. Événements et textures bénéficient de processus de simulation spécifiques.
-
-The simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with the classes of events only and the other with the classes of textures only.
-These two types of sound classes also have specific scheduling procedures during the simulation process.
 
@@
 
-According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment \cite{bregman1994auditory}.
+According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from different sources of a sound environment \cite{bregman1994auditory}.
 Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}.
@@
 
-Maffiolo \cite{maffiolo_caracterisation_1999} distinguishes two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events.
+Maffiolo \cite{maffiolo_caracterisation_1999} distinguish two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events.
 She shows that those processes leads to two abstract cognitive category respectively termed "the event sequences" and "the amorphous sequences".
@@
 They arise from a descriptive analysis based on the identification of the sound sources.
-On the contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}.
+On contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}.
 They result from a holistic analysis based on global acoustic features.
@@
         \myfloatalign
-        \includegraphics[width=.8\linewidth]{gfx/3.pdf}
-       \caption{Hierarchical organisation of the isolated sounds used in the simulation.\gl{TODO : translate and remove subject box}}\label{fig:orgDb}
+        \includegraphics[width=.8\linewidth]{gfx/ch_4/3.png}
+       \caption{Hierachical organisation of the isolated sounds used of the simulation.}\label{fig:orgDb}
 \end{figure}
@@
 
-We term "sample", a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples perceived equivalently.
+We term sample, a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples perceived equivalently.
 
@@
 Rosch \cite{rosch1978cognition}.
-The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources.
-For classes with high level of abstraction that have sub classes, their collection of samples is the union of the collections of the sub classes.
+The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources is.
+For classes with high level of abstraction that have sub classes, its collection of samples is the union of the collections of the sub classes.
 
@@
 Accounting for the perceptual matters detailed previously, two taxonomies are built, one for event sounds and one for texture sounds.
-Four levels of abstraction are considered from the more generic one (level 0), to the more specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}.
+Four levels of abstraction are considered from the more generic one (level 0), to the more specific classes (level 3), leading to a taxonomy close to the one used in \cite{Salamon14}.
 Only three levels of abstraction are considered for the texture sounds.
@@
 
-All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie relative to Full Scale. In our case, the full scale level is set arbitrarily to 1 Volt.
+All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie relative to Full Scale. In our case, the full scale level is of 1 Volt.
 
@@
 
-By a voluntary design choice, the simulation tool does not allow the user to interact and control directly a specific sample.
+By a volountary design choice, the simulation tool do not allow the user to interact and control a specific sample.
 Interaction is done at the track level, a track being a sequence of samples.
@@
 \begin{itemize}
-\item \emph{sound level} ($dB$): for each sample, the sound levels are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
-\item \emph{inter-onset spacing} (second): for the event's track only, and for each sound event sample, the inter-onset spacings are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
+\item \emph{level} ($dB$): the levels are  randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
+\item \emph{inter-onset spacing} (second): for the event's track only, the spacings are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
 \item \emph{start and end time} (second): the subject set the first and last occurrence of a sample of the selected class.
@@
 \begin{itemize}
-\item \emph{event's fades} (seconds): for the event's track only, the subject can set a fade in / fade out duration applied to each sample;
-\item \emph{global fades} (seconds): the subject can set global fade in / fade out duration applied to the entire track.
+\item \emph{event's fades} (seconds): for the event's track only, the subject can set a fade in / fade out duration applied to each samples;
+\item \emph{global fades} (seconds): the subject set the duration of the fade in / fade out at the beginning and the end of the track.
 \end{itemize}
@@
 
-\begin{figure}[!tp]
-		\myfloatalign
-        \subfloat[]
-        {\includegraphics[width=.8\linewidth]{gfx/SSF}\label{fig:ssf}} \par
-        \subfloat[]
-        {\includegraphics[width=\linewidth]{gfx/simscene}\label{fig:simscene}}
-       \caption{Graphical interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{simScene} tool.}
-\end{figure}
-
 %La grande majorité des outils permettant de parcourir une banque de données propose une recherche textuelle sur la base de mots-clé.
@@
 Most browsing tools are based on keyword indexing.
-However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potentially induce biases in the analysis conclusions.
+However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potenitally induce biases in the analysis conclusions.
 Indeed, the subject may select \textit{a priori} sounds referenced as belonging to \emph{park} to build a \emph{calm} soundscape.
@@
 
-Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a plane, each representing one leaf class.
+Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a plane.
 
@@
 The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed in \cite{lafay2016JAES}.
+
+\begin{figure}[t]
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=.8\columnwidth]{gfx/ch_4/SSF}\label{fig:ssf}} \par
+        \subfloat[]
+        {\includegraphics[width=\columnwidth]{gfx/ch_4/simscene}\label{fig:simscene}}
+       \caption{Graphical interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{simScene} tool.}
+\end{figure}
 
@@
 For event's tracks, the horizontal spacing between those rectangles is a function of the time delay between their onsets.
-For tracks of texture, a unique rectangle is displayed as this kind of sounds does not allow spacing with silence.
+For tracks of texture, a unique rectangle is displayed as this kind of sounds do not allow spacing with silence.
 As actual amplitude and spacing values are drawn from random variables, each time the subject changes the value of a parameter, the location and height of the rectangles are updated to reflect the changes in the sequencing of the samples.
@@
 
-\begin{figure}[t]
-        \includegraphics[width=\linewidth]{gfx/5.pdf}
-        \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b).\gl{TODO : translate}}\label{fig:xp1_2}
-\end{figure}
-
 % L'objectif est d'étudier les influences spécifiques des différentes sources sonores qui composent les environnements urbains sur la perception de l'agrément, en utilisant la simulation.
 Pour ce faire, nous planifions nos deux premières expériences (\cf~Figure~\ref{fig:xp1_2}):
-
-Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness.
-For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
-
+%
 % \begin{itemize}
@@
 Chacun compose deux environnements sonores, le premier idéal/agréable et le deuxième non-idéal/désagréable.
+%
 % \item \emph{expérience 1.b d'évaluation}: à l'issue de la simulation, nous n'avons, de fait, qu'une connaissance binaire des propriétés affectives des scènes simulées: idéale (i) et non-idéale (ni).
@@
 L'expérience d'évaluation sert deux buts:
+%
 % \begin{enumerate}
@@
 % \end{enumerate}
+%
 % \end{itemize}
-
-\begin{itemize}
-\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to simulate urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}.
-Each of them has to create two sound environments: one ideal / pleasant and the other one, non ideal / unpleasant.
-\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni).
+%
+% Notre analyse s'appuie sur les données produites par les deux expériences.
+
+Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness.
+For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
+\begin{itemize}
+\item \emph{experiment 1.a (simulation)}: during this experiment, participants are asked to create urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}.
+Each of them creates two environments: one ideal / pleasant and the other one, non ideal / unpleasant.
+\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes: respectively ideal (i) or non ideal (ni).
 Furthermore, this information is given by the creator of the scene.
-The second experimental step aims at investigating more deeply and more broadly our knowledge on the pleasantness of the simulated scenes.
-For that, a second group of participants is asked to evaluate the pleasantness of each scene, on a semantic scale.
+The second experimental step aims at investigating more deeply and more broadly our knowledge on pleasantness.
+For that, a second group of participants are asked to evaluate the pleasantness of each simulated scene, on a semantic scale.
 This experiment has two goals:
 \begin{enumerate}
-\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene.
-\item to detect the presence of \emph{outliers} or ambiguous scenes.
-Indeed, throughout our analyses, the predefined hedonic properties of the scenes (i and ni) are used as reference.
-We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes, \ie~that the less pleasant i-scene remains above the more pleasant ni-scene.
+\item evaluate the respective influence on the pleasantness of the various sources composing the scenes (i or ni)
+\item detect the presence of \emph{outliers} or ambiguous samples.
+For the remainder of the study, the defined hedonic properties (i and ni) are used as references.
+We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes; i.e that the lower i-scene is really above the upper ni-scene,  in terms of perceived pleasantness.
 \end{enumerate}
@@
 
-% Notre analyse s'appuie sur les données produites par les deux expériences.
-
-The data collected by these two experiments (1.a and 1.b) are analyzed conjointly.
+Our analysis is based on the data collected by these two experiments: 1.a and 1.b.
+
+\begin{figure}[t]
+        \myfloatalign
+        \includegraphics[width=\linewidth]{gfx/ch_5/5.png}
+        \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b)}\label{fig:xp1_2}
+\end{figure}
 
@@
 
-The design of this experiment has been validated with a pilot study described in \cite{lafay2014new}.
-
 \subsubsection*{Procedure}
@@
 % Les sujets doivent simuler deux environnements sonores urbains, chacune des scènes devant durer 1 minute. Ils doivent se conformer aux consignes suivantes:
-
-The participants have to simulate two urban sound environments of one minute each. They follow the instructions below:
-
+%
 % \begin{itemize}
@@
 % \end{itemize}
-
-\begin{itemize}
-\item  first simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
-\item second simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
-\end{itemize}
-
+%
 % Tous commencent par simuler l'environnement idéal. Ils ne prennent connaissance de la deuxième consigne qu'à la fin de la première simulation.
-
+%
 % Les sujets sont totalement libres dans le choix des sons, et des paramètres. Ils doivent cependant se soumettre à deux contraintes:
-
-All the participants start by designing the ideal environment; they read the second set of instructions at the end of the first experiment.
-Participants are completely free of their choices concerning sounds and synthesis parameters.
-The created sound environments must nevertheless fulfill the two following constraints :
-
+%
 % \begin{itemize}
@@
 % \end{itemize}
-
+%
+% Ces contraintes font partie de la consigne. Aucun contrôle n'est fait \emph{a priori} dans l'interface de simulation.
+
+The participants have to simulate two urban sound environments of one minute each. They follow the instructions below:
+\begin{itemize}
+\item  first simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
+\item second simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
+\end{itemize}
+
+All the participants start by designing the ideal environment; they read the second set of instruction at the end of the first experiment.
+Participants are completely free of their choice concerning sounds and synthesis parameters.
+They must nevertheless respect the two following constraints :
 \begin{itemize}
 \item the listening point of view is the one of a fixed listener;
-\item the soundscape must be realistic, \ie~physically plausible. For instance, participants are free to insert ten dogs in the soundscape but they cannot insert one dog barking every 10 milliseconds.
-\end{itemize}
-
-% Ces contraintes font partie de la consigne. Aucun contrôle n'est fait \emph{a priori} dans l'interface de simulation.
-
-These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation interface.
+\item the soundscape must be realistic, i.e. physically plausible.
+\end{itemize}
+For instance, the participant can insert ten dogs in the soundscape but she cannot insert one dog barking every 10 milliseconds.
+These constraints are notified in the instructions; no control is done \emph{a priori} in the simulation user interface.
 
@@
 % \end{enumerate}
-
-Each simulation process is decomposed in several steps:
-
-\begin{enumerate}
-\item the simulation: the user is asked to:
-\begin{itemize}
-  \item  select the class of sounds
-	\item  give a name to the selected class of sounds
-	\item  set the parameters of the tracks related to the selected class of sounds, see Section~\ref{sec:simscene_parametre}.
-\end{itemize}
-\item the feedback : writing of a free form comment about the composed soundscape
-\end{enumerate}
-
+%
 % En complément, et une fois les deux scènes sonores réalisées, les sujets sont invités à:
@@
 % \end{itemize}
-
-In addition, once the two sound scenes are completed, participants are invited to:
-
+%
+% Avant de commencer la première simulation, un tutoriel de 20 minutes est proposé aux sujets, afin qu'ils se familiarisent avec le logiciel de simulation, et la banque de données.
+L'expérience est prévue pour durer 2h30.
+
+Each simulation process contains two parts:
+\begin{enumerate}
+\item composition of the simulation: this step is split in three steps:
+\begin{itemize}
+  \item  select the class of sounds
+	\item  give a name to the selected class of sounds
+	\item  set the parameters of the tracks related to the selected class of sounds, see Section~\ref{sec:simscene_parametre}.
+\end{itemize}
+\item production of a free verbalization of the composed soundscape
+\end{enumerate}
+
+In addition, once the two sound scenes are done, participants are invited to
 \begin{itemize}
@@
 
-% Avant de commencer la première simulation, un tutoriel de 20 minutes est proposé aux sujets, afin qu'ils se familiarisent avec le logiciel de simulation, et la banque de données.
-L'expérience est prévue pour durer 2h30.
-
-Before starting the first simulation, a 20-minute tutorial is given to the participants in order to familiarize them with the simulation interface and the sound database.
-The experiment is planned to last about two hours and a half, including breaks that the subjects are allowed to take.
+Before starting the first simulation, a 20-minute tutorial is given to the participants in order to familiarize them with the simulation environment and the sound database.
+The experiment is planned to last about two hours and a half, including breaks that could be taken by the participants.
 
@@
 Ils ne peuvent le modifier par la suite.
-
-All the participants perform the experiment on standard desktop computers with the same hardware and software configurations.
-The audio files are played in diotic conditions using headphones.
-During the tutorial, participants are asked to adjust the sound volume to a comfortable level.
-Once set, they are not allowed to modify it during the remaining of the experiment.
-
+%
 % Tous les sujets réalisent l'expérience simultanément.
@@
 Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
-
-All the participants performed the experiment at the same time. They are equally distributed in three identical quiet rooms. They are not allowed to talk to each other during the experiment.
-
+%
 % Trois expérimentateurs, un dans chaque pièce, sont présents durant la totalité de l'expérience, afin de contrôler le bon déroulement de cette dernière, et de répondre aux éventuelles questions des sujets.
 
-Three experimenters (one in each room) are available during the whole duration of the experiment in order to assist participants with potential hardware and software issues and to answer questions.
+All the participants do the experiment on desktop computers with the same hardware and software configurations.
+The audio files are played in diotic conditions by headphones.
+During the tutorial, participants are asked to adjust the sound level in order to be comfortable.
+Once set, they are not allowed to modify it in the remaining of the experiment.
+All the participants do the experiment at the same time. They are equally split in three identical quiet rooms. They are not allowed to talk to each other during the experiment.
+Three experimenters (one in each room) are present during the whole duration of the experiment in order to assist participants with potential hardware and software issues and to answer questions.
 
@@
 Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
-
-44 students (14 females; averaging $21.6$ years of age, s.d.
-of $2.0$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment.
-All the subjects live in Nantes, France, for at least two years.
-
+%
 % Sur les 44 sujets, 40 réalisent l'expérience avec succès, produisant au final 80 scènes sonores simulées, dont 40 scènes idéales, et 40 scènes non idéales.
@@
 
-Among the 44 participants, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
-the 4 other participants were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.
+44 students (14 females - mean age 21.6, SD = 2.0) from Ecole Centrale de Nantes took part to the experiment. They are all living in downtown Nantes since 2 years or more.
+Among the 44 participants, 40 succeeded, producing in the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
+the 4 other participants were excluded from the process due to a lack of respect or understanding of the instructions, or for exceeding the maximum duration of the experiment.
 
@@
 % En raison de contraintes temporelles, les sujets n'évaluent que des séquences de 30 secondes des scènes simulées, chacune de ces séquences commençant à la seconde 15, et finissant à le seconde 45 de la scène d'une durée initiale de 1 minute.
-
-Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from sec. 15 to sec. 45).
-
+%
 % L'évaluation s'effectue sur une échelle sémantique bipolaire de 7 points, allant de -3 (non-idéale/très désagréable) à +3 (idéale/très agréable).
@@
 Après la notation, ils sont libres de passer à la scène suivante.
-
-The assessment is done with a 7-point bipolar semantic scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
-Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli.
-After the evaluation, they are free to continue to the next scene.
-
+%
 % Pour chaque sujet, les scènes sont présentées dans un ordre aléatoire.
@@
 Ces 10 premières scènes sont rejouées à la fin de l'expérience, et seules les notes données à la deuxième occurrence sont prises en compte.
-
-For each participant, sound scenes are played in a quasi random order.
-5 ideal scenes and 5 non ideal scenes are first sequenced to allow the participants to calibrate their scores.
-These first 10 scenes are played back again at the end of the experiment.
-Only the last evaluations are taken into account.
-
+%
 % L'expérience est prévue pour durer 30 minutes. Les sujets ne connaissent pas la nature des scènes.
 
+Due to temporal constraints, participants only assess 30 seconds of the initial 1-minute simulated scenes composed in Experiment 1.a (from sec. 15 to sec. 45).
+The assessment is done with a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
+Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli.
+After the evaluation, they are free to switch to the next scene.
+For each participant, sound scenes are played in a random order.
+The first 10 scenes allow the participants to calibrate their scores; as they are always composed of 5 ideal scenes and 5 non ideal scenes.
+These first 10 scenes are played back again at the end of the experiment and the last evaluations are the ones that are only taken into account in the results.
 The experiment is planned to last 30 minutes. The participants do not know anything beforehand about the nature of the sound scenes.
@@
 Le niveau sonore de sortie est identique pour tous les sujets.
-
-All the participants perform the experiment on computers with equivalent hardware and software configurations.
+%
+% Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
+%
+% Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de celle-ci, et de répondre aux éventuelles questions des sujets.
+
+All the participants do the experiment on computers with equal hardware and software configurations.
 The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}.
-The stimuli are the scenes obtained by the simulation experiment of experiment 1.a.
+All the sound scenes have been re-composed on the basis of the scores obtained in the simulation experiment (Exp.
+1.a).
 The output sound level is the same for all the participants.
-
-% Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
-
-All the participants perform the experiment simultaneously. They are not allowed to talk to each other during the experiment.
-
-% Un expérimentateur est présent durant la totalité de l'expérience, afin de contrôler le bon déroulement de celle-ci, et de répondre aux éventuelles questions des sujets.
-
-An experimenter is available during the whole duration of the experiment in order to assist participants and answer questions any questions the participants may have.
+All the participants do the experiment simultaneously. They are not allowed to talk to each other during the experiment.
+An experimenter is present during the whole duration of the experiment in order to assist participants and possibly answer questions.
 
@@
 Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
-
-10 students (2 females; averaging $23.1$ years of age, s.d.
-of $1.8$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment.
-All the subjects live in Nantes, France, for at least two years.
-None of them took part to the previous simulation experiment (experiment 1.a).
-
+%
 % Tous les sujets ont réalisé l'expérience avec succès.
 
+10 students (2 females - mean age 23.1, SD = 1.8) from Ecole Centrale de Nantes took part to the experiment.
+All are living in downtown Nantes for 2 years or more.
+None of them took part to the previous simulation experiment (Exp.
+1.a).
 All the participants succeeded in doing the experiment.
 
-\subsection{Data and statistical analysis}
+\subsection{Data and analysis methods}
 \label{sec:xp1_dataAna}
-
-\begin{table}[t]
-\centering
-\begin{tabular}{c c}
-Terms                              & Acronyms                   \\
-\hline
-Sound level                        & $L$                        \\
-Sound level (events)               & $L(E)$                     \\
-Sound level (textures)             & $L(T)$                     \\
-Average pleasantness (per scene)   & $\mathcal{A}_{scene}$      \\
-Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
-                                   &                            \\
-ideal/pleasant                     & i                          \\
-non-ideal/unpleasant               & ni                         \\
-scene ideal/pleasant               & i-scene                    \\
-scene non-ideal/unpleasant         & ni-scene                   \\
-\hline
-\end{tabular}
-\vspace{0.5mm}
-\caption{Acronyms of features used in the analysis of the sensory experiments.}
-\label{tab:acronyme}
-\end{table}
 
@@
 Afin de rester cohérent avec l’expérience 1.b d'évaluation, les descripteurs ne sont pas calculés sur la durée totale des scènes, mais sur la version réduite de 30 secondes (\cf~Section~\ref{sec:xp1b_plan}).
-
-A set of features is attached to each sound scene.
-The analysis is done on the basis of these features.
-A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}.
-In order to be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section~\ref{sec:xp1b_plan}.
-
+%
 % Pour chaque scène sonore, trois types de descripteurs sont considérés:
-
-For each sound scene, three types of features are considered :
-
+%
 % \begin{itemize}
@@
 
-\begin{itemize}
-\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale.
+A set of features is attached to each sound scene.
+The analysis is done on the basis of these features.
+A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}.
+In order be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section \ref{sec:xp1b_plan}.
+For each sound scene, three types of features are considered :
+\begin{itemize}
+\item \emph{perceptual}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale.
 $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants.
 $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant.
-Given the low number of participants in experiment 1.b, we choose not to normalize the pleasantness scores.
-
-\item \emph{semantic features}: we use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene, \textit{i.e.} the sound classes that are present / absent from the scene.
+Given the low number of participants in Exp.
+1.b, we choose not to normalize the pleasantness scores.
+\item \emph{semantic}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene.
 Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise.
-The vector dimension ($n$) depends on the level of abstraction that is considered for the analysis.
+The vector dimension ($n$) depends on the level of abstraction that is considered for analysis.
 For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.
-
-\item \emph{sound levels}: to figure out the sound levels, we draw inspiration from the $L_{Aeq}$.
-In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale).
+\item \emph{structural}: features related to \emph{sound level}.
+To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference.
+In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt.
 The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene.
 An A-filtering module processes the data before the quadratic means are computed.
-We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples, and only the set of texture samples.
-\end{itemize}
-
-%Pour évaluer l'impact spécifique des différentes sources sonores sur l'agrément perçu, nous soumettons nos travaux aux 5 tests/études de significativité présentés ci-après:
-
-In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following significance tests:
-
-\begin{itemize}
-
-%\item \emph{vérification de l'agrément des scènes simulées}: afin de vérifier que la distinction affective imposée entre les i- et ni-scènes se retrouve au niveau de l'agrément perçu, nous observons s'il existe des différences entre les deux types de scènes au niveau de $\mathcal{A}_{scene}$ et $\mathcal{A}_{sujet}$.
-La significativité est évaluée par un test de Student à deux populations indépendantes pour $\mathcal{A}_{scene}$, et par un test de Student à deux populations appariées pour $\mathcal{A}_{sujet}$;
-
-\item \emph{analysis of the perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and ni-scenes during the experiment 1.a.
-To do so, we consider if there exist significative differences for the i- and ni-scenes of the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$.
-The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$;
-
-% GL comprends pas en francais
-
-%\item \emph{analyse des descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels, nous évaluons cette significativité à partir d'un test de Student à deux populations;
-
-\item \emph{analysis of the sound levels}: the goal is to evaluate whether the sound levels differ between the i- and ni-scenes. The significance is measured with a two-sample Student test;
-
-%\item \emph{étude de l'influence des descripteurs structurels sur l'agrément perçu}: afin d'évaluer l'impact potentiel des descripteurs structurels sur l'agrément perçu, nous étudions l'existence de corrélations linéaires entre ces descripteurs et $\mathcal{A}_{scene}$.
-Pour mesurer la corrélation, nous utilisons le coefficient de Pearson.
-Nous adoptons ici une méthodologie couramment utilisée dans l'approche dimensionnelle;
-
-\item \emph{influence of the sound levels on the perceived pleasantness}: the goal is to evaluate if the sound levels impact the perceived pleasantness.
-To do so, we consider linear correlations between these features and $\mathcal{A}_{scene}$.
-The Pearson correlation coefficient is used for that purpose;
-
-%\item \emph{analyse des descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée a un impact sur la composition des scènes en terme de sources sonores, s'il est des classes de sons plus particulièrement utilisées pour simuler un type d'environnement (i ou ni), nous utilisons le V-test.
-Le test est effectué pour chaque niveau d'abstraction, et séparément, pour les classes d'événements et de textures.
-Pour chaque classe $j$ et chaque type d'environnement $k$ ($k=\lbrace i,ni\rbrace$), la valeur $V_{jk}$ du V-test se calcule comme suit:
-
-%où $c$ est le nombre de classes utilisées, $c_k$ le nombre de classes utilisées pour un type d'environnement $k$, $c_j$ le nombre de classes $j$ utilisées, et $c_{jk}$ le nombre de classes $j$ utilisées pour un type d'environnement $k$.
-Le V-test teste l'hypothèse nulle que la proportion $\frac{c_{jk}}{c}$ ne diffère pas significativement de la proportion $\frac{c_{jk}}{c_k}$.
-Si pour un environnement $k$, et une classe $j$, l'hypothèse est rejetée, la classe $j$ est alors typique de l'environnement $k$.
-Les classes typiques sont nommées \textbf{marqueurs sonores};
-
-\item \emph{analysis of the semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni).
-To do so, a V-test is considered \citep{lebart2000statistique} \ml{ref en Anglais ?}.
-With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of time a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
-For each class $j$, and each environment type $k$, an approximation of the statistic value $V_{jk}$ is computed as follows:
+We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples and only the set of texture samples.
+\end{itemize}
+
+In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following tests of significativity:
+
+\begin{itemize}
+\item \emph{checking the pleasantness of the simulated sound scene}: to check that the emotional difference produced by the i- and ni- scenes is also present in terms of perceived pleasantness, we look at the existence of differences between $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$ with respect to the two types of scenes.
+The significativity is evaluated by a Student test on two independent populations for $\mathcal{A}_{scene}$ and by a Student test on two paired populations for $\mathcal{A}_{subject}$;
+\item \emph{analysis of the structural features}: in order to evaluate whether the affective difference between  the i- and ni- scenes has a significative influence  on the nature of the scenes \ie~ if significative differences exist between structural features and perceived pleasantness.
+The significativity is measured with a Student test on two populations;
+\item \emph{study of the influence of the structural features on the perceived pleasantness}: in order to evaluate a possible impact of the structural features on the perceived pleasantness, linear correlations between these features and $\mathcal{A}_{scene}$ are considered.
+The Pearson correlation coefficient is used for that purpose, on the basis of the standard methodology used in multidimensional studies;
+\item \emph{analysis of the semantic features}: in order to determine if the affective difference between  the i- and ni- scenes has an influence on the scenes composition in terms of sound sources, or if specific classes are more frequently used to simulate a given environment (i or ni), a V-test is considered \cite{rakotomalala2008tvpercent}.
+Testing is done at every level of abstraction, and separately for texture and event classes.
+For each class $j$, and each environment type $k$ ($k=\lbrace i,ni\rbrace$), the value $V_{jk}$ is computed as follows:
 
@@
 
-If the null hypothesis is rejected, the class $j$ is said to be typical with respect to the type of scene $k$.
-The typical classes are called \textbf{sound markers}.
-Testing is done for each class, at each level of abstraction, and separately for texture and event classes;
-
-%\item \emph{étude des espaces de représentation induits par les descripteurs sémantiques}: afin d'étudier si une représentation basée uniquement sur la présence ou l'absence de classes de sons permet de séparer les deux types d'environnement, nous considérons l'espace induit par les descripteurs sémantiques $S$.
-$S$ étant un vecteur booléen, nous calculons les distances entre les scènes à partir de la distance de Hamming.
-Considérant les deux vecteurs $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, et $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ de dimension $n$, avec $x=\lbrace 0,1\rbrace$, la distance de Hamming $d_{ham}$ mesure le pourcentage de coordonnées qui diffèrent entre les deux vecteurs:
-
-\item \emph{representation space induced by the semantic features}: the goal is to determine if a representation space of the scenes solely based on the presence / absence of sound sources allows us to distinguish between the two types of scene.
-Considering $S_i$ the semantic feature of the scene $i$, we compute the distances between all the vectors $S_i$.
-A Hamming distance is used.
-Considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$ and $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$, with $x \in\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between the two vectors.
-It is defined as follows:
+where $c$ is the number of available classes, $c_k$ the number of classes for a given type of environment $k$, $c_j$ the number of  $j$ classes, and $c_{jk}$ the number of $j$ classes used for an environment $k$.
+The V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significatively different from the ratio $\frac{c_{jk}}{c_k}$.
+For a given environment $k$, and class $j$, if the hypothesis is rejected, the class $j$ is said to be typical with respect to environment $k$.
+The typical classes are called \textbf{sound markers};
+
+\item \emph{Study of the representation spaces induced by the semantic features}: in order to determine if a representation only based on the presence / absence of sound sources allows us to distinguish between the two types of environment, we investigate the space induced by the semantic features $S$.
+$S$ is a boolean vector and the distance between two scenes is computed with the Hamming distance.
+Considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$, where $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$ is of cardinality $n$, with $x=\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between two vectors, ans is defined as follows:
 
@@
 
-%où $\bigoplus$ désigne l'opérateur du \emph{ou-exclusif}.
-Plus la composition des deux scènes est similaire, et plus ces deux scènes sont proches.
-L'utilisation de la distance de Hamming permet de prendre en compte de manière égale les classes présentes et absentes.
-Pour mesurer la capacité intrinsèque de l'espace à séparer les i- et ni-scènes, nous utilisons une métrique d'\emph{ordre} nommée précision au rang $k$ ($P@k$).
-La $P@k$ mesure la précision obtenue après que $k$ items ont été retrouvés.
-Formellement, pour chaque scène $s_i$, nous calculons le rapport entre le nombre de scènes $s_j$ prises parmi les $k$ plus proches voisines de $s_i$, et partageant le même label que $s_i$, sur le nombre d'items à retrouver ($k$).
-La $P@k$ est alors la moyenne des rapports pour tous les items;
-
 where $\bigoplus$ is the \emph{exclusive-or}.
-Two scenes having similar source compositions will be close in such space.
+The more the composition of the two scenes are similar, the closer they are in this space.
 The use of Hamming distance allows us to take into account equally the presence and absence of classes.
-For measuring the intrinsic ability of the space to discriminate between the i- and ni-scenes, we use a ranking metric named the precision at rank $k$ ($P@k$).
-The $P@k$ computes the precision obtained after the $k$ closest items with respect to a given seed item have been found.
+For measuring the intrinsic ability of the space to split the i- and ni- scene, we use a ranking metric named the precision at rank $k$ ($P@k$) .
+The $P@k$ computes the precision obtained after the $k$ closest items of a given seed item have been found.
 Formally, for each $s_i$ scene (considered as seed), we compute the ratio between the number of $s_j$ scenes taken among the $k$ nearest neighbors of $s_i$ and sharing the same label than $s_i$ versus $k$.
@@
 
-%\item \emph{influence des marqueurs sonores sur l'agrément perçu}: afin d'évaluer les contributions spécifiques de certaines sources sonores, nous évaluons une nouvelle fois l'impact potentiel des descripteurs structurels sur l'agrément perçu, mais en ne tenant compte, cette fois, que des marqueurs sonores pour calculer ces descripteurs.
-%\end{itemize}
-
-\item \emph{influence of the sound markers on the perceived pleasantness}: in order to assess the specific contributions of some sound sources, we again estimate the impact of the sound levels on the perceived pleasantness by taking into account only the sound markers for the computation of those features.
-\end{itemize}
-
-%tous les tests de significativité sont effectués avec un seuil critique $\alpha=0.05$.
-Pour le V-test, étant donné que nous testons beaucoup de classes, une correction de Bonferroni est appliquée.
-Pour les valeurs $p$, dans le cas où $p\geq0.05$, nous indiquons sa valeur.
-Dans le cas où $0.01\leq p<0.05$, nous indiquons seulement $p<0.05$.
-Dans le dernier cas nous indiquons $p<0.01$.
-
-The statistical tests of significance are done with a critical threshold of $\alpha=0.05$.
-For the V-test, considering that a large number of classes is tested, a Bonferroni correction is applied.
+\item \emph{influence of the sound markers on the perceived pleasantness}: in order to assess the specific contributions of some sound sources, we again estimate the impact of the structural features on the perceived pleasantness by taking into account only the sound markers for the computation of those features.
+\end{itemize}
+
+Except for the V-test, the statistical test of significativity are done with a critical threshold of $\alpha=0.05$.
+For the V-test, considering that a great number of classes is tested, a Bonferroni correction is applied.
 For the $p$-value, if $p\geq0.05$, the value is reported.
@@
 
+\begin{table}[t]
+\centering
+\begin{tabular}{c c}
+Terms                         & Acronyms              \\
+\hline
+Level                        & $L$                    \\
+Level (events)            & $L(E)$                 \\
+Level (textures)              & $L(T)$                 \\
+Average pleasantness (per scene)     & $\mathcal{A}_{scene}$  \\
+Average pleasantness (per subject)     & $\mathcal{A}_{sujet}$  \\
+                               &                        \\
+ideal/pleasant                & i                      \\
+non-ideal/unpleasant         & ni                     \\
+scene ideal/pleasant          & i-scene                \\
+scene non-ideal/unpleasant   & ni-scene               \\
+\hline
+\end{tabular}
+\vspace{0.5mm}
+\caption{Acronyms of features used in the analysis of the sensory experiments.}
+\label{tab:acronyme}
+\end{table}
+
 \subsection{Results}
 
-\subsubsection*{Analysis of the perceived pleasantness}
-
-%Dans un premier temps, et afin de garantir la cohérence de nos données, nous vérifions qu'aucune ni-scène n'a un $\mathcal{A}_{scene}$ supérieur à celui d'une i-scène.
-Quatre des scènes ne respectent pas la contrainte.
-Elles et leurs correspondantes i ou ni sont retirées.
-36 i-scènes et 36 ni-scènes restent donc dans le champ de l'analyse.
-Dans un deuxième temps, nous voulons nous assurer que les sujets ont bien perçu une différence d'agrément entre les i- et ni-scènes.
-Pour ce faire, nous observons l'agrément moyen de chaque sujet $\mathcal{A}_{sujet}$, calculé séparément, pour chaque type d'environnement.
-Il apparaît que les i-scènes ont bien été perçues comme significativement plus agréables ($p<0.01$) que les ni-scènes.
-
-First, in order to ensure the coherence of the data, we check that none of the ni-scenes gets a $\mathcal{A}_{scene}$ higher than one of a i-scene.
-4 ni-scenes don't fulfill that constraint: they are thus removed, together with their corresponding i-scenes.
-As a consequence, 36 i-scenes and 36 ni-scenes remain for analysis.
-Second, we verify that participants really perceived a difference in terms of pleasantness between i- and ni-scenes.
+\subsubsection*{Checking of the pleasantness of the synthesized scenes}
+
+First, in order to ensure the coherence of the data, we check that none of the ni- scenes gets a $\mathcal{A}_{scene}$ higher than one of a i- scene.
+Four ni- scenes don’t fullfill that constraint: they are thus removed, together with their corresponding i- scenes.
+As a consequence, 36 i -scenes and 36 ni -scenes remain for analysis.
+Second, we verify that participants really perceived a difference in terms of pleasantness between i- and ni -scenes.
 For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment.
-It indeed appears that the i-scenes were perceived significantly more pleasant ($p < 0.01$) than the ni-scenes.
-
-\subsubsection*{Analysis of the sound levels}
-
-\begin{figure}[t]
-		\myfloatalign
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_1}\label{fig:soundlevela}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_3}\label{fig:soundlevelb}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_5}\label{fig:soundlevelc}}\par
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_2}\label{fig:soundleveld}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_4}\label{fig:soundlevele}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_6}\label{fig:soundlevelf}}
-       \caption{Distributions of the sound levels $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
-\end{figure}
-
-%En premier lieu, nous nous concentrons sur le niveau sonore.
-Les figures~\ref{fig:soundlevela},~\ref{fig:soundlevelb} et~\ref{fig:soundlevelc} affichent les distributions des niveaux $L$, $L(E)$ et $L(T)$.
-Il existe bien une différence de niveau significative entre les i- et ni-scènes ($L$: $p<0.01$), avec un écart des moyennes de -7 $dB$.
-Cette différence affecte aussi bien les événements ($L(E)$: $p<0.01$, écart moyen: -7 $dB$) que les textures ($L(T)$: $p<0.01$, écart moyen: -6 $dB$).
-
-First, our analyses focuses on the sound levels.
-Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the distribution of the levels $L$, $L(E)$ and $L(T)$.
-There is a significant difference in terms of sound levels between i- and ni-scenes ($L$: $p<0.01$).
-This difference is significant for events ($L(E)$: $p<0.01$, mean deviation: -7 $dB$) and for textures ($L(T)$: $p<0.01$, mean deviation: -6 $dB$).
-
-%Nous vérifions, sans surprise, que le niveau des sources sonores est bien un indicateur d'agrément, les ni-scènes ayant tendance à être plus fortes, fait rapporté dans un grand nombre d'études.
-Nous constatons encore que cette différence de niveau s'observe de manière égale pour les événements et les textures sonores.
-
-As expected, the sound level of the sources is indeed a pleasantness indicator, as the ni-scenes tend to be louder.
-This results is also an outcome of a large number of related works.
-We also notice that this difference of sound levels is both significant for events or textures.
-
-%Il apparaît que ce sont les événements qui impactent le plus le niveau global des scènes, l'écart entre $L$ et $L(E)$ n'étant que de 1 $dB$ pour les i-scènes et les ni-scènes.
-Cette observation fait écho aux résultats obtenus par Kuwano~\al~\cite{kuwano_memory_2003}.
-Au cours de leur expérience, les auteurs demandent à leurs sujets d'évaluer une série d'environnements sonores de manière globale, dans un premier temps, puis d'en évaluer le niveau aux instants où chacun identifie une source sonore.
-L'étude montre qu'il n'y a pas de différences significatives entre les jugements globaux et les moyennes des jugements instantanés.
-Pour en revenir à notre expérience, c'est comme si nos sujets avaient inconsciemment tenu compte de cette réalité perceptive lors de la simulation, en faisant porter le niveau sonore global par des sons courts et bien identifiés, \ie~les événements.
-
-It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for i- and ni-scenes.
+It appears that the i- scenes were perceived significatively more pleasant ($p < 0.01$) than the ni- scenes.
+
+\subsubsection*{Analysis of the structural features}
+
+First, our analyses focuses on the sound level.
+Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the levels $L$, $L(E)$ and $L(T)$.
+There is a significative difference in terms of sound level between i- and ni -scenes ($L$: $p<0.01$), with a standard deviation of -7 $dB$.
+This difference is siginificant for events ($L(E)$: $p<0.01$, STD: -7 $dB$) and for textures ($L(T)$: $p<0.01$, STD: -6 $dB$).
+
+As expected, the sound sources sound level is indeed a pleasantness indicator, as the ni -scenes tend to be louder, an outcome of a large number of related works.
+We also notice that this difference of sound level is equally true for events or textures.
+
+It appears that the biggest influence on the global sound level comes from the events, the gap between $L$ and $L(E)$ being only 1 $dB$ for i- and ni- scenes.
 This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}.
 During their experiment, the authors ask their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source.
-The study shows that there is no significant difference between global and averaged instantaneous judgments.
-In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie the events.
-
-%Nous observons, enfin, que le niveau seul ne permet pas de clairement faire la distinction entre les différents types d'environnement.
-En effet, $20\%$ des i-scènes ont un niveau supérieur au niveau minimal des ni-scènes, alors qu'il n'y a pas de recouvrement, si l'on considère l'agrément perçu $\mathcal{A}_{scene}$.
-
-Though, the level alone is not sufficient to differentiate completely between i- and ni-scenes.
-In fact, $20\%$ of the i-scenes have a sound levels higher than the nominal level of the ni-scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
-
-\subsubsection*{Influence of the sound levels on the perceived pleasantness}
+The study shows that there is no significative difference between global and averaged instantaneous judgments.
+In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, \ie the events.
+
+Though, the level alone is not sufficient to differentiate the type of environment.
+In fact, $20\%$ of the i- scenes have a sound level higher than the nominal level of the ni- scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
+
+\begin{figure*}[t]
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_1}\label{fig:soundlevela}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_3}\label{fig:soundlevelb}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_5}\label{fig:soundlevelc}}\par
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_2}\label{fig:soundleveld}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_4}\label{fig:soundlevele}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_6}\label{fig:soundlevelf}}
+       \caption{Distributions of the structural features $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
+\end{figure*}
+
+\subsubsection*{Influence of the structural features on the perceived pleasantness}
+
+In this section, relationships that could exist between structural features and perceived pleasantness are investigated.
+Contrary to the previous section, we do not consider the pleasantness property of a scene on a binary mode (i vs.
+ni): this is the mean pleasantness $\mathcal{A}_{scene}$ that is considered here as a perceptual feature.
+The aim is to investigate the level of correlation between structural features and $\mathcal{A}_{scene}$.
+The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown on Table~\ref{tab:corrStructA}.
+Relationships between $\mathcal{A}_{scene}$ and the structural features are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and ~\ref{fig:soundlevelf}.
+
+Concerning $L$, we can observe a strong negative correlation ($r=-0.77$, $p<0.01$) with $\mathcal{A}_{scene}$, indicating that the higher the sound level is, the more unpleasant the scene is perceived.
+Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship do not occur in the same way for i- and ni- scenes.
+In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni- scenes ($r=-0.78$, $p<0.01$), but is quasi null for i-scenes.
+
+When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the i- scenes tend to be softer than the ni- scenes, thus allowing us to extend to the i-scenes the negative correlation observed for the ni- scenes.
+
+We thus conclude that $L$:
+
+\begin{itemize}
+\item allows us to differentiate between i- and ni- scenes,
+\item characterizes precisely the perceived pleasantness for ni- scenes,
+\item is not a relevant feature for modeling the perceived pleasantness of \textit{a priori} pleasant soundscapes (i- scenes)
+\end{itemize}
+
+The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}.
+For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$).
+Again, we believe that the negative correlation coming from the whole is an artefact due to the level difference between the two types of scenes (i- scenes tend to be softer than ni- scenes).
+Thus, the sound events keep a relative ability to predict the pleasantness of the ni- scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.
+
+To sum up, for an unpleasant environment, sound levels - especially those of the events - negatively influence the perceived pleasantness.
+On the contrary, for a pleasant environment, none of the structural features considered in the study seem to influence the perceived pleasantness.
+
+Those first outcomes tend to show that 1) two modes of perception exist depending on the very nature of the environment (i- or ni-), and 2) each involve independent features.
+The fact that $L$ is not sufficient to characterize the pleasantness of the i- scenes can lead us to conclude that all the sound sources do not equally contribute to the perception of pleasantness.
+Thus, we think that only the level of some of them can influence this percept.
+In order to investigate further in that direction, we analyze in the next section the scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.
 
@@
 \begin{tabular}{l c c c}
-               & all scenes                & i-scenes          & ni-scenes    \\
+               & i-scenes                   & ni-scenes    \\
 \hline
-$L$            & \textbf{-0.77} ($p<0.01$) & -0.32 ($p=0.06$)  & \textbf{-0.78} ($p<0.01$)\\
-$L(E)$         & \textbf{-0.75} ($p<0.01$) & -0.20 ($p=0.24$)  & \textbf{-0.75} ($p<0.01$)\\
-$L(T)$         & \textbf{-0.53} ($p<0.01$) & -0.33 ($p=0.05$)  &  -0.00 ($p=0.99$) \\
+$L$            & -0.32 ($p=0.06$)           & \textbf{-0.78} ($p<0.01$)\\
+$L(E)$         & -0.20 ($p=0.24$)           & \textbf{-0.75} ($p<0.01$)\\
+$L(T)$         & -0.33 ($p=0.05$)           &  -0.00 ($p=0.99$) \\
 \hline
@@
 \vspace{0.5mm}
-\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and sound levels.}
+\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and structural features.}
 \label{tab:corrStructA}
@@
 
-%Dans cette partie, nous analysons les relations fines qui peuvent exister entre les descripteurs structurels, d'une part, et l'agrément perçu, d'autre part.
-Contrairement à la section précédente, où la qualité affective des scènes est représentée de manière binaire (i \vs~ni), nous considérons ici l'agrément moyen $\mathcal{A}_{scene}$ comme descripteur perceptif.
-Il s'agit d'étudier l'existence de potentielles corrélations entre les descripteurs structurels et $\mathcal{A}_{scene}$.
-Les coefficients de corrélations linéaires calculés entre $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ sont présentés dans le tableau~\ref{tab:corrStructA}.
-Les relations entre $\mathcal{A}_{scene}$ et les descripteurs structurels sont illustrées par les figures~\ref{fig:soundleveld},~\ref{fig:soundlevele} et ~\ref{fig:soundlevelf}.
-
-In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated.
-Contrary to the previous section, we do not consider the pleasantness property of a scene on a binary mode (i vs.
-ni): we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature.
-The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$.
-The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown on Table~\ref{tab:corrStructA}.
-Relationships between $\mathcal{A}_{scene}$ and the sound levels are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.
-
-%Concernant $L$, on observe une forte corrélation négative ($r=-0.77$, $p<0.01$) avec $\mathcal{A}_{scene}$, indiquant que plus le niveau sonore est élevé, plus la scène est désagréable.
-Cependant, la figure~\ref{fig:soundleveld} suggère que cette relation ne s'opère pas de la même manière pour les i- et ni-scènes.
-En effet, la corrélation entre $L$ et $\mathcal{A}_{scene}$, pour les ni-scènes, reste élevée ($r=-0.78$, $p<0.01$), mais est inexistante pour les i-scènes.
-
-Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived.
-Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship do not occur in the same way for i- and ni-scenes.
-In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni-scenes ($r=-0.78$, $p<0.01$), but is quasi null for i-scenes.
-
-%Ce résultat, considérant l'ensemble des scènes, s'explique par le fait que les i-scènes ont tendance à être moins fortes que les ni-scènes, donnant ainsi l'illusion de prolonger la corrélation négative observée pour les ni-scènes.
-
-When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the i-scenes tend to be softer than the ni-scenes, thus allowing us to extend artificially to the i-scenes the negative correlation observed for the ni-scenes.
-
-%Nous en concluons que $L$:
-%
-%\begin{itemize}
-%\item permet bien de faire la distinction entre les i- et ni-scènes,
-%\item permet de finement caractériser l'agrément perçu des ni-scènes,
-%\item n'est pas un indicateur pertinent de l'agrément perçu pour des environnements a priori agréables.
-%\end{itemize}
-
-We thus conclude that $L$:
-
-\begin{itemize}
-\item allows us to differentiate between i- and ni-scenes,
-\item characterizes precisely the perceived pleasantness for ni-scenes,
-\item is not a relevant feature for modeling the perceived pleasantness of an \textit{a priori} pleasant soundscape (i-scene).
-\end{itemize}
-
-%Les mêmes constats sont faits concernant $L(E)$ (\cf~Figure~\ref{fig:soundlevele}).
-Pour $L(T)$ (\cf~Figure~\ref{fig:soundlevelf}), la corrélation modérée observée sur les scènes considérées dans leur globalité n'apparaît plus sur les i-scènes et ni-scènes considérées de façon séparée (i-scènes: $r=-0.33$, $p=0.05$, ni-scènes: $r=-0.00$, $p=0.99$).
-On peut penser que la corrélation négative émergeant au niveau l'ensemble est un artefact résultant du fait que le niveau des textures des i-scènes a tendance à être plus bas que celui des ni-scènes.
-Ainsi, si les événements sonores conservent une certaine capacité de prédiction de l'agrément pour les ni-scènes, le niveau des textures n'apporte, lui, que peu d'informations, quel que soit l'environnement.
-
-The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}.
-For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$).
-Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (i-scenes tend to be softer than ni-scenes).
-Thus, the sound events keep a relative ability to predict the pleasantness of the ni-scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.
-
-%En résumé, en présence d'un environnement désagréable, les niveaux sonores, en particulier ceux des événements, ont un impact négatif sur l'agrément.
-En présence d'un environnement agréable, en revanche, aucun des descripteurs structurels considérés ici ne semble influer sur la perception de l'agrément.
-
-To sum up, for an unpleasant environment, sound levels - especially those of the events - negatively influence the perceived pleasantness.
-On the contrary, for a pleasant environment, none of the sound levels considered in the study seem to influence the perceived pleasantness.
-
-%Ces premiers résultats pourraient montrer qu'il existe deux modes de perception, mobilisant chacun des descripteurs indépendants, modes qui s'activent en fonction de la nature de l'environnement (i ou ni).
-%
-%Le fait que $L$ ne permet pas de caractériser l'agrément des i-scènes peut nous amener à penser que toutes les sources sonores ne contribuent pas de manière égale à la perception de l'agrément, mais que seules le niveau de certaines d'entre elles a une réelle influence.
-Afin d'approfondir ce point, nous analysons, dans la section suivante, les scènes d'un point de vue sémantique, \ie~en nous intéressant à la nature des sources qui les composent.
-
-Those first outcomes tend to show that 1) two modes of perception exist depending on the very nature of the environment (i- or ni-), and 2) each involves independent features.
-The fact that $L$ is not sufficient to characterize the pleasantness of the i-scenes can lead us to conclude that all the sound sources do not equally contribute to the perception of pleasantness.
-Thus, we think that only the level of some of them can influence this percept.
-In order to investigate further in that direction, we analyze in the next section the scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.
-
 \subsubsection*{Analysis of the semantic features}
-
-\begin{figure}[!tp]
-		\myfloatalign
-        \subfloat[]
-        {\includegraphics[width=1\linewidth]{gfx/xp1_class_1}\label{fig:soundsourcea}}\par
-        \subfloat[]
-        {\includegraphics[width=.5\linewidth]{gfx/xp1_class_2}\label{fig:soundsourceb}}
-        \subfloat[]
-        {\includegraphics[width=.5\linewidth]{gfx/xp1_class_3}\label{fig:soundsourcec}}
-       \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.\gl{TODO : align}}\label{fig:soundsource}
-\end{figure}
-
-%Nous analysons la composition des scènes en comptant le nombre de sujets ayant utilisé une classe de sons pour simuler un type d'environnement.
-Les résultats sont présentés à la figure~\ref{fig:soundsourcea} pour les événements, et à la figure~\ref{fig:soundsourceb} pour les textures.
-Par souci d'espace, nous choisissons un niveau d'abstraction intermédiaire entre les niveaux 0 et 1, noté $0+$, pour représenter les classes (\cf~Figure~\ref{fig:taxonomie}).
 
@@
 
-%Nous observons une différence notable dans le choix des classes entre les i- et ni-scènes.
-La répartition des classes est très proche de celle obtenue dans une étude similaire sur les environnements sonores urbains idéaux \cite{guastavino2006ideal}, \ie~les classes suggérant la présence humaine et la nature sont très présentes dans les i-scènes, a contrario, les classes désignant des sons mécaniques et/ou de travaux sont principalement utilisées pour les ni-scènes.
-
-We observe a significant difference between i- and ni-scenes for the classes choice.
-The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~classes involving human presence and nature are prevailing in the i-scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni-scenes.
+We observe a significant difference between i- and ni- scenes for the classes choice.
+The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~classes involving human presence and nature are prevailing in the i- scenes, and on the opposite, classes involving mechanical sounds and/or public works are prevailing in the ni- scenes.
 These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.
-
-%Dans une certaine mesure, nos résultats contredisent ce fait.
-La figure~\ref{fig:soundsourcea} montre, en effet, que les classes d'événements de \emph{transports publics} (\emph{bus} et \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) ont été utilisées par les sujets, pour des i-scènes, dans $28\%$ des cas, et pour des ni-scènes, dans $42\%$ des cas.
-Les résultats ne remettent pas en question le fait que les sons de \emph{transports publics} soient bien acceptés: $25\%$ des sujets ont utilisé la classe \emph{bus} pour les i-scènes, un chiffre comparable à celui de la classe \emph{Vélo}, et bien supérieur à celui de toute autre classe de véhicules privés.
-Cependant les classes \emph{transports publics} sont également bien présentes dans les ni-scènes, plus que les classes \emph{voiture} ou \emph{camion} par exemple.
-La classe \emph{transports publics} ne peut donc pas être considérée comme typique d'un environnement sonore urbain idéal.
 
@@
 
-%Cette différence peut s'expliquer par la nature des deux protocoles expérimentaux utilisés.
-Comme nous l'avons fait, Guastavino demande à ses sujets de décrire un environnement.
-Mais ces derniers travaillent de mémoire, alors que nos propres sujets disposent de supports sonores.
-Le fait que nos sujets soient confrontés à la réalité acoustique des sons, pour recréer leurs environnements, peut avoir pour effet de diminuer l'impact du contexte socio-culturel.
-D'autres études utilisant des sons comme stimuli montrent que la classe \emph{bus} peut avoir un effet négatif sur l'appréciation de l'environnement \cite{lavandier2006contribution}.
-
 This difference may be explained by the nature of the experimental protocol used in the two studies.
@@
 
+\begin{figure}[t]
+        \myfloatalign
+        \subfloat[]
+        {\includegraphics[width=1\linewidth]{gfx/ch_5/xp1_class_1}\label{fig:soundsourcea}} \par
+        \subfloat[]
+        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_2}\label{fig:soundsourceb}}
+        \subfloat[]
+        {\includegraphics[width=.4\linewidth]{gfx/ch_5/xp1_class_3}\label{fig:soundsourcec}}
+       \caption{Proportion of synthesized scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.}\label{fig:soundsource}
+\end{figure}
+
 \subsubsection*{Sound markers}
 
-\begin{table}[t]
+We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs with respect to the type of environment (i or ni).
+Let us now investigate whether some of the sound classes are specific to a given environment.
+For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna} separately for each abstraction level.
+Results are presented in Table~\ref{tab:markers}.
+
+Regarding the sound events, 9 markers are identified for all abstraction levels.
+As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i- scenes markers as well as the \emph{church bell} class.
+This latter result may be due to the socio-cultural background of the participants who are mostly European citizens.
+In fact, according to Schafer, a sound that is identified by a person as being an element of his/her environment, is well accepted.
+Sound markers of ni- scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
+
+Regarding the textures, 5 markers are identified.
+For the i- scenes, they are classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}).
+For the ni- scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).
+
+Although the whole set of identified markers are rather intuitive, none of the event classes related to the noise of the motor vehicles are identified as markers, except for the texture class \emph{crossroad}.
+To simulate an unpleasant traffic, participants chose the classes \emph{horn} or \emph{siren}.
+We thus conclude that isolated motor vehicle sounds are understood as being part of the urban environment, and thus their nature is not necessarily linked to an unpleasant soundscape.
+
+\begin{table*}[t]
  \setlength{\tabcolsep}{0.2pt}
@@
 \vspace{0.5mm}
-\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of V-test value. $p\leq0.01$ for all sound markers.}
+\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of $V$ value.}
 \label{tab:markers}
-\end{table}
-
-%Nous avons mis en évidence que, qualitativement, la composition des sources sonores des scènes diffère selon les types d'environnement (i ou ni).
-Nous essayons de voir maintenant si, parmi ces classes, certaines sont typiques d'un environnement en particulier.
-Pour ce faire, nous utilisons le V-test (\cf~Section~\ref{sec:xp1_dataAna}), en considérant séparément chaque niveau d'abstraction.
-Les résultats sont présentés dans le tableau~\ref{tab:markers}.
-
-We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs with respect to the type of environment (i or ni).
-Let us now investigate whether some of the sound classes are specific to a given environment.
-For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna} separately for each abstraction level.
-Results are presented in Table~\ref{tab:markers}.
-
-%Concernant les événements sonores, 9 marqueurs sont identifiés sur l'ensemble des niveaux d'abstraction.
-Comme la figure~\ref{fig:soundsource} le laissait présager, les classes relatives à la présence humaine (\emph{pas homme béton}, \emph{sonnette vélo}), et à la nature (\emph{animaux, oiseaux}, \emph{chants d'oiseaux}) sont des marqueurs de i-scènes.
-Nous notons également la présence de la classe \emph{cloche} dans les marqueurs d'un environnement idéal.
-Ce fait est possiblement dû au \emph{background} socio-culturel des sujets, dans leur grande majorité, des citoyens européens.
-En effet, selon Schafer, un son reconnu par un individu comme faisant partie intégrante de son environnement est bien accepté.
-Les marqueurs de ni-scènes sont des classes faisant référence à des sons de travaux (\emph{travaux}), ou suggérant un trafic dense (\emph{klaxon}, \emph{sirène}).
-
-Regarding the sound events, 9 markers are identified for all abstraction levels.
-As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i-scenes markers as well as the \emph{church bell} class.
-This latter result may be due to the socio-cultural background of the participants who are mostly European citizens.
-In fact, according to Schafer, a sound that is identified by a person as being an element of his/her environment, is well accepted.
-Sound markers of ni-scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
-
-%Concernant les textures sonores, 5 marqueurs sont identifiés.
-Pour les i-scènes, il s'agit de classes faisant référence à des ambiances amorphes, calmes, (\emph{cour-intérieur} et \emph{parc}).
-Pour les ni-scènes, il s'agit, comme pour les événements, de classes faisant référence à des bruits de travaux (\emph{travaux} et \emph{véhicule de travaux}), ainsi que d'une classe faisant référence au trafic (\emph{carrefour}).
-
-Regarding the textures, 5 markers are identified.
-For the i-scenes, they are classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}).
-For the ni-scenes, as for the events, they are classes related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).
-
-%Bien que l'ensemble des marqueurs identifiés soient intuitifs, aucune des classes d'événements faisant directement référence aux bruits de véhicules motorisés n'est un marqueur, exception faite de la classe de textures \emph{carrefour}.
-Pour représenter un trafic désagréable, les sujets ont porté leurs choix sur les classes \emph{klaxon} et \emph{sirène}.
-On peut supposer que les sons isolés de véhicules sont compris comme faisant partie intégrante de l'environnement urbain, et ne sont donc pas particulièrement associés à un environnement désagréable.
-
-Although the whole set of identified markers are rather intuitive, none of the event classes related to the noise of the motor vehicles are identified as markers, except for the texture class \emph{crossroad}.
-To simulate an unpleasant traffic, participants chose the classes \emph{horn} or \emph{siren}.
-We thus conclude that isolated motor vehicle sounds are understood as being part of the urban environment, and thus their nature is not necessarily linked to an unpleasant soundscape.
+\end{table*}
 
 \subsubsection*{Representation space induced by semantic features}
-
-\begin{figure}[t]
-        \myfloatalign
-        \includegraphics[width=\linewidth]{gfx/pa5_1_en}
-       \caption{$P@5$ obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors as a function of the abstraction level.
-The vectors are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers $E_{w/o,m}$.}\label{fig:pa5}
-\end{figure}
-
-%Dans cette partie, nous évaluons la capacité d'une représentation sémantique à séparer les deux types d'environnement.
-Pour ce faire, nous calculons une précision au rang 5 ($p@5$) sur l'espace induit par les descripteurs sémantiques $S$, et ce pour chaque niveau d'abstraction (\cf~Section~\ref{sec:xp1_dataAna}).
-Les vecteurs $S$ sont construits en utilisant toutes les classes ($ET$), les classes d'événements ($E$), les classes de textures ($T$), les classes d'événements ne considérant que les marqueurs sonores ($E_m$), les classes d'événements ne considérant pas les marqueurs sonores $E_{w/o,m}$.
-Nous ne considérons pas les classes de marqueurs de textures, ces dernières étant trop peu nombreuses.
-Pour les mêmes raisons nous ne considérons pas les classes de marqueurs d'événements du niveau d'abstraction $0$.
-Les résultats sont affichés sur la figure~\ref{fig:pa5}.
 
@@
 
-%En ce qui concerne $ET$, la $p@5$ est de $76\%$ pour le niveau d'abstraction 0, et reste supérieure à $86\%$ à partir du niveau d'abstraction 1.
-Ces résultats confirment qu'il est possible de clairement distinguer les deux types d'environnement en se basant seulement sur la présence ou l'absence des classes de sons.
-Nous notons également que, plus le niveau d'abstraction est élevé, plus la capacité de séparer les environnements est importante.
-En d'autres termes, plus nous sommes précis dans notre description de la composition des scènes, plus nous sommes à même d'établir une distinction claire entre les i- et ni-scènes.
-
 Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and stay above $86\%$ for the abstraction level $1$.
@@
 We also notice that the higher the abstraction level is, the more important the ability to make this distinction is.
-In other words, the more precise the scene description is, the more clear the discrimination between i- and ni-scenes is.
-
-%En considérant séparément $E$ et $T$, il apparaît 1) que la $p@5$ obtenue avec $E$ est similaire à celle obtenue avec $ET$, 2) que la $p@5$ obtenue avec $T$ est systématiquement inférieure d'environ $10$ à $15\%$ à celle de $E$.
-Ces résultats indiquent que l'information sémantique permettant de séparer les deux environnements est principalement portée par les événements.
-Ces résultats font, par ailleurs, écho aux travaux de  \cite{maffiolo_caracterisation_1999}, qui montrent que nous analysons de manière descriptive (en identifiant les sources) les scènes événementielles, \ie~composées d'événements sonores (\cf~Section~\ref{sec:simscene_sampleDataSet}).
+In other words, the more precise the scene description is, the more clear the discrimination between i- and ni- scenes is.
 
@@
 
-%Enfin, il apparaît que la $p@5$ obtenue avec $E_{m}$ est égale, voire supérieure à celles obtenues avec $E$ et $ET$, et ce bien qu'une information partielle soit utilisée dans ce cas pour décrire les scènes.
-La dimension des vecteurs de description $S$ pour $E_m$ est en effet inférieure à la dimension des vecteurs $S$ pour $E$, qui est elle même inférieure à celle obtenue dans le cas où toutes les classes sont utilisées ($ET$).
-De plus, dans le cas où les marqueurs ne sont pas pris en compte pour la description ($E_{w/o,m}$), les résultats chutent, passant même en dessous de ceux obtenus en ne considérant que les textures.
-Cela confirme que la majorité de l'information sémantique permettant de faire la distinction entre i-scènes et ni-scènes est incluse dans les marqueurs.
-
 The dimension of the vectors $S$ for $E_m$ is lower than the dimension of vector $S$ for $E$, itself lower than the dimension of vector $S$ obtained when all the classes are considered ($ET$).
@@
 Moreover, in the case when the sound markers are not taken into account for the description ($E_{w/o,m}$), the performance is under those obtained with only the textures as features.
-This confirms that most of the semantic information allowing to differentiate between i- and ni-scenes is included in the markers.
-
-%En résumé, nous déduisons de cette analyse les points suivants:
-%
-%\begin{enumerate}
-%\item contrairement à ce que nous avions constaté avec les descripteurs structurels, une description sémantique de la composition des scènes, en terme de présence/absence de sources sonores, permet de bien distinguer les deux types d'environnement (i ou ni);
-%\item l'information sémantique est majoritairement portée par les classes d'événements sonores;
-%\item parmi les classes d'événements, seule une partie, \ie~les marqueurs sonores, est nécessaire pour faire la distinction entre les i- et ni-scènes.
-%\end{enumerate}
+This confirms that most of the semantic information allowing to differentiate between i- and ni- scenes is included in the markers.
 
@@
 \begin{enumerate}
-\item unlike what we outlined with the sound levels, a semantic description of the scenes constitution in terms of presence / absence of sound sources, allows us to differentiate the two types of environments (i or ni);
+\item unlike what we outlined with the structural features, a semantic description of the scenes constitution in terms of presence / absence of sound sources, allows us to differentiate the two types of environments (i or ni);
 \item the semantic information is mainly contained in the event sound classes;
-\item only a part of the event classes, i.e. the sound markers, are useful to differentiate between the i- and ni-scenes.
+\item only a part of the event classes, i.e. the sound markers, are useful to differentiate between the i- and ni- scenes.
 \end{enumerate}
 
-%Maintenant que nous avons isolé les classes typiques des i- et ni-scènes, et vérifié que la distinction entre ces environnements dépendait de la présence de ces classes, il reste à voir si une description structurelle des scènes, basée uniquement sur ces marqueurs sonores, permet de caractériser l'agrément perçu, mieux qu'une description structurelle globale.
-
-Since we have extracted the typical classes of the i- and ni-scenes and verified that the distinction between them was largely dependent on the presence of these classes, we shall now investigate whether a structural description of the scenes only based on sound markers could characterize the perceived pleasantness, perhaps better than a global structural description.
+Since we have extracted the typical classes of the i- and ni- scenes and verified that the distinction between them was largely dependent on the presence of these classes, we shall now investigate whether a structural description of the scenes only based on sound markers could characterize the perceived pleasantness, perhaps better than a global structural description.
+
+\begin{figure}[t]
+        \myfloatalign
+        \includegraphics[width=\linewidth]{gfx/ch_5/pa5_1_en}
+       \caption{$P@5$ obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors as a function of the abstraction level.
+The vectors are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers $E_{w/o,m}$.}\label{fig:pa5}
+\end{figure}
 
 \subsubsection*{Influence of the sound markers on the perceived pleasantness}
+
+We evaluate the correlations between $\mathcal{A}_{scene}$ and the structural features.
+In this section, the structural features are computed by taking into account only the previously identified sound markers.
+We define $X_m$, the $X$ feature computed by taking into account only the sound markers.
+On the opposite, we define $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
+When the feature characterizes an i- scene (idem for a ni- scene),  only the markers identified for the i- scenes (or the ni- scenes) are considered; we named them, i- markers (or ni- markers).
+Results are shown in Table~\ref{tab:corrMarkers}.
+
+Concerning the sound levels, the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ et $L(T)$ on the other side, see~Figures~\ref{fig:soundlevelMarkera} and~\ref{fig:soundlevelMarkerd}, respectively.
+No matter whether all the classes or only the markers are considered, it appears that:
+
+\begin{enumerate}
+\item a significative difference between levels of i- and ni- scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
+\item the sound level of scenes is mainly related to the sound events, compared to the textures;
+\item the sound level of events has an influence on the perception of pleasantness for ni- scenes, but not for i- scenes;
+\item the sound level of textures does not play any role in the perception of the pleasantness.
+\end{enumerate}
+
+In conclusion, the level of ni- markers has a negative influence on the pleasantness for the ni- scenes.
+On the other hand, the level of i- markers does not influence the perceived pleasantness for the i- scenes.
+
+% En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
+C'est la première fois qu'un indicateur objectif nous permet de préciser l'agrément des environnements agréables.
+Ceci nous amène à conclure que le niveau des classes de sons n'étant pas typiques d'un environnement agréable a un impact négatif sur l'agrément.
+
+Considering the non markers classes, we can observe on the i- scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}.
+It is the first time that an objective feature allows us to define the pleasantness of i- scenes.
+This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.
+
+% Par ailleurs, alors que $L(T)$ ne présentait pas de corrélation pour les ni-scènes, une corrélation négative forte est observée pour $L(T)_b$ ($r-0.73$, $p<0.01$).
+Ce fait indique que les niveaux des classes de textures n'étant pas des marqueurs n'affectent pas l'agrément perçu de la même manière pour les i- et ni-scènes.
+Les niveaux semblent avoir un effet négatif pour les ni-scènes, alors que pour les i-scènes, aucun effet n'est relevé.
+
+Besides, whereas $L(T)$ did not show any significative correlation for ni- scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
+This indicates that the level of non-markers texture classes does not influence the perceived pleasantness in the same way for i- and ni- scenes.
+Sound level seems to have a negative effect for the ni- scenes, and no significant effect for the i- scenes.
+
+% Pour finir, nous considérons un dernier groupe de descripteurs, nommément $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiffa} et~\ref{fig:soundlevelMarkerDiffd}).
+Ces descripteurs expriment la différence entre les niveaux des marqueurs, et ceux des autres classes de sons.
+Ils traduisent l'émergence des marqueurs par rapport à la mixture sonore.
+
+We consider a last group of features, namely $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$.
+These features describe the difference between the markers level and those of the other sound classes, see ~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}.
+They express the saliency of the markers with respect to the sound mixture.
+
+% Pour les i-scènes, une corrélation modérée et positive est observée pour $L_m-L_b$ ($r=0.67$, $p<0.01$) et $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
+Pour les ni-scènes, aucune corrélation n'est observée.
+Dans le cas des i-scènes, ce n'est donc pas le niveau absolu des marqueurs qui importe, mais leur niveau relatif, par rapport aux autres sons qui composent la scène.
+On observe donc pour les environnements idéaux un double mécanisme perceptif:
+
+For the i- scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
+for the ni- scenes no correlation is observed.
+Thus, for the i- scenes, it is not the absolute markers level that is important, but their relative level with regards to the other sounds composing the scene.
+Then, we can induce a double perceptual mechanism for the ideal environments:
+
+\begin{itemize}
+\item the higher the absolute level of sounds not being i- markers, the weaker the pleasantness,
+\item the higher the relative level of i- markers is, with regards to the other sounds, the higher the pleasantness.
+\end{itemize}
+
+For the ni -scenes, the fact that we observe significative correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.
 
@@
 \vspace{0.5mm}
-\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ (Exp. 1.b) and sound levels related to the presence of sound markers.}
+\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ (Exp. 1.b) and structural features related to the presence of sound markers.}
 \label{tab:corrMarkers}
@@
 
-\begin{figure}[t]
+\begin{figure*}[t]
         \myfloatalign
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_7}\label{fig:soundlevelMarkera}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_9}\label{fig:soundlevelNoisea}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_19}\label{fig:soundlevelMarkerDiffa}}\par
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_8}\label{fig:soundlevelMarkerd}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_10}\label{fig:soundlevelNoised}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
-        \caption{Distribution of the relative sound levels  related to the presence of markers $L_m$ (a, d), $L_b$ (b, e) and $L_m-L_b$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
-\end{figure}
-
-%Nous évaluons les corrélations entre $\mathcal{A}_{scene}$ et les descripteurs structurels.
-Pour cette section, les descripteurs structurels sont calculés en tenant compte des marqueurs sonores précédemment identifiés.
-Nous définissons $X_m$ le descripteur $X$ calculé en ne prenant en compte que les sons des marqueurs.
-A l'inverse, nous définissons $X_b$ ($b$: pour «~bruit~») le descripteur $X$ calculé en prenant en compte toutes les classes de sons, excepté les marqueurs.
-Lorsque le descripteur caractérise une i-scène (idem pour une ni-scène), nous ne considérons, pour le calcul, que les marqueurs identifiés pour les i-scènes (ou pour les ni-scènes), que nous nommons i-marqueurs (ou ni-marqueurs).
-Les résultats sont affichés sur le tableau~\ref{tab:corrMarkers}.
-
-We evaluate the correlations between $\mathcal{A}_{scene}$ and the sound levels.
-In this section, the sound levels are computed by taking into account only the previously identified sound markers.
-We define $X_m$, the $X$ feature computed by taking into account only the sound markers.
-On the opposite, we define $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
-When the feature characterizes an i-scene (idem for a ni-scene),  only the markers identified for the i-scenes (or the ni-scenes) are considered; we named them, i- markers (or ni- markers).
-Results are shown in Table~\ref{tab:corrMarkers}.
-
-%Concernant les niveaux sonores (\cf~Figures~\ref{fig:soundlevelMarkera} et~\ref{fig:soundlevelMarkerd}), les mêmes tendances sont observées entre $L_m$, $L(E)_m$ et $L(T)_m$, d'une part, et $L$, $L(E)$ et $L(T)$, d'autre part.
-Que l'on considère uniquement les marqueurs, ou l'ensemble des classes, il s'avère que :
-
-Concerning the sound levels, the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ and $L(T)$ on the other side, see~Figures~\ref{fig:soundlevelMarkera} and~\ref{fig:soundlevelMarkerd}, respectively.
-No matter whether all the classes or only the markers are considered, it appears that:
-
-%\begin{enumerate}
-%\item il existe une différence significative entre les niveaux des i- et ni-scènes ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
-%\item le niveau sonore des scènes est majoritairement porté par les événements sonores, comparé aux textures sonores;
-%\item le niveau sonore des événements a une influence sur la perception de l'agrément pour les ni-scènes, mais pas pour les i-scènes;
-%\item le niveau sonore des textures ne joue aucun rôle dans la perception de l'agrément.
-%\end{enumerate}
-
-\begin{enumerate}
-\item a significant difference between levels of i- and ni-scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
-\item the sound level of scenes is mainly related to the sound events, compared to the textures;
-\item the sound level of events has an influence on the perception of pleasantness for ni-scenes, but not for i-scenes;
-\item the sound level of textures does not play any role in the perception of the pleasantness.
-\end{enumerate}
-
-%En conclusion, le niveau des ni-marqueurs a une influence négative sur l'agrément pour les ni-scènes. En revanche le niveau des i-marqueurs n’impacte pas l'agrément perçu pour les i-scènes.
-
-In conclusion, the level of ni- markers has a negative influence on the pleasantness for the ni-scenes.
-On the other hand, the level of i- markers does not influence the perceived pleasantness for the i-scenes.
-
-% En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
-C'est la première fois qu'un indicateur objectif nous permet de préciser l'agrément des environnements agréables.
-Ceci nous amène à conclure que le niveau des classes de sons n'étant pas typiques d'un environnement agréable a un impact négatif sur l'agrément.
-
-Considering the non markers classes, we can observe on the i-scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}.
-It is the first time that an objective feature allows us to define the pleasantness of i-scenes.
-This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.
-
-% Par ailleurs, alors que $L(T)$ ne présentait pas de corrélation pour les ni-scènes, une corrélation négative forte est observée pour $L(T)_b$ ($r-0.73$, $p<0.01$).
-Ce fait indique que les niveaux des classes de textures n'étant pas des marqueurs n'affectent pas l'agrément perçu de la même manière pour les i- et ni-scènes.
-Les niveaux semblent avoir un effet négatif pour les ni-scènes, alors que pour les i-scènes, aucun effet n'est relevé.
-
-Besides, whereas $L(T)$ did not show any significant correlation for ni-scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
-This indicates that the level of non-markers texture classes does not influence the perceived pleasantness in the same way for i- and ni-scenes.
-Sound level seems to have a negative effect for the ni-scenes, and no significant effect for the i-scenes.
-
-% Pour finir, nous considérons un dernier groupe de descripteurs, nommément $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$ (\cf~Figures~\ref{fig:soundlevelMarkerDiffa} et~\ref{fig:soundlevelMarkerDiffd}).
-Ces descripteurs expriment la différence entre les niveaux des marqueurs, et ceux des autres classes de sons.
-Ils traduisent l'émergence des marqueurs par rapport à la mixture sonore.
-
-We consider a last group of features, namely $L_m-L_b$, $L(E)_m-L(E)_b$ et $L(T)_m-L(T)_b$.
-These features describe the difference between the markers level and those of the other sound classes, see ~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}.
-They express the saliency of the markers with respect to the sound mixture.
-
-% Pour les i-scènes, une corrélation modérée et positive est observée pour $L_m-L_b$ ($r=0.67$, $p<0.01$) et $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
-Pour les ni-scènes, aucune corrélation n'est observée.
-Dans le cas des i-scènes, ce n'est donc pas le niveau absolu des marqueurs qui importe, mais leur niveau relatif, par rapport aux autres sons qui composent la scène.
-On observe donc pour les environnements idéaux un double mécanisme perceptif:
-
-For the i-scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
-for the ni-scenes no correlation is observed.
-Thus, for the i-scenes, it is not the absolute markers level that is important, but their relative level with regards to the other sounds composing the scene.
-Then, we can induce a double perceptual mechanism for the ideal environments:
-
-%\begin{itemize}
-%\item plus le niveau absolu des sons n'étant pas des i-marqueurs est élevé, plus l'agrément est faible,
-%\item plus le niveau relatif des i-marqueurs, par rapport aux autres sons, est élevé, plus l'agrément est élevé.
-%\end{itemize}
-
-\begin{itemize}
-\item the higher the absolute level of sounds not being i- markers, the weaker the pleasantness,
-\item the higher the relative level of i- markers is, with regards to the other sounds, the higher the pleasantness.
-\end{itemize}
-
-%Pour les ni-scènes, le fait que nous observions des corrélations pour $L_m$ et $L(E)_m$, et aucune pour $L_m-L_b$ et $L(E)_m-L(E)_b$, montre que c'est bien le niveau absolu qui importe.
-
-For the ni-scenes, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_7}\label{fig:soundlevelMarkera}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_9}\label{fig:soundlevelNoisea}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_19}\label{fig:soundlevelMarkerDiffa}}\par
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_8}\label{fig:soundlevelMarkerd}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_10}\label{fig:soundlevelNoised}}
+        \subfloat[]
+        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
+        \caption{Distribution of the relative sound levels structural features related to the presence of markers $L_m$ (a, d), $L(E)_m$ (b, e) and $L(T)_m$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
+\end{figure*}
 
@@
 
-%De cette analyse, nous retenons les points suivants:
-
 From this analysis, the following points can be outlined:
@@
 \begin{itemize}
-
-%\item \emph{distinguer les i- et ni-scènes}: les descripteurs sémantiques, ainsi que les descripteurs structurels globaux ($L$, $L(E)$ et $L(T)$), permettent de faire la distinction entre les i-scènes et les ni-scènes.
-La description sémantique semble être plus performante;
-%\item \emph{événements ou textures}: que ce soit pour les descripteurs sémantiques ou structurels, ce sont majoritairement les événements qui permettent de distinguer les deux types d'environnement, les textures n'apportant, au mieux, qu'une information limitée;
-%\item \emph{prédire l'agrément}: si l'on considère les corrélations entre les descripteurs structurels et l'agrément, il semble que la manière de percevoir la qualité de l'environnement diffère en fonction de la nature de ce dernier (i ou ni).
-Il n'apparaît pas envisageable de considérer un même jeu de descripteurs pour prédire, à la fois, l'agrément des i-scènes, et l'agrément des ni-scènes :
-
-\item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni-scenes.
+\item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global structural features ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni- scenes.
 The semantic description seems to be more powerful;
 \item \emph{events or textures}: whatever the feature type, be it semantic or structural, events are mainly useful to differentiate the two types of scenes; textures bring a limited amount of information;
-\item \emph{pleasantness prediction}: considering the correlation between sound levels and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its nature (i or ni).
-From the data gathered in those experiments, it is not conceivable to consider the same set of features to predict the pleasantness of i- and ni-scenes in the same time:
-
-\begin{itemize}
-
-%\item \emph{pour les ni-scènes}, ce sont le niveau global ($L$ et $L(E)$), ou le niveau des marqueurs sonores ($L_m$ et $L(E)_{m}$), qui impactent négativement l'agrément.
-On note ici que prendre en compte les contributions de différentes sources n'améliore pas la capacité de prédiction de l'agrément, par rapport à une analyse holistique de l'environnement.
+\item \emph{pleasantness prediction}: considering the correlation between structural features and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its nature (i or ni).
+From the data gathered in those experiments, it is not conceivable to consider the same set of features to predict the pleasantness of i- and ni- scenes in the same time:
+
+\begin{itemize}
 
@@
 
-%\item \emph{pour les i-scènes}, par contre, prédire l'agrément requiert d'étudier, de manière séparée, les caractéristiques des marqueurs sonores, et celles de l'ensemble des autres sons.
-Ainsi, le niveau des marqueurs pris relativement au bruit ($L(E)_m-L(E)_b$ et $L_m-L_b$) est positivement corrélé à l'agrément, alors que le niveau du bruit ($L_b$ et $L(E)_b$) est, lui, négativement corrélé.
-
 \item \emph{for i-scenes}, on contrary, the pleasantness prediction needs to consider separately the sound markers characteristics and those of the other sounds.
 Thus, the markers level, relative to the noise ($L(E)_m-L(E)_b$ and $L_m-L_b$) is positively correlated to the pleasantness, whereas the noise level ($L_b$ and $L(E)_b$) is negatively correlated.
-
-\end{itemize}
-\end{itemize}
-
-%Le fait que l'agrément des i-scènes ne se corrèle pas avec des descripteurs physiques holistiques, contrairement à l'agrément des ni-scènes, a récemment été déjà observé \cite{gozalo2015relationship}.
-
-The fact that the pleasantness of the i-scenes is not correlated to global physical features, contrary to pleasantness of ni-scenes, has been studied recently \cite{gozalo2015relationship}.
-
-%L'existence de deux modes de perception, mobilisant différents types de descripteurs, et dépendant de la nature (dans notre cas hédonique) des stimuli, est un phénomène qui se rapproche de celui observé pour la perception des textures (\cf~Section~\ref{sec:simscene_sampleDataSet}).
-Le cerveau adapte sa manière de traiter l'information (résumé statistique pour les textures, description fine pour les événements) suite à une prise de décision antérieure quant à la nature du stimulus (à savoir «~est-ce un événement ou une texture ?~»).
-De la même manière, les indicateurs actifs dans le jugement de l'agrément dépendent, eux aussi, d'une identification préalable de la nature hédonique globale de l'environnement (idéale ou non idéale).
+\end{itemize}
+\end{itemize}
+
+The fact that the pleasantness of the i- scenes is not correlated to global physical features, contrary to pleasantness of ni- scenes, has been studied recently \cite{gozalo2015relationship}.
 
@@
 \item \emph{72 km-scenes}: the 72 scenes originally simulated by the subject of experiment 1.a where the sound classes identified as markers are kept.
-The 36 ideal scenes with markers are noted i/km-scenes, and the 36 non ideal scenes with markers are noted ni/km-scenes (km for kept markers).
+The 36 ideal scenes with markers are noted i/km-scenes, and the 36 non ideal scenes with markers are noted ni/km-scenes (k for kept).
 \item \emph{72 rm-scenes}:  72 scenes where the sound classes identified as markers are removed.
-The 36 ideal scenes without markers are noted i/rm-scenes, and the 36 non ideal scenes without markers are noted ni/rm-scenes (r for removed markers).
+The 36 ideal scenes without markers are noted i/rm-scenes, and the 36 non ideal scenes without markers are noted ni/rm-scenes (r for removed).
 \end{itemize}
@@
 
-In order to create rm-scenes with still some sound diversity and no  absence of sound activity within a long period of the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}.
+In order to create km-scenes with still some sound diversity and no  absence of sound activity within a long period of the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}.
 Those classes are:
@@
 
-12 subjects perform the test (4 females; averaging $29.5$ years of age, s.d.
-of $14$ years).
+12 subjects (4 women) perform the test.
 None of them have performed experiments 1.a and 1.b.
+The subjects are 22 to 61 years old (average: 29.5, standard deviation: 14).
 All the subjects live in Nantes, France, for at least two years.
@@
 The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness.
-To do so, we perform an Analysis of Variance.
-We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm).
+To do so, we perform a variance analysis.
+We consider $\mathcal{A}_{sujet}$ as a dependent variable, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm).
 As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to check the validity of the hypothesis that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes.
@@
 \subsubsection{Outliers detection}
-
-\begin{figure}[t]
-        \myfloatalign
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp4_note_13}\label{fig:xp2_Asubject}}
-        \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp4_note_15}\label{fig:xp2_Ascene}}
-        \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness  per scenes$\mathcal{A}_{scene}$ (b) versus scene type.
-Black stars in subfigure (a) stand for the detected outlier, \ie~subject 7.}\label{fig:xp2A}
-\end{figure}
 
@@
 
-Let us consider $\mathcal{A}_{subject}$ for the km-scenes, see Figure~\ref{fig:xp2_Asubject}.
-Close inspection of the answers show that one subject strongly differs from the ones of others.
+Let us consider $\mathcal{A}_{sujet}$ for the km-scenes.
+Close inspection of the answers show that one subject strongly differs with the ones of others.
 This subject evaluated positively close to half of the ni/km-scenes, see Annex~\ref{app:xp2}, subject 7.
@@
 
-In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes and ni/rm-scenes, see Figure~\ref{fig:xp2_Ascene}.
-The repeated measure ANOVA applied to $\mathcal{A}_{subject}$ show a significant effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
+In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes and ni/rm-scenes.
+The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ show a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
 
@@
 
-The \emph{post hoc} analysis exhibits significant differences between all the groups of observations, notably between the i/km- and the i/rm scenes ($p<0.05$) as well as between the ni/km- and the ni/rm-scenes ($p<0.01$).
+The \emph{post hoc} analysis exhibits significative differences between all the groups of observations, notably between the i/km-scenes ($p<0.05$) and the i/rm-scenes ($p<0.01$).
 
@@
 
-These results indicate that the removal of the markers did indeed modify the perception of the scenes by the subjects. Our two hypotheses are thus verified :
+These results indicate that the removal of the markers did indeed modify the perception of the scenes by the subjects. Our two hypotheses are thus verified:
 
@@
 
-The significant interaction shows that the effect of the type of environment influences the effect of the presence/absence of the markers.
+The significative interaction shows that the effect of the type of environment influences the effect of the presence/absence of the markers.
 Indeed, the average of the differences between km-scenes and rm-scenes is larger for the ni-scenes ($1.1$) than for the i-scenes ($0.5$).
@@
 The i-markers do have a positive impact on the perception of an environment.
-The fact that their removal decreases $\mathcal{A}_{scene}$ clearly shows that it is possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
+The fact that their removal decreases $\mathcal{A}_{scene}$ clearly shows that it is possible to improve the perceived pleasantnes of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
 Those conclusions are in line with the positive approach introduced by Schafer in \cite{schafer1977tuning}.
@@
 
-These results tend to confirm that the perception of the qualities of a scene indeed depends primarily on the sound sources that compose it.
+These results tend to confirm that the perception of the qualities of a scene indeed depends above all on the sound sources that compose it.
 The compositional characteristics that are taken into account during the perceptual process appear to vary from one source to the other, from one type of environment to the other.
-This fact leads the authors to believe that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.
+This fact lead the authors to believe that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.
 
@@
 \bibliography{Bibliography}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\onecolumn
+
 \appendix
-\newpage
 \section{Taxonomy of sound classes}
@@
 
-\begin{figure}[!hp]
+\begin{figure}[h]
     \centering
-        {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.9\columnwidth]{gfx/appendix/event_1_en}\label{fig:taxonomieEventa}} \par
+        {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.7\columnwidth]{gfx/ch_5/event_1_en}\label{fig:taxonomieEventa}} \par
          \caption{Taxonomy of sound classes of mechanical events used for the simulation urban soundscapes  with level of abstraction from left to right.}
@@
 
-\begin{figure}[hp]
+\begin{figure}[h]
     \centering
-        {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.9\columnwidth]{gfx/appendix/event_2_en}\label{fig:taxonomieEventb}}
+        {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.7\columnwidth]{gfx/ch_5/event_2_en}\label{fig:taxonomieEventb}}
        \caption{Taxonomy of sound classes of non-mechanical events used for the simulation urban soundscapes with level of abstraction from left to right.}
@@
 
-\begin{figure}[hp]
+\begin{figure}[h]
   \centering
-        {\includegraphics[trim={ 0 0 7cm 0},clip,width=.7\columnwidth]{gfx/appendix/texture_1_en}\label{fig:taxonomieTexturea}} \par
-
-        {\includegraphics[trim={ 0 0 7cm 1cm},clip,width=.7\columnwidth]{gfx/appendix/texture_2_en}\label{fig:taxonomieTextureb}}
+        {\includegraphics[trim={ 0 0 7cm 0},clip,width=.7\columnwidth]{gfx/ch_5/texture_1_en}\label{fig:taxonomieTexturea}} \par
+
+        {\includegraphics[trim={ 0 0 7cm 1cm},clip,width=.7\columnwidth]{gfx/ch_5/texture_2_en}\label{fig:taxonomieTextureb}}
        \caption{Taxonomy of sound classes of textures used for the simulation urban soundscapes with level of abstraction from left to right.}
@@
 \newpage
-\section{Distribution of pleasantness scores given by the subjects during experiment 2}
+\section{Distribution of scores given par the subjects during experiment 2}
 \label{app:xp2}
@@
         \subfloat[subject 1]
-        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_1}\label{fig:xp4_note_1a}}
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_1}\label{fig:xp4_note_1a}}
         \subfloat[subject 2]
-        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_2}\label{fig:xp4_note_1b}}
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_2}\label{fig:xp4_note_1b}}
         \subfloat[subject 3]
-        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_3}\label{fig:xp4_note_1c}}
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_3}\label{fig:xp4_note_1c}}
         \subfloat[subject 4]
-        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_4}\label{fig:xp4_note_1d}} \par
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_4}\label{fig:xp4_note_1d}} \par
         \subfloat[subject 5]
-        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_5}\label{fig:xp4_note_1e}}
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_5}\label{fig:xp4_note_1e}}
         \subfloat[subject 6]
-        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_6}\label{fig:xp4_note_1f}}
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_6}\label{fig:xp4_note_1f}}
         \subfloat[subject 7]
-        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_7}\label{fig:xp4_note_1g}}
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_7}\label{fig:xp4_note_1g}}
         \subfloat[subject 8]
-        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_8}\label{}} \par
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_8}\label{}} \par
         \subfloat[subject 9]
-        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_9}\label{fig:xp4_note_1i}}
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_9}\label{fig:xp4_note_1i}}
         \subfloat[subject 10]
-        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_10}\label{fig:xp4_note_1j}}
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_10}\label{fig:xp4_note_1j}}
         \subfloat[subject 11]
-        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_11}\label{fig:xp4_note_1z}}
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_11}\label{fig:xp4_note_1z}}
         \subfloat[subject 12]
-        {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_12}\label{fig:xp4_note_1l}}
+        {\includegraphics[width=.24\linewidth]{gfx/ch_5/xp4_note_12}\label{fig:xp4_note_1l}}
         \caption{Distribution of scores given par the subjects during experiment 2 for i/am-scenes (green) et ni/am-scenes (red).
-The horizontal axis is, from left to right, the 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant).}\label{fig:xp4_note_1}
+The horizontal axis is, from left to right, the 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant)}\label{fig:xp4_note_1}
 \end{figure}

### 2017-12-21 lafaySimulatedScenePerception.tex
more

--- 2017-12-19 lafaySimulatedScenePerception.tex
@@
 \item \emph{analysis of the semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni).
-To do so, a V-test is considered \citep{lebart2000statistique} \ml{ref en Anglais ?}.
+To do so, a V-test is considered \citep{rakotomalala2008tvpercent}.
 With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of time a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
@@
 Considering only the presence / absence of sound classes thus allows us to nicely discriminate between the two types of environments.
-We also notice that the higher the abstraction level is, thus the more precise the description, the more efective the description is.
+We also notice that the higher the abstraction level is, thus the more precise the description, the more effective the description is.
 
@@
 
-\item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ et $L(T)$), allow us to differentiate between i- and ni-scenes.
+\item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ et $L(T)$), allow us to differentiate reliably between i- and ni-scenes.
 The semantic description seems to be more powerful;
-\item \emph{events or textures}: whatever the feature type, be it semantic or related to the sound pressure level, events are mainly useful to differentiate the two types of scenes; textures bring a limited amount of information;
-\item \emph{pleasantness prediction}: considering the correlation between sound levels and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its nature (i or ni).
-From the data gathered in those experiments, it is not conceivable to consider the same set of features to predict the pleasantness of i- and ni-scenes in the same time:
+\item \emph{events or textures}: whatever the feature type, be it semantic or related to the sound pressure level, events are the most useful components of the scene to differentiate the two types of scenes; textures bring a limited amount of information;
+\item \emph{pleasantness prediction}: considering the correlation between sound levels and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its very nature (i or ni).
+From the data gathered in those experiments, the same set of features cannot be considered to predict the pleasantness of i- and ni-scenes:
 
@@
 
-\item \emph{for ni-scenes}, global levels ($L$ et $L(E)$), or level of sound markers ($L_m$ et $L(E)_{m}$), have a negative influence on pleasantness.
-We can notice here that taking into account the contribution of each of the different sources of the scene does not improve the prediction performance compared to a global analysis of the environment.
+\item \emph{for ni-scenes}, the global levels ($L$ et $L(E)$), or the level of sound markers ($L_m$ et $L(E)_{m}$), have a negative influence on pleasantness.
+Taking into account the contribution of each of the different sources of the scene does not improve the prediction performance compared to a global analysis of the environment.
 
@@
 
-\item \emph{for i-scenes}, on contrary, the pleasantness prediction needs to consider separately the sound markers characteristics and those of the other sounds.
-Thus, the markers level, relative to the noise ($L(E)_m-L(E)_b$ and $L_m-L_b$) is positively correlated to the pleasantness, whereas the noise level ($L_b$ and $L(E)_b$) is negatively correlated.
+\item \emph{for i-scenes}, on contrary, the sound markers characteristics and those of the other sounds have to be considered separately to predict  the pleasantness.
+The markers level, relative to the background noise ($L(E)_m-L(E)_b$ and $L_m-L_b$) is positively correlated to the pleasantness, whereas the noise level ($L_b$ and $L(E)_b$) is negatively correlated.
 
@@
 
-The fact that the pleasantness of the i-scenes is not correlated to global physical features, contrary to pleasantness of ni-scenes, has been studied recently \cite{gozalo2015relationship}.
+The fact that the pleasantness of the i-scenes is not correlated to global physical features, contrary to pleasantness of ni-scenes, has also been studied recently \cite{gozalo2015relationship}.
 
@@
 
-We observe two perceptual modes of operation that involve different types of features, and that rely on the hedonic nature  of the stimuli.
+We can assume two perceptual modes of operation that involve different types of features and rely on the hedonic nature  of the stimuli.
+It thus appears that the features considered for the pleasantness judgment also depends on a preliminary judgment of the global hedonic nature of the environment (ideal or non ideal).
+
 A similar phenomenon is observed for the perception of textures, see Section~\ref{sec:simscene_sampleDataSet}.
-It seems that the brain adapts its way to encode the information (statistic summary for textures, fine description for events) following a previous decision making with regards to the nature of the stimuli, \ie~is it an event or a texture ?
-Similarly, in the light of this study, it appears that the features considered for the pleasantness judgment also depends on a preliminary judgment of the global hedonic nature of the environment (ideal or non ideal).
+It seems that the brain selectively adapts the way it encodes the information (statistic summary for textures, finer description for events) following a previous decision making process based on the nature of the stimuli, \ie~is it an event or a texture ?
 
@@
 
-In order to verify that the pleasantness of the ideal and non ideal scenes is affected by the sound markers, the audio waveform of the simulated scenes are regenerated with or without the classes identified as markers.
-For the i-scenes, some of the i-markers are removed.
-For the ni-scenes, some of the ni-markers are removed.
-To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptual test is then conducted with a protocol close to the one considered in experiment 1.b.
+In order to investigat deeper into the relation between the pleasantness of the ideal and non ideal scenes and the sound markers, the audio waveform of the simulated scenes are regenerated with or without the classes identified as markers.
+To do so, some of the i-markers are removed for the i-scenes and some of the ni-markers are removed for the ni-scenes.
+To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptual test is conducted with a protocol close to the one considered in experiment 1.b.
 
@@
 \begin{itemize}
-\item \emph{for the ni-scenes} we hypothesize that the absence of ni-markers will \textbf{increase} the pleasantness score;
-\item \emph{for the i-scenes} we hypothesize that the absence of the i-markers will \textbf{decrease} the pleasantness score.
+\item \emph{for the ni-scenes}, we hypothesize that the absence of ni-markers will \textbf{increase} the pleasantness score;
+\item \emph{for the i-scenes}, we hypothesize that the absence of the i-markers will \textbf{decrease} the pleasantness score.
 \end{itemize}

### 2017-12-21 lafaySimulatedScenePerception.tex
done with proof reading

@@
 \item \emph{analysis of the perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and ni-scenes during the experiment 1.a.
-To do so, we consider if there exist significative differences for the i- and ni-scenes of the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$.
+To do so, we consider if there exist significant differences for the i- and ni-scenes of the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$.
 The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$;
@@
 Indeed, it is not obvious that the removal of the i-markers, though perceptively positively connoted, decreases the global quality of a soundscape.
-Indeed, this removal also decreases the global level of the scene.
+Indeed, this removal also decreases the global sound level of the scene.
 
@@
 \begin{itemize}
-\item \emph{Church bell}, \emph{bicycle bell}, and \emph{animals} for the i/rm-scenes;
+\item \emph{church bell}, \emph{bicycle bell}, and \emph{animals} for the i/rm-scenes;
 \item \emph{siren}, \emph{car horn} for the ni/rm-scenes.
@@
 
-A supervisor is available during the whole duration of the experiment to monitor the experiment and answer any questions the subjects may have.
+A supervisor is available during the whole duration of the experiment to monitor the experiment and answer to any questions the subjects may have.
 
@@
 We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm).
-As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to check the validity of the hypothesis that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes.
+As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to evaluate that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes.
 The two independent variables are considered as \emph{within-subject} factors.
@@
 
-These results indicate that the removal of the markers did indeed modify the perception of the scenes by the subjects. Our two hypotheses are thus verified :
+These results indicate that the removal of the markers indeed modify the perception of the scenes by the subjects. Our two hypotheses are thus verified :
 
@@
 The i-markers do have a positive impact on the perception of an environment.
-The fact that their removal decreases $\mathcal{A}_{scene}$ clearly shows that it is possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
+The fact that their removal decreases $\mathcal{A}_{scene}$ indicate that it should possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
 Those conclusions are in line with the positive approach introduced by Schafer in \cite{schafer1977tuning}.
@@
 
-This series of three experiments showed that most of the descriptors used in this study, be they semantic or compositional, allows us to distinguish between a said ideal scene and a non ideal one.
+This series of three experiments showed that most of the descriptors used in this study, be they semantic or based on acousitc level, allows us to distinguish between an ideal scene and a non ideal one.
 
@@
 
-These results tend to confirm that the perception of the qualities of a scene indeed depends primarily on the sound sources that compose it.
-The compositional characteristics that are taken into account during the perceptual process appear to vary from one source to the other, from one type of environment to the other.
+These results show that the perception of the qualities of a scene indeed depends primarily on its sound sources which are identified.
+The characteristics that are taken into account during the perceptual process appear to vary from one source to the other, from one type of environment to the other.
 This fact leads the authors to believe that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.
@@
 
-First, the fact that the semantic information (in terms of which sound source is present) and compositional information are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the PE and UE concepts.
+First, the fact that the semantic information (which sound source is present) and compositional information (at which level) are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the PE and UE concepts.
 
@@
 
-Secondly, the fact that the removal of sound markers modifies the perceived pleasantness leads us to believe that the abstract concept related to pleasantness depends on the activation of a network of concrete concepts strongly linked to the sources which are in the case of this study: \emph{bird}, \emph{church bell} and \emph{bicycle bell}.
+Secondly, the fact that the removal of sound markers changes the perceived pleasantness leads us to believe that the abstract concept related to pleasantness depends on the activation of a network of concepts strongly linked to the sources which are in the case of this study: \emph{bird}, \emph{church bell} and \emph{bicycle bell}.
 
@@
 
-Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics that allow us to distinguish between two instances of the same abstract concept linked to the pleasantness are not the same for PE and UE.
+%Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics that allow us to distinguish between two instances of the same abstract concept linked to the pleasantness are not the same for PE and UE.
 
@@
 The outcomes of the three experiments described in this paper demonstrate the usefulness of considering a dedicated simulation tool such as \emph{simScene} in order to scientifically question the perception of soundscapes in an innovative way.
-We also believe that its wider usage could enable urban planning decision-makers to question an entire community about its own representations of the sound environment to which they are exposed, and about the representations of the sound environments to which they would like to be exposed.
+We also believe that its wider usage could enable urban planning decision-makers to question an entire community about its own representations of the sound environment to which they are exposed, and about the representations of the sound environments to which they would like to be exposed to.
 
@@
 
-Future work should consider several other simulation experiments by varying the emotional qualities (quiet, comfortable, troublesome, etc.), but also by specifying specific urban locations (park, square, street, etc.), in order to provide to the scientific community an entire corpora of cognitively informed soundscapes.
+Future work should consider several other simulation experiments by changing the emotional qualities (quiet, comfortable, troublesome, etc.), but also by specifying specific urban locations (park, square, street, etc.), in order to provide to the scientific community an entire corpora of cognitively informed soundscapes.
 
@@
 Finally, one should study the influence of socio-cultural contexts on perception.
-Indeed, if the sound of the church bell is most often a quality environment marker for a Westerner, this do not necessarily hold true for subjects of Eastern, Middle Eastern or other cultures.
+Indeed, if the sound of the church bell is most often a quality environment marker for a Westerner.
+This do not necessarily hold true for subjects of Eastern, Middle Eastern or other cultures.
 
@@
 \begin{itemize}
-\item The simulator can be deployed on a large scale via the Internet thanks to the chosen software architecture;
+\item The simulator can be deployed on a large scale via the Internet thanks to the web-based software architecture;
 \item simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.

### 2018-01-17 lafaySimulatedScenePerception.tex
mathias comments

@@
 \usepackage{graphicx}
-\usepackage{subfig}
+\usepackage[position=top]{subfig}
 \usepackage{caption}
@@
 \allsectionsfont{\bfseries}
+
+\makeatletter
+\newcommand*{\centerfloat}{%
+  \parindent \z@
+  \leftskip \z@ \@plus 1fil \@minus \textwidth
+  \rightskip\leftskip
+  \parfillskip \z@skip}
+\makeatother
 
@@
 
-One of the main goal of soundscape studies is to identify which components of the soundscape influence human perception \cite{aletta2016soundscape}.
-For example, in order to study the link between an environment (urban sound scene) and the induced human sensation (calmness), one mean is to objectify the mental representation of a specific sound scene (a calm urban sound scene).
+One of the main goals of soundscape studies is to identify which components of the soundscape influence human perception \cite{aletta2016soundscape}.
+Establishing a link between an environment (urban sound scene) and the induced human sensation (calmness) is equivalent to instantiating the mental representation of a specific sound scene (a calm urban sound scene).
 Several means have been considered in the literature to perform this task.
@@
 First, a subject can be asked to assess a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}.
-The amount of information that can be gathered strongly depends on the very nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}.
-That being said, those recorded stimuli  can be analyzed, what allows the experimenter to gather a coarse
-grain physical description of them.
+The amount of information that can be gathered strongly depends on the nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}.
+Working with actual stimuli makes it possible to analyze them, thus allowing the experimenter to gather a coarse-grained physical description of them.
 
@@
 
-We propose in this paper to consider the use of a soundscape simulator that the subject can manage to objectify his/her representation of a given sound environment.
-We believe that the use of such a device allows us to gain the benefits of the two above cited approaches.
-As the subject is asked to produces audio data (the signal of the simulated scene), it allows the experimenter to study a precise modal version of the mental representation of the subject that is both characterized semantically and physically, respectively in terms of the sound sources and their levels.
+We propose in this paper to consider the use of a soundscape simulator that the subject can employ to objectify his/her representation of a given sound environment.
+We believe that the use of such a device allows us to gain the benefits of the two abovementioned approaches.
+As the subject is asked to produce audio data (the signal of the simulated scene), it allows the experimenter to study a precise modal version of the subject's mental representation that is characterized both semantically (the nature of sound sources) and physically (\eg~the levels of sound sources).
 
@@
 
-We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent research studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,
-szeremeta2009analysis}.
+We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,szeremeta2009analysis}.
 Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}.
@@
 
-For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed in this article can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment of interest.
-With such a material, not only the type of sound sources is indeed available for study, but also the exact level and audio waveform for each source together with the structural characteristics of the scene, that is the temporal distribution of the sound events.
+For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed here can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment.
+With such material, not only is the type of sound sources available for study, but also the exact level and audio waveform for each source together with the structural properties of the scene, that is, the temporal distribution of the sound events.
 
@@
 
-To demonstrate the potential of the proposed approach in its ability to question the human perception of the sound environment, we study in this paper the notion of perceptual pleasantness of urban environmental scenes.
-Results and outcomes of a series of three experiments that foster on the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:
+To demonstrate the potential of the proposed approach in its ability to question the human perception of sound environments, we study in this paper the notion of perceptual pleasantness of urban environmental scenes.
+Results and outcomes of a series of three experiments that build upon the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:
 
@@
 \begin{enumerate}
-\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non ideal soundscapes that are considered as material for the following experiments
-\item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale
+\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non ideal soundscapes that are considered as material for the following experiments;
+\item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
 \item \emph{experiment 2, pleasantness evaluation after modification of the scenes}: the subjects judge the pleasantness of the simulated scenes on a semantic scale as in 1.b, but some scenes are modified beforehand, \ie~some specific sounds classes that are identified as having a significant impact on perceived pleasantness are removed.
@@
 
-In our approach, the simulator developed for this study\footnote{simScene, available online \url{https://bitbucket.org/mlagrange/simscene}} gives a monophonic representation of the scene, with the added benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are both expressive and useful for analysis.
+In our approach, the simulator developed for this study\footnote{simScene, available online \url{https://bitbucket.org/mlagrange/simscene}} only yields a monophonic representation of the scene, but that simplification comes with the benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are both expressive and useful for analysis.
 
@@
 The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}.
-The experiments  1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section \ref{sec:simulation}.
-Experiment 2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section \ref{sec:modification}.
+Experiments~1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section~\ref{sec:simulation}.
+Experiment~2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section~\ref{sec:modification}.
 Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.
@@
 It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}.
-It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{web-audio} standard that allows the manipulation of digital audio data within the browser\footnote{\emph{web-audio}: \url{www.w3.org/TR/webaudio}}.
-The interface for selecting the sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.
+It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{Web Audio} standard that allows the manipulation of digital audio data within the browser\footnote{\emph{Web Audio}: \url{www.w3.org/TR/webaudio}}.
+The interface for selecting sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.
 
@@
 The user can then manipulate some parameters detailed in Section~\ref{sec:simscene_parametre} to control the time and magnitude distribution of the occurrences of the sound sources.
-Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) to provide some comments about the simulated scene.
+Text fields are also available for the user to 1) name each track, 2) name the entire scene, 3) provide free comments about the simulated scene.
 
@@
 
-We choose not to include any musical content in the sound database, as the study of the pleasantness of a given style or genre of music is beyond the scope of the study.
+We choose not to include any musical content in the sound database, as the study of the pleasantness of a given style or genre of music is beyond the scope of this study.
 
@@
 
-The literature usually distinguishes between:
-
-\begin{itemize}
-\item {a sound event}: an isolated sound, sparsely distributed whose acoustical characteristics may change with respect to time
-\item {a sound texture}: an isolated sound of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
+ A commonly accepted distinction consists in separating:
+
+\begin{itemize}
+\item {sound events}: isolated sounds, precisely located in time, whose acoustical characteristics may change with respect to time;
+\item {sound textures}: isolated sounds of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
 \end{itemize}
@@
 
-The simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with the classes of events only and the other with the classes of textures only.
+%Les différences morphologiques entre événements et textures sonores ont des conséquences sur la perception des scènes.
+
+In order to account for the possiblity that the morphological differences between those two classes may have some important consequences on the perception of the scenes, the simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with classes of events only and the other with classes of textures only.
 These two types of sound classes also have specific scheduling procedures during the simulation process.
-
-%Les différences morphologiques entre événements et textures sonores ont des conséquences sur la perception des scènes.
-
-We do so because we assume that the morphological differences between those two classes have some important consequences on the perception of the scenes.
 
@@
 According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment \cite{bregman1994auditory}.
-Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}.
-In this case, the source is a perceptual entity as events that are not isolated are grouped into a stream.
+Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}, while events that cannot be isolated are grouped into a same stream.
 
@@
 
-Maffiolo \cite{maffiolo_caracterisation_1999} distinguishes two distinct categorization processes, that differentiate themselves with the capacity of the listener to identify sound events.
-She shows that those processes leads to two abstract cognitive category respectively termed "the event sequences" and "the amorphous sequences".
+Maffiolo \cite{maffiolo_caracterisation_1999} distinguishes two separated categorization processes, either of which is triggered depending on the listener's ability to identify sound events.
+She shows that those processes lead to two abstract cognitive categories respectively termed "event sequences" and "amorphous sequences".
 
@@
 
-The event sequences are composed of salient events that can easily be recognized such as \emph{car start} or \emph{male speech}.
+Event sequences are composed of salient events that can easily be recognized, such as \emph{car start} or \emph{male speech}.
 They arise from a descriptive analysis based on the identification of the sound sources.
-On the contrary, the amorphous sequences are sound environments where distinct events can hardly be found, such as \emph{traffic hubhub}.
+On the contrary, amorphous sequences are sound environments where distinct events cannot be readily identified, such as \emph{traffic hubhub}.
 They result from a holistic analysis based on global acoustic features.
@@
 
-Concerning sound textures, \ie~sounds that have stable characteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can decide not to abstract the physical information available using some of the summary statistics discarding lower level information \cite{nelken2013ear}.
+Concerning sound textures, \ie~sounds that have stable characteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can opt for an abstract, statistical representation of the perceived information, discarding precise physical properties of the sound \cite{nelken2013ear}.
 
@@
 
-At the light of those results, it appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli \cite{nelken2013ear,mcdermott2013summary}.
-We thus consider in this study the soundscape as a "skeleton of events on a bed of textures" as coined in \cite{nelken2013ear}.
+In the light of those results, it appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli \cite{nelken2013ear,mcdermott2013summary}.
+We thus consider in this study the soundscape as a ``\,skeleton of events on a bed of textures\,'' as coined in \cite{nelken2013ear}.
 
@@
 
-We term "sample", a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples perceived equivalently.
+We call ``\,sample\,'' a recording of an isolated sound, be it an event or a texture. Each sound class is implemented as a collection of samples judged to be perceptually equivalent.
 
@@
 
-The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure that is close to the one of the vertical axis of the categorical organization proposed by E.
+The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure similar to the vertical axis of the categorical organization proposed by E.
 Rosch \cite{rosch1978cognition}.
 The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources.
-For classes with high level of abstraction that have sub classes, their collection of samples is the union of the collections of the sub classes.
+For classes with a high level of abstraction that have sub classes, their collection of samples is the union of the collections of the sub classes.
 
@@
 
-Accounting for the perceptual matters detailed previously, two taxonomies are built, one for event sounds and one for texture sounds.
-Four levels of abstraction are considered from the more generic one (level 0), to the more specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}.
+Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds.
+Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}.
 Only three levels of abstraction are considered for the texture sounds.
@@
 
-Original sounds have been recorded using a shotgun microphone \emph{AT8035}\footnote{\emph{Micro canon AT8035}: \url{eu.audio-technica.com}} plugged to a \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}} recorder.
+Original sounds have been recorded using a shotgun microphone \emph{AT8035}\footnote{\emph{AT8035 shotgun microphone}: \url{eu.audio-technica.com}} plugged to a \emph{ZOOM H4n}\footnote{\emph{ZOOM H4n} : \url{www.zoom.co.jp/english/products/h4n}} recorder.
 The use of such a microphone allowed us to isolate as much as possible sound events of interest from the urban background.
@@
 
-By a voluntary design choice, the simulation tool does not allow the user to interact and control directly a specific sample.
+By a deliberate design choice, the simulation tool does not allow the user to interact with and control directly a specific sample.
 Interaction is done at the track level, a track being a sequence of samples.
@@
 \begin{itemize}
-\item \emph{sound level} ($dB$): for each sample, the sound levels are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
-\item \emph{inter-onset spacing} (second): for the event's track only, and for each sound event sample, the inter-onset spacings are randomly sampled according to a normal distribution parametrized by the subject in terms of mean value and variance;
-\item \emph{start and end time} (second): the subject set the first and last occurrence of a sample of the selected class.
+\item \emph{sound level} ($dB$): for each sample, the sound levels are drawn randomly following a normal distribution parameterized by the subject in terms of mean value and variance;
+\item \emph{inter-onset spacing} (second): for event tracks only, and for each sound event sample, the inter-onset spacings are drawn randomly following a normal distribution parameterized by the subject in terms of mean value and variance;
+\item \emph{start and end time} (second): the subject sets the start and end times between which the texture or sequence of repeated events occurs.
 \end{itemize}
@@
 \begin{itemize}
-\item \emph{event's fades} (seconds): for the event's track only, the subject can set a fade in / fade out duration applied to each sample;
-\item \emph{global fades} (seconds): the subject can set global fade in / fade out duration applied to the entire track.
+\item \emph{event fades} (seconds): for the event tracks only, the subject can set a fade in / fade out duration applied to each sample;
+\item \emph{global fades} (seconds): the subject can set global fade in and fade out durations applied to the entire track.
 \end{itemize}
@@
 
-The samples of texture are sequenced without time spacing, therefore the parameters \emph{event's fade} and \emph{inter-onset spacing} are not available for this kind of track.
+Texture samples are sequenced without time spacing, therefore the parameters \emph{event fade} and \emph{inter-onset spacing} are not available for this kind of track.
 
@@
         {\includegraphics[width=\linewidth]{gfx/simscene}\label{fig:simscene}}
-       \caption{Graphical interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{simScene} tool.}
+       \caption{\emph{SimScene} graphical interfaces for the selection of sound classes (a) and their sequencing (b).}
 \end{figure}
@@
 Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user.
-Most browsing tools are based on keyword indexing.
-However, for sensory experiment that study the objectivation of the mental representation of the subject, this mean is problematic as the availability of a verbal description of the sound can influence the choice of the subject and potentially induce biases in the analysis conclusions.
-Indeed, the subject may select \textit{a priori} sounds referenced as belonging to \emph{park} to build a \emph{calm} soundscape.
+Most browsing tools are based on keyword indexing; however, for sensory experiments that study the objectivation of a subject's mental representations, this means is problematic: indeed, the availability of a verbal description of the sound can influence the subject's choice, and potentially induce biases in the analysis conclusions.
+For example, a subject may automatically select sounds referenced as belonging to a \emph{park} environment to build a \emph{calm} soundscape, rather than focusing on their perception.
 
@@
 
-Therefore, the selection interface considered in this study is text free and composed of spatially organized circles on a plane, each representing one leaf class.
+Therefore, the selection interface considered in this study is text free and designed so as to force the user to rely on listening.
 
@@
 
-The Figure~\ref{fig:ssf} shows the interface used for the selection of events.
-The classes of the higher level of abstraction are plotted with circles.
-The spatial location of the circles is chosen according to the hierarchical organization of the sound database: sub-classes belonging to the same class are close to each others, and so on until the user reaches the lower level classes displayed in grey which are directly linked to a collection of samples.
+Figure~\ref{fig:ssf} shows the interface used for the selection of events.
+Each circle corresponds to a sound class, the lowest level of abstraction (leaves) colored in grey.
+The spatial location of those circles is chosen according to the hierarchical organization of the sound database: sub-classes belonging to the same class are close to each others, and so on until the user reaches the leaf classes, which are directly linked to a collection of samples.
 
@@
 
-Each of those classes have a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle.
+Each of those classes has a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle.
 The subject can browse the database by listening to those prototype sounds.
@@
 As shown on Figure~\ref{fig:simscene}, the simulation interface displays a schematic of the scene under creation.
-Each track is represented as an horizontal band with a temporal axis.
-Each sample of this track is displayed by a rectangle whose height is proportional to the amplitude of the sample.
-For event's tracks, the horizontal spacing between those rectangles is a function of the time delay between their onsets.
-For tracks of texture, a unique rectangle is displayed as this kind of sounds does not allow spacing with silence.
-As actual amplitude and spacing values are drawn from random variables, each time the subject changes the value of a parameter, the location and height of the rectangles are updated to reflect the changes in the sequencing of the samples.
+Each track is represented as a horizontal strip with a temporal axis.
+Each sample of this track is displayed as a rectangle whose height is proportional to the amplitude of the sample.
+For event tracks, the horizontal spacing between those rectangles is a function of the time delay between their onsets.
+For texture tracks, a unique rectangle is displayed as this kind of sounds does not allow spacing with silence.
+As the actual amplitude and spacing values are drawn from random variables, each time the subject changes the value of a parameter, the location and height of the rectangles are updated to reflect the changes in the sequencing of the samples.
 The subject can listen to the resulting sound scene at any time.
@@
         \includegraphics[width=\linewidth]{gfx/5.pdf}
-        \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b).\gl{TODO : translate}}\label{fig:xp1_2}
+        \caption{Experimental protocol of the simulation experiment (1.a) and pleasantness evaluation experiment (1.b).\gl{TODO : translate}}\label{fig:xp1_2}
 \end{figure}
@@
 \begin{itemize}
-\item \emph{experiment 1.a (simulation)}: during this experiment, subjects are asked to simulate urban sound environments, using the simulation tool \emph{Simscene}, see Section~\ref{sec:simulator}.
-Each of them has to create two sound environments: one ideal / pleasant and the other one, non ideal / unpleasant.
+\item \emph{experiment 1.a (simulation)}: during this experiment, subjects are asked to simulate urban sound environments using \emph{Simscene} (see Section~\ref{sec:simulator}).
+Each of them has to create two sound environments: one ideal~/ pleasant, and the other non ideal~/ unpleasant.
 \item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni).
@@
 The second experimental step aims at investigating more deeply and more broadly our knowledge on the pleasantness of the simulated scenes.
-For that, a second group of subjects is asked to evaluate the pleasantness of each scene, on a semantic scale.
+For that, a second group of subjects is asked to evaluate the pleasantness of each scene produced during (1.a), on a semantic scale.
 This experiment has two goals:
 \begin{enumerate}
-\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene.
+\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene;
 \item to detect the presence of \emph{outliers} or ambiguous scenes.
 Indeed, throughout our analyses, the predefined hedonic properties of the scenes (i and ni) are used as reference.
-We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes, \ie~that the less pleasant i-scene remains above the more pleasant ni-scene.
+We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes, \ie~that the least pleasant i-scene remains more highly rated on average than the most pleasant ni-scene.
 \end{enumerate}
@@
 
-The subjects have to simulate two urban sound environments of one minute each. They follow the instructions below:
+The subjects are asked to simulate two urban sound environments of one minute each, following these instructions:
 
@@
 All the subjects start by designing the ideal environment; they read the second set of instructions at the end of the first experiment.
-subjects are completely free of their choices concerning sounds and synthesis parameters.
+Subjects are completely free of their choices concerning sounds and synthesis parameters.
 The created sound environments must nevertheless fulfill the two following constraints :
@@
 \begin{itemize}
-\item the listening point of view is the one of a fixed listener;
+\item the listening point of view is that of a fixed listener;
 \item the soundscape must be realistic, \ie~physically plausible. For instance, subjects are free to insert ten dogs in the soundscape but they cannot insert one dog barking every 10 milliseconds.
@@
 
-Each simulation process is decomposed in several steps:
+Each simulation process is decomposed into several steps:
 
 \begin{enumerate}
-\item the simulation: the user is asked to:
-\begin{itemize}
-  \item  select the class of sounds
-	\item  give a name to the selected class of sounds
-	\item  set the parameters of the tracks related to the selected class of sounds, see Section~\ref{sec:simscene_parametre}.
-\end{itemize}
-\item the feedback : writing of a free form comment about the composed soundscape
+\item Simulation, where the user is asked to:
+  \begin{itemize}
+  \item  select sound classes,
+  \item  give each of them a name,
+  \item  set the parameters of the tracks related to the selected sound classes of sounds, see Section~\ref{sec:simscene_parametre}.
+  \end{itemize}
+\item Feedback : writing of a free form comment about the composed soundscape.
 \end{enumerate}
@@
 \begin{itemize}
-\item  point out the sound sources that were missing for the composition
-\item  give a comment about the ergonomics of the simulation environment
-\item   give a comment about the ergonomics of the selection tool
+\item  point out the sound sources that were missing for the composition;
+\item  give a comment about the ergonomics of the simulation environment;
+\item  give a comment about the ergonomics of the selection tool.
 \end{itemize}
@@
 
-All the subjects performed the experiment at the same time. They are equally distributed in three identical quiet rooms. They are not allowed to talk to each other during the experiment.
+All the subjects performed the experiment at the same time. They are equally distributed in three identical quiet rooms, and are not allowed to talk to each other during the experiment.
 
@@
 
-Three experimenters (one in each room) are available during the whole duration of the experiment in order to assist subjects with potential hardware and software issues and to answer questions.
+Three experimenters (one in each room) are available during the whole duration of the experiment in order to assist subjects with potential hardware and software issues, and to answer questions.
 
@@
 
-44 students (14 females; averaging $21.6$ years of age, s.d.
-of $2.0$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment.
-All the subjects live in Nantes, France, for at least two years.
+44 students (30 male, 14 female; averaging $21.6$ years of age, s.d.
+of $2.0$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part in the experiment.
+All the subjects had been living in Nantes, France, for at least two years at the time of the experiment.
 
@@
 
-Due to temporal constraints, subjects only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from sec. 15 to sec. 45).
+Due to temporal constraints, subjects only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from second 15 to sec. 45).
 
@@
 
-10 students (2 females; averaging $23.1$ years of age, s.d.
+10 students (8 male, 2 female; averaging $23.1$ years of age, s.d.
 of $1.8$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment.
-All the subjects live in Nantes, France, for at least two years.
+All the subjects had been living in Nantes, France, for at least two years at the time of the experiment.
 None of them took part to the previous simulation experiment (experiment 1.a).
@@
 
-A set of features is attached to each sound scene.
-The analysis is done on the basis of these features.
-A summary of these features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}.
-In order to be consistent with the evaluation of experiment 1.b, features are not computed in the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b, see Section~\ref{sec:xp1b_plan}.
+A set of features, upon which the analysis is conducted, is attached to each sound scene.
+A summary of those features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}.
+In order to be consistent with the evaluation of experiment 1.b, features are not computed on the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b (Section~\ref{sec:xp1b_plan}).
 
@@
 The vector dimension ($n$) depends on the level of abstraction that is considered for the analysis.
-For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.
-
-\item \emph{sound levels}: to figure out the sound levels, we draw inspiration from the $L_{Aeq}$.
+For instance, for the abstraction level 1, that includes 44 classes of sounds ($n = 44$).
+
+\item \emph{structural features}: while \emph{SimScene} allows us to acces a variety of information about the scene structure (such as the density of events), we only focus in this first study on the sound levels.
+To figure those out, we draw inspiration from the $L_{Aeq}$.
 In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale).
@@
 
-In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we put the data to the five following significance tests:
+In order to evaluate the specific impact of the various sound sources on the perceived pleasantness, we run the data through the five following significance tests:
 
@@
 
-\item \emph{analysis of the perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and ni-scenes during the experiment 1.a.
+\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and ni-scenes during the experiment 1.a.
 To do so, we consider if there exist significant differences for the i- and ni-scenes of the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$.
-The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$;
+The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.
 
@@
 
-\item \emph{analysis of the sound levels}: the goal is to evaluate whether the sound levels differ between the i- and ni-scenes. The significance is measured with a two-sample Student test;
+\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the i- and ni-scenes. The significance is measured with a two-sample Student test.
 
@@
 
-\item \emph{influence of the sound levels on the perceived pleasantness}: the goal is to evaluate if the sound levels impact the perceived pleasantness.
+\item \emph{Influence of sound levels on perceived pleasantness}: the goal is to evaluate if the sound levels impact the perceived pleasantness.
 To do so, we consider linear correlations between these features and $\mathcal{A}_{scene}$.
-The Pearson correlation coefficient is used for that purpose;
+The Pearson correlation coefficient is used for that purpose.
 
@@
 
-\item \emph{analysis of the semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni).
+\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni).
 To do so, a V-test is considered \citep{rakotomalala2008tvpercent}.
-With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of time a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
-For each class $j$, and each environment type $k$, an approximation of the statistic value $V_{jk}$ is computed as follows:
+With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
+For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:
 
@@
 If the null hypothesis is rejected, the class $j$ is said to be typical with respect to the type of scene $k$.
-The typical classes are called \textbf{sound markers}.
-Testing is done for each class, at each level of abstraction, and separately for texture and event classes;
+Such typical classes are called \textbf{sound markers}.
+Testing is done for each class, at each level of abstraction, and separately for texture and event classes.
 
@@
 
-\item \emph{representation space induced by the semantic features}: the goal is to determine if a representation space of the scenes solely based on the presence / absence of sound sources allows us to distinguish between the two types of scene.
-Considering $S_i$ the semantic feature of the scene $i$, we compute the distances between all the vectors $S_i$.
-A Hamming distance is used.
-Considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$ and $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$, with $x \in\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between the two vectors.
+\item \emph{Representation space induced by the semantic features}: the goal is to determine if a representation space of the scenes solely based on the presence / absence of sound sources allows us to distinguish between the two types of scene.
+Denoting as $S_i$ the semantic features of scene $i$, we compute the distances between all $S_i$ vectors.
+A Hamming distance is used: considering two $n$-dimension vectors $S_1=(x_{1,1},x_{1,2},\ldots,x_{1,n})$ and $S_2=(x_{2,1},x_{2,2},\ldots,x_{2,n})$, with $x \in\lbrace 0,1\rbrace$, the Hamming distance $d_{ham}$ measures the proportion of coordinates that differ between the two vectors.
 It is defined as follows:
@@
 
-where $\bigoplus$ is the \emph{exclusive-or}.
+where $\bigoplus$ is the \emph{exclusive-or} operator.
 Two scenes having similar source compositions will be close in such space.
-The use of Hamming distance allows us to take into account equally the presence and absence of classes.
-For measuring the intrinsic ability of the space to discriminate between the i- and ni-scenes, we use a ranking metric named the precision at rank $k$ ($P@k$).
+Using the Hamming distance allows us to take into account equally the presence and absence of classes.
+In order to measure the intrinsic ability of the space to discriminate between i- and ni-scenes, we use a ranking metric named the precision at rank $k$ ($P@k$).
 The $P@k$ computes the precision obtained after the $k$ closest items with respect to a given seed item have been found.
-Formally, for each $s_i$ scene (considered as seed), we compute the ratio between the number of $s_j$ scenes taken among the $k$ nearest neighbors of $s_i$ and sharing the same label than $s_i$ versus $k$.
-the $P@k$ is then the average of this ratio for all the items considered as search seeds;
+Formally, for each $s_i$ scene (considered as seed), we compute the proportion of $s_j$ scenes in the $k$ nearest neighbors of $s_i$ that share the same label as $s_i$.
+The $P@k$ is then the average of this ratio for all the items considered as search seeds.
 
@@
 
-\item \emph{influence of the sound markers on the perceived pleasantness}: in order to assess the specific contributions of some sound sources, we again estimate the impact of the sound levels on the perceived pleasantness by taking into account only the sound markers for the computation of those features.
+\item \emph{Influence of the sound markers on the perceived pleasantness}: in order to assess the specific contributions of some sound sources, we again estimate the impact of the sound levels on the perceived pleasantness by taking into account only the sound markers for the computation of those features.
 \end{itemize}
@@
 
-The statistical tests of significance are done with a critical threshold of $\alpha=0.05$.
+All statistical significance tests are conducted with a critical threshold of $\alpha=0.05$.
 For the V-test, considering that a large number of classes is tested, a Bonferroni correction is applied.
@@
 
-\subsubsection*{Analysis of the perceived pleasantness}
+\subsubsection*{Analysis of perceived pleasantness}
 
@@
 First, in order to ensure the coherence of the data, we check that none of the ni-scenes gets a $\mathcal{A}_{scene}$ higher than one of a i-scene.
-4 ni-scenes don't fulfill that constraint: they are thus removed, together with their corresponding i-scenes.
+4 ni-scenes do not fulfill that constraint: they are thus removed, together with their corresponding i-scenes.
 As a consequence, 36 i-scenes and 36 ni-scenes remain for analysis.
@@
 
-\subsubsection*{Analysis of the sound levels}
+\subsubsection*{Analysis of sound levels}
 
@@
 
-First, our analyses focuses on the sound levels.
-Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the distribution of the levels $L$, $L(E)$ and $L(T)$.
+First, our analysis focuses on the sound levels.
+Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the distributions of levels $L$, $L(E)$ and $L(T)$.
 There is a significant difference in terms of sound levels between i- and ni-scenes ($L$: $p<0.01$).
@@
 This results is also an outcome of a large number of related works.
-We also notice that this difference of sound levels is both significant for events or textures.
+We also notice that this difference of sound levels is significant for both events and textures.
 
@@
 
-Though, the level alone is not sufficient to differentiate completely between i- and ni-scenes.
-In fact, $20\%$ of the i-scenes have a sound levels higher than the nominal level of the ni-scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
-
-\subsubsection*{Influence of the sound levels on the perceived pleasantness}
+However, level alone is not sufficient to fully differentiate between i- and ni-scenes.
+In fact, $20\%$ of the i-scenes have a sound levels higher than the lowest level of the ni-scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
+
+\subsubsection*{Influence of sound levels on the perceived pleasantness}
 
@@
 In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated.
-Contrary to the previous section, we do not consider the pleasantness property of a scene on a binary mode (i vs.
-ni): we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature.
+Contrary to the previous test, we do not limit ourselves to a binary i-scenes \emph{vs.} ni-scenes disctinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature.
 The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$.
@@
 Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived.
-Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship do not occur in the same way for i- and ni-scenes.
+Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for i- and ni-scenes.
 In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni-scenes ($r=-0.78$, $p<0.01$), but is quasi null for i-scenes.
@@
 \item allows us to differentiate between i- and ni-scenes,
-\item characterizes precisely the perceived pleasantness for ni-scenes,
+\item characterizes precisely the perceived pleasantness for ni-scenes (unsurprisingly, an unpleasant scene gets all the more unpleasant as it gets louder),
 \item is not a relevant feature for modeling the perceived pleasantness of an \textit{a priori} pleasant soundscape (i-scene).
@@
 Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (i-scenes tend to be softer than ni-scenes).
-Thus, the sound events keep a relative ability to predict the pleasantness of the ni-scenes, but the level of the textures do not bring a lot of information whatever the type of environment is.
+Thus, while sound event levels maintain a relative ability to predict the pleasantness of the ni-scenes, texture levels do not bring much information, whatever the type of environment is.
 
@@
 
-To sum up, for an unpleasant environment, sound levels - especially those of the events - negatively influence the perceived pleasantness.
+To sum up, for an unpleasant environment, sound levels --~especially those of events~-- negatively influence the perceived pleasantness.
 On the contrary, for a pleasant environment, none of the sound levels considered in the study seem to influence the perceived pleasantness.
@@
 
-Those first outcomes tend to show that 1) two modes of perception exist depending on the very nature of the environment (i- or ni-), and 2) each involves independent features.
-The fact that $L$ is not sufficient to characterize the pleasantness of the i-scenes can lead us to conclude that all the sound sources do not equally contribute to the perception of pleasantness.
-Thus, we think that only the level of some of them can influence this percept.
-In order to investigate further in that direction, we analyze in the next section the scenes on a semantic point of view, \ie we take an interest in the nature of the sources they are made of.
+Those first outcomes tend to show that 1) two modes of perception exist depending on the nature of the environment (i or ni), and 2) each involves distinct, independent features.
+The fact that $L$ is not sufficient to characterize the pleasantness of the i-scenes can lead us to conclude that all sound sources do not equally contribute to the perception of pleasantness.
+We thus put forward the hypothesis that only the level of some of them can influence this perception.
+In order to investigate further in that direction, we analyze in the next section the scenes from a semantic point of view, \ie we take an interest in the nature of the sources they are made of.
 
@@
 \begin{figure}[!tp]
-		\myfloatalign
-        \subfloat[]
-        {\includegraphics[width=1\linewidth]{gfx/xp1_class_1}\label{fig:soundsourcea}}\par
-        \subfloat[]
-        {\includegraphics[width=.5\linewidth]{gfx/xp1_class_2}\label{fig:soundsourceb}}
-        \subfloat[]
-        {\includegraphics[width=.5\linewidth]{gfx/xp1_class_3}\label{fig:soundsourcec}}
-       \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.\gl{TODO : align}}\label{fig:soundsource}
+  \centerfloat
+  \subfloat[]{\includegraphics[width=\textwidth]{gfx/xp1_class_1}\label{fig:soundsourcea}}
+
+  \subfloat[]{\includegraphics[width=.5\linewidth]{gfx/xp1_class_2}\label{fig:soundsourceb}}
+  \subfloat[]{\includegraphics[width=.5\linewidth]{gfx/xp1_class_3}\label{fig:soundsourcec}}
+                             \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.}\label{fig:soundsource}
 \end{figure}
@@
 
-We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate the environment.
+We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate a given type of environment (i or ni).
 Results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
-For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$ is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.
+For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$, is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.
 
@@
 To a certain extent, our results contradict this result.
-In fact,  Figure~\ref{fig:soundsourcea} show that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the subjects for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes.
-Those results do not state the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the i-scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes.
-Though, \emph{public transport} classes are also strongly present in the ni-scenes, for instance more than \emph{light vehicle} or \emph{trucks} classes.
+In fact,  Figure~\ref{fig:soundsourcea} shows that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the subjects for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes.
+Those results do not question the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the i-scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes.
+However, \emph{public transport} classes are also strongly present in the ni-scenes, for instance more than \emph{light vehicle} or \emph{truck} classes.
 On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.
@@
 As in our study, Guastavino asks the subjects to describe an environment.
-She asks them to perform this task starting only from their memories whereas, in our case, subjects perform the same task using actual sound samples that they can listen to.
+She asks them to perform this task starting only from their memories, whereas in our case subjects perform the same task using actual sound samples that they can listen to.
 The fact that subjects in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact.
@@
 
-We have shown that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs i- or ni-scenes.
+We have shown that, from a qualitative point of view, the composition of the scenes in terms of sound sources differs between i- or ni-scenes.
 We now investigate whether some of the sound classes are specific to a given environment.
@@
 Regarding the textures, 5 markers are identified.
-For the i-scenes, those classes related to passive or quiet ambiances (\emph{courtyard}, \emph{park}).
-The marker classes for the ni-scenes are, as for the events, related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroad}).
+For the i-scenes, those classes related to subdued or quiet ambiances (\emph{courtyard}, \emph{park}).
+The marker classes for the ni-scenes are, as for the events, related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroads}).
 
@@
 
-Although the whole set of identified markers are rather intuitive, none of the event classes related to the noise of the motor vehicles are identified as markers, except for the texture class \emph{crossroad}.
+Although the whole set of identified markers are rather intuitive, none of the event classes related to the noise of motor vehicles are identified as markers, except for the texture class \emph{crossroads}.
 To generate an unpleasant traffic, subjects chose the classes \emph{horn} or \emph{siren}.
@@
 In this section, we evaluate at which level a semantic representation of the scenes allows us to discriminate between the two types of environment.
-For this purpose, a rank 5-precision ($p@5$) is computed on the space induced by the semantic features $S$, and for each abstraction level, see Section~\ref{sec:xp1_dataAna}.
+For this purpose, a rank 5-precision ($p@5$) is computed on the space induced by the semantic features $S$, and for each abstraction level (see Section~\ref{sec:xp1_dataAna}).
 The vectors $S$ are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers ($E_{w/o,m}$).
-Texture classes corresponding to sound markers are not numerous enough to reliably compute the metric.
-They are thus not considered.
+Texture classes corresponding to sound markers are not numerous enough to reliably compute the metric, and are thus not considered.
 For the same reasons, event classes corresponding to sound markers at abstraction level $0$ are also discarded.
@@
 
-Concerning $ET$, the $p@5$ is $76\%$ for the abstraction level $0$, and remains above $86\%$ for higher abstraction levels.
+Concerning $ET$, the $p@5$ is $76\%$ at abstraction level $0$ (the most abstract), and remains above $86\%$ for higher abstraction levels.
 Considering only the presence / absence of sound classes thus allows us to nicely discriminate between the two types of environments.
-We also notice that the higher the abstraction level is, thus the more precise the description, the more effective the description is.
+We also notice that the less abstract (and therefore more precise) the description is, the more effective it is to predict agreement.
 
@@
 $S$ being a boolean vector, the smaller the dimension, the lower the amount of information it can carry.
-Despite this, it appears that the $p@5$ obtained with $E_m$ is equal – or above – to the ones obtained with $E$ or $ET$, although only a partial information is used in that case to describe the scenes.
-On the opposite, if the sound markers are not taken into account for the description ($E_{w/o,m}$), the performance is below the one achieved when considering only the textures as features.
+Despite this, it appears that the $p@5$ obtained with $E_m$ is equal – or superior – to the ones obtained with $E$ or $ET$, although only a partial information is used in that case to describe the scenes.
+Reciprocally, if the sound markers are not taken into account for the description ($E_{w/o,m}$), the performance is below the one achieved when considering only the textures as features.
 Thus, most of the semantic information allowing to differentiate between i- and ni-scenes is included in the markers.
@@
 
-The outcomes of this analysis are:
+To sum up, the outcomes of this analysis are:
 
 \begin{enumerate}
-\item unlike what we outlined with the sound levels, a semantic description of the scenes composition in terms of presence / absence of sound sources, allows us to differentiate reliably the two types of environments (i or ni);
+\item unlike what we outlined with the sound levels, a semantic description of the scenes composition in terms of presence~/ absence of sound sources allows us to reliably differentiate between the two types of environments (i or ni);
 \item the semantic information is mainly contained in the event sound classes;
@@
 
-\subsubsection*{Influence of the sound markers on the perceived pleasantness}
+\subsubsection*{Influence of sound marker levels on the perceived pleasantness}
 
@@
 In this section, the sound levels are computed by taking into account only the previously identified sound markers.
-We define $X_m$, the $X$ feature computed by taking into account only the sound markers.
-On the opposite, we define $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
-When the feature characterizes an i-scene (idem for a ni-scene),  only the markers identified for the i-scenes (or the ni-scenes) are considered; we named them, i-markers (or ni- markers).
-Results are shown in Table~\ref{tab:corrMarkers}.
+We define $X_m$, the $X$ feature computed by taking into account only the sound markers, and $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
+When the feature characterizes an i-scene (resp.
+ni-scene), only the markers identified for the i-scenes (resp.
+ni-scenes) are considered.
+We henceforth call \emph{i-markers} and \emph{ni-markers} the two types of markers.
+Results are shown on Table~\ref{tab:corrMarkers}.
 
@@
 
-Alos, whereas $L(T)$ did not show any significant correlation for ni-scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
-This indicates that the level of non-markers texture classes does not influence the perceived pleasantness in the same way for i- and ni-scenes.
+Moreover, whereas $L(T)$ did not show any significant correlation for ni-scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
+This indicates that the level of non-marker texture classes does not influence the perceived pleasantness in the same way for i- and ni-scenes.
 Sound level seems to have a negative effect for the ni-scenes, and no significant effect for the i-scenes.
@@
 Thus, for the i-scenes, it is not the absolute markers level that is important, but their relative level with respect to the other sounds composing the scene.
-A double perceptual mechanism for the ideal environments: can thus be observed:
+A double perceptual mechanism for the ideal environments can thus be observed:
 
@@
 
-\item \emph{differentiating the i- and ni-scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ et $L(T)$), allow us to differentiate reliably between i- and ni-scenes.
+\item \emph{differentiating i- and ni-scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ et $L(T)$), allow us to differentiate reliably between i- and ni-scenes.
 The semantic description seems to be more powerful;
@@
 
-\item \emph{for i-scenes}, on contrary, the sound markers characteristics and those of the other sounds have to be considered separately to predict  the pleasantness.
-The markers level, relative to the background noise ($L(E)_m-L(E)_b$ and $L_m-L_b$) is positively correlated to the pleasantness, whereas the noise level ($L_b$ and $L(E)_b$) is negatively correlated.
+\item \emph{for i-scenes}, on the contrary, the sound markers characteristics and those of the other sounds have to be considered separately to predict  the pleasantness.
+The markers level relative to the background noise ($L(E)_m-L(E)_b$ and $L_m-L_b$) is positively correlated to the pleasantness, whereas the noise level ($L_b$ and $L(E)_b$) is negatively correlated.
 
@@
 
-The fact that the pleasantness of the i-scenes is not correlated to global physical features, contrary to pleasantness of ni-scenes, has also been studied recently \cite{gozalo2015relationship}.
+The fact that the pleasantness of the i-scenes is not correlated to global physical features, contrary to the pleasantness of ni-scenes, has also been studied recently \cite{gozalo2015relationship}.
 
@@
 
+%\mr{Honnêtement, je trouve que là les conclusions vont au-delà de ce que l'expérience permet de soutenir.
+Les résultats pourraient tout aussi bien s'expliquer de manière plus simple, sans parler de modes de perception différents, en supposant que le volume sonore a toujours un impact négatif sur l'agrément, mais que si un accroissement du volume sonore est concentré sur des marqueurs positifs, l'impact positif de ceux-ci contrebalance l'impact négatif du volume.}
+
+Another hypothesis would be that the volume somehow act as an hedonic "gain" factor.
+If the volume of a negative marker is high, it lowers the overall pleasantness.
+If the volume of a positive marker is high, it highers the overall pleasantness.
+Evidently, the positive gain is expected to saturate at a given level and will quickly decreases as the level raises above a given threshold.
+
 \section{Experiment 2: modification of the semantic content}
@@
 
-The previous experiment demonstrated that, among the classes of sounds occurring in urban soundscapes, those with markers are specific to some environments.
+The previous experiment demonstrated that, among the classes of sounds occurring in urban soundscapes, those gathering markers are specific to some environments.
 Those sound markers seem to have a great impact on perception.
@@
 
-In order to investigat deeper into the relation between the pleasantness of the ideal and non ideal scenes and the sound markers, the audio waveform of the simulated scenes are regenerated with or without the classes identified as markers.
-To do so, some of the i-markers are removed for the i-scenes and some of the ni-markers are removed for the ni-scenes.
+In order to investigate deeper into the relation between the pleasantness of the ideal and non ideal scenes and the sound markers, the audio waveform of the simulated scenes are regenerated with or without the classes identified as markers.
+To do so, i-markers are removed from the i-scenes, and ni-markers are removed from the ni-scenes.
 To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptual test is conducted with a protocol close to the one considered in experiment 1.b.
@@
 
-If the first hypothesis is rather intuitive, the second is less.
-Indeed, it is not obvious that the removal of the i-markers, though perceptively positively connoted, decreases the global quality of a soundscape.
-Indeed, this removal also decreases the global sound level of the scene.
-
 %Néanmoins, comme nous l'avons vu, le niveau global n'est qu'un indicateur partiel de l'agrément pour les environnements sonores idéaux.
@@
 
-Though, as discussed before, the global amplitude level can only be considered as a partial indicator of pleasantness of the ideal soundscapes.
+If the first hypothesis is rather intuitive, the second is less.
+Indeed, it is not obvious that the removal of the i-markers, though perceptively positively connoted, will decrease the global quality of a soundscape, since this removal also decreases the global sound level of the scene.
+However, as discussed before, the global amplitude level can only be considered as a partial indicator of pleasantness of the ideal soundscapes.
 Furthermore, the level of i-markers positively impact the pleasantness.
@@
 
-Thus, only a fraction of the i-markers and ni-markers are removed in the rm-scenes.
+Thus, only part of the i-markers and ni-markers are removed from the rm-scenes.
 
@@
 The evaluation is done on a 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant).
-Before scoring a scene, subjects must listen to the first 20 seconds.
+Before rating a scene, subjects must listen to the first 20 seconds.
 After scoring, they are free to move on to the next scene.
@@
 
-All the subjects perform the test on the same type of computers. They listen to the stimuli with semi-open \emph{Beyer-Dynamic DT 990 Pro} headphones. The sound level is the same for all subjects.
-
 %Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
 
-All the subjects perform the test at the same time, in a quiet environment. They are not allowed to talk to each other during the experiment.
+All the subjects perform the test at the same time, on the same type of computers, in a quiet environment.
+They listen to the stimuli with semi-open \emph{Beyer-Dynamic DT 990 Pro} headphones, with the sound level set to be the same for all the subjects.
+They are not allowed to talk to each other during the experiment.
 
@@
 
-A supervisor is available during the whole duration of the experiment to monitor the experiment and answer to any questions the subjects may have.
+A supervisor is available during the whole duration of the experiment to monitor it and answer to any question the subjects may have.
 
@@
 
-12 subjects perform the test (4 females; averaging $29.5$ years of age, s.d.
+12 subjects perform the test (8 male, 4 female; averaging $29.5$ years of age, s.d.
 of $14$ years).
 None of them have performed experiments 1.a and 1.b.
-All the subjects live in Nantes, France, for at least two years.
+All the subjects had been living in Nantes, France, for at least two years at the time of the experiment.
 All the subjects performed the test successfully.
@@
 The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness.
-To do so, we perform an Analysis of Variance.
-We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (i/ni), and the presence/absence of markers (km/rm).
-As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to evaluate that there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes.
+To do so, we perform an variance analysis.
+We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (i/ni) and the presence/absence of markers (km/rm).
+As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to evaluate whether there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes.
 The two independent variables are considered as \emph{within-subject} factors.
@@
 
-All the significance tests are done using a critical threshold $\alpha=0.05$.
+All the significance tests are performed with critical threshold $\alpha=0.05$.
 
@@
         {\includegraphics[width=.33\linewidth]{gfx/xp4_note_15}\label{fig:xp2_Ascene}}
-        \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness  per scenes$\mathcal{A}_{scene}$ (b) versus scene type.
+        \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness per scene$\mathcal{A}_{scene}$ (b) versus scene type.
 Black stars in subfigure (a) stand for the detected outlier, \ie~subject 7.}\label{fig:xp2A}
@@
 
-Let us consider $\mathcal{A}_{subject}$ for the km-scenes, see Figure~\ref{fig:xp2_Asubject}.
-Close inspection of the answers show that one subject strongly differs from the ones of others.
-This subject evaluated positively close to half of the ni/km-scenes, see Annex~\ref{app:xp2}, subject 7.
-This subject also gave a score above 0 for 58\% of the ni/km-scenes, contrary to an average of 11\% for the remaining of the subjects.
+Let us consider $\mathcal{A}_{subject}$ for the km-scenes (Figure~\ref{fig:xp2_Asubject}).
+Close inspection of the answers show that one subject's judgments strongly differ from the others'.
+This subject evaluated positively close to half of the ni/km-scenes (see Annex~\ref{app:xp2}, subject 7), and gave a score above 0 for 58\% of the ni/km-scenes, contrary to an average of 11\% for the other subjects.
 
 Furthermore, this subject used the whole scale (from -5 to 5) to score both the i-scenes and the ni-scenes.
-Those behaviors strongly differ with the ones of the others subjects be they from this experiment or the previous ones.
+Those behaviors strongly differ with the ones of the other subjects, be they from this experiment or the previous ones.
 Subject 7 is thus discarded from the analysis.
@@
 
-In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes and ni/rm-scenes, see Figure~\ref{fig:xp2_Ascene}.
-The repeated measure ANOVA applied to $\mathcal{A}_{subject}$ show a significant effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
+In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes and ni/rm-scenes (Figure~\ref{fig:xp2_Ascene}).
+The repeated measure ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
 
@@
 
-The \emph{post hoc} analysis exhibits significant differences between all the groups of observations, notably between the i/km- and the i/rm scenes ($p<0.05$) as well as between the ni/km- and the ni/rm-scenes ($p<0.01$).
+The \emph{post hoc} analysis exhibits significant differences between all groups of observations, notably between the i/km- and the i/rm scenes ($p<0.05$) as well as between the ni/km- and the ni/rm-scenes ($p<0.01$).
 
@@
 \begin{itemize}
-\item the removal of the ni-markers improved the pleasantness of the ni-scenes;
-\item the removal of the i-markers reduced the pleasantness of the i-scenes.
+\item the removal of ni-markers improved the pleasantness of the ni-scenes;
+\item the removal of i-markers reduced the pleasantness of the i-scenes.
 \end{itemize}
@@
 
-The significant interaction shows that the effect of the type of environment influences the effect of the presence/absence of the markers.
-Indeed, the average of the differences between km-scenes and rm-scenes is larger for the ni-scenes ($1.1$) than for the i-scenes ($0.5$).
+The significant interaction shows that the type of environment influences the effect of the presence/absence of the markers.
+Indeed, the average difference between km-scenes and rm-scenes is larger for the ni-scenes ($1.1$) than for the i-scenes ($0.5$).
 
@@
 
-This experiment shows that the presence of markers identified during the analysis of experiment 1 does have an impact on the perceived pleasantness.
+This experiment shows that the presence of the markers identified during the analysis of experiment 1 does have an impact on the perceived pleasantness.
 The removal of the ni-markers has a positive effect on the perception of the ni-scenes.
-Perhaps more surprisingly the removal of the i-markers slightly decreases the perception of the i-scenes.
-This is even more striking because, due to the removal of the markers, the acoustic pressure level of the i/km-scenes is higher than the one of the i/rm-scenes.
+Perhaps more surprisingly the removal of the i-markers slightly decreases the perception of the i-scenes: this is a more striking observation since, due to the removal of the markers, the acoustic pressure level of the i/km-scenes is higher than the one of the i/rm-scenes.
 
@@
 
-The i-markers do have a positive impact on the perception of an environment.
+This strongly confirms that i-markers do have a positive impact on the perception of an environment.
 The fact that their removal decreases $\mathcal{A}_{scene}$ indicate that it should possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
@@
 
-This series of three experiments showed that most of the descriptors used in this study, be they semantic or based on acousitc level, allows us to distinguish between an ideal scene and a non ideal one.
+This series of three experiments showed that most of the descriptors used in this study, be they of a semantic or acoustic nature, allow us to distinguish between an ideal scene and a non ideal one.
 
@@
 That being said, we observe that the physical characteristics correlated with the perceived pleasantness clearly differ depending on the type of scenes.
-In the case of ideal scenes, it is above all the emergence of sound markers that determines the perceived quality, whereas in the case of non-ideal scenes, it is the overall sound level that prominently influences the perceived quality.
+In the case of ideal scenes, it is above all the emergence of sound markers that determines the perceived quality, whereas in the case of non-ideal scenes, it is the overall sound level that prominently influences it.
 
@@
 
-These results show that the perception of the qualities of a scene indeed depends primarily on its sound sources which are identified.
+These results show that the perception of the qualities of a scene indeed depends primarily on its identifiable sound sources.
 The characteristics that are taken into account during the perceptual process appear to vary from one source to the other, from one type of environment to the other.
@@
 
-Finally, these works allow us to conjecture as to the nature of the mental representations of the concepts "pleasant (urban sound) environment" (PE) and "unpleasant (urban sound) environment" (UE).
+Finally, this work allows us to conjecture as to the nature of the mental representations of the concepts "pleasant (urban sound) environment" (PE) and "unpleasant (urban sound) environment" (UE).
 
@@
 
-Secondly, the fact that the removal of sound markers changes the perceived pleasantness leads us to believe that the abstract concept related to pleasantness depends on the activation of a network of concepts strongly linked to the sources which are in the case of this study: \emph{bird}, \emph{church bell} and \emph{bicycle bell}.
+Second, the fact that the removal of sound markers changes the perceived pleasantness leads us to believe that the abstract concept related to pleasantness depends on the activation of a network of concepts strongly linked to the sources which are in the case of this study: \emph{bird}, \emph{church bell} and \emph{bicycle bell}.
 
@@
 
-It would also be interesting to use the alteration capabilities of the proposed paradigm in order to study further the effects caused by the voluntary modification of a characteristic of a scene, as during the suppression of the sound markers practiced in experiment 2.
+There are also many more avenues of research to fully explore the capabilities of the proposed paradigm; first by taking into account a wider range of structural features (for example the density or regularity of events), and second by studying further the effects caused by the voluntary modification of scene composition, as during the suppression of the sound markers practiced in experiment 2.
 
@@
 Finally, one should study the influence of socio-cultural contexts on perception.
-Indeed, if the sound of the church bell is most often a quality environment marker for a Westerner.
-This do not necessarily hold true for subjects of Eastern, Middle Eastern or other cultures.
+Indeed, if the sound of the church bell is most often a quality environment marker for a Westerner, this does not necessarily hold true for subjects of Eastern, Middle Eastern or other cultures.
 
@@
         {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_12}\label{fig:xp4_note_1l}}
-        \caption{Distribution of scores given par the subjects during experiment 2 for i/am-scenes (green) et ni/am-scenes (red).
+        \caption{Distribution of scores given by the subjects during experiment 2 for i/am-scenes (green) et ni/am-scenes (red).
 The horizontal axis is, from left to right, the 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant).}\label{fig:xp4_note_1}

### 2018-01-19 lafaySimulatedScenePerception.tex
plus
incorporating jfp comments

@@
 We propose in this paper to consider the use of a soundscape simulator that the subject can employ to objectify his/her representation of a given sound environment.
-We believe that the use of such a device allows us to gain the benefits of the two abovementioned approaches.
+We believe that the use of such a device allows us to gain the benefits of the two above mentioned approaches.
 As the subject is asked to produce audio data (the signal of the simulated scene), it allows the experimenter to study a precise modal version of the subject's mental representation that is characterized both semantically (the nature of sound sources) and physically (\eg~the levels of sound sources).
@@
 
-In order to account for the possiblity that the morphological differences between those two classes may have some important consequences on the perception of the scenes, the simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with classes of events only and the other with classes of textures only.
+In order to account for the possibility that the morphological differences between those two classes may have some important consequences on the perception of the scenes, the simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with classes of events only and the other with classes of textures only.
 These two types of sound classes also have specific scheduling procedures during the simulation process.
@@
 Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds.
-Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}.
+Four \ml{Three ?} levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}.
 Only three levels of abstraction are considered for the texture sounds.
@@
 
-All the subjects perform the experiment on standard desktop computers with the same hardware and software configurations.
-The audio files are played in diotic conditions using headphones.
-During the tutorial, subjects are asked to adjust the sound volume to a comfortable level.
-Once set, they are not allowed to modify it during the remaining of the experiment.
+All the subjects performed the experiment on standard desktop computers with the same hardware and software configurations.
+The audio files were played in diotic conditions using headphones.
+During the tutorial, subjects were asked to adjust the sound volume to a comfortable level.
+Once set, they were not allowed to modify it during the remaining of the experiment.
 
@@
 
-All the subjects performed the experiment at the same time. They are equally distributed in three identical quiet rooms, and are not allowed to talk to each other during the experiment.
+All the subjects performed the experiment at the same time. They were equally distributed in three identical quiet rooms, and are not allowed to talk to each other during the experiment.
 
@@
 
-Three experimenters (one in each room) are available during the whole duration of the experiment in order to assist subjects with potential hardware and software issues, and to answer questions.
-
-\subsubsection*{subjects}
+Three experimenters (one in each room) were available during the whole duration of the experiment in order to assist subjects with potential hardware and software issues, and to answer questions.
+
+\subsubsection*{Subjects}
 
@@
 Only the last evaluations are taken into account.
+Each participant evaluates all the sound scenes.
 
@@
 The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}.
-The stimuli are the scenes obtained by the simulation experiment of experiment 1.a.
+The stimuli are the scenes obtained in experiment 1.a.
 The output sound level is the same for all the subjects.
@@
 
-\subsubsection*{subjects}
+\subsubsection*{Subjects}
 
@@
 10 students (8 male, 2 female; averaging $23.1$ years of age, s.d.
-of $1.8$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment.
+of $1.8$ years) from \emph{Ecole Centrale de Nantes} took part to the experiment.
 All the subjects had been living in Nantes, France, for at least two years at the time of the experiment.
@@
 Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores.
+$\mathcal{A}_{subject}$ is computed for i-scenes and ni-scenes.
 
@@
 The vector dimension ($n$) depends on the level of abstraction that is considered for the analysis.
-For instance, for the abstraction level 1, that includes 44 classes of sounds ($n = 44$).
+For instance, for the abstraction level 1, that includes 44 classes of sounds, the dimension is thus $44$ ($n = 44$).
 
@@
 The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.
+
+\jfp{je ne comprends pas, pour moi il faut faire, pour chaque sujet de l'expé 1, un test de différence entre Aniscène  et Aiscène
+a quoi sert Asubject ici ?}
 
@@
 It indeed appears that the i-scenes were perceived significantly more pleasant ($p < 0.01$) than the ni-scenes.
+
+\jfp{a la limite on pourrait ne pas écarter arbitrairement de scène et ne garder que celles pour lesquelles les iscenes sont signif.
++ plaisante que les ni-scene.
+ca revient au même.
+c'est les test qui les écarte.
+il faudrait peut être justifier pourquoi on fait cela  : pour ne pas fausser les résultats de l'analyse et vérifier que les i scènes sont effectivement perçues + plaisantes que les ni.}
 
@@
 The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$.
-The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown on Table~\ref{tab:corrStructA}.
+The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown in Table~\ref{tab:corrStructA}.
 Relationships between $\mathcal{A}_{scene}$ and the sound levels are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.
@@
 Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for i- and ni-scenes.
-In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni-scenes ($r=-0.78$, $p<0.01$), but is quasi null for i-scenes.
+In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni-scenes ($r=-0.78$, $p<0.01$), but is not significant ($p=0.06$) for i-scenes.
 
@@
   \subfloat[]{\includegraphics[width=.5\linewidth]{gfx/xp1_class_3}\label{fig:soundsourcec}}
-                             \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.}\label{fig:soundsource}
+                             \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.
+\jfp{pourquoi la classe "start passage way", présente en figure A9, n'est pas présente sur ce graphe ?}}\label{fig:soundsource}
 \end{figure}
@@
 We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate a given type of environment (i or ni).
-Results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
+For the 36 i and ni-scenes considered, results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
 For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$, is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.
@@
 \vspace{0.5mm}
-\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked as decreasing order of V-test value. $p\leq0.01$ for all sound markers.}
+\caption{Event and texture classes identified as sound markers.
+In each cell, markers are ranked as decreasing order of V-test value \ml{ shown between parenthesis}.
+$p\leq0.01$ for all sound markers.}
 \label{tab:markers}
@@
   \caption{$P@5$ obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors as a function of the abstraction level.
-The vectors are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers $E_{w/o,m}$.}\label{fig:pa5}
+The vectors are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers $E_{w/o,m}$.
+Baseline results are achieved by considering random vectors as input.}\label{fig:pa5}
 \end{figure}
@@
 This indicates that the level of non-marker texture classes does not influence the perceived pleasantness in the same way for i- and ni-scenes.
-Sound level seems to have a negative effect for the ni-scenes, and no significant effect for the i-scenes.
+Sound level of textures seems to have a negative effect for the ni-scenes, and no significant effect for the i-scenes.
 
@@
 
-On contrary, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ for the ni-scenes, shows that this is indeed the absolute level that matters.
+On the contrary, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ for the ni-scenes, shows that this is indeed the absolute level that matters.
 
@@
 
-%\mr{Honnêtement, je trouve que là les conclusions vont au-delà de ce que l'expérience permet de soutenir.
+\mr{Honnêtement, je trouve que là les conclusions vont au-delà de ce que l'expérience permet de soutenir.
 Les résultats pourraient tout aussi bien s'expliquer de manière plus simple, sans parler de modes de perception différents, en supposant que le volume sonore a toujours un impact négatif sur l'agrément, mais que si un accroissement du volume sonore est concentré sur des marqueurs positifs, l'impact positif de ceux-ci contrebalance l'impact négatif du volume.}
 
-Another hypothesis would be that the volume somehow act as an hedonic "gain" factor.
+\ml{Ma traduction : Another hypothesis would be that the volume somehow act as an hedonic "gain" factor.
 If the volume of a negative marker is high, it lowers the overall pleasantness.
 If the volume of a positive marker is high, it highers the overall pleasantness.
-Evidently, the positive gain is expected to saturate at a given level and will quickly decreases as the level raises above a given threshold.
+Evidently, the positive gain is expected to saturate at a given level and will quickly decreases as the level raises above a given threshold.}
 
@@
 The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness.
-To do so, we perform an variance analysis.
+To do so, we perform a variance analysis.
 We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (i/ni) and the presence/absence of markers (km/rm).
@@
 \emph{Post hoc} analyses are done using the Tukey-Kramer procedure.
+\jfp{l'analyse est inutile a mon avis, il n'y a que 2 niveaux pour chaque facteur de l'ANOVA.}
 
@@
         \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness per scene$\mathcal{A}_{scene}$ (b) versus scene type.
-Black stars in subfigure (a) stand for the detected outlier, \ie~subject 7.}\label{fig:xp2A}
+Black stars in subfigure (a) stand for the detected outlier, \ie~subject 7.
+\ml{maj figure: sujet->subject, et changer am et sm pour km et rm}}\label{fig:xp2A}
 \end{figure}
@@
 
+\jfp{je crois que l'analyse post-hoc n'est pas nécessaire lorsqu'il n'y a que 2 niveaux pour les facteurs de l'ANOVA.
+les conclusions seront strictement les mêmes que pour l'ANOVA.c'est utile à partir de 3 niveaux, pour savoir quelle moyenne diffère signif.
+des autres.}
+
 The \emph{post hoc} analysis exhibits significant differences between all groups of observations, notably between the i/km- and the i/rm scenes ($p<0.05$) as well as between the ni/km- and the ni/rm-scenes ($p<0.01$).
@@
 This strongly confirms that i-markers do have a positive impact on the perception of an environment.
-The fact that their removal decreases $\mathcal{A}_{scene}$ indicate that it should possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
+The fact that their removal decreases $\mathcal{A}_{scene}$ indicate that it should be possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
 Those conclusions are in line with the positive approach introduced by Schafer in \cite{schafer1977tuning}.
@@
 \item The simulator can be deployed on a large scale via the Internet thanks to the web-based software architecture;
-\item simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
+\item Simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
 \end{itemize}

### 2018-01-29 lafaySimulatedScenePerception.tex
Réponses aux différents commentaires

@@
 Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds.
-Four \ml{Three ?} levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}.
+Four \ml{Three ?} \gl{non 4} levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}.
 Only three levels of abstraction are considered for the texture sounds.
@@
 
-All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie relative to Full Scale. In our case, the full scale level is set arbitrarily to 1 Volt.
+All samples are normalized to the same $RMS$ level of $-12$ $dB$ FS, \ie~relative to Full Scale. In our case, the full scale level is set arbitrarily to 1 Volt.
 
@@
 
-\jfp{je ne comprends pas, pour moi il faut faire, pour chaque sujet de l'expé 1, un test de différence entre Aniscène  et Aiscène
-a quoi sert Asubject ici ?}
+\jfp{je ne comprends pas, pour moi il faut faire, pour chaque sujet de l'expé 1, un test de différence entre Aniscène  et Aiscène a quoi sert Asubject ici ?} \gl{Les deux sont possibles : Ascene mesure des différences structurelles, Asubject des différences perceptives}
 
@@
 c'est les test qui les écarte.
-il faudrait peut être justifier pourquoi on fait cela  : pour ne pas fausser les résultats de l'analyse et vérifier que les i scènes sont effectivement perçues + plaisantes que les ni.}
+il faudrait peut être justifier pourquoi on fait cela  : pour ne pas fausser les résultats de l'analyse et vérifier que les i scènes sont effectivement perçues + plaisantes que les ni.} \gl{Oui.
+Mais ça revient au même de toute façon}
 
@@
 The study shows that there is no significant difference between global and averaged instantaneous judgments.
-In our case, the result can be interpreted as if the subjects had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie the events.
+In our case, the result can be interpreted as if the subjects had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie~the events.
 
@@
                              \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.
-\jfp{pourquoi la classe "start passage way", présente en figure A9, n'est pas présente sur ce graphe ?}}\label{fig:soundsource}
+\jfp{pourquoi la classe "start passage way", présente en figure A9, n'est pas présente sur ce graphe ?} \gl{Elle n'est pas parlante}}\label{fig:soundsource}
 \end{figure}
@@
 Those results are in line with \cite{maffiolo_caracterisation_1999}.
-As discussed in Section~\ref{sec:simscene_sampleDataSet}, it seems that humans analyze the event scenes which are composed of several sound events in a causal manner, \ie by identifying the sources.
+As discussed in Section~\ref{sec:simscene_sampleDataSet}, it seems that humans analyze the event scenes which are composed of several sound events in a causal manner, \ie~by identifying the sources.
 
@@
 If the volume of a positive marker is high, it highers the overall pleasantness.
-Evidently, the positive gain is expected to saturate at a given level and will quickly decreases as the level raises above a given threshold.}
+Evidently, the positive gain is expected to saturate at a given level and will quickly decreases as the level raises above a given threshold.} \gl{@mathieu, pourquoi pas.
+Mais je pense qu'on peut laisser également le premier paragraphe, ça permettra aux reviewer de commenter qqch :).
+Quid de mettre les deux.}
 
@@
 \emph{Post hoc} analyses are done using the Tukey-Kramer procedure.
-\jfp{l'analyse est inutile a mon avis, il n'y a que 2 niveaux pour chaque facteur de l'ANOVA.}
+\jfp{l'analyse est inutile a mon avis, il n'y a que 2 niveaux pour chaque facteur de l'ANOVA.} \gl{2 niveaux par facteur, mais 4 niveaux en tout}
 

### 2018-01-30 lafaySimulatedScenePerception.tex
removing comments

@@
 
-\newcommand{\gl}[1]{\textcolor{magenta}{GL : #1}}
-\newcommand{\mr}[1]{\textcolor{green}{MR : #1}}
-\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
-\newcommand{\jfp}[1]{\textcolor{blue}{JFP : #1}}
+%\newcommand{\gl}[1]{\textcolor{magenta}{GL : #1}}
+%\newcommand{\mr}[1]{\textcolor{green}{MR : #1}}
+%\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
+%\newcommand{\jfp}[1]{\textcolor{blue}{JFP : #1}}
 \newcommand{\ie}{\emph{i.\,e.}}
@@
         \includegraphics[width=.8\linewidth]{gfx/3.pdf}
-       \caption{Hierarchical organisation of the isolated sounds used in the simulation.\gl{TODO : translate and remove subject box}}\label{fig:orgDb}
+       \caption{Hierarchical organisation of the isolated sounds used in the simulation.}\label{fig:orgDb}
 \end{figure}
@@
 Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds.
-Four \ml{Three ?} \gl{non 4} levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}.
+Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}.
 Only three levels of abstraction are considered for the texture sounds.
@@
         \includegraphics[width=\linewidth]{gfx/5.pdf}
-        \caption{Experimental protocol of the simulation experiment (1.a) and pleasantness evaluation experiment (1.b).\gl{TODO : translate}}\label{fig:xp1_2}
+        \caption{Experimental protocol of the simulation experiment (1.a) and pleasantness evaluation experiment (1.b).}\label{fig:xp1_2}
 \end{figure}
@@
 
-\jfp{je ne comprends pas, pour moi il faut faire, pour chaque sujet de l'expé 1, un test de différence entre Aniscène  et Aiscène a quoi sert Asubject ici ?} \gl{Les deux sont possibles : Ascene mesure des différences structurelles, Asubject des différences perceptives}
+%\jfp{je ne comprends pas, pour moi il faut faire, pour chaque sujet de l'expé 1, un test de différence entre Aniscène  et Aiscène a quoi sert Asubject ici ?} \gl{Les deux sont possibles : Ascene mesure des différences structurelles, Asubject des différences perceptives}
 
@@
 
-\jfp{a la limite on pourrait ne pas écarter arbitrairement de scène et ne garder que celles pour lesquelles les iscenes sont signif.
+%\jfp{a la limite on pourrait ne pas écarter arbitrairement de scène et ne garder que celles pour lesquelles les iscenes sont signif.
 + plaisante que les ni-scene.
@@
                              \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.
-\jfp{pourquoi la classe "start passage way", présente en figure A9, n'est pas présente sur ce graphe ?} \gl{Elle n'est pas parlante}}\label{fig:soundsource}
+}\label{fig:soundsource}
 \end{figure}
+
+%\jfp{pourquoi la classe "start passage way", présente en figure A9, n'est pas présente sur ce graphe ?} \gl{Elle n'est pas parlante}
 
@@
 \vspace{0.5mm}
-\caption{Event and texture classes identified as sound markers.
-In each cell, markers are ranked as decreasing order of V-test value \ml{ shown between parenthesis}.
-$p\leq0.01$ for all sound markers.}
+\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked as decreasing order of V-test value, shown between parenthesis. $p\leq0.01$ for all sound markers.}
 \label{tab:markers}
@@
 
-\mr{Honnêtement, je trouve que là les conclusions vont au-delà de ce que l'expérience permet de soutenir.
+%\mr{Honnêtement, je trouve que là les conclusions vont au-delà de ce que l'expérience permet de soutenir.
 Les résultats pourraient tout aussi bien s'expliquer de manière plus simple, sans parler de modes de perception différents, en supposant que le volume sonore a toujours un impact négatif sur l'agrément, mais que si un accroissement du volume sonore est concentré sur des marqueurs positifs, l'impact positif de ceux-ci contrebalance l'impact négatif du volume.}
 
-\ml{Ma traduction : Another hypothesis would be that the volume somehow act as an hedonic "gain" factor.
+Another hypothesis would be that the volume somehow act as an hedonic "gain" factor.
 If the volume of a negative marker is high, it lowers the overall pleasantness.
 If the volume of a positive marker is high, it highers the overall pleasantness.
-Evidently, the positive gain is expected to saturate at a given level and will quickly decreases as the level raises above a given threshold.} \gl{@mathieu, pourquoi pas.
-Mais je pense qu'on peut laisser également le premier paragraphe, ça permettra aux reviewer de commenter qqch :).
-Quid de mettre les deux.}
+Evidently, the positive gain is expected to saturate at a given level and will quickly decreases as the level raises above a given threshold.
+
+%\gl{@mathieu, pourquoi pas. Mais je pense qu'on peut laisser également le premier paragraphe, ça permettra aux reviewer de commenter qqch :). Quid de mettre les deux.}
 
@@
 \emph{Post hoc} analyses are done using the Tukey-Kramer procedure.
-\jfp{l'analyse est inutile a mon avis, il n'y a que 2 niveaux pour chaque facteur de l'ANOVA.} \gl{2 niveaux par facteur, mais 4 niveaux en tout}
+%\jfp{l'analyse est inutile a mon avis, il n'y a que 2 niveaux pour chaque facteur de l'ANOVA.} \gl{2 niveaux par facteur, mais 4 niveaux en tout}
 
@@
         \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness per scene$\mathcal{A}_{scene}$ (b) versus scene type.
-Black stars in subfigure (a) stand for the detected outlier, \ie~subject 7.
-\ml{maj figure: sujet->subject, et changer am et sm pour km et rm}}\label{fig:xp2A}
+Black stars in subfigure (a) stand for the detected outlier, \ie~subject 7.}\label{fig:xp2A}
 \end{figure}
+
+% {maj figure: changer am et sm pour km et rm}
 
@@
 
-\jfp{je crois que l'analyse post-hoc n'est pas nécessaire lorsqu'il n'y a que 2 niveaux pour les facteurs de l'ANOVA.
+%\jfp{je crois que l'analyse post-hoc n'est pas nécessaire lorsqu'il n'y a que 2 niveaux pour les facteurs de l'ANOVA.
 les conclusions seront strictement les mêmes que pour l'ANOVA.c'est utile à partir de 3 niveaux, pour savoir quelle moyenne diffère signif.

### 2018-01-31 lafaySimulatedScenePerception.tex
dot in title
typos

@@
 
-    \address{$^{1)}$ Laboratoire des Sciences du Numérique, Nantes, France.\\
+    \address{$^{1)}$ Laboratoire des Sciences du Numérique de Nantes-CNRS-\'Ecole Centrale de Nantes, Nantes, France.\\
     \hspace*{8pt}mathieu.lagrange@cnrs.fr\\
@@
 
-    \title{Investigating soundscapes perception through acoustic scenes simulation.}
+    \title{Investigating soundscapes perception through acoustic scenes simulation}
 
@@
         {\includegraphics[width=.33\linewidth]{gfx/xp4_note_15}\label{fig:xp2_Ascene}}
-        \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness per scene$\mathcal{A}_{scene}$ (b) versus scene type.
+        \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness per scene $\mathcal{A}_{scene}$ (b) versus scene type.
 Black stars in subfigure (a) stand for the detected outlier, \ie~subject 7.}\label{fig:xp2A}

### 2018-01-31 lafaySimulatedScenePerception.tex
fig size

@@
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp4_note_13}\label{fig:xp2_Asubject}}
+        {\includegraphics[width=.48\linewidth]{gfx/xp4_note_13}\label{fig:xp2_Asubject}}
         \subfloat[]
-        {\includegraphics[width=.33\linewidth]{gfx/xp4_note_15}\label{fig:xp2_Ascene}}
+        {\includegraphics[width=.48\linewidth]{gfx/xp4_note_15}\label{fig:xp2_Ascene}}
         \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness per scene $\mathcal{A}_{scene}$ (b) versus scene type.

### 2018-06-21 v2/responseToReviewers.tex
response

\documentclass[10pt]{article}

\title{Reply to reviewer concerning submission """}

\begin{document}

\maketitle

\section{Answer to Reviewer 1}

\begin{enumerate}

\item \emph{The utility of this tool beyond the experiments presented here is not apparent because the flexibility of the system was not demonstrated, for example:
- SimScene itself was only used in the first experiment, and other applications of it were not considered. Design choices vary widely between researchers and domains, meaning that the specific instantiations here may not be generally useful.
- It is not clear how this tool would be expanded to consider other acoustic parameters of interest, beyond level. It might be useful to consider potential relationships to other soundscape softwares such as TAPESTREA, and how the exploration of other acoustic parameters could be made intuitive for experiment participants.}

$\rightarrow$

\item \emph{In the last sentence of the abstract, the distinction of 'physical descriptors' vs. 'global descriptors'  does not have enough context here to be informative.}

$\rightarrow$

\item \emph{In section 2.1 (pg. 5), the citations (particularly 27-31) do not seem relevant for justifying the event/texture distinction in the simulator.}

$\rightarrow$

\end{enumerate}

\end{document}

### 2018-06-21 v1/lafaySimulatedScenePerception.tex
response

--- 2018-01-31 lafaySimulatedScenePerception.tex
@@
 the 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.
+The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available \url{http://soundthings.org/research/urbanSoundScape/XP2014}.
+The resulting waveforms are available for download \url{https://archive.org/details/soundSimulatedUrbanScene}.
 

### 2018-06-26 v2/responseToReviewers.tex
adressing editor comments

@@
 
-\title{Reply to reviewer concerning submission """}
+\title{Reply to reviewers concerning submission BR-Org-18-034 "Investigating soundscapes perception through acoustic scenes simulation"}
 
@@
 \maketitle
+
+As a preamble, we would like to thank the editor and the reviewer for their comments and suggestions.
+Following these comments, we made several changes to the article, which are summarized here.
+The next sections list our answers to each of the reviewer’s comments, with references to the revised manuscript (page, column, and paragraph) where appropriate.
+
+\section{Answer to Academic Editor}
+
+\begin{enumerate}
+
+\item \emph{please follow APA (American Psychological Association) style guidelines (w.r.t. in-text citations, etc.).}
+
+$\rightarrow$ The manuscript now follows the APA citation style. The text have been modified accordingly.
+
+\item \emph{ Please avoid the over-use of acronyms as it makes reading of your article unnecessarily difficult (examples: P@5, ni, i, etc.).
+This sort of lab jargon is not transparent to the wider readership of BRM.
+}
+
+$\rightarrow$ Those abbreviations are now in full text, except for some figures for the sake of clarity, in which case the acronyms are detailed in the caption.
+
+\item \emph{I would also like to see comparisons between your approach and other state-of-the-art soundscape analyses (e.g. TAPESTREA, see comment from the reviewer below).}
+
+$\rightarrow$ TODO
+
+\item \emph{You might consider providing a link to a github site with examples of your stimuli. It is hard to imagine what these sound scenes sound like without some real examples.}
+
+The resulting waveforms are available for download \url{https://archive.org/details/soundSimulatedUrbanScene}.
+The link is now in the introduction.
+The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available \url{http://soundthings.org/research/urbanSoundScape/XP2014}.
+
+$\rightarrow$
+
+\end{enumerate}
 
@@
 
+\item \emph{In the last sentence of the abstract, the distinction of 'physical descriptors' vs. 'global descriptors'  does not have enough context here to be informative.}
+
+$\rightarrow$
+
+\item \emph{In section 2.1 (pg. 5), the citations (particularly 27-31) do not seem relevant for justifying the event/texture distinction in the simulator.}
+
+$\rightarrow$
+
 \end{enumerate}

### 2018-06-26 lafaySimulatedScenePerception.tex
adressing editor comments

@@
 \usepackage{color}
+%\usepackage[natbibapa]{apacite}
 
@@
 \begin{enumerate}
-\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non ideal soundscapes that are considered as material for the following experiments;
+\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non ideal soundscapes that are considered as material for the following experiments; The resulting waveforms are available for download \url{https://archive.org/details/soundSimulatedUrbanScene}.
 \item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
@@
 \begin{enumerate}
-\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (i or ni) thanks to a finer quantification of the pleasantness of the scene;
+\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (ideal or non ideal) thanks to a finer quantification of the pleasantness of the scene;
 \item to detect the presence of \emph{outliers} or ambiguous scenes.
-Indeed, throughout our analyses, the predefined hedonic properties of the scenes (i and ni) are used as reference.
-We thus need to ensure beforehand that no ambiguity exists between extreme cases of i and ni scenes, \ie~that the least pleasant i-scene remains more highly rated on average than the most pleasant ni-scene.
+Indeed, throughout our analyses, the predefined hedonic properties of the scenes (ideal or not) are used as reference.
+We thus need to ensure beforehand that no ambiguity exists between extreme cases of ideal and non ideal scenes, \ie~that the least pleasant ideal scene remains more highly rated on average than the most pleasant non ideal scene.
 \end{enumerate}
@@
 the 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.
+The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available \url{http://soundthings.org/research/urbanSoundScape/XP2014}.
 
@@
 Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
-                                   &                            \\
-ideal/pleasant                     & i                          \\
-non-ideal/unpleasant               & ni                         \\
-scene ideal/pleasant               & i-scene                    \\
-scene non-ideal/unpleasant         & ni-scene                   \\
+%                                    &                            \\
+% ideal/pleasant                     & i                          \\
+% non-ideal/unpleasant               & ni                         \\
+% scene ideal/pleasant               & i-scene                    \\
+% scene non-ideal/unpleasant         & ni-scene                   \\
 \hline
@@
 Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores.
-$\mathcal{A}_{subject}$ is computed for i-scenes and ni-scenes.
+$\mathcal{A}_{subject}$ is computed for ideal scenes and non ideal scenes.
 
@@
 
-\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and ni-scenes during the experiment 1.a.
-To do so, we consider if there exist significant differences for the i- and ni-scenes of the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$.
+\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and non ideal scenes during the experiment 1.a.
+To do so, we consider if there exist significant differences for the i- and non ideal scenes of the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$.
 The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.
@@
 
-\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the i- and ni-scenes. The significance is measured with a two-sample Student test.
+\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the i- and non ideal scenes. The significance is measured with a two-sample Student test.
 
@@
 Using the Hamming distance allows us to take into account equally the presence and absence of classes.
-In order to measure the intrinsic ability of the space to discriminate between i- and ni-scenes, we use a ranking metric named the precision at rank $k$ ($P@k$).
+In order to measure the intrinsic ability of the space to discriminate between i- and non ideal scenes, we use a ranking metric named the precision at rank $k$ ($P@k$).
 The $P@k$ computes the precision obtained after the $k$ closest items with respect to a given seed item have been found.
@@
 
-First, in order to ensure the coherence of the data, we check that none of the ni-scenes gets a $\mathcal{A}_{scene}$ higher than one of a i-scene.
-4 ni-scenes do not fulfill that constraint: they are thus removed, together with their corresponding i-scenes.
-As a consequence, 36 i-scenes and 36 ni-scenes remain for analysis.
-Second, we verify that subjects really perceived a difference in terms of pleasantness between i- and ni-scenes.
+First, in order to ensure the coherence of the data, we check that none of the non ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of a ideal scene.
+4 non ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes.
+As a consequence, 36 ideal scenes and 36 non ideal scenes remain for analysis.
+Second, we verify that subjects really perceived a difference in terms of pleasantness between i- and non ideal scenes.
 For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment.
-It indeed appears that the i-scenes were perceived significantly more pleasant ($p < 0.01$) than the ni-scenes.
+It indeed appears that the ideal scenes were perceived significantly more pleasant ($p < 0.01$) than the non ideal scenes.
 
@@
         {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_6}\label{fig:soundlevelf}}
-       \caption{Distributions of the sound levels $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
+       \caption{Distributions of the sound levels $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (i: ideal, ni: non ideal) (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
 \end{figure}
@@
 Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the distributions of levels $L$, $L(E)$ and $L(T)$.
-There is a significant difference in terms of sound levels between i- and ni-scenes ($L$: $p<0.01$).
+There is a significant difference in terms of sound levels between ideal and non ideal scenes ($L$: $p<0.01$).
 This difference is significant for events ($L(E)$: $p<0.01$, mean deviation: -7 $dB$) and for textures ($L(T)$: $p<0.01$, mean deviation: -6 $dB$).
@@
 
-As expected, the sound level of the sources is indeed a pleasantness indicator, as the ni-scenes tend to be louder.
+As expected, the sound level of the sources is indeed a pleasantness indicator, as the non ideal scenes tend to be louder.
 This results is also an outcome of a large number of related works.
@@
 
-It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for i- and ni-scenes.
+It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for ideal and non ideal scenes.
 This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}.
@@
 
-However, level alone is not sufficient to fully differentiate between i- and ni-scenes.
-In fact, $20\%$ of the i-scenes have a sound levels higher than the lowest level of the ni-scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
+However, level alone is not sufficient to fully differentiate between ideal and non ideal scenes.
+In fact, $20\%$ of the ideal scenes have a sound levels higher than the lowest level of the non ideal scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
 
@@
 \begin{tabular}{l c c c}
-               & all scenes                & i-scenes          & ni-scenes    \\
+               & all scenes                & ideal scenes          & non ideal scenes    \\
 \hline
@@
 In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated.
-Contrary to the previous test, we do not limit ourselves to a binary i-scenes \emph{vs.} ni-scenes disctinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature.
+Contrary to the previous test, we do not limit ourselves to a binary ideal scenes \emph{vs.} non ideal scenes disctinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature.
 The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$.
@@
 Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived.
-Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for i- and ni-scenes.
-In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for ni-scenes ($r=-0.78$, $p<0.01$), but is not significant ($p=0.06$) for i-scenes.
+Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for ideal and non ideal scenes.
+In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for non ideal scenes ($r=-0.78$, $p<0.01$), but is not significant ($p=0.06$) for ideal scenes.
 
@@
 
-When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the i-scenes tend to be softer than the ni-scenes, thus allowing us to extend artificially to the i-scenes the negative correlation observed for the ni-scenes.
+When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the ideal scenes tend to be softer than the non ideal scenes, thus allowing us to extend artificially to the ideal scenes the negative correlation observed for the non ideal scenes.
 
@@
 \begin{itemize}
-\item allows us to differentiate between i- and ni-scenes,
-\item characterizes precisely the perceived pleasantness for ni-scenes (unsurprisingly, an unpleasant scene gets all the more unpleasant as it gets louder),
-\item is not a relevant feature for modeling the perceived pleasantness of an \textit{a priori} pleasant soundscape (i-scene).
+\item allows us to differentiate between i- and non ideal scenes,
+\item characterizes precisely the perceived pleasantness for non ideal scenes (unsurprisingly, an unpleasant scene gets all the more unpleasant as it gets louder),
+\item is not a relevant feature for modeling the perceived pleasantness of an \textit{a priori} pleasant soundscape (ideal scene).
 \end{itemize}
@@
 The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}.
-For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (i-scenes: $r=-0.33$, $p=0.05$, ni-scenes: $r=-0.00$, $p=0.99$).
-Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (i-scenes tend to be softer than ni-scenes).
-Thus, while sound event levels maintain a relative ability to predict the pleasantness of the ni-scenes, texture levels do not bring much information, whatever the type of environment is.
+For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (ideal scenes: $r=-0.33$, $p=0.05$, non ideal scenes: $r=-0.00$, $p=0.99$).
+Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (ideal scenes tend to be softer than non ideal scenes).
+Thus, while sound event levels maintain a relative ability to predict the pleasantness of the non ideal scenes, texture levels do not bring much information, whatever the type of environment is.
 
@@
 Those first outcomes tend to show that 1) two modes of perception exist depending on the nature of the environment (i or ni), and 2) each involves distinct, independent features.
-The fact that $L$ is not sufficient to characterize the pleasantness of the i-scenes can lead us to conclude that all sound sources do not equally contribute to the perception of pleasantness.
+The fact that $L$ is not sufficient to characterize the pleasantness of the ideal scenes can lead us to conclude that all sound sources do not equally contribute to the perception of pleasantness.
 We thus put forward the hypothesis that only the level of some of them can influence this perception.
@@
 We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate a given type of environment (i or ni).
-For the 36 i and ni-scenes considered, results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
+For the 36 i and non ideal scenes considered, results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
 For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$, is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.
@@
 
-We observe a noticable difference in terms of class choices between the i- and ni-scenes.
-The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~ on one hand, classes involving human presence and nature are prevailing in the i-scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the ni-scenes.
+We observe a noticeable difference in terms of class choices between the i- and non ideal scenes.
+The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~ on one hand, classes involving human presence and nature are prevailing in the ideal scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the non ideal scenes.
 These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.
@@
 To a certain extent, our results contradict this result.
-In fact,  Figure~\ref{fig:soundsourcea} shows that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the subjects for $28\%$ of the i-scenes, and for $42\%$ of the ni-scenes.
-Those results do not question the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the i-scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes.
-However, \emph{public transport} classes are also strongly present in the ni-scenes, for instance more than \emph{light vehicle} or \emph{truck} classes.
+In fact,  Figure~\ref{fig:soundsourcea} shows that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the subjects for $28\%$ of the ideal scenes, and for $42\%$ of the non ideal scenes.
+Those results do not question the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the ideal scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes.
+However, \emph{public transport} classes are also strongly present in the non ideal scenes, for instance more than \emph{light vehicle} or \emph{truck} classes.
 On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.
@@
 Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
-level & i-scenes & ni-scenes \\
+level & ideal scenes & non ideal scenes \\
 \hline
@@
   & \multicolumn{2}{c}{Texture sound markers}      \\
-  & i-scenes & ni-scenes \\
+  & ideal scenes & non ideal scenes \\
 \hline
@@
 
-We have shown that, from a qualitative point of view, the composition of the scenes in terms of sound sources differs between i- or ni-scenes.
+We have shown that, from a qualitative point of view, the composition of the scenes in terms of sound sources differs between i- or non ideal scenes.
 We now investigate whether some of the sound classes are specific to a given environment.
@@
 Regarding the sound events, 9 markers are identified for all abstraction levels.
-As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are i-scenes markers as well as the \emph{church bell} class.
+As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are ideal scenes markers as well as the \emph{church bell} class.
 This latter result may be due to the socio-cultural background of the subjects who are mostly European citizens.
 In fact, according to Schafer, a sound that is identified by a person as being an important element of his/her environment, is well accepted.
-Sound markers of ni-scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
+Sound markers of non ideal scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
 
@@
 Regarding the textures, 5 markers are identified.
-For the i-scenes, those classes related to subdued or quiet ambiances (\emph{courtyard}, \emph{park}).
-The marker classes for the ni-scenes are, as for the events, related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroads}).
+For the ideal scenes, those classes related to subdued or quiet ambiances (\emph{courtyard}, \emph{park}).
+The marker classes for the non ideal scenes are, as for the events, related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroads}).
 
@@
   \includegraphics[width=\linewidth]{gfx/pa5_1_en}
-  \caption{$P@5$ obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors as a function of the abstraction level.
+  \caption{Rank 5-precision ($P@5$) obtained by considering the dissimilarity matrix computed from the paired Hamming distances of the semantic features vectors as a function of the abstraction level.
 The vectors are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers $E_{w/o,m}$.
@@
 In this section, we evaluate at which level a semantic representation of the scenes allows us to discriminate between the two types of environment.
-For this purpose, a rank 5-precision ($p@5$) is computed on the space induced by the semantic features $S$, and for each abstraction level (see Section~\ref{sec:xp1_dataAna}).
+For this purpose, a rank 5-precision is computed on the space induced by the semantic features $S$, and for each abstraction level (see Section~\ref{sec:xp1_dataAna}).
 The vectors $S$ are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers ($E_{w/o,m}$).
@@
 
-Concerning $ET$, the $p@5$ is $76\%$ at abstraction level $0$ (the most abstract), and remains above $86\%$ for higher abstraction levels.
+Concerning $ET$, the rank 5-precision is $76\%$ at abstraction level $0$ (the most abstract), and remains above $86\%$ for higher abstraction levels.
 Considering only the presence / absence of sound classes thus allows us to nicely discriminate between the two types of environments.
@@
 
-Considering $E$ and $T$ separately, it appears that: 1) the $p@5$ obtained with $E$ is similar to the one obtained with $ET$ ; 2) the $p@5$ obtained with $T$ is always lower than the one obtained with $E$, by $10\%$ to $15\%$.
+Considering $E$ and $T$ separately, it appears that: 1) the rank 5-precision obtained with $E$ is similar to the one obtained with $ET$ ; 2) the precision obtained with $T$ is always lower than the one obtained with $E$, by $10\%$ to $15\%$.
 Those results indicate that the semantic information that is discriminative is mostly carried by the events.
@@
 $S$ being a boolean vector, the smaller the dimension, the lower the amount of information it can carry.
-Despite this, it appears that the $p@5$ obtained with $E_m$ is equal – or superior – to the ones obtained with $E$ or $ET$, although only a partial information is used in that case to describe the scenes.
+Despite this, it appears that the rank 5-precision obtained with $E_m$ is equal – or superior – to the ones obtained with $E$ or $ET$, although only a partial information is used in that case to describe the scenes.
 Reciprocally, if the sound markers are not taken into account for the description ($E_{w/o,m}$), the performance is below the one achieved when considering only the textures as features.
-Thus, most of the semantic information allowing to differentiate between i- and ni-scenes is included in the markers.
+Thus, most of the semantic information allowing to differentiate between ideal and non ideal scenes is included in the markers.
 
@@
 \item the semantic information is mainly contained in the event sound classes;
-\item only a part of the event classes, i.e. the sound markers, are useful to differentiate between the i- and ni-scenes.
+\item only a part of the event classes, i.e. the sound markers, are useful to differentiate between the ideal and non ideal scenes.
 \end{enumerate}
@@
 
-Since we have extracted the typical classes of the i- and ni-scenes and verified that the distinction between them was largely dependent on the presence of these classes, we shall now investigate whether a description of the scenes only based on the sound pressure level of sound markers could characterize the perceived pleasantness, perhaps better than a globally computed sound level.
+Since we have extracted the typical classes of the ideal and non ideal scenes and verified that the distinction between them was largely dependent on the presence of these classes, we shall now investigate whether a description of the scenes only based on the sound pressure level of sound markers could characterize the perceived pleasantness, perhaps better than a globally computed sound level.
 
@@
 \begin{tabular}{l r r}
-                  &   i-scenes                  & ni-scenes \\
+                  &   ideal scenes                  & non ideal scenes \\
 \hline
@@
         {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
-        \caption{Distribution of the relative sound levels related to the presence of markers $L_m$ (a, d), $L_b$ (b, e) and $L_m-L_b$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
+        \caption{Distribution of the relative sound levels related to the presence of markers $L_m$ (a, d), $L_b$ (b, e) and $L_m-L_b$ (c, f), versus scene type (i: ideal, ni: non ideal) (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
 \end{figure}
@@
 We define $X_m$, the $X$ feature computed by taking into account only the sound markers, and $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
-When the feature characterizes an i-scene (resp.
-ni-scene), only the markers identified for the i-scenes (resp.
-ni-scenes) are considered.
-We henceforth call \emph{i-markers} and \emph{ni-markers} the two types of markers.
+When the feature characterizes an ideal scene (resp.
+non ideal scene), only the markers identified for the ideal scenes (resp.
+non ideal scenes) are considered.
+We henceforth call \emph{ideal markers} and \emph{non ideal markers} the two types of markers.
 Results are shown on Table~\ref{tab:corrMarkers}.
@@
 \begin{enumerate}
-\item a significant difference between levels of i- and ni-scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
+\item a significant difference between levels of ideal and non ideal scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
 \item the sound level of scenes is mainly related to the sound events, compared to the textures;
-\item the sound level of events has an influence on the perception of pleasantness for ni-scenes, but not for i-scenes;
+\item the sound level of events has an influence on the perception of pleasantness for non ideal scenes, but not for ideal scenes;
 \item the sound level of textures does not play any role in the perception of the pleasantness.
@@
 
-To conclude, the level of ni- markers has a negative influence on the pleasantness for the ni-scenes.
-On the other hand, the level of i- markers does not influence the perceived pleasantness for the i-scenes.
+To conclude, the level of ni- markers has a negative influence on the pleasantness for the non ideal scenes.
+On the other hand, the level of i- markers does not influence the perceived pleasantness for the ideal scenes.
 
@@
 
-Considering the non markers classes, we can observe on the i-scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}.
-It is the first time that an objective feature allows us to define the pleasantness of i-scenes.
+Considering the non markers classes, we can observe on the ideal scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}.
+It is the first time that an objective feature allows us to define the pleasantness of ideal scenes.
 This leads us to conclude that the level of non-typical sound classes of a pleasant environment has a negative influence on the pleasantness.
@@
 
-Moreover, whereas $L(T)$ did not show any significant correlation for ni-scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
-This indicates that the level of non-marker texture classes does not influence the perceived pleasantness in the same way for i- and ni-scenes.
-Sound level of textures seems to have a negative effect for the ni-scenes, and no significant effect for the i-scenes.
+Moreover, whereas $L(T)$ did not show any significant correlation for non ideal scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
+This indicates that the level of non-marker texture classes does not influence the perceived pleasantness in the same way for i- and non ideal scenes.
+Sound level of textures seems to have a negative effect for the non ideal scenes, and no significant effect for the ideal scenes.
 
@@
 
-For the i-scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
-for the ni-scenes no correlation is observed.
-Thus, for the i-scenes, it is not the absolute markers level that is important, but their relative level with respect to the other sounds composing the scene.
+For the ideal scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
+for the non ideal scenes no correlation is observed.
+Thus, for the ideal scenes, it is not the absolute markers level that is important, but their relative level with respect to the other sounds composing the scene.
 A double perceptual mechanism for the ideal environments can thus be observed:
@@
 \begin{itemize}
-\item the higher the absolute level of sounds not being i-markers, the weaker the pleasantness,
-\item the higher the relative level of i-markers is, with respect to the remaining sounds, the higher the pleasantness.
+\item the higher the absolute level of sounds not being ideal markers, the weaker the pleasantness,
+\item the higher the relative level of ideal markers is, with respect to the remaining sounds, the higher the pleasantness.
 \end{itemize}
@@
 
-On the contrary, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ for the ni-scenes, shows that this is indeed the absolute level that matters.
+On the contrary, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ for the non ideal scenes, shows that this is indeed the absolute level that matters.
 
@@
 
-\item \emph{differentiating i- and ni-scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ et $L(T)$), allow us to differentiate reliably between i- and ni-scenes.
+\item \emph{differentiating i- and non ideal scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ et $L(T)$), allow us to differentiate reliably between ideal and non ideal scenes.
 The semantic description seems to be more powerful;
@@
 \item \emph{pleasantness prediction}: considering the correlation between sound levels and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its very nature (i or ni).
-From the data gathered in those experiments, the same set of features cannot be considered to predict the pleasantness of i- and ni-scenes:
+From the data gathered in those experiments, the same set of features cannot be considered to predict the pleasantness of ideal and non ideal scenes:
 
@@
 
-\item \emph{for ni-scenes}, the global levels ($L$ et $L(E)$), or the level of sound markers ($L_m$ et $L(E)_{m}$), have a negative influence on pleasantness.
+\item \emph{for non ideal scenes}, the global levels ($L$ et $L(E)$), or the level of sound markers ($L_m$ et $L(E)_{m}$), have a negative influence on pleasantness.
 Taking into account the contribution of each of the different sources of the scene does not improve the prediction performance compared to a global analysis of the environment.
@@
 
-\item \emph{for i-scenes}, on the contrary, the sound markers characteristics and those of the other sounds have to be considered separately to predict  the pleasantness.
+\item \emph{for ideal scenes}, on the contrary, the sound markers characteristics and those of the other sounds have to be considered separately to predict  the pleasantness.
 The markers level relative to the background noise ($L(E)_m-L(E)_b$ and $L_m-L_b$) is positively correlated to the pleasantness, whereas the noise level ($L_b$ and $L(E)_b$) is negatively correlated.
@@
 
-The fact that the pleasantness of the i-scenes is not correlated to global physical features, contrary to the pleasantness of ni-scenes, has also been studied recently \cite{gozalo2015relationship}.
+The fact that the pleasantness of the ideal scenes is not correlated to global physical features, contrary to the pleasantness of non ideal scenes, has also been studied recently \cite{gozalo2015relationship}.
 
@@
 In order to investigate deeper into the relation between the pleasantness of the ideal and non ideal scenes and the sound markers, the audio waveform of the simulated scenes are regenerated with or without the classes identified as markers.
-To do so, i-markers are removed from the i-scenes, and ni-markers are removed from the ni-scenes.
+To do so, ideal markers are removed from the ideal scenes, and non ideal markers are removed from the non ideal scenes.
 To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptual test is conducted with a protocol close to the one considered in experiment 1.b.
@@
 \begin{itemize}
-\item \emph{for the ni-scenes}, we hypothesize that the absence of ni-markers will \textbf{increase} the pleasantness score;
-\item \emph{for the i-scenes}, we hypothesize that the absence of the i-markers will \textbf{decrease} the pleasantness score.
+\item \emph{for the non ideal scenes}, we hypothesize that the absence of non ideal markers will \textbf{increase} the pleasantness score;
+\item \emph{for the ideal scenes}, we hypothesize that the absence of the ideal markers will \textbf{decrease} the pleasantness score.
 \end{itemize}
@@
 If the first hypothesis is rather intuitive, the second is less.
-Indeed, it is not obvious that the removal of the i-markers, though perceptively positively connoted, will decrease the global quality of a soundscape, since this removal also decreases the global sound level of the scene.
+Indeed, it is not obvious that the removal of the ideal markers, though perceptively positively connoted, will decrease the global quality of a soundscape, since this removal also decreases the global sound level of the scene.
 However, as discussed before, the global amplitude level can only be considered as a partial indicator of pleasantness of the ideal soundscapes.
-Furthermore, the level of i-markers positively impact the pleasantness.
+Furthermore, the level of ideal markers positively impact the pleasantness.
 For those reasons, the validation of the second hypothesis is of high interest.
@@
 \begin{itemize}
-\item \emph{72 km-scenes}: the 72 scenes originally simulated by the subject of experiment 1.a where the sound classes identified as markers are kept.
-The 36 ideal scenes with markers are noted i/km-scenes, and the 36 non ideal scenes with markers are noted ni/km-scenes (km for kept markers).
-\item \emph{72 rm-scenes}:  72 scenes where the sound classes identified as markers are removed.
-The 36 ideal scenes without markers are noted i/rm-scenes, and the 36 non ideal scenes without markers are noted ni/rm-scenes (r for removed markers).
+\item \emph{72 scenes with markers}: the 72 scenes originally simulated by the subject of experiment 1.a where the sound classes identified as markers are kept.
+\item \emph{72 scenes without markers}:  72 scenes where the sound classes identified as markers are removed.
 \end{itemize}
@@
 \begin{itemize}
-\item \emph{church bell}, \emph{bicycle bell}, and \emph{animals} for the i/rm-scenes;
-\item \emph{siren}, \emph{car horn} for the ni/rm-scenes.
+\item \emph{church bell}, \emph{bicycle bell}, and \emph{animals} for the ideal scenes without markers;
+\item \emph{siren}, \emph{car horn} for the non ideal scenes without markers.
 \end{itemize}
@@
 
-Thus, only part of the i-markers and ni-markers are removed from the rm-scenes.
+Thus, only part of the ideal markers and non ideal markers are removed from the rm-scenes.
 
@@
 The first 10 scenes allow the subject to calibrate their scores.
-Those calibration scenes are  5 i/km-scenes and 5 ni/km-scenes.
+Those calibration scenes are  5 ideal scenes with markers and 5 nideal scenes with markers.
 These first 10 scenes are replayed at the end of the experiment, and only the scores given at the second occurrence are taken into account.
@@
 
-The aim here is to validate the hypothesis that the removal of i-markers and ni-markers have an impact on the perceived pleasantness.
+The aim here is to validate the hypothesis that the removal of ideal markers and non ideal markers have an impact on the perceived pleasantness.
 To do so, we perform a variance analysis.
@@
         {\includegraphics[width=.48\linewidth]{gfx/xp4_note_15}\label{fig:xp2_Ascene}}
-        \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness per scene $\mathcal{A}_{scene}$ (b) versus scene type.
+        \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness per scene $\mathcal{A}_{scene}$ (b) versus scene type (i: ideal, ni: non ideal, km: with markers, rm: without markers).
 Black stars in subfigure (a) stand for the detected outlier, \ie~subject 7.}\label{fig:xp2A}
@@
 Close inspection of the answers show that one subject's judgments strongly differ from the others'.
-This subject evaluated positively close to half of the ni/km-scenes (see Annex~\ref{app:xp2}, subject 7), and gave a score above 0 for 58\% of the ni/km-scenes, contrary to an average of 11\% for the other subjects.
-
-Furthermore, this subject used the whole scale (from -5 to 5) to score both the i-scenes and the ni-scenes.
+This subject evaluated positively close to half of the non ideal scenes with markers (see Annex~\ref{app:xp2}, subject 7), and gave a score above 0 for 58\% of the nideal scenes with markers, contrary to an average of 11\% for the other subjects.
+
+Furthermore, this subject used the whole scale (from -5 to 5) to score both the ideal scenes and the non ideal scenes.
 Those behaviors strongly differ with the ones of the other subjects, be they from this experiment or the previous ones.
@@
 
-In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes and ni/rm-scenes (Figure~\ref{fig:xp2_Ascene}).
+In this section, we study the scores given by the subject while listening to the several types of scenes, namely the ideal scenes with markers, nideal scenes with markers, ideal scenes without markers and non ideal scenes without markers (Figure~\ref{fig:xp2_Ascene}).
 The repeated measure ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
@@
 
-The \emph{post hoc} analysis exhibits significant differences between all groups of observations, notably between the i/km- and the i/rm scenes ($p<0.05$) as well as between the ni/km- and the ni/rm-scenes ($p<0.01$).
+The \emph{post hoc} analysis exhibits significant differences between all groups of observations, notably between the i/km- and the i/rm scenes ($p<0.05$) as well as between the ni/km- and the non ideal scenes without markers ($p<0.01$).
 
@@
 \begin{itemize}
-\item the removal of ni-markers improved the pleasantness of the ni-scenes;
-\item the removal of i-markers reduced the pleasantness of the i-scenes.
+\item the removal of non ideal markers improved the pleasantness of the non ideal scenes;
+\item the removal of ideal markers reduced the pleasantness of the ideal scenes.
 \end{itemize}
@@
 The significant interaction shows that the type of environment influences the effect of the presence/absence of the markers.
-Indeed, the average difference between km-scenes and rm-scenes is larger for the ni-scenes ($1.1$) than for the i-scenes ($0.5$).
+Indeed, the average difference between km-scenes and rm-scenes is larger for the non ideal scenes ($1.1$) than for the ideal scenes ($0.5$).
 
@@
 This experiment shows that the presence of the markers identified during the analysis of experiment 1 does have an impact on the perceived pleasantness.
-The removal of the ni-markers has a positive effect on the perception of the ni-scenes.
-Perhaps more surprisingly the removal of the i-markers slightly decreases the perception of the i-scenes: this is a more striking observation since, due to the removal of the markers, the acoustic pressure level of the i/km-scenes is higher than the one of the i/rm-scenes.
+The removal of the non ideal markers has a positive effect on the perception of the non ideal scenes.
+Perhaps more surprisingly the removal of the ideal markers slightly decreases the perception of the ideal scenes: this is a more striking observation since, due to the removal of the markers, the acoustic pressure level of the ideal scenes with markers is higher than the one of the ideal scenes without markers.
 
@@
 
-This strongly confirms that i-markers do have a positive impact on the perception of an environment.
+This strongly confirms that ideal markers do have a positive impact on the perception of an environment.
 The fact that their removal decreases $\mathcal{A}_{scene}$ indicate that it should be possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
@@
 
-\bibliographystyle{elsarticle-num}
+\bibliographystyle{model5-names}
+\biboptions{authoryear}
 \bibliography{Bibliography}

### 2018-06-26 v2/responseToReviewers.tex
adressing editor comments

@@
 
-$\rightarrow$ Those abbreviations are now in full text, except for some figures for the sake of clarity, in which case the acronyms are detailed in the caption.
+$\rightarrow$ Those abbreviations are now in full text, except for some figures for the sake of clarity, in which case the acronyms used are detailed in the caption.
 
@@
 
-$\rightarrow$ TODO
+$\rightarrow$ The work of Misra \& al is indeed very relevant. A new Section (number 6) is now dedicated to this comparison with state-of-the-art approaches.
 
@@
 
-The resulting waveforms are available for download \url{https://archive.org/details/soundSimulatedUrbanScene}.
+$\rightarrow$ The resulting waveforms are available for download \url{https://archive.org/details/soundSimulatedUrbanScene}.
 The link is now in the introduction.
 The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available \url{http://soundthings.org/research/urbanSoundScape/XP2014}.
-
-$\rightarrow$
 
@@
 - SimScene itself was only used in the first experiment, and other applications of it were not considered.
-Design choices vary widely between researchers and domains, meaning that the specific instantiations here may not be generally useful.
-- It is not clear how this tool would be expanded to consider other acoustic parameters of interest, beyond level.
-It might be useful to consider potential relationships to other soundscape softwares such as TAPESTREA, and how the exploration of other acoustic parameters could be made intuitive for experiment participants.}
+Design choices vary widely between researchers and domains, meaning that the specific instantiations here may not be generally useful.}
 
@@
 
+\item \emph{It is not clear how this tool would be expanded to consider other acoustic parameters of interest, beyond level.
+It might be useful to consider potential relationships to other soundscape softwares such as TAPESTREA, and how the exploration of other acoustic parameters could be made intuitive for experiment participants.}
+
+$\rightarrow$ The work of Misra \& al is indeed very relevant. A new Section (number 6) is now dedicated to this comparison with state-of-the-art approaches.
+
 \item \emph{In the last sentence of the abstract, the distinction of 'physical descriptors' vs. 'global descriptors'  does not have enough context here to be informative.}
 
-$\rightarrow$
+$\rightarrow$ This sentence have be rewritten to be more explicit: "acoustic pressure levels computed for specific sound sources better characterize the appraisal than the acoustic pressure level computed over the overall soundscape."
 
@@
 
-$\rightarrow$
-
-\item \emph{In the last sentence of the abstract, the distinction of 'physical descriptors' vs. 'global descriptors'  does not have enough context here to be informative.}
-
-$\rightarrow$
-
-\item \emph{In section 2.1 (pg. 5), the citations (particularly 27-31) do not seem relevant for justifying the event/texture distinction in the simulator.}
-
-$\rightarrow$
+$\rightarrow$ The
 

### 2018-06-26 lafaySimulatedScenePerception.tex
adressing editor comments

@@
 A rating experiment of the soundscape pleasantness using a 7-point bipolar semantic scale is conducted to further refine the analysis of the simulated urban acoustic scenes.
-Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.
+Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) acoustic pressure levels computed for specific sound sources better characterize the appraisal than the acoustic pressure level computed over the overall soundscape.
   \end{abstract}
@@
 
-\subsection{Discussion}
+%\subsection{Discussion}
 
@@
 %Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics that allow us to distinguish between two instances of the same abstract concept linked to the pleasantness are not the same for PE and UE.
+
+\section{Related computational approaches}
+
+Considering sound scene synthesis or composition as part of an experimental protocol requires the availability of software resources that are easier and easier to master.
+While most processing framework that have been considered are developed in native language for efficiency purposes like Marsyas \cite{tzanetakis2000marsyas} or Clam \cite{amatriain2006clam}, many advanced sound manipulation.
+
+Closer to our sound model is the Tapestrea framework \cite{misra2007musical} that allows the manipulation sounds trough wavelet and sinusoidal modeling to allow sound manipulations for doing acousmatic composition or sound
+design.
+The main issue with considering this framework for soundscape perception studies is that the sounds are processed using sound models that would strongly decrease the ecological validity of the study.
+That being said, the sequencing model is quite close to the one considered in this study.
+
+The experiments described here does not introduce yet another sound manipulation framework and is thus not designed to be extensible. Rather it is considered here to demonstrate the following:
+\begin{itemize}
+  \item the use of Webaudio javascript library is useful to design advanced experimental protocols that consider sound manipulation over the web
+  \item the soundscape model considered is useful to question important research questions within the soundscape community  but simple enough to be grasped by novice participants.
+\end{itemize}
+
+Extensions to other sound parameters beyond acoustic levels and .
+However, the control of those new parameters would need a specific user experience study to maintain a high degree of usability of the controlling interface.
 

### 2018-06-27 lafaySimulatedScenePerception.tex
done with reformating citations

@@
 
-One of the main goals of soundscape studies is to identify which components of the soundscape influence human perception \cite{aletta2016soundscape}.
+As asserted by~\cite{aletta2016soundscape}, one of the main goals of soundscape studies is to identify which components of the soundscape influence human perception.
 Establishing a link between an environment (urban sound scene) and the induced human sensation (calmness) is equivalent to instantiating the mental representation of a specific sound scene (a calm urban sound scene).
@@
 
-First, a subject can be asked to assess a given sound following a given perceptual scale (\eg~calmness, pleasantness) \cite{axelsson2005soundscape,davies2013perception,cain2013development}.
+First, a subject can be asked to assess a given sound following a given perceptual scale (\eg~calmness, pleasantness), see~\cite{axelsson2005soundscape,davies2013perception,cain2013development}.
 The amount of information that can be gathered strongly depends on the nature of the available stimuli, be they recorded sound scenes as part of a within laboratory experiment, or \emph{in situ}.
@@
 
-Second, a subject can be asked to describe a given sound environment \cite{guastavino2006ideal, dubois2006cognitive}.
+Second, a subject can be asked to describe a given sound environment, see~\cite{guastavino2006ideal, dubois2006cognitive}.
 A large amount of quantitative and semantic information is then collected about the subject's representation of this type of sound environment.
@@
 
-We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene \cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,szeremeta2009analysis}.
-Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene \cite{gozalo2015relationship,ricciardi2015sound}.
+We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene, see~\cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,szeremeta2009analysis}.
+Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene, as found by~\cite{gozalo2015relationship,ricciardi2015sound}.
 
@@
 
-To the best of our knowledge, only Bruce~\cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception.
+To the best of our knowledge, only~\cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception.
 They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location.
@@
 It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{Web Audio} standard that allows the manipulation of digital audio data within the browser\footnote{\emph{Web Audio}: \url{www.w3.org/TR/webaudio}}.
-The interface for selecting sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} \cite{d32011} visualization library.
+The interface for selecting sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} visualization library proposed by~\cite{d32011}.
 
@@
 
-The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} that study how humans discriminate different kind of urban soundscapes.
+The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts written by~\cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} that study how humans discriminate different kind of urban soundscapes.
 
@@
 \item {sound events}: isolated sounds, precisely located in time, whose acoustical characteristics may change with respect to time;
-\item {sound textures}: isolated sounds of long duration whose acoustical characteristics are stable with respect to time \cite{saint1995classification}.
+\item {sound textures}: isolated sounds of long duration whose acoustical characteristics are stable with respect to time, as expressed by~\cite{saint1995classification}.
 \end{itemize}
@@
 
-According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment \cite{bregman1994auditory}.
+According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment~\cite{bregman1994auditory}.
 Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}, while events that cannot be isolated are grouped into a same stream.
 
-%Maffiolo \cite{maffiolo_caracterisation_1999} met en évidence deux processus de catégorisation distincts, engagés en fonction de la capacité de l'auditeur à identifier des événements sonores.
+%Maffiolo~\cite{maffiolo_caracterisation_1999} met en évidence deux processus de catégorisation distincts, engagés en fonction de la capacité de l'auditeur à identifier des événements sonores.
 Elle montre l'existence de deux catégories cognitives abstraites d'environnements sonores respectivement appelées: «~les séquences événementielles~» et «~les séquences amorphes~».
 
-Maffiolo \cite{maffiolo_caracterisation_1999} distinguishes two separated categorization processes, either of which is triggered depending on the listener's ability to identify sound events.
+Maffiolo~\cite{maffiolo_caracterisation_1999} distinguishes two separated categorization processes, either of which is triggered depending on the listener's ability to identify sound events.
 She shows that those processes lead to two abstract cognitive categories respectively termed "event sequences" and "amorphous sequences".
@@
 Elles profitent, elles, d'une analyse holistique, effectuée à partir d'indicateurs acoustiques (objectifs) globaux.
-%Cette différence de traitement entre scènes événementielles et scènes amorphes est relevée également dans \cite{guastavino2006ideal,dubois2006cognitive}.
+%Cette différence de traitement entre scènes événementielles et scènes amorphes est relevée également dans~\cite{guastavino2006ideal,dubois2006cognitive}.
 
@@
 
-Concerning sound textures, \ie~sounds that have stable characteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can opt for an abstract, statistical representation of the perceived information, discarding precise physical properties of the sound \cite{nelken2013ear}.
+Concerning sound textures, \ie~sounds that have stable characteristics over time,~\cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can opt for an abstract, statistical representation of the perceived information, discarding precise physical properties of the sound~\cite{nelken2013ear}.
 
@@
 
-In the light of those results, it appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli \cite{nelken2013ear,mcdermott2013summary}.
-We thus consider in this study the soundscape as a ``\,skeleton of events on a bed of textures\,'' as coined in \cite{nelken2013ear}.
+In the light of those results, it appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli~\cite{nelken2013ear,mcdermott2013summary}.
+We thus consider in this study the soundscape as a ``\,skeleton of events on a bed of textures\,'' as coined in~\cite{nelken2013ear}.
 
@@
 
-The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure similar to the vertical axis of the categorical organization proposed by E.
-Rosch \cite{rosch1978cognition}.
+The sound classes are organized hierarchically (\cf~Figure~\ref{fig:orgDb}) according to a structure similar to the vertical axis of the categorical organization proposed by~\cite{rosch1978cognition}.
 The lower the level of abstraction, the more precise the description of the class and the more perceptually similar the sound sources.
@@
 Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds.
-Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in \cite{Salamon14}.
+Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used by~\cite{Salamon14}.
 Only three levels of abstraction are considered for the texture sounds.
@@
 The subject can browse the database by listening to those prototype sounds.
-The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed in \cite{lafay2016JAES}.
+The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed by~\cite{lafay2016JAES}.
 
@@
 %La scène sonore est ainsi vue comme une somme de sources sonores, ou, si l'on admet la distinction opérée entre événements et textures sonores, «~un squelette d'événements sur un lit de textures~» \cite{nelken2013ear}.
-Cet outil de simulation est présenté en détail dans \cite{rossignol2015simscene}.
-
-As such, the underlying model of the scene is a sum of sound sources. Further details about the simulation interface can be found in \cite{rossignol2015simscene}.
+Cet outil de simulation est présenté en détail dans~\cite{rossignol2015simscene}.
+
+As such, the underlying model of the scene is a sum of sound sources. The simulation interface is more toroughly described by~\cite{rossignol2015simscene}.
 
@@
 
-The design of this experiment has been validated with a pilot study described in \cite{lafay2014new}.
+The design of this experiment has been validated with a pilot study described by~\cite{lafay2014new}.
 
@@
 \item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni).
-To do so, a V-test is considered \citep{rakotomalala2008tvpercent}.
+To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}.
 With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
@@
 It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for ideal and non ideal scenes.
-This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}.
+This observation is in agreement with the results obtained by~\cite{kuwano_memory_2003}.
 During their experiment, the authors ask their subjects to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source.
@@
 We observe a noticeable difference in terms of class choices between the i- and non ideal scenes.
-The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes \cite{guastavino2006ideal}, \ie~ on one hand, classes involving human presence and nature are prevailing in the ideal scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the non ideal scenes.
-These results confirm a previously observed fact: the semantic nature of the sound sources play an important role in the assessment of the environment \cite{raimbault2005urban,dubois2006cognitive}.
+The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes by~\cite{guastavino2006ideal}, \ie~ on one hand, classes involving human presence and nature are prevailing in the ideal scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the non ideal scenes.
+These results confirm a fact previously observed by~\cite{raimbault2005urban,dubois2006cognitive}: the semantic nature of the sound sources play an important role in the assessment of the environment.
 
@@
 
-Nevertheless, we notice some differences with \cite{guastavino2006ideal}: Guastavino’s results show that sounds of \emph{public transportation} are specific of ideal urban soundscapes.
+Nevertheless, we notice some differences with results of~\cite{guastavino2006ideal} which show that sounds of \emph{public transportation} are specific of ideal urban soundscapes.
 The authors interpret this by the fact that the perception of pleasantness is, among other things, due to socio-cultural factors.
@@
 The fact that subjects in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact.
-Other studies that considered sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment \cite{lavandier2006contribution}.
+Other studies that considered sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment, as studied by~\cite{lavandier2006contribution}.
 
@@
 Those results indicate that the semantic information that is discriminative is mostly carried by the events.
-Those results are in line with \cite{maffiolo_caracterisation_1999}.
+Those results are in line with results of~\cite{maffiolo_caracterisation_1999}.
 As discussed in Section~\ref{sec:simscene_sampleDataSet}, it seems that humans analyze the event scenes which are composed of several sound events in a causal manner, \ie~by identifying the sources.
@@
 
-The fact that the pleasantness of the ideal scenes is not correlated to global physical features, contrary to the pleasantness of non ideal scenes, has also been studied recently \cite{gozalo2015relationship}.
+The fact that the pleasantness of the ideal scenes is not correlated to global physical features, contrary to the pleasantness of non ideal scenes, has also been studied recently by~\cite{gozalo2015relationship}.
 
@@
 The fact that their removal decreases $\mathcal{A}_{scene}$ indicate that it should be possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
-Those conclusions are in line with the positive approach introduced by Schafer in \cite{schafer1977tuning}.
+Those conclusions are in line with the positive approach introduced by~\cite{schafer1977tuning}.
 
@@
 Considering sound scene synthesis or composition as part of an experimental protocol requires the availability of software resources that are easier and easier to master.
-While most processing framework that have been considered are developed in native language for efficiency purposes like Marsyas \cite{tzanetakis2000marsyas} or Clam \cite{amatriain2006clam}, many advanced sound manipulation.
-
-Closer to our sound model is the Tapestrea framework \cite{misra2007musical} that allows the manipulation sounds trough wavelet and sinusoidal modeling to allow sound manipulations for doing acousmatic composition or sound
+While most processing framework that have been considered are developed in native language for efficiency purposes like Marsyas developed by~\cite{tzanetakis2000marsyas} or Clam developed by~\cite{amatriain2006clam}, many advanced sound manipulation.
+
+Closer to our sound model is the Tapestrea framework proposed by~\cite{misra2007musical} that allows the manipulation sounds trough wavelet and sinusoidal modeling to allow sound manipulations for doing acousmatic composition or sound
 design.

### 2018-06-27 v2/responseToReviewers.tex
diffing

@@
 \documentclass[10pt]{article}
+\usepackage{url}
 
@@
 
-$\rightarrow$
+$\rightarrow$ We definitely agree.
+As now discussed in more detail in Section 6 of the manuscript, the proposed software tool is not designed for genericity by we believe that the proposed specific design strategies and use of recent software toolkits can be considered as inspirations for desiging other experimental protocols questionning soundscape perception.
 
@@
 
-$\rightarrow$ The
+$\rightarrow$ The references to the ASA theory and statistical processing of textures are now removed.
+The work of Nelken and de Cheveigné is still cited as, in our humble opinion, it describes a soundscape model that is very close to the one used in the study described in this paper.
 

### 2018-06-28 lafaySimulatedScenePerception.tex
remove some acronyms

--- 2018-06-26 lafaySimulatedScenePerception.tex
@@
 
-Finally, this work allows us to conjecture as to the nature of the mental representations of the concepts "pleasant (urban sound) environment" (PE) and "unpleasant (urban sound) environment" (UE).
+Finally, this work allows us to conjecture as to the nature of the mental representations of the concepts "pleasant urban sound environment" and "unpleasant urban sound environment" .
 
@@
 
-First, the fact that the semantic information (which sound source is present) and compositional information (at which level) are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the PE and UE concepts.
+First, the fact that the semantic information (which sound source is present) and compositional information (at which level) are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the above cited concepts.
 

### 2018-07-15 lafaySimulatedScenePerception.tex
Add GL revisions

--- 2018-06-27 lafaySimulatedScenePerception.tex
@@
 
+It indeed appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli.
+
+\cite{maffiolo_caracterisation_1999} distinguishes two separated categorization processes, either of which is triggered depending on the listener's ability to identify sound events.
+She shows that those processes lead to two abstract cognitive categories respectively termed "event sequences" and "amorphous sequences".
+Event sequences are composed of salient events that can easily be recognized, such as \emph{car start} or \emph{male speech}.
+They arise from a descriptive analysis based on the identification of the sound sources.
+On the contrary, amorphous sequences are sound environments where distinct events cannot be readily identified, such as \emph{traffic hubhub}.
+They result from a holistic analysis based on global acoustic features.
+
+Concerning sound textures, \ie~sounds that have stable characteristics over time, \cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can opt for an abstract, statistical representation of the perceived information, discarding precise physical properties of the sound.
+
+In order to account for the possibility that the morphological differences between soud events and sound textures may have some important consequences on the perception of the scenes, the simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with classes of events only and the other with classes of textures only.
+These two types of sound classes also have specific scheduling procedures during the simulation process.
+
+We consider in this study the soundscape as a ``\,skeleton of events on a bed of textures\,'' as coined by~\cite{nelken2013ear}.
+
 % et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures. Événements et textures bénéficient de processus de simulation spécifiques.
@@
 %Les différences morphologiques entre événements et textures sonores ont des conséquences sur la perception des scènes.
-
-In order to account for the possibility that the morphological differences between those two classes may have some important consequences on the perception of the scenes, the simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with classes of events only and the other with classes of textures only.
-These two types of sound classes also have specific scheduling procedures during the simulation process.
 
@@
 
-According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment~\cite{bregman1994auditory}.
+%According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment~\cite{bregman1994auditory}.
 Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}, while events that cannot be isolated are grouped into a same stream.
@@
 Elle montre l'existence de deux catégories cognitives abstraites d'environnements sonores respectivement appelées: «~les séquences événementielles~» et «~les séquences amorphes~».
-
-Maffiolo~\cite{maffiolo_caracterisation_1999} distinguishes two separated categorization processes, either of which is triggered depending on the listener's ability to identify sound events.
-She shows that those processes lead to two abstract cognitive categories respectively termed "event sequences" and "amorphous sequences".
 
@@
 
-Event sequences are composed of salient events that can easily be recognized, such as \emph{car start} or \emph{male speech}.
-They arise from a descriptive analysis based on the identification of the sound sources.
-On the contrary, amorphous sequences are sound environments where distinct events cannot be readily identified, such as \emph{traffic hubhub}.
-They result from a holistic analysis based on global acoustic features.
-
 %Pour ce qui est des textures, \ie~des sons possédants des caractéristiques stables au cours du temps, \cite{mcdermott2011sound,mcdermott2013summary} montrent que le cerveau peut décider sciemment de dégrader l'information physique perçue, en la résumant de manière statistique, faisant fi de toute autre représentation plus détaillée \cite{nelken2013ear}.
 
-Concerning sound textures, \ie~sounds that have stable characteristics over time,~\cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can opt for an abstract, statistical representation of the perceived information, discarding precise physical properties of the sound~\cite{nelken2013ear}.
+%Concerning sound textures, \ie~sounds that have stable characteristics over time,~\cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can opt for an abstract, statistical representation of the perceived information, discarding precise physical properties of the sound~\cite{nelken2013ear}.
 
@@
 
-In the light of those results, it appears that the processing of auditory information comprises some sort of decision concerning the nature of the stimuli~\cite{nelken2013ear,mcdermott2013summary}.
-We thus consider in this study the soundscape as a ``\,skeleton of events on a bed of textures\,'' as coined in~\cite{nelken2013ear}.
+% mcdermott2013summary
 
@@
 \item to detect the presence of \emph{outliers} or ambiguous scenes.
-Indeed, throughout our analyses, the predefined hedonic properties of the scenes (ideal or not) are used as reference.
+Indeed, throughout our analyses, the predefined hedonic properties of the scenes (ideal or non ideal) are used as reference.
 We thus need to ensure beforehand that no ambiguity exists between extreme cases of ideal and non ideal scenes, \ie~that the least pleasant ideal scene remains more highly rated on average than the most pleasant non ideal scene.
@@
 
-We have shown that, from a qualitative point of view, the composition of the scenes in terms of sound sources differs between i- or non ideal scenes.
+We have shown that, from a qualitative point of view, the composition of the scenes in terms of sound sources differs between ideal or non ideal scenes.
 We now investigate whether some of the sound classes are specific to a given environment.
@@
 
-Considering $E$ and $T$ separately, it appears that: 1) the rank 5-precision obtained with $E$ is similar to the one obtained with $ET$ ; 2) the precision obtained with $T$ is always lower than the one obtained with $E$, by $10\%$ to $15\%$.
+Considering $E$ and $T$ separately, it appears that: 1) the rank 5-precision obtained with $E$ is similar to the one obtained with $ET$ ; 2) the rank 5-precision obtained with $T$ is always lower than the one obtained with $E$, by $10\%$ to $15\%$.
 Those results indicate that the semantic information that is discriminative is mostly carried by the events.
@@
 In this section, the sound levels are computed by taking into account only the previously identified sound markers.
-We define $X_m$, the $X$ feature computed by taking into account only the sound markers, and $X_b$, the $X$ feature computed by taking into account all the sound classes, except the sound markers.
+We define $L_m$ (resp.
+$L(E)_m$ and $L(T)_m$), the sound level computed by taking into account only the sound markers, and $L_b$ (resp.
+$L(E)_b$ and $L(T)_b$), the sound level computed by taking into account all the sound classes, except the sound markers.
 When the feature characterizes an ideal scene (resp.
@@
 
-Notwithstanding the presence or absence of markers, the km-scenes and rm-scenes are exactly the same.
+Notwithstanding the presence or absence of markers, the scenes with and without markers are exactly the same.
 
@@
 
-In order to create rm-scenes with still some sound diversity and no  absence of sound activity within a long period of the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}.
+In order to create the scenes without markers with still some sound diversity and no absence of sound activity within a long period of the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}.
 Those classes are:
@@
 The first 10 scenes allow the subject to calibrate their scores.
-Those calibration scenes are  5 ideal scenes with markers and 5 nideal scenes with markers.
+Those calibration scenes are  5 ideal scenes with markers and 5 non ideal scenes with markers.
 These first 10 scenes are replayed at the end of the experiment, and only the scores given at the second occurrence are taken into account.
@@
 To do so, we perform a variance analysis.
-We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (i/ni) and the presence/absence of markers (km/rm).
-As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to evaluate whether there exist a significant difference of perceived pleasantness between km-scenes and rm-scenes.
+We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (ideal or non ideal) and the presence/absence of markers.
+As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to evaluate whether there exist a significant difference of perceived pleasantness.
 The two independent variables are considered as \emph{within-subject} factors.
-The factors being of only two levels each (type: i/ni,  marker: km/rm), the sphericity hypothesis does not need to be checked.
+The factors being of only two levels each (type: ideal or non ideal,  marker: with or without), the sphericity hypothesis does not need to be checked.
 \emph{Post hoc} analyses are done using the Tukey-Kramer procedure.
@@
 
-Let us consider $\mathcal{A}_{subject}$ for the km-scenes (Figure~\ref{fig:xp2_Asubject}).
+Let us consider $\mathcal{A}_{subject}$ for the scenes with markers (Figure~\ref{fig:xp2_Asubject}).
 Close inspection of the answers show that one subject's judgments strongly differ from the others'.
-This subject evaluated positively close to half of the non ideal scenes with markers (see Annex~\ref{app:xp2}, subject 7), and gave a score above 0 for 58\% of the nideal scenes with markers, contrary to an average of 11\% for the other subjects.
+This subject evaluated positively close to half of the non ideal scenes with markers (see Annex~\ref{app:xp2}, subject 7), and gave a score above 0 for 58\% of the non ideal scenes with markers, contrary to an average of 11\% for the other subjects.
 
@@
 
-In this section, we study the scores given by the subject while listening to the several types of scenes, namely the ideal scenes with markers, nideal scenes with markers, ideal scenes without markers and non ideal scenes without markers (Figure~\ref{fig:xp2_Ascene}).
-The repeated measure ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
+In this section, we study the scores given by the subject while listening to the several types of scenes, namely the ideal scenes with markers, non ideal scenes with markers, ideal scenes without markers and non ideal scenes without markers (Figure~\ref{fig:xp2_Ascene}).
+The repeated measure ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the type environment (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
 
@@
 
-The \emph{post hoc} analysis exhibits significant differences between all groups of observations, notably between the i/km- and the i/rm scenes ($p<0.05$) as well as between the ni/km- and the non ideal scenes without markers ($p<0.01$).
+The \emph{post hoc} analysis exhibits significant differences between all groups of observations, notably between the ideal scenes with markers and the ideal scenes without markers ($p<0.05$) as well as between the non ideal scenes with markers and the non ideal scenes without markers ($p<0.01$).
 
@@
 The significant interaction shows that the type of environment influences the effect of the presence/absence of the markers.
-Indeed, the average difference between km-scenes and rm-scenes is larger for the non ideal scenes ($1.1$) than for the ideal scenes ($0.5$).
+Indeed, the average difference between the scenes with markers and the scenes without markers is larger for the non ideal scenes ($1.1$) than for the ideal scenes ($0.5$).
 

### 2018-07-15 lafaySimulatedScenePerception.tex
replace all i and ni

@@
 Each of them has to create two sound environments: one ideal~/ pleasant, and the other non ideal~/ unpleasant.
-\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal (i) or non ideal (ni).
+\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal or non ideal.
 Furthermore, this information is given by the creator of the scene.
@@
 
-\item \emph{semantic features}: we use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene, \textit{i.e.} the sound classes that are present / absent from the scene.
+\item \emph{semantic features}: we use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene, \ie~the sound classes that are present / absent from the scene.
 Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise.
@@
 
-\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  i- and non ideal scenes during the experiment 1.a.
-To do so, we consider if there exist significant differences for the i- and non ideal scenes of the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$.
+\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  ideal and non ideal scenes during the experiment 1.a.
+To do so, we consider if there exist significant differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ of the ideal and non ideal scenes.
 The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.
@@
 
-\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the i- and non ideal scenes. The significance is measured with a two-sample Student test.
+\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the ideal and non ideal scenes. The significance is measured with a two-sample Student test.
 
@@
 
-\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni).
+\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (ideal or non ideal).
 To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}.
-With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
+With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of environment $k$ ($k=\lbrace ideal, non ideal\rbrace$), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
 For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:
@@
 Using the Hamming distance allows us to take into account equally the presence and absence of classes.
-In order to measure the intrinsic ability of the space to discriminate between i- and non ideal scenes, we use a ranking metric named the precision at rank $k$ ($P@k$).
+In order to measure the intrinsic ability of the space to discriminate between ideal and non ideal scenes, we use a ranking metric named the precision at rank $k$ ($P@k$).
 The $P@k$ computes the precision obtained after the $k$ closest items with respect to a given seed item have been found.
@@
 As a consequence, 36 ideal scenes and 36 non ideal scenes remain for analysis.
-Second, we verify that subjects really perceived a difference in terms of pleasantness between i- and non ideal scenes.
+Second, we verify that subjects really perceived a difference in terms of pleasantness between ideal and non ideal scenes.
 For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment.
@@
 \begin{itemize}
-\item allows us to differentiate between i- and non ideal scenes,
+\item allows us to differentiate between ideal and non ideal scenes,
 \item characterizes precisely the perceived pleasantness for non ideal scenes (unsurprisingly, an unpleasant scene gets all the more unpleasant as it gets louder),
@@
 
-Those first outcomes tend to show that 1) two modes of perception exist depending on the nature of the environment (i or ni), and 2) each involves distinct, independent features.
+Those first outcomes tend to show that 1) two modes of perception exist depending on the nature of the environment (ideal or non ideal), and 2) each involves distinct, independent features.
 The fact that $L$ is not sufficient to characterize the pleasantness of the ideal scenes can lead us to conclude that all sound sources do not equally contribute to the perception of pleasantness.
 We thus put forward the hypothesis that only the level of some of them can influence this perception.
-In order to investigate further in that direction, we analyze in the next section the scenes from a semantic point of view, \ie we take an interest in the nature of the sources they are made of.
+In order to investigate further in that direction, we analyze in the next section the scenes from a semantic point of view, \ie~we take an interest in the nature of the sources they are made of.
 
@@
   \subfloat[]{\includegraphics[width=.5\linewidth]{gfx/xp1_class_3}\label{fig:soundsourcec}}
-                             \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.
+                             \caption{Proportion of simulated scenes (i: ideal, ni: non ideal) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.
 }\label{fig:soundsource}
@@
 
-We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate a given type of environment (i or ni).
-For the 36 i and non ideal scenes considered, results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
+We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate a given type of environment (ideal or non ideal).
+For the 36 ideal and 36 non ideal scenes considered, results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
 For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$, is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.
@@
 
-We observe a noticeable difference in terms of class choices between the i- and non ideal scenes.
-The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes by~\cite{guastavino2006ideal}, \ie~ on one hand, classes involving human presence and nature are prevailing in the ideal scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the non ideal scenes.
+We observe a noticeable difference in terms of class choices between the ideal and non ideal scenes.
+The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes by~\cite{guastavino2006ideal}, \ie~on one hand, classes involving human presence and nature are prevailing in the ideal scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the non ideal scenes.
 These results confirm a fact previously observed by~\cite{raimbault2005urban,dubois2006cognitive}: the semantic nature of the sound sources play an important role in the assessment of the environment.
@@
 \begin{enumerate}
-\item unlike what we outlined with the sound levels, a semantic description of the scenes composition in terms of presence~/ absence of sound sources allows us to reliably differentiate between the two types of environments (i or ni);
+\item unlike what we outlined with the sound levels, a semantic description of the scenes composition in terms of presence~/ absence of sound sources allows us to reliably differentiate between the two types of environments (ideal or non ideal);
 \item the semantic information is mainly contained in the event sound classes;
-\item only a part of the event classes, i.e. the sound markers, are useful to differentiate between the ideal and non ideal scenes.
+\item only a part of the event classes, \ie~the sound markers, are useful to differentiate between the ideal and non ideal scenes.
 \end{enumerate}
@@
 To conclude, the level of ni- markers has a negative influence on the pleasantness for the non ideal scenes.
-On the other hand, the level of i- markers does not influence the perceived pleasantness for the ideal scenes.
+On the other hand, the level of the ideal markers does not influence the perceived pleasantness for the ideal scenes.
 
@@
 Moreover, whereas $L(T)$ did not show any significant correlation for non ideal scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
-This indicates that the level of non-marker texture classes does not influence the perceived pleasantness in the same way for i- and non ideal scenes.
+This indicates that the level of non-marker texture classes does not influence the perceived pleasantness in the same way for ideal and non ideal scenes.
 Sound level of textures seems to have a negative effect for the non ideal scenes, and no significant effect for the ideal scenes.
@@
 
-\item \emph{differentiating i- and non ideal scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ et $L(T)$), allow us to differentiate reliably between ideal and non ideal scenes.
+\item \emph{differentiating ideal and non ideal scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ et $L(T)$), allow us to differentiate reliably between ideal and non ideal scenes.
 The semantic description seems to be more powerful;
 \item \emph{events or textures}: whatever the feature type, be it semantic or related to the sound pressure level, events are the most useful components of the scene to differentiate the two types of scenes; textures bring a limited amount of information;
-\item \emph{pleasantness prediction}: considering the correlation between sound levels and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its very nature (i or ni).
+\item \emph{pleasantness prediction}: considering the correlation between sound levels and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its very nature (ideal or non ideal).
 From the data gathered in those experiments, the same set of features cannot be considered to predict the pleasantness of ideal and non ideal scenes:
@@
 
-Thus, only part of the ideal markers and non ideal markers are removed from the rm-scenes.
+Thus, only part of the ideal markers and non ideal markers are removed from the scenes without markers.
 
@@
         {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_12}\label{fig:xp4_note_1l}}
-        \caption{Distribution of scores given by the subjects during experiment 2 for i/am-scenes (green) et ni/am-scenes (red).
+        \caption{Distribution of scores given by the subjects during experiment 2 for the ideal scenes with markers (green) et non ideal scenes with markers (red).
 The horizontal axis is, from left to right, the 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant).}\label{fig:xp4_note_1}

### 2018-07-16 v2/responseToReviewers.tex
integrates jfp comments

@@
 $\rightarrow$ We definitely agree.
-As now discussed in more detail in Section 6 of the manuscript, the proposed software tool is not designed for genericity by we believe that the proposed specific design strategies and use of recent software toolkits can be considered as inspirations for desiging other experimental protocols questionning soundscape perception.
+As now discussed in more detail in Section 6 of the manuscript, the proposed software tool is not designed for genericity but we believe that the proposed specific design strategies and use of recent software toolkits can be considered as inspirations for desiging other experimental protocols questionning soundscape perception.
 

### 2018-07-16 lafaySimulatedScenePerception.tex
integrates jfp comments

@@
 \begin{enumerate}
-\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non ideal soundscapes that are considered as material for the following experiments; The resulting waveforms are available for download \url{https://archive.org/details/soundSimulatedUrbanScene}.
+\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non-ideal soundscapes that are considered as material for the following experiments; The resulting waveforms are available for download \url{https://archive.org/details/soundSimulatedUrbanScene}.
 \item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
@@
 Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds.
-Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used by~\cite{Salamon14}.
-Only three levels of abstraction are considered for the texture sounds.
+Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in~\cite{Salamon14}, see Figure~\ref{fig:taxonomieEb}.
+Only three levels of abstraction are considered for the texture sounds, see Figure~\ref{fig:taxonomieT}.
 
@@
 
-As such, the underlying model of the scene is a sum of sound sources. The simulation interface is more toroughly described by~\cite{rossignol2015simscene}.
+As such, the underlying model of the scene is a sum of sound sources. The simulation interface is more thoroughly described by~\cite{rossignol2015simscene}.
 
@@
 \begin{itemize}
-\item \emph{experiment 1.a (simulation)}: during this experiment, subjects are asked to simulate urban sound environments using \emph{Simscene} (see Section~\ref{sec:simulator}).
-Each of them has to create two sound environments: one ideal~/ pleasant, and the other non ideal~/ unpleasant.
-\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal or non ideal.
+\item \emph{experiment 1.a (simulation)}: during this experiment, subjects are asked to create simulated urban sound environments using \emph{Simscene} (see Section~\ref{sec:simulator}).
+Each of them has to create two sound environments: one ideal~/ pleasant, and the other non-ideal~/ unpleasant.
+\item \emph{experiment 1.b (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal or non-ideal.
 Furthermore, this information is given by the creator of the scene.
@@
 \begin{enumerate}
-\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (ideal or non ideal) thanks to a finer quantification of the pleasantness of the scene;
+\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (ideal or non-ideal) thanks to a finer quantification of the pleasantness of the scene;
 \item to detect the presence of \emph{outliers} or ambiguous scenes.
-Indeed, throughout our analyses, the predefined hedonic properties of the scenes (ideal or non ideal) are used as reference.
-We thus need to ensure beforehand that no ambiguity exists between extreme cases of ideal and non ideal scenes, \ie~that the least pleasant ideal scene remains more highly rated on average than the most pleasant non ideal scene.
+Indeed, throughout our analyses, the predefined hedonic properties of the scenes (ideal or non-ideal) are used as reference.
+We thus need to ensure beforehand that no ambiguity exists between extreme cases of ideal and non-ideal scenes, \ie~that the least pleasant ideal scene remains more highly rated on average than the most pleasant non-ideal scene.
 \end{enumerate}
@@
 \item  first simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
-\item second simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
+\item second simulation: create a \textbf{plausible urban soundscape} which is non-ideal, according to you (where you would not like to live);
 \end{itemize}
@@
   \item  select sound classes,
-  \item  give each of them a name,
+  \item  give to each of them a name,
   \item  set the parameters of the tracks related to the selected sound classes of sounds, see Section~\ref{sec:simscene_parametre}.
@@
 
-All the subjects performed the experiment at the same time. They were equally distributed in three identical quiet rooms, and are not allowed to talk to each other during the experiment.
+All the subjects performed the experiment at the same time. They were equally distributed in three identical quiet rooms, and were not allowed to talk to each other during the experiment.
 
@@
 
-Among the 44 subjects, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes).
-the 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.
+Among the 44 subjects, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non-ideal scenes).
+The 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.
 The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available \url{http://soundthings.org/research/urbanSoundScape/XP2014}.
@@
 
-The assessment is done with a 7-point bipolar semantic scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant).
+The assessment is done with a 7-point bipolar semantic scale going from -3 (non-ideal / unpleasant) to +3 (ideal / pleasant).
 Before evaluating a scene, subjects must listen to the first 20 seconds of the stimuli.
@@
 For each participant, sound scenes are played in a quasi random order.
-5 ideal scenes and 5 non ideal scenes are first sequenced to allow the subjects to calibrate their scores.
+5 ideal scenes and 5 non-ideal scenes are first sequenced to allow the subjects to calibrate their scores.
 These first 10 scenes are played back again at the end of the experiment.
@@
 
-All the subjects perform the experiment simultaneously. They are not allowed to talk to each other during the experiment.
+All the subjects perform the experiment simultaneously. They were not allowed to talk to each other during the experiment.
 
@@
 Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores.
-$\mathcal{A}_{subject}$ is computed for ideal scenes and non ideal scenes.
+$\mathcal{A}_{subject}$ is computed for ideal scenes and non-ideal scenes.
 
@@
 
-\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  ideal and non ideal scenes during the experiment 1.a.
-To do so, we consider if there exist significant differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ of the ideal and non ideal scenes.
+\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  ideal and non-ideal scenes during the experiment 1.a.
+To do so, we consider if there exists significant differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ of the ideal and non-ideal scenes.
 The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.
@@
 
-\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the ideal and non ideal scenes. The significance is measured with a two-sample Student test.
+\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the ideal and non-ideal scenes. The significance is measured with a two-sample Student test.
 
@@
 
-\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (ideal or non ideal).
+\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (ideal or non-ideal).
 To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}.
-With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of environment $k$ ($k=\lbrace ideal, non ideal\rbrace$), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
+With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of environment $k$ ($k \in \lbrace ideal, non-ideal\rbrace$), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
 For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:
@@
 Using the Hamming distance allows us to take into account equally the presence and absence of classes.
-In order to measure the intrinsic ability of the space to discriminate between ideal and non ideal scenes, we use a ranking metric named the precision at rank $k$ ($P@k$).
+In order to measure the intrinsic ability of the space to discriminate between ideal and non-ideal scenes, we use a ranking metric named the precision at rank $k$ ($P@k$).
 The $P@k$ computes the precision obtained after the $k$ closest items with respect to a given seed item have been found.
@@
 
-First, in order to ensure the coherence of the data, we check that none of the non ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of a ideal scene.
-4 non ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes.
-As a consequence, 36 ideal scenes and 36 non ideal scenes remain for analysis.
-Second, we verify that subjects really perceived a difference in terms of pleasantness between ideal and non ideal scenes.
+First, in order to ensure the coherence of the data, we check that none of the non-ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of a ideal scene.
+4 non-ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes.
+As a consequence, 36 ideal scenes and 36 non-ideal scenes remain for analysis.
+Second, we verify that subjects really perceived a difference in terms of pleasantness between ideal and non-ideal scenes.
 For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment.
-It indeed appears that the ideal scenes were perceived significantly more pleasant ($p < 0.01$) than the non ideal scenes.
+It indeed appears that the ideal scenes were perceived significantly more pleasant ($p < 0.01$) than the non-ideal scenes for all the participants.
 
@@
         {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_6}\label{fig:soundlevelf}}
-       \caption{Distributions of the sound levels $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (i: ideal, ni: non ideal) (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
+       \caption{Distributions of the sound levels $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (i: ideal, ni: non-ideal) (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
 \end{figure}
@@
 Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the distributions of levels $L$, $L(E)$ and $L(T)$.
-There is a significant difference in terms of sound levels between ideal and non ideal scenes ($L$: $p<0.01$).
+There is a significant difference in terms of sound levels between ideal and non-ideal scenes ($L$: $p<0.01$).
 This difference is significant for events ($L(E)$: $p<0.01$, mean deviation: -7 $dB$) and for textures ($L(T)$: $p<0.01$, mean deviation: -6 $dB$).
@@
 
-As expected, the sound level of the sources is indeed a pleasantness indicator, as the non ideal scenes tend to be louder.
+As expected, the sound level of the sources is indeed a pleasantness indicator, as the non-ideal scenes tend to be louder.
 This results is also an outcome of a large number of related works.
@@
 
-It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for ideal and non ideal scenes.
+It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for ideal and non-ideal scenes.
 This observation is in agreement with the results obtained by~\cite{kuwano_memory_2003}.
@@
 
-However, level alone is not sufficient to fully differentiate between ideal and non ideal scenes.
-In fact, $20\%$ of the ideal scenes have a sound levels higher than the lowest level of the non ideal scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
+However, level alone is not sufficient to fully differentiate between ideal and non-ideal scenes.
+In fact, $20\%$ of the ideal scenes have a sound levels higher than the lowest level of the non-ideal scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
 
@@
 \begin{tabular}{l c c c}
-               & all scenes                & ideal scenes          & non ideal scenes    \\
+               & all scenes                & ideal scenes          & non-ideal scenes    \\
 \hline
@@
 In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated.
-Contrary to the previous test, we do not limit ourselves to a binary ideal scenes \emph{vs.} non ideal scenes disctinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature.
+Contrary to the previous test, we do not limit ourselves to a binary ideal scenes \emph{vs.} non-ideal scenes disctinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature.
 The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$.
@@
 Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived.
-Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for ideal and non ideal scenes.
-In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for non ideal scenes ($r=-0.78$, $p<0.01$), but is not significant ($p=0.06$) for ideal scenes.
+Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for ideal and non-ideal scenes.
+In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for non-ideal scenes ($r=-0.78$, $p<0.01$), but is not significant ($p=0.06$) for ideal scenes.
 
@@
 
-When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the ideal scenes tend to be softer than the non ideal scenes, thus allowing us to extend artificially to the ideal scenes the negative correlation observed for the non ideal scenes.
+When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the ideal scenes tend to be softer than the non-ideal scenes, thus allowing us to extend artificially to the ideal scenes the negative correlation observed for the non-ideal scenes.
 
@@
 \begin{itemize}
-\item allows us to differentiate between ideal and non ideal scenes,
-\item characterizes precisely the perceived pleasantness for non ideal scenes (unsurprisingly, an unpleasant scene gets all the more unpleasant as it gets louder),
+\item allows us to differentiate between ideal and non-ideal scenes,
+\item characterizes precisely the perceived pleasantness for non-ideal scenes (unsurprisingly, an unpleasant scene gets all the more unpleasant as it gets louder),
 \item is not a relevant feature for modeling the perceived pleasantness of an \textit{a priori} pleasant soundscape (ideal scene).
@@
 The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}.
-For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (ideal scenes: $r=-0.33$, $p=0.05$, non ideal scenes: $r=-0.00$, $p=0.99$).
-Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (ideal scenes tend to be softer than non ideal scenes).
-Thus, while sound event levels maintain a relative ability to predict the pleasantness of the non ideal scenes, texture levels do not bring much information, whatever the type of environment is.
+For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (ideal scenes: $r=-0.33$, $p=0.05$, non-ideal scenes: $r=-0.00$, $p=0.99$).
+Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (ideal scenes tend to be softer than non-ideal scenes).
+Thus, while sound event levels maintain a relative ability to predict the pleasantness of the non-ideal scenes, texture levels do not bring much information, whatever the type of environment is.
 
@@
 
-Those first outcomes tend to show that 1) two modes of perception exist depending on the nature of the environment (ideal or non ideal), and 2) each involves distinct, independent features.
+Those first outcomes tend to show that 1) two modes of perception exist depending on the nature of the environment (ideal or non-ideal), and 2) each involves distinct, independent features.
 The fact that $L$ is not sufficient to characterize the pleasantness of the ideal scenes can lead us to conclude that all sound sources do not equally contribute to the perception of pleasantness.
@@
   \subfloat[]{\includegraphics[width=.5\linewidth]{gfx/xp1_class_3}\label{fig:soundsourcec}}
-                             \caption{Proportion of simulated scenes (i: ideal, ni: non ideal) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.
+                             \caption{Proportion of simulated scenes (i: ideal, ni: non-ideal) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.
 }\label{fig:soundsource}
@@
 
-We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate a given type of environment (ideal or non ideal).
-For the 36 ideal and 36 non ideal scenes considered, results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
+We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate a given type of environment (ideal or non-ideal).
+For the 36 ideal and 36 non-ideal scenes considered, results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
 For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$, is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.
@@
 
-We observe a noticeable difference in terms of class choices between the ideal and non ideal scenes.
-The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes by~\cite{guastavino2006ideal}, \ie~on one hand, classes involving human presence and nature are prevailing in the ideal scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the non ideal scenes.
+We observe a noticeable difference in terms of class choices between the ideal and non-ideal scenes.
+The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes by~\cite{guastavino2006ideal}, \ie~on one hand, classes involving human presence and nature are prevailing in the ideal scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the non-ideal scenes.
 These results confirm a fact previously observed by~\cite{raimbault2005urban,dubois2006cognitive}: the semantic nature of the sound sources play an important role in the assessment of the environment.
@@
 
-Nevertheless, we notice some differences with results of~\cite{guastavino2006ideal} which show that sounds of \emph{public transportation} are specific of ideal urban soundscapes.
+Nevertheless, we notice some differences with the results of~\cite{guastavino2006ideal} which show that sounds of \emph{public transportation} are specific of ideal urban soundscapes.
 The authors interpret this by the fact that the perception of pleasantness is, among other things, due to socio-cultural factors.
@@
 To a certain extent, our results contradict this result.
-In fact,  Figure~\ref{fig:soundsourcea} shows that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the subjects for $28\%$ of the ideal scenes, and for $42\%$ of the non ideal scenes.
+In fact,  Figure~\ref{fig:soundsourcea} shows that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the subjects for $28\%$ of the ideal scenes, and for $42\%$ of the non-ideal scenes.
 Those results do not question the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the ideal scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes.
-However, \emph{public transport} classes are also strongly present in the non ideal scenes, for instance more than \emph{light vehicle} or \emph{truck} classes.
+However, \emph{public transport} classes are also strongly present in the non-ideal scenes, for instance more than \emph{light vehicle} or \emph{truck} classes.
 On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.
@@
 Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
-level & ideal scenes & non ideal scenes \\
+level & ideal scenes & non-ideal scenes \\
 \hline
@@
   & \multicolumn{2}{c}{Texture sound markers}      \\
-  & ideal scenes & non ideal scenes \\
+  & ideal scenes & non-ideal scenes \\
 \hline
@@
 
-We have shown that, from a qualitative point of view, the composition of the scenes in terms of sound sources differs between ideal or non ideal scenes.
+We have shown that, from a qualitative point of view, the composition of the scenes in terms of sound sources differs between ideal or non-ideal scenes.
 We now investigate whether some of the sound classes are specific to a given environment.
@@
 In fact, according to Schafer, a sound that is identified by a person as being an important element of his/her environment, is well accepted.
-Sound markers of non ideal scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
+Sound markers of non-ideal scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
 
@@
 For the ideal scenes, those classes related to subdued or quiet ambiances (\emph{courtyard}, \emph{park}).
-The marker classes for the non ideal scenes are, as for the events, related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroads}).
+The marker classes for the non-ideal scenes are, as for the events, related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroads}).
 
@@
 Reciprocally, if the sound markers are not taken into account for the description ($E_{w/o,m}$), the performance is below the one achieved when considering only the textures as features.
-Thus, most of the semantic information allowing to differentiate between ideal and non ideal scenes is included in the markers.
+Thus, most of the semantic information allowing to differentiate between ideal and non-ideal scenes is included in the markers.
 
@@
 \begin{enumerate}
-\item unlike what we outlined with the sound levels, a semantic description of the scenes composition in terms of presence~/ absence of sound sources allows us to reliably differentiate between the two types of environments (ideal or non ideal);
+\item unlike what we outlined with the sound levels, a semantic description of the scenes composition in terms of presence~/ absence of sound sources allows us to reliably differentiate between the two types of environments (ideal or non-ideal);
 \item the semantic information is mainly contained in the event sound classes;
-\item only a part of the event classes, \ie~the sound markers, are useful to differentiate between the ideal and non ideal scenes.
+\item only a part of the event classes, \ie~the sound markers, are useful to differentiate between the ideal and non-ideal scenes.
 \end{enumerate}
@@
 
-Since we have extracted the typical classes of the ideal and non ideal scenes and verified that the distinction between them was largely dependent on the presence of these classes, we shall now investigate whether a description of the scenes only based on the sound pressure level of sound markers could characterize the perceived pleasantness, perhaps better than a globally computed sound level.
+Since we have extracted the typical classes of the ideal and non-ideal scenes and verified that the distinction between them was largely dependent on the presence of these classes, we shall now investigate whether a description of the scenes only based on the sound pressure level of sound markers could characterize the perceived pleasantness, perhaps better than a globally computed sound level.
 
@@
 \begin{tabular}{l r r}
-                  &   ideal scenes                  & non ideal scenes \\
+                  &   ideal scenes                  & non-ideal scenes \\
 \hline
@@
         {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
-        \caption{Distribution of the relative sound levels related to the presence of markers $L_m$ (a, d), $L_b$ (b, e) and $L_m-L_b$ (c, f), versus scene type (i: ideal, ni: non ideal) (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
+        \caption{Distribution of the relative sound levels related to the presence of markers $L_m$ (a, d), $L_b$ (b, e) and $L_m-L_b$ (c, f), versus scene type (i: ideal, ni: non-ideal) (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
 \end{figure}
@@
 When the feature characterizes an ideal scene (resp.
-non ideal scene), only the markers identified for the ideal scenes (resp.
-non ideal scenes) are considered.
-We henceforth call \emph{ideal markers} and \emph{non ideal markers} the two types of markers.
+non-ideal scene), only the markers identified for the ideal scenes (resp.
+non-ideal scenes) are considered.
+We henceforth call \emph{ideal markers} and \emph{non-ideal markers} the two types of markers.
 Results are shown on Table~\ref{tab:corrMarkers}.
@@
 \begin{enumerate}
-\item a significant difference between levels of ideal and non ideal scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
+\item a significant difference between levels of ideal and non-ideal scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
 \item the sound level of scenes is mainly related to the sound events, compared to the textures;
-\item the sound level of events has an influence on the perception of pleasantness for non ideal scenes, but not for ideal scenes;
+\item the sound level of events has an influence on the perception of pleasantness for non-ideal scenes, but not for ideal scenes;
 \item the sound level of textures does not play any role in the perception of the pleasantness.
@@
 
-To conclude, the level of ni- markers has a negative influence on the pleasantness for the non ideal scenes.
+To conclude, the level of ni- markers has a negative influence on the pleasantness for the non-ideal scenes.
 On the other hand, the level of the ideal markers does not influence the perceived pleasantness for the ideal scenes.
@@
 
-Moreover, whereas $L(T)$ did not show any significant correlation for non ideal scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
-This indicates that the level of non-marker texture classes does not influence the perceived pleasantness in the same way for ideal and non ideal scenes.
-Sound level of textures seems to have a negative effect for the non ideal scenes, and no significant effect for the ideal scenes.
+Moreover, whereas $L(T)$ did not show any significant correlation for non-ideal scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$).
+This indicates that the level of non-marker texture classes does not influence the perceived pleasantness in the same way for ideal and non-ideal scenes.
+Sound level of textures seems to have a negative effect for the non-ideal scenes, and no significant effect for the ideal scenes.
 
@@
 For the ideal scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
-for the non ideal scenes no correlation is observed.
+for the non-ideal scenes no correlation is observed.
 Thus, for the ideal scenes, it is not the absolute markers level that is important, but their relative level with respect to the other sounds composing the scene.
@@
 
-On the contrary, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ for the non ideal scenes, shows that this is indeed the absolute level that matters.
+On the contrary, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ for the non-ideal scenes, shows that this is indeed the absolute level that matters.
 
@@
 
-\item \emph{differentiating ideal and non ideal scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ et $L(T)$), allow us to differentiate reliably between ideal and non ideal scenes.
+\item \emph{differentiating ideal and non-ideal scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ et $L(T)$), allow us to differentiate reliably between ideal and non-ideal scenes.
 The semantic description seems to be more powerful;
 \item \emph{events or textures}: whatever the feature type, be it semantic or related to the sound pressure level, events are the most useful components of the scene to differentiate the two types of scenes; textures bring a limited amount of information;
-\item \emph{pleasantness prediction}: considering the correlation between sound levels and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its very nature (ideal or non ideal).
-From the data gathered in those experiments, the same set of features cannot be considered to predict the pleasantness of ideal and non ideal scenes:
+\item \emph{pleasantness prediction}: considering the correlation between sound levels and pleasantness, it seems that the way subjects perceived the quality of a given environment depends on its very nature (ideal or non-ideal).
+From the data gathered in those experiments, the same set of features cannot be considered to predict the pleasantness of ideal and non-ideal scenes:
 
@@
 
-\item \emph{for non ideal scenes}, the global levels ($L$ et $L(E)$), or the level of sound markers ($L_m$ et $L(E)_{m}$), have a negative influence on pleasantness.
+\item \emph{for non-ideal scenes}, the global levels ($L$ et $L(E)$), or the level of sound markers ($L_m$ et $L(E)_{m}$), have a negative influence on pleasantness.
 Taking into account the contribution of each of the different sources of the scene does not improve the prediction performance compared to a global analysis of the environment.
@@
 
-The fact that the pleasantness of the ideal scenes is not correlated to global physical features, contrary to the pleasantness of non ideal scenes, has also been studied recently by~\cite{gozalo2015relationship}.
+The fact that the pleasantness of the ideal scenes is not correlated to global physical features, contrary to the pleasantness of non-ideal scenes, has also been studied recently by~\cite{gozalo2015relationship}.
 
@@
 We can assume two perceptual modes of operation that involve different types of features and rely on the hedonic nature  of the stimuli.
-It thus appears that the features considered for the pleasantness judgment also depends on a preliminary judgment of the global hedonic nature of the environment (ideal or non ideal).
+It thus appears that the features considered for the pleasantness judgment also depends on a preliminary judgment of the global hedonic nature of the environment (ideal or non-ideal).
 
@@
 
-Another hypothesis would be that the volume somehow act as an hedonic "gain" factor.
+Another hypothesis would be that the volume somehow acts as an hedonic "gain" factor.
 If the volume of a negative marker is high, it lowers the overall pleasantness.
-If the volume of a positive marker is high, it highers the overall pleasantness.
+If the volume of a positive marker is high, it increases the overall pleasantness.
 Evidently, the positive gain is expected to saturate at a given level and will quickly decreases as the level raises above a given threshold.
@@
 
-In order to investigate deeper into the relation between the pleasantness of the ideal and non ideal scenes and the sound markers, the audio waveform of the simulated scenes are regenerated with or without the classes identified as markers.
-To do so, ideal markers are removed from the ideal scenes, and non ideal markers are removed from the non ideal scenes.
+In order to investigate deeper into the relation between the pleasantness of the ideal and non-ideal scenes and the sound markers, the audio waveform of the simulated scenes are regenerated with or without the classes identified as markers.
+To do so, ideal markers are removed from the ideal scenes, and non-ideal markers are removed from the non-ideal scenes.
 To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptual test is conducted with a protocol close to the one considered in experiment 1.b.
@@
 
-The objective of this experiment is to study if the removal of the previously identified markers have an impact on the perceived pleasantness. Two hypothesis are thus formulated:
+The objective of this experiment is to study if the removal of the previously identified markers has an impact on the perceived pleasantness. Two hypothesis are thus formulated:
 
@@
 \begin{itemize}
-\item \emph{for the non ideal scenes}, we hypothesize that the absence of non ideal markers will \textbf{increase} the pleasantness score;
+\item \emph{for the non-ideal scenes}, we hypothesize that the absence of non-ideal markers will \textbf{increase} the pleasantness score;
 \item \emph{for the ideal scenes}, we hypothesize that the absence of the ideal markers will \textbf{decrease} the pleasantness score.
@@
 \item \emph{church bell}, \emph{bicycle bell}, and \emph{animals} for the ideal scenes without markers;
-\item \emph{siren}, \emph{car horn} for the non ideal scenes without markers.
+\item \emph{siren}, \emph{car horn} for the non-ideal scenes without markers.
 \end{itemize}
@@
 
-Thus, only part of the ideal markers and non ideal markers are removed from the scenes without markers.
+Thus, only part of the ideal markers and non-ideal markers are removed from the scenes without markers.
 
@@
 The first 10 scenes allow the subject to calibrate their scores.
-Those calibration scenes are  5 ideal scenes with markers and 5 non ideal scenes with markers.
+Those calibration scenes are  5 ideal scenes with markers and 5 non-ideal scenes with markers.
 These first 10 scenes are replayed at the end of the experiment, and only the scores given at the second occurrence are taken into account.
@@
 They listen to the stimuli with semi-open \emph{Beyer-Dynamic DT 990 Pro} headphones, with the sound level set to be the same for all the subjects.
-They are not allowed to talk to each other during the experiment.
+They were not allowed to talk to each other during the experiment.
 
@@
 
-The aim here is to validate the hypothesis that the removal of ideal markers and non ideal markers have an impact on the perceived pleasantness.
-To do so, we perform a variance analysis.
-We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (ideal or non ideal) and the presence/absence of markers.
-As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA is used to evaluate whether there exist a significant difference of perceived pleasantness.
+The aim here is to validate the hypothesis that the removal of ideal markers and non-ideal markers has a significant effect on the perceived pleasantness.
+To do so, we perform an analysis of variance (ANOVA).
+We consider the pleasantness rating of each subject on each scene as a dependent variable, and as independent variables, the type of environment (ideal or non-ideal) and the presence/absence of markers.
+As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA with interaction is used to evaluate whether there exist a significant difference of perceived pleasantness.
 The two independent variables are considered as \emph{within-subject} factors.
-The factors being of only two levels each (type: ideal or non ideal,  marker: with or without), the sphericity hypothesis does not need to be checked.
+The factors being of only two levels each (type: ideal or non-ideal,  marker: with or without), the sphericity hypothesis does not need to be checked.
 \emph{Post hoc} analyses are done using the Tukey-Kramer procedure.
 %\jfp{l'analyse est inutile a mon avis, il n'y a que 2 niveaux pour chaque facteur de l'ANOVA.} \gl{2 niveaux par facteur, mais 4 niveaux en tout}
+
+% $\mathcal{A}_{subject}$
 
@@
         {\includegraphics[width=.48\linewidth]{gfx/xp4_note_15}\label{fig:xp2_Ascene}}
-        \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness per scene $\mathcal{A}_{scene}$ (b) versus scene type (i: ideal, ni: non ideal, km: with markers, rm: without markers).
+        \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness per scene $\mathcal{A}_{scene}$ (b) versus scene type (i: ideal, ni: non-ideal, km: with markers, rm: without markers).
 Black stars in subfigure (a) stand for the detected outlier, \ie~subject 7.}\label{fig:xp2A}
@@
 Close inspection of the answers show that one subject's judgments strongly differ from the others'.
-This subject evaluated positively close to half of the non ideal scenes with markers (see Annex~\ref{app:xp2}, subject 7), and gave a score above 0 for 58\% of the non ideal scenes with markers, contrary to an average of 11\% for the other subjects.
-
-Furthermore, this subject used the whole scale (from -5 to 5) to score both the ideal scenes and the non ideal scenes.
+This subject evaluated positively close to half of the non-ideal scenes with markers (see Annex~\ref{app:xp2}, subject 7), and gave a score above 0 for 58\% of the non-ideal scenes with markers, contrary to an average of 11\% for the other subjects.
+
+Furthermore, this subject used the whole scale (from -5 to 5) to score both the ideal scenes and the non-ideal scenes.
 Those behaviors strongly differ with the ones of the other subjects, be they from this experiment or the previous ones.
@@
 
-In this section, we study the scores given by the subject while listening to the several types of scenes, namely the ideal scenes with markers, non ideal scenes with markers, ideal scenes without markers and non ideal scenes without markers (Figure~\ref{fig:xp2_Ascene}).
-The repeated measure ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the type environment (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
+In this section, we study the scores given by the subject while listening to the several types of scenes, namely the ideal scenes with markers, non-ideal scenes with markers, ideal scenes without markers and non-ideal scenes without markers (Figure~\ref{fig:xp2_Ascene}).
+The repeated measure ANOVA applied to the pleasantness rating shows a significant effect of the type environment (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
+
+%  $\mathcal{A}_{subject}$
 
@@
 
-The \emph{post hoc} analysis exhibits significant differences between all groups of observations, notably between the ideal scenes with markers and the ideal scenes without markers ($p<0.05$) as well as between the non ideal scenes with markers and the non ideal scenes without markers ($p<0.01$).
+The \emph{post hoc} analysis exhibits significant differences between all groups of observations, notably between the ideal scenes with markers and the ideal scenes without markers ($p<0.05$) as well as between the non-ideal scenes with markers and the non-ideal scenes without markers ($p<0.01$).
 
@@
 
-These results indicate that the removal of the markers indeed modify the perception of the scenes by the subjects. Our two hypotheses are thus verified :
+These results indicate that the removal of the markers indeed modify the perception of the scenes by the subjects, see Figure~\ref{fig:xp2A}. Our two hypotheses are thus verified :
 
@@
 \begin{itemize}
-\item the removal of non ideal markers improved the pleasantness of the non ideal scenes;
+\item the removal of non-ideal markers improved the pleasantness of the non-ideal scenes;
 \item the removal of ideal markers reduced the pleasantness of the ideal scenes.
@@
 The significant interaction shows that the type of environment influences the effect of the presence/absence of the markers.
-Indeed, the average difference between the scenes with markers and the scenes without markers is larger for the non ideal scenes ($1.1$) than for the ideal scenes ($0.5$).
+Indeed, the average difference between the scenes with markers and the scenes without markers is larger for the non-ideal scenes ($1.1$) than for the ideal scenes ($0.5$).
 
@@
 This experiment shows that the presence of the markers identified during the analysis of experiment 1 does have an impact on the perceived pleasantness.
-The removal of the non ideal markers has a positive effect on the perception of the non ideal scenes.
+The removal of the non-ideal markers has a positive effect on the perception of the non-ideal scenes.
 Perhaps more surprisingly the removal of the ideal markers slightly decreases the perception of the ideal scenes: this is a more striking observation since, due to the removal of the markers, the acoustic pressure level of the ideal scenes with markers is higher than the one of the ideal scenes without markers.
@@
 This strongly confirms that ideal markers do have a positive impact on the perception of an environment.
-The fact that their removal decreases $\mathcal{A}_{scene}$ indicate that it should be possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
+The fact that their removal decrease $\mathcal{A}_{scene}$ indicates that it should be possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
 Those conclusions are in line with the positive approach introduced by~\cite{schafer1977tuning}.
@@
 
-This series of three experiments showed that most of the descriptors used in this study, be they of a semantic or acoustic nature, allow us to distinguish between an ideal scene and a non ideal one.
+This series of three experiments showed that most of the descriptors used in this study, be they of a semantic or acoustic nature, allow us to distinguish between an ideal scene and a non-ideal one.
 
@@
 \begin{itemize}
-\item in the case of non ideal scenes, one should focus on reducing the acoustic pressure level, whether globally, or by discarding specific sources such as \emph{sirens} or \emph{car horns}
+\item in the case of non-ideal scenes, one should focus on reducing the acoustic pressure level, whether globally, or by discarding specific sources such as \emph{sirens} or \emph{car horns}
 \item in the case of ideal scenes, one should first identify which sources are pleasant to the targeted community, second lower the volume of the other sound sources, and, if possible, raise the contribution or add positive sound markers.
@@
 
-Considering sound scene synthesis or composition as part of an experimental protocol requires the availability of software resources that are easier and easier to master.
-While most processing framework that have been considered are developed in native language for efficiency purposes like Marsyas developed by~\cite{tzanetakis2000marsyas} or Clam developed by~\cite{amatriain2006clam}, many advanced sound manipulation.
-
-Closer to our sound model is the Tapestrea framework proposed by~\cite{misra2007musical} that allows the manipulation sounds trough wavelet and sinusoidal modeling to allow sound manipulations for doing acousmatic composition or sound
+Considering sound scene synthesis or composition as part of an experimental protocol requires the availability of software resources that are simple to manipulate for subjects with little training, and possibly running on many types of hardware.
+While most processing frameworks that have been considered are developed in native language for efficiency purposes like Marsyas developed by~\cite{tzanetakis2000marsyas} or Clam developed by~\cite{amatriain2006clam}, many advanced sound manipulations can now be performed on modern web browsers using the webaudio liibrary.
+
+Closer to our sound model is the Tapestrea framework proposed by~\cite{misra2007musical} that allows the manipulation of sounds trough wavelet and sinusoidal modeling for acousmatic composition or sound
 design.
@@
 
-Extensions to other sound parameters beyond acoustic levels and .
-However, the control of those new parameters would need a specific user experience study to maintain a high degree of usability of the controlling interface.
+Extensions to other sound parameters beyond acoustic levels and sound events distribution could be integrated as the software is open source.
+However, the control of those new parameters would need a specific user experience study to maintain a sufficient degree of usability of the controlling interface.
 
@@
 
-One interesting avenue in this direction would be to study the impact of adding positively appreciated sound markers to a non ideal scene in order to study the hypothesis that this type of addition would improve the perceived quality of the scene.
+One interesting avenue in this direction would be to study the impact of adding positively appreciated sound markers to a non-ideal scene in order to study the hypothesis that this type of addition would improve the perceived quality of the scene.
 
@@
         {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_12}\label{fig:xp4_note_1l}}
-        \caption{Distribution of scores given by the subjects during experiment 2 for the ideal scenes with markers (green) et non ideal scenes with markers (red).
+        \caption{Distribution of scores given by the subjects during experiment 2 for the ideal scenes with markers (green) et non-ideal scenes with markers (red).
 The horizontal axis is, from left to right, the 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant).}\label{fig:xp4_note_1}

### 2018-07-16 v2/responseToReviewers.tex
update response

@@
 
-$\rightarrow$ The references to the ASA theory and statistical processing of textures are now removed.
-The work of Nelken and de Cheveigné is still cited as, in our humble opinion, it describes a soundscape model that is very close to the one used in the study described in this paper.
+$\rightarrow$ The references to the ASA theory are now removed.
+The work of McDermott and Nelken and de Cheveigné are still cited as, in our humble opinion, they have some relevance to the matter discussed in this section.
 

### 2018-07-16 lafaySimulatedScenePerception.tex
Update section 4.3

@@
 
-As asserted by~\cite{aletta2016soundscape}, one of the main goals of soundscape studies is to identify which components of the soundscape influence human perception.
+One of the main goals of soundscape studies is to identify which components of the soundscape influence human perception, see~\cite{aletta2016soundscape}.
 Establishing a link between an environment (urban sound scene) and the induced human sensation (calmness) is equivalent to instantiating the mental representation of a specific sound scene (a calm urban sound scene).
@@
 We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene, see~\cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,szeremeta2009analysis}.
-Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene, as found by~\cite{gozalo2015relationship,ricciardi2015sound}.
+Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene, see~\cite{gozalo2015relationship,ricciardi2015sound}.
 
@@
 \begin{enumerate}
-\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non-ideal soundscapes that are considered as material for the following experiments; The resulting waveforms are available for download \url{https://archive.org/details/soundSimulatedUrbanScene}.
+\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non-ideal soundscapes that are considered as material for the following experiments; The resulting waveforms are available for download\footnote{Soundscapes, available online \url{https://archive.org/details/soundSimulatedUrbanScene}}.
 \item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
@@
 
-The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts written by~\cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} that study how humans discriminate different kind of urban soundscapes.
+The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts written that study how humans discriminate different kind of urban soundscape, see~\cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} .
 
@@
 \item {sound events}: isolated sounds, precisely located in time, whose acoustical characteristics may change with respect to time;
-\item {sound textures}: isolated sounds of long duration whose acoustical characteristics are stable with respect to time, as expressed by~\cite{saint1995classification}.
+\item {sound textures}: isolated sounds of long duration whose acoustical characteristics are stable with respect to time, see~\cite{saint1995classification}.
 \end{itemize}
@@
 The 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.
-The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available \url{http://soundthings.org/research/urbanSoundScape/XP2014}.
+The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available online\footnote{Available online \url{http://soundthings.org/research/urbanSoundScape/XP2014}}.
 
@@
 
-All the subjects perform the experiment on computers with equivalent hardware and software configurations.
-The audio files are played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}.
-The stimuli are the scenes obtained in experiment 1.a.
-The output sound level is the same for all the subjects.
+All the subjects performed the experiment on computers with equivalent hardware and software configurations.
+The audio files were played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}.
+The stimuli were the scenes obtained in experiment 1.a.
+The output sound level was the same for all the subjects.
 
@@
 
-All the subjects perform the experiment simultaneously. They were not allowed to talk to each other during the experiment.
+All the subjects performed the experiment simultaneously. They were not allowed to talk to each other during the experiment.
 
@@
 
-An experimenter is available during the whole duration of the experiment in order to assist subjects and answer questions any questions the subjects may have.
+An experimenter was available during the whole duration of the experiment in order to assist subjects and answer any questions the subjects might have.
 
@@
 Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
-%                                    &                            \\
-% ideal/pleasant                     & i                          \\
-% non-ideal/unpleasant               & ni                         \\
-% scene ideal/pleasant               & i-scene                    \\
-% scene non-ideal/unpleasant         & ni-scene                   \\
+                                   &                            \\
+                                   &                           \\
+                                   &                          \\
+                                   &                     \\
+                                   &                    \\
 \hline
@@
 \vspace{0.5mm}
-\caption{Acronyms of features used in the analysis of the sensory experiments.}
+\caption{Acronyms of features used in the analysis of the experiments.}
 \label{tab:acronyme}
@@
 \item \emph{Influence of sound levels on perceived pleasantness}: the goal is to evaluate if the sound levels impact the perceived pleasantness.
-To do so, we consider linear correlations between these features and $\mathcal{A}_{scene}$.
+To do so, we consider linear correlations between this feature and $\mathcal{A}_{scene}$.
 The Pearson correlation coefficient is used for that purpose.
@@
 To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}.
-With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of environment $k$ ($k \in \lbrace ideal, non-ideal\rbrace$), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
+With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of environment $k$ (ideal or non-ideal), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
 For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:
@@
 For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment.
-It indeed appears that the ideal scenes were perceived significantly more pleasant ($p < 0.01$) than the non-ideal scenes for all the participants.
+It indeed appears that the ideal scenes were perceived significantly more pleasant ($p < 0.01$) than the non-ideal scenes.
 
@@
 The fact that subjects in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact.
-Other studies that considered sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment, as studied by~\cite{lavandier2006contribution}.
+Other studies that considered sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment, see~\cite{lavandier2006contribution}.
 
@@
 Those results are in line with results of~\cite{maffiolo_caracterisation_1999}.
-As discussed in Section~\ref{sec:simscene_sampleDataSet}, it seems that humans analyze the event scenes which are composed of several sound events in a causal manner, \ie~by identifying the sources.
+As discussed in Section~\ref{sec:simscene_sampleDataSet}, it seems that humans analyze the event scenes which are composed of several sound events in a descriptive manner, \ie~by identifying the sources.
 
@@
 
-Concerning the sound levels, the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ and $L(T)$ on the other side, see~Figures~\ref{fig:soundlevelMarkera} and~\ref{fig:soundlevelMarkerd}, respectively.
+Concerning the sound levels, the same trends are measured between $L_m$, $L(E)_m$ and $L(T)_m$ on one side, and $L$, $L(E)$ and $L(T)$ on the other side, see~Figures~\ref{fig:soundlevelMarkera} and~\ref{fig:soundlevelMarkerd}.
 No matter whether all the classes or only the markers are considered, it appears that:
@@
 
-To conclude, the level of ni- markers has a negative influence on the pleasantness for the non-ideal scenes.
+To conclude, the level of non-ideal markers has a negative influence on the pleasantness for the non-ideal scenes.
 On the other hand, the level of the ideal markers does not influence the perceived pleasantness for the ideal scenes.
@@
 
-Considering the non markers classes, we can observe on the ideal scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see Figures~\ref{fig:soundlevelNoisea}  and ~\ref{fig:soundlevelNoised}.
+Considering the non markers classes, we can observe on the ideal scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see~Figures~\ref{fig:soundlevelNoisea} and~\ref{fig:soundlevelNoised}.
 It is the first time that an objective feature allows us to define the pleasantness of ideal scenes.
@@
 A last group of features is now considered, namely $L_m-L_b$, $L(E)_m-L(E)_b$ and $L(T)_m-L(T)_b$.
-These features describe the difference between the markers level and those of the other sound classes, see ~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}.
+These features describe the difference between the markers level and those of the other sound classes, see~Figures~\ref{fig:soundlevelMarkerDiffa} and~\ref{fig:soundlevelMarkerDiffd}.
 They express the saliency of the markers with respect to the sound mixture.
@@
 For the ideal scenes, a moderate positive correlation is observed for $L_m-L_b$ ($r=0.67$, $p<0.01$) and $L(E)_m-L(E)_b$ ($r=0.66$, $p<0.01$).
-for the non-ideal scenes no correlation is observed.
+For the non-ideal scenes no correlation is observed.
 Thus, for the ideal scenes, it is not the absolute markers level that is important, but their relative level with respect to the other sounds composing the scene.
@@
 \item the higher the absolute level of sounds not being ideal markers, the weaker the pleasantness,
-\item the higher the relative level of ideal markers is, with respect to the remaining sounds, the higher the pleasantness.
+\item the higher the relative level of ideal markers compared to the remaining sounds, the higher the pleasantness.
 \end{itemize}
@@
 
-The aim here is to validate the hypothesis that the removal of ideal markers and non-ideal markers has a significant effect on the perceived pleasantness.
+The aim here is to validate the hypothesis that the removal of ideal and non-ideal markers has a significant effect on the perceived pleasantness.
 To do so, we perform an analysis of variance (ANOVA).
-We consider the pleasantness rating of each subject on each scene as a dependent variable, and as independent variables, the type of environment (ideal or non-ideal) and the presence/absence of markers.
-As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA with interaction is used to evaluate whether there exist a significant difference of perceived pleasantness.
+We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (ideal or non-ideal) and the presence/absence of markers.
+As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA with interaction is used to evaluate whether there exist a significant difference of perceived pleasantness between the scenes with and without markers.
 The two independent variables are considered as \emph{within-subject} factors.
@@
 In this section, we study the scores given by the subject while listening to the several types of scenes, namely the ideal scenes with markers, non-ideal scenes with markers, ideal scenes without markers and non-ideal scenes without markers (Figure~\ref{fig:xp2_Ascene}).
-The repeated measure ANOVA applied to the pleasantness rating shows a significant effect of the type environment (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
+The repeated measure ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the type environment (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
 
@@
 
-%\subsection{Discussion}
+\subsection{Discussion}
 

### 2018-07-23 lafaySimulatedScenePerception.tex
nm comments

@@
     \hspace*{8pt}mathieu.lagrange@cnrs.fr\\
-    $^{2)}$ STMS Ircam-CNRS-UPMC Institut de Recherche et Coordination Acoustique/Musique, Paris, France}
+    $^{2)}$ STMS Ircam-CNRS-SU Institut de Recherche et Coordination Acoustique/Musique, Paris, France}
 
@@
 
-In order to account for the possibility that the morphological differences between soud events and sound textures may have some important consequences on the perception of the scenes, the simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with classes of events only and the other with classes of textures only.
+In order to account for the possibility that the morphological differences between sound events and sound textures may have some important consequences on the perception of the scenes, the simulation tool \emph{Simscene} follows this distinction and considers two distinct sound databases: one with classes of events only and the other with classes of textures only.
 These two types of sound classes also have specific scheduling procedures during the simulation process.
@@
         \myfloatalign
-        \includegraphics[width=.8\linewidth]{gfx/3.pdf}
+        \includegraphics[width=\linewidth]{gfx/3.pdf}
        \caption{Hierarchical organisation of the isolated sounds used in the simulation.}\label{fig:orgDb}
@@
 Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user.
-Most browsing tools are based on keyword indexing; however, for sensory experiments that study the objectivation of a subject's mental representations, this means is problematic: indeed, the availability of a verbal description of the sound can influence the subject's choice, and potentially induce biases in the analysis conclusions.
+Most browsing tools are based on keyword indexing; however, for sensory experiments that study the objectivation of a subject's mental representations, this can be problematic as the availability of a verbal description of the sound can influence the subject's choice, and potentially induce biases in the analysis conclusions.
 For example, a subject may automatically select sounds referenced as belonging to a \emph{park} environment to build a \emph{calm} soundscape, rather than focusing on their perception.
@@
 of $2.0$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part in the experiment.
-All the subjects had been living in Nantes, France, for at least two years at the time of the experiment.
+All the subjects had been living in Nantes, France, for at least two years at the time of the experiment and reported normal hearing.
 
@@
 
-Due to temporal constraints, subjects only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from second 15 to sec. 45).
+Due to temporal constraints, subjects only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from sec. 15 to sec. 45).
 
@@
 of $1.8$ years) from \emph{Ecole Centrale de Nantes} took part to the experiment.
-All the subjects had been living in Nantes, France, for at least two years at the time of the experiment.
+All the subjects had been living in Nantes, France, for at least two years at the time of the experiment and reported normal hearing.
 None of them took part to the previous simulation experiment (experiment 1.a).
@@
 If the null hypothesis is rejected, the class $j$ is said to be typical with respect to the type of scene $k$.
-Such typical classes are called \textbf{sound markers}.
+Such typical classes are called \textbf{sound markers}, in reference to the work of \cite{schafer1993soundscape}.
 Testing is done for each class, at each level of abstraction, and separately for texture and event classes.
@@
         {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_6}\label{fig:soundlevelf}}
-       \caption{Distributions of the sound levels $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (i: ideal, ni: non-ideal) (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
+       \caption{Distributions of the sound levels $L$ (a, d), $L(E)$ (b, e) and $L(T)$ (c, f), with respect to scene type (i: ideal, ni: non-ideal) (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
 \end{figure}
@@
 \begin{enumerate}
-\item a significant difference between levels of ideal and non-ideal scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
+\item a significant difference between levels of ideal and non-ideal scenes exists ($L_m$, $L(E)_m$ and $L(T)_m$: $p<0.01$);
 \item the sound level of scenes is mainly related to the sound events, compared to the textures;
@@
 None of them have performed experiments 1.a and 1.b.
-All the subjects had been living in Nantes, France, for at least two years at the time of the experiment.
+All the subjects had been living in Nantes, France, for at least two years at the time of the experiment and reported normal hearing.
 All the subjects performed the test successfully.
@@
 \begin{itemize}
-\item The simulator can be deployed on a large scale via the Internet thanks to the web-based software architecture;
+\item The simulator can be deployed on a large scale via the Internet thanks to the web-based software architecture, provided that the sound databases are adapted to the cultural background of the cohort;
 \item Simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.

### 2018-09-03 lafaySimulatedScenePerception.tex
GL relecture

@@
 %Nous pensons que la simulation permet de faire un lien élégant et efficace entre ces deux approches.
-Il permet en effet d'obtenir du sujet une version modale, \ie~établie sur des dimensions physiques (le signal de la scène simulée), de sa représentation mentale de l'objet; les scènes générées étant à la fois caractérisées sémantiquement (\eg~sources sonores présentes) et physiquement (\eg~niveaux des sources sonores).
+Elle permet en effet d'obtenir du sujet une version modale, \ie~établie sur des dimensions physiques (le signal de la scène simulée), de sa représentation mentale de l'objet; les scènes générées étant à la fois caractérisées sémantiquement (\eg~sources sonores présentes) et physiquement (\eg~niveaux des sources sonores).
 
@@
 For several reasons that are detailed in this paper, the use of a soundscape simulator such as the one proposed here can lead to interesting outcomes as it allows the experimenter to separately study  the influence of the sound sources that compose a sound environment.
-With such material, not only is the type of sound sources available for study, but also the exact level and audio waveform for each source together with the structural properties of the scene, that is, the temporal distribution of the sound events.
+With such material, not only is the type of sound sources available for study, but also the exact sound level and audio waveform for each source together with the structural properties of the scene, that is, the temporal distribution of the sound events.
 
@@
 Experiment~2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section~\ref{sec:modification}.
-Conclusions and discussion about future work follow in Section \ref{sec:conclusion}.
+Conclusions and discussion about future work follow in Section~\ref{sec:conclusion}.
 
@@
 
-The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts written that study how humans discriminate different kind of urban soundscape, see~\cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories,maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive,leobon_analyse_1986,brown2011towards} .
+The chosen typology is established based on the category/classes of sounds found while reviewing several articles or thesis manuscripts that study how humans discriminate different kind of urban soundscape, see~\cite{leobon_analyse_1986,maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,beaumont2004pertinence,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,polack2008perceptive,niessen2010categories,brown2011towards} .
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsubsection*{Events and Textures}
-
-%La littérature fait généralement la distinction entre :
-% \begin{itemize}
-% \item {un événement sonore}: un son isolé, ponctuel, dont les caractéristiques physiques varient au cours du temps;
-% \item {une texture sonore}: un son, isolé, long, dont les caractéristiques physiques restent stables au cours du temps \cite{saint1995classification}.
-% \end{itemize}
 
@@
 
-% et utilise deux banques de données distinctes : l'une regroupant des classes d’événements, l'autre des classes de textures. Événements et textures bénéficient de processus de simulation spécifiques.
-
-%Les différences morphologiques entre événements et textures sonores ont des conséquences sur la perception des scènes.
-
-%Lorsque des événements émergent d'un environnement sonore, le cerveau traite l'information des différentes sources de manière séparée \cite{bregman1994auditory}.
-Plusieurs flux auditifs sont ainsi générés, \ie~un pour chaque séquence d'événements émis par la même source \cite{carlyon2004brain}.
-Inversement, quand le cerveau ne parvient pas à isoler des événements, les éléments constituants de la scène sont agglomérés dans un même flux.
-
-%According to the Auditory Scene Analysis (ASA) theory, the brain processes separately events emerging from the different sources of a sound environment~\cite{bregman1994auditory}.
-Several auditory streams are created, one for each sequence of events emitted by the same source \cite{carlyon2004brain}, while events that cannot be isolated are grouped into a same stream.
-
-%Maffiolo~\cite{maffiolo_caracterisation_1999} met en évidence deux processus de catégorisation distincts, engagés en fonction de la capacité de l'auditeur à identifier des événements sonores.
-Elle montre l'existence de deux catégories cognitives abstraites d'environnements sonores respectivement appelées: «~les séquences événementielles~» et «~les séquences amorphes~».
-
-%Les séquences événementielles sont des environnements composés d'événements saillants et identifiables (\emph{démarrage de voiture}, \emph{voix d'homme}).
-Elles profitent d'une analyse descriptive basée sur l'identification des sources sonores.
-Les séquences amorphes sont des environnements dont il est difficile d'isoler des éléments distincts («~brouhaha de rue~», «~brouhaha de trafic~»).
-Elles profitent, elles, d'une analyse holistique, effectuée à partir d'indicateurs acoustiques (objectifs) globaux.
-%Cette différence de traitement entre scènes événementielles et scènes amorphes est relevée également dans~\cite{guastavino2006ideal,dubois2006cognitive}.
-
-%Pour ce qui est des textures, \ie~des sons possédants des caractéristiques stables au cours du temps, \cite{mcdermott2011sound,mcdermott2013summary} montrent que le cerveau peut décider sciemment de dégrader l'information physique perçue, en la résumant de manière statistique, faisant fi de toute autre représentation plus détaillée \cite{nelken2013ear}.
-
-%Concerning sound textures, \ie~sounds that have stable characteristics over time,~\cite{mcdermott2011sound,mcdermott2013summary} demonstrates that the human brain can opt for an abstract, statistical representation of the perceived information, discarding precise physical properties of the sound~\cite{nelken2013ear}.
-
-%Au regard de ces résultats, il apparaît que le processus de traitement de l'information sonore comprend une prise de décision quant à la nature des stimuli \cite{nelken2013ear,mcdermott2013summary}, laquelle va ensuite influer sur la manière d'extraire et d'analyser l'information.
-
-% mcdermott2013summary
-
 \subsubsection*{Taxonomy}
@@
 
-Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds.
-Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in~\cite{Salamon14}, see Figure~\ref{fig:taxonomieEb}.
+Accounting for the previously detailed perceptual matters, two taxonomies are built, one for sound events and one for sound textures.
+Up to four levels of abstraction are considered from the most generic classes(level 0) to the most specific classes (level 3), leading to a taxonomy close to the one used in~\cite{Salamon14}, see Appendix~\ref{app:taxonomie}.
 Only three levels of abstraction are considered for the texture sounds, see Figure~\ref{fig:taxonomieT}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Texture samples are sequenced without time spacing, therefore the parameters \emph{event fade} and \emph{inter-onset spacing} are not available for this kind of track.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Dans le cadre d'une expérience sensorielle visant à objectiver la représentation interne du sujet, cette approche est problématique.
-
-%En effet la description verbale du son, si elle est accessible au sujet, peut potentiellement influencer sa sélection.
+En effet la description verbale du son, si elle est accessible au sujet, peut potentiellement influencer sa sélection.
 Dans les faits, pour construire une scène environnementale «~calme~», le sujet sélectionne a priori les sons référencés sous le vocable \emph{parc}.
@@
 Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user.
-Most browsing tools are based on keyword indexing; however, for sensory experiments that study the objectivation of a subject's mental representations, this can be problematic as the availability of a verbal description of the sound can influence the subject's choice, and potentially induce biases in the analysis conclusions.
+Most browsing tools are based on keyword indexing; however, for sensory experiments that study the objectivation of a subject's mental representations, this may be problematic as the availability of a verbal description of the sound can influence the subject's choice, and potentially induce biases in the analysis conclusions.
 For example, a subject may automatically select sounds referenced as belonging to a \emph{park} environment to build a \emph{calm} soundscape, rather than focusing on their perception.
@@
 As such, the underlying model of the scene is a sum of sound sources. The simulation interface is more thoroughly described by~\cite{rossignol2015simscene}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 % Chaque processus de simulation comprend deux parties :
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsubsection*{Apparatus}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsubsection*{Subjects}
@@
 44 students (30 male, 14 female; averaging $21.6$ years of age, s.d.
-of $2.0$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part in the experiment.
+of $2.0$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) performed the experiment.
 All the subjects had been living in Nantes, France, for at least two years at the time of the experiment and reported normal hearing.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Planning of experiment 1.b}
@@
 
-Due to temporal constraints, subjects only assess 30 seconds of the initial 1-minute simulated scenes generated in experiment 1.a (from sec. 15 to sec. 45).
+The subjects evaluate the 80 simulated scenes generated in experiment 1.a.
+Due to temporal constraints, subjects only assess 30 seconds of the initial 1-minute simulated scenes (from sec.
+15 to sec.
+45).
 
@@
 
-All the subjects performed the experiment on computers with equivalent hardware and software configurations.
+All the subjects performed the experiment on standard desktop computers with the same hardware and software configurations.
 The audio files were played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}.
@@
 
-All the subjects performed the experiment simultaneously. They were not allowed to talk to each other during the experiment.
+All the subjects performed the experiment simultaneously in a quiet environment. They were not allowed to talk to each other during the experiment.
 
@@
 
-An experimenter was available during the whole duration of the experiment in order to assist subjects and answer any questions the subjects might have.
+An experimenter was available during the whole duration of the experiment in order to assist subjects and to answer questions.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 10 students (8 male, 2 female; averaging $23.1$ years of age, s.d.
-of $1.8$ years) from \emph{Ecole Centrale de Nantes} took part to the experiment.
+of $1.8$ years) from \emph{Ecole Centrale de Nantes} performed the experiment.
 All the subjects had been living in Nantes, France, for at least two years at the time of the experiment and reported normal hearing.
-None of them took part to the previous simulation experiment (experiment 1.a).
+None of them took part in the previous simulation experiment (experiment 1.a).
 
@@
 All the subjects succeeded in doing the experiment.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
-                                   &                            \\
-                                   &                           \\
-                                   &                          \\
-                                   &                     \\
-                                   &                    \\
 \hline
@@
 A set of features, upon which the analysis is conducted, is attached to each sound scene.
-A summary of those features (and the corresponding acronyms) is presented in Table ~\ref{tab:acronyme}.
+A summary of those features (and the corresponding acronyms) is presented in Table~\ref{tab:acronyme}.
 In order to be consistent with the evaluation of experiment 1.b, features are not computed on the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b (Section~\ref{sec:xp1b_plan}).
@@
 For each sound scene, three types of features are considered :
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale.
-$\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the subjects.
-$\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant.
+$\mathcal{A}_{scene}$ is the average pleasantness for each scene, computed as the average of all the scores given by all the subjects to a specific scene.
+$\mathcal{A}_{subject}$ is the average pleasantness for each subject, computed as the average of all the scores given by a specific subject to all the scenes.
+$\mathcal{A}_{subject}$ is computed for ideal scenes and non-ideal scenes separately.
 Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores.
-$\mathcal{A}_{subject}$ is computed for ideal scenes and non-ideal scenes.
 
@@
 
-\item \emph{structural features}: while \emph{SimScene} allows us to acces a variety of information about the scene structure (such as the density of events), we only focus in this first study on the sound levels.
+\item \emph{structural features}: while \emph{SimScene} allows us to access a variety of information about the scene structure (such as the density of events), we only focus in this first study on the sound levels.
 To figure those out, we draw inspiration from the $L_{Aeq}$.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %Pour évaluer l'impact spécifique des différentes sources sonores sur l'agrément perçu, nous soumettons nos travaux aux 5 tests/études de significativité présentés ci-après:
@@
 \begin{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the ideal and non-ideal scenes during the experiment 1.a.
-To do so, we consider if there exists significant differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ of the ideal and non-ideal scenes.
+To do so, we consider if there exists significant differences between the ideal and non-ideal scenes for $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$.
 The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.
 
-%\jfp{je ne comprends pas, pour moi il faut faire, pour chaque sujet de l'expé 1, un test de différence entre Aniscène  et Aiscène a quoi sert Asubject ici ?} \gl{Les deux sont possibles : Ascene mesure des différences structurelles, Asubject des différences perceptives}
-
-% GL comprends pas en francais
-
-%\item \emph{analyse des descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels, nous évaluons cette significativité à partir d'un test de Student à deux populations;
-
-\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the ideal and non-ideal scenes. The significance is measured with a two-sample Student test.
+%\item \emph{analyse des descripteurs structurels}: afin d'évaluer si la distinction affective imposée entre les i- et ni-scènes impacte de manière significative la nature des scènes, \ie~s'il existe des différences significatives entre les descripteurs structurels.
+Nous évaluons cette significativité à partir d'un test de Student à deux populations;
+
+\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels ($L$, $L(E)$ and $L(T)$) differ between the ideal and non-ideal scenes.
+The significance is measured with a two-sample Student test.
 
@@
 
-\item \emph{Influence of sound levels on perceived pleasantness}: the goal is to evaluate if the sound levels impact the perceived pleasantness.
-To do so, we consider linear correlations between this feature and $\mathcal{A}_{scene}$.
+\item \emph{Influence of sound levels on perceived pleasantness}: the goal is to evaluate if the sound levels ($L$, $L(E)$ and $L(T)$) impact the perceived pleasantness.
+To do so, we consider linear correlations between those features and $\mathcal{A}_{scene}$.
 The Pearson correlation coefficient is used for that purpose.
@@
 Le V-test teste l'hypothèse nulle que la proportion $\frac{c_{jk}}{c}$ ne diffère pas significativement de la proportion $\frac{c_{jk}}{c_k}$.
-Si pour un environnement $k$, et une classe $j$, l'hypothèse est rejetée, la classe $j$ est alors typique de l'environnement $k$.
-Les classes typiques sont nommées \textbf{marqueurs sonores};
-
-\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (ideal or non-ideal).
+
+\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used in a given type of environment (ideal or non-ideal).
 To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}.
-With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of environment $k$ (ideal or non-ideal), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
+With $c$ being the total number of classes used to simulate all the scenes, $c_k$ the number of classes used to simulate the scenes of a given type of environment $k$ (ideal or non-ideal), $c_j$ the number of times a class $j$ has been used to simulate all the scenes, and $c_{jk}$ the number of time a class $j$ has been used to simulate the scenes of a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
 For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:
@@
 
-If the null hypothesis is rejected, the class $j$ is said to be typical with respect to the type of scene $k$.
+% Si pour un environnement $k$, et une classe $j$, l'hypothèse est rejetée, la classe $j$ est alors typique de l'environnement $k$. Les classes typiques sont nommées \textbf{marqueurs sonores};
+
+If the null hypothesis is rejected, the class $j$ is said to be typical with respect to the type of environnement $k$.
 Such typical classes are called \textbf{sound markers}, in reference to the work of \cite{schafer1993soundscape}.
@@
 
-\item \emph{Representation space induced by the semantic features}: the goal is to determine if a representation space of the scenes solely based on the presence / absence of sound sources allows us to distinguish between the two types of scene.
+\item \emph{Representation space induced by the semantic features}: the goal is to determine if a representation space of the scenes solely based on the presence / absence of sound sources allows us to distinguish between the two types of scenes.
 Denoting as $S_i$ the semantic features of scene $i$, we compute the distances between all $S_i$ vectors.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Results}
@@
 For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment.
-It indeed appears that the ideal scenes were perceived significantly more pleasant ($p < 0.01$) than the non-ideal scenes.
-
-%\jfp{a la limite on pourrait ne pas écarter arbitrairement de scène et ne garder que celles pour lesquelles les iscenes sont signif.
-+ plaisante que les ni-scene.
-ca revient au même.
-c'est les test qui les écarte.
-il faudrait peut être justifier pourquoi on fait cela  : pour ne pas fausser les résultats de l'analyse et vérifier que les i scènes sont effectivement perçues + plaisantes que les ni.} \gl{Oui.
-Mais ça revient au même de toute façon}
+It indeed appears that the ideal scenes were perceived significantly more pleasant ($p<0.01$) than the non-ideal scenes.
 
@@
 Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the distributions of levels $L$, $L(E)$ and $L(T)$.
-There is a significant difference in terms of sound levels between ideal and non-ideal scenes ($L$: $p<0.01$).
+There is a significant difference in terms of sound levels between ideal and non-ideal scenes ($L$: $p<0.01$, mean deviation: -7 $dB$).
 This difference is significant for events ($L(E)$: $p<0.01$, mean deviation: -7 $dB$) and for textures ($L(T)$: $p<0.01$, mean deviation: -6 $dB$).
@@
 We also notice that this difference of sound levels is significant for both events and textures.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %Nous observons, enfin, que le niveau seul ne permet pas de clairement faire la distinction entre les différents types d'environnement.
@@
 
-However, level alone is not sufficient to fully differentiate between ideal and non-ideal scenes.
+However, sound level alone is not sufficient to fully differentiate between ideal and non-ideal scenes.
 In fact, $20\%$ of the ideal scenes have a sound levels higher than the lowest level of the non-ideal scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
@@
 \hline
-$L$            & \textbf{-0.77} ($p<0.01$) & -0.32 ($p=0.06$)  & \textbf{-0.78} ($p<0.01$)\\
-$L(E)$         & \textbf{-0.75} ($p<0.01$) & -0.20 ($p=0.24$)  & \textbf{-0.75} ($p<0.01$)\\
+$L$            & \textbf{-0.77} ($p<0.01$) & -0.32 ($p=0.06$)  & \textbf{-0.78} ($p<0.01$) \\
+$L(E)$         & \textbf{-0.75} ($p<0.01$) & -0.20 ($p=0.24$)  & \textbf{-0.75} ($p<0.01$) \\
 $L(T)$         & \textbf{-0.53} ($p<0.01$) & -0.33 ($p=0.05$)  &  -0.00 ($p=0.99$)         \\
@@
 In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated.
-Contrary to the previous test, we do not limit ourselves to a binary ideal scenes \emph{vs.} non-ideal scenes disctinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature.
+Contrary to the previous test, we do not limit ourselves to a binary ideal scenes \emph{vs.} non-ideal scenes distinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature.
 The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$.
 The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown in Table~\ref{tab:corrStructA}.
-Relationships between $\mathcal{A}_{scene}$ and the sound levels are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.
+Relationships between $\mathcal{A}_{scene}$ and the sound levels are depicted in Figures~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.
 
@@
 Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for ideal and non-ideal scenes.
-In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for non-ideal scenes ($r=-0.78$, $p<0.01$), but is not significant ($p=0.06$) for ideal scenes.
+In fact, the correlation between $L$ and $\mathcal{A}_{scene}$ remains high for non-ideal scenes ($r=-0.78$, $p<0.01$), but is not significant ($r=-0.32$, $p=0.06$) for ideal scenes.
 
@@
 When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the ideal scenes tend to be softer than the non-ideal scenes, thus allowing us to extend artificially to the ideal scenes the negative correlation observed for the non-ideal scenes.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}.
-For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (ideal scenes: $r=-0.33$, $p=0.05$, non-ideal scenes: $r=-0.00$, $p=0.99$).
+For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes ($r=-0.53$, $p<0.01$) disappears when separate scenes are considered (ideal scenes: $r=-0.33$, $p=0.05$, non-ideal scenes: $r=-0.00$, $p=0.99$).
 Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (ideal scenes tend to be softer than non-ideal scenes).
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %En résumé, en présence d'un environnement désagréable, les niveaux sonores, en particulier ceux des événements, ont un impact négatif sur l'agrément.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %Ces premiers résultats pourraient montrer qu'il existe deux modes de perception, mobilisant chacun des descripteurs indépendants, modes qui s'activent en fonction de la nature de l'environnement (i ou ni).
-%
-%Le fait que $L$ ne permet pas de caractériser l'agrément des i-scènes peut nous amener à penser que toutes les sources sonores ne contribuent pas de manière égale à la perception de l'agrément, mais que seules le niveau de certaines d'entre elles a une réelle influence.
+Le fait que $L$ ne permet pas de caractériser l'agrément des i-scènes peut nous amener à penser que toutes les sources sonores ne contribuent pas de manière égale à la perception de l'agrément, mais que seules le niveau de certaines d'entre elles a une réelle influence.
 Afin d'approfondir ce point, nous analysons, dans la section suivante, les scènes d'un point de vue sémantique, \ie~en nous intéressant à la nature des sources qui les composent.
@@
 In order to investigate further in that direction, we analyze in the next section the scenes from a semantic point of view, \ie~we take an interest in the nature of the sources they are made of.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-%\jfp{pourquoi la classe "start passage way", présente en figure A9, n'est pas présente sur ce graphe ?} \gl{Elle n'est pas parlante}
-
 %Nous analysons la composition des scènes en comptant le nombre de sujets ayant utilisé une classe de sons pour simuler un type d'environnement.
@@
 For the 36 ideal and 36 non-ideal scenes considered, results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures.
-For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$, is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.
+For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$, is used to depict classes, see~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.
 
@@
 These results confirm a fact previously observed by~\cite{raimbault2005urban,dubois2006cognitive}: the semantic nature of the sound sources play an important role in the assessment of the environment.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 As in our study, Guastavino asks the subjects to describe an environment.
-She asks them to perform this task starting only from their memories, whereas in our case subjects perform the same task using actual sound samples that they can listen to.
+But she asks them to perform this task starting only from their memories, whereas in our case subjects perform the same task using actual sound samples that they can listen to.
 The fact that subjects in our experiment are faced to the acoustic reality of the sounds for composing the environment may have decreased the socio-cultural impact.
 Other studies that considered sounds as stimuli have shown that the \emph{bus} class can have a negative influence on the assessment of the environment, see~\cite{lavandier2006contribution}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-In this section, we evaluate at which level a semantic representation of the scenes allows us to discriminate between the two types of environment.
-For this purpose, a rank 5-precision is computed on the space induced by the semantic features $S$, and for each abstraction level (see Section~\ref{sec:xp1_dataAna}).
+In this section, we evaluate at which level a semantic representation of the scenes allows us to discriminate between the two types of environments.
+For this purpose, a rank 5-precision is computed on the space induced by the semantic features $S$, and for each abstraction level (see~Section~\ref{sec:xp1_dataAna}).
 The vectors $S$ are built by using all the classes ($ET$), only the event classes ($E$), only the texture classes ($T$), only the event classes corresponding to sound markers ($E_m$), or only the event classes excluding sound markers ($E_{w/o,m}$).
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %En ce qui concerne $ET$, la $p@5$ est de $76\%$ pour le niveau d'abstraction 0, et reste supérieure à $86\%$ à partir du niveau d'abstraction 1.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %En considérant séparément $E$ et $T$, il apparaît 1) que la $p@5$ obtenue avec $E$ est similaire à celle obtenue avec $ET$, 2) que la $p@5$ obtenue avec $T$ est systématiquement inférieure d'environ $10$ à $15\%$ à celle de $E$.
@@
 As discussed in Section~\ref{sec:simscene_sampleDataSet}, it seems that humans analyze the event scenes which are composed of several sound events in a descriptive manner, \ie~by identifying the sources.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Since we have extracted the typical classes of the ideal and non-ideal scenes and verified that the distinction between them was largely dependent on the presence of these classes, we shall now investigate whether a description of the scenes only based on the sound pressure level of sound markers could characterize the perceived pleasantness, perhaps better than a globally computed sound level.
+Since we have extracted the typical classes of the ideal and non-ideal scenes and verified that the distinction between those two types of scenes was largely dependent on the presence of these classes, we shall now investigate whether a description of the scenes only based on the sound pressure level of these sound markers could characterize the perceived pleasantness, perhaps better than a globally computed sound level.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %\begin{enumerate}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %En conclusion, le niveau des ni-marqueurs a une influence négative sur l'agrément pour les ni-scènes. En revanche le niveau des i-marqueurs n’impacte pas l'agrément perçu pour les i-scènes.
@@
 To conclude, the level of non-ideal markers has a negative influence on the pleasantness for the non-ideal scenes.
-On the other hand, the level of the ideal markers does not influence the perceived pleasantness for the ideal scenes.
+On the other hand, the level of ideal markers does not influence the perceived pleasantness for the ideal scenes.
 
@@
 
-Considering the non markers classes, we can observe on the ideal scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$), see~Figures~\ref{fig:soundlevelNoisea} and~\ref{fig:soundlevelNoised}.
+Considering the non markers classes, we can observe on the ideal scenes results, a weak negative correlation for $L_b$  ($r=-52$, $p<0.01$) and $L(E)_b$ ($r=-51$, $p<0.01$), see~Figures~\ref{fig:soundlevelNoisea} and~\ref{fig:soundlevelNoised}.
 It is the first time that an objective feature allows us to define the pleasantness of ideal scenes.
@@
 Sound level of textures seems to have a negative effect for the non-ideal scenes, and no significant effect for the ideal scenes.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-On the contrary, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ for the non-ideal scenes, shows that this is indeed the absolute level that matters.
+On the contrary, the fact that we observe for the non-ideal scenes both significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$, shows that this is indeed the absolute level that matters for non ideal environments.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\item \emph{differentiating ideal and non-ideal scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ et $L(T)$), allow us to differentiate reliably between ideal and non-ideal scenes.
+\item \emph{differentiating ideal and non-ideal scenes}: the semantic features, and the global sound levels ($L$, $L(E)$ and $L(T)$), allow us to differentiate reliably between ideal and non-ideal scenes.
 The semantic description seems to be more powerful;
@@
 
-\item \emph{for non-ideal scenes}, the global levels ($L$ et $L(E)$), or the level of sound markers ($L_m$ et $L(E)_{m}$), have a negative influence on pleasantness.
+\item \emph{for non-ideal scenes}, the global levels ($L$ and $L(E)$), or the level of sound markers ($L_m$ and $L(E)_{m}$), have a negative influence on pleasantness.
 Taking into account the contribution of each of the different sources of the scene does not improve the prediction performance compared to a global analysis of the environment.
@@
 \end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-%\gl{@mathieu, pourquoi pas. Mais je pense qu'on peut laisser également le premier paragraphe, ça permettra aux reviewer de commenter qqch :). Quid de mettre les deux.}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \subsection{Objective}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-In order to investigate deeper into the relation between the pleasantness of the ideal and non-ideal scenes and the sound markers, the audio waveform of the simulated scenes are regenerated with or without the classes identified as markers.
+In order to investigate deeper into the relation between the pleasantness of the ideal and non-ideal scenes and the sound markers, the audio waveform of the simulated scenes are regenerated without the classes identified as markers.
 To do so, ideal markers are removed from the ideal scenes, and non-ideal markers are removed from the non-ideal scenes.
 To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptual test is conducted with a protocol close to the one considered in experiment 1.b.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %Si la première hypothèse est intuitive, la deuxième l'est moins.
@@
 Cette suppression aura, de surcroît, pour effet de diminuer le niveau sonore global de la scène.
-
-%Néanmoins, comme nous l'avons vu, le niveau global n'est qu'un indicateur partiel de l'agrément pour les environnements sonores idéaux.
+Néanmoins, comme nous l'avons vu, le niveau global n'est qu'un indicateur partiel de l'agrément pour les environnements sonores idéaux.
 Qui plus est, cet indicateur, lorsque qu'il décrit le niveau des i-marqueurs, impacte de manière positive la qualité de la scène.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Planning of experiment 2}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %La banque de données de stimuli compte 144 séquences de 30 secondes. Ces 144 séquences comprennent:
+
+There are 144 stimuli of 30 seconds duration. More precisely:
+
 % \begin{itemize}
@@
 
-There are 144 stimuli of 30 seconds duration. More precisely:
-
 \begin{itemize}
-\item \emph{72 scenes with markers}: the 72 scenes originally simulated by the subject of experiment 1.a where the sound classes identified as markers are kept.
+\item \emph{72 scenes with markers}: the 72 scenes originally simulated by the subjects of experiment 1.a where the sound classes identified as markers are kept.
 \item \emph{72 scenes without markers}:  72 scenes where the sound classes identified as markers are removed.
@@
 Those classes are:
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Thus, only part of the ideal markers and non-ideal markers are removed from the scenes without markers.
-
-\subsubsection*{Experiment}
+Thus, only part of the ideal and non-ideal markers are removed from the scenes without markers.
+
+\subsubsection*{Procedure}
 
@@
 
+The subjects evaluate the 144 scenes.
+The evaluation is done on a 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant).
+Before rating a scene, subjects must listen to the first 20 seconds.
+After scoring, they are free to move on to the next scene.
+
 %Pour chaque sujet, les scènes sont présentées dans un ordre aléatoire.
@@
 
-%L'expérience est prévue pour durer 1 heure. Les sujets ne connaissent pas la nature des scènes.
-
-The subjects evaluate the 144 scenes.
-The evaluation is done on a 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant).
-Before rating a scene, subjects must listen to the first 20 seconds.
-After scoring, they are free to move on to the next scene.
-
 For each subject, the scenes are presented in a random order.
@@
 
+%L'expérience est prévue pour durer 1 heure. Les sujets ne connaissent pas la nature des scènes.
+
 The experiment is scheduled to last 1 hour. The subjects do not know the nature of the scenes.
 
-\subsubsection*{Experimental protocol}
+\subsubsection*{Apparatus}
 
@@
 
+All the subjects performed the experiment on standard desktop computers with the same hardware and software configurations.
+The audio files were played in diotic conditions by semi-open headphones \emph{Beyerdynamic DT 990 Pro}.
+The output sound level was the same for all the subjects.
+
 %Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
 
-All the subjects perform the test at the same time, on the same type of computers, in a quiet environment.
-They listen to the stimuli with semi-open \emph{Beyer-Dynamic DT 990 Pro} headphones, with the sound level set to be the same for all the subjects.
-They were not allowed to talk to each other during the experiment.
+All the subjects performed the experiment simultaneously in a quiet environment. They were not allowed to talk to each other during the experiment.
 
@@
 
-A supervisor is available during the whole duration of the experiment to monitor it and answer to any question the subjects may have.
+An experimenter was available during the whole duration of the experiment in order to assist subjects and to answer questions.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Les sujets sont âgés de 22 à 61 ans (moyenne: 29.5, écart-type: 14).
-Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.Tous les sujets ont réalisé l'expérience avec succès.
-
-12 subjects perform the test (8 male, 4 female; averaging $29.5$ years of age, s.d.
+Tous les sujets sont Nantais, et vivent dans cette ville depuis deux ans ou plus.
+
+12 subjects performed the experiment (8 male, 4 female; averaging $29.5$ years of age, s.d.
 of $14$ years).
-None of them have performed experiments 1.a and 1.b.
 All the subjects had been living in Nantes, France, for at least two years at the time of the experiment and reported normal hearing.
-All the subjects performed the test successfully.
-
-\subsection{Data and analysis method}
+None of them took part in the previous experiments 1.a and 1.b.
+
+%Tous les sujets ont réalisé l'expérience avec succès.
+
+All the subjects succeeded in doing the experiment.
+
+\subsection{Data and statistical analysis}
 
@@
 To do so, we perform an analysis of variance (ANOVA).
-We consider $\mathcal{A}_{subject}$ as a dependent variable, and as independent variables, the type of environment (ideal or non-ideal) and the presence/absence of markers.
-As each subject evaluated the whole set of scenes, a two factors repeated measure ANOVA with interaction is used to evaluate whether there exist a significant difference of perceived pleasantness between the scenes with and without markers.
+We consider $\mathcal{A}_{subject}$ as a dependent variable, and the type of environment (ideal or non-ideal) as well as the presence/absence of markers as independent variables.
+As each subject evaluated the whole set of scenes, a two factors repeated measures ANOVA with interaction is used to evaluate whether there exist a significant difference of perceived pleasantness between the scenes with and without markers.
 The two independent variables are considered as \emph{within-subject} factors.
@@
 \emph{Post hoc} analyses are done using the Tukey-Kramer procedure.
-%\jfp{l'analyse est inutile a mon avis, il n'y a que 2 niveaux pour chaque facteur de l'ANOVA.} \gl{2 niveaux par facteur, mais 4 niveaux en tout}
-
-% $\mathcal{A}_{subject}$
 
@@
 
-All the significance tests are performed with critical threshold $\alpha=0.05$.
+All statistical significance tests are conducted with a critical threshold of $\alpha=0.05$
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-% {maj figure: changer am et sm pour km et rm}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Il apparaît que les réponses d'un des sujets diffèrent des autres.
-Ce dernier a évalué positivement près de la moitié des ni/am-scènes (\cf~Annexe~\ref{app:xp2} : sujet 7).
+Ce dernier a évalué positivement plus de la moitié des ni/am-scènes (\cf~Annexe~\ref{app:xp2} : sujet 7).
 Le sujet a donné à 58\% des ni/am-scènes une note supérieure à 0, contre une moyenne de 11\% pour les autres sujets.
@@
 
-Let us consider $\mathcal{A}_{subject}$ for the scenes with markers (Figure~\ref{fig:xp2_Asubject}).
-Close inspection of the answers show that one subject's judgments strongly differ from the others'.
-This subject evaluated positively close to half of the non-ideal scenes with markers (see Annex~\ref{app:xp2}, subject 7), and gave a score above 0 for 58\% of the non-ideal scenes with markers, contrary to an average of 11\% for the other subjects.
+Let us consider $\mathcal{A}_{subject}$ for the scenes with markers, see~Figure~\ref{fig:xp2_Asubject}.
+Close inspection of the answers shows that one subject's judgments strongly differ from the others'.
+This subject evaluated positively more than half of the non-ideal scenes with markers (see Annex~\ref{app:xp2}, subject 7).
+He gave a score above 0 for 58\% of the non-ideal scenes with markers, contrary to an average of 11\% for the other subjects.
 
@@
 
-In this section, we study the scores given by the subject while listening to the several types of scenes, namely the ideal scenes with markers, non-ideal scenes with markers, ideal scenes without markers and non-ideal scenes without markers (Figure~\ref{fig:xp2_Ascene}).
-The repeated measure ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the type environment (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
-
-%  $\mathcal{A}_{subject}$
+In this section, we study the scores given by the subject while listening to the different types of scenes, namely the ideal scenes with markers, non-ideal scenes with markers, ideal scenes without markers and non-ideal scenes without markers, see Figure~\ref{fig:xp2_Ascene}.
+The repeated measures ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the type environment (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-%\jfp{je crois que l'analyse post-hoc n'est pas nécessaire lorsqu'il n'y a que 2 niveaux pour les facteurs de l'ANOVA.
-les conclusions seront strictement les mêmes que pour l'ANOVA.c'est utile à partir de 3 niveaux, pour savoir quelle moyenne diffère signif.
-des autres.}
-
 The \emph{post hoc} analysis exhibits significant differences between all groups of observations, notably between the ideal scenes with markers and the ideal scenes without markers ($p<0.05$) as well as between the non-ideal scenes with markers and the non-ideal scenes without markers ($p<0.01$).
@@
 These results indicate that the removal of the markers indeed modify the perception of the scenes by the subjects, see Figure~\ref{fig:xp2A}. Our two hypotheses are thus verified :
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 %L'interaction significative montre que l'effet du type d'environnement influe sur l'effet de l'absence/présence des marqueurs.
-En effet la moyenne des écarts entre am- et sm-scènes est plus importante pour les ni-scènes ($1.1$) que pour les i-scènes ($0.5$).
+En effet la moyenne des écarts de $\mathcal{A}_{subject}$ entre les am- et sm-scènes est plus importante pour les ni-scènes ($1.1$) que pour les i-scènes ($0.5$).
 
 The significant interaction shows that the type of environment influences the effect of the presence/absence of the markers.
-Indeed, the average difference between the scenes with markers and the scenes without markers is larger for the non-ideal scenes ($1.1$) than for the ideal scenes ($0.5$).
+Indeed, the average difference of $\mathcal{A}_{subject}$ between the scenes with markers and the scenes without markers is larger for the non-ideal scenes ($1.1$) than for the ideal scenes ($0.5$).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %Les i-marqueurs ont donc bien un effet bénéfique sur la perception d'un environnement.
@@
 This strongly confirms that ideal markers do have a positive impact on the perception of an environment.
-The fact that their removal decrease $\mathcal{A}_{scene}$ indicates that it should be possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
+The fact that their removal decreases $\mathcal{A}_{scene}$ indicates that it should be possible to improve the perceived pleasantness of a given urban area by the addition of sounds commonly considered as pleasant such as \emph{bird calls}.
 Those conclusions are in line with the positive approach introduced by~\cite{schafer1977tuning}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \label{sec:conclusion}
+
 %Les expériences ont montré que la majorité des descripteurs utilisés, qu'ils soient sémantiques ou structurels, permettent de faire la distinction entre une scène idéale et une scène non-idéale.
@@
 This fact leads the authors to believe that there is little hope to find a holistic physical descriptor that can adequately account for the affective qualities of all types of sound environment.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %Ces travaux nous permettent enfin de conjecturer quant à la nature des représentations mentales des concepts «~environnement sonore urbain agréable~» (EA) et «~environnement sonore urbain désagréable~» (ED).
@@
 
-First, the fact that the semantic information (which sound source is present) and compositional information (at which level) are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the above cited concepts.
+First, the fact that the semantic information (which sound source is present) and structural information (at which level) are different for ideal and non-ideal scenes leads us to believe that these two types of information characterize the above cited concepts.
 
@@
 
-%Troisièmement, le fait que les descripteurs structurels corrélés à l'agrément diffèrent en fonction du type de scènes nous porte à croire que les caractéristiques structurelles permettant de distinguer deux instances d'un même concept abstrait lié à l'agrément ne sont pas les mêmes pour EA et ED.
-
-%Third, the fact that the compositional descriptors correlated with the perceived pleasantness change according to the type of scenes leads us to believe that the compositional characteristics that allow us to distinguish between two instances of the same abstract concept linked to the pleasantness are not the same for PE and UE.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Conclusion}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %Dans la continuité des travaux réalisés, il conviendrait de multiplier les expériences de simulation, en faisant varier les qualités affectives (calme, confortable, gênante, \etc), mais aussi en spécifiant des lieux particuliers (parc, place, rue, \etc), afin d'élaborer des corpus entiers de scènes cognitivement renseignées de paysages sonores.
 
-Future work should consider several other simulation experiments by changing the emotional qualities (quiet, comfortable, troublesome, etc.), but also by specifying specific urban locations (park, square, street, etc.), in order to provide to the scientific community an entire corpora of cognitively informed soundscapes.
+Future work should consider several other simulation experiments by changing the emotional qualities (quiet, comfortable, troublesome, etc.), but also by targeting specific urban locations (park, square, street, etc.), in order to provide to the scientific community an entire corpora of cognitively informed soundscapes.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-%Outre les possibilités déjà évoquées, le protocole de simulation présenté ici ainsi que son implémentation apporte encore dans ce cas deux avantages:
-
-Once again, besides the interesting possibilities already mentioned, the simulation protocol presented here as well as its implementation brings in this case two important advantages:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%   DEBUT RELECTURE APPROFONDIE   %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%Outre les possibilités déjà évoquées, le protocole de simulation présenté ici ainsi que son implémentation apporte encore dans ce cas trois avantages:
+
+Once again, besides the interesting possibilities already mentioned, the simulation protocol presented here as well as its implementation brings in this case three advantages:
 
@@
 % \item le simulateur peut être déployé à large échelle via internet du fait de l'architecture logicielle choisie pour son développement;
-% \item les scènes simulées peuvent être analysées sans avoir à tenir compte des différentes langues maternelles des sujets, la nature sémantique des classes de sons utilisées étant connue \emph{a priori} par l'expérimentateur et sans avoir besoin d'analyser la scène sonore pour idientifier les sources, leurs présences dans la scène étant directement exploitable.
+% \item les scènes simulées peuvent être analysées sans avoir à tenir compte des différentes langues maternelles des sujets, la nature sémantique des classes de sons utilisées étant connue \emph{a priori} par l'expérimentateur
+% \item les scènes simulées peuvent être analysées sans avoir besoin d'annoter la scène sonore pour idientifier les sources, leurs présences dans la scène étant directement exploitable.
 % \end{itemize}
@@
 \item The simulator can be deployed on a large scale via the Internet thanks to the web-based software architecture, provided that the sound databases are adapted to the cultural background of the cohort;
-\item Simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, the semantic nature of the classes of sounds used being known \emph{a priori} by the experimenter and without the need to analyze and annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
+\item Simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, the semantic nature of the sound classes used being known \emph{a priori} by the experimenter;
+\item Simulated scenes can be analyzed without the need to annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
 \end{itemize}
 
-%Bien évidemment, ces approches nécessitent d'accroître la taille des banques de sons isolés disponibles, un effort conséquent, mais nécessaire, qui contribuera grandement aux nombreux domaines de recherche ayant trait aux scènes sonores.
-
-%\nm{a ce propos, il n'est pas fait mention des sons manquants pour les subjects (cf. protocole expe de 1a);
-%ne faudrait -il pas en dire 2 mots ce qui alimenterait aussi la discussion sur la dependance de la qualite (et l'efficacite) de la simulation a  la constitution de la base de donnees de textures et d'evenemetns ?}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2018-09-04 lafaySimulatedScenePerception.tex
Relecture Mathias

@@
 
-We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene, see~\cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,szeremeta2009analysis}.
-Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene, see~\cite{gozalo2015relationship,ricciardi2015sound}.
+We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent studies demonstrate that all sound sources do not contribute equally to the perception of the sound scene, see~\cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,szeremeta2009analysis}.
+Thus, much attention is given to the specific contributions of the different sources to the notion of emotional quality of the scene, see~\cite{gozalo2015relationship,ricciardi2015sound}.
 
@@
 \begin{enumerate}
-\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non-ideal soundscapes that are considered as material for the following experiments; The resulting waveforms are available for download\footnote{Soundscapes, available online \url{https://archive.org/details/soundSimulatedUrbanScene}}.
+\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non-ideal soundscapes that are considered as material for the following experiments (the resulting waveforms are available for download\footnote{Soundscapes, available online \url{https://archive.org/details/soundSimulatedUrbanScene}});
 \item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
@@
 They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location.
-The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics.
-A lack of diversity in terms of sound sources is nonetheless mentioned by the authors that limit the strength of the outcomes given in this study.
+The authors show that the addition or removal of the sources globally follows social or semantic considerations more than their acoustical characteristics.
+A lack of diversity in terms of sound sources is nonetheless mentioned by the authors as a limiting factor to the strength of the outcomes given in this study.
 
@@
 
-The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}.
+The remaining of this paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}.
 Experiments~1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section~\ref{sec:simulation}.
@@
 
-Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user.
+Once the typology and the set of sounds are available, an important design issue is the need for a suitable way to display the sound dataset to the user.
 Most browsing tools are based on keyword indexing; however, for sensory experiments that study the objectivation of a subject's mental representations, this may be problematic as the availability of a verbal description of the sound can influence the subject's choice, and potentially induce biases in the analysis conclusions.
@@
 
-Therefore, the selection interface considered in this study is text free and designed so as to force the user to rely on listening.
+Therefore, the selection interface considered in this study is text-free and designed so as to force the user to rely on listening.
 
@@
 Figure~\ref{fig:ssf} shows the interface used for the selection of events.
-Each circle corresponds to a sound class, the lowest level of abstraction (leaves) colored in grey.
+Each circle corresponds to a sound class, with the lowest level of abstraction (leaves) colored in grey.
 The spatial location of those circles is chosen according to the hierarchical organization of the sound database: sub-classes belonging to the same class are close to each others, and so on until the user reaches the leaf classes, which are directly linked to a collection of samples.
@@
 The subject can browse the database by listening to those prototype sounds.
-The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed by~\cite{lafay2016JAES}.
+The efficiency of this interface compared to several others designs has been evaluated and the outcomes are discussed by~\cite{lafay2016JAES}.
 
@@
 Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness.
-For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
+For that, the first two experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):
 
@@
   \item  select sound classes,
-  \item  give to each of them a name,
+  \item  give each of them a name,
   \item  set the parameters of the tracks related to the selected sound classes of sounds, see Section~\ref{sec:simscene_parametre}.
@@
 
-In addition, once the two sound scenes are completed, subjects are invited to:
+In addition, once the two sound scenes are completed, the subjects are invited to:
 
@@
 
-Before starting the first simulation, a 20-minute tutorial is given to the subjects in order to familiarize them with the simulation interface and the sound database.
+Before starting the first simulation, a 20-minute tutorial is given in order to familiarize the subjects with the simulation interface and the sound database.
 The experiment is planned to last about two hours and a half, including breaks that the subjects are allowed to take.
@@
 
-Among the 44 subjects, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non-ideal scenes).
+Among the 44 subjects, 40 succeeded, producing in the end 80 simulated sound scenes (40 ideal scenes, 40 non-ideal scenes).
 The 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.
@@
 The assessment is done with a 7-point bipolar semantic scale going from -3 (non-ideal / unpleasant) to +3 (ideal / pleasant).
-Before evaluating a scene, subjects must listen to the first 20 seconds of the stimuli.
+Before evaluating a scene, the subjects must listen at least to the first 20 seconds of the stimuli.
 After the evaluation, they are free to continue to the next scene.
@@
 \begin{tabular}{c c}
-Terms                              & Acronyms                   \\
+Terms                              & Abbreviations              \\
 \hline
@@
 \vspace{0.5mm}
-\caption{Acronyms of features used in the analysis of the experiments.}
+\caption{Abbreviations of features used in the analysis of the experiments.}
 \label{tab:acronyme}
@@
 A set of features, upon which the analysis is conducted, is attached to each sound scene.
-A summary of those features (and the corresponding acronyms) is presented in Table~\ref{tab:acronyme}.
+A summary of those features (and the corresponding abbreviations) is presented in Table~\ref{tab:acronyme}.
 In order to be consistent with the evaluation of experiment 1.b, features are not computed on the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b (Section~\ref{sec:xp1b_plan}).
@@
 
-\item \emph{semantic features}: we use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene, \ie~the sound classes that are present / absent from the scene.
+\item \emph{semantic features}: we use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the classes of sounds involved in the scene, \ie~the sound classes that are present / absent from the scene.
 Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise.
@@
 \item \emph{structural features}: while \emph{SimScene} allows us to access a variety of information about the scene structure (such as the density of events), we only focus in this first study on the sound levels.
-To figure those out, we draw inspiration from the $L_{Aeq}$.
+To figure those out, we draw inspiration from the $L_{Aeq}$ measure.
 In our case, it is a scalar computed from the signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale).
@@
 
-\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the ideal and non-ideal scenes during the experiment 1.a.
-To do so, we consider if there exists significant differences between the ideal and non-ideal scenes for $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$.
+\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the ideal and non-ideal scenes during experiment~1.a.
+To do so, we consider if there exist significant differences between the ideal and non-ideal scenes for $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$.
 The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.
@@
 
-\item \emph{Influence of sound levels on perceived pleasantness}: the goal is to evaluate if the sound levels ($L$, $L(E)$ and $L(T)$) impact the perceived pleasantness.
+\item \emph{Influence of sound levels on perceived pleasantness}: the goal is to evaluate whether the sound levels ($L$, $L(E)$ and $L(T)$) affect the perceived pleasantness.
 To do so, we consider linear correlations between those features and $\mathcal{A}_{scene}$.
@@
 
-\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used in a given type of environment (ideal or non-ideal).
+\item \emph{Analysis of semantic features}: the goal is to evaluate whether specific classes are more frequently used in a given type of environment (ideal or non-ideal).
 To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}.
-With $c$ being the total number of classes used to simulate all the scenes, $c_k$ the number of classes used to simulate the scenes of a given type of environment $k$ (ideal or non-ideal), $c_j$ the number of times a class $j$ has been used to simulate all the scenes, and $c_{jk}$ the number of time a class $j$ has been used to simulate the scenes of a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
+With $c$ being the total number of classes used to simulate all the scenes, $c_k$ the number of classes used to simulate the scenes of a given type of environment $k$ (ideal or non-ideal), $c_j$ the number of times a class $j$ has been used to simulate all the scenes, and $c_{jk}$ the number of times a class $j$ has been used to simulate the scenes of a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$.
 For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:
@@
 where $\bigoplus$ is the \emph{exclusive-or} operator.
-Two scenes having similar source compositions will be close in such space.
+Two scenes having similar source compositions will be close in such a space.
 Using the Hamming distance allows us to take into account equally the presence and absence of classes.
 In order to measure the intrinsic ability of the space to discriminate between ideal and non-ideal scenes, we use a ranking metric named the precision at rank $k$ ($P@k$).
-The $P@k$ computes the precision obtained after the $k$ closest items with respect to a given seed item have been found.
-Formally, for each $s_i$ scene (considered as seed), we compute the proportion of $s_j$ scenes in the $k$ nearest neighbors of $s_i$ that share the same label as $s_i$.
+The $P@k$ computes the precision obtained after the $k$ closest items to a given seed item have been found.
+Formally, for each scene $s_i$ (considered as seed), we compute the proportion of $s_j$ scenes in the $k$ nearest neighbors of $s_i$ that share the same label as $s_i$.
 The $P@k$ is then the average of this ratio for all the items considered as search seeds.
@@
 For the V-test, considering that a large number of classes is tested, a Bonferroni correction is applied.
-For the $p$-value, if $p\geq0.05$, the value is reported.
-If $0.01\leq p<0.05$, we only report $p<0.05$, otherwise we report $p<0.01$.
+For the $p$-value, if $p\geq0.05$, the value is reported; if $0.01\leq p<0.05$, we only report $p<0.05$, otherwise we report $p<0.01$.
 
@@
 
-First, in order to ensure the coherence of the data, we check that none of the non-ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of a ideal scene.
-4 non-ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes.
+First, in order to ensure the coherence of the data, we check that none of the non-ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of an ideal scene.
+Four non-ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes.
 As a consequence, 36 ideal scenes and 36 non-ideal scenes remain for analysis.
@@
 As expected, the sound level of the sources is indeed a pleasantness indicator, as the non-ideal scenes tend to be louder.
-This results is also an outcome of a large number of related works.
+This result is also an outcome of a large number of related works.
 We also notice that this difference of sound levels is significant for both events and textures.
@@
 
-It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for ideal and non-ideal scenes.
+It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ between ideal and non-ideal scenes.
 This observation is in agreement with the results obtained by~\cite{kuwano_memory_2003}.
@@
 However, sound level alone is not sufficient to fully differentiate between ideal and non-ideal scenes.
-In fact, $20\%$ of the ideal scenes have a sound levels higher than the lowest level of the non-ideal scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
+In fact, $20\%$ of the ideal scenes have sound levels higher than the lowest level of the non-ideal scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.
 
@@
 Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived.
-Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for ideal and non-ideal scenes.
+However, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for ideal and non-ideal scenes.
 In fact, the correlation between $L$ and $\mathcal{A}_{scene}$ remains high for non-ideal scenes ($r=-0.78$, $p<0.01$), but is not significant ($r=-0.32$, $p=0.06$) for ideal scenes.
@@
 
-To sum up, for an unpleasant environment, sound levels --~especially those of events~-- negatively influence the perceived pleasantness.
+To sum up, for an unpleasant environment, sound levels --\,especially those of events\,-- negatively influence the perceived pleasantness.
 On the contrary, for a pleasant environment, none of the sound levels considered in the study seem to influence the perceived pleasantness.
@@
 We thus put forward the hypothesis that only the level of some of them can influence this perception.
-In order to investigate further in that direction, we analyze in the next section the scenes from a semantic point of view, \ie~we take an interest in the nature of the sources they are made of.
+In order to investigate further in that direction, we analyze in the next section the scenes from a semantic point of view, \ie~we take an interest in the nature of the sources they are composed of.
 
@@
 The authors interpret this by the fact that the perception of pleasantness is, among other things, due to socio-cultural factors.
-Thus, in our representation of the world, sounds of public transportation are positively connoted and tend to be more accepted than sounds of personal vehicles.
+Thus, in our representation of the world, sounds of public transportation would be positively connoted and tend to be more accepted than sounds of personal vehicles.
 
@@
 
-Regarding the sound events, 9 markers are identified for all abstraction levels.
+Regarding sound events, 9 markers are identified for all abstraction levels.
 As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are ideal scenes markers as well as the \emph{church bell} class.
@@
 In fact, according to Schafer, a sound that is identified by a person as being an important element of his/her environment, is well accepted.
-Sound markers of non-ideal scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).
+Sound markers of non-ideal scenes are classes related to construction sites (\emph{construction works}), or suggesting intense traffic (\emph{horn}, \emph{siren}).
 
@@
 
-Regarding the textures, 5 markers are identified.
-For the ideal scenes, those classes related to subdued or quiet ambiances (\emph{courtyard}, \emph{park}).
-The marker classes for the non-ideal scenes are, as for the events, related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroads}).
+Regarding textures, 5 markers are identified.
+For the ideal scenes, those are classes related to subdued or quiet ambiances (\emph{courtyard}, \emph{park}).
+The marker classes for the non-ideal scenes are, as for the events, related to construction sites (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroads}).
 
@@
 
-Concerning $ET$, the rank 5-precision is $76\%$ at abstraction level $0$ (the most abstract), and remains above $86\%$ for higher abstraction levels.
-Considering only the presence / absence of sound classes thus allows us to nicely discriminate between the two types of environments.
+Concerning $ET$, the rank 5-precision is $76\%$ at abstraction level $0$ (the most abstract), and remains above $86\%$ for subsequent abstraction levels.
+Considering only the presence / absence of sound classes thus allows us to properly discriminate between the two types of environments.
 We also notice that the less abstract (and therefore more precise) the description is, the more effective it is to predict agreement.
@@
 $S$ being a boolean vector, the smaller the dimension, the lower the amount of information it can carry.
-Despite this, it appears that the rank 5-precision obtained with $E_m$ is equal – or superior – to the ones obtained with $E$ or $ET$, although only a partial information is used in that case to describe the scenes.
+Despite this, it appears that the rank 5-precision obtained with $E_m$ is equal –\,or superior\,– to the ones obtained with $E$ or $ET$, although only a partial information is used in that case to describe the scenes.
 Reciprocally, if the sound markers are not taken into account for the description ($E_{w/o,m}$), the performance is below the one achieved when considering only the textures as features.
-Thus, most of the semantic information allowing to differentiate between ideal and non-ideal scenes is included in the markers.
+Thus, most if not all of the semantic information allowing to differentiate between ideal and non-ideal scenes is included in the markers.
 
@@
 
-On the contrary, the fact that we observe for the non-ideal scenes both significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$, shows that this is indeed the absolute level that matters for non ideal environments.
+On the contrary, the fact that we observe for the non-ideal scenes both significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$, shows that it is indeed the absolute level that matters for non ideal environments.
 
@@
 
-In order to investigate deeper into the relation between the pleasantness of the ideal and non-ideal scenes and the sound markers, the audio waveform of the simulated scenes are regenerated without the classes identified as markers.
+In order to investigate deeper into the relation between the pleasantness of ideal and non-ideal scenes and the sound markers, the audio waveforms of the simulated scenes are regenerated without the classes identified as markers.
 To do so, ideal markers are removed from the ideal scenes, and non-ideal markers are removed from the non-ideal scenes.
@@
 
-In order to create the scenes without markers with still some sound diversity and no absence of sound activity within a long period of the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}.
+In order to create the marker-less scenes with still some sound diversity and no absence of sound activity for long periods within the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}.
 Those classes are:
@@
 
-In this section, we study the scores given by the subject while listening to the different types of scenes, namely the ideal scenes with markers, non-ideal scenes with markers, ideal scenes without markers and non-ideal scenes without markers, see Figure~\ref{fig:xp2_Ascene}.
-The repeated measures ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the type environment (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
+In this section, we study the scores given by the subjects while listening to the different types of scenes, namely the ideal scenes with markers, non-ideal scenes with markers, ideal scenes without markers and non-ideal scenes without markers, see Figure~\ref{fig:xp2_Ascene}.
+The repeated measures ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the environment type (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
 
@@
 
-These results indicate that the removal of the markers indeed modify the perception of the scenes by the subjects, see Figure~\ref{fig:xp2A}. Our two hypotheses are thus verified :
+Those results indicate that the removal of the markers indeed modify the perception of the scenes by the subjects, see Figure~\ref{fig:xp2A}. Our two hypotheses are thus verified :
 
@@
 Considering sound scene synthesis or composition as part of an experimental protocol requires the availability of software resources that are simple to manipulate for subjects with little training, and possibly running on many types of hardware.
-While most processing frameworks that have been considered are developed in native language for efficiency purposes like Marsyas developed by~\cite{tzanetakis2000marsyas} or Clam developed by~\cite{amatriain2006clam}, many advanced sound manipulations can now be performed on modern web browsers using the webaudio liibrary.
+While most processing frameworks that have been considered are developed in native language for efficiency purposes like Marsyas developed by~\cite{tzanetakis2000marsyas} or Clam developed by~\cite{amatriain2006clam}, many advanced sound manipulations can now be performed on modern web browsers using the webaudio library.
 
@@
   \item the use of Webaudio javascript library is useful to design advanced experimental protocols that consider sound manipulation over the web
-  \item the soundscape model considered is useful to question important research questions within the soundscape community  but simple enough to be grasped by novice participants.
+  \item the considered soundscape model is useful to question important research questions within the soundscape community but simple enough to be grasped by novice participants.
 \end{itemize}
@@
 \begin{itemize}
-\item The simulator can be deployed on a large scale via the Internet thanks to the web-based software architecture, provided that the sound databases are adapted to the cultural background of the cohort;
-\item Simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, the semantic nature of the sound classes used being known \emph{a priori} by the experimenter;
-\item Simulated scenes can be analyzed without the need to annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
+\item The simulator can be deployed on a large scale via the Internet thanks to the web-based software architecture, provided that the sound databases are adapted to the cultural background of the test subject population;
+\item Simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, since the semantic natures of the used sound classes are known \emph{a priori} by the experimenter;
+\item Simulated scenes can be analyzed without the need to annotate the sound scene to identify the sources, since their occurrences in the simulated scene are directly available.
 \end{itemize}

### 2018-09-04 lafaySimulatedScenePerception.tex
fixing typo

@@
 Accounting for the previously detailed perceptual matters, two taxonomies are built, one for sound events and one for sound textures.
-Up to four levels of abstraction are considered from the most generic classes(level 0) to the most specific classes (level 3), leading to a taxonomy close to the one used in~\cite{Salamon14}, see Appendix~\ref{app:taxonomie}.
+Up to four levels of abstraction are considered from the most generic classes (level 0) to the most specific classes (level 3), leading to a taxonomy close to the one used in~\cite{Salamon14}, see Appendix~\ref{app:taxonomie}.
 Only three levels of abstraction are considered for the texture sounds, see Figure~\ref{fig:taxonomieT}.

### 2018-09-04 lafaySimulatedScenePerception.tex
fixing typo

@@
 
-On the contrary, the fact that we observe for the non-ideal scenes both significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$, shows that it is indeed the absolute level that matters for non ideal environments.
+On the contrary, the fact that we observe for the non-ideal scenes both significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$, shows that it is indeed the absolute level that matters for non-ideal environments.
 

### 2018-09-04 lafaySimulatedScenePerception.tex
fixing typo

@@
 The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes by~\cite{guastavino2006ideal}, \ie~on one hand, classes involving human presence and nature are prevailing in the ideal scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the non-ideal scenes.
-These results confirm a fact previously observed by~\cite{raimbault2005urban,dubois2006cognitive}: the semantic nature of the sound sources play an important role in the assessment of the environment.
+These results confirm a fact previously observed by~\cite{raimbault2005urban} and~\cite{dubois2006cognitive}: the semantic nature of the sound sources play an important role in the assessment of the environment.
 

### 2018-09-04 lafaySimulatedScenePerception.tex
fixing typo

@@
 In this section, we study the scores given by the subjects while listening to the different types of scenes, namely the ideal scenes with markers, non-ideal scenes with markers, ideal scenes without markers and non-ideal scenes without markers, see Figure~\ref{fig:xp2_Ascene}.
-The repeated measures ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the environment type (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
+The repeated measures ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the environment type ($F[1,10]=175$, $p<0.01$), of the presence/absence of markers ($F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
 

### 2018-09-04 v2/responseToReviewers.tex
fix resp

@@
 
-\title{Reply to reviewers concerning submission BR-Org-18-034 "Investigating soundscapes perception through acoustic scenes simulation"}
+\title{Reply to academic editor and reviewer concerning submission BR-Org-18-034 "Investigating soundscapes perception through acoustic scenes simulation"}
 
@@
 Following these comments, we made several changes to the article, which are summarized here.
-The next sections list our answers to each of the reviewer’s comments, with references to the revised manuscript (page, column, and paragraph) where appropriate.
+The next sections list our answers to each of their comments, with references to the revised manuscript where appropriate.
 
@@
 
-$\rightarrow$ The manuscript now follows the APA citation style. The text have been modified accordingly.
+$\rightarrow$ The manuscript now follows the APA citation style. The text has been modified accordingly.
 
@@
 
-$\rightarrow$ Those abbreviations are now in full text, except for some figures for the sake of clarity, in which case the acronyms used are detailed in the caption.
+$\rightarrow$ Those abbreviations are now in full text, except in for some figures for the sake of clarity, in which case the used acronyms are detailed in the caption.
 
@@
 The link is now in the introduction.
-The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available \url{http://soundthings.org/research/urbanSoundScape/XP2014}.
+The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available at \url{http://soundthings.org/research/urbanSoundScape/XP2014}.
 
@@
 
-\section{Answer to Reviewer 1}
+\section{Answer to Reviewer}
 
@@
 $\rightarrow$ We definitely agree.
-As now discussed in more detail in Section 6 of the manuscript, the proposed software tool is not designed for genericity but we believe that the proposed specific design strategies and use of recent software toolkits can be considered as inspirations for desiging other experimental protocols questionning soundscape perception.
+As now discussed in more detail in Section 6 of the manuscript, the proposed software tool is not designed for genericity but we believe that the proposed specific design strategies and use of recent software toolkits can be considered as inspirations for designing other experimental protocols questioning soundscape perception.
 

### 2018-09-05 lafaySimulatedScenePerception.tex
GL dernière relecture

@@
 \begin{enumerate}
-\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non-ideal soundscapes that are considered as material for the following experiments (the resulting waveforms are available for download\footnote{Soundscapes, available online \url{https://archive.org/details/soundSimulatedUrbanScene}});
+\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non-ideal soundscapes that are considered as material for the following experiments (the resulting waveforms are available for download\footnote{Soundscapes: \url{https://archive.org/details/soundSimulatedUrbanScene}});
 \item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
@@
 
-In our approach, the simulator developed for this study\footnote{simScene, available online \url{https://bitbucket.org/mlagrange/simscene}} only yields a monophonic representation of the scene, but that simplification comes with the benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are both expressive and useful for analysis.
+In our approach, the simulator developed for this study\footnote{simScene: \url{https://bitbucket.org/mlagrange/simscene}} only yields a monophonic representation of the scene, but that simplification comes with the benefit of a wider range of available sound sources and scheduling parameters in order to provide outputs which are both expressive and useful for analysis.
 
@@
 
-\emph{Simscene} is an online digital audio tool whose first version has been developed as part of the HOULE project\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
+\emph{Simscene} is an online digital audio tool whose first version has been developed as part of the HOULE project\footnote{\emph{HOULE project}: \url{houle.ircam.fr}}.
 It has been designed to run on the popular web browsers \emph{Chrome} and \emph{Firefox}.
-It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js} : \url{angularjs.org}} and the \emph{Web Audio} standard that allows the manipulation of digital audio data within the browser\footnote{\emph{Web Audio}: \url{www.w3.org/TR/webaudio}}.
+It is fully written in javascript using the \emph{angular.js} library\footnote{\emph{angular.js}: \url{angularjs.org}} and the \emph{Web Audio} standard that allows the manipulation of digital audio data within the browser\footnote{\emph{Web Audio}: \url{www.w3.org/TR/webaudio}}.
 The interface for selecting sound sources (\cf~Section~\ref{sec:simscene_ssf}) uses the popular \emph{D3.js} visualization library proposed by~\cite{d32011}.
@@
 Accounting for the previously detailed perceptual matters, two taxonomies are built, one for sound events and one for sound textures.
-Up to four levels of abstraction are considered from the most generic classes (level 0) to the most specific classes (level 3), leading to a taxonomy close to the one used in~\cite{Salamon14}, see Appendix~\ref{app:taxonomie}.
+Up to four levels of abstraction are considered from the most generic classes (level 0) to the most specific classes (level 3), leading to a taxonomy close to the one used in~\cite{Salamon14}, see~\ref{app:taxonomie}.
 Only three levels of abstraction are considered for the texture sounds, see Figure~\ref{fig:taxonomieT}.
@@
 The 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.
-The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available online\footnote{Available online \url{http://soundthings.org/research/urbanSoundScape/XP2014}}.
+The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available online\footnote{Experiment 1.a: \url{http://soundthings.org/research/urbanSoundScape/XP2014}}.
 
@@
         {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.9\columnwidth]{gfx/appendix/event_1_en}\label{fig:taxonomieEventa}} \par
-         \caption{Taxonomy of sound classes of mechanical events used for the simulation urban soundscapes with level of abstraction from left to right.}
+         \caption{Taxonomy of sound classes of mechanical events used for the simulation of urban soundscapes, with level of abstraction from left to right.}
        \label{fig:taxonomieEa}
@@
         {\includegraphics[trim={ 0 0 0cm 1cm},clip,width=.9\columnwidth]{gfx/appendix/event_2_en}\label{fig:taxonomieEventb}}
-       \caption{Taxonomy of sound classes of non-mechanical events used for the simulation urban soundscapes with level of abstraction from left to right.}
+       \caption{Taxonomy of sound classes of non-mechanical events used for the simulation of urban soundscapes, with level of abstraction from left to right.}
        \label{fig:taxonomieEb}
@@
         {\includegraphics[trim={ 0 0 7cm 1cm},clip,width=.7\columnwidth]{gfx/appendix/texture_2_en}\label{fig:taxonomieTextureb}}
-       \caption{Taxonomy of sound classes of textures used for the simulation urban soundscapes with level of abstraction from left to right.}
+       \caption{Taxonomy of sound classes of textures used for the simulation of urban soundscapes, with level of abstraction from left to right.}
        \label{fig:taxonomieT}
@@
         {\includegraphics[width=.24\linewidth]{gfx/appendix/xp4_note_12}\label{fig:xp4_note_1l}}
-        \caption{Distribution of scores given by the subjects during experiment 2 for the ideal scenes with markers (green) et non-ideal scenes with markers (red).
+        \caption{Distribution of scores given by the subjects during experiment 2 for the ideal scenes with markers (green) and the non-ideal scenes with markers (red).
 The horizontal axis is, from left to right, the 11-point bipolar semantic scale ranging from -5 (non-ideal / very unpleasant) to +5 (ideal / very pleasant).}\label{fig:xp4_note_1}

### 2018-09-17 lafaySimulatedScenePerception.tex
final typo fixes

@@
 
-%Un des objectifs premiers des études sur les paysages sonores est d'identifier les informations contenues dans l'environnement qui influent sur les sensations perçues \cite{aletta2016soundscape}.
-Questionner le lien entre sensation (\eg~calme) et environnement (\eg~scène sonore urbaine), revient à objectiver la représentation mentale que se fait un sujet d'une scène sonore en particulier (\eg~une scène sonore urbaine calme).
+%Un des objectifs premiers des études sur les paysages sonores est d'identifier les informations contenues dans l'environmentent qui influent sur les sensations perçues \cite{aletta2016soundscape}.
+Questionner le lien entre sensation (\eg~calme) et environmentent (\eg~scène sonore urbaine), revient à objectiver la représentation mentale que se fait un sujet d'une scène sonore en particulier (\eg~une scène sonore urbaine calme).
 
@@
 
-%Si on demande à un sujet d'évaluer un environnement donné suivant une échelle perceptive donnée (\eg~calme, agrément) \cite{axelsson2005soundscape,davies2013perception,cain2013development}, le gain d'information est conditionné par la nature des stimuli disponibles, qu'il s'agisse de scènes sonores enregistrées (expérience en laboratoire), ou réelles (expérience \emph{in situ}).
+%Si on demande à un sujet d'évaluer un environmentent donné suivant une échelle perceptive donnée (\eg~calme, agrément) \cite{axelsson2005soundscape,davies2013perception,cain2013development}, le gain d'information est conditionné par la nature des stimuli disponibles, qu'il s'agisse de scènes sonores enregistrées (expérience en laboratoire), ou réelles (expérience \emph{in situ}).
 Cependant, le fait que ces stimuli existent permet d'en établir une description physique précise.
@@
 
-%Si, à l'inverse, on demande à un sujet de décrire un environnement donné (\eg~décrivez une scène sonore urbaine «~calme~») \cite{guastavino2006ideal, dubois2006cognitive}, on collecte une information riche, à la fois quantitative et sémantique, de la représentation qu'il se fait de cet environnement.
+%Si, à l'inverse, on demande à un sujet de décrire un environmentent donné (\eg~décrivez une scène sonore urbaine «~calme~») \cite{guastavino2006ideal, dubois2006cognitive}, on collecte une information riche, à la fois quantitative et sémantique, de la représentation qu'il se fait de cet environmentent.
 Néanmoins, en l'absence de données sonores, on ne peut la caractériser physiquement.
@@
 Les recherches portent désormais sur l'étude des contributions spécifiques des différentes sources sur la notion de qualité affective de la scène \cite{gozalo2015relationship,ricciardi2015sound}.
-Pour de nombreuses raisons qui seront explicités tout au long de cet article, ces recherches peuvent tirés profit de ce type d'outil qui à l'avantage de permettre d'analyser séparément les influences des éléments sonores qui composent l'environnement.
+Pour de nombreuses raisons qui seront explicités tout au long de cet article, ces recherches peuvent tirés profit de ce type d'outil qui à l'avantage de permettre d'analyser séparément les influences des éléments sonores qui composent l'environmentent.
 
@@
 %Afin de répondre à ces problématiques, nous proposons d'utiliser un outil de simulation permettant à un sujet de simuler des paysages sonores à partir d'un corpus de sons enregistrés.
-Pour montrer l'intérêt de l'utilisation d'un tel outil pour questionnner la perception de l'environnement sonore, nous choisissons, comme cadre applicatif, le problème de l'agrément perçu dans les environnements sonores urbains.
+Pour montrer l'intérêt de l'utilisation d'un tel outil pour questionnner la perception de l'environmentent sonore, nous choisissons, comme cadre applicatif, le problème de l'agrément perçu dans les environmentents sonores urbains.
 Nous présentons les résultats d'une série d'expériences s'appuyant sur la simulation, et visant, chacune, à comprendre comment les différentes sources sonores qui composent une scène influent sur la perception de l'agrément:
@@
 To demonstrate the potential of the proposed approach in its ability to question the human perception of sound environments, we study in this paper the notion of perceptual pleasantness of urban environmental scenes.
-Results and outcomes of a series of three experiments that build upon the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:
+Results and outcomes of a series of experiments that build upon the use of the simulator are studied in order to better comprehend how different sound sources typically present in a urban scene impact pleasantness:
 
 % \begin{enumerate}
-% \item \emph{expérience 1.a, simulation}:  les sujets simulent les environnements idéaux/non idéaux qui serviront de stimuli pour les étapes suivantes;
+% \item \emph{expérience 1.a, simulation}:  les sujets simulent les environmentents idéaux/non idéaux qui serviront de stimuli pour les étapes suivantes;
 % \item \emph{expérience 1.b, évaluation de l'agrément}: les sujets doivent évaluer l'agrément des scènes simulées sur une échelle sémantique;
@@
 %A notre connaissance, seuls Bruce~\al \cite{bruce2009development,bruce2014effects} se sont servis de la simulation dans une apporche similaire.
-Ils proposent un outil permettant d'agir sur un environnement en ajoutant ou supprimant des sources sonores spécifiques d'une part, en modifiant le niveau sonore des sources, et leurs positions spatiales d'autre part.
-Leurs sujets recréent un un environnement urbain en manipulant un jeu de sources sonores pré-établi.
+Ils proposent un outil permettant d'agir sur un environmentent en ajoutant ou supprimant des sources sonores spécifiques d'une part, en modifiant le niveau sonore des sources, et leurs positions spatiales d'autre part.
+Leurs sujets recréent un un environmentent urbain en manipulant un jeu de sources sonores pré-établi.
 Les auteurs montrent que l'inclusion ou l'exclusion des sources dépendent plus de considérations sociales/sémantiques, que de leurs caractéristiques physiques.
@@
 
-To the best of our knowledge, only~\cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception.
+To the best of our knowledge, only~\cite{bruce2009development,bruce2014effects} considered the use of a simulator to investigate soundscape perception.
 They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location.
@@
 
-%\emph{Simscene} est un environnement de travail audio-numérique dont la première version a été développée dans le cadre du projet HOULE\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
+%\emph{Simscene} est un environmentent de travail audio-numérique dont la première version a été développée dans le cadre du projet HOULE\footnote{\emph{Projet HOULE} : \url{houle.ircam.fr}}.
 Il est prévu pour fonctionner sur les navigateurs internet \emph{Chrome} et \emph{Firefox}.
@@
 
-%Afin de créer un corpus de classes de sons de référence pour la simulation, nous réalisons une typologie des sons environnementaux urbains.
+%Afin de créer un corpus de classes de sons de référence pour la simulation, nous réalisons une typologie des sons environmententaux urbains.
 
@@
 En effet la description verbale du son, si elle est accessible au sujet, peut potentiellement influencer sa sélection.
-Dans les faits, pour construire une scène environnementale «~calme~», le sujet sélectionne a priori les sons référencés sous le vocable \emph{parc}.
+Dans les faits, pour construire une scène environmententale «~calme~», le sujet sélectionne a priori les sons référencés sous le vocable \emph{parc}.
 
@@
 
-% L'objectif est d'étudier les influences spécifiques des différentes sources sonores qui composent les environnements urbains sur la perception de l'agrément, en utilisant la simulation.
+% L'objectif est d'étudier les influences spécifiques des différentes sources sonores qui composent les environmentents urbains sur la perception de l'agrément, en utilisant la simulation.
 Pour ce faire, nous planifions nos deux premières expériences (\cf~Figure~\ref{fig:xp1_2}):
@@
 % \begin{itemize}
-% \item \emph{expérience 1.a de simulation}: au cours de cette expérience, les sujets doivent simuler des environnements sonores urbains, en utilisant l'outil de simulation \emph{Simscene} (\cf~Section~\ref{sec:simscene}).
-Chacun compose deux environnements sonores, le premier idéal/agréable et le deuxième non-idéal/désagréable.
+% \item \emph{expérience 1.a de simulation}: au cours de cette expérience, les sujets doivent simuler des environmentents sonores urbains, en utilisant l'outil de simulation \emph{Simscene} (\cf~Section~\ref{sec:simscene}).
+Chacun compose deux environmentents sonores, le premier idéal/agréable et le deuxième non-idéal/désagréable.
 % \item \emph{expérience 1.b d'évaluation}: à l'issue de la simulation, nous n'avons, de fait, qu'une connaissance binaire des propriétés affectives des scènes simulées: idéale (i) et non-idéale (ni).
@@
 % \begin{enumerate}
-% \item évaluer l'influence respective des différentes sources sur l'agrément, pour chaque type d'environnement (i ou ni);
+% \item évaluer l'influence respective des différentes sources sur l'agrément, pour chaque type d'environmentent (i ou ni);
 % \item détecter la présence de cas extrêmes ou ambigus (\emph{outlier}) dans les scènes simulées.
@@
 
-\subsection{Planning of Experiment 1.a}
+\subsection{Experiment 1.a: Methods}
 \label{sec:xp1a_plan}
@@
 
-% Les sujets doivent simuler deux environnements sonores urbains, chacune des scènes devant durer 1 minute. Ils doivent se conformer aux consignes suivantes:
+% Les sujets doivent simuler deux environmentents sonores urbains, chacune des scènes devant durer 1 minute. Ils doivent se conformer aux consignes suivantes:
 
@@
 
-% Tous commencent par simuler l'environnement idéal. Ils ne prennent connaissance de la deuxième consigne qu'à la fin de la première simulation.
+% Tous commencent par simuler l'environmentent idéal. Ils ne prennent connaissance de la deuxième consigne qu'à la fin de la première simulation.
 
@@
 % Tous les sujets réalisent l'expérience simultanément.
-Ils sont répartis de manière égale dans trois pièces identiques, toutes possédant un environnement calme.
+Ils sont répartis de manière égale dans trois pièces identiques, toutes possédant un environmentent calme.
 Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
@@
 
-\subsection{Planning of experiment 1.b}
+\subsection{Experiment 1.b: Methods}
 \label{sec:xp1b_plan}
@@
 
-% Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
+% Tous les sujets réalisent l'expérience simultanément, dans un environmentent calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
 
@@
 
-%\item \emph{analyse des descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée a un impact sur la composition des scènes en terme de sources sonores, s'il est des classes de sons plus particulièrement utilisées pour simuler un type d'environnement (i ou ni), nous utilisons le V-test.
+%\item \emph{analyse des descripteurs sémantiques}: afin d'apprécier si la distinction affective imposée a un impact sur la composition des scènes en terme de sources sonores, s'il est des classes de sons plus particulièrement utilisées pour simuler un type d'environmentent (i ou ni), nous utilisons le V-test.
 Le test est effectué pour chaque niveau d'abstraction, et séparément, pour les classes d'événements et de textures.
-Pour chaque classe $j$ et chaque type d'environnement $k$ ($k=\lbrace i,ni\rbrace$), la valeur $V_{jk}$ du V-test se calcule comme suit:
-
-%où $c$ est le nombre de classes utilisées, $c_k$ le nombre de classes utilisées pour un type d'environnement $k$, $c_j$ le nombre de classes $j$ utilisées, et $c_{jk}$ le nombre de classes $j$ utilisées pour un type d'environnement $k$.
+Pour chaque classe $j$ et chaque type d'environmentent $k$ ($k=\lbrace i,ni\rbrace$), la valeur $V_{jk}$ du V-test se calcule comme suit:
+
+%où $c$ est le nombre de classes utilisées, $c_k$ le nombre de classes utilisées pour un type d'environmentent $k$, $c_j$ le nombre de classes $j$ utilisées, et $c_{jk}$ le nombre de classes $j$ utilisées pour un type d'environmentent $k$.
 Le V-test teste l'hypothèse nulle que la proportion $\frac{c_{jk}}{c}$ ne diffère pas significativement de la proportion $\frac{c_{jk}}{c_k}$.
@@
 
-% Si pour un environnement $k$, et une classe $j$, l'hypothèse est rejetée, la classe $j$ est alors typique de l'environnement $k$. Les classes typiques sont nommées \textbf{marqueurs sonores};
-
-If the null hypothesis is rejected, the class $j$ is said to be typical with respect to the type of environnement $k$.
+% Si pour un environmentent $k$, et une classe $j$, l'hypothèse est rejetée, la classe $j$ est alors typique de l'environmentent $k$. Les classes typiques sont nommées \textbf{marqueurs sonores};
+
+If the null hypothesis is rejected, the class $j$ is said to be typical with respect to the type of environmentent $k$.
 Such typical classes are called \textbf{sound markers}, in reference to the work of \cite{schafer1993soundscape}.
@@
 
-%\item \emph{étude des espaces de représentation induits par les descripteurs sémantiques}: afin d'étudier si une représentation basée uniquement sur la présence ou l'absence de classes de sons permet de séparer les deux types d'environnement, nous considérons l'espace induit par les descripteurs sémantiques $S$.
+%\item \emph{étude des espaces de représentation induits par les descripteurs sémantiques}: afin d'étudier si une représentation basée uniquement sur la présence ou l'absence de classes de sons permet de séparer les deux types d'environmentent, nous considérons l'espace induit par les descripteurs sémantiques $S$.
 $S$ étant un vecteur booléen, nous calculons les distances entre les scènes à partir de la distance de Hamming.
@@
 Dans un deuxième temps, nous voulons nous assurer que les sujets ont bien perçu une différence d'agrément entre les i- et ni-scènes.
-Pour ce faire, nous observons l'agrément moyen de chaque sujet $\mathcal{A}_{sujet}$, calculé séparément, pour chaque type d'environnement.
+Pour ce faire, nous observons l'agrément moyen de chaque sujet $\mathcal{A}_{sujet}$, calculé séparément, pour chaque type d'environmentent.
 Il apparaît que les i-scènes ont bien été perçues comme significativement plus agréables ($p<0.01$) que les ni-scènes.
@@
 As expected, the sound level of the sources is indeed a pleasantness indicator, as the non-ideal scenes tend to be louder.
-This result is also an outcome of a large number of related works.
+This result is also an outcome of a large number of related studies.
 We also notice that this difference of sound levels is significant for both events and textures.
@@
 Cette observation fait écho aux résultats obtenus par Kuwano~\al~\cite{kuwano_memory_2003}.
-Au cours de leur expérience, les auteurs demandent à leurs sujets d'évaluer une série d'environnements sonores de manière globale, dans un premier temps, puis d'en évaluer le niveau aux instants où chacun identifie une source sonore.
+Au cours de leur expérience, les auteurs demandent à leurs sujets d'évaluer une série d'environmentents sonores de manière globale, dans un premier temps, puis d'en évaluer le niveau aux instants où chacun identifie une source sonore.
 L'étude montre qu'il n'y a pas de différences significatives entre les jugements globaux et les moyennes des jugements instantanés.
@@
 
-%Nous observons, enfin, que le niveau seul ne permet pas de clairement faire la distinction entre les différents types d'environnement.
+%Nous observons, enfin, que le niveau seul ne permet pas de clairement faire la distinction entre les différents types d'environmentent.
 En effet, $20\%$ des i-scènes ont un niveau supérieur au niveau minimal des ni-scènes, alors qu'il n'y a pas de recouvrement, si l'on considère l'agrément perçu $\mathcal{A}_{scene}$.
@@
 %\item permet de finement caractériser l'agrément perçu des ni-scènes,
-%\item n'est pas un indicateur pertinent de l'agrément perçu pour des environnements a priori agréables.
+%\item n'est pas un indicateur pertinent de l'agrément perçu pour des environmentents a priori agréables.
 %\end{itemize}
@@
 On peut penser que la corrélation négative émergeant au niveau l'ensemble est un artefact résultant du fait que le niveau des textures des i-scènes a tendance à être plus bas que celui des ni-scènes.
-Ainsi, si les événements sonores conservent une certaine capacité de prédiction de l'agrément pour les ni-scènes, le niveau des textures n'apporte, lui, que peu d'informations, quel que soit l'environnement.
+Ainsi, si les événements sonores conservent une certaine capacité de prédiction de l'agrément pour les ni-scènes, le niveau des textures n'apporte, lui, que peu d'informations, quel que soit l'environmentent.
 
@@
 Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (ideal scenes tend to be softer than non-ideal scenes).
-Thus, while sound event levels maintain a relative ability to predict the pleasantness of the non-ideal scenes, texture levels do not bring much information, whatever the type of environment is.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%En résumé, en présence d'un environnement désagréable, les niveaux sonores, en particulier ceux des événements, ont un impact négatif sur l'agrément.
-En présence d'un environnement agréable, en revanche, aucun des descripteurs structurels considérés ici ne semble influer sur la perception de l'agrément.
+Thus, while sound event levels maintain a relative ability to predict the pleasantness of the non-ideal scenes, texture levels do not provide much information, independent of the environment.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIN RELECTURE APPROFONDIE    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%En résumé, en présence d'un environmentent désagréable, les niveaux sonores, en particulier ceux des événements, ont un impact négatif sur l'agrément.
+En présence d'un environmentent agréable, en revanche, aucun des descripteurs structurels considérés ici ne semble influer sur la perception de l'agrément.
 
@@
 
-%Ces premiers résultats pourraient montrer qu'il existe deux modes de perception, mobilisant chacun des descripteurs indépendants, modes qui s'activent en fonction de la nature de l'environnement (i ou ni).
+%Ces premiers résultats pourraient montrer qu'il existe deux modes de perception, mobilisant chacun des descripteurs indépendants, modes qui s'activent en fonction de la nature de l'environmentent (i ou ni).
 Le fait que $L$ ne permet pas de caractériser l'agrément des i-scènes peut nous amener à penser que toutes les sources sonores ne contribuent pas de manière égale à la perception de l'agrément, mais que seules le niveau de certaines d'entre elles a une réelle influence.
@@
 
-%Nous analysons la composition des scènes en comptant le nombre de sujets ayant utilisé une classe de sons pour simuler un type d'environnement.
+%Nous analysons la composition des scènes en comptant le nombre de sujets ayant utilisé une classe de sons pour simuler un type d'environmentent.
 Les résultats sont présentés à la figure~\ref{fig:soundsourcea} pour les événements, et à la figure~\ref{fig:soundsourceb} pour les textures.
@@
 %Nous observons une différence notable dans le choix des classes entre les i- et ni-scènes.
-La répartition des classes est très proche de celle obtenue dans une étude similaire sur les environnements sonores urbains idéaux \cite{guastavino2006ideal}, \ie~les classes suggérant la présence humaine et la nature sont très présentes dans les i-scènes, a contrario, les classes désignant des sons mécaniques et/ou de travaux sont principalement utilisées pour les ni-scènes.
+La répartition des classes est très proche de celle obtenue dans une étude similaire sur les environmentents sonores urbains idéaux \cite{guastavino2006ideal}, \ie~les classes suggérant la présence humaine et la nature sont très présentes dans les i-scènes, a contrario, les classes désignant des sons mécaniques et/ou de travaux sont principalement utilisées pour les ni-scènes.
 
@@
 Cependant les classes \emph{transports publics} sont également bien présentes dans les ni-scènes, plus que les classes \emph{voiture} ou \emph{camion} par exemple.
-La classe \emph{transports publics} ne peut donc pas être considérée comme typique d'un environnement sonore urbain idéal.
+La classe \emph{transports publics} ne peut donc pas être considérée comme typique d'un environmentent sonore urbain idéal.
 
@@
 %Cette différence peut s'expliquer par la nature des deux protocoles expérimentaux utilisés.
-Comme nous l'avons fait, Guastavino demande à ses sujets de décrire un environnement.
+Comme nous l'avons fait, Guastavino demande à ses sujets de décrire un environmentent.
 Mais ces derniers travaillent de mémoire, alors que nos propres sujets disposent de supports sonores.
-Le fait que nos sujets soient confrontés à la réalité acoustique des sons, pour recréer leurs environnements, peut avoir pour effet de diminuer l'impact du contexte socio-culturel.
-D'autres études utilisant des sons comme stimuli montrent que la classe \emph{bus} peut avoir un effet négatif sur l'appréciation de l'environnement \cite{lavandier2006contribution}.
+Le fait que nos sujets soient confrontés à la réalité acoustique des sons, pour recréer leurs environmentents, peut avoir pour effet de diminuer l'impact du contexte socio-culturel.
+D'autres études utilisant des sons comme stimuli montrent que la classe \emph{bus} peut avoir un effet négatif sur l'appréciation de l'environmentent \cite{lavandier2006contribution}.
 
@@
 
-%Nous avons mis en évidence que, qualitativement, la composition des sources sonores des scènes diffère selon les types d'environnement (i ou ni).
-Nous essayons de voir maintenant si, parmi ces classes, certaines sont typiques d'un environnement en particulier.
+%Nous avons mis en évidence que, qualitativement, la composition des sources sonores des scènes diffère selon les types d'environmentent (i ou ni).
+Nous essayons de voir maintenant si, parmi ces classes, certaines sont typiques d'un environmentent en particulier.
 Pour ce faire, nous utilisons le V-test (\cf~Section~\ref{sec:xp1_dataAna}), en considérant séparément chaque niveau d'abstraction.
@@
 Comme la figure~\ref{fig:soundsource} le laissait présager, les classes relatives à la présence humaine (\emph{pas homme béton}, \emph{sonnette vélo}), et à la nature (\emph{animaux, oiseaux}, \emph{chants d'oiseaux}) sont des marqueurs de i-scènes.
-Nous notons également la présence de la classe \emph{cloche} dans les marqueurs d'un environnement idéal.
+Nous notons également la présence de la classe \emph{cloche} dans les marqueurs d'un environmentent idéal.
 Ce fait est possiblement dû au \emph{background} socio-culturel des sujets, dans leur grande majorité, des citoyens européens.
-En effet, selon Schafer, un son reconnu par un individu comme faisant partie intégrante de son environnement est bien accepté.
+En effet, selon Schafer, un son reconnu par un individu comme faisant partie intégrante de son environmentent est bien accepté.
 Les marqueurs de ni-scènes sont des classes faisant référence à des sons de travaux (\emph{travaux}), ou suggérant un trafic dense (\emph{klaxon}, \emph{sirène}).
@@
 Pour représenter un trafic désagréable, les sujets ont porté leurs choix sur les classes \emph{klaxon} et \emph{sirène}.
-On peut supposer que les sons isolés de véhicules sont compris comme faisant partie intégrante de l'environnement urbain, et ne sont donc pas particulièrement associés à un environnement désagréable.
+On peut supposer que les sons isolés de véhicules sont compris comme faisant partie intégrante de l'environmentent urbain, et ne sont donc pas particulièrement associés à un environmentent désagréable.
 
@@
 
-%Dans cette partie, nous évaluons la capacité d'une représentation sémantique à séparer les deux types d'environnement.
+%Dans cette partie, nous évaluons la capacité d'une représentation sémantique à séparer les deux types d'environmentent.
 Pour ce faire, nous calculons une précision au rang 5 ($p@5$) sur l'espace induit par les descripteurs sémantiques $S$, et ce pour chaque niveau d'abstraction (\cf~Section~\ref{sec:xp1_dataAna}).
@@
 %En ce qui concerne $ET$, la $p@5$ est de $76\%$ pour le niveau d'abstraction 0, et reste supérieure à $86\%$ à partir du niveau d'abstraction 1.
-Ces résultats confirment qu'il est possible de clairement distinguer les deux types d'environnement en se basant seulement sur la présence ou l'absence des classes de sons.
-Nous notons également que, plus le niveau d'abstraction est élevé, plus la capacité de séparer les environnements est importante.
+Ces résultats confirment qu'il est possible de clairement distinguer les deux types d'environmentent en se basant seulement sur la présence ou l'absence des classes de sons.
+Nous notons également que, plus le niveau d'abstraction est élevé, plus la capacité de séparer les environmentents est importante.
 En d'autres termes, plus nous sommes précis dans notre description de la composition des scènes, plus nous sommes à même d'établir une distinction claire entre les i- et ni-scènes.
@@
 %En considérant séparément $E$ et $T$, il apparaît 1) que la $p@5$ obtenue avec $E$ est similaire à celle obtenue avec $ET$, 2) que la $p@5$ obtenue avec $T$ est systématiquement inférieure d'environ $10$ à $15\%$ à celle de $E$.
-Ces résultats indiquent que l'information sémantique permettant de séparer les deux environnements est principalement portée par les événements.
+Ces résultats indiquent que l'information sémantique permettant de séparer les deux environmentents est principalement portée par les événements.
 Ces résultats font, par ailleurs, écho aux travaux de  \cite{maffiolo_caracterisation_1999}, qui montrent que nous analysons de manière descriptive (en identifiant les sources) les scènes événementielles, \ie~composées d'événements sonores (\cf~Section~\ref{sec:simscene_sampleDataSet}).
@@
 %\begin{enumerate}
-%\item contrairement à ce que nous avions constaté avec les descripteurs structurels, une description sémantique de la composition des scènes, en terme de présence/absence de sources sonores, permet de bien distinguer les deux types d'environnement (i ou ni);
+%\item contrairement à ce que nous avions constaté avec les descripteurs structurels, une description sémantique de la composition des scènes, en terme de présence/absence de sources sonores, permet de bien distinguer les deux types d'environmentent (i ou ni);
 %\item l'information sémantique est majoritairement portée par les classes d'événements sonores;
@@
 
-%Maintenant que nous avons isolé les classes typiques des i- et ni-scènes, et vérifié que la distinction entre ces environnements dépendait de la présence de ces classes, il reste à voir si une description structurelle des scènes, basée uniquement sur ces marqueurs sonores, permet de caractériser l'agrément perçu, mieux qu'une description structurelle globale.
+%Maintenant que nous avons isolé les classes typiques des i- et ni-scènes, et vérifié que la distinction entre ces environmentents dépendait de la présence de ces classes, il reste à voir si une description structurelle des scènes, basée uniquement sur ces marqueurs sonores, permet de caractériser l'agrément perçu, mieux qu'une description structurelle globale.
 
@@
 % En considérant maintenant les classes non marqueurs (\cf~Figures~\ref{fig:soundlevelNoisea} et~\ref{fig:soundlevelNoised}), nous remarquons, sur les i-scènes, une corrélation négative modérée/faible pour $L_b$  ($r=-52$, $p<0.01$) et $L(E)_b$ ($r=-51$, $p<0.01$).
-C'est la première fois qu'un indicateur objectif nous permet de préciser l'agrément des environnements agréables.
-Ceci nous amène à conclure que le niveau des classes de sons n'étant pas typiques d'un environnement agréable a un impact négatif sur l'agrément.
+C'est la première fois qu'un indicateur objectif nous permet de préciser l'agrément des environmentents agréables.
+Ceci nous amène à conclure que le niveau des classes de sons n'étant pas typiques d'un environmentent agréable a un impact négatif sur l'agrément.
 
@@
 Dans le cas des i-scènes, ce n'est donc pas le niveau absolu des marqueurs qui importe, mais leur niveau relatif, par rapport aux autres sons qui composent la scène.
-On observe donc pour les environnements idéaux un double mécanisme perceptif:
+On observe donc pour les environmentents idéaux un double mécanisme perceptif:
 
@@
 La description sémantique semble être plus performante;
-%\item \emph{événements ou textures}: que ce soit pour les descripteurs sémantiques ou structurels, ce sont majoritairement les événements qui permettent de distinguer les deux types d'environnement, les textures n'apportant, au mieux, qu'une information limitée;
-%\item \emph{prédire l'agrément}: si l'on considère les corrélations entre les descripteurs structurels et l'agrément, il semble que la manière de percevoir la qualité de l'environnement diffère en fonction de la nature de ce dernier (i ou ni).
+%\item \emph{événements ou textures}: que ce soit pour les descripteurs sémantiques ou structurels, ce sont majoritairement les événements qui permettent de distinguer les deux types d'environmentent, les textures n'apportant, au mieux, qu'une information limitée;
+%\item \emph{prédire l'agrément}: si l'on considère les corrélations entre les descripteurs structurels et l'agrément, il semble que la manière de percevoir la qualité de l'environmentent diffère en fonction de la nature de ce dernier (i ou ni).
 Il n'apparaît pas envisageable de considérer un même jeu de descripteurs pour prédire, à la fois, l'agrément des i-scènes, et l'agrément des ni-scènes :
@@
 %\item \emph{pour les ni-scènes}, ce sont le niveau global ($L$ et $L(E)$), ou le niveau des marqueurs sonores ($L_m$ et $L(E)_{m}$), qui impactent négativement l'agrément.
-On note ici que prendre en compte les contributions de différentes sources n'améliore pas la capacité de prédiction de l'agrément, par rapport à une analyse holistique de l'environnement.
+On note ici que prendre en compte les contributions de différentes sources n'améliore pas la capacité de prédiction de l'agrément, par rapport à une analyse holistique de l'environmentent.
 
@@
 Le cerveau adapte sa manière de traiter l'information (résumé statistique pour les textures, description fine pour les événements) suite à une prise de décision antérieure quant à la nature du stimulus (à savoir «~est-ce un événement ou une texture ?~»).
-De la même manière, les indicateurs actifs dans le jugement de l'agrément dépendent, eux aussi, d'une identification préalable de la nature hédonique globale de l'environnement (idéale ou non idéale).
+De la même manière, les indicateurs actifs dans le jugement de l'agrément dépendent, eux aussi, d'une identification préalable de la nature hédonique globale de l'environmentent (idéale ou non idéale).
 
@@
 
-%L'expérience précédente a montré que, parmi les classes de sons peuplant le monde sonore, celles regroupant les marqueurs sont caractéristiques de certains types d'environnement.
-Ces marqueurs sonores semblent avoir un impact particulier sur la perception de leurs environnements.
+%L'expérience précédente a montré que, parmi les classes de sons peuplant le monde sonore, celles regroupant les marqueurs sont caractéristiques de certains types d'environmentent.
+Ces marqueurs sonores semblent avoir un impact particulier sur la perception de leurs environmentents.
 C'est ce dernier point qui est étudié dans cette expérience.
@@
 %Si la première hypothèse est intuitive, la deuxième l'est moins.
-En effet, il n’apparaît pas évident que la suppression des i-marqueurs, bien que s'agissant de sons positivement connotés, diminue la qualité globale d'un environnement.
+En effet, il n’apparaît pas évident que la suppression des i-marqueurs, bien que s'agissant de sons positivement connotés, diminue la qualité globale d'un environmentent.
 Cette suppression aura, de surcroît, pour effet de diminuer le niveau sonore global de la scène.
-Néanmoins, comme nous l'avons vu, le niveau global n'est qu'un indicateur partiel de l'agrément pour les environnements sonores idéaux.
+Néanmoins, comme nous l'avons vu, le niveau global n'est qu'un indicateur partiel de l'agrément pour les environmentents sonores idéaux.
 Qui plus est, cet indicateur, lorsque qu'il décrit le niveau des i-marqueurs, impacte de manière positive la qualité de la scène.
@@
 
-\subsection{Planning of experiment 2}
+\subsection{Experiment 2: Methods}
 
@@
 
-%Tous les sujets réalisent l'expérience simultanément, dans un environnement calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
+%Tous les sujets réalisent l'expérience simultanément, dans un environmentent calme. Ils n'ont pas le droit de s'adresser la parole pendant l'expérience.
 
@@
 Pour ce faire, nous utilisons l'analyse de variance.
-Nous considérons, comme variable dépendante, $\mathcal{A}_{sujet}$, et, comme variables indépendantes, le type d'environnement (i/ni), et la présence/absence de marqueurs (am/sm).
+Nous considérons, comme variable dépendante, $\mathcal{A}_{sujet}$, et, comme variables indépendantes, le type d'environmentent (i/ni), et la présence/absence de marqueurs (am/sm).
 Chaque sujet devant évaluer la totalité des stimuli, une ANOVA à mesures répétées à deux facteurs est utilisée afin vérifier s'il existe des différences significatives d'agrément perçu.
@@
 %Dans cette section nous étudions comment les sujets ont perçu les différents types de scènes, nommément: i/am-, ni/am-, i/sm- et ni/sm-scène.
-L'ANOVA à mesures répétées pratiquée sur $\mathcal{A}_{sujet}$ montre un effet significatif du type d'environnement (i/ni: $F[1,10]=175$, $p<0.01$), de la présence/absence des marqueurs (am/sm: $F[1,10]=7$, $p<0.05$), ainsi que de l'interaction entre les deux facteurs ($F[1,10]=67$, $p<0.01$).
+L'ANOVA à mesures répétées pratiquée sur $\mathcal{A}_{sujet}$ montre un effet significatif du type d'environmentent (i/ni: $F[1,10]=175$, $p<0.01$), de la présence/absence des marqueurs (am/sm: $F[1,10]=7$, $p<0.05$), ainsi que de l'interaction entre les deux facteurs ($F[1,10]=67$, $p<0.01$).
 
@@
 
-%L'interaction significative montre que l'effet du type d'environnement influe sur l'effet de l'absence/présence des marqueurs.
+%L'interaction significative montre que l'effet du type d'environmentent influe sur l'effet de l'absence/présence des marqueurs.
 En effet la moyenne des écarts de $\mathcal{A}_{subject}$ entre les am- et sm-scènes est plus importante pour les ni-scènes ($1.1$) que pour les i-scènes ($0.5$).
@@
 
-%Les i-marqueurs ont donc bien un effet bénéfique sur la perception d'un environnement.
+%Les i-marqueurs ont donc bien un effet bénéfique sur la perception d'un environmentent.
 Le fait que leur suppression diminue $\mathcal{A}_{scene}$ montre clairement qu'il est possible d'améliorer la qualité sonore d'un lieu en ajoutant des sons bien acceptés comme \emph{oiseau}.
@@
 
-This series of three experiments showed that most of the descriptors used in this study, be they of a semantic or acoustic nature, allow us to distinguish between an ideal scene and a non-ideal one.
+This series of experiments showed that most of the descriptors used in this study, be they of a semantic or acoustic nature, allow us to distinguish between an ideal scene and a non-ideal one.
 
@@
 
-%Ces résultats tendent à confirmer que la perception des qualités d'une scène dépend avant tout des sources sonores qui la composent, les caractéristiques structurelles mobilisées dans le processus perceptif semblant varier d'une source à l'autre, et d'un type d'environnement à l'autre.
-Ce fait montre qu'il est illusoire d'envisager qu'un descripteur physique holistique puisse rendre compte, de manière pertinente, des qualités affectives de tous types d'environnement.
+%Ces résultats tendent à confirmer que la perception des qualités d'une scène dépend avant tout des sources sonores qui la composent, les caractéristiques structurelles mobilisées dans le processus perceptif semblant varier d'une source à l'autre, et d'un type d'environmentent à l'autre.
+Ce fait montre qu'il est illusoire d'envisager qu'un descripteur physique holistique puisse rendre compte, de manière pertinente, des qualités affectives de tous types d'environmentent.
 
@@
 
-%Cet état de fait peut potentiellement influer sur les stratégies à adopter pour améliorer la qualité de l’environnement sonore:
+%Cet état de fait peut potentiellement influer sur les stratégies à adopter pour améliorer la qualité de l’environmentent sonore:
 
@@
 
-%Ces travaux nous permettent enfin de conjecturer quant à la nature des représentations mentales des concepts «~environnement sonore urbain agréable~» (EA) et «~environnement sonore urbain désagréable~» (ED).
-
-Finally, this work allows us to conjecture as to the nature of the mental representations of the concepts "pleasant urban sound environment" and "unpleasant urban sound environment" .
+%Ces travaux nous permettent enfin de conjecturer quant à la nature des représentations mentales des concepts «~environmentent sonore urbain agréable~» (EA) et «~environmentent sonore urbain désagréable~» (ED).
+
+The present results allow us to conjecture as to the nature of the mental representations of the concepts "pleasant urban sound environment" and "unpleasant urban sound environment" .
 
@@
 
-Closer to our sound model is the Tapestrea framework proposed by~\cite{misra2007musical} that allows the manipulation of sounds trough wavelet and sinusoidal modeling for acousmatic composition or sound
+Closer to our sound model is the Tapestrea framework proposed by~\cite{misra2007musical} that allows the manipulation of sounds through wavelet and sinusoidal modeling for acoustic composition or sound
 design.
@@
 \begin{itemize}
-  \item the use of Webaudio javascript library is useful to design advanced experimental protocols that consider sound manipulation over the web
+  \item the use of Webaudio javascript library is useful to design advanced experimental protocols that consider sound manipulation over the Internet
   \item the considered soundscape model is useful to question important research questions within the soundscape community but simple enough to be grasped by novice participants.
@@
 
-%Au terme de notre étude, nous pensons que la simulation est un outil dont le développement pourrait permettre aux décideurs en matière d'urbanisme d'interroger toute une communauté sur ses représentations propres des environnements sonores auxquels elle est exposée, et, pourquoi pas, sur les représentations des environnements sonores auxquels elle voudrait être exposée.
-
-The outcomes of the three experiments described in this paper demonstrate the usefulness of considering a dedicated simulation tool such as \emph{simScene} in order to scientifically question the perception of soundscapes in an innovative way.
+%Au terme de notre étude, nous pensons que la simulation est un outil dont le développement pourrait permettre aux décideurs en matière d'urbanisme d'interroger toute une communauté sur ses représentations propres des environmentents sonores auxquels elle est exposée, et, pourquoi pas, sur les représentations des environmentents sonores auxquels elle voudrait être exposée.
+
+The outcomes of the these experiments described in this paper demonstrate the usefulness of considering a dedicated simulation tool such as \emph{simScene} in order to scientifically question the perception of soundscapes in an innovative way.
 We also believe that its wider usage could enable urban planning decision-makers to question an entire community about its own representations of the sound environment to which they are exposed, and about the representations of the sound environments to which they would like to be exposed to.
@@
 %Enfin, l'on pourrait encore étudier l'influence des contextes socio-culturels sur la perception.
-Dans les faits, si le son de cloche est le plus souvent un marqueur d'environnement de qualité pour un occidental, cela ne se vérifie pas nécessairement auprès de sujets de culture orientale, moyen-orientale ou autre.
+Dans les faits, si le son de cloche est le plus souvent un marqueur d'environmentent de qualité pour un occidental, cela ne se vérifie pas nécessairement auprès de sujets de culture orientale, moyen-orientale ou autre.
 
